### 2014-11-13 ms/ms.tex
First commit

\documentclass[11pt, oneside]{article}

%
% Packages
%

\usepackage{geometry}
\geometry{letterpaper}
\usepackage[parfill]{parskip}    			% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{topcapt}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{natbib}

%
% Title and authors
%

\title{(working title) Local Physiological Adaptation in \textit{Mimulus cardinalis}}
\author{Christopher D. Muir and Amy L. Angert}
%\date{}							% Activate to display a given date or no date

%
% Start document
%

\begin{document}
\maketitle

\section*{Abstract}

\section*{Introduction}

% What's the big picture (this is TOO big):
It is taken for granted that genes are the foundation of evolution. If we get to the genetic basis of adaptive evolution, then we have understood the process at its deepest level. I believe this view is misguided. Genetic change is, to the best of our knowledge, the basis of evolution, but this does not imply that by understanding genetic variation we can predict phenotypic evolution. Rather, variation and constraint on phenotypic evolution are often determined by what goes on outside. Organisms must survive and reproduce in the real world of physical laws and interactions with other organisms. Physiology, the study of organismal function, tells us about the variation and constraint on ways in which organisms can meet these challenges. ...

% this paper. good start!
Local adaptation is one of the most ubiquitous observations in evolutionary ecology. In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection. Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species. To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...

1. `intrinsic' physiological variation
2. variation in stress response
3. what ecological factors explain divergence (i.e. putative selective agents underlying local adaptation)

 % one paragraph general review of physiology, environment, and local adaptation

 % one paragraph review of physiology, environment, and local adaptation in Mimulus

 % one paragraph wrapping up, motivating this study

\section*{Methods}

\subsection*{Population Selection}

I chose 16 populations from throughout the range of \textit{M. cardinalis} (Table 1) that have been previously studied. Seeds were collected in the field [Amy - explain seed collection].

\begin{table}[htbp]
   \centering
   \topcaption{Table 1: 16 Focal populations}
   \begin{tabular}{@{} lllllll @{}}
      \toprule
      Name    & Region & Demo? & Pop gen? & Lat & Lon & Alt \\
      \midrule
	Hauser Creek						& South Margin		& yes	& yes	& 32.65822	& -116.53235	& 892 \\
	Cottonwood Creek					& South Margin		& yes	& no		& 32.80122	& -116.50194	& 1206 \\
	Sweetwater River / Cuyamaca Rancho	& South Margin		& yes	& yes	& 32.89928	& -116.5849	& 1223 \\
	Grade Road - to Palomar Mountain SP	& South Margin		& no		& no		& 33.31392	& -116.87129	& 1500 \\
	Whitewater Canyon					& Transverse		& yes	& yes	& 33.99329	& -116.66267	& 696 \\
	Mill Creek							& Transverse		& yes	& no		& 34.07808	& -116.87558	& 1992 \\
	West Fork Mojave River				& Transverse		& yes	& no		& 34.28425	& -117.37539	& 1087 \\
	North Fork Middle Tule				&				& yes	&		& 36.20081	& -118.65092	& 1284 \\
	Paradise Creek						& South Sierras		& yes	& yes	& 36.51776	& -118.75877	& 887 \\
	Redwood Creek					& South Sierras		& yes	& yes	& 36.69096	& -118.90961	& 1683.72 \\
	Wawona							& Central Sierras	& yes	& yes	& 37.539		& -119.654	& 1208 \\
	Rainbow Pool (RP)					& Central Sierras	& yes	& no		& 37.8188		& -120.00743	& 862.2792 \\
	Middle Yuba (Oregon Creek)			& North Sierras		& yes	& yes	& 39.39442	& -121.08302	& 425.196 \\
	Little Jameson						& North Sierras		& yes	& yes	& 39.74298	& -120.70401	& 1592 \\
	Deep Creek						& North Coast		& yes	& yes	& 41.66546	& -123.11341	& 694 \\
	Rock Creek (North Fork Umpqua)		& North Margin		& yes	& yes	& 43.37375	& -122.9575	& 311.2008 \\
	\bottomrule
   \end{tabular}
\end{table}

\subsection*{Plant propagation}

% Spring 2014 cohort
On 14 April, 2014, 3-5 seeds per family were sown directly on sand (Quikrete Play Sand, Georgia, USA) watered to field capacity in RLC4 Ray Leach cone-tainers placed in RL98 98-well trays (Stuewe \& Sons, Inc., Oregon, USA). We used pure sand both to facilitate root-washing and because \textit{M. cardinalis} typically grows in sandy, riparian soils (A. Angert, pers. obs.). Two jumbo-sized cotton balls at the bottom of cone-tainers prevented sand from washing out. Cone-tainers were continuously bottom-watered during germination by placing them in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) filled part way with water, placed on benches in greenhouses at the University British Columbia campus Vancouver, Canada (LAT LON of UBC). Misters thoroughly wetted the top of the sand every two hours during the day. Most seeds germinated between 1 and 2 weeks, but we allowed 3 weeks before transferring seedlings to growth chambers. Germination was recorded daily from one to two weeks after sowing, and every few days thereafter. On 5 May (21 days after sowing), seedlings were transferred to one of two MODEL Growth Chambers (Conviron, Manitoba, Canada). We thinned seedlings to one plant per cone-tainer, leaving the center-most, largest plant. Where possible, extra seedlings were transplanted to cone-tainers that did not have any germinants. In total X\% of cone-tainers had at least one germinant. After transplanting, 743 of 768 (96.7\%) had plants that could be used in the experiment. We allowed one week at constant, non stressful conditions (20 degree day, 16 degree night) for plants to acclimate to growth chambers before starting treatments.

\subsection*{Treatments}

To do this, I will use 4 watering treatments on two populations on two soils. The watering treatments are: constant water, constant level/decreasing frequency, decreasing level/constant frequency, no water. I will monitor soil water potential in pots with no plants. Basically, I'm just trying to figure out whether one or another method is easier or more effective, and if I see any obvious response from plants in term of growth.

\subsubsection*{Temperature}

We simulated typical growing season (June and July) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (High Temp) and Little Jameson (Low Temp). We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA (CITE). Daily temperatures from ClimateWNA are usually highly correlated with the air temperature recorded from data loggers in the field at these sites (A. Angert, unpub. data). Hence, the ClimateWNA temperature profiles are likely to similar to actual thermal regimes experienced by \textit{M. cardinalis} in nature. To create realistic temperature regimes, we calculated the mean temperature trend from June to July using LOESS \citep{Cleveland_etal_1992}. The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ). The `VARselect' function in the vars package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals. We fit and simulated from the VAR(2) using the package dse \citep{Gilbert_2006?} in R. Simulated data closely resembled the autocorrelation and between-site correlation of the actual data. From simulated mean temperature, we next selected minimum and maximum daily temperatures. Mean, min, and max temperature were highly correlated at both sites. We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models. For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature. During the middle of the day, chamber temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.

\subsubsection*{Water}

			Wet: daily/constant irrigation with cooled water
			Dry: start at week \#, gradually decrease watering frequency/level (see Nancy Emery's protocols)

			Are there any data from California about soil water content through the season?

\subsection*{Monitoring environment}

		Cross check with Poorter paper \\
		Light (PAR sensors), temp/humidity (HOBOs), soil moisture

\subsection*{Traits}

\subsubsection*{Structure}
\subsubsection*{Function}
\subsubsection*{Performance}

Growth rate: leaf expansion rate AND stem elongation rate

\subsubsection*{Floral}

\subsection*{Intrinsic variation}
	% population intercept?
\subsection*{Variation in stress response} % was high temp really stressful?
	% growth rate
	% photosynthetic
	% trait plasticity
	% variation in mortality (do I trust having multiple data points from growth measurements?)

\subsection*{Environmental correlates}
	% latitude
	% temperature (mean annual, July/june)
	% stream characterisitics - what data does Amy have from surveys?

	% approaches: direct correlation, bedazzle?

\end{document}

Vegetative: leaf symmetry or other leaf shape parameters that probably reflect developmental constraint but not affect function \\
Floral: Size, shape, herkogamy. Decide with Megan \\

	\subsection{Structure}

	Leaf traits method: \\
	Day 1: \\
	- select focal leaf (e.g. second or third leaf); \\
	- place in H20, dark, cool overnight. \\
	Day 2: \\
	- wet weight; \\
	- scan; \\
	- place in drying oven. \\

		\begin{enumerate}
			\item{Stomatal density, guard cell length, ratio from leaf opposite focal leaf (above}
			\item{Leaf size/shape (from scans)}
			\item{Leaf thickness/SLA}
			\item{Colorimetry (from scans, spec)}
			\item{stem anthocyanin (qualitative)}
		\end{enumerate}

	OPEN QUESTIONS: \\
	save wet samples in preservative (e.g. for leaf venation?) \\
	chlorophyll content? \\
	branching architecture (anything in plant protocols?)

	\subsection{Function}

		\subsubsection{during experiment}

		\begin{enumerate}
			\item{Instantaneous photosynthetic rate, $g_\text{s}$, water-use efficiency.}
				\subitem{3 min per plant per LICOR}
				\subitem{Do as many plants as possible in 1-2 weeks}
				\subitem{Before and after drought?}
			\item{Respiration}
			\item{Whole plant water-use (gravimetric)}
				\subitem{}
			\item{Leaf temperature (subset)}
			\item{Temperature response curve (subset)}
				\subitem{see protocol here: http://prometheuswiki.publish.csiro.au/tiki-index.php?page=Temperature+response+of+photosynthesis+using+a+Li-Cor+6400}
			\item{Light response curve}
				\subitem{Sampling? Only in cool, wet?}
				\subitem{At what temperature?}

		\end{enumerate}

		\subsubsection{after experiment}

		\begin{enumerate}
			\item{CN content}
			\item{carbon isotope discrimination}
		\end{enumerate}

	\subsection{Development time and Life history}

	\begin{enumerate}
		\item{Flowering time: census daily}
		\item{Starch accumulation: collect (subset) of plants}
		\item{Rhizome formation: collect on subset of plants harvested for dry mass}
	\end{enumerate}

	\subsection{Floral (2nd/3rd/etc flower?)}

	\begin{enumerate}
		\item{Flower size}
		\item{Herkogamy}
	\end{enumerate}

\section{Performance}

	Possibilities: \\
	1. (sub-)weekly census of plant height. $\approx$ 0.5 min per plant, so allow 1 day \\

	$$RGR_\text{height} = \frac{\text{ln} H_2 - \text{ln} H_1}{t_2 - t_1}$$

	2. early development (pre-treatment) leaf-area growth with camera \\
	3. harvest dry biomass before and after treatments (get whole plant leaf area using scanner at same time to ground truth camera)

	$$RGR_\text{mass} = \frac{\text{ln} M_2 - \text{ln} M_1}{t_2 - t_1}$$

\section{Determining ecological gradients that matter}

Performance tradeoffs
Isolation by ecology

\section{Tests of divergent selection}

	\subsection{Trait-environment correlation}

	We will account for relatedness by using genetic distance matrix in linear models.

	\subsection{$P_\text{ST}$ -- $F_\text{ST}$}

	Contrast trait classes with Fst distribution

	\subsection{Isolation by ecology}

	Is \textbf{G} better predicted by \textbf{E} than geographic distance? Use Gideon's method?

	\subsection{Performance tradeoffs between treatments}

\section{Ecological strategy, physiological mechanism, and tradeoff}

	\subsection{Drought escape}
		Pattern: \\
			Flowering time shorter in drier/S populations \\
			More between population variation in flowering time more variable than predicted by Fst \\
			... \\

	\subsection{Drought avoidance}
		Pattern: \\
			Maybe this should really be dehydration avoidance? \\
			Populations from drier/S avoid dehydration by limiting water loss; more variable than predicted by Fst \\

			Mechanism \\
			closing stomata earlier during soil drying

		Mechanisms: \\
			lower water-use \\

	\subsection{Drought tolerance}
		Pattern: \\
			drier/S populations have greater tolerance to low tissue water content

		Mechanisms: \\
			greater WUE \\
			ability to photosynthesize during drought \\

\section{Extension: incipient speciation as a byproduct or divergence?}

Collect pollen to measure viability.

\end{document}

### 2014-11-13 Analysis/MasterAnalysis.R

@@ -1,0 +1,110 @@
+#
+#	Master Analysis of M. cardinalis Growth Chamber Experiment (Spring 2014)
+#
+#	Title:	 Local Physiological adaptaiton in Mimulus cardinalist
+#	Authors: Christopher D. Muir and Amy L. Angert
+#	Journal: Evolution?
+#
+
+### Preliminaries
+
+#
+##	Libraries
+#
+
+	library(lme4)
+	library(MCMCglmm)
+	library(MASS)
+	library(car)
+	library(pbkrtest)
+	
+#
+##	Set working directory
+#
+
+	setwd("~/Google Drive/CardLocalAdaptation")
+
+#
+##	Population Environmental data
+#
+	# This needs to be updated with PRISM, other data, etc. Lat, lon should be correct though
+	popenv <- read.csv("Data/PopulationEnvData.csv")
+	
+#
+##	Read in processed datasets
+#
+
+	data <- read.csv("Data/MasterDatasheet_out.csv")
+	
+### JUST USING GROWTH NOW UNTIL OTHER DATA ARE READY
+### Analysis 1: Variation in 'intrinsic' traits (significant main effect of Population)
+### Analysis 2: Variation in plasticity (significant Treatment x Population interaction)
+
+	# LLL Growth rate
+	# Using Absolute growth rates for now as this seems most sensible. Need to test that this correlates with final biomass size.
+	
+	# approach 1: stepAIC, no Family effect
+	fit1 <- stepAIC(lm(lll_AbsGrowth ~ Population * TempTrt * WaterTrt, data = data))
+	Anova(fit1, type = 2)
+
+	# approach 2: mixed model with family	
+	fit1 <- lmer(lll_AbsGrowth ~ Population * TempTrt * WaterTrt + (1|Family), data = data)
+	
+	# 3-way interaction
+	fit2 <- update(fit1, . ~ . - Population:TempTrt:WaterTrt)
+	KRmodcomp(fit1, fit2) # n.s.
+
+	# Population x Treatment interactions
+	fit3 <- update(fit2, . ~ . - Population:TempTrt)
+	fit4 <- update(fit2, . ~ . - Population:WaterTrt)
+	KRmodcomp(fit2, fit3) # n.s.
+	KRmodcomp(fit2, fit4) # marginally significant. keep.
+	
+	# Main effects
[…]

### 2014-11-13 Analysis/PopulationMap.R

@@ -1,0 +1,208 @@
+#
+# Niche modeling in Mimulus cardinalis
+#
+
+setwd("~/Dropbox/CardNiche")
+
+# Records from GBIF (these already include UCJEPS, so maybe just use those?)
+
+card <- gbif("Mimulus", "cardinalis")
+card <- card[-which(card$lon > -80), ]
+nrow(card)
+
+# Jespon herbarium records
+# Accessed July 3, 2013
+
+card <- read.csv("~/Documents/Postdoc/CardNiche/card_jepson_records_minimal.csv", colClasses = c("factor", "integer", "integer", "factor", "character", "numeric", "numeric"))
+with(card, plot(longitude, latitude))
+
+# convert elevation strings to meters
+
+x <- which(regexpr("[0-9]+ [-] [0-9]+ m", card$elevation) != -1)
+
+le <- as.numeric(unlist(strsplit(card$elevation[x], " - "))[seq(1, 2 * length(x), 2)])
+ue <- as.numeric(unlist(strsplit(unlist(strsplit(card$elevation[x], " - "))[seq(2, 2 * length(x), 2)], " m")))
+card$elevation[x] <- (le + ue) / 2
+
+x <- which(regexpr("[0-9]+ [-] [0-9]+ ft", card$elevation) != -1)
+le <- as.numeric(unlist(strsplit(card$elevation[x], " - "))[seq(1, 2 * length(x), 2)])
+ue <- as.numeric(unlist(strsplit(unlist(strsplit(card$elevation[x], " - "))[seq(2, 2 * length(x), 2)], " ft")))
+card$elevation[x] <- (le + ue) / 2
+
+x <- regexpr(" ft", card$elevation)
+
+card$elevation[x != -1] <- as.numeric(substr(card$elevation[x != -1], 1, x[x != -1] - 1)) * 0.3048
+
+x <- regexpr(" m", card$elevation)
+
+card$elevation[x != -1] <- as.numeric(substr(card$elevation[x != -1], 1, x[x != -1] - 1))
+
+card$elevation <- as.numeric(card$elevation)
+
+#
+## Anget lab seed collections
+#
+
+library(raster)
+library(dismo)
+
+card2 <- read.csv("~/Dropbox/CardNiche/CardPopulations.csv")
+
+# get rid of distant pop
+card2 <- card2[card2$Lon < -115 & !is.na(card2$Lon), ]
[…]

### 2014-11-13 Data/ProcessData.R

@@ -1,0 +1,455 @@
+setwd("~/Google Drive/CardLocalAdaptation/Data")
+
+library(lme4)
+library(reshape2)
+library(lubridate)
+library(MASS)
+
+#
+#	Preparing datasheet for LLL analysis
+#
+
+	data <- read.csv("MasterDatasheet_in.csv")
+	cols2keep <- c("Block", "Col", "Row", "Population", "Family", "Transplant", "TempTrt",
+		"WaterTrt", "LLL_0512", "LLL_0515", "LLL_0520", "LLL_0523", "LLL_0526", "LLL_0529",
+		"LLL_0602", "LLL_0605", "LLL_0609", "LLL_0612")
+	data$Row <- as.character(data$Row)
+	data$Population[data$Population == "Sweetwater River"] <- "Cuyamaca Rancho"
+	data$Population <- factor(as.character(data$Population))
+	data <- data[1:768, cols2keep]
+	data$indiv <- gl(768, 1)
+	data$TempTrt <- factor(as.character(data$TempTrt))
+	data$WaterTrt <- factor(as.character(data$WaterTrt))
+
+	# Get rid of transplants
+	data <- subset(data, data$Transplant == "N")
+
+	# Remove all < 1 mm, call NA
+	data$LLL_0512 <- as.numeric(gsub("< 1", "NA", data$LLL_0512))
+	data$LLL_0515 <- as.numeric(gsub("< 1", "NA", data$LLL_0515))
+	data$LLL_0520 <- as.numeric(gsub("< 1", "NA", data$LLL_0520))
+	data$LLL_0523 <- as.numeric(gsub("< 1", "NA", data$LLL_0523))
+	data$LLL_0526 <- as.numeric(gsub("< 1", "NA", data$LLL_0526))
+
+#
+#	LLL growth rates
+#
+
+	# Melt data.frame
+	lll.data <- melt(data, value.name = "LLL")
+	lll.data$Day <- mdy(paste(substr(lll.data$variable, 5, 8), "2014", sep = ""))
+	lll.data$DayN <- (as.numeric(lll.data$Day) - 1399852800) / 86400
+	
+	# Get rid of NAs
+	lll.data <- lll.data[!is.na(lll.data$LLL), ]
+	
+	# Remove those with less than 5 datapoints (removes 18 observations)
+	# Remove those with less than 4 datapoints (removes 6 observations)
[…]

### 2014-11-14 ms/ms.tex
Edited manuscript. LMER analysis of growth rate data. Added functions.R

@@
 \usepackage{natbib}
+
+\usepackage{color}
+
+%
+% Commenting
+%
+
+\newcommand{\cdm}[1]{{ \color{magenta} [{\bf{CDM:}} {\em#1}]}} % Chris' comments are in magenta.
+\newcommand{\ala}[1]{{ \color{blue} [{\bf{CDM:}} {\em#1}]}} % Amy's comments are in blue.
 
@@
 % What's the big picture (this is TOO big):
-It is taken for granted that genes are the foundation of evolution.
+% It is taken for granted that genes are the foundation of evolution.
 If we get to the genetic basis of adaptive evolution, then we have understood the process at its deepest level.
@@
 
-% this paper. good start!
-Local adaptation is one of the most ubiquitous observations in evolutionary ecology.
-In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
+Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
+Local adaptation is most often examined within species where selection in different environments leads to ecologically relevant trait divergence between populations.
+Despite local adaptation leading to population divergence, many traits are conserved within species.
+\cdm{this is not quite right.
+What I really want to say is that some traits divergence genetically, others are plastic, but plasticity is conserved because...?
+why?
+no spatial variation in selection for plasticity?}.
+Can we predict which types of traits are likely to be different between populations and which are likely to be constant?
+A long-standing hypothesis is that the grain of environmental variation determines whether organisms respond through population genetic differentiation or phenotypic plasticity \citep{}.
+\cdm{What I want to get to is whether plasticity is conserved.
+The problem is, just because there is selection for plasticity, this doesn't need to imply that the reaction norm is actually the same across populations.
+But perhaps we can oversimplify here?}.
+
+When the environment is fine grained
+
+, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
 Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
@@
 
-I chose 16 populations from throughout the range of \textit{M. cardinalis} (Table 1) that have been previously studied. Seeds were collected in the field [Amy - explain seed collection].
+I chose 16 populations from throughout the range of \textit{M.
+cardinalis} (Table 1) that have been previously studied.
+Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
 
@@
 	Cottonwood Creek					& South Margin		& yes	& no		& 32.80122	& -116.50194	& 1206 \\
-	Sweetwater River / Cuyamaca Rancho	& South Margin		& yes	& yes	& 32.89928	& -116.5849	& 1223 \\
+	Sweetwater River					& South Margin		& yes	& yes	& 32.89928	& -116.5849	& 1223 \\
 	Grade Road - to Palomar Mountain SP	& South Margin		& no		& no		& 33.31392	& -116.87129	& 1500 \\
@@
 	West Fork Mojave River				& Transverse		& yes	& no		& 34.28425	& -117.37539	& 1087 \\
-	North Fork Middle Tule				&				& yes	&		& 36.20081	& -118.65092	& 1284 \\
+	North Fork Middle Tule				& South Sierras		& yes	& yes?	& 36.20081	& -118.65092	& 1284 \\
 	Paradise Creek						& South Sierras		& yes	& yes	& 36.51776	& -118.75877	& 887 \\
@@
 
+\section*{Results}
+
+% ANOVA table for LLL growth rate (need to label tables)
+\begin{table*}[htbp]
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\textbf{caption} \\ \\
+\begin{center}
+\begin{tabular}{lccccc}
+\toprule
+
+\input{Tables/Table_lllGrowthAnova.txt}
+
+\bottomrule
+\end{tabular}
+\end{center}
+\end{table*}
+
 \end{document}
-
-Vegetative: leaf symmetry or other leaf shape parameters that probably reflect developmental constraint but not affect function \\
-Floral: Size, shape, herkogamy. Decide with Megan \\
 

### 2014-11-14 Analysis/MasterAnalysis.R

@@ -17,6 +17,7 @@
 	library(MASS)
 	library(car)
 	library(pbkrtest)
+	library(lmerTest)
 	
 #
 ##	Set working directory
@@ -24,6 +25,12 @@
 
 	setwd("~/Google Drive/CardLocalAdaptation")
 
+#
+##	Source custom functions
+#
+
+	source("Analysis/functions.R")
+	
 #
 ##	Population Environmental data
 #
@@ -48,25 +55,32 @@
 	Anova(fit1, type = 2)
 
 	# approach 2: mixed model with family	
-	fit1 <- lmer(lll_AbsGrowth ~ Population * TempTrt * WaterTrt + (1|Family), data = data)
-	
-	# 3-way interaction
-	fit2 <- update(fit1, . ~ . - Population:TempTrt:WaterTrt)
-	KRmodcomp(fit1, fit2) # n.s.
+	mm <- lmer(lll_AbsGrowth ~ Population * TempTrt * WaterTrt + (1|Family),
+		data = data)
+	fit1 <- step(mm) #includes pop, temp, h2o, population x water
+	# fit1 <- step(fit1, ddf = "Kenward-Roger") # same result using Satterthwaite ddf
 
-	# Population x Treatment interactions
-	fit3 <- update(fit2, . ~ . - Population:TempTrt)
-	fit4 <- update(fit2, . ~ . - Population:WaterTrt)
-	KRmodcomp(fit2, fit3) # n.s.
-	KRmodcomp(fit2, fit4) # marginally significant. keep.
+	# Dyanamic table of mm output for ms.tex
+	fit1$anova.table
+	exportTable(file = "ms/Tables/Table_lllGrowthAnova.txt", data = data.frame(rnorm(5), rnorm(5)), colnames = c("Test1", "Test2"), rownames = LETTERS[1:5])
+
+
+	# modify code to print anova tables in latex
+	mapply(function(X, Y)
+	{
+		tmp <- anova(X, fit1a)
+		p <- tmp[, "Pr(>F)"][2]
+		oom <- ceiling(-log10(p)) # order of magnitude
+		p <- if(p > 0.001){round(p, 2)}else{paste(round(p * 10 ^ oom, 2), 
+			sprintf("$\\times10^{-%s}$", oom))}
+		cat(sprintf('%s & %s & %s & %s & %s & %s \\\\', Y, tmp$Df[2], 
[…]

### 2014-11-14 Analysis/functions.R

@@ -1,0 +1,22 @@
+exportTable <- function(file, data, colnames, rownames, ...)
+{
+	# Check that there are the right number of columns and rows
+	if(length(colnames) != ncol(data)) stop("number of columns in data does not match length of column names")
+
+	if(length(rownames) != nrow(data)) stop("number of rows in data does not match length of row names")
+
+	table <- file(file, open = "w")
+	cat(" &", colnames[1], file = table)
+	for (i in 2:length(colnames)) cat(" &", colnames[i], file = table)
+	cat(" \\\\ \\midrule\n", file = table)
+	
+	for (i in 1:length(rownames))
+	{
+		cat(rownames[i], "&", data[i, 1], file = table)
+		for (j in 2:length(colnames)) cat(" &", data[i, j], file = table)
+		cat(" \\\\\n", file = table)
+	}
+	
+	close(con = table)	
+	
+}

### 2014-11-15 Analysis/MasterAnalysis.R
Worked on ANOVA tables for growth rate data

@@ -58,53 +58,73 @@
 	mm <- lmer(lll_AbsGrowth ~ Population * TempTrt * WaterTrt + (1|Family),
 		data = data)
 	fit1 <- step(mm) #includes pop, temp, h2o, population x water
-	# fit1 <- step(fit1, ddf = "Kenward-Roger") # same result using Satterthwaite ddf
+	# fit1 <- step(mm, ddf = "Kenward-Roger") # same result using Satterthwaite ddf
 
 	# Dyanamic table of mm output for ms.tex
-	fit1$anova.table
-	exportTable(file = "ms/Tables/Table_lllGrowthAnova.txt", data = data.frame(rnorm(5), rnorm(5)), colnames = c("Test1", "Test2"), rownames = LETTERS[1:5])
-
-
-	# modify code to print anova tables in latex
-	mapply(function(X, Y)
-	{
-		tmp <- anova(X, fit1a)
-		p <- tmp[, "Pr(>F)"][2]
-		oom <- ceiling(-log10(p)) # order of magnitude
-		p <- if(p > 0.001){round(p, 2)}else{paste(round(p * 10 ^ oom, 2), 
-			sprintf("$\\times10^{-%s}$", oom))}
-		cat(sprintf('%s & %s & %s & %s & %s & %s \\\\', Y, tmp$Df[2], 
-			round(tmp[, "Sum of Sq"][2], 3), round(tmp[, "Sum of Sq"][2] / tmp$Df[2], 3), 
-			round(tmp$F[2], 2), p), file = table2)
-	}, listFita, predVar)
-		
+	tab <- fit1$anova.table[c("Population", "TempTrt", "WaterTrt", "Population:TempTrt",
+		"Population:WaterTrt", "TempTrt:WaterTrt", "Population:TempTrt:WaterTrt"), ]
 	
+	# Modify P-values for scientific notation
+	p <- -expm1(pf(tab$F, tab$NumDF, tab$DenDF, log = T))
+	oom <- ceiling(-log10(p)) # order of magnitude
+	p <- mapply(function(X1, X2) if(X1 > 0.01){round(X1, 2)} else {
+		paste(round(X1 * 10 ^ X2, 2), sprintf("$\\times10^{-%s}$", X2))}, p, oom)
+	
+	dat <- data.frame(SS = round(tab[, "Sum Sq"], 1), MS = round(tab[, "Mean Sq"], 1),
[…]

### 2014-11-15 ms/ms.tex
Worked on ANOVA tables for growth rate data
Worked on analysis, made growth rate ANOVA tables, figures correlating growth with latitude, made refs.bib file, and added some to methods section of ms.tex

@@
 \usepackage{natbib}
-
 \usepackage{color}
@@
 \newcommand{\cdm}[1]{{ \color{magenta} [{\bf{CDM:}} {\em#1}]}} % Chris' comments are in magenta.
-\newcommand{\ala}[1]{{ \color{blue} [{\bf{CDM:}} {\em#1}]}} % Amy's comments are in blue.
+\newcommand{\ala}[1]{{ \color{blue} [{\bf{ALA:}} {\em#1}]}} % Amy's comments are in blue.
 
@@
 			Wet: daily/constant irrigation with cooled water
-			Dry: start at week \#, gradually decrease watering frequency/level (see Nancy Emery's protocols)
+			Dry: start at week \#, gradually decrease watering frequency/level
 
@@
 \subsubsection*{Structure}
+
+leaf traits, biomass allometry
+
 \subsubsection*{Function}
+
+photosynthesis
+
 \subsubsection*{Performance}
@@
 Growth rate: leaf expansion rate AND stem elongation rate
-
-\subsubsection*{Floral}
+Biomass (show correlation between growth rate and biomass)
 
 \subsection*{Intrinsic variation}
-	% population intercept?
-\subsection*{Variation in stress response} % was high temp really stressful?
-	% growth rate
-	% photosynthetic
-	% trait plasticity
-	% variation in mortality (do I trust having multiple data points from growth measurements?)
+
+\cdm{which traits: growth rate, biomass, biomass allometry, leaf traits, photosynthesis?, mortality?}
+We tested for Population, Treatment, and Population $\times$ Treatment interactions using mixed model ANOVA with Family included as a random factor.
+Models were fit by restricted maximum likelihood using lmer from the R package lme4 version 1.0-5 \citep{Bates_etal_2013}.
+Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package lmerTest version 2.0-11 \citep{Kuznetsova_etal_2014}.
+Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation.
+Significant Population effect indicate intrinsic trait differences; significant Population $\times$ Treatment effects indicate population differences in plasticity.
+
+	%
 
@@
 \selectfont
-\textbf{caption} \\ \\
+\textbf{ANOVA table, Longest leaf length growth} \\ \\
 \begin{center}
-\begin{tabular}{lccccc}
+\begin{tabular}{lcccccc}
 \toprule
@@
 \end{table*}
+
+% ANOVA table for height growth rate (need to label tables)
+\begin{table*}[htbp]
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\textbf{ANOVA table, height growth} \\ \\
+\begin{center}
+\begin{tabular}{lcccccc}
+\toprule
+
+\input{Tables/Table_heightGrowthAnova.txt}
+
+\bottomrule
+\end{tabular}
+\end{center}
+\end{table*}
+
+\subsection*{Latitude versus growth rate}
+
+\begin{figure*}
+\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_LLL_Lat.pdf}}
+\label{fig:Fig1}
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\textbf{Figure \#} 
+\end{figure*}
+
+\begin{figure*}
+\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_Height_Lat.pdf}}
+\label{fig:Fig1}
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\textbf{Figure \#} 
+\end{figure*}
+
+%--------------------------------------------------
+% References
+%--------------------------------------------------
+
+\setlength{\bibsep}{6pt}
+\bigskip
+
+\bibliography{refs}
+\bibliographystyle{evolution}
 

### 2014-11-15 Analysis/MasterAnalysis.R

@@ -35,6 +35,7 @@
 ##	Population Environmental data
 #
 	# This needs to be updated with PRISM, other data, etc. Lat, lon should be correct though
+	# Populations are ordered alphabetically by site to match output from stats
 	popenv <- read.csv("Data/PopulationEnvData.csv")
 	
 #
@@ -50,11 +51,7 @@
 	# LLL Growth rate
 	# Using Absolute growth rates for now as this seems most sensible. Need to test that this correlates with final biomass size.
 	
-	# approach 1: stepAIC, no Family effect
-	fit1 <- stepAIC(lm(lll_AbsGrowth ~ Population * TempTrt * WaterTrt, data = data))
-	Anova(fit1, type = 2)
-
-	# approach 2: mixed model with family	
+	# Mixed model ANOVA with family as random factor fit using step down procedure
 	mm <- lmer(lll_AbsGrowth ~ Population * TempTrt * WaterTrt + (1|Family),
 		data = data)
 	fit1 <- step(mm) #includes pop, temp, h2o, population x water
@@ -82,18 +79,46 @@
 	exportTable(file = "ms/Tables/Table_lllGrowthAnova.txt", data = dat, 
 		colnames = colnames, rownames = rownames)
 			
-	# Plot latitudinal variation
+	# Plot latitudinal variation in Intercept and Population:WaterTrt
 	
-	#### NEED TO DO ####
-
+	# best fit model based on stepdown elimination procedure
+	fit1 <- lmer(lll_AbsGrowth ~ Population + TempTrt + WaterTrt + Population:WaterTrt +
+		(1|Family), data = data)
+	
+	# Intercept: least-square coefficients ('betas') and 95% CIs
+	betas <- lsmeans(fit1)
+	
+	# Rows with Population coefficients
+	x <- grep("Population  [a-zA-Z]+", rownames(betas$lsmeans.table))
+	pdf("ms/Figures/Figure_LLL_Lat.pdf", 5, 5)
+	par(mar = c(5, 5, 1, 1))
+	plot(popenv$Lat, betas$lsmeans.table$Estimate[x], xlim = c(30, 45), ylim = c(1.5, 3.1),
+		xlab = "Latitude of Origin", ylab = expression(paste("Leaf elongation rate (", 
[…]

### 2014-11-17 Data/ProcessData.R
Cleaned up MasterDatasheet_in.csv and worked on ProcessData.R

@@ -9,20 +9,17 @@ library(MASS)
 #	Preparing datasheet for LLL analysis
 #
 
-	data <- read.csv("MasterDatasheet_in.csv")
-	cols2keep <- c("Block", "Col", "Row", "Population", "Family", "Transplant", "TempTrt",
+	data <- read.csv("MasterDatasheet_in.csv", nrows = 768)
+	cols2keep <- c("indiv", "Block", "Col", "Row", "Population", "Family", "TempTrt",
 		"WaterTrt", "LLL_0512", "LLL_0515", "LLL_0520", "LLL_0523", "LLL_0526", "LLL_0529",
 		"LLL_0602", "LLL_0605", "LLL_0609", "LLL_0612")
 	data$Row <- as.character(data$Row)
 	data$Population[data$Population == "Sweetwater River"] <- "Cuyamaca Rancho"
 	data$Population <- factor(as.character(data$Population))
-	data <- data[1:768, cols2keep]
 	data$indiv <- gl(768, 1)
+	data <- data[data$UseForLLLGrowth, cols2keep]
 	data$TempTrt <- factor(as.character(data$TempTrt))
 	data$WaterTrt <- factor(as.character(data$WaterTrt))
-
-	# Get rid of transplants
-	data <- subset(data, data$Transplant == "N")
 
 	# Remove all < 1 mm, call NA
 	data$LLL_0512 <- as.numeric(gsub("< 1", "NA", data$LLL_0512))
@@ -42,20 +39,19 @@ library(MASS)
 	
 	# Get rid of NAs
 	lll.data <- lll.data[!is.na(lll.data$LLL), ]
-	
-	# Remove those with less than 5 datapoints (removes 18 observations)
-	# Remove those with less than 4 datapoints (removes 6 observations)
-	lll.data <- lll.data[!(lll.data$indiv %in% which(table(lll.data$indiv) < 4)), ]
-	
+		
 	# Refactor indiv
 	lll.data$indiv <- as.factor(as.character(lll.data$indiv))
 
 	#
 	# Empirical Bayes Approach: get coefficients for every plant
 	#
-	
+
 	mm <- lmer(log(LLL) ~ poly(DayN, 2, raw = T) + (poly(DayN, 2, raw = T)|indiv),
 		data = lll.data)
+	# Logistic not a very good fit. Something else worth trying?
[…]

### 2014-11-20 Data/ProcessData.R
Data quality control and added data to MasterDatasheet_out.csv

@@ -4,7 +4,7 @@ library(lme4)
 library(reshape2)
 library(lubridate)
 library(MASS)
-
+	
 #
 #	Preparing datasheet for LLL analysis
 #
@@ -234,28 +234,24 @@ library(MASS)
 	# Should be true if tmp and data are in same order
 	all(tmp$indiv == data$indiv)
 
-	tmp$mass <- with(tmp, LeavesDW_g + ShootsDW_g + RootsDW_g)
-	plot(data$lll_AbsGrowth, tmp$mass, col = data$TempTrt, log = "y")
-	plot(data$lll_RelGrowth, tmp$mass, col = data$TempTrt, log = "y")
-
-	plot(data$height_AbsGrowth, tmp$mass, col = data$TempTrt, log = "y")
-	plot(data$height_RelGrowth, tmp$mass, col = data$TempTrt, log = "y")
-
-
-	tmp <- subset(tmp, nchar(as.character(tmp$HarvestDate)) == 6 & !is.na(tmp$mass))
-	tmp$HarvestDate <- dmy(paste(as.character(tmp$HarvestDate), "-2014", sep = ""))
-	tmp$HarvestDate <- (as.numeric(tmp$HarvestDate) - 1399852800) / 86400
+	# add biomass and flowers at harvest as well as harvest date [STILL NEEDS QC]
+	data$LeavesDW_g <- ifelse(tmp$UseForBiomass, tmp$LeavesDW_g, NA)
+	data$ShootsDW_g <- ifelse(tmp$UseForBiomass, tmp$ShootsDW_g, NA)
+	data$RootsDW_g <- ifelse(tmp$UseForBiomass, tmp$RootsDW_g, NA)
+	data$Biomass <- with(data, LeavesDW_g + ShootsDW_g + RootsDW_g)
+	data$NFlwr <- ifelse(tmp$UseForBiomass, tmp$NFlwr, NA)
+	data$NBud <- ifelse(tmp$UseForBiomass, tmp$Nbud, NA)# change column name to NBud in datasheet
+	data$HarvestDate <- ifelse(nchar(as.character(tmp$HarvestDate)) == 6, 
+		(as.numeric(dmy(paste(as.character(tmp$HarvestDate), "-2014", sep = ""))) - 
+		1399852800) / 86400, NA)
 	
-	with(tmp, plot(HarvestDate, mass, log = "y", col = TempTrt))
[…]

### 2014-11-21 Analysis/MasterAnalysis.R
Tested for population differences in germination and initial size.

@@ -18,6 +18,7 @@
 	library(car)
 	library(pbkrtest)
 	library(lmerTest)
+	library(survival)
 	
 #
 ##	Set working directory
@@ -43,7 +44,122 @@
 #
 
 	data <- read.csv("Data/MasterDatasheet_out.csv")
+
+#
+##	Difference in germiantion timing (yes) and intitial size(?)?
+#
+
+	# Create Surv object
+	germ <- with(subset(data, !is.na(data$MinGermDay)), Surv(MinGermDay, MaxGermDay,	
+		type = "interval2"))
+		
+	fit1 <- survreg(germ ~ Population * WaterTrt * TempTrt, 
+		data = subset(data, !is.na(data$MinGermDay)))
+	Anova(fit1, type = 2)
+	with(subset(data, !is.na(data$MinGermDay)), plot(predict(fit1), Biomass, log = "y", col = TempTrt))
+	with(subset(data, !is.na(data$MinGermDay)), plot(predict(fit1), LLL_0512, log = "y"))	
+	data$low <- ifelse(is.na(data$LLL_0512), -Inf, log(data$LLL_0512))
+	data$high <- ifelse(is.na(data$LLL_0512), log(0.24), log(data$LLL_0512))
+	mm1 <- MCMCglmm(cbind(low, high) ~ Population * WaterTrt * TempTrt, 
+		random = ~ Family, data = data, family = "cengaussian", nitt = 1.1e5, thin = 1e2,
+		burnin = 1e4)
+	mm2 <- MCMCglmm(cbind(low, high) ~ Population + WaterTrt + TempTrt + 
+		Population:WaterTrt + Population:TempTrt + WaterTrt:TempTrt, 
+		random = ~ Family, data = data, family = "cengaussian", nitt = 1.1e5, thin = 1e2,
+		burnin = 1e4)
+	mm2$DIC - mm1$DIC # drop 3-way interaction
 	
+	# Test for WaterTrt:TempTrt
+	mm3 <- MCMCglmm(cbind(low, high) ~ Population + WaterTrt + TempTrt + 
+		Population:WaterTrt + Population:TempTrt, 
+		random = ~ Family, data = data, family = "cengaussian", nitt = 1.1e5, thin = 1e2,
+		burnin = 1e4)
+	mm3$DIC - mm2$DIC # WaterTrt:TempTrt n.s.
+
[…]

### 2014-11-21 Analysis/MasterAnalysis.R
Analysis of population differences in germination rate and initial size.

@@ -46,20 +46,23 @@
 	data <- read.csv("Data/MasterDatasheet_out.csv")
 
 #
-##	Difference in germiantion timing (yes) and intitial size(?)?
+##	Difference in germiantion timing (yes) and intitial size (no?)?
 #
 
+	##### Germination
 	# Create Surv object
 	germ <- with(subset(data, !is.na(data$MinGermDay)), Surv(MinGermDay, MaxGermDay,	
 		type = "interval2"))
-		
-	fit1 <- survreg(germ ~ Population * WaterTrt * TempTrt, 
-		data = subset(data, !is.na(data$MinGermDay)))
-	Anova(fit1, type = 2)
-	with(subset(data, !is.na(data$MinGermDay)), plot(predict(fit1), Biomass, log = "y", col = TempTrt))
-	with(subset(data, !is.na(data$MinGermDay)), plot(predict(fit1), LLL_0512, log = "y"))	
+
+	fit1 <- survreg(germ ~ Population + frailty(Family), 
+		data = subset(data, !is.na(data$MinGermDay)), dist = "gaussian")
+	plot(popenv$Lat, c(coef(fit1)[1], coef(fit1)[1] + coef(fit1)[2:16]))
+	anova(fit1) # significant family and population effects
+
+	##### Initial size (LLL on May 12)
 	data$low <- ifelse(is.na(data$LLL_0512), -Inf, log(data$LLL_0512))
 	data$high <- ifelse(is.na(data$LLL_0512), log(0.24), log(data$LLL_0512))
+	set.seed(1121014)
 	mm1 <- MCMCglmm(cbind(low, high) ~ Population * WaterTrt * TempTrt, 
 		random = ~ Family, data = data, family = "cengaussian", nitt = 1.1e5, thin = 1e2,
 		burnin = 1e4)
@@ -151,15 +154,17 @@
 		burnin = 1e4)
 	mm14$DIC - mm12$DIC # n.s.
 
+	mm15 <- MCMCglmm(cbind(low, high) ~ 1, data = data, family = "cengaussian", 
+		nitt = 1.1e5, thin = 1e2, burnin = 1e4)
+	mm15$DIC - mm14$DIC # n.s.
+
 	#mm14 is best (almost lowest DIC)
-
-
-	# Test for WaterTrt
-	mm7 <- MCMCglmm(cbind(low, high) ~ Population + TempTrt + 
[…]

### 2014-11-22 Analysis/MasterAnalysis.R
Quantified effects of germination day on growth rate and made new figures and tables.

@@ -54,8 +54,10 @@
 	germ <- with(subset(data, !is.na(data$MinGermDay)), Surv(MinGermDay, MaxGermDay,	
 		type = "interval2"))
 
-	fit1 <- survreg(germ ~ Population + frailty(Family), 
-		data = subset(data, !is.na(data$MinGermDay)), dist = "gaussian")
+	# lognormal produced best fit to data
+	fit1 <- survreg(germ ~ Population + frailty(Family, sparse = F), 
+		data = subset(data, !is.na(data$MinGermDay)), dist = "lognormal")
+		
 	plot(popenv$Lat, c(coef(fit1)[1], coef(fit1)[1] + coef(fit1)[2:16]))
 	anova(fit1) # significant family and population effects
 
@@ -173,14 +175,15 @@
 	# Using Absolute growth rates for now as this seems most sensible. Need to test that this correlates with final biomass size.
 	
 	# Mixed model ANOVA with family as random factor fit using step down procedure
-	mm <- lmer(lll_AbsGrowth ~ Population * TempTrt * WaterTrt + (1|Family),
+	mm <- lmer(lll_AbsGrowth ~ AvgGermDay + Population * TempTrt * WaterTrt + (1|Family),
 		data = data)
 	fit1 <- step(mm, reduce.random = F) #includes pop, temp, h2o, population x water
-	# fit1 <- step(mm, ddf = "Kenward-Roger") # same result using Satterthwaite ddf
+	# fit1 <- step(mm, ddf = "Kenward-Roger", reduce.random = F) # same result using Satterthwaite ddf
 
 	# Dyanamic table of mm output for ms.tex
-	tab <- fit1$anova.table[c("Population", "TempTrt", "WaterTrt", "Population:TempTrt",
-		"Population:WaterTrt", "TempTrt:WaterTrt", "Population:TempTrt:WaterTrt"), ]
+	tab <- fit1$anova.table[c("AvgGermDay", "Population", "TempTrt", "WaterTrt", 
+		"Population:TempTrt", "Population:WaterTrt", "TempTrt:WaterTrt", 
+		"Population:TempTrt:WaterTrt"), ]
 	
 	# Modify P-values for scientific notation
 	p <- -expm1(pf(tab$F, tab$NumDF, tab$DenDF, log = T))
@@ -193,19 +196,19 @@
 		F = round(tab[, "F.value"], 1), P = p,
 		stringsAsFactors = F)
[…]

### 2014-11-22 Data/ProcessData.R

@@ -240,7 +240,7 @@ library(MASS)
 	data$RootsDW_g <- ifelse(tmp$UseForBiomass, tmp$RootsDW_g, NA)
 	data$Biomass <- with(data, LeavesDW_g + ShootsDW_g + RootsDW_g)
 	data$NFlwr <- ifelse(tmp$UseForBiomass, tmp$NFlwr, NA)
-	data$NBud <- ifelse(tmp$UseForBiomass, tmp$Nbud, NA)# change column name to NBud in datasheet
+	data$NBud <- ifelse(tmp$UseForBiomass, tmp$NBud, NA)
 	data$HarvestDate <- ifelse(nchar(as.character(tmp$HarvestDate)) == 6, 
 		(as.numeric(dmy(paste(as.character(tmp$HarvestDate), "-2014", sep = ""))) - 
 		1399852800) / 86400, NA)
@@ -248,6 +248,7 @@ library(MASS)
 	# add germination date
 	data$MinGermDay <- tmp$MinGermDay
 	data$MaxGermDay <- tmp$MaxGermDay
+	data$AvgGermDay <- (tmp$MinGermDay + tmp$MaxGermDay) / 2
 
 	# add Mortality
 	data$DiedFromStress <- tmp$DiedFromStress

### 2014-11-22 ms/ms.tex
Made figure for germination day and updated ms.tex

@@
 \usepackage{topcapt}
+\usepackage[labelfont =  {sf, bf}, textfont = sf, format = plain]{caption}
 \usepackage{amssymb}
@@
 \usepackage{color}
+
+% Make helvetica the default sans-serif font
+\renewcommand\sfdefault{phv}
 
@@
 \maketitle
+\listoffigures
+\listoftables
 
@@
 % ANOVA table for LLL growth rate (need to label tables)
-\begin{table*}[htbp]
+\begin{table}[htbp]
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-\textbf{ANOVA table, Longest leaf length growth} \\ \\
+	\caption[ANOVA table, leaf expansion rate]{CAPTION}
 	\begin{center}
@@
 	\end{center}
-\end{table*}
+\end{table}
 
 % ANOVA table for height growth rate (need to label tables)
-\begin{table*}[htbp]
+\begin{table}[htbp]
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-\textbf{ANOVA table, height growth} \\ \\
+	\caption[ANOVA table, stem expansion rate]{CAPTION}
 	\begin{center}
@@
 	\end{center}
-\end{table*}
+\end{table}
 
@@
 
-\begin{figure*}
-\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_LLL_Lat.pdf}}
-\label{fig:Fig1}
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-\textbf{Figure \#} 
-\end{figure*}
+	\caption[Southern populations germinate sooner.]{CAPTION}
+	\label{fig:Fig_DoG}
+\end{figure}
 
-\begin{figure*}
-\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_Height_Lat.pdf}}
-\label{fig:Fig1}
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-\textbf{Figure \#} 
-\end{figure*}
+	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
+	\label{fig:Fig_LLL}
+\end{figure}
+
+\begin{figure}
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Southern populations grow faster (stem elongation rate).]{CAPTION}
+	\label{fig:Fig_height}
+\end{figure}
 

### 2014-11-22 Analysis/MasterAnalysis.R

@@ -57,9 +57,34 @@
 	# lognormal produced best fit to data
 	fit1 <- survreg(germ ~ Population + frailty(Family, sparse = F), 
 		data = subset(data, !is.na(data$MinGermDay)), dist = "lognormal")
-		
-	plot(popenv$Lat, c(coef(fit1)[1], coef(fit1)[1] + coef(fit1)[2:16]))
 	anova(fit1) # significant family and population effects
+	
+	# Figure of germination time by latitude
+	png("ms/Figures/Figure_DoG_Lat.png", 5, 5, units = "in", res = 1e3)
+	par(mar = c(5, 5, 1, 1))
+	plot(popenv$Lat, c(coef(fit1)[1], coef(fit1)[1] + coef(fit1)[2:16]), xlim = c(30, 45), 
+		ylim = c(1.65, 2.65), xlab = "Latitude of Origin", 
+		ylab = "Day of Germination [log scale]", 
+		cex.lab = 1.5, type = "n", axes = F, frame.plot = T)
+	lowCI <- c(confint(fit1)[1, 1], coef(fit1)[1] + confint(fit1)[2:16, 1])
+	uppCI <- c(confint(fit1)[1, 2], coef(fit1)[1] + confint(fit1)[2:16, 2])
+	arrows(popenv$Lat, lowCI, popenv$Lat, uppCI, length = 0.05, angle = 90, code = 3)
+	points(popenv$Lat, c(coef(fit1)[1], coef(fit1)[1] + coef(fit1)[2:16]), col = "white",
+		bg = "black", pch = 21, cex = 1.5)
+	
+	axis(1, at = seq(30, 45, 5), labels = expression(30*degree, 35*degree, 40*degree,
+		45*degree), lwd = 0, lwd.ticks = 1)	
+	axis(2, lwd = 0, lwd.ticks = 1, las = 1, at = log(seq(6, 14, 2)),
+		labels = seq(6, 14, 2))	
+	
+	fit <- lm(c(coef(fit1)[1], coef(fit1)[1] + coef(fit1)[2:16]) ~ Lat, data = popenv)
+	points(range(popenv$Lat), predict(fit, new = data.frame(Lat = range(popenv$Lat))),
+		type = "l", lwd = 2)
+	
+	text(grconvertX(0.05, "npc"), grconvertY(0.95, "npc"), adj = c(0, 1),
[…]

### 2014-11-23 Analysis/MasterAnalysis.R
Finished supplemental table on initial size (LLL_0512) and updated exportTable function

@@ -192,6 +192,24 @@
 	load(file = "Analysis/InitSizeModels.RData")
 	# Make supp table of results?
 	
+	rownames <- c("Population + Water + Temperature + Population:Water + Population:Temperature + Water:Temperature + Population:Water:Temperature",
+	"Population + Water + Temperature + Population:Water + Population:Temperature + Water:Temperature",
+	"Population + Water + Temperature + Population:Water + Population:Temperature",
+	"Population + Water + Temperature + Population:Water + Water:Temperature",
+	"Population + Water + Temperature + Population:Temperature + Water:Temperature",
+	"Population + Water + Temperature + Population:Water",
+	"Population + Water + Temperature + Water:Temperature",
+	"Population + Water + Temperature", "Population + Water",
+	"Population + Temperature", "Water + Temperature", "Population", "Water", "-", "-")
+
+	colnames <- c("Random", "DIC")
+	dat <- data.frame(Random = c(rep("Family", 14), "-"), 
+		DIC = round(sapply(InitSizeModels, function(X) X$DIC), 1),
+		stringsAsFactors = F)
+
+	exportTable(file = "ms/Tables/Table_InitialSize.txt", data = dat, 
+		colnames = colnames, rownames = rownames, h1 = "Model")
+
 ### JUST USING GROWTH NOW UNTIL OTHER DATA ARE READY
 ### Analysis 1: Variation in 'intrinsic' traits (significant main effect of Population)
 ### Analysis 2: Variation in plasticity (significant Treatment x Population interaction)

### 2014-11-23 Analysis/functions.R

@@ -1,4 +1,4 @@
-exportTable <- function(file, data, colnames, rownames, ...)
+exportTable <- function(file, data, colnames, rownames, h1 = "", ...)
 {
 	# Check that there are the right number of columns and rows
 	if(length(colnames) != ncol(data)) stop("number of columns in data does not match length of column names")
@@ -6,14 +6,20 @@ exportTable <- function(file, data, colnames, rownames, ...)
 	if(length(rownames) != nrow(data)) stop("number of rows in data does not match length of row names")
 
 	table <- file(file, open = "w")
-	cat(" &", colnames[1], file = table)
-	for (i in 2:length(colnames)) cat(" &", colnames[i], file = table)
+	cat(h1, "&", colnames[1], file = table)
+	if (length(colnames) > 1) 
+	{
+		for (i in 2:length(colnames)) cat(" &", colnames[i], file = table)
+	}
 	cat(" \\\\ \\midrule\n", file = table)
 	
 	for (i in 1:length(rownames))
 	{
 		cat(rownames[i], "&", data[i, 1], file = table)
-		for (j in 2:length(colnames)) cat(" &", data[i, j], file = table)
+		if (length(colnames) > 1) 
+		{
+			for (j in 2:length(colnames)) cat(" &", data[i, j], file = table)
+		}
 		cat(" \\\\\n", file = table)
 	}
 	

### 2014-11-24 ms/ms.tex
Finished supplemental table on initial size (LLL_0512) and updated exportTable function
Added more Methods detail to text and added some references.

@@
 \usepackage{color}
+\usepackage{array}
+\usepackage{gensymb}
 
@@
 Two jumbo-sized cotton balls at the bottom of cone-tainers prevented sand from washing out.
-Cone-tainers were continuously bottom-watered during germination by placing them in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) filled part way with water, placed on benches in greenhouses at the University British Columbia campus Vancouver, Canada (LAT LON of UBC).
+Cone-tainers were continuously bottom-watered during germination by placing them in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) filled part way with water, placed on benches in greenhouses at the University British Columbia campus in Vancouver, Canada (49\degree 15' N, 123\degree 15' W).
 Misters thoroughly wetted the top of the sand every two hours during the day.
@@
 We thinned seedlings to one plant per cone-tainer, leaving the center-most, largest plant.
-Where possible, extra seedlings were transplanted to cone-tainers that did not have any germinants.
-In total X\% of cone-tainers had at least one germinant.
-After transplanting, 743 of 768 (96.7\%) had plants that could be used in the experiment.
-We allowed one week at constant, non stressful conditions (20 degree day, 16 degree night) for plants to acclimate to growth chambers before starting treatments.
+702 of 768 (91.4\%) had plants that could be used in the experiment.
+We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments.
+The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments (Table S\#).
 
@@
 
+\cdm{Not sure where this goes} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package survival version 2.37-7 \citep{Therneau_2014}.
+The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
+The signifcance of the Population effect was determined using analysis of deviance.
+
 \subsection*{Intrinsic variation}
@@
 Significant Population effect indicate intrinsic trait differences; significant Population $\times$ Treatment effects indicate population differences in plasticity.
+For growth rate, we also accounted for differences in germination rate by including day of germination as a factor.
 
@@
 \bibliographystyle{evolution}
+
+%--------------------------------------------------
+% Supplementary Information
+%--------------------------------------------------
+
+% DIC table for initial size (LLL on 5/12)
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, leaf expansion rate]{Initial size of seedlings did not vary among Populations, Families, or Treatments.
+We used a censored Gaussian model of initial size at the outset of the experiment (longest leaf length of the first true leaves).
+The model was censored because we could not accurately measure leaves less than 0.25 mm with digital callipers (217 of 702, 30.9\%, were too small).
+We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package MCMCglmm version 2.17 \citep{Hadfield_2010}.
+We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in.
+We step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
+	\begin{center}
+	\begin{tabular}{>{\everypar{\hangindent1cm}{}\raggedright}p{6cm}lc}
+	\toprule
+
+	\input{Tables/Table_InitialSize.txt}
+
+	\bottomrule
+	\end{tabular}
+	\end{center}
+\end{table}
 

### 2014-11-25 Data/ProcessData.R
Added leaf trait data.

@@ -253,13 +253,24 @@ library(MASS)
 	# add Mortality
 	data$DiedFromStress <- tmp$DiedFromStress
 	
+	# add Leaf traits and compute new ones
+	data$LeafCollectionDate <- rep(NA, nrow(data))
+	x <- grep("[0-9][0-9]-[a-zA-Z]+", as.character(tmp$LeafCollectionDate))
+	data$LeafCollectionDate[x] <- 
+		(as.numeric(dmy(paste(as.character(tmp$LeafCollectionDate[x]), "-2014", sep = "")))
+		- 1399852800) / 86400
+	data$UppTrichomeDens <- tmp$UppTrichomeNo / tmp$Length_mm # No per mm
+	data$LowTrichomeDens <- tmp$LowTrichomeNo / tmp$Length_mm # No per mm
+	data$LeafArea_cm2 <- tmp$LeafArea_cm2
+	data$LMA <- 1e4 * tmp$LeafDW_g / tmp$LeafArea_cm2 # in g per m^2
+	data$LDMC <- tmp$LeafDW_g / tmp$LeafFW_g
+
 #
 #	Export for Master Analysis
 #
 
 	data <- data[which(!(is.na(data$lll_AbsGrowth) | is.na(data$height_AbsGrowth))), ]
 	write.csv(data, "MasterDatasheet_out.csv")
-
 	
 ### stuff to move possibly	
 	#

### 2015-04-21 ms/ms.Rnw
New git ignore and vastly updated ms folder

\documentclass[11pt, oneside]{article}

%
% Packages
%

\usepackage{geometry}
\geometry{letterpaper}
\usepackage[parfill]{parskip}      		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{topcapt}
\usepackage[labelfont =  {sf, bf}, textfont = sf, format = plain]{caption}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{color}
\usepackage{array}
\usepackage{gensymb}

% Make helvetica the default sans-serif font
\renewcommand\sfdefault{phv}

% Package command for citing R packages
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}

%
% Commenting
%

\newcommand{\cdm}[1]{{ \color{magenta} [{\bf{CDM:}} {\em#1}]}} % Chris' comments are in magenta.
\newcommand{\ala}[1]{{ \color{blue} [{\bf{ALA:}} {\em#1}]}} % Amy's comments are in blue.

%
% Where to find files
%

<<Front Matter>>=

  # The path to the ms directory may need to be changed for different users
  # Manuscript files
  pathMS <- "~/Google Drive/CardLocalAdaptation/ms"

  # Data files for paper
  pathDat <- paste(pathMS, "/Data", sep = "")

  # Figures for paper
  pathFig <- paste(pathMS, "/Figures", sep = "")

  # Tables for paper
  pathTab <- paste(pathMS, "/Tables", sep = "")

  # Small function to make statistical significance asterisks
  sigStar <- function(pvalue)
  {
    if (pvalue >= 0.05) return("")
    if (pvalue < 0.05 & pvalue >= 0.01) return("*")
    if (pvalue < 0.01 & pvalue >= 0.001) return("**")
    if (pvalue < 0.001) return("***")
  }

  # Color palette
  # Primary color:

   primShade0 <- rgb(1, 0, 0.286)
   primShade1 <- rgb(1, 0.643, 0.745)
   primShade2 <- rgb(1, 0.439, 0.6)
   primShade3 <- rgb(0.765, 0, 0.22)
   primShade4 <- rgb(0.588, 0, 0.169)

  # Secondary color (1):

  sec1shade0 <- rgb(0.008, 0.957, 1)
  sec1shade1 <- rgb(0.643, 0.984, 1)
  sec1shade2 <- rgb(0.443, 0.976, 1)
  sec1shade3 <- rgb(0, 0.49, 0.514)
  sec1shade4 <- rgb(0, 0.376, 0.396)

  # Secondary color (2):

  sec2shade0 <- rgb(0.988, 1, 0)
  sec2shade1 <- rgb(0.996, 1, 0.643)
  sec2shade2 <- rgb(0.992, 1, 0.439)
  sec2shade3 <- rgb(0.835, 0.847, 0)
  sec2shade4 <- rgb(0.643, 0.651, 0)

@

%
% Title and authors
%

\title{(working title) Local Physiological Adaptation in \textit{Mimulus cardinalis}}
\author{Christopher D. Muir and Amy L. Angert}
%\date{}							% Activate to display a given date or no date

%
% Start document
%

\begin{document}
\maketitle
\listoffigures
\listoftables

\section*{Abstract}

% write/analyze
% proposal for summer
  % local fitness set concave
  % range-wide fitness set convex
  % interaction between shape of fitness set and gene flow?
% evo-phys proposal

\section*{Introduction}

Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it. Local adaptation within species most often involves genetic differences in ecologically relevant traits caused by selection to different environments, leading to predictable trait-environment correlations that persist in a common garden. Although local adaptation is well-supported by data from reciprocal transplant experiments, trait-environment associations, and ecological genetics/genomics, predicting which traits respond to selection exceeds our understanding. In particular, the role of physiological traits is poorly understood, even though physiology explains adaptation to the abiotic environment, which we need to understand as a baseline for anticipating how organisms will respond to climate change. A related problem is identifying the exact selective agents, such as abiotic factors like soil water availability or biotic interactions [etc.]... Here, we simultaneously investigate which traits and causal selective agents explain range-wide patterns of local physiological adaptation to climate in \textit{Mimulus}, a model genus for local adaptation studies.

% niche-tracking versus niche-shifting
One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve. For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.

% Selective agents
We also know little about the selective agents, such as climatic factors or biotic interactions, that are most important in driving adaptive divergence. In certain situations, such as pollinator shifts and edaphic adaptation, it is straightforward to identify putative selective agents and test predicted mechanisms. However, it is more challenging for cases in which traits vary with less specific factors like mean annual precipitation or latitude. For example, body size clines (larger size at higher latitude) have evolved repeatedly on three contients in \textit{Drosophila subobscura}, strongly indicating selection. But which one or more of the many climatic factors that covary with latitude actually controls selection on body size? Selection might be even more complex for species that live over multiple seasons, meaning that environmental variation rather than the average environment may be a selective agent.

% Motivate using Mimulus
Linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation have been conducted using species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources. ... Conspicuous absence of phys mechanism

% Final paragraph
1. `intrinsic' physiological variation
2. variation in stress response
3. what ecological factors explain divergence (i.e. putative selective agents underlying local adaptation)

Conclude that observations are best explained by stabilizing selection on the fundamental niche, but spatially varying divergent selection on coordinated life history and physiological traits.

\section*{Methods}

\subsection*{Population Selection}

We used 16 populations from throughout the range of \textit{M. cardinalis} (Table 1). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.

<<focClim, echo = FALSE>>=
# Read in climate data for 16 focal populations to use in making Table 1
load('~/Google Drive/CardLocalAdaptation/Data/Climate/Robjects/focClim.RData') # focClim

@

\begin{table}[ht]
   \centering
   \topcaption{16 Focal populations}
   \begin{tabular}{@{} lllllll @{}}
      \toprule
  Name& Region  & Demo?& Pop gen?& Latitude  & Longtiude  & Elevation (mas) \\
      \midrule
	HAU & South Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "HAU"], 3)}	& 
    \Sexpr{round(focClim$Longitude[focClim$ID == "HAU"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "HAU"]}   \\
	CTC	& South Margin		& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "CTC"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "CTC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "CTC"]}   \\
	CUR	& South Margin		& yes	& yes & \Sexpr{round(focClim$Latitude[focClim$ID == "CUR"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "CUR"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "CUR"]}   \\
	GRP & South Margin		& no	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "GRP"], 3)} &
    \Sexpr{round(focClim$Longitude[focClim$ID == "GRP"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "GRP"]}   \\
	WWC &	Transverse		  & yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "WWC"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "WWC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WWC"]}   \\
	MIL	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "MIL"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "MIL"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "MIL"]}   \\
	WFM	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "WFM"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "WFM"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WFM"]}   \\
	NMT	& South Sierras		& yes	& yes?& \Sexpr{round(focClim$Latitude[focClim$ID == "NMT"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "NMT"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "NMT"]}   \\
	PRD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "PRD"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "PRD"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "PRD"]}   \\
	RWD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "RWD"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "RWD"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RWD"]}   \\
	WNA	& Central Sierras	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "WNA"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "WNA"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WNA"]}   \\
	RBW	& Central Sierras	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "RBW"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "RBW"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RBW"]}   \\
	MYU	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "MYU"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "MYU"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "MYU"]}   \\
	LIJ	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "LIJ"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "LIJ"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "LIJ"]}   \\
	DPC	& North Coast	  	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "DPC"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "DPC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "DPC"]}   \\
	RCC	& North Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "RCC"], 3)} & 
    \Sexpr{round(focClim$Longitude[focClim$ID == "RCC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RCC"]}   \\
	\bottomrule
   \end{tabular}
\end{table}

\subsection*{Plant propagation}

On 14 April, 2014, 3-5 seeds per family were sown directly on sand (Quikrete Play Sand, Georgia, USA) watered to field capacity in RLC4 Ray Leach cone-tainers placed in RL98 98-well trays (Stuewe \& Sons, Inc., Oregon, USA). We used pure sand both to facilitate root-washing and because \textit{M. cardinalis} typically grows in sandy, riparian soils (A. Angert, pers. obs.). Two jumbo-sized cotton balls at the bottom of cone-tainers prevented sand from washing out. Cone-tainers were continuously bottom-watered during germination by placing them in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) filled part way with water, placed on benches in greenhouses at the University British Columbia campus in Vancouver, Canada (49\degree 15' N, 123\degree 15' W). Misters thoroughly wetted the top of the sand every two hours during the day. Most seeds germinated between 1 and 2 weeks, but we allowed 3 weeks before transferring seedlings to growth chambers. Germination was recorded daily from one to two weeks after sowing, and every few days thereafter. On 5 May (21 days after sowing), seedlings were transferred to one of two MODEL Growth Chambers (Conviron, Manitoba, Canada). We thinned seedlings to one plant per cone-tainer, leaving the center-most, largest plant. 702 of 768 (91.4\%) had plants that could be used in the experiment. We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments. The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments (Table S\#).

\paragraph*{Monitoring environment}

  	Cross check with Poorter paper \\
		Light (PAR sensors), temp/humidity (HOBOs), soil moisture

\subsection*{Treatments}

We used 2 watering treatments on two populations on two soils. The watering treatments are: well-watered, drought.

\subsubsection*{Temperature}

We simulated typical growing season (June and July) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (High Temp) and Little Jameson (Low Temp). We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}. Daily temperatures from ClimateWNA are usually highly correlated with the air temperature recorded from data loggers in the field at these sites (A. Angert, unpub. data). Hence, the ClimateWNA temperature profiles are likely to similar to actual thermal regimes experienced by \textit{M. cardinalis} in nature. To create realistic temperature regimes, we calculated the mean temperature trend from June to July using LOESS \citep{Cleveland_etal_1992}. The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ). The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals. We fit and simulated from the VAR(2) using the package \pkg{dse} \citep{Gilbert_2014} in R. Simulated data closely resembled the autocorrelation and between-site correlation of the actual data. From simulated mean temperature, we next selected minimum and maximum daily temperatures. Mean, min, and max temperature were highly correlated at both sites. We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models. For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature. During the middle of the day, chamber temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.

\subsubsection*{Water}

      April 14 germinate
      Before May 20 (36 days): 7.5 min (complete flood) every 2 hours
      between May 20 to June 10 (57 days), down to 30 seconds, 1x per day
			Wet: daily/constant irrigation with cooled water
			Dry: start at week \#, gradually decrease watering frequency/level

			\cdm{Are there any data from California about soil water content through the season?}

<<ExperimentalDesign>>=
library(scales) # for color transparency
exptlDes <- read.csv(paste(pathDat, "/ExperimentalDesign.csv", sep = ""))

pdf(paste(pathFig, "/ExperimentalDesignBase.pdf", sep = ""), 7, 5)
par(mai = rep(1, 4))
plot(0, 0, type = "n", xlim = c(0, max(exptlDes$DayN)), ylim = c(-1, 41), 
		 axes = F, frame.plot = T, cex.lab = 1.5, xlab = "Key treatments and measurements", 
     ylab = expression(paste("Temperature (", phantom()*degree, "C)")))

# Temperature axis
axis(2, at = seq(0, 40, 10), lwd = 0, lwd.ticks = 1, las = 1,
	labels = c(expression(0*degree), expression(10*degree), expression(20*degree), 
						 expression(30*degree), expression(40*degree)))
# Axis of time on top
mtext("Days after sowing", side = 3, line = 3.5, cex = 1.5)
mtext("(Date for climate simulation)", side = 3, line = 2.25, cex = 1.5)
axis(3, at = seq(0, 90, 30), lwd = 0, lwd.ticks = 1)
axis(3, at = c(17, 47, 78), lwd = 0, lwd.ticks = 1, line = -0.5, 
     labels = c("(June)", "(July)", "(August)"))

# Indicate sowing, beginning of temperature and drought treatments
abline(v = 0, col = "grey", lwd = 2, lty = 2)
mtext("Sow", side = 1, at = 0, line = 0)
abline(v = 17, col = "grey", lwd = 2, lty = 2)
mtext("Temp\nTreatment", side = 1, at = 17, line = 1)
abline(v = 37, col = "grey", lwd = 2, lty = 2)
mtext("Drought\nTreatment", side = 1, at = 37, line = 1)

# Polygon of Hot treatment, min and max
with(exptlDes[18:94, ], polygon(c(17:93, 93:17), c(Tmin_hot, rev(Tmax_hot)),
  border = NA, col = alpha(primShade0, 0.25)))

# Polygon of Cool treatment, min and max
with(exptlDes[18:94, ], polygon(c(17:93, 93:17), c(Tmin_cool, rev(Tmax_cool)),
  border = NA, col = alpha(sec1shade0, 0.25)))

# Lines with average temperature
with(exptlDes[18:94, ], points(17:93, Tavg_hot, lwd = 2, type = "l", 
	col = primShade3))
with(exptlDes[18:94, ], points(17:93, Tavg_cool, lwd = 2, type = "l", 
	col = sec1shade3))

# Rug indicating measurment timing
rect(6, -2, 17, 0, border = "white", col = "grey")
with(exptlDes, text(11.5, -1, labels = "Germ", col = "white", font = 3, 
  cex = 0.9))

with(exptlDes, rect(min(which(LERmeas)), -2, max(which(LERmeas)), 0, 
	border = "white", col = "grey"))
with(exptlDes, text(min((which(LERmeas)) + max(which(LERmeas))) / 2, -1, 
	labels = "LER", col = "white", font = 3, cex = 0.9))

with(exptlDes, rect(min(which(SERmeas)), 0, max(which(SERmeas)), 2,
	border = "white", col = "grey"))
with(exptlDes, text(min((which(SERmeas)) + max(which(SERmeas))) / 2, 1, 
	labels = "SER", col = "white", font = 3, cex = 0.9))

with(exptlDes, rect(min(which(Harvest)), -2, max(which(Harvest)), 0, 
	border = "white", col = "grey"))
with(exptlDes, text(min((which(Harvest)) + max(which(Harvest))) / 2, -1, 
	labels = "Harvest", col = "white", font = 3, cex = 0.9))

with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 5, 
	labels = "Photo", col = "grey", font = 3, cex = 0.9))
with(exptlDes, rect(min(which(PhotoMeas)), 2, max(which(PhotoMeas)), 4, 
	border = "white", col = "grey"))
#with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 3, 
#	labels = "Photo", col = "white", font = 3, cex = 0.9))

dev.off()

# Legend
pdf(paste(pathFig, "/ExperimentalDesignLegend.pdf", sep = ""), 1.5, 2)
par(mai = rep(0, 4))
plot(0, 0, type = "n", xlim = c(0, 1.5), ylim = c(0, 1), axes = F)

polygon(c(0, 0.5, 1, 1, 0.5, 0), c(0.8, 1, 1, 0.55, 0.55, 0.35), border = NA, 
				col = alpha(primShade0, 0.25))
polygon(c(0, 0.5, 1, 1, 0.5, 0), c(0.6, 0.45, 0.45, 0, 0, 0.15), border = NA, 
				col = alpha(sec1shade0, 0.25))
points(c(0, 0.5, 1), c(0.575, 0.775, 0.775), lwd = 2, type = "l", col = primShade3)
points(c(0, 0.5, 1), c(0.375, 0.225, 0.225), lwd = 2, type = "l", col = sec1shade3)
text(0.5, 0.9, labels = "Hot", col = primShade3)
text(0.5, 0.1, labels = "Cool", col = sec1shade3)
text(1.25, 1, labels = "Max", col = alpha(primShade0, 0.25))
text(1.25, 0.55, labels = "Min", col = alpha(primShade0, 0.25))
text(1.25, 0.775, labels = "Mean", col = primShade3)

text(1.25, 0, labels = "Min", col = alpha(sec1shade0, 0.25))
text(1.25, 0.45, labels = "Max", col = alpha(sec1shade0, 0.25))
text(1.25, 0.225, labels = "Mean", col = sec1shade3)

dev.off()

@

\subsection*{Measurements: Growth and photosynthesis}

\begin{table}[ht]
   \centering
   \topcaption{Traits}
   \begin{tabular}{@{} ll @{}}
      \toprule
  Trait & Units \\
      \midrule
  Day of germination  & day \\
  Leaf expansion rate  &  mm day$^{-1}$  \\
  Shoot elongation rate  &  mm or cm day$^{-1}$  \\
  Harvest dry mass &  g  \\
  Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
  Mortality &  Probability?  \\
	    \bottomrule
   \end{tabular}
\end{table}

\paragraph{Day of germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}. The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function. The signifcance of the Population effect was determined using analysis of deviance.

<<Germination rate>>=

  library(survival)
  pathMS <- "~/Google Drive/CardLocalAdaptation/ms"
  pathDat <- paste(pathMS, "/Data", sep = "")

  master <- read.csv(paste(pathDat, "/MasterDatasheet_out.csv", sep = ""), 
    row.names = 1)

  # Create Surv object
	germ <- with(subset(master, !is.na(master$MinGermDay)), Surv(MinGermDay, MaxGermDay,	
		type = "interval2"))

	# lognormal produced best fit to data
	fitGerm <- survreg(germ ~ PopID + frailty(Family, sparse = F), 
		data = subset(master, !is.na(master$MinGermDay)), dist = "lognormal")
	suppressWarnings(anova(fitGerm)) # significant family and population effects

@

\paragraph{Growth rate: leaf expansion and shoot elongation}

We censused longest leaf length $>1$ mm on 10 days (twice per week) between 12 May and 12 June (28 to 59 days after sowing). We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.  We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing). Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.

<<Leaf expansion rate>>=

  library(lme4)
  library(lmerTest)

  # New (blank) datasheet on which to add data
  newdata <- read.csv(paste(pathDat, "/MasterDatasheet_in.csv", sep = ""))

  # Read in LER data
  LERdata <- read.csv(paste(pathDat, "/LERdata.csv", sep = ""))

  #
	# Empirical Bayes Approach: get coefficients for every plant
	#

	# suppressWarnings(mmLER <- lmer(log(LLL) ~ poly(DayN, 2, raw = T) + 
    # (poly(DayN, 2, raw = T)|indiv), data = LERdata))
  # save(mmLER, file = paste(pathDat, "/mmLER.RData", sep = ""))
  load(file = paste(pathDat, "/mmLER.RData", sep = ""))

  # Extract coefficients and add to datasheet
	Y <- t(unlist(fixef(mmLER)) + t(ranef(mmLER)$indiv))[match(newdata$indiv, 
		rownames(ranef(mmLER)$indiv)), ]
	colnames(Y) <- c("b0_LER", "b1_LER", "b2_LER")
  newdata <- cbind(newdata, Y)

  # Model-based LER
	# Timeframe for Cool treatment: Day 0 - 31
	# Timeframe for Hot treatment: Day 0 - 24
	newdata$LERstart <- exp(newdata$b0_LER)
	newdata$LERend <- exp(newdata$b0_LER + 
					newdata$b1_LER * ifelse(newdata$TempTrt == "Cool", 31, 24) + 
					newdata$b2_LER * ifelse(newdata$TempTrt == "Cool", 31 ^ 2, 24 ^ 2))
	newdata$LER_AbsGrowth <- (newdata$LERend - newdata$LERstart) / 
		ifelse(newdata$TempTrt == "Cool", 31, 24)

  # Mixed model ANOVA with family as random factor fit using step down procedure
	mm <- lmer(LER_AbsGrowth ~ AvgGermDay + PopID * TempTrt * WaterTrt + (1|Family),
		data = master)
	fitLER <- step(mm, reduce.random = F) # includes pop, temp, h2o, population x water
	# fitLER <- step(mm, ddf = "Kenward-Roger", reduce.random = F) # same result using Satterthwaite ddf

@

<<Shoot elongation rate>>=

  # Read in SER data
  SERdata <- read.csv(paste(pathDat, "/SERdata.csv", sep = ""))

  #
  # Empirical Bayes Approach: get coefficients for every plant
	#

	# suppressWarnings(mmSER <- lmer(log(height + 1) ~ poly(DayN, 2, raw = T) + 
    # (poly(DayN, 2, raw = T)|indiv), data = SERdata))
  # save(mmSER, file = paste(pathDat, "/mmSER.RData", sep = ""))
  load(file = paste(pathDat, "/mmSER.RData", sep = ""))

  # Extract coefficients and add to datasheet
  Y <- t(unlist(fixef(mmSER)) + t(ranef(mmSER)$indiv))[match(newdata$indiv, 
		rownames(ranef(mmSER)$indiv)), ]
	colnames(Y) <- c("b0_SER", "b1_SER", "b2_SER")
  newdata <- cbind(newdata, Y)

  # Model-based SER
	# Timeframe for Cool treatment: Day 0 - 31
	# Timeframe for Hot treatment: Day 0 - 24
	newdata$SERstart <- exp(newdata$b0_SER)
	newdata$SERend <- exp(newdata$b0_SER + 
					newdata$b1_SER * ifelse(newdata$TempTrt == "Cool", 31, 24) + 
					newdata$b2_SER * ifelse(newdata$TempTrt == "Cool", 31 ^ 2, 24 ^ 2))
	newdata$SER_AbsGrowth <- (newdata$SERend - newdata$SERstart) / 
		ifelse(newdata$TempTrt == "Cool", 31, 24)

  mm <- lmer(SER_AbsGrowth ~ AvgGermDay + PopID * TempTrt * WaterTrt + 
		(1|Family), data = master)
	fitSER <- step(mm, reduce.random = F) #includes pop, temp, h2o, temp x water
@

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity. During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ \cdm{get exact temps used}) so that all plants within a temperature treatment were measured under the same conditions.

<<Photosynthesis>>=

  # Photosynthetic rate
  mm <- lmer(Photo ~ PopID * TempTrt * WaterTrt + (1|Family),
    data = subset(master, !is.na(master$Photo)))
  fitPhoto <- step(mm, reduce.random = F)

  # Intrinsic photosynthetic rate
  # = residuals of Conductance - Photosynthesis regression
  # Removed one outlier (may change if data are rearranged!)
	mm <- lmer(resPhoto ~ PopID * TempTrt * WaterTrt + (1|Family),
    data = subset(master, !is.na(master$Photo))[-52, ])
	fitIntrnPhoto <- step(mm, reduce.random = F)

@

\paragraph{Mortality}
We assayed mortality during twice-weekly growth measurements. We could not get GLMM with Family effects to converge, so we used GLM with a quasibinomial error structure and assessed signifiance using Type 2 Analysis of Deviance with the R package \pkg{car}.

<<Mortality>>=

  library(car)

  fitMort <- glm(DiedFromStress ~ PopID * TempTrt * WaterTrt,
    data = subset(master, !is.na(master$DiedFromStress)), family = "quasibinomial")

  Anova(fitMort, type = 2)
@

\paragraph{Biomass (show correlation between growth rate and biomass?)}

\subsection*{Intrinsic variation and plasticity}

\cdm{which traits: germination, growth rate, biomass, biomass allometry, leaf traits, photosynthesis?, mortality?}
We tested for Population, Treatment, and Population $\times$ Treatment interactions using mixed model ANOVA with Family included as a random factor. Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}. Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-11 \citep{Kuznetsova_etal_2014}. Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation. Significant Population effect indicate intrinsic trait differences; significant Population $\times$ Treatment effects indicate population differences in plasticity. For growth rate, we also accounted for differences in germination rate by including day of germination as a factor.

\subsection*{Principal components of germination, growth, and photosynthesis}

<<Principal Components>>=

  popMeans <- data.frame(
    germ = c(coef(fitGerm)[1], coef(fitGerm)[1] + coef(fitGerm)[2:16]),
    LER = fitLER$lsmeans.table[1:16, "Estimate"],
    SER = fitSER$lsmeans.table[1:16, "Estimate"],
    photo = fitIntrnPhoto$lsmeans.table[1:16, "Estimate"])

  traitPC <- prcomp(popMeans, center = T, scale. = T)
  focClim$PC1 <- traitPC$x[, "PC1"]

  # Percent variance explained by first principal component
  pve <- paste(round(100 * traitPC$sdev[1] ^ 2 / sum(traitPC$sdev ^ 2), 1), "%")

  # Plot latitude versus PC1
  pdf(paste(pathFig, "/Figure_PC1vLat.pdf", sep = ""), 5, 5)
  par(mar = c(5, 5, 1, 1))
  plot(focClim$Lat, focClim$PC1, xlim = c(32, 44), ylim = c(-4, 4), axes = F,
    frame.plot = T, type = "n", xlab = "", ylab = "")
  title(xlab = "Latitude", cex.lab = 1.5)
  title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
    list(x = pve)))
  axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), expression(36*degree),
    expression(40*degree), expression(44*degree)), lwd = 0, lwd.ticks = 1) 
  axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
  fitLat <- lm(PC1 ~ Latitude, data = focClim)

  # polygon of confidence intervals
  x <- seq(min(focClim$Latitude), max(focClim$Latitude), l = 1e3)
  u <- predict(fitLat, new = data.frame(Latitude = x))
  s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitLat) %*% c(1, X)))
  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), upperCI = qnorm(0.975, u, s))
  polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
    col = "grey80", border = NA)
  lines(x, confInt$lowerCI, lwd = 2, lty = 2)
  lines(x, confInt$upperCI, lwd = 2, lty = 2)
  lines(range(focClim$Lat), rev(range(predict(fitLat))), lwd = 2)
  points(focClim$Lat, focClim$PC1, pch = 21, col = "black", bg = primShade0, cex = 1.5,
    lwd = 2)
  dev.off()

     # plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
  #plot(focClim$Lat, traitPC$x[, "PC1"])
  #cor.test(focClim$Lat, traitPC$x[, "PC1"])
  #cor.test(popMeans[, 4], traitPC$x[, "PC1"])
  #plot(focClim$bio16_cva, traitPC$x[, "PC1"])
@

\subsection*{Selective agents and environmental correlates}

<<VSURF>>=

library(VSURF)
  # Load VSURF results of climatic variables associated with latitude at cardinalis occurence points
  # Note that occurences have been randomly thinned to account for variation in sampling effort
  load(paste(pathDat, "/clim2lat_fit_t1.RData", sep = "")) # fit_t1
  vsurfLat <- fit_t1

  # Figure of variable importance for latitude
  pdf(paste(pathFig, "/Figure_clim2lat_imp.pdf", sep = ""), 4, 5)
  par(mar = c(5, 6, 1, 1))
  plot(0, 0, type = "n", xlim = c(0, ceiling(vsurfLat$ord.imp$x)[1]), ylim = c(1, 
    length(vsurfLat$varselect.pred)), axes = F, frame.plot = T, xlab = "Importance", 
    ylab = "", cex.lab = 1.5)
  for (i in 1:length(vsurfLat$varselect.pred)) 
  {
     lines(x = c(-1, length(vsurfLat$varselect.pred)), y = c(i, i), col = "grey")
  }
  points(vsurfLat$ord.imp$x[1:length(vsurfLat$varselect.pred)], 
    length(vsurfLat$varselect.pred):1, pch = 21, col = "black", bg = "white", cex = 1.5)
	mtext(climVars[vsurfLat$varselect.pred], side = 2, 
    at = length(vsurfLat$varselect.pred):1, las = 1, line = 1)
	axis(1, lwd = 0, lwd.ticks = 1)
	dev.off()

  # Columns with climatic variables to test
  load(paste(pathDat, "/Exploratory_climate_variables.RData", sep = "")) #climVars

  # Variable selection using random forests (VSURF)
  linmod.PC1 <- as.formula(paste("PC1 ~ ", paste(climVars, collapse = " + "), sep = ""))
  set.seed(69646)
  vsurfPC1 <- VSURF.parallel(linmod.PC1, data = focClim[, c("PC1", climVars)], 
    ncores = 3)

  # Plot most important climatic predictor versus PC1
  pdf(paste(pathFig, "/Figure_bio16cva_v_Lat.pdf", sep = ""), 5, 5)
  par(mar = c(5, 5, 1, 1))
  plot(focClim$bio16_cva, focClim$PC1, xlim = c(0.25, 0.55), ylim = c(-4, 4), axes = F,
    frame.plot = T, type = "n", xlab = "", ylab = "")
  title(xlab = "Precipitation of Wettest Quarter", cex.lab = 1.5)
  title(xlab = "(Coefficient of variation)", cex.lab = 1, line = 4)
  title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
    list(x = pve)))
  axis(1, lwd = 0, lwd.ticks = 1) 
  axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
  fitBio16cva <- lm(PC1 ~ bio16_cva, data = focClim)

  # polygon of confidence intervals
  x <- seq(min(focClim$bio16_cva), max(focClim$bio16_cva), l = 1e3)
  u <- predict(fitBio16cva, new = data.frame(bio16_cva = x))
  s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitBio16cva) %*% c(1, X)))
  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), upperCI = qnorm(0.975, u, s))
  polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
    col = "grey80", border = NA)
  lines(x, confInt$lowerCI, lwd = 2, lty = 2)
  lines(x, confInt$upperCI, lwd = 2, lty = 2)
  lines(range(focClim$bio16_cva), range(predict(fitBio16cva)), lwd = 2)
  points(focClim$bio16_cva, focClim$PC1, pch = 21, col = "black", bg = primShade0, 
    cex = 1.5, lwd = 2)
  dev.off()

  # Overlap between latitude predictors and phenotype predictors
  imp2 <- climVars[vsurfPC1$varselect.pred]
  nImpPC1 <- length(imp2)

  vsurfLat$varselect.pred
  imp1 <- climVars[vsurfLat$varselect.pred]
  nImpLat <- length(imp1)

  # Figure of variable importance for traits
	pdf(paste(pathFig, "/Figure_clim2pc1_imp.pdf", sep = ""), 4, 5)
	par(mar = c(5, 1, 1, 6))

	plot(0, 0, type = "n", xlim = c(0, 30), ylim = c(1, nImpLat), 
		axes = F, frame.plot = T, xlab = "Importance", ylab = "", cex.lab = 1.5)
	for (i in (nImpLat:(nImpLat - nImpPC1 + 1))) lines(x = c(-100, 100), y = c(i, i), 
		col = "grey")
	points(100 * vsurfPC1$ord.imp$x[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), pch = 21, 
		col = "black", bg = "white", cex = 1.5)
	mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, 
    at = nImpLat:(nImpLat - nImpPC1 + 1), 
		font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
	axis(1, lwd = 0, lwd.ticks = 1, labels = seq(0, 3, 0.5), at = seq(0, 30, 5))
  dev.off()
@

(this also goes into the goal of trying to figure out ecological selective agent driving trait variation). Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude? Some hypotheses:
\begin{itemize}
  \item{H1: Latitude encompasses multivariate climatic dimension.}
  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
\end{itemize}

Predictions:
\begin{itemize}
  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]). Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
\end{itemize}

Methods specific to latitude vs. trait correlation:
We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We used a comprehensive set of \cdm{358 in total before thinning} recent (since 2000) known \textit{M. cardinalis} occurences recorded from herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included X environmental variables from ClimateWNA and 17? bioclimatic variables (Table S\#). For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation). Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.

Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration. We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation. We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations. Since \textit{M. cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}. Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.

To identify climatic variables that were closely associated with latitude in \textit{M. cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}. In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.

Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions. If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.

Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection. Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species. To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...

% Methods for spatial-average selection
For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.

\begin{table}[ht]
   \centering
   \topcaption{Climatic variables}
   \begin{tabular}{@{} ll @{}}
      \toprule
  Climate variable & Abbreviation \\
      \midrule
  % climateWNA VARIABLES:
	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
	NFFD	& number of frost-free days                           \\
	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
	Eref	& Hargreaves reference evaporation (mm) \\
	CMD	  & Hargreaves climativ moisture deficit (mm) \\
	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
	RH    &	mean annual relative humidity \\
  %BIOCLIM VARIABLES:
  bio1	&	Annual Mean Temperature	\\
	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
	bio4	&	Temperature Seasonality (standard deviation *100)	\\
	bio5	&	Max Temperature of Warmest Month	\\
	bio6	&	Min Temperature of Coldest Month	\\
	bio7	&	Temperature Annual Range (bio5-bio6)	\\
	bio8	&	Mean Temperature of Wettest Quarter	\\
	bio9	&	Mean Temperature of Driest Quarter	\\
	bio10	&	Mean Temperature of Warmest Quarter	\\
	bio11	&	Mean Temperature of Coldest Quarter	\\
	bio12	&	Annual Precipitation	\\
	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
	bio16	&	Precipitation of Wettest Quarter	\\
	bio17	&	Precipitation of Driest Quarter	\\
	bio18	&	Precipitation of Warmest Quarter	\\
	bio19	&	Precipitation of Coldest Quarter	\\
	    \bottomrule
   \end{tabular}
\end{table}

\section*{Results}

\begin{table}[ht]
   \centering
   \topcaption{Population, treatment effects}
   \begin{tabular}{@{} llllllll @{}}
      \toprule
                                   & Germination & Leaf expansion & Stem elongation & Photosynthesis & Intrin. Photo & Mortality \\
                                   & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
      \midrule
  Population                       &
    \Sexpr{sigStar(anova(fitGerm)["Pr(>Chi)"]["PopID",])}  &
    \Sexpr{sigStar(fitLER$anova.table["PopID", "Pr(>F)"])} &
    \Sexpr{sigStar(fitSER$anova.table["PopID", "Pr(>F)"])} &
    \Sexpr{sigStar(fitPhoto$anova.table["PopID", "Pr(>F)"])} & 
    \Sexpr{sigStar(fitIntrnPhoto$anova.table["PopID", "Pr(>F)"])} & 
    \Sexpr{sigStar(Anova(fitMort, type = 2)["PopID", "Pr(>Chisq)"])} & \\
  Temperature                      &
    NA                                                          &
    \Sexpr{sigStar(fitLER$anova.table["TempTrt", "Pr(>F)"])}    &
    \Sexpr{sigStar(fitSER$anova.table["TempTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitPhoto$anova.table["TempTrt", "Pr(>F)"])} & 
    \Sexpr{sigStar(fitIntrnPhoto$anova.table["TempTrt", "Pr(>F)"])} &     \Sexpr{sigStar(Anova(fitMort, type = 2)["TempTrt", "Pr(>Chisq)"])} & \\
  Water                            &
    NA                                                          &
    \Sexpr{sigStar(fitLER$anova.table["WaterTrt", "Pr(>F)"])}   &
    \Sexpr{sigStar(fitSER$anova.table["WaterTrt", "Pr(>F)"])}   &
    \Sexpr{sigStar(fitPhoto$anova.table["WaterTrt", "Pr(>F)"])} & 
    \Sexpr{sigStar(fitIntrnPhoto$anova.table["WaterTrt", "Pr(>F)"])} &     
    \Sexpr{sigStar(Anova(fitMort, type = 2)["WaterTrt", "Pr(>Chisq)"])} & \\
  Pop $\times$ Temp                &
    NA                                                          &
    \Sexpr{sigStar(fitLER$anova.table["PopID:TempTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitSER$anova.table["PopID:TempTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitPhoto$anova.table["PopID:TempTrt", "Pr(>F)"])} & 
    \Sexpr{sigStar(fitIntrnPhoto$anova.table["PopID:TempTrt", "Pr(>F)"])} &     
    \Sexpr{sigStar(Anova(fitMort, type = 2)["PopID:TempTrt", "Pr(>Chisq)"])} & \\
  Pop $\times$ Water               &
     NA                                                          &
    \Sexpr{sigStar(fitLER$anova.table["PopID:WaterTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitSER$anova.table["PopID:WaterTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitPhoto$anova.table["PopID:WaterTrt", "Pr(>F)"])} & 
    \Sexpr{sigStar(fitIntrnPhoto$anova.table["PopID:WaterTrt", "Pr(>F)"])} &     
    \Sexpr{sigStar(Anova(fitMort, type = 2)["PopID:WaterTrt", "Pr(>Chisq)"])} & \\
  Temp $\times$ Water              &
    NA                                                          &
    \Sexpr{sigStar(fitLER$anova.table["TempTrt:WaterTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitSER$anova.table["TempTrt:WaterTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitPhoto$anova.table["TempTrt:WaterTrt", "Pr(>F)"])} & 
    \Sexpr{sigStar(fitIntrnPhoto$anova.table["TempTrt:WaterTrt", "Pr(>F)"])} &     
    \Sexpr{sigStar(Anova(fitMort, type = 2)["TempTrt:WaterTrt", "Pr(>Chisq)"])} & \\
  Pop $\times$ Temp $\times$ Water &
    NA                                                          &
    \Sexpr{sigStar(fitLER$anova.table["PopID:TempTrt:WaterTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitSER$anova.table["PopID:TempTrt:WaterTrt", "Pr(>F)"])} &
    \Sexpr{sigStar(fitPhoto$anova.table["PopID:TempTrt:WaterTrt", "Pr(>F)"])} & 
    \Sexpr{sigStar(fitIntrnPhoto$anova.table["PopID:TempTrt:WaterTrt", "Pr(>F)"])} &     
    \Sexpr{sigStar(Anova(fitMort, type = 2)["PopID:TempTrt:WaterTrt", "Pr(>Chisq)"])} & \\
      \bottomrule
   \end{tabular}
\end{table}

\subsection*{Coordinated latitudinal clines in germination, growth, and photosynthesis}

\subsection*{Little evidence population by treatment interactions}

\subsection*{Latitude and phenotypic clines are associated with more variable precipitation}

% ANOVA table for LLL growth rate (need to label tables)
\begin{table}[htbp]
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[ANOVA table, leaf expansion rate]{CAPTION}
	\begin{center}
	\begin{tabular}{lcccccc}
	\toprule

	\input{Tables/Table_lllGrowthAnova.txt}

	\bottomrule
	\end{tabular}
	\end{center}
\end{table}

% ANOVA table for height growth rate (need to label tables)
\begin{table}[htbp]
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[ANOVA table, stem expansion rate]{CAPTION}
	\begin{center}
	\begin{tabular}{lcccccc}
	\toprule

	\input{Tables/Table_heightGrowthAnova.txt}

	\bottomrule
	\end{tabular}
	\end{center}
\end{table}

\subsection*{Latitude versus growth rate}

\begin{figure}[h!]
	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations germinate sooner.]{CAPTION}
	\label{fig:Fig_DoG}
\end{figure}

\begin{figure}[h!]
	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
	\label{fig:Fig_LLL}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations grow faster (stem elongation rate).]{CAPTION}
	\label{fig:Fig_height}
\end{figure}

\section*{Discussion}

% stuff about environmental grain
How organisms adapt depends on the grain of the environment, ..., and gene flow between populations

Possibilities:
1. no trait variation: no environmental variation, high gene flow, generalist (convex)
2. variation in mean:
3. variation in plasticity: convex?,

our data are most consistent with:
- concave tradeoffs, fine-grained (coarse+concave would predict alt morphs), high gene flow

%--------------------------------------------------
% References
%--------------------------------------------------

\setlength{\bibsep}{6pt}
\bigskip

\bibliography{refs}
\bibliographystyle{evolution}

%--------------------------------------------------
% Supplementary Information
%--------------------------------------------------

% DIC table for initial size (LLL on 5/12)
\begin{table}[htbp]
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[ANOVA table, leaf expansion rate]{Initial size of seedlings did not vary among Populations, Families, or Treatments. We used a censored Gaussian model of initial size at the outset of the experiment (longest leaf length of the first true leaves). The model was censored because we could not accurately measure leaves less than 0.25 mm with digital callipers (217 of 702, 30.9\%, were too small). We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package \pkg{MCMCglmm} version 2.17 \citep{Hadfield_2010}. We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in. We step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
	\begin{center}
	\begin{tabular}{>{\everypar{\hangindent1cm}{}\raggedright}p{6cm}lc}
	\toprule

	\input{Tables/Table_InitialSize.txt}

	\bottomrule
	\end{tabular}
	\end{center}
\end{table}

\end{document}

	\subsection{Structure}

	Leaf traits method: \\
	Day 1: \\
	- select focal leaf (e.g. second or third leaf); \\
	- place in H20, dark, cool overnight. \\
	Day 2: \\
	- wet weight; \\
	- scan; \\
	- place in drying oven. \\

		\begin{enumerate}
			\item{Stomatal density, guard cell length, ratio from leaf opposite focal leaf (above}
			\item{Leaf size/shape (from scans)}
			\item{Leaf thickness/SLA}
			\item{Colorimetry (from scans, spec)}
			\item{stem anthocyanin (qualitative)}
		\end{enumerate}

	OPEN QUESTIONS: \\
	save wet samples in preservative (e.g. for leaf venation?) \\
	chlorophyll content? \\
	branching architecture (anything in plant protocols?)

	\subsection{Function}

		\subsubsection{during experiment}

		\begin{enumerate}
			\item{Instantaneous photosynthetic rate, $g_\text{s}$, water-use efficiency.}
				\subitem{3 min per plant per LICOR}
				\subitem{Do as many plants as possible in 1-2 weeks}
				\subitem{Before and after drought?}
			\item{Respiration}
			\item{Whole plant water-use (gravimetric)}
				\subitem{}
			\item{Leaf temperature (subset)}
			\item{Temperature response curve (subset)}
				\subitem{see protocol here: http://prometheuswiki.publish.csiro.au/tiki-index.php?page=Temperature+response+of+photosynthesis+using+a+Li-Cor+6400}
			\item{Light response curve}
				\subitem{Sampling? Only in cool, wet?}
				\subitem{At what temperature?}

		\end{enumerate}

		\subsubsection{after experiment}

		\begin{enumerate}
			\item{CN content}
			\item{carbon isotope discrimination}
		\end{enumerate}

	\subsection{Development time and Life history}

	\begin{enumerate}
		\item{Flowering time: census daily}
		\item{Starch accumulation: collect (subset) of plants}
		\item{Rhizome formation: collect on subset of plants harvested for dry mass}
	\end{enumerate}

	\subsection{Floral (2nd/3rd/etc flower?)}

	\begin{enumerate}
		\item{Flower size}
		\item{Herkogamy}
	\end{enumerate}

\section{Performance}

	Possibilities: \\
	1. (sub-)weekly census of plant height. $\approx$ 0.5 min per plant, so allow 1 day \\

	$$RGR_\text{height} = \frac{\text{ln} H_2 - \text{ln} H_1}{t_2 - t_1}$$

	2. early development (pre-treatment) leaf-area growth with camera \\
	3. harvest dry biomass before and after treatments (get whole plant leaf area using scanner at same time to ground truth camera)

	$$RGR_\text{mass} = \frac{\text{ln} M_2 - \text{ln} M_1}{t_2 - t_1}$$

\section{Determining ecological gradients that matter}

Performance tradeoffs
Isolation by ecology

\section{Tests of divergent selection}

	\subsection{Trait-environment correlation}

	We will account for relatedness by using genetic distance matrix in linear models.

	\subsection{$P_\text{ST}$ -- $F_\text{ST}$}

	Contrast trait classes with Fst distribution

	\subsection{Isolation by ecology}

	Is \textbf{G} better predicted by \textbf{E} than geographic distance? Use Gideon's method?

	\subsection{Performance tradeoffs between treatments}

\section{Ecological strategy, physiological mechanism, and tradeoff}

	\subsection{Drought escape}
		Pattern: \\
			Flowering time shorter in drier/S populations \\
			More between population variation in flowering time more variable than predicted by Fst \\
			... \\

	\subsection{Drought avoidance}
		Pattern: \\
			Maybe this should really be dehydration avoidance? \\
			Populations from drier/S avoid dehydration by limiting water loss; more variable than predicted by Fst \\

			Mechanism \\
			closing stomata earlier during soil drying

		Mechanisms: \\
			lower water-use \\

	\subsection{Drought tolerance}
		Pattern: \\
			drier/S populations have greater tolerance to low tissue water content

		Mechanisms: \\
			greater WUE \\
			ability to photosynthesize during drought \\

\section{Extension: incipient speciation as a byproduct or divergence?}

Collect pollen to measure viability.

\end{document}

### 2015-04-21 ms/ms.tex
New git ignore and vastly updated ms folder

@@
-\documentclass[11pt, oneside]{article}
+\documentclass[11pt, oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 
+% Package command for citing R packages
+\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
+
 %
@@
 \newcommand{\ala}[1]{{ \color{blue} [{\bf{ALA:}} {\em#1}]}} % Amy's comments are in blue.
+
+%
+% Where to find files
+%
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+  \hlcom{# The path to the ms directory may need to be changed for different users}
+  \hlcom{# Manuscript files}
+  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"}
+
+  \hlcom{# Data files for paper}
+  \hlstd{pathDat} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Data"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+
+  \hlcom{# Figures for paper}
+  \hlstd{pathFig} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Figures"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+
+  \hlcom{# Tables for paper}
+  \hlstd{pathTab} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Tables"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+
+  \hlcom{# Small function to make statistical significance asterisks}
+  \hlstd{sigStar} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{pvalue}\hlstd{)}
+  \hlstd{\{}
+    \hlkwa{if} \hlstd{(pvalue} \hlopt{>=} \hlnum{0.05}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{""}\hlstd{)}
+    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.05} \hlopt{&} \hlstd{pvalue} \hlopt{>=} \hlnum{0.01}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"*"}\hlstd{)}
+    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.01} \hlopt{&} \hlstd{pvalue} \hlopt{>=} \hlnum{0.001}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"**"}\hlstd{)}
+    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.001}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"***"}\hlstd{)}
+  \hlstd{\}}
+
+  \hlcom{# Color palette}
+  \hlcom{# Primary color:}
+
+   \hlstd{primShade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.286}\hlstd{)}
+   \hlstd{primShade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0.643}\hlstd{,} \hlnum{0.745}\hlstd{)}
+   \hlstd{primShade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0.439}\hlstd{,} \hlnum{0.6}\hlstd{)}
+   \hlstd{primShade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.765}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.22}\hlstd{)}
+   \hlstd{primShade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.588}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.169}\hlstd{)}
+
+  \hlcom{# Secondary color (1):}
+
+  \hlstd{sec1shade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.008}\hlstd{,} \hlnum{0.957}\hlstd{,} \hlnum{1}\hlstd{)}
+  \hlstd{sec1shade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.643}\hlstd{,} \hlnum{0.984}\hlstd{,} \hlnum{1}\hlstd{)}
+  \hlstd{sec1shade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.443}\hlstd{,} \hlnum{0.976}\hlstd{,} \hlnum{1}\hlstd{)}
+  \hlstd{sec1shade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.49}\hlstd{,} \hlnum{0.514}\hlstd{)}
+  \hlstd{sec1shade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.376}\hlstd{,} \hlnum{0.396}\hlstd{)}
+
+  \hlcom{# Secondary color (2):}
+
+  \hlstd{sec2shade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.988}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0}\hlstd{)}
+  \hlstd{sec2shade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.996}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.643}\hlstd{)}
+  \hlstd{sec2shade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.992}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.439}\hlstd{)}
+  \hlstd{sec2shade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.835}\hlstd{,} \hlnum{0.847}\hlstd{,} \hlnum{0}\hlstd{)}
+  \hlstd{sec2shade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.643}\hlstd{,} \hlnum{0.651}\hlstd{,} \hlnum{0}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 %
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
+% write/analyze
+% proposal for summer
+  % local fitness set concave
+  % range-wide fitness set convex
+  % interaction between shape of fitness set and gene flow?
+% evo-phys proposal
+
 \section*{Introduction}
 
-% What's the big picture (this is TOO big):
-% It is taken for granted that genes are the foundation of evolution.
-If we get to the genetic basis of adaptive evolution, then we have understood the process at its deepest level.
-I believe this view is misguided.
-Genetic change is, to the best of our knowledge, the basis of evolution, but this does not imply that by understanding genetic variation we can predict phenotypic evolution.
-Rather, variation and constraint on phenotypic evolution are often determined by what goes on outside.
-Organisms must survive and reproduce in the real world of physical laws and interactions with other organisms.
-Physiology, the study of organismal function, tells us about the variation and constraint on ways in which organisms can meet these challenges.
+Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
+Local adaptation within species most often involves genetic differences in ecologically relevant traits caused by selection to different environments, leading to predictable trait-environment correlations that persist in a common garden.
+Although local adaptation is well-supported by data from reciprocal transplant experiments, trait-environment associations, and ecological genetics/genomics, predicting which traits respond to selection exceeds our understanding.
+In particular, the role of physiological traits is poorly understood, even though physiology explains adaptation to the abiotic environment, which we need to understand as a baseline for anticipating how organisms will respond to climate change.
+A related problem is identifying the exact selective agents, such as abiotic factors like soil water availability or biotic interactions [etc.]...
+Here, we simultaneously investigate which traits and causal selective agents explain range-wide patterns of local physiological adaptation to climate in \textit{Mimulus}, a model genus for local adaptation studies.
+
+% niche-tracking versus niche-shifting
+One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve.
+For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.
+
+% Selective agents
+We also know little about the selective agents, such as climatic factors or biotic interactions, that are most important in driving adaptive divergence.
+In certain situations, such as pollinator shifts and edaphic adaptation, it is straightforward to identify putative selective agents and test predicted mechanisms.
+However, it is more challenging for cases in which traits vary with less specific factors like mean annual precipitation or latitude.
+For example, body size clines (larger size at higher latitude) have evolved repeatedly on three contients in \textit{Drosophila subobscura}, strongly indicating selection.
+But which one or more of the many climatic factors that covary with latitude actually controls selection on body size?
+Selection might be even more complex for species that live over multiple seasons, meaning that environmental variation rather than the average environment may be a selective agent.
+
+% Motivate using Mimulus
+Linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions.
+Many classic and contemporary studies of local adaptation have been conducted using species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources.
 ...
+Conspicuous absence of phys mechanism
 
-Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
-Local adaptation is most often examined within species where selection in different environments leads to ecologically relevant trait divergence between populations.
-Despite local adaptation leading to population divergence, many traits are conserved within species.
-\cdm{this is not quite right.
-What I really want to say is that some traits divergence genetically, others are plastic, but plasticity is conserved because...?
-why?
-no spatial variation in selection for plasticity?}.
-Can we predict which types of traits are likely to be different between populations and which are likely to be constant?
-A long-standing hypothesis is that the grain of environmental variation determines whether organisms respond through population genetic differentiation or phenotypic plasticity \citep{}.
-\cdm{What I want to get to is whether plasticity is conserved.
-The problem is, just because there is selection for plasticity, this doesn't need to imply that the reaction norm is actually the same across populations.
-But perhaps we can oversimplify here?}.
-
-When the environment is fine grained
-
-, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
-Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
-To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
-
+% Final paragraph
 1. `intrinsic' physiological variation
@@
 
- % one paragraph general review of physiology, environment, and local adaptation
-
- % one paragraph review of physiology, environment, and local adaptation in Mimulus
-
- % one paragraph wrapping up, motivating this study
+Conclude that observations are best explained by stabilizing selection on the fundamental niche, but spatially varying divergent selection on coordinated life history and physiological traits.
 
@@
 
-I chose 16 populations from throughout the range of \textit{M.
-cardinalis} (Table 1) that have been previously studied.
-Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
+We used 16 populations from throughout the range of \textit{M. cardinalis} (Table 1). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
 
-\begin{table}[htbp]
+\begin{table}[ht]
    \centering
-   \topcaption{Table 1: 16 Focal populations}
+   \topcaption{16 Focal populations}
    \begin{tabular}{@{} lllllll @{}}
       \toprule
-      Name    & Region & Demo? & Pop gen? & Lat & Lon & Alt \\
+  Name& Region  & Demo?& Pop gen?& Latitude  & Longtiude  & Elevation (mas) \\
       \midrule
-	Hauser Creek						& South Margin		& yes	& yes	& 32.65822	& -116.53235	& 892 \\
-	Cottonwood Creek					& South Margin		& yes	& no		& 32.80122	& -116.50194	& 1206 \\
-	Sweetwater River					& South Margin		& yes	& yes	& 32.89928	& -116.5849	& 1223 \\
-	Grade Road - to Palomar Mountain SP	& South Margin		& no		& no		& 33.31392	& -116.87129	& 1500 \\
-	Whitewater Canyon					& Transverse		& yes	& yes	& 33.99329	& -116.66267	& 696 \\
-	Mill Creek							& Transverse		& yes	& no		& 34.07808	& -116.87558	& 1992 \\
-	West Fork Mojave River				& Transverse		& yes	& no		& 34.28425	& -117.37539	& 1087 \\
-	North Fork Middle Tule				& South Sierras		& yes	& yes?	& 36.20081	& -118.65092	& 1284 \\
-	Paradise Creek						& South Sierras		& yes	& yes	& 36.51776	& -118.75877	& 887 \\
-	Redwood Creek					& South Sierras		& yes	& yes	& 36.69096	& -118.90961	& 1683.72 \\
-	Wawona							& Central Sierras	& yes	& yes	& 37.539		& -119.654	& 1208 \\
-	Rainbow Pool (RP)					& Central Sierras	& yes	& no		& 37.8188		& -120.00743	& 862.2792 \\
-	Middle Yuba (Oregon Creek)			& North Sierras		& yes	& yes	& 39.39442	& -121.08302	& 425.196 \\
-	Little Jameson						& North Sierras		& yes	& yes	& 39.74298	& -120.70401	& 1592 \\
-	Deep Creek						& North Coast		& yes	& yes	& 41.66546	& -123.11341	& 694 \\
-	Rock Creek (North Fork Umpqua)		& North Margin		& yes	& yes	& 43.37375	& -122.9575	& 311.2008 \\
+	HAU & South Margin		& yes	& yes	& 	& 
+    	&    \\
+	CTC	& South Margin		& yes	& no	&  & 
+    	&    \\
+	CUR	& South Margin		& yes	& yes &  & 
+    	&    \\
+	GRP & South Margin		& no	& no	&  &
+    	&    \\
+	WWC &	Transverse		  & yes	& yes	&  & 
+    	&    \\
+	MIL	& Transverse	  	& yes	& no	&  & 
+    	&    \\
+	WFM	& Transverse	  	& yes	& no	&  & 
+    	&    \\
+	NMT	& South Sierras		& yes	& yes?&  & 
+    	&    \\
+	PRD	& South Sierras		& yes	& yes	&  & 
+    	&    \\
+	RWD	& South Sierras		& yes	& yes	&  & 
+    	&    \\
+	WNA	& Central Sierras	& yes	& yes	&  & 
+    	&    \\
+	RBW	& Central Sierras	& yes	& no	&  & 
+    	&    \\
+	MYU	& North Sierras		& yes	& yes	&  & 
+    	&    \\
+	LIJ	& North Sierras		& yes	& yes	&  & 
+    	&    \\
+	DPC	& North Coast	  	& yes	& yes	&  & 
+    	&    \\
+	RCC	& North Margin		& yes	& yes	&  & 
+    	&    \\
 	\bottomrule
@@
 
-% Spring 2014 cohort
 On 14 April, 2014, 3-5 seeds per family were sown directly on sand (Quikrete Play Sand, Georgia, USA) watered to field capacity in RLC4 Ray Leach cone-tainers placed in RL98 98-well trays (Stuewe \& Sons, Inc., Oregon, USA).
@@
 
+\paragraph*{Monitoring environment}
+
+  	Cross check with Poorter paper \\
+		Light (PAR sensors), temp/humidity (HOBOs), soil moisture
+
 \subsection*{Treatments}
 
-To do this, I will use 4 watering treatments on two populations on two soils.
-The watering treatments are: constant water, constant level/decreasing frequency, decreasing level/constant frequency, no water.
-I will monitor soil water potential in pots with no plants.
-Basically, I'm just trying to figure out whether one or another method is easier or more effective, and if I see any obvious response from plants in term of growth.
+We used 2 watering treatments on two populations on two soils. The watering treatments are: well-watered, drought.
 
@@
 We simulated typical growing season (June and July) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (High Temp) and Little Jameson (Low Temp).
-We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA (CITE).
+We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
 Daily temperatures from ClimateWNA are usually highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
@@
 The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ).
-The `VARselect' function in the vars package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
-We fit and simulated from the VAR(2) using the package dse \citep{Gilbert_2006?} in R.
+The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
+We fit and simulated from the VAR(2) using the package \pkg{dse} \citep{Gilbert_2014} in R.
 Simulated data closely resembled the autocorrelation and between-site correlation of the actual data.
@@
 
-			Are there any data from California about soil water content through the season?
+			\cdm{Are there any data from California about soil water content through the season?}
 
-\subsection*{Monitoring environment}
+\subsubsection*{Growth and photosynthesis}
 
-		Cross check with Poorter paper \\
-		Light (PAR sensors), temp/humidity (HOBOs), soil moisture
+\begin{table}[ht]
+   \centering
+   \topcaption{Traits}
+   \begin{tabular}{@{} ll @{}}
+      \toprule
+  Trait & Units \\
+      \midrule
+  Day of germination  & day \\
+  Leaf expansion rate  &  mm day$^{-1}$  \\
+  Shoot elongation rate  &  mm or cm day$^{-1}$  \\
+  Harvest dry mass &  g  \\
+  Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
+  Mortality &  Probability?  \\
+	    \bottomrule
+   \end{tabular}
+\end{table}
 
-\subsection*{Traits}
-
-\subsubsection*{Structure}
-
-leaf traits, biomass allometry
-
-\subsubsection*{Function}
-
-photosynthesis
-
-\subsubsection*{Performance}
-
-Growth rate: leaf expansion rate AND stem elongation rate
-Biomass (show correlation between growth rate and biomass)
-
-\cdm{Not sure where this goes} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package survival version 2.37-7 \citep{Therneau_2014}.
+\paragraph{Day of germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}.
 The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
@@
 
-\subsection*{Intrinsic variation}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+  \hlkwd{library}\hlstd{(survival)}
+  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"}
+  \hlstd{pathDat} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Data"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
 
-\cdm{which traits: growth rate, biomass, biomass allometry, leaf traits, photosynthesis?, mortality?}
+  \hlstd{master} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/MasterDatasheet_out.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),}
+    \hlkwc{row.names} \hlstd{=} \hlnum{1}\hlstd{)}
+
+  \hlcom{# Create Surv object}
+        \hlstd{germ} \hlkwb{<-} \hlkwd{with}\hlstd{(}\hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{MinGermDay)),} \hlkwd{Surv}\hlstd{(MinGermDay, MaxGermDay,}
+                \hlkwc{type} \hlstd{=} \hlstr{"interval2"}\hlstd{))}
+
+        \hlcom{# lognormal produced best fit to data}
+        \hlstd{fitGerm} \hlkwb{<-} \hlkwd{survreg}\hlstd{(germ} \hlopt{~} \hlstd{PopID} \hlopt{+} \hlkwd{frailty}\hlstd{(Family,} \hlkwc{sparse} \hlstd{= F),}
+                \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{MinGermDay)),} \hlkwc{dist} \hlstd{=} \hlstr{"lognormal"}\hlstd{)}
+        \hlkwd{suppressWarnings}\hlstd{(}\hlkwd{anova}\hlstd{(fitGerm))} \hlcom{# significant family and population effects}
+\end{alltt}
+\begin{verbatim}
+##                                   Df Deviance Resid. Df    -2*LL
+## NULL                              NA       NA  700.0000 2796.417
+## PopID                        15.0000 151.9495  685.0000 2644.468
+## frailty(Family, sparse = F) 106.9738 254.7233  578.0262 2389.745
+##                                 Pr(>Chi)
+## NULL                                  NA
+## PopID                       9.887489e-25
+## frailty(Family, sparse = F) 4.530519e-14
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\paragraph{Growth rate: leaf expansion and shoot elongation}
+
+We censused longest leaf length $>1$ mm on 10 days (twice per week) between 12 May and 12 June (28 to 59 days after sowing).
+We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.
+We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing).
+Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+  \hlkwd{library}\hlstd{(lme4)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: Matrix\\\#\# Loading required package: Rcpp}}\begin{alltt}
+  \hlkwd{library}\hlstd{(lmerTest)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'lmerTest'\\\#\# \\\#\# The following object is masked from 'package:lme4':\\\#\# \\\#\#\ \ \ \  lmer\\\#\# \\\#\# The following object is masked from 'package:stats':\\\#\# \\\#\#\ \ \ \  step}}\begin{alltt}
+  \hlcom{# New (blank) datasheet on which to add data}
+  \hlstd{newdata} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/MasterDatasheet_in.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+
+  \hlcom{# Read in LER data}
+  \hlstd{LERdata} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/LERdata.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+
+  \hlcom{#}
+        \hlcom{# Empirical Bayes Approach: get coefficients for every plant}
+        \hlcom{#}
+
+        \hlcom{# suppressWarnings(mmLER <- lmer(log(LLL) ~ poly(DayN, 2, raw = T) + }
+    \hlcom{# (poly(DayN, 2, raw = T)|indiv), data = LERdata))}
+  \hlcom{# save(mmLER, file = paste(pathDat, "/mmLER.RData", sep = ""))}
+  \hlkwd{load}\hlstd{(}\hlkwc{file} \hlstd{=} \hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/mmLER.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+
+  \hlcom{# Extract coefficients and add to datasheet}
+        \hlstd{Y} \hlkwb{<-} \hlkwd{t}\hlstd{(}\hlkwd{unlist}\hlstd{(}\hlkwd{fixef}\hlstd{(mmLER))} \hlopt{+} \hlkwd{t}\hlstd{(}\hlkwd{ranef}\hlstd{(mmLER)}\hlopt{$}\hlstd{indiv))[}\hlkwd{match}\hlstd{(newdata}\hlopt{$}\hlstd{indiv,}
+                \hlkwd{rownames}\hlstd{(}\hlkwd{ranef}\hlstd{(mmLER)}\hlopt{$}\hlstd{indiv)), ]}
+        \hlkwd{colnames}\hlstd{(Y)} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"b0_LER"}\hlstd{,} \hlstr{"b1_LER"}\hlstd{,} \hlstr{"b2_LER"}\hlstd{)}
+  \hlstd{newdata} \hlkwb{<-} \hlkwd{cbind}\hlstd{(newdata, Y)}
+
+  \hlcom{# Model-based LER}
+        \hlcom{# Timeframe for Cool treatment: Day 0 - 31}
+        \hlcom{# Timeframe for Hot treatment: Day 0 - 24}
+        \hlstd{newdata}\hlopt{$}\hlstd{LERstart} \hlkwb{<-} \hlkwd{exp}\hlstd{(newdata}\hlopt{$}\hlstd{b0_LER)}
+        \hlstd{newdata}\hlopt{$}\hlstd{LERend} \hlkwb{<-} \hlkwd{exp}\hlstd{(newdata}\hlopt{$}\hlstd{b0_LER} \hlopt{+}
+                                        \hlstd{newdata}\hlopt{$}\hlstd{b1_LER} \hlopt{*} \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31}\hlstd{,} \hlnum{24}\hlstd{)} \hlopt{+}
+                                        \hlstd{newdata}\hlopt{$}\hlstd{b2_LER} \hlopt{*} \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31} \hlopt{^} \hlnum{2}\hlstd{,} \hlnum{24} \hlopt{^} \hlnum{2}\hlstd{))}
+        \hlstd{newdata}\hlopt{$}\hlstd{LER_AbsGrowth} \hlkwb{<-} \hlstd{(newdata}\hlopt{$}\hlstd{LERend} \hlopt{-} \hlstd{newdata}\hlopt{$}\hlstd{LERstart)} \hlopt{/}
+                \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31}\hlstd{,} \hlnum{24}\hlstd{)}
+
+  \hlcom{# Mixed model ANOVA with family as random factor fit using step down procedure}
+        \hlstd{mm} \hlkwb{<-} \hlkwd{lmer}\hlstd{(LER_AbsGrowth} \hlopt{~} \hlstd{AvgGermDay} \hlopt{+} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt} \hlopt{+} \hlstd{(}\hlnum{1}\hlopt{|}\hlstd{Family),}
+                \hlkwc{data} \hlstd{= master)}
+        \hlstd{fitLER} \hlkwb{<-} \hlkwd{step}\hlstd{(mm,} \hlkwc{reduce.random} \hlstd{= F)} \hlcom{# includes pop, temp, h2o, population x water}
+        \hlcom{# fitLER <- step(mm, ddf = "Kenward-Roger", reduce.random = F) # same result using Satterthwaite ddf}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+  \hlcom{# Read in SER data}
+  \hlstd{SERdata} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/SERdata.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+
+  \hlcom{#}
+  \hlcom{# Empirical Bayes Approach: get coefficients for every plant}
+        \hlcom{#}
+
+        \hlcom{# suppressWarnings(mmSER <- lmer(log(height + 1) ~ poly(DayN, 2, raw = T) + }
+    \hlcom{# (poly(DayN, 2, raw = T)|indiv), data = SERdata))}
+  \hlcom{# save(mmSER, file = paste(pathDat, "/mmSER.RData", sep = ""))}
+  \hlkwd{load}\hlstd{(}\hlkwc{file} \hlstd{=} \hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/mmSER.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+
+  \hlcom{# Extract coefficients and add to datasheet}
+  \hlstd{Y} \hlkwb{<-} \hlkwd{t}\hlstd{(}\hlkwd{unlist}\hlstd{(}\hlkwd{fixef}\hlstd{(mmSER))} \hlopt{+} \hlkwd{t}\hlstd{(}\hlkwd{ranef}\hlstd{(mmSER)}\hlopt{$}\hlstd{indiv))[}\hlkwd{match}\hlstd{(newdata}\hlopt{$}\hlstd{indiv,}
+                \hlkwd{rownames}\hlstd{(}\hlkwd{ranef}\hlstd{(mmSER)}\hlopt{$}\hlstd{indiv)), ]}
+        \hlkwd{colnames}\hlstd{(Y)} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"b0_SER"}\hlstd{,} \hlstr{"b1_SER"}\hlstd{,} \hlstr{"b2_SER"}\hlstd{)}
+  \hlstd{newdata} \hlkwb{<-} \hlkwd{cbind}\hlstd{(newdata, Y)}
+
+  \hlcom{# Model-based SER}
+        \hlcom{# Timeframe for Cool treatment: Day 0 - 31}
+        \hlcom{# Timeframe for Hot treatment: Day 0 - 24}
+        \hlstd{newdata}\hlopt{$}\hlstd{SERstart} \hlkwb{<-} \hlkwd{exp}\hlstd{(newdata}\hlopt{$}\hlstd{b0_SER)}
+        \hlstd{newdata}\hlopt{$}\hlstd{SERend} \hlkwb{<-} \hlkwd{exp}\hlstd{(newdata}\hlopt{$}\hlstd{b0_SER} \hlopt{+}
+                                        \hlstd{newdata}\hlopt{$}\hlstd{b1_SER} \hlopt{*} \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31}\hlstd{,} \hlnum{24}\hlstd{)} \hlopt{+}
+                                        \hlstd{newdata}\hlopt{$}\hlstd{b2_SER} \hlopt{*} \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31} \hlopt{^} \hlnum{2}\hlstd{,} \hlnum{24} \hlopt{^} \hlnum{2}\hlstd{))}
+        \hlstd{newdata}\hlopt{$}\hlstd{SER_AbsGrowth} \hlkwb{<-} \hlstd{(newdata}\hlopt{$}\hlstd{SERend} \hlopt{-} \hlstd{newdata}\hlopt{$}\hlstd{SERstart)} \hlopt{/}
+                \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31}\hlstd{,} \hlnum{24}\hlstd{)}
+
+  \hlstd{mm} \hlkwb{<-} \hlkwd{lmer}\hlstd{(SER_AbsGrowth} \hlopt{~} \hlstd{AvgGermDay} \hlopt{+} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt} \hlopt{+}
+                \hlstd{(}\hlnum{1}\hlopt{|}\hlstd{Family),} \hlkwc{data} \hlstd{= master)}
+        \hlstd{fitSER} \hlkwb{<-} \hlkwd{step}\hlstd{(mm,} \hlkwc{reduce.random} \hlstd{= F)} \hlcom{#includes pop, temp, h2o, temp x water}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\paragraph{Photosynthesis}
+During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of 329 plants evenly spread between treatments and families within populations.
+The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
+All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity.
+During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ \cdm{get exact temps used}) so that all plants within a temperature treatment were measured under the same conditions.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+  \hlcom{# Photosynthetic rate}
+  \hlstd{mm} \hlkwb{<-} \hlkwd{lmer}\hlstd{(Photo} \hlopt{~} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt} \hlopt{+} \hlstd{(}\hlnum{1}\hlopt{|}\hlstd{Family),}
+    \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{Photo)))}
+  \hlstd{fitPhoto} \hlkwb{<-} \hlkwd{step}\hlstd{(mm,} \hlkwc{reduce.random} \hlstd{= F)}
+
+  \hlcom{# Intrinsic photosynthetic rate}
+  \hlcom{# = residuals of Conductance - Photosynthesis regression}
+  \hlcom{# Removed one outlier (may change if data are rearranged!)}
+        \hlstd{mm} \hlkwb{<-} \hlkwd{lmer}\hlstd{(resPhoto} \hlopt{~} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt} \hlopt{+} \hlstd{(}\hlnum{1}\hlopt{|}\hlstd{Family),}
+    \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{Photo))[}\hlopt{-}\hlnum{52}\hlstd{, ])}
+        \hlstd{fitIntrnPhoto} \hlkwb{<-} \hlkwd{step}\hlstd{(mm,} \hlkwc{reduce.random} \hlstd{= F)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\paragraph{Mortality}
+We assayed mortality during twice-weekly growth measurements.
+We could not get GLMM with Family effects to converge, so we used GLM with a quasibinomial error structure and assessed signifiance using Type 2 Analysis of Deviance with the R package \pkg{car}.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+  \hlkwd{library}\hlstd{(car)}
+
+  \hlstd{fitMort} \hlkwb{<-} \hlkwd{glm}\hlstd{(DiedFromStress} \hlopt{~} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt,}
+    \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{DiedFromStress)),} \hlkwc{family} \hlstd{=} \hlstr{"quasibinomial"}\hlstd{)}
+
+  \hlkwd{Anova}\hlstd{(fitMort,} \hlkwc{type} \hlstd{=} \hlnum{2}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## Analysis of Deviance Table (Type II tests)
+## 
+## Response: DiedFromStress
+##                        LR Chisq Df Pr(>Chisq)    
+## PopID                    38.169 15  0.0008519 ***
+## TempTrt                 247.763  1  < 2.2e-16 ***
+## WaterTrt                 79.508  1  < 2.2e-16 ***
+## PopID:TempTrt            26.188 15  0.0360962 *  
+## PopID:WaterTrt           11.547 15  0.7129658    
+## TempTrt:WaterTrt         48.543  1  3.231e-12 ***
+## PopID:TempTrt:WaterTrt    8.685 15  0.8934160    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\paragraph{Biomass (show correlation between growth rate and biomass?)}
+
+\subsection*{Intrinsic variation and plasticity}
+
+\cdm{which traits: germination, growth rate, biomass, biomass allometry, leaf traits, photosynthesis?, mortality?}
 We tested for Population, Treatment, and Population $\times$ Treatment interactions using mixed model ANOVA with Family included as a random factor.
-Models were fit by restricted maximum likelihood using lmer from the R package lme4 version 1.0-5 \citep{Bates_etal_2013}.
-Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package lmerTest version 2.0-11 \citep{Kuznetsova_etal_2014}.
+Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}.
+Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-11 \citep{Kuznetsova_etal_2014}.
 Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation.
@@
 
-	%
+\subsection*{Principal components of germination, growth, and photosynthesis}
 
-\subsection*{Environmental correlates}
-	% latitude
-	% temperature (mean annual, July/june)
-	% stream characterisitics - what data does Amy have from surveys?
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+  \hlstd{popMeans} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}
+    \hlkwc{germ} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{1}\hlstd{],} \hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{1}\hlstd{]} \hlopt{+} \hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{2}\hlopt{:}\hlnum{16}\hlstd{]),}
+    \hlkwc{LER} \hlstd{= fitLER}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{],}
+    \hlkwc{SER} \hlstd{= fitSER}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{],}
+    \hlkwc{photo} \hlstd{= fitIntrnPhoto}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{])}
 
-	% approaches: direct correlation, bedazzle?
+  \hlstd{traitPC} \hlkwb{<-} \hlkwd{prcomp}\hlstd{(popMeans,} \hlkwc{center} \hlstd{= T,} \hlkwc{scale.} \hlstd{= T)}
+  \hlstd{focClim}\hlopt{$}\hlstd{PC1} \hlkwb{<-} \hlstd{traitPC}\hlopt{$}\hlstd{x[,} \hlstr{"PC1"}\hlstd{]}
+
+  \hlcom{# Percent variance explained by first principal component}
+  \hlstd{pve} \hlkwb{<-} \hlkwd{paste}\hlstd{(}\hlkwd{round}\hlstd{(}\hlnum{100} \hlopt{*} \hlstd{traitPC}\hlopt{$}\hlstd{sdev[}\hlnum{1}\hlstd{]} \hlopt{^} \hlnum{2} \hlopt{/} \hlkwd{sum}\hlstd{(traitPC}\hlopt{$}\hlstd{sdev} \hlopt{^} \hlnum{2}\hlstd{),} \hlnum{1}\hlstd{),} \hlstr{"%"}\hlstd{)}
+
+  \hlcom{# Plot latitude versus PC1}
+  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_PC1vLat.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{5}\hlstd{,} \hlnum{5}\hlstd{)}
+  \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
+  \hlkwd{plot}\hlstd{(focClim}\hlopt{$}\hlstd{Lat, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{32}\hlstd{,} \hlnum{44}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{axes} \hlstd{= F,}
+    \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
+  \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Latitude"}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
+  \hlkwd{title}\hlstd{(}\hlkwc{ylab} \hlstd{=} \hlkwd{substitute}\hlstd{(}\hlkwd{plain}\hlstd{(Principal}\hlopt{~}\hlstd{Component}\hlopt{~}\hlnum{1}\hlstd{)}\hlopt{~}\hlkwd{group}\hlstd{(}\hlstr{"("}\hlstd{, x,} \hlstr{")"}\hlstd{),}
+    \hlkwd{list}\hlstd{(}\hlkwc{x} \hlstd{= pve)))}
+  \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{32}\hlstd{,} \hlnum{44}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{labels} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{expression}\hlstd{(}\hlnum{32}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{36}\hlopt{*}\hlstd{degree),}
+    \hlkwd{expression}\hlstd{(}\hlnum{40}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{44}\hlopt{*}\hlstd{degree)),} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
+  \hlkwd{axis}\hlstd{(}\hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{,} \hlnum{2}\hlstd{),} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
+  \hlstd{fitLat} \hlkwb{<-} \hlkwd{lm}\hlstd{(PC1} \hlopt{~} \hlstd{Latitude,} \hlkwc{data} \hlstd{= focClim)}
+
+  \hlcom{# polygon of confidence intervals}
+  \hlstd{x} \hlkwb{<-} \hlkwd{seq}\hlstd{(}\hlkwd{min}\hlstd{(focClim}\hlopt{$}\hlstd{Latitude),} \hlkwd{max}\hlstd{(focClim}\hlopt{$}\hlstd{Latitude),} \hlkwc{l} \hlstd{=} \hlnum{1e3}\hlstd{)}
+  \hlstd{u} \hlkwb{<-} \hlkwd{predict}\hlstd{(fitLat,} \hlkwc{new} \hlstd{=} \hlkwd{data.frame}\hlstd{(}\hlkwc{Latitude} \hlstd{= x))}
+  \hlstd{s} \hlkwb{<-} \hlkwd{sqrt}\hlstd{(}\hlkwd{sapply}\hlstd{(x,} \hlkwa{function}\hlstd{(}\hlkwc{X}\hlstd{)} \hlkwd{t}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X))} \hlopt{%*%} \hlkwd{vcov}\hlstd{(fitLat)} \hlopt{%*%} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X)))}
+  \hlstd{confInt} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwc{lowerCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.025}\hlstd{, u, s),} \hlkwc{upperCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.975}\hlstd{, u, s))}
+  \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(x,} \hlkwd{rev}\hlstd{(x)),} \hlkwd{c}\hlstd{(confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwd{rev}\hlstd{(confInt}\hlopt{$}\hlstd{upperCI)),}
+    \hlkwc{col} \hlstd{=} \hlstr{"grey80"}\hlstd{,} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{)}
+  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
+  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{upperCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
+  \hlkwd{lines}\hlstd{(}\hlkwd{range}\hlstd{(focClim}\hlopt{$}\hlstd{Lat),} \hlkwd{rev}\hlstd{(}\hlkwd{range}\hlstd{(}\hlkwd{predict}\hlstd{(fitLat))),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
+  \hlkwd{points}\hlstd{(focClim}\hlopt{$}\hlstd{Lat, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{= primShade0,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{,}
+    \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
+  \hlkwd{dev.off}\hlstd{()}
+\end{alltt}
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+\begin{alltt}
+     \hlcom{# plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)}
+  \hlcom{#plot(focClim$Lat, traitPC$x[, "PC1"])}
+  \hlcom{#cor.test(focClim$Lat, traitPC$x[, "PC1"])}
+  \hlcom{#cor.test(popMeans[, 4], traitPC$x[, "PC1"])}
+  \hlcom{#plot(focClim$bio16_cva, traitPC$x[, "PC1"])}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\subsection*{Selective agents and environmental correlates}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{library}\hlstd{(VSURF)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: randomForest\\\#\# randomForest 4.6-10\\\#\# Type rfNews() to see new features/changes/bug fixes.\\\#\# Loading required package: rpart\\\#\# Loading required package: doParallel\\\#\# Loading required package: foreach\\\#\# Loading required package: iterators\\\#\# Loading required package: parallel}}\begin{alltt}
+  \hlcom{# Load VSURF results of climatic variables associated with latitude at cardinalis occurence points}
+  \hlcom{# Note that occurences have been randomly thinned to account for variation in sampling effort}
+  \hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/clim2lat_fit_t1.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))} \hlcom{# fit_t1}
+  \hlstd{vsurfLat} \hlkwb{<-} \hlstd{fit_t1}
+
+  \hlcom{# Figure of variable importance for latitude}
+  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_clim2lat_imp.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{4}\hlstd{,} \hlnum{5}\hlstd{)}
+  \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{6}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
+  \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlkwd{ceiling}\hlstd{(vsurfLat}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x)[}\hlnum{1}\hlstd{]),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,}
+    \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)),} \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{xlab} \hlstd{=} \hlstr{"Importance"}\hlstd{,}
+    \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
+  \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlnum{1}\hlopt{:}\hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred))}
+  \hlstd{\{}
+     \hlkwd{lines}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{1}\hlstd{,} \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(i, i),} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
+  \hlstd{\}}
+  \hlkwd{points}\hlstd{(vsurfLat}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)],}
+    \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)}\hlopt{:}\hlnum{1}\hlstd{,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
+        \hlkwd{mtext}\hlstd{(climVars[vsurfLat}\hlopt{$}\hlstd{varselect.pred],} \hlkwc{side} \hlstd{=} \hlnum{2}\hlstd{,}
+    \hlkwc{at} \hlstd{=} \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)}\hlopt{:}\hlnum{1}\hlstd{,} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{1}\hlstd{)}
+\end{alltt}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in as.graphicsAnnot(text): object 'climVars' not found}}\begin{alltt}
+        \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
+        \hlkwd{dev.off}\hlstd{()}
+\end{alltt}
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+\begin{alltt}
+  \hlcom{# Columns with climatic variables to test}
+  \hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/Exploratory_climate_variables.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))} \hlcom{#climVars}
+
+  \hlcom{# Variable selection using random forests (VSURF)}
+  \hlstd{linmod.PC1} \hlkwb{<-} \hlkwd{as.formula}\hlstd{(}\hlkwd{paste}\hlstd{(}\hlstr{"PC1 ~ "}\hlstd{,} \hlkwd{paste}\hlstd{(climVars,} \hlkwc{collapse} \hlstd{=} \hlstr{" + "}\hlstd{),} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+  \hlkwd{set.seed}\hlstd{(}\hlnum{69646}\hlstd{)}
+  \hlstd{vsurfPC1} \hlkwb{<-} \hlkwd{VSURF.parallel}\hlstd{(linmod.PC1,} \hlkwc{data} \hlstd{= focClim[,} \hlkwd{c}\hlstd{(}\hlstr{"PC1"}\hlstd{, climVars)],}
+    \hlkwc{ncores} \hlstd{=} \hlnum{3}\hlstd{)}
+\end{alltt}
+
+{\ttfamily\noindent\color{warningcolor}{\#\# Warning in VSURF.parallel.formula(linmod.PC1, data = focClim[, c("{}PC1"{}, : VSURF with a formula-type call outputs selected variables\\\#\# which are indices of the input matrix based on the formula:\\\#\# you may reorder these to get indices of the original data}}\begin{alltt}
+  \hlcom{# Plot most important climatic predictor versus PC1}
+  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_bio16cva_v_Lat.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{5}\hlstd{,} \hlnum{5}\hlstd{)}
+  \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
+  \hlkwd{plot}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0.25}\hlstd{,} \hlnum{0.55}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{axes} \hlstd{= F,}
+    \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
+  \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Precipitation of Wettest Quarter"}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
+  \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"(Coefficient of variation)"}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{)}
+  \hlkwd{title}\hlstd{(}\hlkwc{ylab} \hlstd{=} \hlkwd{substitute}\hlstd{(}\hlkwd{plain}\hlstd{(Principal}\hlopt{~}\hlstd{Component}\hlopt{~}\hlnum{1}\hlstd{)}\hlopt{~}\hlkwd{group}\hlstd{(}\hlstr{"("}\hlstd{, x,} \hlstr{")"}\hlstd{),}
+    \hlkwd{list}\hlstd{(}\hlkwc{x} \hlstd{= pve)))}
+  \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
+  \hlkwd{axis}\hlstd{(}\hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{,} \hlnum{2}\hlstd{),} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
+  \hlstd{fitBio16cva} \hlkwb{<-} \hlkwd{lm}\hlstd{(PC1} \hlopt{~} \hlstd{bio16_cva,} \hlkwc{data} \hlstd{= focClim)}
+
+  \hlcom{# polygon of confidence intervals}
+  \hlstd{x} \hlkwb{<-} \hlkwd{seq}\hlstd{(}\hlkwd{min}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwd{max}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwc{l} \hlstd{=} \hlnum{1e3}\hlstd{)}
+  \hlstd{u} \hlkwb{<-} \hlkwd{predict}\hlstd{(fitBio16cva,} \hlkwc{new} \hlstd{=} \hlkwd{data.frame}\hlstd{(}\hlkwc{bio16_cva} \hlstd{= x))}
+  \hlstd{s} \hlkwb{<-} \hlkwd{sqrt}\hlstd{(}\hlkwd{sapply}\hlstd{(x,} \hlkwa{function}\hlstd{(}\hlkwc{X}\hlstd{)} \hlkwd{t}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X))} \hlopt{%*%} \hlkwd{vcov}\hlstd{(fitBio16cva)} \hlopt{%*%} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X)))}
+  \hlstd{confInt} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwc{lowerCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.025}\hlstd{, u, s),} \hlkwc{upperCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.975}\hlstd{, u, s))}
+  \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(x,} \hlkwd{rev}\hlstd{(x)),} \hlkwd{c}\hlstd{(confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwd{rev}\hlstd{(confInt}\hlopt{$}\hlstd{upperCI)),}
+    \hlkwc{col} \hlstd{=} \hlstr{"grey80"}\hlstd{,} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{)}
+  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
+  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{upperCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
+  \hlkwd{lines}\hlstd{(}\hlkwd{range}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwd{range}\hlstd{(}\hlkwd{predict}\hlstd{(fitBio16cva)),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
+  \hlkwd{points}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{= primShade0,}
+    \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
+  \hlkwd{dev.off}\hlstd{()}
+\end{alltt}
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+\begin{alltt}
+  \hlcom{# Overlap between latitude predictors and phenotype predictors}
+  \hlstd{imp2} \hlkwb{<-} \hlstd{climVars[vsurfPC1}\hlopt{$}\hlstd{varselect.pred]}
+  \hlstd{nImpPC1} \hlkwb{<-} \hlkwd{length}\hlstd{(imp2)}
+
+  \hlstd{vsurfLat}\hlopt{$}\hlstd{varselect.pred}
+\end{alltt}
+\begin{verbatim}
+##  [1] 45 43 42 20 47 41 46 39 19 13
+\end{verbatim}
+\begin{alltt}
+  \hlstd{imp1} \hlkwb{<-} \hlstd{climVars[vsurfLat}\hlopt{$}\hlstd{varselect.pred]}
+  \hlstd{nImpLat} \hlkwb{<-} \hlkwd{length}\hlstd{(imp1)}
+
+  \hlcom{# Figure of variable importance for traits}
+        \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_clim2pc1_imp.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{4}\hlstd{,} \hlnum{5}\hlstd{)}
+        \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{6}\hlstd{))}
+
+        \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{30}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, nImpLat),}
+                \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{xlab} \hlstd{=} \hlstr{"Importance"}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
+        \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlstd{(nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{)))} \hlkwd{lines}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{100}\hlstd{,} \hlnum{100}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(i, i),}
+                \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
+        \hlkwd{points}\hlstd{(}\hlnum{100} \hlopt{*} \hlstd{vsurfPC1}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlstd{nImpPC1], nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{),} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,}
+                \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
+        \hlkwd{mtext}\hlstd{(climVars[vsurfPC1}\hlopt{$}\hlstd{varselect.pred],} \hlkwc{side} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{1}\hlstd{,}
+    \hlkwc{at} \hlstd{= nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{),}
+                \hlkwc{font} \hlstd{=} \hlkwd{ifelse}\hlstd{(vsurfPC1}\hlopt{$}\hlstd{varselect.pred} \hlopt{%in%} \hlstd{vsurfLat}\hlopt{$}\hlstd{varselect.pred,}     \hlnum{2}\hlstd{,} \hlnum{1}\hlstd{))}
+        \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{3}\hlstd{,} \hlnum{0.5}\hlstd{),} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{30}\hlstd{,} \hlnum{5}\hlstd{))}
+  \hlkwd{dev.off}\hlstd{()}
+\end{alltt}
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
+Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
+Some hypotheses:
+\begin{itemize}
+  \item{H1: Latitude encompasses multivariate climatic dimension.}
+  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
+\end{itemize}
+
+Predictions:
+\begin{itemize}
+  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
+Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
+  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
+  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
+\end{itemize}
+
+Methods specific to latitude vs. trait correlation:
+We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+cardinalis}.
+We used a comprehensive set of \cdm{358 in total before thinning} recent (since 2000) known \textit{M.
+cardinalis} occurences recorded from herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
+Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
+For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
+In total, we included X environmental variables from ClimateWNA and 17?
+bioclimatic variables (Table S\#).
+For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
+Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
+
+Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
+We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation.
+We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
+Since \textit{M.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.
+
+To identify climatic variables that were closely associated with latitude in \textit{M.
+cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
+In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.
+
+Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions.
+If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.
+
+Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
+, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
+Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
+To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
+
+% Methods for spatial-average selection
+For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.
+
+\begin{table}[ht]
+   \centering
+   \topcaption{Climatic variables}
+   \begin{tabular}{@{} ll @{}}
+      \toprule
+  Climate variable & Abbreviation \\
+      \midrule
+  % climateWNA VARIABLES:
+	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
+	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
+	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
+	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
+	NFFD	& number of frost-free days                           \\
+	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
+	Eref	& Hargreaves reference evaporation (mm) \\
+	CMD	  & Hargreaves climativ moisture deficit (mm) \\
+	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
+	RH    &	mean annual relative humidity \\
+  %BIOCLIM VARIABLES:
+  bio1	&	Annual Mean Temperature	\\
+	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
+	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
+	bio4	&	Temperature Seasonality (standard deviation *100)	\\
+	bio5	&	Max Temperature of Warmest Month	\\
+	bio6	&	Min Temperature of Coldest Month	\\
+	bio7	&	Temperature Annual Range (bio5-bio6)	\\
+	bio8	&	Mean Temperature of Wettest Quarter	\\
+	bio9	&	Mean Temperature of Driest Quarter	\\
+	bio10	&	Mean Temperature of Warmest Quarter	\\
+	bio11	&	Mean Temperature of Coldest Quarter	\\
+	bio12	&	Annual Precipitation	\\
+	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
+	bio16	&	Precipitation of Wettest Quarter	\\
+	bio17	&	Precipitation of Driest Quarter	\\
+	bio18	&	Precipitation of Warmest Quarter	\\
+	bio19	&	Precipitation of Coldest Quarter	\\
+	    \bottomrule
+   \end{tabular}
+\end{table}
 
 \section*{Results}
+
+\begin{table}[ht]
+   \centering
+   \topcaption{Population, treatment effects}
+   \begin{tabular}{@{} llllllll @{}}
+      \toprule
+                                   & Germination & Leaf expansion & Stem elongation & Photosynthesis & Intrin. Photo & Mortality \\
+                                   & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
+      \midrule
+  Population                       &
+    ***  &
+    *** &
+    *** &
+    *** & 
+    *** & 
+    *** & \\
+  Temperature                      &
+    NA                                                          &
+    ***    &
+    *** &
+    *** & 
+     &     *** & \\
+  Water                            &
+    NA                                                          &
+    ***   &
+    ***   &
+    *** & 
+     &     
+    *** & \\
+  Pop $\times$ Temp                &
+    NA                                                          &
+     &
+     &
+    * & 
+     &     
+    * & \\
+  Pop $\times$ Water               &
+     NA                                                          &
+    * &
+     &
+     & 
+     &     
+     & \\
+  Temp $\times$ Water              &
+    NA                                                          &
+     &
+    *** &
+     & 
+     &     
+    *** & \\
+  Pop $\times$ Temp $\times$ Water &
+    NA                                                          &
+     &
+     &
+     & 
+     &     
+     & \\
+      \bottomrule
+   \end{tabular}
+\end{table}
+
+\subsection*{Coordinated latitudinal clines in germination, growth, and photosynthesis}
+
+\subsection*{Little evidence population by treatment interactions}
+
+\subsection*{Latitude and phenotypic clines are associated with more variable precipitation}
 
@@
 
+\section*{Discussion}
+
+% stuff about environmental grain
+How organisms adapt depends on the grain of the environment, ..., and gene flow between populations
+
+Possibilities:
+1. no trait variation: no environmental variation, high gene flow, generalist (convex)
+2. variation in mean:
+3. variation in plasticity: convex?,
+
+our data are most consistent with:
+- concave tradeoffs, fine-grained (coarse+concave would predict alt morphs), high gene flow
+
 %--------------------------------------------------
@@
 The model was censored because we could not accurately measure leaves less than 0.25 mm with digital callipers (217 of 702, 30.9\%, were too small).
-We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package MCMCglmm version 2.17 \citep{Hadfield_2010}.
+We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package \pkg{MCMCglmm} version 2.17 \citep{Hadfield_2010}.
 We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in.

### 2015-04-21 Analysis/EnvironmentalAnalysis.R
Updated Analysis folder

@@ -1,9 +1,19 @@
 library(raster)
 library(rgdal)
+library(geoR)
+
+# notes from discussion with Amy 12/17
+# two things:
+#1. Why is latitude a better predictor?
+#	a. maybe it's some multivariate thing (see if climate PCs are better than lat or single cliamte var alone)
+#	b. maybe need different time averages
+
+#2. Average selective environment
+#	
 
 ### Occurences
 	# Import
-	occCard <- read.csv("~/Google Drive/cardinalisENM/data files/herb.dat.lnpreds.csv")
+	occCard <- read.csv("~/Google Drive/cardinalisENM/data files/all.records.aug.31.csv")
 	
 	# Convert to SpatialPoints and exclude absences
 	occCard <- SpatialPoints(coords = occCard[occCard$PRESABS == 1, c("Longitude", 
@@ -11,24 +21,83 @@ library(rgdal)
 	
 	# Reproject to match raster layers
 	occCard <- spTransform(occCard, CRSobj = CRS("+proj=lcc +lat_1=49 +lat_2=77 +lat_0=0 +lon_0=-95 +x_0=0 +y_0=0 +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"))
-	
+
 ### 8 environmental layers used in modeling
-	
-	# This doesn't work because file path is wrong
-	# load("/Users/cdmuir/Google Drive/cardinalisENM/R objects/wc0.5/pred.dom.Rda")
-	
-	# This recreates the same thing with correct local path
-	pred.dom <- stack(paste("~/Google Drive/cardinalisENM/R objects/wc0.5/", c("bio2.grd", 
-		"bio3.grd", "bio4.grd", "bio10.grd", "bio11.grd", "bio12.grd", "bio14.grd", 
-		"bio15.grd"), sep = ""))
+	### GET 8 LAYERS FOR FOCAL POPS!
+	pred <- stack(paste("~/Google Drive/CardLocalAdaptation/Data/Climate/Raw bios (untransformed)/", c("bio2.grd", "bio3.grd", "bio4.grd", "bio10.grd", "bio11.grd", "bio12.grd", "bio14.grd", "bio15.grd"), sep = ""))
 
[…]

### 2015-04-21 Analysis/MasterAnalysis.R

@@ -1,10 +1,15 @@
 #
 #	Master Analysis of M. cardinalis Growth Chamber Experiment (Spring 2014)
 #
-#	Title:	 Local Physiological adaptaiton in Mimulus cardinalist
+#	Title:	 Local Physiological adaptaiton in Mimulus cardinalis
 #	Authors: Christopher D. Muir and Amy L. Angert
 #	Journal: Evolution?
 #
+
+
+## PART 1:	TRAIT VARIATION
+## PART 2: 	SELECTIVE AGENT
+## PART 3	???
 
 ### Preliminaries
 
@@ -19,6 +24,7 @@
 	library(pbkrtest)
 	library(lmerTest)
 	library(survival)
+	library(VSURF)
 	
 #
 ##	Set working directory
@@ -33,11 +39,11 @@
 	source("Analysis/functions.R")
 	
 #
-##	Population Environmental data
+##	Focal population climate data
 #
-	# This needs to be updated with PRISM, other data, etc. Lat, lon should be correct though
+	# This data.frame will change as climate data pipeline is updated
 	# Populations are ordered alphabetically by site to match output from stats
-	popenv <- read.csv("Data/PopulationEnvData.csv")
+	load('~/Google Drive/CardLocalAdaptation/Data/Climate/Robjects/focClim.RData')
 	
 #
 ##	Read in processed datasets
@@ -54,32 +60,35 @@
 	germ <- with(subset(data, !is.na(data$MinGermDay)), Surv(MinGermDay, MaxGermDay,	
 		type = "interval2"))
 
-	# lognormal produced best fit to data
+	# lognormal produced best fit to data (ID = Population effect)
 	fit1 <- survreg(germ ~ Population + frailty(Family, sparse = F), 
 		data = subset(data, !is.na(data$MinGermDay)), dist = "lognormal")
 	anova(fit1) # significant family and population effects
 	
+	# Add population germination effects to climate data.frame for subsequent analysis
+	focClim$DoG <- c(coef(fit1)[1], coef(fit1)[1] + coef(fit1)[2:16])
+	
 	# Figure of germination time by latitude
-	png("ms/Figures/Figure_DoG_Lat.png", 5, 5, units = "in", res = 1e3)
[…]

### 2015-04-21 Analysis/prelim environmental analysis.R

@@ -1,0 +1,41 @@
+setwd("~/Google Drive/CardLocalAdaptation/Data/Climate/ClimateWNA")
+library(raster)
+# environmental
+# growing degree days, summer precip, drought indicies
+# variation within and between sites
+
+#make big stack
+files <- list.files("NORM_6190_Bioclim_ASCII", pattern = ".asc")
+bioclim <- stack(paste("NORM_6190_Bioclim_ASCII/", files, sep = "")[1])
+
+latlon <- SpatialPoints(popenv[, c("Lon", "Lat")], proj4string = CRS("+proj=longlat +datum=WGS84"))
+
+
+popenv$ggd <- 
+tmp <- raster::extract(bioclim, latlon[1, ], buffer = 1e6, cellnumbers = T) # buffer is in meters
+
+r <- bioclim[[1]]
+r[] <- NA
+r[tmp[[1]][, 1]] <- 1
+plot(bioclim)
+
+with(popenv, plot(Lat, ggd))
+	plot(popenv$ggd, betas$lsmeans.table$Estimate[x], log = "")
+	
+cor(cbind(popenv[, 75], betas$lsmeans.table$Estimate[x]))
+
+# get all card points to get average selective environment along with latitude.
+
+
+# PRISM file name format:
+PRISM_ppt_stable_4kmD1_YYYYMMDD_bil
+PRISM_tmax_stable_4kmD1_YYYYMMDD_bil
+PRISM_tmin_stable_4kmD1_YYYYMMDD_bil
+PRISM_tmean_stable_4kmD1_YYYYMMDD_bil
+
+setwd("~/Google Drive/CardLocalAdaptation/Data/Climate/PRISM_daily")
+files <- list.files(pattern = ".zip")[1:2]
+for (i in files)
+{
+	system(paste("unzip", i))
+}

### 2015-04-21 Data/ProcessData.R
testing

@@ -21,6 +21,10 @@ library(MASS)
 	data$TempTrt <- factor(as.character(data$TempTrt))
 	data$WaterTrt <- factor(as.character(data$WaterTrt))
 
+  # Write MasterDatasheet_in.csv for ms directory with just identifying info (no trait data)
+  write.csv(data[, c("indiv", "Block", "Col", "Row", "Population", "Family", "TempTrt", 
+  	"WaterTrt")], file = "../ms/Data/MasterDatasheet_in.csv", row.names = F)
+
 	# Remove all < 1 mm, call NA
 	data$LLL_0512 <- as.numeric(gsub("< 1", "NA", data$LLL_0512))
 	data$LLL_0515 <- as.numeric(gsub("< 1", "NA", data$LLL_0515))
@@ -33,51 +37,54 @@ library(MASS)
 #
 
 	# Melt data.frame
-	lll.data <- melt(data, value.name = "LLL")
-	lll.data$Day <- mdy(paste(substr(lll.data$variable, 5, 8), "2014", sep = ""))
-	lll.data$DayN <- (as.numeric(lll.data$Day) - 1399852800) / 86400
+	LERdata <- melt(data, value.name = "LLL")
+	LERdata$Day <- mdy(paste(substr(LERdata$variable, 5, 8), "2014", sep = ""))
+	LERdata$DayN <- (as.numeric(LERdata$Day) - 1399852800) / 86400
 	
 	# Get rid of NAs
-	lll.data <- lll.data[!is.na(lll.data$LLL), ]
+	LERdata <- LERdata[!is.na(LERdata$LLL), ]
 		
 	# Refactor indiv
-	lll.data$indiv <- as.factor(as.character(lll.data$indiv))
+	LERdata$indiv <- as.factor(as.character(LERdata$indiv))
+
+  # Export version for paper
+  write.csv(LERdata, file = "~/Google Drive/CardLocalAdaptation/ms/Data/LERdata.csv")
 
 	#
 	# Empirical Bayes Approach: get coefficients for every plant
 	#
 
-	mm <- lmer(log(LLL) ~ poly(DayN, 2, raw = T) + (poly(DayN, 2, raw = T)|indiv),
-		data = lll.data)
+	mmLER <- lmer(log(LLL) ~ poly(DayN, 2, raw = T) + (poly(DayN, 2, raw = T)|indiv),
+		data = LERdata)
[…]

### 2015-04-21 Analysis/EnvironmentalAnalysis.R
Removed Data and Analysis folder. Added GitHub instructions in ms.

@@ -1,163 +1,0 @@
-library(raster)
-library(rgdal)
-library(geoR)
-
-# notes from discussion with Amy 12/17
-# two things:
-#1. Why is latitude a better predictor?
-#	a. maybe it's some multivariate thing (see if climate PCs are better than lat or single cliamte var alone)
-#	b. maybe need different time averages
-
-#2. Average selective environment
-#	
-
-### Occurences
-	# Import
-	occCard <- read.csv("~/Google Drive/cardinalisENM/data files/all.records.aug.31.csv")
-	
-	# Convert to SpatialPoints and exclude absences
-	occCard <- SpatialPoints(coords = occCard[occCard$PRESABS == 1, c("Longitude", 
-		"Latitude")], proj4string = CRS("+proj=longlat +ellps=WGS84"))
-	
-	# Reproject to match raster layers
-	occCard <- spTransform(occCard, CRSobj = CRS("+proj=lcc +lat_1=49 +lat_2=77 +lat_0=0 +lon_0=-95 +x_0=0 +y_0=0 +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"))
-
-### 8 environmental layers used in modeling
-	### GET 8 LAYERS FOR FOCAL POPS!
-	pred <- stack(paste("~/Google Drive/CardLocalAdaptation/Data/Climate/Raw bios (untransformed)/", c("bio2.grd", "bio3.grd", "bio4.grd", "bio10.grd", "bio11.grd", "bio12.grd", "bio14.grd", "bio15.grd"), sep = ""))
-
-	logPred <- stack(paste("~/Google Drive/CardLocalAdaptation/Data/Climate/Predictor set (log transformed)/", c("bio2.grd", "bio3.grd", "bio4.grd", "bio10.grd", "bio11.grd", "bio12.grd", "bio14.grd", "bio15.grd"), sep = ""))
-
-### Get 8 layers for focal sites
-	popenv <- read.csv("~/Google Drive/CardLocalAdaptation/Data/PopulationEnvData.csv")
-
-	# Convert to SpatialPoints
-	tmp <- SpatialPoints(coords = popenv[, c("Lon", "Lat")], proj4string = CRS("+proj=longlat +ellps=WGS84"))
-
-	# Reproject to match raster layers
[…]

### 2015-04-21 Analysis/MasterAnalysis.R

@@ -1,534 +1,0 @@
-#
-#	Master Analysis of M. cardinalis Growth Chamber Experiment (Spring 2014)
-#
-#	Title:	 Local Physiological adaptaiton in Mimulus cardinalis
-#	Authors: Christopher D. Muir and Amy L. Angert
-#	Journal: Evolution?
-#
-
-
-## PART 1:	TRAIT VARIATION
-## PART 2: 	SELECTIVE AGENT
-## PART 3	???
-
-### Preliminaries
-
-#
-##	Libraries
-#
-
-	library(lme4)
-	library(MCMCglmm)
-	library(MASS)
-	library(car)
-	library(pbkrtest)
-	library(lmerTest)
-	library(survival)
-	library(VSURF)
-	
-#
-##	Set working directory
-#
-
-	setwd("~/Google Drive/CardLocalAdaptation")
-
-#
-##	Source custom functions
-#
-
-	source("Analysis/functions.R")
-	
-#
-##	Focal population climate data
-#
-	# This data.frame will change as climate data pipeline is updated
-	# Populations are ordered alphabetically by site to match output from stats
-	load('~/Google Drive/CardLocalAdaptation/Data/Climate/Robjects/focClim.RData')
-	
-#
-##	Read in processed datasets
-#
-
-	data <- read.csv("Data/MasterDatasheet_out.csv")
-
-#
-##	Difference in germiantion timing (yes) and intitial size (no?)?
-#
-
-	##### Germination
-	# Create Surv object
-	germ <- with(subset(data, !is.na(data$MinGermDay)), Surv(MinGermDay, MaxGermDay,	
-		type = "interval2"))
-
-	# lognormal produced best fit to data (ID = Population effect)
-	fit1 <- survreg(germ ~ Population + frailty(Family, sparse = F), 
-		data = subset(data, !is.na(data$MinGermDay)), dist = "lognormal")
-	anova(fit1) # significant family and population effects
-	
-	# Add population germination effects to climate data.frame for subsequent analysis
-	focClim$DoG <- c(coef(fit1)[1], coef(fit1)[1] + coef(fit1)[2:16])
-	
-	# Figure of germination time by latitude
-	pdf("ms/Figures/Figure_DoG_Lat.pdf", 5, 5)
[…]

### 2015-04-21 Analysis/PopulationMap.R

@@ -1,208 +1,0 @@
-#
-# Niche modeling in Mimulus cardinalis
-#
-
-setwd("~/Dropbox/CardNiche")
-
-# Records from GBIF (these already include UCJEPS, so maybe just use those?)
-
-card <- gbif("Mimulus", "cardinalis")
-card <- card[-which(card$lon > -80), ]
-nrow(card)
-
-# Jespon herbarium records
-# Accessed July 3, 2013
-
-card <- read.csv("~/Documents/Postdoc/CardNiche/card_jepson_records_minimal.csv", colClasses = c("factor", "integer", "integer", "factor", "character", "numeric", "numeric"))
-with(card, plot(longitude, latitude))
-
-# convert elevation strings to meters
-
-x <- which(regexpr("[0-9]+ [-] [0-9]+ m", card$elevation) != -1)
-
-le <- as.numeric(unlist(strsplit(card$elevation[x], " - "))[seq(1, 2 * length(x), 2)])
-ue <- as.numeric(unlist(strsplit(unlist(strsplit(card$elevation[x], " - "))[seq(2, 2 * length(x), 2)], " m")))
-card$elevation[x] <- (le + ue) / 2
-
-x <- which(regexpr("[0-9]+ [-] [0-9]+ ft", card$elevation) != -1)
-le <- as.numeric(unlist(strsplit(card$elevation[x], " - "))[seq(1, 2 * length(x), 2)])
-ue <- as.numeric(unlist(strsplit(unlist(strsplit(card$elevation[x], " - "))[seq(2, 2 * length(x), 2)], " ft")))
-card$elevation[x] <- (le + ue) / 2
-
-x <- regexpr(" ft", card$elevation)
-
-card$elevation[x != -1] <- as.numeric(substr(card$elevation[x != -1], 1, x[x != -1] - 1)) * 0.3048
-
-x <- regexpr(" m", card$elevation)
-
-card$elevation[x != -1] <- as.numeric(substr(card$elevation[x != -1], 1, x[x != -1] - 1))
-
-card$elevation <- as.numeric(card$elevation)
-
-#
-## Anget lab seed collections
-#
-
-library(raster)
-library(dismo)
-
-card2 <- read.csv("~/Dropbox/CardNiche/CardPopulations.csv")
-
-# get rid of distant pop
-card2 <- card2[card2$Lon < -115 & !is.na(card2$Lon), ]
[…]

### 2015-04-21 Analysis/functions.R

@@ -1,28 +1,0 @@
-exportTable <- function(file, data, colnames, rownames, h1 = "", ...)
-{
-	# Check that there are the right number of columns and rows
-	if(length(colnames) != ncol(data)) stop("number of columns in data does not match length of column names")
-
-	if(length(rownames) != nrow(data)) stop("number of rows in data does not match length of row names")
-
-	table <- file(file, open = "w")
-	cat(h1, "&", colnames[1], file = table)
-	if (length(colnames) > 1) 
-	{
-		for (i in 2:length(colnames)) cat(" &", colnames[i], file = table)
-	}
-	cat(" \\\\ \\midrule\n", file = table)
-	
-	for (i in 1:length(rownames))
-	{
-		cat(rownames[i], "&", data[i, 1], file = table)
-		if (length(colnames) > 1) 
-		{
-			for (j in 2:length(colnames)) cat(" &", data[i, j], file = table)
-		}
-		cat(" \\\\\n", file = table)
-	}
-	
-	close(con = table)	
-	
-}

### 2015-04-21 Analysis/prelim environmental analysis.R

@@ -1,41 +1,0 @@
-setwd("~/Google Drive/CardLocalAdaptation/Data/Climate/ClimateWNA")
-library(raster)
-# environmental
-# growing degree days, summer precip, drought indicies
-# variation within and between sites
-
-#make big stack
-files <- list.files("NORM_6190_Bioclim_ASCII", pattern = ".asc")
-bioclim <- stack(paste("NORM_6190_Bioclim_ASCII/", files, sep = "")[1])
-
-latlon <- SpatialPoints(popenv[, c("Lon", "Lat")], proj4string = CRS("+proj=longlat +datum=WGS84"))
-
-
-popenv$ggd <- 
-tmp <- raster::extract(bioclim, latlon[1, ], buffer = 1e6, cellnumbers = T) # buffer is in meters
-
-r <- bioclim[[1]]
-r[] <- NA
-r[tmp[[1]][, 1]] <- 1
-plot(bioclim)
-
-with(popenv, plot(Lat, ggd))
-	plot(popenv$ggd, betas$lsmeans.table$Estimate[x], log = "")
-	
-cor(cbind(popenv[, 75], betas$lsmeans.table$Estimate[x]))
-
-# get all card points to get average selective environment along with latitude.
-
-
-# PRISM file name format:
-PRISM_ppt_stable_4kmD1_YYYYMMDD_bil
-PRISM_tmax_stable_4kmD1_YYYYMMDD_bil
-PRISM_tmin_stable_4kmD1_YYYYMMDD_bil
-PRISM_tmean_stable_4kmD1_YYYYMMDD_bil
-
-setwd("~/Google Drive/CardLocalAdaptation/Data/Climate/PRISM_daily")
-files <- list.files(pattern = ".zip")[1:2]
-for (i in files)
-{
-	system(paste("unzip", i))
-}

### 2015-04-21 Data/ProcessData.R

@@ -1,532 +1,0 @@
-setwd("~/Google Drive/CardLocalAdaptation/Data")
-
-library(lme4)
-library(reshape2)
-library(lubridate)
-library(MASS)
-	
-#
-#	Preparing datasheet for LLL analysis
-#
-
-	data <- read.csv("MasterDatasheet_in.csv", nrows = 768)
-	cols2keep <- c("indiv", "Block", "Col", "Row", "Population", "Family", "TempTrt",
-		"WaterTrt", "LLL_0512", "LLL_0515", "LLL_0520", "LLL_0523", "LLL_0526", 
-		"LLL_0529", "LLL_0602", "LLL_0605", "LLL_0609", "LLL_0612")
-	data$Row <- as.character(data$Row)
-	data$Population[data$Population == "Sweetwater River"] <- "Cuyamaca Rancho"
-	data$Population <- factor(as.character(data$Population))
-	data$indiv <- gl(768, 1)
-	data <- data[data$UseForLLLGrowth, cols2keep]
-	data$TempTrt <- factor(as.character(data$TempTrt))
-	data$WaterTrt <- factor(as.character(data$WaterTrt))
-
-  # Write MasterDatasheet_in.csv for ms directory with just identifying info (no trait data)
-  write.csv(data[, c("indiv", "Block", "Col", "Row", "Population", "Family", "TempTrt", 
-  	"WaterTrt")], file = "../ms/Data/MasterDatasheet_in.csv", row.names = F)
-
-	# Remove all < 1 mm, call NA
-	data$LLL_0512 <- as.numeric(gsub("< 1", "NA", data$LLL_0512))
-	data$LLL_0515 <- as.numeric(gsub("< 1", "NA", data$LLL_0515))
-	data$LLL_0520 <- as.numeric(gsub("< 1", "NA", data$LLL_0520))
-	data$LLL_0523 <- as.numeric(gsub("< 1", "NA", data$LLL_0523))
-	data$LLL_0526 <- as.numeric(gsub("< 1", "NA", data$LLL_0526))
-
-#
-#	LLL growth rates
-#
-
-	# Melt data.frame
-	LERdata <- melt(data, value.name = "LLL")
-	LERdata$Day <- mdy(paste(substr(LERdata$variable, 5, 8), "2014", sep = ""))
-	LERdata$DayN <- (as.numeric(LERdata$Day) - 1399852800) / 86400
[…]

### 2015-04-22 ms/ms.Rnw
Worked on manuscript Results

@@
 
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ExptlDes.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Experimental Design]{CAPTION}
+	\label{fig:Fig_ExptlDes}
+\end{figure}
+
 \subsection*{Measurements: Growth and photosynthesis}
@@
   # Plot latitude versus PC1
-  pdf(paste(pathFig, "/Figure_PC1vLat.pdf", sep = ""), 5, 5)
-  par(mar = c(5, 5, 1, 1))
+  pdf(paste(pathFig, "/Figure_PC1vLat.pdf", sep = ""), 5.5, 5)
+  par(mai = c(1, 1.5, 0.25, 0.25))
   plot(focClim$Lat, focClim$PC1, xlim = c(32, 44), ylim = c(-4, 4), axes = F,
@@
     list(x = pve)))
-  axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), expression(36*degree),
-    expression(40*degree), expression(44*degree)), lwd = 0, lwd.ticks = 1) 
+  axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), 
+		expression(36*degree), expression(40*degree), expression(44*degree)), 
+		lwd = 0, lwd.ticks = 1) 
   axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
   fitLat <- lm(PC1 ~ Latitude, data = focClim)
+
+	# Label y-axis from slow to fast growth
+	mtext(c("Slow\nGrowth", "Fast\nGrowth"), 2, at = c(-4, 4), cex = 1.25, 
+				font = 3, las = 1, line = 4, adj = 0.5)
 
@@
 
+As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
+temperature or precipitation) that covary with latitude.
+
 <<VSURF>>=
@@
   vsurfLat <- fit_t1
+
+	# Load climVars
+	load(paste(pathDat, "/ExploratoryCimateVariables.RData", sep = ""))
 
@@
 
-  # Columns with climatic variables to test
-  load(paste(pathDat, "/Exploratory_climate_variables.RData", sep = "")) #climVars
-
   # Variable selection using random forests (VSURF)
   linmod.PC1 <- as.formula(paste("PC1 ~ ", paste(climVars, collapse = " + "), sep = ""))
-  set.seed(69646)
+  set.seed(69647)
   vsurfPC1 <- VSURF.parallel(linmod.PC1, data = focClim[, c("PC1", climVars)], 
@@
 	RH    &	mean annual relative humidity \\
-  %BIOCLIM VARIABLES:
+  % BIOCLIM VARIABLES:
   bio1	&	Annual Mean Temperature	\\
@@
 
+%--------------------------------------------------
+% Results
+%--------------------------------------------------
+
 \section*{Results}
+
+\subsection*{Coordinated latitudinal clines in germination, growth, and photosynthesis}
+
+Genetic differences between populations affecting originsmal function and performance point to traits that may be involved in local adaptation.
+Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M.
+cardinalis} (Table~\ref{table:Table_AnovaSummary}).
+A single principal component captured \Sexpr{pve}$\%$ of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
+As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of Population, Treatment, Population x Treatment interactions for varies traits
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
     \Sexpr{sigStar(fitPhoto$anova.table["TempTrt", "Pr(>F)"])} & 
-    \Sexpr{sigStar(fitIntrnPhoto$anova.table["TempTrt", "Pr(>F)"])} &     \Sexpr{sigStar(Anova(fitMort, type = 2)["TempTrt", "Pr(>Chisq)"])} & \\
+    \Sexpr{sigStar(fitIntrnPhoto$anova.table["TempTrt", "Pr(>F)"])} &
+    \Sexpr{sigStar(Anova(fitMort, type = 2)["TempTrt", "Pr(>Chisq)"])} & \\
   Water                            &
@@
    \end{tabular}
+   \label{table:Table_AnovaSummary}
 \end{table}
 
-\subsection*{Coordinated latitudinal clines in germination, growth, and photosynthesis}
-
-\subsection*{Little evidence population by treatment interactions}
-
-\subsection*{Latitude and phenotypic clines are associated with more variable precipitation}
-
-% ANOVA table for LLL growth rate (need to label tables)
-\begin{table}[htbp]
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of latitude versus PC1 (growth axis)
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{CAPTION}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-	\input{Tables/Table_lllGrowthAnova.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-% ANOVA table for height growth rate (need to label tables)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, stem expansion rate]{CAPTION}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-	\input{Tables/Table_heightGrowthAnova.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-\subsection*{Latitude versus growth rate}
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations germinate sooner.]{CAPTION}
-	\label{fig:Fig_DoG}
+	\caption[Southern populations grow faster]{CAPTION}
+	\label{fig:Fig_PC1vLat}
 \end{figure}
 
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
-	\label{fig:Fig_LLL}
-\end{figure}
-
-\begin{figure}
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (stem elongation rate).]{CAPTION}
-	\label{fig:Fig_height}
-\end{figure}
+\subsection*{Remarkably little evidence for variation in plasticity}
+
+Genotype $\times$ environment interactions are also a common signature of local adaptation [CITE].
+For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
+We found remarkably little evidence for this pattern in \textit{M.
+cardinalis}.
+There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
+Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
+Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
+However, the fact that the Population and Treatment effects were highly significant ($P \lle 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
+
+\subsection*{Latitude and phenotypic clines are associated with more variable precipitation}
+
+Latitude explains variation
+Presumably selective agent is not latitude, but something related to it
+Exploratory approach to identify putative selective agent to test in future studies
+We used a machine learning approach...identified interannual variation in precipitation
+Obviously, cannot identify alternative variables that were not included in analysis.
+
+%--------------------------------------------------
+% Discussion
+%--------------------------------------------------
 
 \section*{Discussion}
+
+1.
+How does local adaptation occur?
+It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
+For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
+
+2. Why does local adaptation succeed? shallow environmental gradients lead to convex tradeoffs not heavily constrained by gene flow.
+
+3. Selective agent - unpredictability. Explain
+
+4.
+Why does local adaptation fail?
+concave, high gene flow, demography.
+Qualify that our metrics of plasticity are coarse, but at least suggest that GxE is much weaker than G for the environmental gradients evaluated here.
+
+5.
+Future work and broad conclusions.
+Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
 
@@
 
+% ANOVA table for LER (need to label tables)
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, leaf expansion rate]{CAPTION}
+	\begin{center}
+	\begin{tabular}{lcccccc}
+	\toprule
+
+	\input{Tables/Table_lllGrowthAnova.txt}
+
+	\bottomrule
+	\end{tabular}
+	\end{center}
+\end{table}
+
+% ANOVA table for SER (need to label tables)
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, stem expansion rate]{CAPTION}
+	\begin{center}
+	\begin{tabular}{lcccccc}
+	\toprule
+
+	\input{Tables/Table_heightGrowthAnova.txt}
+
+	\bottomrule
+	\end{tabular}
+	\end{center}
+\end{table}
+
+% Figures showing latitudinal clines with specific traits
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Southern populations germinate sooner.]{CAPTION}
+	\label{fig:Fig_DoG}
+\end{figure}
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
+	\label{fig:Fig_LLL}
+\end{figure}
+
+\begin{figure}
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Southern populations grow faster (stem elongation rate).]{CAPTION}
+	\label{fig:Fig_height}
+\end{figure}
+
 \end{document}

### 2015-04-22 ms/ms.tex
Worked on manuscript Results

@@
 
+      April 14 germinate
+      Before May 20 (36 days): 7.5 min (complete flood) every 2 hours
+      between May 20 to June 10 (57 days), down to 30 seconds, 1x per day
 			Wet: daily/constant irrigation with cooled water
@@
 
-\subsubsection*{Growth and photosynthesis}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{library}\hlstd{(scales)} \hlcom{# for color transparency}
+\hlstd{exptlDes} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/ExperimentalDesign.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+
+\hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/ExperimentalDesignBase.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{7}\hlstd{,} \hlnum{5}\hlstd{)}
+\hlkwd{par}\hlstd{(}\hlkwc{mai} \hlstd{=} \hlkwd{rep}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{4}\hlstd{))}
+\hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlkwd{max}\hlstd{(exptlDes}\hlopt{$}\hlstd{DayN)),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{1}\hlstd{,} \hlnum{41}\hlstd{),}
+                 \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{,} \hlkwc{xlab} \hlstd{=} \hlstr{"Key treatments and measurements"}\hlstd{,}
+     \hlkwc{ylab} \hlstd{=} \hlkwd{expression}\hlstd{(}\hlkwd{paste}\hlstd{(}\hlstr{"Temperature ("}\hlstd{,} \hlkwd{phantom}\hlstd{()}\hlopt{*}\hlstd{degree,} \hlstr{"C)"}\hlstd{)))}
+
+\hlcom{# Temperature axis}
+\hlkwd{axis}\hlstd{(}\hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{10}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,}
+        \hlkwc{labels} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{expression}\hlstd{(}\hlnum{0}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{10}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{20}\hlopt{*}\hlstd{degree),}
+                                                 \hlkwd{expression}\hlstd{(}\hlnum{30}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{40}\hlopt{*}\hlstd{degree)))}
+\hlcom{# Axis of time on top}
+\hlkwd{mtext}\hlstd{(}\hlstr{"Days after sowing"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{3.5}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
+\hlkwd{mtext}\hlstd{(}\hlstr{"(Date for climate simulation)"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{2.25}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
+\hlkwd{axis}\hlstd{(}\hlnum{3}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{90}\hlstd{,} \hlnum{30}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
+\hlkwd{axis}\hlstd{(}\hlnum{3}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{17}\hlstd{,} \hlnum{47}\hlstd{,} \hlnum{78}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlopt{-}\hlnum{0.5}\hlstd{,}
+     \hlkwc{labels} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"(June)"}\hlstd{,} \hlstr{"(July)"}\hlstd{,} \hlstr{"(August)"}\hlstd{))}
+
+\hlcom{# Indicate sowing, beginning of temperature and drought treatments}
+\hlkwd{abline}\hlstd{(}\hlkwc{v} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
+\hlkwd{mtext}\hlstd{(}\hlstr{"Sow"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{0}\hlstd{)}
+\hlkwd{abline}\hlstd{(}\hlkwc{v} \hlstd{=} \hlnum{17}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
+\hlkwd{mtext}\hlstd{(}\hlstr{"Temp\textbackslash{}nTreatment"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlnum{17}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{1}\hlstd{)}
+\hlkwd{abline}\hlstd{(}\hlkwc{v} \hlstd{=} \hlnum{37}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
+\hlkwd{mtext}\hlstd{(}\hlstr{"Drought\textbackslash{}nTreatment"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlnum{37}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{1}\hlstd{)}
+
+\hlcom{# Polygon of Hot treatment, min and max}
+\hlkwd{with}\hlstd{(exptlDes[}\hlnum{18}\hlopt{:}\hlnum{94}\hlstd{, ],} \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{17}\hlopt{:}\hlnum{93}\hlstd{,} \hlnum{93}\hlopt{:}\hlnum{17}\hlstd{),} \hlkwd{c}\hlstd{(Tmin_hot,} \hlkwd{rev}\hlstd{(Tmax_hot)),}
+  \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(primShade0,} \hlnum{0.25}\hlstd{)))}
+
+\hlcom{# Polygon of Cool treatment, min and max}
+\hlkwd{with}\hlstd{(exptlDes[}\hlnum{18}\hlopt{:}\hlnum{94}\hlstd{, ],} \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{17}\hlopt{:}\hlnum{93}\hlstd{,} \hlnum{93}\hlopt{:}\hlnum{17}\hlstd{),} \hlkwd{c}\hlstd{(Tmin_cool,} \hlkwd{rev}\hlstd{(Tmax_cool)),}
+  \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(sec1shade0,} \hlnum{0.25}\hlstd{)))}
+
+\hlcom{# Lines with average temperature}
+\hlkwd{with}\hlstd{(exptlDes[}\hlnum{18}\hlopt{:}\hlnum{94}\hlstd{, ],} \hlkwd{points}\hlstd{(}\hlnum{17}\hlopt{:}\hlnum{93}\hlstd{, Tavg_hot,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"l"}\hlstd{,}
+        \hlkwc{col} \hlstd{= primShade3))}
+\hlkwd{with}\hlstd{(exptlDes[}\hlnum{18}\hlopt{:}\hlnum{94}\hlstd{, ],} \hlkwd{points}\hlstd{(}\hlnum{17}\hlopt{:}\hlnum{93}\hlstd{, Tavg_cool,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"l"}\hlstd{,}
+        \hlkwc{col} \hlstd{= sec1shade3))}
+
+\hlcom{# Rug indicating measurment timing}
+\hlkwd{rect}\hlstd{(}\hlnum{6}\hlstd{,} \hlopt{-}\hlnum{2}\hlstd{,} \hlnum{17}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlnum{11.5}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Germ"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,}
+  \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
+
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{rect}\hlstd{(}\hlkwd{min}\hlstd{(}\hlkwd{which}\hlstd{(LERmeas)),} \hlopt{-}\hlnum{2}\hlstd{,} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(LERmeas)),} \hlnum{0}\hlstd{,}
+        \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{))}
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlkwd{min}\hlstd{((}\hlkwd{which}\hlstd{(LERmeas))} \hlopt{+} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(LERmeas)))} \hlopt{/} \hlnum{2}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{,}
+        \hlkwc{labels} \hlstd{=} \hlstr{"LER"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
+
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{rect}\hlstd{(}\hlkwd{min}\hlstd{(}\hlkwd{which}\hlstd{(SERmeas)),} \hlnum{0}\hlstd{,} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(SERmeas)),} \hlnum{2}\hlstd{,}
+        \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{))}
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlkwd{min}\hlstd{((}\hlkwd{which}\hlstd{(SERmeas))} \hlopt{+} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(SERmeas)))} \hlopt{/} \hlnum{2}\hlstd{,} \hlnum{1}\hlstd{,}
+        \hlkwc{labels} \hlstd{=} \hlstr{"SER"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
+
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{rect}\hlstd{(}\hlkwd{min}\hlstd{(}\hlkwd{which}\hlstd{(Harvest)),} \hlopt{-}\hlnum{2}\hlstd{,} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(Harvest)),} \hlnum{0}\hlstd{,}
+        \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{))}
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlkwd{min}\hlstd{((}\hlkwd{which}\hlstd{(Harvest))} \hlopt{+} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(Harvest)))} \hlopt{/} \hlnum{2}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{,}
+        \hlkwc{labels} \hlstd{=} \hlstr{"Harvest"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
+
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlkwd{min}\hlstd{((}\hlkwd{which}\hlstd{(PhotoMeas))} \hlopt{+} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(PhotoMeas)))} \hlopt{/} \hlnum{2}\hlstd{,} \hlnum{5}\hlstd{,}
+        \hlkwc{labels} \hlstd{=} \hlstr{"Photo"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
+\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{rect}\hlstd{(}\hlkwd{min}\hlstd{(}\hlkwd{which}\hlstd{(PhotoMeas)),} \hlnum{2}\hlstd{,} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(PhotoMeas)),} \hlnum{4}\hlstd{,}
+        \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{))}
+\hlcom{#with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 3, }
+\hlcom{#	labels = "Photo", col = "white", font = 3, cex = 0.9))}
+
+\hlkwd{dev.off}\hlstd{()}
+\end{alltt}
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+\begin{alltt}
+\hlcom{# Legend}
+\hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/ExperimentalDesignLegend.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{1.5}\hlstd{,} \hlnum{2}\hlstd{)}
+\hlkwd{par}\hlstd{(}\hlkwc{mai} \hlstd{=} \hlkwd{rep}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{4}\hlstd{))}
+\hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{1.5}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{1}\hlstd{),} \hlkwc{axes} \hlstd{= F)}
+
+\hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwd{c}\hlstd{(}\hlnum{0.8}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.55}\hlstd{,} \hlnum{0.55}\hlstd{,} \hlnum{0.35}\hlstd{),} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{,}
+                                \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(primShade0,} \hlnum{0.25}\hlstd{))}
+\hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwd{c}\hlstd{(}\hlnum{0.6}\hlstd{,} \hlnum{0.45}\hlstd{,} \hlnum{0.45}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.15}\hlstd{),} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{,}
+                                \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(sec1shade0,} \hlnum{0.25}\hlstd{))}
+\hlkwd{points}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{1}\hlstd{),} \hlkwd{c}\hlstd{(}\hlnum{0.575}\hlstd{,} \hlnum{0.775}\hlstd{,} \hlnum{0.775}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"l"}\hlstd{,} \hlkwc{col} \hlstd{= primShade3)}
+\hlkwd{points}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{1}\hlstd{),} \hlkwd{c}\hlstd{(}\hlnum{0.375}\hlstd{,} \hlnum{0.225}\hlstd{,} \hlnum{0.225}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"l"}\hlstd{,} \hlkwc{col} \hlstd{= sec1shade3)}
+\hlkwd{text}\hlstd{(}\hlnum{0.5}\hlstd{,} \hlnum{0.9}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Hot"}\hlstd{,} \hlkwc{col} \hlstd{= primShade3)}
+\hlkwd{text}\hlstd{(}\hlnum{0.5}\hlstd{,} \hlnum{0.1}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Cool"}\hlstd{,} \hlkwc{col} \hlstd{= sec1shade3)}
+\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{1}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Max"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(primShade0,} \hlnum{0.25}\hlstd{))}
+\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0.55}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Min"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(primShade0,} \hlnum{0.25}\hlstd{))}
+\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0.775}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Mean"}\hlstd{,} \hlkwc{col} \hlstd{= primShade3)}
+
+\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Min"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(sec1shade0,} \hlnum{0.25}\hlstd{))}
+\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0.45}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Max"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(sec1shade0,} \hlnum{0.25}\hlstd{))}
+\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0.225}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Mean"}\hlstd{,} \hlkwc{col} \hlstd{= sec1shade3)}
+
+\hlkwd{dev.off}\hlstd{()}
+\end{alltt}
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ExptlDes.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Experimental Design]{CAPTION}
+	\label{fig:Fig_ExptlDes}
+\end{figure}
+
+\subsection*{Measurements: Growth and photosynthesis}
 
@@
   \hlcom{# Plot latitude versus PC1}
-  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_PC1vLat.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{5}\hlstd{,} \hlnum{5}\hlstd{)}
-  \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
+  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_PC1vLat.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{5.5}\hlstd{,} \hlnum{5}\hlstd{)}
+  \hlkwd{par}\hlstd{(}\hlkwc{mai} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{1.5}\hlstd{,} \hlnum{0.25}\hlstd{,} \hlnum{0.25}\hlstd{))}
   \hlkwd{plot}\hlstd{(focClim}\hlopt{$}\hlstd{Lat, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{32}\hlstd{,} \hlnum{44}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{axes} \hlstd{= F,}
@@
     \hlkwd{list}\hlstd{(}\hlkwc{x} \hlstd{= pve)))}
-  \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{32}\hlstd{,} \hlnum{44}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{labels} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{expression}\hlstd{(}\hlnum{32}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{36}\hlopt{*}\hlstd{degree),}
-    \hlkwd{expression}\hlstd{(}\hlnum{40}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{44}\hlopt{*}\hlstd{degree)),} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
+  \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{32}\hlstd{,} \hlnum{44}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{labels} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{expression}\hlstd{(}\hlnum{32}\hlopt{*}\hlstd{degree),}
+                \hlkwd{expression}\hlstd{(}\hlnum{36}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{40}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{44}\hlopt{*}\hlstd{degree)),}
+                \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
   \hlkwd{axis}\hlstd{(}\hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{,} \hlnum{2}\hlstd{),} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
   \hlstd{fitLat} \hlkwb{<-} \hlkwd{lm}\hlstd{(PC1} \hlopt{~} \hlstd{Latitude,} \hlkwc{data} \hlstd{= focClim)}
+
+        \hlcom{# Label y-axis from slow to fast growth}
+        \hlkwd{mtext}\hlstd{(}\hlkwd{c}\hlstd{(}\hlstr{"Slow\textbackslash{}nGrowth"}\hlstd{,} \hlstr{"Fast\textbackslash{}nGrowth"}\hlstd{),} \hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{cex} \hlstd{=} \hlnum{1.25}\hlstd{,}
+                                \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{adj} \hlstd{=} \hlnum{0.5}\hlstd{)}
 
@@
 \subsection*{Selective agents and environmental correlates}
+
+As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
+temperature or precipitation) that covary with latitude.
 
@@
 
+%--------------------------------------------------
+% Results
+%--------------------------------------------------
+
 \section*{Results}
+
+\subsection*{Coordinated latitudinal clines in germination, growth, and photosynthesis}
+
+Genetic differences between populations affecting originsmal function and performance point to traits that may be involved in local adaptation.
+Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M.
+cardinalis} (Table~\ref{Table_AnovaSummary}).
+A single principal component captured 74.2 %\% of among population trait variation, defining an axis of variation from fast to slow growth (Fig~\ref{Fig_PC1vLat}).
+As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of Population, Treatment, Population x Treatment interactions for varies traits
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
     *** & 
-     &     *** & \\
+     &
+    *** & \\
   Water                            &
@@
 	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{CAPTION}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-	\input{Tables/Table_lllGrowthAnova.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-% ANOVA table for height growth rate (need to label tables)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, stem expansion rate]{CAPTION}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-	\input{Tables/Table_heightGrowthAnova.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-\subsection*{Latitude versus growth rate}
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations germinate sooner.]{CAPTION}
-	\label{fig:Fig_DoG}
+	\caption[Southern populations grow faster]{CAPTION}
+	\label{fig:Fig_PC1vLat}
 \end{figure}
 
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
-	\label{fig:Fig_LLL}
-\end{figure}
-
-\begin{figure}
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (stem elongation rate).]{CAPTION}
-	\label{fig:Fig_height}
-\end{figure}
+\subsection*{Remarkably little evidence for variation in plasticity}
+
+Genotype by environment interactions are also a common signature of local adaptation [CITE].
+For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
+We found remarkably little evidence for this pattern in \textit{M.
+cardinalis}.
+
+Do not think this is a power issue because statistical signifiance of treatment and populations effects were very high.
+
+\subsection*{Latitude and phenotypic clines are associated with more variable precipitation}
+
+%--------------------------------------------------
+% Discussion
+%--------------------------------------------------
 
 \section*{Discussion}
+
+1.
+How does local adaptation occur?
+It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
+For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
+
+2. Why does local adaptation succeed? shallow environmental gradients lead to convex tradeoffs not heavily constrained by gene flow.
+
+3. Selective agent - unpredictability. Explain
+
+4.
+Why does local adaptation fail?
+concave, high gene flow, demography.
+Qualify that our metrics of plasticity are coarse, but at least suggest that GxE is much weaker than G for the environmental gradients evaluated here.
+
+5.
+Future work and broad conclusions.
+Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
 

### 2015-04-22 ms/ms.Rnw
Worked on climate variable importance figure

@@
   par(mar = c(5, 6, 1, 1))
-  plot(0, 0, type = "n", xlim = c(0, ceiling(vsurfLat$ord.imp$x)[1]), ylim = c(1, 
-    length(vsurfLat$varselect.pred)), axes = F, frame.plot = T, xlab = "Importance", 
-    ylab = "", cex.lab = 1.5)
+  plot(0, 0, type = "n", xlim = c(0, ceiling(vsurfLat$ord.imp$x)[1]), 
+		ylim = c(1, length(vsurfLat$varselect.pred)), axes = F, frame.plot = T, 
+		xlab = "", ylab = "")
+	title(xlab = "Climate-Latitude\nImportance", line = 4, cex.lab = 1.5)
   for (i in 1:length(vsurfLat$varselect.pred)) 
@@
     length(vsurfLat$varselect.pred):1, pch = 21, col = "black", bg = "white", cex = 1.5)
-	mtext(climVars[vsurfLat$varselect.pred], side = 2, 
-    at = length(vsurfLat$varselect.pred):1, las = 1, line = 1)
+	# mtext(climVars[vsurfLat$varselect.pred], side = 2, 
+    # at = length(vsurfLat$varselect.pred):1, las = 1, line = 1)
 	axis(1, lwd = 0, lwd.ticks = 1)
@@
 
-	plot(0, 0, type = "n", xlim = c(0, 30), ylim = c(1, nImpLat), 
-		axes = F, frame.plot = T, xlab = "Importance", ylab = "", cex.lab = 1.5)
-	for (i in (nImpLat:(nImpLat - nImpPC1 + 1))) lines(x = c(-100, 100), y = c(i, i), 
-		col = "grey")
-	points(100 * vsurfPC1$ord.imp$x[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), pch = 21, 
-		col = "black", bg = "white", cex = 1.5)
-	mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, 
-    at = nImpLat:(nImpLat - nImpPC1 + 1), 
-		font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
-	axis(1, lwd = 0, lwd.ticks = 1, labels = seq(0, 3, 0.5), at = seq(0, 30, 5))
+	plot(0, 0, type = "n", xlim = c(0, 0.3), ylim = c(1, nImpLat), 
+		axes = F, frame.plot = T, xlab = "", ylab = "")
+	title(xlab = "Climate-Trait PC1\nImportance", line = 4, cex.lab = 1.5)
+	for (i in (nImpLat:(nImpLat - nImpPC1 + 1))) 
+	{
+		lines(x = c(-100, 100), y = c(i, i), col = "grey")
+	}
+	points(vsurfPC1$ord.imp$x[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), 
+		pch = 21, col = "black", bg = "white", cex = 1.5)
+	# mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, 
+  #  at = nImpLat:(nImpLat - nImpPC1 + 1), 
+	#	font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
+	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.3, 0.1))
   dev.off()
@@
 Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Treatment effects were highly significant ($P \lle 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
+However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
 
@@
 Exploratory approach to identify putative selective agent to test in future studies
-We used a machine learning approach...identified interannual variation in precipitation
+We used a machine learning approach...identified interannual variation in precipitation (Fig~\ref{fig:Fig_ClimVarImp})
 Obviously, cannot identify alternative variables that were not included in analysis.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of climatic variable importance
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimVarImp.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{CAPTION}
+	\label{fig:Fig_ClimVarImp}
+\end{figure}
 

### 2015-04-22 ms/ms.tex
Worked on climate variable importance figure

@@
 
+        \hlcom{# Load climVars}
+        \hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/ExploratoryCimateVariables.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+
   \hlcom{# Figure of variable importance for latitude}
@@
   \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{6}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
-  \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlkwd{ceiling}\hlstd{(vsurfLat}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x)[}\hlnum{1}\hlstd{]),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,}
-    \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)),} \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{xlab} \hlstd{=} \hlstr{"Importance"}\hlstd{,}
-    \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
+  \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlkwd{ceiling}\hlstd{(vsurfLat}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x)[}\hlnum{1}\hlstd{]),}
+                \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)),} \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,}
+                \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
+        \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Climate-Latitude\textbackslash{}nImportance"}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
   \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlnum{1}\hlopt{:}\hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred))}
@@
     \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)}\hlopt{:}\hlnum{1}\hlstd{,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlkwd{mtext}\hlstd{(climVars[vsurfLat}\hlopt{$}\hlstd{varselect.pred],} \hlkwc{side} \hlstd{=} \hlnum{2}\hlstd{,}
-    \hlkwc{at} \hlstd{=} \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)}\hlopt{:}\hlnum{1}\hlstd{,} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{1}\hlstd{)}
-\end{alltt}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in as.graphicsAnnot(text): object 'climVars' not found}}\begin{alltt}
+        \hlcom{# mtext(climVars[vsurfLat$varselect.pred], side = 2, }
+    \hlcom{# at = length(vsurfLat$varselect.pred):1, las = 1, line = 1)}
         \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
@@
 \begin{alltt}
-  \hlcom{# Columns with climatic variables to test}
-  \hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/Exploratory_climate_variables.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))} \hlcom{#climVars}
-
   \hlcom{# Variable selection using random forests (VSURF)}
   \hlstd{linmod.PC1} \hlkwb{<-} \hlkwd{as.formula}\hlstd{(}\hlkwd{paste}\hlstd{(}\hlstr{"PC1 ~ "}\hlstd{,} \hlkwd{paste}\hlstd{(climVars,} \hlkwc{collapse} \hlstd{=} \hlstr{" + "}\hlstd{),} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-  \hlkwd{set.seed}\hlstd{(}\hlnum{69646}\hlstd{)}
+  \hlkwd{set.seed}\hlstd{(}\hlnum{69647}\hlstd{)}
   \hlstd{vsurfPC1} \hlkwb{<-} \hlkwd{VSURF.parallel}\hlstd{(linmod.PC1,} \hlkwc{data} \hlstd{= focClim[,} \hlkwd{c}\hlstd{(}\hlstr{"PC1"}\hlstd{, climVars)],}
@@
 
-        \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{30}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, nImpLat),}
-                \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{xlab} \hlstd{=} \hlstr{"Importance"}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlstd{(nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{)))} \hlkwd{lines}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{100}\hlstd{,} \hlnum{100}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(i, i),}
-                \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
-        \hlkwd{points}\hlstd{(}\hlnum{100} \hlopt{*} \hlstd{vsurfPC1}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlstd{nImpPC1], nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{),} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,}
-                \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlkwd{mtext}\hlstd{(climVars[vsurfPC1}\hlopt{$}\hlstd{varselect.pred],} \hlkwc{side} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{1}\hlstd{,}
-    \hlkwc{at} \hlstd{= nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{),}
-                \hlkwc{font} \hlstd{=} \hlkwd{ifelse}\hlstd{(vsurfPC1}\hlopt{$}\hlstd{varselect.pred} \hlopt{%in%} \hlstd{vsurfLat}\hlopt{$}\hlstd{varselect.pred,}     \hlnum{2}\hlstd{,} \hlnum{1}\hlstd{))}
-        \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{3}\hlstd{,} \hlnum{0.5}\hlstd{),} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{30}\hlstd{,} \hlnum{5}\hlstd{))}
+        \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.3}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, nImpLat),}
+                \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
+        \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Climate-Trait PC1\textbackslash{}nImportance"}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
+        \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlstd{(nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{)))}
+        \hlstd{\{}
+                \hlkwd{lines}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{100}\hlstd{,} \hlnum{100}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(i, i),} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
+        \hlstd{\}}
+        \hlkwd{points}\hlstd{(vsurfPC1}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlstd{nImpPC1], nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{),}
+                \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
+        \hlcom{# mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, }
+  \hlcom{#  at = nImpLat:(nImpLat - nImpPC1 + 1), }
+        \hlcom{#	font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))}
+        \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.3}\hlstd{,} \hlnum{0.1}\hlstd{))}
   \hlkwd{dev.off}\hlstd{()}
@@
 Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis} (Table~\ref{Table_AnovaSummary}).
-A single principal component captured 74.2 %\% of among population trait variation, defining an axis of variation from fast to slow growth (Fig~\ref{Fig_PC1vLat}).
+cardinalis} (Table~\ref{table:Table_AnovaSummary}).
+A single principal component captured 74.2 %$\%$ of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
 As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
@@
 
-Genotype by environment interactions are also a common signature of local adaptation [CITE].
+Genotype $\times$ environment interactions are also a common signature of local adaptation [CITE].
 For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
@@
 cardinalis}.
-
-Do not think this is a power issue because statistical signifiance of treatment and populations effects were very high.
+There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
+Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
+Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
+However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
 
 \subsection*{Latitude and phenotypic clines are associated with more variable precipitation}
+
+Latitude explains variation
+Presumably selective agent is not latitude, but something related to it
+Exploratory approach to identify putative selective agent to test in future studies
+We used a machine learning approach...identified interannual variation in precipitation (Fig~\ref{fig:Fig_ClimVarImp})
+Obviously, cannot identify alternative variables that were not included in analysis.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of climatic variable importance
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimVarImp.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{CAPTION}
+	\label{fig:Fig_ClimVarImp}
+\end{figure}
 

### 2015-04-23 ms/ms.Rnw
Pushing from linux

@@
   # Manuscript files
-  pathMS <- "~/Google Drive/CardLocalAdaptation/ms"
+  pathMS <- "~/Google Drive/CardLocalAdaptation/ms" # Chris' MacBook
+  pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop
 
@@
 # Read in climate data for 16 focal populations to use in making Table 1
+# need to make this available from ms directory
 load('~/Google Drive/CardLocalAdaptation/Data/Climate/Robjects/focClim.RData') # focClim

### 2015-04-24 ms/ms.Rnw
Small changes

@@
   pathMS <- "~/Google Drive/CardLocalAdaptation/ms" # Chris' MacBook
-  pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop
+  # pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop
 

### 2015-04-24 ms/ms.Rnw
Worked on Methods.

@@
 1. `intrinsic' physiological variation
-2. variation in stress response
+2. variation in plasticity
 3. what ecological factors explain divergence (i.e. putative selective agents underlying local adaptation)
@@
 On 5 May (21 days after sowing), seedlings were transferred to one of two MODEL Growth Chambers (Conviron, Manitoba, Canada).
-We thinned seedlings to one plant per cone-tainer, leaving the center-most, largest plant.
+We thinned seedlings to one plant per cone-tainer, leaving the center-most plant.
 702 of 768 (91.4\%) had plants that could be used in the experiment.
@@
 
-\paragraph*{Monitoring environment}
-
-  	Cross check with Poorter paper \\
-		Light (PAR sensors), temp/humidity (HOBOs), soil moisture
-
 \subsection*{Treatments}
 
-We used 2 watering treatments on two populations on two soils. The watering treatments are: well-watered, drought.
-
-\subsubsection*{Temperature}
-
-We simulated typical growing season (June and July) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (High Temp) and Little Jameson (Low Temp).
+We imposed four treatments, a fully-factorial cross of two temperature treatments and two watering treatments.
+Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment and another to the Cool treatment.
+Within each chamber, there were two Well-watered blocks and two Drought blocks.
+A detailed description of treatments is given below \cdm{should some be moved to supplement?} and summarized in Fig~\ref{fig:Fig_ExptlDes}.
+The irradiance in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$.
+The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature treatments (data not shown).
+
+\subsubsection*{Temperature treatments}
+
+We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool).
 We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
-Daily temperatures from ClimateWNA are usually highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
+Daily temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
 Angert, unpub.
 data).
-Hence, the ClimateWNA temperature profiles are likely to similar to actual thermal regimes experienced by \textit{M.
+Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M.
 cardinalis} in nature.
-To create realistic temperature regimes, we calculated the mean temperature trend from June to July using LOESS \citep{Cleveland_etal_1992}.
+We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}.
 The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ).
 The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
-We fit and simulated from the VAR(2) using the package \pkg{dse} \citep{Gilbert_2014} in R.
+We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R.
 Simulated data closely resembled the autocorrelation and between-site correlation of the actual data.
@@
 For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
-During the middle of the day, chamber temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
+During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
 
@@
 
-      April 14 germinate
+For watering treatments, we simulated two extreme types of streams where \textit{M.
+cardinalis} grows.
+In the well-watered treatment, we simulated a large stream with that never goes dry during the summer growing season.
+In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down as the winter snow pack melts.
+In both treatments, plants were bottom-watered using with water from chilled by MAKE AND MODEL OF CHILLER to 7.5 degrees???
+Check temp!
+MARKER - continue here!!!
 Before May 20 (36 days): 7.5 min (complete flood) every 2 hours

### 2015-04-27 ms/ms.Rnw
Worked on ms

@@
 # Read in climate data for 16 focal populations to use in making Table 1
-# need to make this available from ms directory
-load('~/Google Drive/CardLocalAdaptation/Data/Climate/Robjects/focClim.RData') # focClim
+# [MAYBE SHOULD RENAMED "ExtendedTable1" or something?]
+load(paste(pathDat, "/focClim.RData", sep = ""))
 
@@
 
-\subsubsection*{Water}
+\subsubsection*{Watering treatments}
 
@@
 cardinalis} grows.
-In the well-watered treatment, we simulated a large stream with that never goes dry during the summer growing season.
+In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
 In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down as the winter snow pack melts.
-In both treatments, plants were bottom-watered using with water from chilled by MAKE AND MODEL OF CHILLER to 7.5 degrees???
-Check temp!
-MARKER - continue here!!!
-Before May 20 (36 days): 7.5 min (complete flood) every 2 hours
-      between May 20 to June 10 (57 days), down to 30 seconds, 1x per day
-			Wet: daily/constant irrigation with cooled water
-			Dry: start at week \#, gradually decrease watering frequency/level
+In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER.
+Plants in the well-watered treatment were fully saturated every two hours during the day.
+Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing).
+Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.
 
@@
 
-\subsection*{Measurements: Growth and photosynthesis}
+\subsection*{Growth and photosynthesis}
 
@@
    \end{tabular}
+   \label{table:Table_traits}
 \end{table}
@@
 
-\cdm{which traits: germination, growth rate, biomass, biomass allometry, leaf traits, photosynthesis?, mortality?}
-We tested for Population, Treatment, and Population $\times$ Treatment interactions using mixed model ANOVA with Family included as a random factor.
+For all traits (Table~\ref{table:Table_traits}) we tested for Population, Treatment, and Population $\times$ Treatment interactions.
+We interpreted significant Population effects to indicate intrinsic variation and Population by Treatment effects to indicate variation in plasticity.
+As mentioned above, we survival and GLM models for germination rate and mortality, respectively.
+For all other traits, we used mixed model ANOVAs with Family included as a random factor.
 Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}.
@@
 \subsection*{Principal components of germination, growth, and photosynthesis}
-
+[maybe this goes in Results?]
+We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components. Spec
 <<Principal Components>>=
@@
 temperature or precipitation) that covary with latitude.
+
+MARKER - continue here!!!
+
+(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
+Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
+Some hypotheses:
+\begin{itemize}
+  \item{H1: Latitude encompasses multivariate climatic dimension.}
+  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
+\end{itemize}
+
+Predictions:
+\begin{itemize}
+  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
+Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
+  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
+  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
+\end{itemize}
+
+Methods specific to latitude vs. trait correlation:
+We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+cardinalis}.
+We used a comprehensive set of \cdm{358 in total before thinning} recent (since 2000) known \textit{M.
+cardinalis} occurences recorded from herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
+Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
+For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
+In total, we included X environmental variables from ClimateWNA and 17?
+bioclimatic variables (Table S\#).
+For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
+Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
+
+Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
+We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation.
+We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
+Since \textit{M.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.
+
+To identify climatic variables that were closely associated with latitude in \textit{M.
+cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
+In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.
+
+Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions.
+If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.
+
+Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
+, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
+Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
+To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
+
+% Methods for spatial-average selection
+For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.
 
@@
 @
-
-(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
-Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
-Some hypotheses:
-\begin{itemize}
-  \item{H1: Latitude encompasses multivariate climatic dimension.}
-  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
-\end{itemize}
-
-Predictions:
-\begin{itemize}
-  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
-Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
-  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
-  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
-\end{itemize}
-
-Methods specific to latitude vs. trait correlation:
-We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
-cardinalis}.
-We used a comprehensive set of \cdm{358 in total before thinning} recent (since 2000) known \textit{M.
-cardinalis} occurences recorded from herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
-Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included X environmental variables from ClimateWNA and 17?
-bioclimatic variables (Table S\#).
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
-Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
-
-Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
-We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation.
-We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.
-
-To identify climatic variables that were closely associated with latitude in \textit{M.
-cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
-In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.
-
-Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions.
-If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.
-
-Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
-, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
-Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
-To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
-
-% Methods for spatial-average selection
-For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.
 

### 2015-04-27 ms/ms.Rnw
Worked on ms

@@
 
+<<thinnedOccClimate>>=
+# R data.frame of 178 M. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data
+load(paste(pathDat, "/thinnedOccClimate.RData", sep = ""))
+@
+
 \subsection*{Selective agents and environmental correlates}
@@
 temperature or precipitation) that covary with latitude.
-
-MARKER - continue here!!!
-
-(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
-Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
-Some hypotheses:
-\begin{itemize}
-  \item{H1: Latitude encompasses multivariate climatic dimension.}
-  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
-\end{itemize}
-
-Predictions:
-\begin{itemize}
-  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
-Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
-  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
-  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
-\end{itemize}
-
-Methods specific to latitude vs. trait correlation:
 We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
 cardinalis}.
-We used a comprehensive set of \cdm{358 in total before thinning} recent (since 2000) known \textit{M.
-cardinalis} occurences recorded from herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+We did this by looking for overlap between climatic variables that best predict latitude of \textit{M.
+cadinalis} populations and climatic variables that best predict trait variation across our 16 focal populations.
+We used a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
+cardinalis} occurences.
+These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
 For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
@@
 In total, we included X environmental variables from ClimateWNA and 17?
-bioclimatic variables (Table S\#).
+bioclimatic variables (Table~\ref{table:TableS_}).
 For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
@@
 
+(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
+Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
+Some hypotheses:
+\begin{itemize}
+  \item{H1: Latitude encompasses multivariate climatic dimension.}
+  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
+\end{itemize}
+
+Predictions:
+\begin{itemize}
+  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
+Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
+  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
+  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
+\end{itemize}
+
 \end{document}

### 2015-04-27 ms/ms.tex
Worked on ms

@@
   \hlcom{# Manuscript files}
-  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"}
+  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"} \hlcom{# Chris' MacBook}
+  \hlcom{# pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop}
 
@@
 For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
-During the middle of the day, chamber temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
-
-\subsubsection*{Water}
-
-      April 14 germinate
-      Before May 20 (36 days): 7.5 min (complete flood) every 2 hours
-      between May 20 to June 10 (57 days), down to 30 seconds, 1x per day
-			Wet: daily/constant irrigation with cooled water
-			Dry: start at week \#, gradually decrease watering frequency/level
+During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
+
+\subsubsection*{Watering treatments}
+
+For watering treatments, we simulated two extreme types of streams where \textit{M.
+cardinalis} grows.
+In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
+In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down as the winter snow pack melts.
+In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER.
+Plants in the well-watered treatment were fully saturated every two hours during the day.
+Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing).
+Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.
 
@@
 \subsection*{Principal components of germination, growth, and photosynthesis}
-
+[maybe this goes in Results?]
+We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components. Spec
 \begin{knitrout}
@@
 
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# R data.frame of 178 M. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data}
+\hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/thinnedOccClimate.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
 \subsection*{Selective agents and environmental correlates}
@@
 temperature or precipitation) that covary with latitude.
+We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+cardinalis}.
+We did this by looking for overlap between climatic variables that best predict latitude of \textit{M.
+cadinalis} populations and climatic variables that best predict trait variation across our 16 focal populations.
+We used a representative set of 178 recent (since 2000) known \textit{M.
+cardinalis} occurences.
+These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
+Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
+For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
+In total, we included X environmental variables from ClimateWNA and 17?
+bioclimatic variables (Table~\ref{table:TableS_}).
+For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
+Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
+
+Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
+We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation.
+We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
+Since \textit{M.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.
+
+To identify climatic variables that were closely associated with latitude in \textit{M.
+cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
+In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.
+
+Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions.
+If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.
+
+Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
+, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
+Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
+To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
+
+% Methods for spatial-average selection
+For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.
 
@@
 \end{knitrout}
-
-(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
-Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
-Some hypotheses:
-\begin{itemize}
-  \item{H1: Latitude encompasses multivariate climatic dimension.}
-  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
-\end{itemize}
-
-Predictions:
-\begin{itemize}
-  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
-Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
-  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
-  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
-\end{itemize}
-
-Methods specific to latitude vs. trait correlation:
-We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
-cardinalis}.
-We used a comprehensive set of \cdm{358 in total before thinning} recent (since 2000) known \textit{M.
-cardinalis} occurences recorded from herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
-Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included X environmental variables from ClimateWNA and 17?
-bioclimatic variables (Table S\#).
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
-Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
-
-Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
-We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation.
-We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.
-
-To identify climatic variables that were closely associated with latitude in \textit{M.
-cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
-In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.
-
-Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions.
-If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.
-
-Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
-, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
-Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
-To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
-
-% Methods for spatial-average selection
-For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.
 

### 2015-04-28 ms/ms.Rnw
Added VSURF for spatially-averaged climate

@@
 
+# Read in spatially-averaged (10 km buffer) for 16 focal populations
+load(paste(pathDat, "/focClimSA.RData", sep = ""))
+
 @
@@
       \midrule
-	HAU & South Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "HAU"], 3)}	& 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "HAU"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "HAU"]}   \\
-	CTC	& South Margin		& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "CTC"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "CTC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "CTC"]}   \\
-	CUR	& South Margin		& yes	& yes & \Sexpr{round(focClim$Latitude[focClim$ID == "CUR"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "CUR"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "CUR"]}   \\
-	GRP & South Margin		& no	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "GRP"], 3)} &
-    \Sexpr{round(focClim$Longitude[focClim$ID == "GRP"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "GRP"]}   \\
-	WWC &	Transverse		  & yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "WWC"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "WWC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WWC"]}   \\
-	MIL	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "MIL"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "MIL"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "MIL"]}   \\
-	WFM	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "WFM"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "WFM"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WFM"]}   \\
-	NMT	& South Sierras		& yes	& yes?& \Sexpr{round(focClim$Latitude[focClim$ID == "NMT"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "NMT"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "NMT"]}   \\
-	PRD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "PRD"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "PRD"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "PRD"]}   \\
-	RWD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "RWD"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "RWD"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RWD"]}   \\
-	WNA	& Central Sierras	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "WNA"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "WNA"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "WNA"]}   \\
-	RBW	& Central Sierras	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$ID == "RBW"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "RBW"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RBW"]}   \\
-	MYU	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "MYU"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "MYU"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "MYU"]}   \\
-	LIJ	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "LIJ"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "LIJ"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "LIJ"]}   \\
-	DPC	& North Coast	  	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "DPC"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "DPC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "DPC"]}   \\
-	RCC	& North Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$ID == "RCC"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$ID == "RCC"], 3)}	& \Sexpr{focClim$Elevation[focClim$ID == "RCC"]}   \\
+	HAU & South Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Hauser Creek"], 3)}	& 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Hauser Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Hauser Creek"]}   \\
+	CTC	& South Margin		& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Cottonwood Creek"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Cottonwood Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Cottonwood Creek"]}   \\
+	CUR	& South Margin		& yes	& yes & \Sexpr{round(focClim$Latitude[focClim$PopID == "Cuyamaca Rancho"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Cuyamaca Rancho"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Cuyamaca Rancho"]}   \\
+	GRP & South Margin		& no	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Grade Road Palomar"], 3)} &
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Grade Road Palomar"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Grade Road Palomar"]}   \\
+	WWC &	Transverse		  & yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Whitewater Canyon"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Whitewater Canyon"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Whitewater Canyon"]}   \\
+	MIL	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Mill Creek"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Mill Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Mill Creek"]}   \\
+	WFM	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "West Fork Mojave River"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "West Fork Mojave River"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "West Fork Mojave River"]}   \\
+	NMT	& South Sierras		& yes	& yes?& \Sexpr{round(focClim$Latitude[focClim$PopID == "North Fork Middle Tule"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "North Fork Middle Tule"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "North Fork Middle Tule"]}   \\
+	PRD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Paradise Creek"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Paradise Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Paradise Creek"]}   \\
+	RWD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Redwood Creek"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Redwood Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Redwood Creek"]}   \\
+	WNA	& Central Sierras	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Wawona"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Wawona"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Wawona"]}   \\
+	RBW	& Central Sierras	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Rainbow Pool"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Rainbow Pool"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Rainbow Pool"]}   \\
+	MYU	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Middle Yuba"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Middle Yuba"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Middle Yuba"]}   \\
+	LIJ	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Little Jameson"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Little Jameson"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Little Jameson"]}   \\
+	DPC	& North Coast	  	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Deep Creek"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Deep Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Deep Creek"]}   \\
+	RCC	& North Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Rock Creek"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "Rock Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Rock Creek"]}   \\
 	\bottomrule
@@
   focClim$PC1 <- traitPC$x[, "PC1"]
+  focClimSA$PC1 <- traitPC$x[, "PC1"]
 
@@
 
-As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
+As with previous studies in \textit{Mimulus} and many other species [CITES, Huey_etal, mosquitoes, what else?], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
 temperature or precipitation) that covary with latitude.
@@
 We did this by looking for overlap between climatic variables that best predict latitude of \textit{M.
-cadinalis} populations and climatic variables that best predict trait variation across our 16 focal populations.
+cadinalis} occurence records and climatic variables that best predict trait variation across our 16 focal populations.
 We used a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
@@
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included X environmental variables from ClimateWNA and 17?
-bioclimatic variables (Table~\ref{table:TableS_}).
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
+In total, we included \Sexpr{length(climVars) / 2} cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
 Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
@@
     ncores = 3)
+  vsurfPC1_SA <- VSURF.parallel(linmod.PC1, data = focClimSA[, c("PC1", climVars)], 
+    ncores = 3)
 
@@
   # Overlap between latitude predictors and phenotype predictors
+#### NEED TO COMPARE RESULTS FROM POINT DATA AND SPATIALLY AVERAGED CLIMATE!
   imp2 <- climVars[vsurfPC1$varselect.pred]
+  # imp2 <- climVars[vsurfPC1_SA$varselect.pred]
   nImpPC1 <- length(imp2)
@@
    \end{tabular}
+   \label{table:TableS_ClimVars}
 \end{table}

### 2015-04-28 ms/ms.tex
Added VSURF for spatially-averaged climate

@@
       \midrule
-	HAU & South Margin		& yes	& yes	& 	& 
-    	&    \\
-	CTC	& South Margin		& yes	& no	&  & 
-    	&    \\
-	CUR	& South Margin		& yes	& yes &  & 
-    	&    \\
-	GRP & South Margin		& no	& no	&  &
-    	&    \\
-	WWC &	Transverse		  & yes	& yes	&  & 
-    	&    \\
-	MIL	& Transverse	  	& yes	& no	&  & 
-    	&    \\
-	WFM	& Transverse	  	& yes	& no	&  & 
-    	&    \\
-	NMT	& South Sierras		& yes	& yes?&  & 
-    	&    \\
-	PRD	& South Sierras		& yes	& yes	&  & 
-    	&    \\
-	RWD	& South Sierras		& yes	& yes	&  & 
-    	&    \\
-	WNA	& Central Sierras	& yes	& yes	&  & 
-    	&    \\
-	RBW	& Central Sierras	& yes	& no	&  & 
-    	&    \\
-	MYU	& North Sierras		& yes	& yes	&  & 
-    	&    \\
-	LIJ	& North Sierras		& yes	& yes	&  & 
-    	&    \\
-	DPC	& North Coast	  	& yes	& yes	&  & 
-    	&    \\
-	RCC	& North Margin		& yes	& yes	&  & 
-    	&    \\
+	HAU & South Margin		& yes	& yes	& 32.657	& 
+    \ensuremath{-116.532}	& 799   \\
+	CTC	& South Margin		& yes	& no	& 32.609 & 
+    \ensuremath{-116.7}	& 267   \\
+	CUR	& South Margin		& yes	& yes & 32.9 & 
+    \ensuremath{-116.585}	& 1180   \\
+	GRP & South Margin		& no	& no	& 33.314 &
+    \ensuremath{-116.871}	& 1577   \\
+	WWC &	Transverse		  & yes	& yes	& 33.994 & 
+    \ensuremath{-116.665}	& 705   \\
+	MIL	& Transverse	  	& yes	& no	& 34.077 & 
+    \ensuremath{-116.873}	& 2050   \\
+	WFM	& Transverse	  	& yes	& no	& 34.284 & 
+    \ensuremath{-117.378}	& 1120   \\
+	NMT	& South Sierras		& yes	& yes?& 36.201 & 
+    \ensuremath{-118.651}	& 1314   \\
+	PRD	& South Sierras		& yes	& yes	& 36.518 & 
+    \ensuremath{-118.759}	& 926   \\
+	RWD	& South Sierras		& yes	& yes	& 36.691 & 
+    \ensuremath{-118.91}	& 1727   \\
+	WNA	& Central Sierras	& yes	& yes	& 37.541 & 
+    \ensuremath{-119.649}	& 1224   \\
+	RBW	& Central Sierras	& yes	& no	& 37.819 & 
+    \ensuremath{-120.007}	& 876   \\
+	MYU	& North Sierras		& yes	& yes	& 39.397 & 
+    \ensuremath{-121.082}	& 455   \\
+	LIJ	& North Sierras		& yes	& yes	& 39.743 & 
+    \ensuremath{-120.704}	& 1603   \\
+	DPC	& North Coast	  	& yes	& yes	& 41.668 & 
+    \ensuremath{-123.11}	& 707   \\
+	RCC	& North Margin		& yes	& yes	& 43.374 & 
+    \ensuremath{-122.957}	& 326   \\
 	\bottomrule
@@
 
-{\ttfamily\noindent\color{warningcolor}{\#\# Warning in VSURF.parallel.formula(linmod.PC1, data = focClim[, c("{}PC1"{}, : VSURF with a formula-type call outputs selected variables\\\#\# which are indices of the input matrix based on the formula:\\\#\# you may reorder these to get indices of the original data}}\begin{alltt}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in model.frame.default(formula = ord.sd \textasciitilde{} ., data = u, na.action = function (x) : variable lengths differ (found for 'u')}}\begin{alltt}
   \hlcom{# Plot most important climatic predictor versus PC1}
@@
   \hlstd{imp2} \hlkwb{<-} \hlstd{climVars[vsurfPC1}\hlopt{$}\hlstd{varselect.pred]}
+\end{alltt}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'vsurfPC1' not found}}\begin{alltt}
   \hlstd{nImpPC1} \hlkwb{<-} \hlkwd{length}\hlstd{(imp2)}
-
+\end{alltt}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'imp2' not found}}\begin{alltt}
   \hlstd{vsurfLat}\hlopt{$}\hlstd{varselect.pred}
@@
         \hlstd{\}}
+\end{alltt}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): object 'nImpPC1' not found}}\begin{alltt}
         \hlkwd{points}\hlstd{(vsurfPC1}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlstd{nImpPC1], nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{),}
                 \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
+\end{alltt}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in points(vsurfPC1\$ord.imp\$x[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + : object 'vsurfPC1' not found}}\begin{alltt}
         \hlcom{# mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, }

### 2015-04-28 ms/ms.Rnw
Worked on Results.

@@
 cadinalis} occurence records and climatic variables that best predict trait variation across our 16 focal populations.
-We used a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
+For brevity, we refer to these as Climate-Latitude and Climate-Trait variables.
+We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}).
+From VSURF models, we kept only variables selected for prediction, the most stringent criteria.
+
+For these analyses, we compiled a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
 cardinalis} occurences.
@@
 For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
-Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
-
-Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
-We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation.
-We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.
-
-To identify climatic variables that were closely associated with latitude in \textit{M.
-cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
-In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.
-
-Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions.
-If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.
-
-Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
-, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
-Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
-To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
-
-% Methods for spatial-average selection
-For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.
+Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.
 
@@
 
-\subsection*{Coordinated latitudinal clines in germination, growth, and photosynthesis}
+\subsection*{A coordinated latitudinal cline in germination, growth, and photosynthesis}
 
@@
 
-\subsection*{Latitude and phenotypic clines are associated with more variable precipitation}
+\subsection*{Climatic variability best explains phenotypic divergence}
 
@@
 
+\subsection*{Spatially-averaged climate}
+
+% Methods for spatial-average selection
+Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
+We therefore calculated spatially averaged climate means and coefficients of variation.
+We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
+Since \textit{M.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
+
 \end{document}

### 2015-04-28 ms/ms.Rnw
Worked on Resuls and Discussion.

@@
 [maybe this goes in Results?]
-We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components. Spec
+We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components. 
 <<Principal Components>>=
@@
 
-Latitude explains variation
-Presumably selective agent is not latitude, but something related to it
-Exploratory approach to identify putative selective agent to test in future studies
-We used a machine learning approach...identified interannual variation in precipitation (Fig~\ref{fig:Fig_ClimVarImp})
-Obviously, cannot identify alternative variables that were not included in analysis.
+Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
+For \textit{M.
+cardinalis}, interannual variation in precipitation over the past 30 years is very closely related to the latitude of recently-recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
+The two most important Climate-Latitude variables were also strong predictors of position along the fast-slow growth axis (Fig.~\ref{fig:Fig_ClimVarImp}B).
+This overlap between Climate-Latitude and Climate-Trait variables suggests that interannual variation in precipitation is an important selective agent in \textit{M.
+cardinalis}.
+Specifically, we hypothesize that more frequent droughts (greater precipitation cv) in Southern populations selects for an `annual-ized' life history, as we detail in the Discussion.
+We must qualify these results because our analysis obviously cannot rule out that alternative variables not included in the analysis may be more important.
 
@@
 
-1.
+The observation that populations are locally adapted to their environment is old and well-documented in many species.
+New frontiers in local adaptation research aim at a general understanding of why local adaptation sometimes suceeds and other times fails, why certain traits respond to selection whereas others are constrained, and what are the important environmental drivers of local adaptation [CITES such as Kawecki].
+Understanding why and how populations adapt to different environments is especially significant for predicting how species' ranges will shift under climate change.
+In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint.
+...
+
+<- add to above: 1.
 How does local adaptation occur?
@@
 - concave tradeoffs, fine-grained (coarse+concave would predict alt morphs), high gene flow
+
+%--------------------------------------------------
+% Acknowledgements
+%--------------------------------------------------
+
+\section*{Acknowledgements}
+Erin Warkman and Lisa Lin helped collect data.
 
@@
 	\end{center}
+	\lable{tables:TableS_InitialSize}
 \end{table}

### 2015-04-29 ms/ms.Rnw
worked on Discussion

@@
 The observation that populations are locally adapted to their environment is old and well-documented in many species.
-New frontiers in local adaptation research aim at a general understanding of why local adaptation sometimes suceeds and other times fails, why certain traits respond to selection whereas others are constrained, and what are the important environmental drivers of local adaptation [CITES such as Kawecki].
-Understanding why and how populations adapt to different environments is especially significant for predicting how species' ranges will shift under climate change.
-In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint.
-...
-
-<- add to above: 1.
-How does local adaptation occur?
-It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
-For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
+New frontiers in local adaptation research seek a general understanding of why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
+Understanding why and how populations have adapted to different environments in the past is especially significant npw because it the scientific basis for predicting how species' ranges will shift under climate change [CITES, Catullo_etal_2015].
+In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint related to local adaptation.
+In particular, our data suggest that photosynthesis and intrinsic growth rate vary among populations, whereas the fundamental abiotic niche is more conserved, and that climatic variation may be a more important selective agent than the average.
+Next, we tie these results into the broader threads of evolutionary theory in order to figure out what factors best explain our data.
+
+As long as environments vary and there exists genetic variation in fitness, why shouldn't local adaptation suceed?
+Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain local adaptation.
+In \textit{M.
+cardinalis} we found substantial genetic variation among populations along a phenotypic axis from fast to slow growth (Fig.~\ref{fig:Fig_PC1vLat}).
+Th...
+fitness tradeoffs are not too strong, demographic symmetry, and attenuation of gene flow.
 
@@
 
+Although evolutionary theory ... But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
+
 3. Selective agent - unpredictability. Explain
 
+(also 4.) Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years, could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
+We found no evidence for this; all populations responded to drought and temperature treatments similarly (Table~\ref{table:Table_AnovaSummary}).
+However, evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
+Riparian habitats are highly heterogeneous at small spatial scales.
+Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought if there is little direct precipitation during the growing season, as in Meditteranean climates of Western North America.
+But alleles that confer greater drought tolerance may be quite costly in well-watered soils, leading to strong fitness tradeoffs.
+Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
+Demography and gene flow may reinforce niche conservatism.
+A new mutant with increased drought tolerance that can survive at the margin of a population.
+But if .
+Finally, gene flow, which is generally high among \textit{M.
+cardinalis} populations within the same ecoregion \citep{Paul_etal_2015}, will thwart local adaptation and reinforce specialization.
+Spatial grain of the environment, demographic asymmetry, and gene flow conspire to thwart local adaptation via altered fundamental niche.
 4.
@@
 Qualify that our metrics of plasticity are coarse, but at least suggest that GxE is much weaker than G for the environmental gradients evaluated here.
+
+highly connected by gene flow, outcrossing (self-compatible but strongly herkogamous)
+
+?
+It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
+For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
 

### 2015-04-29 ms/ms.Rnw
Worked on Resuls and Discussion.

@@
 \usepackage{gensymb}
+\usepackage{setspace}
+\newcommand{\stretchy}{1.5}
+\usepackage{lineno}
 
@@
 
-% write/analyze
-% proposal for summer
-  % local fitness set concave
-  % range-wide fitness set convex
-  % interaction between shape of fitness set and gene flow?
-% evo-phys proposal
+\setstretch{\stretchy}
 
 \section*{Introduction}
+
+% Toggle these commands to get version for doing word count
+\linenumbers
+%\pagenumbering{gobble}
+%\raggedright
 
@@
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included \Sexpr{length(climVars) / 2} cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+In total, we included %\Sexpr{length(climVars) / 2}%
+cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
 For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
@@
 @
-
-\begin{table}[ht]
-   \centering
-   \topcaption{Climatic variables}
-   \begin{tabular}{@{} ll @{}}
-      \toprule
-  Climate variable & Abbreviation \\
-      \midrule
-  % climateWNA VARIABLES:
-	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
-	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
-	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
-	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
-	NFFD	& number of frost-free days                           \\
-	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
-	Eref	& Hargreaves reference evaporation (mm) \\
-	CMD	  & Hargreaves climativ moisture deficit (mm) \\
-	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
-	RH    &	mean annual relative humidity \\
-  % BIOCLIM VARIABLES:
-  bio1	&	Annual Mean Temperature	\\
-	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
-	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
-	bio4	&	Temperature Seasonality (standard deviation *100)	\\
-	bio5	&	Max Temperature of Warmest Month	\\
-	bio6	&	Min Temperature of Coldest Month	\\
-	bio7	&	Temperature Annual Range (bio5-bio6)	\\
-	bio8	&	Mean Temperature of Wettest Quarter	\\
-	bio9	&	Mean Temperature of Driest Quarter	\\
-	bio10	&	Mean Temperature of Warmest Quarter	\\
-	bio11	&	Mean Temperature of Coldest Quarter	\\
-	bio12	&	Annual Precipitation	\\
-	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
-	bio16	&	Precipitation of Wettest Quarter	\\
-	bio17	&	Precipitation of Driest Quarter	\\
-	bio18	&	Precipitation of Warmest Quarter	\\
-	bio19	&	Precipitation of Coldest Quarter	\\
-	    \bottomrule
-   \end{tabular}
-   \label{table:TableS_ClimVars}
-\end{table}
 
@@
 	\selectfont
-	\caption[Southern populations grow faster]{CAPTION}
+	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude.
+Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods).
+The line and 95\% confidence intervals (grey polygon) were estimated using standard linear regression.}
 	\label{fig:Fig_PC1vLat}
@@
 	\selectfont
-	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{CAPTION}
+	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{Interannual variation in precipitation is closely correlated with latitude and trait variation.
+A.
+Using Random Forest regression, we identified 10 climatic variables significantly (high importance) associated with latitude of \textit{M.
+cardinalis} occurences.
+B.
+The two most important Climate-Latitude variables were also the two most important Climate-Trait variables.
+Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
+Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp) or Precipitation (Prec); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Dry Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
 	\label{fig:Fig_ClimVarImp}
@@
 The observation that populations are locally adapted to their environment is old and well-documented in many species.
-New frontiers in local adaptation research seek a general understanding of why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
-Understanding why and how populations have adapted to different environments in the past is especially significant npw because it the scientific basis for predicting how species' ranges will shift under climate change [CITES, Catullo_etal_2015].
-In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint related to local adaptation.
+New frontiers in local adaptation research seek a general explanation for why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
+Understanding why and how populations have adapted to different environments in the past is especially significant now as a scientific basis for predicting how species' ranges will shift under climate change [CITES, Catullo_etal_2015].
+In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint.
 In particular, our data suggest that photosynthesis and intrinsic growth rate vary among populations, whereas the fundamental abiotic niche is more conserved, and that climatic variation may be a more important selective agent than the average.
@@
 
-As long as environments vary and there exists genetic variation in fitness, why shouldn't local adaptation suceed?
-Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain local adaptation.
+As long as environments vary and there exists genetic variation in fitness, why shouldn't local adapted genotypes evolve?
+Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001}.
 In \textit{M.
-cardinalis} we found substantial genetic variation among populations along a phenotypic axis from fast to slow growth (Fig.~\ref{fig:Fig_PC1vLat}).
-Th...
+cardinalis} we found substantial genetic variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
+If this variation is adaptive, it could be maintained as long as fitness tradeoffs are not too Th...
 fitness tradeoffs are not too strong, demographic symmetry, and attenuation of gene flow.
@@
 
+\begin{table}[ht]
+   \centering
+   \topcaption{Climatic variables}
+   \begin{tabular}{@{} ll @{}}
+      \toprule
+  Climate variable & Abbreviation \\
+      \midrule
+  % climateWNA VARIABLES:
+	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
+	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
+	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
+	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
+	NFFD	& number of frost-free days                           \\
+	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
+	Eref	& Hargreaves reference evaporation (mm) \\
+	CMD	  & Hargreaves climativ moisture deficit (mm) \\
+	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
+	RH    &	mean annual relative humidity \\
+  % BIOCLIM VARIABLES:
+  bio1	&	Annual Mean Temperature	\\
+	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
+	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
+	bio4	&	Temperature Seasonality (standard deviation *100)	\\
+	bio5	&	Max Temperature of Warmest Month	\\
+	bio6	&	Min Temperature of Coldest Month	\\
+	bio7	&	Temperature Annual Range (bio5-bio6)	\\
+	bio8	&	Mean Temperature of Wettest Quarter	\\
+	bio9	&	Mean Temperature of Driest Quarter	\\
+	bio10	&	Mean Temperature of Warmest Quarter	\\
+	bio11	&	Mean Temperature of Coldest Quarter	\\
+	bio12	&	Annual Precipitation	\\
+	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
+	bio16	&	Precipitation of Wettest Quarter	\\
+	bio17	&	Precipitation of Driest Quarter	\\
+	bio18	&	Precipitation of Warmest Quarter	\\
+	bio19	&	Precipitation of Coldest Quarter	\\
+	    \bottomrule
+   \end{tabular}
+   \label{table:TableS_ClimVars}
+\end{table}
+
 % ANOVA table for LER (need to label tables)

### 2015-04-29 ms/ms.tex
Worked on Resuls and Discussion.

--- 2015-04-27 ms/ms.tex
@@
-\documentclass[11pt, oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
+\documentclass[11pt, oneside]{article}
 
@@
 %
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
 \begin{document}
@@
 [maybe this goes in Results?]
-We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components. Spec
+We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components. 
 \begin{knitrout}
@@
   \hlstd{focClim}\hlopt{$}\hlstd{PC1} \hlkwb{<-} \hlstd{traitPC}\hlopt{$}\hlstd{x[,} \hlstr{"PC1"}\hlstd{]}
+  \hlstd{focClimSA}\hlopt{$}\hlstd{PC1} \hlkwb{<-} \hlstd{traitPC}\hlopt{$}\hlstd{x[,} \hlstr{"PC1"}\hlstd{]}
 
@@
 \end{knitrout}
-
-\subsection*{Selective agents and environmental correlates}
-
-As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
-temperature or precipitation) that covary with latitude.
-We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
-cardinalis}.
-We did this by looking for overlap between climatic variables that best predict latitude of \textit{M.
-cadinalis} populations and climatic variables that best predict trait variation across our 16 focal populations.
-We used a representative set of 178 recent (since 2000) known \textit{M.
-cardinalis} occurences.
-These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
-Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included X environmental variables from ClimateWNA and 17?
-bioclimatic variables (Table~\ref{table:TableS_}).
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
-Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
-
-Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
-We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation.
-We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.
-
-To identify climatic variables that were closely associated with latitude in \textit{M.
-cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
-In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.
-
-Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions.
-If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.
-
-Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
-, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
-Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
-To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
-
-% Methods for spatial-average selection
-For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{library}\hlstd{(VSURF)}
-\end{alltt}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: randomForest\\\#\# randomForest 4.6-10\\\#\# Type rfNews() to see new features/changes/bug fixes.\\\#\# Loading required package: rpart\\\#\# Loading required package: doParallel\\\#\# Loading required package: foreach\\\#\# Loading required package: iterators\\\#\# Loading required package: parallel}}\begin{alltt}
-  \hlcom{# Load VSURF results of climatic variables associated with latitude at cardinalis occurence points}
-  \hlcom{# Note that occurences have been randomly thinned to account for variation in sampling effort}
-  \hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/clim2lat_fit_t1.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))} \hlcom{# fit_t1}
-  \hlstd{vsurfLat} \hlkwb{<-} \hlstd{fit_t1}
-
-        \hlcom{# Load climVars}
-        \hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/ExploratoryCimateVariables.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-
-  \hlcom{# Figure of variable importance for latitude}
-  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_clim2lat_imp.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{4}\hlstd{,} \hlnum{5}\hlstd{)}
-  \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{6}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
-  \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlkwd{ceiling}\hlstd{(vsurfLat}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x)[}\hlnum{1}\hlstd{]),}
-                \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)),} \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,}
-                \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
-        \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Climate-Latitude\textbackslash{}nImportance"}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
-  \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlnum{1}\hlopt{:}\hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred))}
-  \hlstd{\{}
-     \hlkwd{lines}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{1}\hlstd{,} \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(i, i),} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
-  \hlstd{\}}
-  \hlkwd{points}\hlstd{(vsurfLat}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)],}
-    \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)}\hlopt{:}\hlnum{1}\hlstd{,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlcom{# mtext(climVars[vsurfLat$varselect.pred], side = 2, }
-    \hlcom{# at = length(vsurfLat$varselect.pred):1, las = 1, line = 1)}
-        \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-        \hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\begin{alltt}
-  \hlcom{# Variable selection using random forests (VSURF)}
-  \hlstd{linmod.PC1} \hlkwb{<-} \hlkwd{as.formula}\hlstd{(}\hlkwd{paste}\hlstd{(}\hlstr{"PC1 ~ "}\hlstd{,} \hlkwd{paste}\hlstd{(climVars,} \hlkwc{collapse} \hlstd{=} \hlstr{" + "}\hlstd{),} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-  \hlkwd{set.seed}\hlstd{(}\hlnum{69647}\hlstd{)}
-  \hlstd{vsurfPC1} \hlkwb{<-} \hlkwd{VSURF.parallel}\hlstd{(linmod.PC1,} \hlkwc{data} \hlstd{= focClim[,} \hlkwd{c}\hlstd{(}\hlstr{"PC1"}\hlstd{, climVars)],}
-    \hlkwc{ncores} \hlstd{=} \hlnum{3}\hlstd{)}
-\end{alltt}
-
-{\ttfamily\noindent\color{warningcolor}{\#\# Warning in VSURF.parallel.formula(linmod.PC1, data = focClim[, c("{}PC1"{}, : VSURF with a formula-type call outputs selected variables\\\#\# which are indices of the input matrix based on the formula:\\\#\# you may reorder these to get indices of the original data}}\begin{alltt}
-  \hlcom{# Plot most important climatic predictor versus PC1}
-  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_bio16cva_v_Lat.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{5}\hlstd{,} \hlnum{5}\hlstd{)}
-  \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
-  \hlkwd{plot}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0.25}\hlstd{,} \hlnum{0.55}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{axes} \hlstd{= F,}
-    \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
-  \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Precipitation of Wettest Quarter"}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
-  \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"(Coefficient of variation)"}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{)}
-  \hlkwd{title}\hlstd{(}\hlkwc{ylab} \hlstd{=} \hlkwd{substitute}\hlstd{(}\hlkwd{plain}\hlstd{(Principal}\hlopt{~}\hlstd{Component}\hlopt{~}\hlnum{1}\hlstd{)}\hlopt{~}\hlkwd{group}\hlstd{(}\hlstr{"("}\hlstd{, x,} \hlstr{")"}\hlstd{),}
-    \hlkwd{list}\hlstd{(}\hlkwc{x} \hlstd{= pve)))}
-  \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-  \hlkwd{axis}\hlstd{(}\hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{,} \hlnum{2}\hlstd{),} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-  \hlstd{fitBio16cva} \hlkwb{<-} \hlkwd{lm}\hlstd{(PC1} \hlopt{~} \hlstd{bio16_cva,} \hlkwc{data} \hlstd{= focClim)}
-
-  \hlcom{# polygon of confidence intervals}
-  \hlstd{x} \hlkwb{<-} \hlkwd{seq}\hlstd{(}\hlkwd{min}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwd{max}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwc{l} \hlstd{=} \hlnum{1e3}\hlstd{)}
-  \hlstd{u} \hlkwb{<-} \hlkwd{predict}\hlstd{(fitBio16cva,} \hlkwc{new} \hlstd{=} \hlkwd{data.frame}\hlstd{(}\hlkwc{bio16_cva} \hlstd{= x))}
-  \hlstd{s} \hlkwb{<-} \hlkwd{sqrt}\hlstd{(}\hlkwd{sapply}\hlstd{(x,} \hlkwa{function}\hlstd{(}\hlkwc{X}\hlstd{)} \hlkwd{t}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X))} \hlopt{%*%} \hlkwd{vcov}\hlstd{(fitBio16cva)} \hlopt{%*%} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X)))}
-  \hlstd{confInt} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwc{lowerCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.025}\hlstd{, u, s),} \hlkwc{upperCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.975}\hlstd{, u, s))}
-  \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(x,} \hlkwd{rev}\hlstd{(x)),} \hlkwd{c}\hlstd{(confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwd{rev}\hlstd{(confInt}\hlopt{$}\hlstd{upperCI)),}
-    \hlkwc{col} \hlstd{=} \hlstr{"grey80"}\hlstd{,} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{)}
-  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{upperCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{lines}\hlstd{(}\hlkwd{range}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwd{range}\hlstd{(}\hlkwd{predict}\hlstd{(fitBio16cva)),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{points}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{= primShade0,}
-    \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\begin{alltt}
-  \hlcom{# Overlap between latitude predictors and phenotype predictors}
-  \hlstd{imp2} \hlkwb{<-} \hlstd{climVars[vsurfPC1}\hlopt{$}\hlstd{varselect.pred]}
-  \hlstd{nImpPC1} \hlkwb{<-} \hlkwd{length}\hlstd{(imp2)}
-
-  \hlstd{vsurfLat}\hlopt{$}\hlstd{varselect.pred}
-\end{alltt}
-\begin{verbatim}
-##  [1] 45 43 42 20 47 41 46 39 19 13
-\end{verbatim}
-\begin{alltt}
-  \hlstd{imp1} \hlkwb{<-} \hlstd{climVars[vsurfLat}\hlopt{$}\hlstd{varselect.pred]}
-  \hlstd{nImpLat} \hlkwb{<-} \hlkwd{length}\hlstd{(imp1)}
-
-  \hlcom{# Figure of variable importance for traits}
-        \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_clim2pc1_imp.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{4}\hlstd{,} \hlnum{5}\hlstd{)}
-        \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{6}\hlstd{))}
-
-        \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.3}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, nImpLat),}
-                \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
-        \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Climate-Trait PC1\textbackslash{}nImportance"}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlstd{(nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{)))}
-        \hlstd{\{}
-                \hlkwd{lines}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{100}\hlstd{,} \hlnum{100}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(i, i),} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
-        \hlstd{\}}
-        \hlkwd{points}\hlstd{(vsurfPC1}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlstd{nImpPC1], nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{),}
-                \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlcom{# mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, }
-  \hlcom{#  at = nImpLat:(nImpLat - nImpPC1 + 1), }
-        \hlcom{#	font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))}
-        \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.3}\hlstd{,} \hlnum{0.1}\hlstd{))}
-  \hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\begin{table}[ht]
-   \centering
-   \topcaption{Climatic variables}
-   \begin{tabular}{@{} ll @{}}
-      \toprule
-  Climate variable & Abbreviation \\
-      \midrule
-  % climateWNA VARIABLES:
-	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
-	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
-	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
-	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
-	NFFD	& number of frost-free days                           \\
-	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
-	Eref	& Hargreaves reference evaporation (mm) \\
-	CMD	  & Hargreaves climativ moisture deficit (mm) \\
-	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
-	RH    &	mean annual relative humidity \\
-  % BIOCLIM VARIABLES:
-  bio1	&	Annual Mean Temperature	\\
-	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
-	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
-	bio4	&	Temperature Seasonality (standard deviation *100)	\\
-	bio5	&	Max Temperature of Warmest Month	\\
-	bio6	&	Min Temperature of Coldest Month	\\
-	bio7	&	Temperature Annual Range (bio5-bio6)	\\
-	bio8	&	Mean Temperature of Wettest Quarter	\\
-	bio9	&	Mean Temperature of Driest Quarter	\\
-	bio10	&	Mean Temperature of Warmest Quarter	\\
-	bio11	&	Mean Temperature of Coldest Quarter	\\
-	bio12	&	Annual Precipitation	\\
-	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
-	bio16	&	Precipitation of Wettest Quarter	\\
-	bio17	&	Precipitation of Driest Quarter	\\
-	bio18	&	Precipitation of Warmest Quarter	\\
-	bio19	&	Precipitation of Coldest Quarter	\\
-	    \bottomrule
-   \end{tabular}
-\end{table}
-
-%--------------------------------------------------
-% Results
-%--------------------------------------------------
-
-\section*{Results}
-
-\subsection*{Coordinated latitudinal clines in germination, growth, and photosynthesis}
-
-Genetic differences between populations affecting originsmal function and performance point to traits that may be involved in local adaptation.
-Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis} (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured 74.2 %$\%$ of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
-As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Table of Population, Treatment, Population x Treatment interactions for varies traits
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}[ht]
-   \centering
-   \topcaption{Population, treatment effects}
-   \begin{tabular}{@{} llllllll @{}}
-      \toprule
-                                   & Germination & Leaf expansion & Stem elongation & Photosynthesis & Intrin. Photo & Mortality \\
-                                   & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
-      \midrule
-  Population                       &
-    ***  &
-    *** &
-    *** &
-    *** & 
-    *** & 
-    *** & \\
-  Temperature                      &
-    NA                                                          &
-    ***    &
-    *** &
-    *** & 
-     &
-    *** & \\
-  Water                            &
-    NA                                                          &
-    ***   &
-    ***   &
-    *** & 
-     &     
-    *** & \\
-  Pop $\times$ Temp                &
-    NA                                                          &
-     &
-     &
-    * & 
-     &     
-    * & \\
-  Pop $\times$ Water               &
-     NA                                                          &
-    * &
-     &
-     & 
-     &     
-     & \\
-  Temp $\times$ Water              &
-    NA                                                          &
-     &
-    *** &
-     & 
-     &     
-    *** & \\
-  Pop $\times$ Temp $\times$ Water &
-    NA                                                          &
-     &
-     &
-     & 
-     &     
-     & \\
-      \bottomrule
-   \end{tabular}
-   \label{table:Table_AnovaSummary}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of latitude versus PC1 (growth axis)
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster]{CAPTION}
-	\label{fig:Fig_PC1vLat}
-\end{figure}
-
-\subsection*{Remarkably little evidence for variation in plasticity}
-
-Genotype $\times$ environment interactions are also a common signature of local adaptation [CITE].
-For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
-We found remarkably little evidence for this pattern in \textit{M.
-cardinalis}.
-There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
-Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
-Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
-
-\subsection*{Latitude and phenotypic clines are associated with more variable precipitation}
-
-Latitude explains variation
-Presumably selective agent is not latitude, but something related to it
-Exploratory approach to identify putative selective agent to test in future studies
-We used a machine learning approach...identified interannual variation in precipitation (Fig~\ref{fig:Fig_ClimVarImp})
-Obviously, cannot identify alternative variables that were not included in analysis.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of climatic variable importance
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimVarImp.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{CAPTION}
-	\label{fig:Fig_ClimVarImp}
-\end{figure}
-
-%--------------------------------------------------
-% Discussion
-%--------------------------------------------------
-
-\section*{Discussion}
-
-1.
-How does local adaptation occur?
-It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
-For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
-
-2. Why does local adaptation succeed? shallow environmental gradients lead to convex tradeoffs not heavily constrained by gene flow.
-
-3. Selective agent - unpredictability. Explain
-
-4.
-Why does local adaptation fail?
-concave, high gene flow, demography.
-Qualify that our metrics of plasticity are coarse, but at least suggest that GxE is much weaker than G for the environmental gradients evaluated here.
-
-5.
-Future work and broad conclusions.
-Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
-
-% stuff about environmental grain
-How organisms adapt depends on the grain of the environment, ..., and gene flow between populations
-
-Possibilities:
-1. no trait variation: no environmental variation, high gene flow, generalist (convex)
-2. variation in mean:
-3. variation in plasticity: convex?,
-
-our data are most consistent with:
-- concave tradeoffs, fine-grained (coarse+concave would predict alt morphs), high gene flow
-
-%--------------------------------------------------
-% References
-%--------------------------------------------------
-
-\setlength{\bibsep}{6pt}
-\bigskip
-
-\bibliography{refs}
-\bibliographystyle{evolution}
-
-%--------------------------------------------------
-% Supplementary Information
-%--------------------------------------------------
-
-% DIC table for initial size (LLL on 5/12)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{Initial size of seedlings did not vary among Populations, Families, or Treatments.
-We used a censored Gaussian model of initial size at the outset of the experiment (longest leaf length of the first true leaves).
-The model was censored because we could not accurately measure leaves less than 0.25 mm with digital callipers (217 of 702, 30.9\%, were too small).
-We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package \pkg{MCMCglmm} version 2.17 \citep{Hadfield_2010}.
-We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in.
-We step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
-	\begin{center}
-	\begin{tabular}{>{\everypar{\hangindent1cm}{}\raggedright}p{6cm}lc}
-	\toprule
-
-	\input{Tables/Table_InitialSize.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-% ANOVA table for LER (need to label tables)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{CAPTION}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-	\input{Tables/Table_lllGrowthAnova.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-% ANOVA table for SER (need to label tables)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, stem expansion rate]{CAPTION}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-	\input{Tables/Table_heightGrowthAnova.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-% Figures showing latitudinal clines with specific traits
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations germinate sooner.]{CAPTION}
-	\label{fig:Fig_DoG}
-\end{figure}
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
-	\label{fig:Fig_LLL}
-\end{figure}
-
-\begin{figure}
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (stem elongation rate).]{CAPTION}
-	\label{fig:Fig_height}
-\end{figure}
-
-\end{document}
-
-	\subsection{Structure}
-
-	Leaf traits method: \\
-	Day 1: \\
-	- select focal leaf (e.g. second or third leaf); \\
-	- place in H20, dark, cool overnight. \\
-	Day 2: \\
-	- wet weight; \\
-	- scan; \\
-	- place in drying oven. \\
-
-		\begin{enumerate}
-			\item{Stomatal density, guard cell length, ratio from leaf opposite focal leaf (above}
-			\item{Leaf size/shape (from scans)}
-			\item{Leaf thickness/SLA}
-			\item{Colorimetry (from scans, spec)}
-			\item{stem anthocyanin (qualitative)}
-		\end{enumerate}
-
-	OPEN QUESTIONS: \\
-	save wet samples in preservative (e.g. for leaf venation?) \\
-	chlorophyll content? \\
-	branching architecture (anything in plant protocols?)
-
-	\subsection{Function}
-
-		\subsubsection{during experiment}
-
-		\begin{enumerate}
-			\item{Instantaneous photosynthetic rate, $g_\text{s}$, water-use efficiency.}
-				\subitem{3 min per plant per LICOR}
-				\subitem{Do as many plants as possible in 1-2 weeks}
-				\subitem{Before and after drought?}
-			\item{Respiration}
-			\item{Whole plant water-use (gravimetric)}
-				\subitem{}
-			\item{Leaf temperature (subset)}
-			\item{Temperature response curve (subset)}
-				\subitem{see protocol here: http://prometheuswiki.publish.csiro.au/tiki-index.php?page=Temperature+response+of+photosynthesis+using+a+Li-Cor+6400}
-			\item{Light response curve}
-				\subitem{Sampling? Only in cool, wet?}
-				\subitem{At what temperature?}
-
-		\end{enumerate}
-
-		\subsubsection{after experiment}
-
-		\begin{enumerate}
-			\item{CN content}
-			\item{carbon isotope discrimination}
-		\end{enumerate}
-
-	\subsection{Development time and Life history}
-
-	\begin{enumerate}
-		\item{Flowering time: census daily}
-		\item{Starch accumulation: collect (subset) of plants}
-		\item{Rhizome formation: collect on subset of plants harvested for dry mass}
-	\end{enumerate}
-
-	\subsection{Floral (2nd/3rd/etc flower?)}
-
-	\begin{enumerate}
-		\item{Flower size}
-		\item{Herkogamy}
-	\end{enumerate}
-
-\section{Performance}
-
-	Possibilities: \\
-	1. (sub-)weekly census of plant height. $\approx$ 0.5 min per plant, so allow 1 day \\
-
-	$$RGR_\text{height} = \frac{\text{ln} H_2 - \text{ln} H_1}{t_2 - t_1}$$
-
-	2. early development (pre-treatment) leaf-area growth with camera \\
-	3. harvest dry biomass before and after treatments (get whole plant leaf area using scanner at same time to ground truth camera)
-
-	$$RGR_\text{mass} = \frac{\text{ln} M_2 - \text{ln} M_1}{t_2 - t_1}$$
-
-\section{Determining ecological gradients that matter}
-
-Performance tradeoffs
-Isolation by ecology
-
-\section{Tests of divergent selection}
-
-	\subsection{Trait-environment correlation}
-
-	We will account for relatedness by using genetic distance matrix in linear models.
-
-	\subsection{$P_\text{ST}$ -- $F_\text{ST}$}
-
-	Contrast trait classes with Fst distribution
-
-	\subsection{Isolation by ecology}
-
-	Is \textbf{G} better predicted by \textbf{E} than geographic distance? Use Gideon's method?
-
-	\subsection{Performance tradeoffs between treatments}
-
-\section{Ecological strategy, physiological mechanism, and tradeoff}
-
-	\subsection{Drought escape}
-		Pattern: \\
-			Flowering time shorter in drier/S populations \\
-			More between population variation in flowering time more variable than predicted by Fst \\
-			... \\
-
-	\subsection{Drought avoidance}
-		Pattern: \\
-			Maybe this should really be dehydration avoidance? \\
-			Populations from drier/S avoid dehydration by limiting water loss; more variable than predicted by Fst \\
-
-			Mechanism \\
-			closing stomata earlier during soil drying
-
-		Mechanisms: \\
-			lower water-use \\
-
-	\subsection{Drought tolerance}
-		Pattern: \\
-			drier/S populations have greater tolerance to low tissue water content
-
-		Mechanisms: \\
-			greater WUE \\
-			ability to photosynthesize during drought \\
-
-\section{Extension: incipient speciation as a byproduct or divergence?}
-
-Collect pollen to measure viability.
-
-(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
-Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
-Some hypotheses:
-\begin{itemize}
-  \item{H1: Latitude encompasses multivariate climatic dimension.}
-  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
-\end{itemize}
-
-Predictions:
-\begin{itemize}
-  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
-Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
-  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
-  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
-\end{itemize}
-
-\end{document}

### 2015-04-29 ms/ms.Rnw
Worked on Discussion.

@@
 % Toggle these commands to get version for doing word count
-\linenumbers
-%\pagenumbering{gobble}
-%\raggedright
+% \linenumbers
+% \pagenumbering{gobble}
+% \raggedright
 
@@
 
-As with previous studies in \textit{Mimulus} and many other species [CITES, Huey_etal, mosquitoes, what else?], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
+As with previous studies in \textit{Mimulus} and many other species [CITES, \citep{Huey_etal}, mosquitoes, what else?], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
 temperature or precipitation) that covary with latitude.
@@
 From VSURF models, we kept only variables selected for prediction, the most stringent criteria.
-
-For these analyses, we compiled a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
-cardinalis} occurences.
-These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
-Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included %\Sexpr{length(climVars) / 2}%
-cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
-Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.
 
@@
 
+For these analyses, we compiled a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
+cardinalis} occurences.
+These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
+Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
+For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
+In total, we included \Sexpr{length(climVars) / 2} cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
+Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.
 %--------------------------------------------------
@@
       \toprule
-                                   & Germination & Leaf expansion & Stem elongation & Photosynthesis & Intrin. Photo & Mortality \\
+                                   & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
                                    & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
@@
 New frontiers in local adaptation research seek a general explanation for why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
-Understanding why and how populations have adapted to different environments in the past is especially significant now as a scientific basis for predicting how species' ranges will shift under climate change [CITES, Catullo_etal_2015].
+Understanding why and how populations have adapted to different environments in the past is especially significant now as a scientific basis for predicting how species' ranges will shift under climate change [CITES, \citep{Catullo_etal_2015}].
 In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint.
@@
 Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001}.
+Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strong asymmetric, and/or maladaptive gene flow is too high.
 In \textit{M.
 cardinalis} we found substantial genetic variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
-If this variation is adaptive, it could be maintained as long as fitness tradeoffs are not too Th...
-fitness tradeoffs are not too strong, demographic symmetry, and attenuation of gene flow.
-
-2. Why does local adaptation succeed? shallow environmental gradients lead to convex tradeoffs not heavily constrained by gene flow.
-
-Although evolutionary theory ... But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
+If this variation is adaptive, it suggests that the fitness tradeoff between doing well in low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
+That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatial heterogenous selection.
+We also know from previous work that population size does not vary strongly with latitude \cdm{Amy - is this correct?
+JPs pop gen data seemed to suggest little difference in effective pop size across the range.
+Is this also true of census pop size?}.
+Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range.
 
 3. Selective agent - unpredictability. Explain
+Reiterate result
+
+But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
 
@@
 	\end{center}
-	\lable{tables:TableS_InitialSize}
+	\label{tables:TableS_InitialSize}
 \end{table}
@@
 	\selectfont
-	\caption[ANOVA table, stem expansion rate]{CAPTION}
+	\caption[ANOVA table, shoot elongation rate]{CAPTION}
 	\begin{center}
@@
 	\selectfont
-	\caption[Southern populations grow faster (stem elongation rate).]{CAPTION}
+	\caption[Southern populations grow faster (shoot elongation rate).]{CAPTION}
 	\label{fig:Fig_height}

### 2015-04-29 ms/ms.tex
Worked on Discussion.

--- 2015-04-27 ms/ms.tex
@@
 
-% write/analyze
-% proposal for summer
-  % local fitness set concave
-  % range-wide fitness set convex
-  % interaction between shape of fitness set and gene flow?
-% evo-phys proposal
+\setstretch{\stretchy}
 
 \section*{Introduction}
+
+% Toggle these commands to get version for doing word count
+% \linenumbers
+% \pagenumbering{gobble}
+% \raggedright
 
@@
 
-As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
+As with previous studies in \textit{Mimulus} and many other species [CITES, \citep{Huey_etal}, mosquitoes, what else?], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
 temperature or precipitation) that covary with latitude.
@@
 We did this by looking for overlap between climatic variables that best predict latitude of \textit{M.
-cadinalis} populations and climatic variables that best predict trait variation across our 16 focal populations.
-We used a representative set of 178 recent (since 2000) known \textit{M.
-cardinalis} occurences.
-These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
-Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included X environmental variables from ClimateWNA and 17?
-bioclimatic variables (Table~\ref{table:TableS_}).
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation).
-Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.
-
-Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
-We therefore calculated spatially averaged \cdm{maybe spatially-integrated is better term?} climate means and coefficients of variation.
-We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and CVs were weighted by the predicted suitability of each point, using ENM models from \cite{Angert_ENM}.
-
-To identify climatic variables that were closely associated with latitude in \textit{M.
-cardinalis}, we used a "Variable Selection Using Random Forest" (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
-In this algorithm, climatic variables are regressed against latitude, and the most important variables were selected after a three-stage selection process that progressively excludes unimportant variables.
-
-Predictions: looked for climatic variables (means or CVs) that were strong predictors of both population latitudes (from all card pops) and trait variation in 16 focal popultions.
-If spatial average environment is important, then expect greater predictive power and closer association with latitude-predictors than point averages.
-
-Evaluated fit on test set (correlation between predicted versus actual latitude) and on pseudoabsence set.
-, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection.
-Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species.
-To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...
-
-% Methods for spatial-average selection
-For 16 focal populations, I took the average climate within a 10 km buffer, clipped to streams based on the National Hydrological Database \citep{NHD} weighted by climatic suitability determined [multimodel average/ensemble?] from ecological niche modeling \citep{Angert_ENM}.
+cadinalis} occurence records and climatic variables that best predict trait variation across our 16 focal populations.
+For brevity, we refer to these as Climate-Latitude and Climate-Trait variables.
+We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}).
+From VSURF models, we kept only variables selected for prediction, the most stringent criteria.
 
@@
 {\ttfamily\noindent\color{warningcolor}{\#\# Warning in VSURF.parallel.formula(linmod.PC1, data = focClim[, c("{}PC1"{}, : VSURF with a formula-type call outputs selected variables\\\#\# which are indices of the input matrix based on the formula:\\\#\# you may reorder these to get indices of the original data}}\begin{alltt}
+  \hlstd{vsurfPC1_SA} \hlkwb{<-} \hlkwd{VSURF.parallel}\hlstd{(linmod.PC1,} \hlkwc{data} \hlstd{= focClimSA[,} \hlkwd{c}\hlstd{(}\hlstr{"PC1"}\hlstd{, climVars)],}
+    \hlkwc{ncores} \hlstd{=} \hlnum{3}\hlstd{)}
+\end{alltt}
+
+{\ttfamily\noindent\color{warningcolor}{\#\# Warning in VSURF.parallel.formula(linmod.PC1, data = focClimSA[, c("{}PC1"{}, : VSURF with a formula-type call outputs selected variables\\\#\# which are indices of the input matrix based on the formula:\\\#\# you may reorder these to get indices of the original data}}\begin{alltt}
   \hlcom{# Plot most important climatic predictor versus PC1}
@@
   \hlcom{# Overlap between latitude predictors and phenotype predictors}
+\hlcom{#### NEED TO COMPARE RESULTS FROM POINT DATA AND SPATIALLY AVERAGED CLIMATE!}
   \hlstd{imp2} \hlkwb{<-} \hlstd{climVars[vsurfPC1}\hlopt{$}\hlstd{varselect.pred]}
+  \hlcom{# imp2 <- climVars[vsurfPC1_SA$varselect.pred]}
   \hlstd{nImpPC1} \hlkwb{<-} \hlkwd{length}\hlstd{(imp2)}
@@
 \end{knitrout}
+
+For these analyses, we compiled a representative set of 178 recent (since 2000) known \textit{M.
+cardinalis} occurences.
+These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
+Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
+For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
+In total, we included 24 cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
+Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.
+%--------------------------------------------------
+% Results
+%--------------------------------------------------
+
+\section*{Results}
+
+\subsection*{A coordinated latitudinal cline in germination, growth, and photosynthesis}
+
+Genetic differences between populations affecting originsmal function and performance point to traits that may be involved in local adaptation.
+Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M.
+cardinalis} (Table~\ref{table:Table_AnovaSummary}).
+A single principal component captured 74.2 %$\%$ of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
+As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of Population, Treatment, Population x Treatment interactions for varies traits
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}[ht]
+   \centering
+   \topcaption{Population, treatment effects}
+   \begin{tabular}{@{} llllllll @{}}
+      \toprule
+                                   & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
+                                   & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
+      \midrule
+  Population                       &
+    ***  &
+    *** &
+    *** &
+    *** & 
+    *** & 
+    *** & \\
+  Temperature                      &
+    NA                                                          &
+    ***    &
+    *** &
+    *** & 
+     &
+    *** & \\
+  Water                            &
+    NA                                                          &
+    ***   &
+    ***   &
+    *** & 
+     &     
+    *** & \\
+  Pop $\times$ Temp                &
+    NA                                                          &
+     &
+     &
+    * & 
+     &     
+    * & \\
+  Pop $\times$ Water               &
+     NA                                                          &
+    * &
+     &
+     & 
+     &     
+     & \\
+  Temp $\times$ Water              &
+    NA                                                          &
+     &
+    *** &
+     & 
+     &     
+    *** & \\
+  Pop $\times$ Temp $\times$ Water &
+    NA                                                          &
+     &
+     &
+     & 
+     &     
+     & \\
+      \bottomrule
+   \end{tabular}
+   \label{table:Table_AnovaSummary}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of latitude versus PC1 (growth axis)
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude.
+Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods).
+The line and 95\% confidence intervals (grey polygon) were estimated using standard linear regression.}
+	\label{fig:Fig_PC1vLat}
+\end{figure}
+
+\subsection*{Remarkably little evidence for variation in plasticity}
+
+Genotype $\times$ environment interactions are also a common signature of local adaptation [CITE].
+For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
+We found remarkably little evidence for this pattern in \textit{M.
+cardinalis}.
+There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
+Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
+Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
+However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
+
+\subsection*{Climatic variability best explains phenotypic divergence}
+
+Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
+For \textit{M.
+cardinalis}, interannual variation in precipitation over the past 30 years is very closely related to the latitude of recently-recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
+The two most important Climate-Latitude variables were also strong predictors of position along the fast-slow growth axis (Fig.~\ref{fig:Fig_ClimVarImp}B).
+This overlap between Climate-Latitude and Climate-Trait variables suggests that interannual variation in precipitation is an important selective agent in \textit{M.
+cardinalis}.
+Specifically, we hypothesize that more frequent droughts (greater precipitation cv) in Southern populations selects for an `annual-ized' life history, as we detail in the Discussion.
+We must qualify these results because our analysis obviously cannot rule out that alternative variables not included in the analysis may be more important.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of climatic variable importance
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimVarImp.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{Interannual variation in precipitation is closely correlated with latitude and trait variation.
+A.
+Using Random Forest regression, we identified 10 climatic variables significantly (high importance) associated with latitude of \textit{M.
+cardinalis} occurences.
+B.
+The two most important Climate-Latitude variables were also the two most important Climate-Trait variables.
+Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
+Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp) or Precipitation (Prec); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Dry Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
+	\label{fig:Fig_ClimVarImp}
+\end{figure}
+
+%--------------------------------------------------
+% Discussion
+%--------------------------------------------------
+
+\section*{Discussion}
+
+The observation that populations are locally adapted to their environment is old and well-documented in many species.
+New frontiers in local adaptation research seek a general explanation for why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
+Understanding why and how populations have adapted to different environments in the past is especially significant now as a scientific basis for predicting how species' ranges will shift under climate change [CITES, \citep{Catullo_etal_2015}].
+In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint.
+In particular, our data suggest that photosynthesis and intrinsic growth rate vary among populations, whereas the fundamental abiotic niche is more conserved, and that climatic variation may be a more important selective agent than the average.
+Next, we tie these results into the broader threads of evolutionary theory in order to figure out what factors best explain our data.
+
+As long as environments vary and there exists genetic variation in fitness, why shouldn't local adapted genotypes evolve?
+Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001}.
+In \textit{M.
+cardinalis} we found substantial genetic variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
+If this variation is adaptive, it could be maintained as long as fitness tradeoffs are not too Th...
+fitness tradeoffs are not too strong, demographic symmetry, and attenuation of gene flow.
+
+2. Why does local adaptation succeed? shallow environmental gradients lead to convex tradeoffs not heavily constrained by gene flow.
+
+Although evolutionary theory ... But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
+
+3. Selective agent - unpredictability. Explain
+
+(also 4.) Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years, could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
+We found no evidence for this; all populations responded to drought and temperature treatments similarly (Table~\ref{table:Table_AnovaSummary}).
+However, evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
+Riparian habitats are highly heterogeneous at small spatial scales.
+Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought if there is little direct precipitation during the growing season, as in Meditteranean climates of Western North America.
+But alleles that confer greater drought tolerance may be quite costly in well-watered soils, leading to strong fitness tradeoffs.
+Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
+Demography and gene flow may reinforce niche conservatism.
+A new mutant with increased drought tolerance that can survive at the margin of a population.
+But if .
+Finally, gene flow, which is generally high among \textit{M.
+cardinalis} populations within the same ecoregion \citep{Paul_etal_2015}, will thwart local adaptation and reinforce specialization.
+Spatial grain of the environment, demographic asymmetry, and gene flow conspire to thwart local adaptation via altered fundamental niche.
+4.
+Why does local adaptation fail?
+concave, high gene flow, demography.
+Qualify that our metrics of plasticity are coarse, but at least suggest that GxE is much weaker than G for the environmental gradients evaluated here.
+
+highly connected by gene flow, outcrossing (self-compatible but strongly herkogamous)
+
+?
+It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
+For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
+
+5.
+Future work and broad conclusions.
+Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
+
+% stuff about environmental grain
+How organisms adapt depends on the grain of the environment, ..., and gene flow between populations
+
+Possibilities:
+1. no trait variation: no environmental variation, high gene flow, generalist (convex)
+2. variation in mean:
+3. variation in plasticity: convex?,
+
+our data are most consistent with:
+- concave tradeoffs, fine-grained (coarse+concave would predict alt morphs), high gene flow
+
+%--------------------------------------------------
+% Acknowledgements
+%--------------------------------------------------
+
+\section*{Acknowledgements}
+Erin Warkman and Lisa Lin helped collect data.
+
+%--------------------------------------------------
+% References
+%--------------------------------------------------
+
+\setlength{\bibsep}{6pt}
+\bigskip
+
+\bibliography{refs}
+\bibliographystyle{evolution}
+
+%--------------------------------------------------
+% Supplementary Information
+%--------------------------------------------------
+
+% DIC table for initial size (LLL on 5/12)
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, leaf expansion rate]{Initial size of seedlings did not vary among Populations, Families, or Treatments.
+We used a censored Gaussian model of initial size at the outset of the experiment (longest leaf length of the first true leaves).
+The model was censored because we could not accurately measure leaves less than 0.25 mm with digital callipers (217 of 702, 30.9\%, were too small).
+We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package \pkg{MCMCglmm} version 2.17 \citep{Hadfield_2010}.
+We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in.
+We step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
+	\begin{center}
+	\begin{tabular}{>{\everypar{\hangindent1cm}{}\raggedright}p{6cm}lc}
+	\toprule
+
+	\input{Tables/Table_InitialSize.txt}
+
+	\bottomrule
+	\end{tabular}
+	\end{center}
+	\label{tables:TableS_InitialSize}
+\end{table}
 
@@
    \end{tabular}
-\end{table}
-
-%--------------------------------------------------
-% Results
-%--------------------------------------------------
-
-\section*{Results}
-
-\subsection*{Coordinated latitudinal clines in germination, growth, and photosynthesis}
-
-Genetic differences between populations affecting originsmal function and performance point to traits that may be involved in local adaptation.
-Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis} (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured 74.2 %$\%$ of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
-As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Table of Population, Treatment, Population x Treatment interactions for varies traits
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}[ht]
-   \centering
-   \topcaption{Population, treatment effects}
-   \begin{tabular}{@{} llllllll @{}}
-      \toprule
-                                   & Germination & Leaf expansion & Stem elongation & Photosynthesis & Intrin. Photo & Mortality \\
-                                   & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
-      \midrule
-  Population                       &
-    ***  &
-    *** &
-    *** &
-    *** & 
-    *** & 
-    *** & \\
-  Temperature                      &
-    NA                                                          &
-    ***    &
-    *** &
-    *** & 
-     &
-    *** & \\
-  Water                            &
-    NA                                                          &
-    ***   &
-    ***   &
-    *** & 
-     &     
-    *** & \\
-  Pop $\times$ Temp                &
-    NA                                                          &
-     &
-     &
-    * & 
-     &     
-    * & \\
-  Pop $\times$ Water               &
-     NA                                                          &
-    * &
-     &
-     & 
-     &     
-     & \\
-  Temp $\times$ Water              &
-    NA                                                          &
-     &
-    *** &
-     & 
-     &     
-    *** & \\
-  Pop $\times$ Temp $\times$ Water &
-    NA                                                          &
-     &
-     &
-     & 
-     &     
-     & \\
-      \bottomrule
-   \end{tabular}
-   \label{table:Table_AnovaSummary}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of latitude versus PC1 (growth axis)
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster]{CAPTION}
-	\label{fig:Fig_PC1vLat}
-\end{figure}
-
-\subsection*{Remarkably little evidence for variation in plasticity}
-
-Genotype $\times$ environment interactions are also a common signature of local adaptation [CITE].
-For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
-We found remarkably little evidence for this pattern in \textit{M.
-cardinalis}.
-There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
-Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
-Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
-
-\subsection*{Latitude and phenotypic clines are associated with more variable precipitation}
-
-Latitude explains variation
-Presumably selective agent is not latitude, but something related to it
-Exploratory approach to identify putative selective agent to test in future studies
-We used a machine learning approach...identified interannual variation in precipitation (Fig~\ref{fig:Fig_ClimVarImp})
-Obviously, cannot identify alternative variables that were not included in analysis.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of climatic variable importance
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimVarImp.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{CAPTION}
-	\label{fig:Fig_ClimVarImp}
-\end{figure}
-
-%--------------------------------------------------
-% Discussion
-%--------------------------------------------------
-
-\section*{Discussion}
-
-1.
-How does local adaptation occur?
-It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
-For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
-
-2. Why does local adaptation succeed? shallow environmental gradients lead to convex tradeoffs not heavily constrained by gene flow.
-
-3. Selective agent - unpredictability. Explain
-
-4.
-Why does local adaptation fail?
-concave, high gene flow, demography.
-Qualify that our metrics of plasticity are coarse, but at least suggest that GxE is much weaker than G for the environmental gradients evaluated here.
-
-5.
-Future work and broad conclusions.
-Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
-
-% stuff about environmental grain
-How organisms adapt depends on the grain of the environment, ..., and gene flow between populations
-
-Possibilities:
-1. no trait variation: no environmental variation, high gene flow, generalist (convex)
-2. variation in mean:
-3. variation in plasticity: convex?,
-
-our data are most consistent with:
-- concave tradeoffs, fine-grained (coarse+concave would predict alt morphs), high gene flow
-
-%--------------------------------------------------
-% References
-%--------------------------------------------------
-
-\setlength{\bibsep}{6pt}
-\bigskip
-
-\bibliography{refs}
-\bibliographystyle{evolution}
-
-%--------------------------------------------------
-% Supplementary Information
-%--------------------------------------------------
-
-% DIC table for initial size (LLL on 5/12)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{Initial size of seedlings did not vary among Populations, Families, or Treatments.
-We used a censored Gaussian model of initial size at the outset of the experiment (longest leaf length of the first true leaves).
-The model was censored because we could not accurately measure leaves less than 0.25 mm with digital callipers (217 of 702, 30.9\%, were too small).
-We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package \pkg{MCMCglmm} version 2.17 \citep{Hadfield_2010}.
-We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in.
-We step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
-	\begin{center}
-	\begin{tabular}{>{\everypar{\hangindent1cm}{}\raggedright}p{6cm}lc}
-	\toprule
-
-	\input{Tables/Table_InitialSize.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
+   \label{table:TableS_ClimVars}
 \end{table}
@@
 	\selectfont
-	\caption[Southern populations grow faster (stem elongation rate).]{CAPTION}
+	\caption[Southern populations grow faster (shoot elongation rate).]{CAPTION}
 	\label{fig:Fig_height}
 \end{figure}
+
+\subsection*{Spatially-averaged climate}
+
+% Methods for spatial-average selection
+Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
+We therefore calculated spatially averaged climate means and coefficients of variation.
+We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
+Since \textit{M.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
 

### 2015-04-29 ms/ms.Rnw
Worked on Discussion.

@@
 
+% Maybe focus in on tradeoffs as sine qua non of specialization vs generalization, then mention that demography and/or gene flow could reinforce?
+
 \section*{Discussion}
@@
 In particular, our data suggest that photosynthesis and intrinsic growth rate vary among populations, whereas the fundamental abiotic niche is more conserved, and that climatic variation may be a more important selective agent than the average.
-Next, we tie these results into the broader threads of evolutionary theory in order to figure out what factors best explain our data.
+Next, we tie these results into the broader threads of evolutionary theory in order to figure out what factors explain the trait variation we see and that we do not.
 
@@
 
-3. Selective agent - unpredictability. Explain
-Reiterate result
-
-But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
-
-(also 4.) Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years, could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
+Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years (see next section), could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
 We found no evidence for this; all populations responded to drought and temperature treatments similarly (Table~\ref{table:Table_AnovaSummary}).
-However, evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
-Riparian habitats are highly heterogeneous at small spatial scales.
-Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought if there is little direct precipitation during the growing season, as in Meditteranean climates of Western North America.
-But alleles that confer greater drought tolerance may be quite costly in well-watered soils, leading to strong fitness tradeoffs.
+We hypothesize that evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
+Riparian habitats where \textit{M.
+cardinalis} live are highly heterogeneous at small spatial scales.
+Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Meditteranean climates of Western North America.
+But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and vice versa, leading to strong fitness tradeoffs.
 Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
 Demography and gene flow may reinforce niche conservatism.
-A new mutant with increased drought tolerance that can survive at the margin of a population.
-But if .
+A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher resource environments.
 Finally, gene flow, which is generally high among \textit{M.
 cardinalis} populations within the same ecoregion \citep{Paul_etal_2015}, will thwart local adaptation and reinforce specialization.
-Spatial grain of the environment, demographic asymmetry, and gene flow conspire to thwart local adaptation via altered fundamental niche.
-4.
-Why does local adaptation fail?
-concave, high gene flow, demography.
-Qualify that our metrics of plasticity are coarse, but at least suggest that GxE is much weaker than G for the environmental gradients evaluated here.
-
-highly connected by gene flow, outcrossing (self-compatible but strongly herkogamous)
-
-?
-It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
-For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
+Thus, the spatial grain of the environment, demographic asymmetry, and gene flow may conspire to constrain local adaptation via altered fundamental niche.
+
+4. Selective agent - unpredictability. Explain
+Reiterate result
+other supporting evidence: 
+- ENM models are better for habitat suitability for S pops when using data from <5 years; better for Northern populations using 30 yr (M. Bayly \& A. Angert, unpub. data)
+- greater variance in recruit size in S pops
+-
+
+There is a lot of interest in understanding how organisms will respond to changes in climatic variation, not just changes in the average climate.
+Our data indeed suggest that variation may be more important than the mean.
+
+But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
 
@@
 Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
+Next, we are testing whether phenotypic variation and constraint can be predicted by the shape of tradeoffs, as predicted by evolutionary theory,
 

### 2015-04-29 ms/ms.Rnw
Worked on figures

@@
       \midrule
-	HAU & South Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Hauser Creek"], 3)}	& 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Hauser Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Hauser Creek"]}   \\
-	CTC	& South Margin		& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Cottonwood Creek"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Cottonwood Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Cottonwood Creek"]}   \\
-	CUR	& South Margin		& yes	& yes & \Sexpr{round(focClim$Latitude[focClim$PopID == "Cuyamaca Rancho"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Cuyamaca Rancho"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Cuyamaca Rancho"]}   \\
-	GRP & South Margin		& no	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Grade Road Palomar"], 3)} &
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Grade Road Palomar"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Grade Road Palomar"]}   \\
-	WWC &	Transverse		  & yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Whitewater Canyon"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Whitewater Canyon"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Whitewater Canyon"]}   \\
-	MIL	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Mill Creek"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Mill Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Mill Creek"]}   \\
-	WFM	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "West Fork Mojave River"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "West Fork Mojave River"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "West Fork Mojave River"]}   \\
-	NMT	& South Sierras		& yes	& yes?& \Sexpr{round(focClim$Latitude[focClim$PopID == "North Fork Middle Tule"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "North Fork Middle Tule"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "North Fork Middle Tule"]}   \\
-	PRD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Paradise Creek"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Paradise Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Paradise Creek"]}   \\
-	RWD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Redwood Creek"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Redwood Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Redwood Creek"]}   \\
-	WNA	& Central Sierras	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Wawona"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Wawona"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Wawona"]}   \\
-	RBW	& Central Sierras	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Rainbow Pool"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Rainbow Pool"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Rainbow Pool"]}   \\
-	MYU	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Middle Yuba"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Middle Yuba"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Middle Yuba"]}   \\
-	LIJ	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Little Jameson"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Little Jameson"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Little Jameson"]}   \\
-	DPC	& North Coast	  	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Deep Creek"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Deep Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Deep Creek"]}   \\
-	RCC	& North Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "Rock Creek"], 3)} & 
-    \Sexpr{round(focClim$Longitude[focClim$PopID == "Rock Creek"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "Rock Creek"]}   \\
+	HAU & South Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "HAU"], 3)}	& 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "HAU"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "HAU"]}   \\
+	CTC	& South Margin		& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "CTC"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "CTC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "CTC"]}   \\
+	CUR	& South Margin		& yes	& yes & \Sexpr{round(focClim$Latitude[focClim$PopID == "CUR"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "CUR"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "CUR"]}   \\
+	GRP & South Margin		& no	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "GRP"], 3)} &
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "GRP"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "GRP"]}   \\
+	WWC &	Transverse		  & yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "WWC"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "WWC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WWC"]}   \\
+	MIL	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MIL"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "MIL"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "MIL"]}   \\
+	WFM	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "WFM"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "WFM"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WFM"]}   \\
+	NMT	& South Sierras		& yes	& yes?& \Sexpr{round(focClim$Latitude[focClim$PopID == "NMT"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "NMT"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "NMT"]}   \\
+	PRD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "PRD"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "PRD"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "PRD"]}   \\
+	RWD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RWD"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "RWD"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RWD"]}   \\
+	WNA	& Central Sierras	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "WNA"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "WNA"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WNA"]}   \\
+	RBW	& Central Sierras	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RBW"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "RBW"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RBW"]}   \\
+	MYU	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MYU"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "MYU"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "MYU"]}   \\
+	LIJ	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "LIJ"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "LIJ"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "LIJ"]}   \\
+	DPC	& North Coast	  	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "DPC"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "DPC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "DPC"]}   \\
+	RCC	& North Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RCC"], 3)} & 
+    \Sexpr{round(focClim$Longitude[focClim$PopID == "RCC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RCC"]}   \\
 	\bottomrule
@@
 						 expression(30*degree), expression(40*degree)))
+
+# Watering axis
+axis(4, at = c(0, 40), lwd = 0, lwd.ticks = 1, las = 1, labels = c(1, 12))
+mtext(expression(paste("Watering Freq (no. Day", phantom()^-1, ")")), 4, 
+	cex = 1.5, line = 3, las = 3)
+
 # Axis of time on top
@@
 	border = "white", col = "grey"))
-with(exptlDes, text(min((which(SERmeas)) + max(which(SERmeas))) / 2, 1, 
+with(exptlDes, text(min((which(SERmeas)) + max(which(SERmeas))) / 2 - 5, 1, 
 	labels = "SER", col = "white", font = 3, cex = 0.9))
@@
 #	labels = "Photo", col = "white", font = 3, cex = 0.9))
+
+# Polygon of Drought treatment
+# Rescale watering frequency to vary from 0 to 40
+exptlDes$WaterFreq_dry <- exptlDes$WaterFreq_dry * (40 / 11) - 40 / 11
+
+# with(exptlDes, points(DayN, WaterFreq_dry, type = "l", lwd = 12, 
+	# col = alpha("white", 0.5)))
+with(exptlDes, points(DayN, WaterFreq_dry, type = "l", lwd = 3, col = sec2shade4))
 

### 2015-04-29 ms/ms.tex
Worked on figures

@@
 Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001}.
+Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strong asymmetric, and/or maladaptive gene flow is too high.
 In \textit{M.
 cardinalis} we found substantial genetic variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
-If this variation is adaptive, it could be maintained as long as fitness tradeoffs are not too Th...
-fitness tradeoffs are not too strong, demographic symmetry, and attenuation of gene flow.
-
-2. Why does local adaptation succeed? shallow environmental gradients lead to convex tradeoffs not heavily constrained by gene flow.
-
-Although evolutionary theory ... But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
-
-3. Selective agent - unpredictability. Explain
-
-(also 4.) Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years, could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
+If this variation is adaptive, it suggests that the fitness tradeoff between doing well in low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
+That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatial heterogenous selection.
+We also know from previous work that population size does not vary strongly with latitude \cdm{Amy - is this correct?
+JPs pop gen data seemed to suggest little difference in effective pop size across the range.
+Is this also true of census pop size?}.
+Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range.
+
+Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years (see next section), could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
 We found no evidence for this; all populations responded to drought and temperature treatments similarly (Table~\ref{table:Table_AnovaSummary}).
-However, evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
-Riparian habitats are highly heterogeneous at small spatial scales.
-Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought if there is little direct precipitation during the growing season, as in Meditteranean climates of Western North America.
-But alleles that confer greater drought tolerance may be quite costly in well-watered soils, leading to strong fitness tradeoffs.
+We hypothesize that evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
+Riparian habitats where \textit{M.
+cardinalis} live are highly heterogeneous at small spatial scales.
+Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Meditteranean climates of Western North America.
+But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and vice versa, leading to strong fitness tradeoffs.
 Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
 Demography and gene flow may reinforce niche conservatism.
-A new mutant with increased drought tolerance that can survive at the margin of a population.
-But if .
+A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher resource environments.
 Finally, gene flow, which is generally high among \textit{M.
 cardinalis} populations within the same ecoregion \citep{Paul_etal_2015}, will thwart local adaptation and reinforce specialization.
-Spatial grain of the environment, demographic asymmetry, and gene flow conspire to thwart local adaptation via altered fundamental niche.
-4.
-Why does local adaptation fail?
-concave, high gene flow, demography.
-Qualify that our metrics of plasticity are coarse, but at least suggest that GxE is much weaker than G for the environmental gradients evaluated here.
-
-highly connected by gene flow, outcrossing (self-compatible but strongly herkogamous)
-
-?
-It's complex (gene flow, shape of tradeoffs, demography) and although there is abundant evidence for local adaptaiton, we do not a general sense for when local adaptation succeeds or fails.
-For this, we need a system in which there are axes of variation and constraint, traits with known physiological function?, and selective agents
+Thus, the spatial grain of the environment, demographic asymmetry, and gene flow may conspire to constrain local adaptation via altered fundamental niche.
+
+4. Selective agent - unpredictability. Explain
+Reiterate result
+other supporting evidence: 
+- ENM models are better for habitat suitability for S pops when using data from <5 years; better for Northern populations using 30 yr (M. Bayly \& A. Angert, unpub. data)
+- greater variance in recruit size in S pops
+-
+
+There is a lot of interest in understanding how organisms will respond to changes in climatic variation, not just changes in the average climate.
+Our data indeed suggest that variation may be more important than the mean.
+
+But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
 

### 2015-04-30 ms/ms.Rnw
Finished rough draft

@@
 
-<<Front Matter>>=
+<<Front Matter, echo=FALSE, results='hide'>>=
 
@@
 
-\title{(working title) Local Physiological Adaptation in \textit{Mimulus cardinalis}}
-\author{Christopher D. Muir and Amy L. Angert}
-%\date{}							% Activate to display a given date or no date
+\title{Grow with the flow: faster growth is associated with more variable precipitation in a perennial herb}
+\author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
+\date{}
 
@@
 \maketitle
+
+$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada
+
+$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca
+
+Original Article
+
+Key words: local adaptation, photosynthesis, growth rate, climatic variance, \textit{Mimulus}
+
+Word counts: TBA
+
+3 figures; 3 tables; ? supporting files
+
+Data will be archived on Dryad upon acceptance.
+
+\clearpage
+
 \listoffigures
@@
 Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
-Local adaptation within species most often involves genetic differences in ecologically relevant traits caused by selection to different environments, leading to predictable trait-environment correlations that persist in a common garden.
-Although local adaptation is well-supported by data from reciprocal transplant experiments, trait-environment associations, and ecological genetics/genomics, predicting which traits respond to selection exceeds our understanding.
+Local adaptation within species most often involves genetic differences in ecologically relevant traits caused by selection to different environments, leading to predictable trait-environment correlations that persist in a common environment.
+Although local adaptation is well-supported by data from reciprocal transplant experiments, trait-environment associations, and ecological genetics/genomics, predicting which traits respond to selection is beyond our current understanding.
 In particular, the role of physiological traits is poorly understood, even though physiology explains adaptation to the abiotic environment, which we need to understand as a baseline for anticipating how organisms will respond to climate change.
-A related problem is identifying the exact selective agents, such as abiotic factors like soil water availability or biotic interactions [etc.]...
-Here, we simultaneously investigate which traits and causal selective agents explain range-wide patterns of local physiological adaptation to climate in \textit{Mimulus}, a model genus for local adaptation studies.
+A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures.
+Here, we  investigate which traits and causal selective agents explain range-wide patterns of local physiological adaptation to climate in \textit{Mimulus}, a model genus for local adaptation studies.
 
@@
 However, it is more challenging for cases in which traits vary with less specific factors like mean annual precipitation or latitude.
-For example, body size clines (larger size at higher latitude) have evolved repeatedly on three contients in \textit{Drosophila subobscura}, strongly indicating selection.
+For example, body size clines (larger size at higher latitude) have evolved repeatedly on three contients in \textit{Drosophila subobscura}, strongly indicating selection \citep{Huey_etal_2000}.
 But which one or more of the many climatic factors that covary with latitude actually controls selection on body size?
-Selection might be even more complex for species that live over multiple seasons, meaning that environmental variation rather than the average environment may be a selective agent.
+Selection might be even more complex for perennial species that live over multiple seasons, meaning that environmental variation rather than the average environment may be a selective agent.
 
@@
 Linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions.
-Many classic and contemporary studies of local adaptation have been conducted using species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources.
-...
-Conspicuous absence of phys mechanism
-
-% Final paragraph
-1. `intrinsic' physiological variation
-2. variation in plasticity
-3. what ecological factors explain divergence (i.e. putative selective agents underlying local adaptation)
-
-Conclude that observations are best explained by stabilizing selection on the fundamental niche, but spatially varying divergent selection on coordinated life history and physiological traits.
+Many classic and contemporary studies of local adaptation have been conducted using species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources [willis, lowry?].
+Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008}.
+
+Two common signatures of local adaptation are nonrandom genetic differences in ecologically important traits and variation in plasticity (genotype by environment interaction).
+In this study, we found strong genetically-based differences in photosynthesis and growth, but no evidence for variation in plasticity in response to temperature and drought treatments.
+Interannual variation in precipitaiton is associated intrinsic variation in these traits, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+cardinalis}.
 
@@
 
-We used 16 populations from throughout the range of \textit{M. cardinalis} (Table 1). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
-
-<<focClim, echo = FALSE>>=
+We used 16 populations from throughout the range of \textit{M. cardinalis} (Table~\ref{table:Table_FocalPops}). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
+
+<<focClim, echo = FALSE, results='hide', warning=FALSE>>=
 # Read in climate data for 16 focal populations to use in making Table 1
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of focal population
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{table}[ht]
    \centering
-   \topcaption{16 Focal populations}
+   \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
    \begin{tabular}{@{} lllllll @{}}
@@
 	\end{tabular}
+	\label{table:Table_FocalPops}
 \end{table}
@@
 
-<<ExperimentalDesign>>=
+<<ExperimentalDesign, echo=FALSE, results='hide', warning=FALSE>>=
 library(scales) # for color transparency
@@
 
-# Polygon of Drought treatment
+# Line of Drought treatment
 # Rescale watering frequency to vary from 0 to 40
@@
 	# col = alpha("white", 0.5)))
-with(exptlDes, points(DayN, WaterFreq_dry, type = "l", lwd = 3, col = sec2shade4))
+with(exptlDes, points(DayN, WaterFreq_dry, type = "l", lwd = 3, col = "black"))
 
@@
 # Legend
-pdf(paste(pathFig, "/ExperimentalDesignLegend.pdf", sep = ""), 1.5, 2)
+pdf(paste(pathFig, "/ExperimentalDesignLegend.pdf", sep = ""), 2, 2)
 par(mai = rep(0, 4))
-plot(0, 0, type = "n", xlim = c(0, 1.5), ylim = c(0, 1), axes = F)
-
+plot(0, 0, type = "n", xlim = c(-0.5, 1.5), ylim = c(0, 1), axes = F)
+
+# Temperature
 polygon(c(0, 0.5, 1, 1, 0.5, 0), c(0.8, 1, 1, 0.55, 0.55, 0.35), border = NA, 
@@
 points(c(0, 0.5, 1), c(0.375, 0.225, 0.225), lwd = 2, type = "l", col = sec1shade3)
+
+# Drought
+lines(c(0, 1), c(1, 0), lwd = 2)
+text(-0.25, 0.95, labels = "Water\nFreq")
 text(0.5, 0.9, labels = "Hot", col = primShade3)
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure summarizing experimental design
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{figure}[h!]
@@
 	\selectfont
-	\caption[Experimental Design]{CAPTION}
+	\caption[Experimental Design]{Overview of experimental treatments and timing of key trait measurements.
+All plants germinated within 21 days of sowing.
+At that time, we began temperature treatments (left axis), simulating a typical June-August weather pattern at Hot (red) and Cold (blue) sites.
+The bold lines track the average daily temperatures.
+Within each day, there was a maximum daytime temperture (top of translucent polygons) and minimum nighttime temperature (bottom of translucent polygons).
+The drought treatment commenced later by ramping down the frequency of bottom-watering episodes (black line; right axis).
+Grey boxes on the bottom of the plot outline the period of key measurements described in the Methods.}
 	\label{fig:Fig_ExptlDes}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of key traits
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{table}[ht]
    \centering
-   \topcaption{Traits}
+   \topcaption{Key traits measured in this study.}
    \begin{tabular}{@{} ll @{}}
@@
 
-<<Germination rate>>=
+<<Germination rate, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
 
@@
 
-<<Leaf expansion rate>>=
+<<Leaf expansion rate, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
 
@@
 
-<<Shoot elongation rate>>=
+<<Shoot elongation rate, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
 
@@
 
-<<Photosynthesis>>=
+<<Photosynthesis, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
 
@@
 
-<<Mortality>>=
+<<Mortality, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
 
@@
 \subsection*{Principal components of germination, growth, and photosynthesis}
-[maybe this goes in Results?]
 We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components.
-<<Principal Components>>=
+The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast and slow growing populations.
+
+<<Principal Components, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
 
@@
 
-<<thinnedOccClimate>>=
+<<thinnedOccClimate, echo=FALSE, results='hide', warning=FALSE>>=
 # R data.frame of 178 M. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data
@@
 
-<<VSURF>>=
+<<VSURF, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
 
@@
 Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.
+
 %--------------------------------------------------
@@
    \centering
-   \topcaption{Population, treatment effects}
+   \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects.
+We used different statistical modeling for the diverse traits assayed -- glm: generalized linear model using R \citep{R}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
+Note that temperature and water treatments were imposed after germination, hence are not application to this trait.
+Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}
+   \resizebox{6.5in}{!}{
    \begin{tabular}{@{} llllllll @{}}
       \toprule
-                                   & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
-                                   & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
+    {\raggedleft Trait}                          & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
+    Statistical model              & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
       \midrule
@@
       \bottomrule
-   \end{tabular}
+   \end{tabular} }
    \label{table:Table_AnovaSummary}
@@
 
-4. Selective agent - unpredictability. Explain
-Reiterate result
-other supporting evidence: 
-- ENM models are better for habitat suitability for S pops when using data from <5 years; better for Northern populations using 30 yr (M. Bayly \& A. Angert, unpub. data)
-- greater variance in recruit size in S pops
--
-
+Based on the available data, interannual variation in annual or winter precipitation (these are closely correlated in Mediterranean climates) may be the selective agent driving variation in growth and photosynthesis.
+Variation in precipitation was best predicted latitude of recent \textit{M.
+cardinalis} occurences and trait variation along the fast-slow growth continuum (Fig.~\ref{fig:Fig_ClimVarImp}).
+A life history tradeoff between allocation to growth in the current year at the expense of future years could explain this pattern.
+In southern populations with more frequent droughts capable of killing rhizomes, a more annualized strategy could be favored.
+Conversely, in more predictable northern environments, lifetime fitness may be optimized when a significant fraction of assimilate is allocated below ground for future years.
+Although this hypothesis remains to be directly tested, a few indepdent lines of evidence are consistent with it.
+Preliminary surveys suggest that northern populations not only grow slower, but also produce greater numbers of rhizomes (C.D.
+Muir, unpub.
+data), suggesting an allocation tradeoff.
+Ecological niche models also show that occurence of southern populations is best predicted by recent climate (< 5 years), whereas northern occurences are best predicted by climate over the previous 30 years (M.
+Bayly \& A.
+Angert, unpub.
+data).
+Finally, demographic surveys of natural populations show greater variation in the size recruits in southern populations, suggesting higher maximum growth rates under natural conditions (M.
+Bayly \& A.
+Angert, unpub.
+data).
 There is a lot of interest in understanding how organisms will respond to changes in climatic variation, not just changes in the average climate.
@@
 
-But what we really need to understand is why certain types of traits are repeatedly targets of natural selection, whereas other traits are constrained.
-
-5.
-Future work and broad conclusions.
+[Future work and broad conclusions].
 Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
-Next, we are testing whether phenotypic variation and constraint can be predicted by the shape of tradeoffs, as predicted by evolutionary theory,
-
-% stuff about environmental grain
-How organisms adapt depends on the grain of the environment, ..., and gene flow between populations
-
-Possibilities:
-1. no trait variation: no environmental variation, high gene flow, generalist (convex)
-2. variation in mean:
-3. variation in plasticity: convex?,
-
-our data are most consistent with:
-- concave tradeoffs, fine-grained (coarse+concave would predict alt morphs), high gene flow
+Next, we are testing whether phenotypic variation and constraint can be predicted by the shape of tradeoffs, as predicted by evolutionary theory.
+\cdm{I will work more on this once I am more comfortable with the rest of the discussion}.
 
@@
 \section*{Acknowledgements}
-Erin Warkman and Lisa Lin helped collect data.
+Erin Warkman and Lisa Lin helped collect data. CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program. \cdm{Please fill in other relevant funding info}
 
@@
 
+\clearpage
+
 %--------------------------------------------------
-% Supplementary Information
+% Supporting Information
 %--------------------------------------------------
+
+\section*{Supporting Information}
+
+% Modify and restart table/figure numbering for appendixes
+\renewcommand\thefigure{S\arabic{figure}}    
+\renewcommand\thetable{S\arabic{table}}    
+\renewcommand\theequation{S\arabic{equation}}    
+\setcounter{table}{0}    
+\setcounter{equation}{0}
+\setcounter{figure}{0}
 

### 2015-04-30 ms/ms.tex
Finished rough draft

@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-  \hlcom{# The path to the ms directory may need to be changed for different users}
-  \hlcom{# Manuscript files}
-  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"} \hlcom{# Chris' MacBook}
-  \hlcom{# pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop}
-
-  \hlcom{# Data files for paper}
-  \hlstd{pathDat} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Data"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
-
-  \hlcom{# Figures for paper}
-  \hlstd{pathFig} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Figures"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
-
-  \hlcom{# Tables for paper}
-  \hlstd{pathTab} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Tables"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
-
-  \hlcom{# Small function to make statistical significance asterisks}
-  \hlstd{sigStar} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{pvalue}\hlstd{)}
-  \hlstd{\{}
-    \hlkwa{if} \hlstd{(pvalue} \hlopt{>=} \hlnum{0.05}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{""}\hlstd{)}
-    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.05} \hlopt{&} \hlstd{pvalue} \hlopt{>=} \hlnum{0.01}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"*"}\hlstd{)}
-    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.01} \hlopt{&} \hlstd{pvalue} \hlopt{>=} \hlnum{0.001}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"**"}\hlstd{)}
-    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.001}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"***"}\hlstd{)}
-  \hlstd{\}}
-
-  \hlcom{# Color palette}
-  \hlcom{# Primary color:}
-
-   \hlstd{primShade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.286}\hlstd{)}
-   \hlstd{primShade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0.643}\hlstd{,} \hlnum{0.745}\hlstd{)}
-   \hlstd{primShade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0.439}\hlstd{,} \hlnum{0.6}\hlstd{)}
-   \hlstd{primShade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.765}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.22}\hlstd{)}
-   \hlstd{primShade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.588}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.169}\hlstd{)}
-
-  \hlcom{# Secondary color (1):}
-
-  \hlstd{sec1shade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.008}\hlstd{,} \hlnum{0.957}\hlstd{,} \hlnum{1}\hlstd{)}
-  \hlstd{sec1shade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.643}\hlstd{,} \hlnum{0.984}\hlstd{,} \hlnum{1}\hlstd{)}
-  \hlstd{sec1shade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.443}\hlstd{,} \hlnum{0.976}\hlstd{,} \hlnum{1}\hlstd{)}
-  \hlstd{sec1shade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.49}\hlstd{,} \hlnum{0.514}\hlstd{)}
-  \hlstd{sec1shade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.376}\hlstd{,} \hlnum{0.396}\hlstd{)}
-
-  \hlcom{# Secondary color (2):}
-
-  \hlstd{sec2shade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.988}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0}\hlstd{)}
-  \hlstd{sec2shade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.996}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.643}\hlstd{)}
-  \hlstd{sec2shade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.992}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.439}\hlstd{)}
-  \hlstd{sec2shade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.835}\hlstd{,} \hlnum{0.847}\hlstd{,} \hlnum{0}\hlstd{)}
-  \hlstd{sec2shade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.643}\hlstd{,} \hlnum{0.651}\hlstd{,} \hlnum{0}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
 %
@@
 
-We used 16 populations from throughout the range of \textit{M. cardinalis} (Table 1). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
+We used 16 populations from throughout the range of \textit{M. cardinalis} (Table~\ref{table:Table_FocalPops}). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of focal population
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
    \centering
-   \topcaption{16 Focal populations}
+   \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
    \begin{tabular}{@{} lllllll @{}}
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{library}\hlstd{(scales)} \hlcom{# for color transparency}
-\hlstd{exptlDes} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/ExperimentalDesign.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-
-\hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/ExperimentalDesignBase.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{7}\hlstd{,} \hlnum{5}\hlstd{)}
-\hlkwd{par}\hlstd{(}\hlkwc{mai} \hlstd{=} \hlkwd{rep}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{4}\hlstd{))}
-\hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlkwd{max}\hlstd{(exptlDes}\hlopt{$}\hlstd{DayN)),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{1}\hlstd{,} \hlnum{41}\hlstd{),}
-                 \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{,} \hlkwc{xlab} \hlstd{=} \hlstr{"Key treatments and measurements"}\hlstd{,}
-     \hlkwc{ylab} \hlstd{=} \hlkwd{expression}\hlstd{(}\hlkwd{paste}\hlstd{(}\hlstr{"Temperature ("}\hlstd{,} \hlkwd{phantom}\hlstd{()}\hlopt{*}\hlstd{degree,} \hlstr{"C)"}\hlstd{)))}
-
-\hlcom{# Temperature axis}
-\hlkwd{axis}\hlstd{(}\hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{10}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,}
-        \hlkwc{labels} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{expression}\hlstd{(}\hlnum{0}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{10}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{20}\hlopt{*}\hlstd{degree),}
-                                                 \hlkwd{expression}\hlstd{(}\hlnum{30}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{40}\hlopt{*}\hlstd{degree)))}
-\hlcom{# Axis of time on top}
-\hlkwd{mtext}\hlstd{(}\hlstr{"Days after sowing"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{3.5}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
-\hlkwd{mtext}\hlstd{(}\hlstr{"(Date for climate simulation)"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{2.25}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
-\hlkwd{axis}\hlstd{(}\hlnum{3}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{90}\hlstd{,} \hlnum{30}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-\hlkwd{axis}\hlstd{(}\hlnum{3}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{17}\hlstd{,} \hlnum{47}\hlstd{,} \hlnum{78}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlopt{-}\hlnum{0.5}\hlstd{,}
-     \hlkwc{labels} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"(June)"}\hlstd{,} \hlstr{"(July)"}\hlstd{,} \hlstr{"(August)"}\hlstd{))}
-
-\hlcom{# Indicate sowing, beginning of temperature and drought treatments}
-\hlkwd{abline}\hlstd{(}\hlkwc{v} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-\hlkwd{mtext}\hlstd{(}\hlstr{"Sow"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{0}\hlstd{)}
-\hlkwd{abline}\hlstd{(}\hlkwc{v} \hlstd{=} \hlnum{17}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-\hlkwd{mtext}\hlstd{(}\hlstr{"Temp\textbackslash{}nTreatment"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlnum{17}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{1}\hlstd{)}
-\hlkwd{abline}\hlstd{(}\hlkwc{v} \hlstd{=} \hlnum{37}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-\hlkwd{mtext}\hlstd{(}\hlstr{"Drought\textbackslash{}nTreatment"}\hlstd{,} \hlkwc{side} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlnum{37}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{1}\hlstd{)}
-
-\hlcom{# Polygon of Hot treatment, min and max}
-\hlkwd{with}\hlstd{(exptlDes[}\hlnum{18}\hlopt{:}\hlnum{94}\hlstd{, ],} \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{17}\hlopt{:}\hlnum{93}\hlstd{,} \hlnum{93}\hlopt{:}\hlnum{17}\hlstd{),} \hlkwd{c}\hlstd{(Tmin_hot,} \hlkwd{rev}\hlstd{(Tmax_hot)),}
-  \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(primShade0,} \hlnum{0.25}\hlstd{)))}
-
-\hlcom{# Polygon of Cool treatment, min and max}
-\hlkwd{with}\hlstd{(exptlDes[}\hlnum{18}\hlopt{:}\hlnum{94}\hlstd{, ],} \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{17}\hlopt{:}\hlnum{93}\hlstd{,} \hlnum{93}\hlopt{:}\hlnum{17}\hlstd{),} \hlkwd{c}\hlstd{(Tmin_cool,} \hlkwd{rev}\hlstd{(Tmax_cool)),}
-  \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(sec1shade0,} \hlnum{0.25}\hlstd{)))}
-
-\hlcom{# Lines with average temperature}
-\hlkwd{with}\hlstd{(exptlDes[}\hlnum{18}\hlopt{:}\hlnum{94}\hlstd{, ],} \hlkwd{points}\hlstd{(}\hlnum{17}\hlopt{:}\hlnum{93}\hlstd{, Tavg_hot,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"l"}\hlstd{,}
-        \hlkwc{col} \hlstd{= primShade3))}
-\hlkwd{with}\hlstd{(exptlDes[}\hlnum{18}\hlopt{:}\hlnum{94}\hlstd{, ],} \hlkwd{points}\hlstd{(}\hlnum{17}\hlopt{:}\hlnum{93}\hlstd{, Tavg_cool,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"l"}\hlstd{,}
-        \hlkwc{col} \hlstd{= sec1shade3))}
-
-\hlcom{# Rug indicating measurment timing}
-\hlkwd{rect}\hlstd{(}\hlnum{6}\hlstd{,} \hlopt{-}\hlnum{2}\hlstd{,} \hlnum{17}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlnum{11.5}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Germ"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,}
-  \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
-
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{rect}\hlstd{(}\hlkwd{min}\hlstd{(}\hlkwd{which}\hlstd{(LERmeas)),} \hlopt{-}\hlnum{2}\hlstd{,} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(LERmeas)),} \hlnum{0}\hlstd{,}
-        \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{))}
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlkwd{min}\hlstd{((}\hlkwd{which}\hlstd{(LERmeas))} \hlopt{+} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(LERmeas)))} \hlopt{/} \hlnum{2}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{,}
-        \hlkwc{labels} \hlstd{=} \hlstr{"LER"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
-
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{rect}\hlstd{(}\hlkwd{min}\hlstd{(}\hlkwd{which}\hlstd{(SERmeas)),} \hlnum{0}\hlstd{,} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(SERmeas)),} \hlnum{2}\hlstd{,}
-        \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{))}
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlkwd{min}\hlstd{((}\hlkwd{which}\hlstd{(SERmeas))} \hlopt{+} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(SERmeas)))} \hlopt{/} \hlnum{2}\hlstd{,} \hlnum{1}\hlstd{,}
-        \hlkwc{labels} \hlstd{=} \hlstr{"SER"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
-
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{rect}\hlstd{(}\hlkwd{min}\hlstd{(}\hlkwd{which}\hlstd{(Harvest)),} \hlopt{-}\hlnum{2}\hlstd{,} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(Harvest)),} \hlnum{0}\hlstd{,}
-        \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{))}
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlkwd{min}\hlstd{((}\hlkwd{which}\hlstd{(Harvest))} \hlopt{+} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(Harvest)))} \hlopt{/} \hlnum{2}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{,}
-        \hlkwc{labels} \hlstd{=} \hlstr{"Harvest"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
-
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{text}\hlstd{(}\hlkwd{min}\hlstd{((}\hlkwd{which}\hlstd{(PhotoMeas))} \hlopt{+} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(PhotoMeas)))} \hlopt{/} \hlnum{2}\hlstd{,} \hlnum{5}\hlstd{,}
-        \hlkwc{labels} \hlstd{=} \hlstr{"Photo"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{,} \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{0.9}\hlstd{))}
-\hlkwd{with}\hlstd{(exptlDes,} \hlkwd{rect}\hlstd{(}\hlkwd{min}\hlstd{(}\hlkwd{which}\hlstd{(PhotoMeas)),} \hlnum{2}\hlstd{,} \hlkwd{max}\hlstd{(}\hlkwd{which}\hlstd{(PhotoMeas)),} \hlnum{4}\hlstd{,}
-        \hlkwc{border} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{))}
-\hlcom{#with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 3, }
-\hlcom{#	labels = "Photo", col = "white", font = 3, cex = 0.9))}
-
-\hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\begin{alltt}
-\hlcom{# Legend}
-\hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/ExperimentalDesignLegend.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{1.5}\hlstd{,} \hlnum{2}\hlstd{)}
-\hlkwd{par}\hlstd{(}\hlkwc{mai} \hlstd{=} \hlkwd{rep}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{4}\hlstd{))}
-\hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{1.5}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{1}\hlstd{),} \hlkwc{axes} \hlstd{= F)}
-
-\hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwd{c}\hlstd{(}\hlnum{0.8}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.55}\hlstd{,} \hlnum{0.55}\hlstd{,} \hlnum{0.35}\hlstd{),} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{,}
-                                \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(primShade0,} \hlnum{0.25}\hlstd{))}
-\hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwd{c}\hlstd{(}\hlnum{0.6}\hlstd{,} \hlnum{0.45}\hlstd{,} \hlnum{0.45}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.15}\hlstd{),} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{,}
-                                \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(sec1shade0,} \hlnum{0.25}\hlstd{))}
-\hlkwd{points}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{1}\hlstd{),} \hlkwd{c}\hlstd{(}\hlnum{0.575}\hlstd{,} \hlnum{0.775}\hlstd{,} \hlnum{0.775}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"l"}\hlstd{,} \hlkwc{col} \hlstd{= primShade3)}
-\hlkwd{points}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.5}\hlstd{,} \hlnum{1}\hlstd{),} \hlkwd{c}\hlstd{(}\hlnum{0.375}\hlstd{,} \hlnum{0.225}\hlstd{,} \hlnum{0.225}\hlstd{),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"l"}\hlstd{,} \hlkwc{col} \hlstd{= sec1shade3)}
-\hlkwd{text}\hlstd{(}\hlnum{0.5}\hlstd{,} \hlnum{0.9}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Hot"}\hlstd{,} \hlkwc{col} \hlstd{= primShade3)}
-\hlkwd{text}\hlstd{(}\hlnum{0.5}\hlstd{,} \hlnum{0.1}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Cool"}\hlstd{,} \hlkwc{col} \hlstd{= sec1shade3)}
-\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{1}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Max"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(primShade0,} \hlnum{0.25}\hlstd{))}
-\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0.55}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Min"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(primShade0,} \hlnum{0.25}\hlstd{))}
-\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0.775}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Mean"}\hlstd{,} \hlkwc{col} \hlstd{= primShade3)}
-
-\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Min"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(sec1shade0,} \hlnum{0.25}\hlstd{))}
-\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0.45}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Max"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{alpha}\hlstd{(sec1shade0,} \hlnum{0.25}\hlstd{))}
-\hlkwd{text}\hlstd{(}\hlnum{1.25}\hlstd{,} \hlnum{0.225}\hlstd{,} \hlkwc{labels} \hlstd{=} \hlstr{"Mean"}\hlstd{,} \hlkwc{col} \hlstd{= sec1shade3)}
-
-\hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure summarizing experimental design
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-  \hlkwd{library}\hlstd{(survival)}
-  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"}
-  \hlstd{pathDat} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Data"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
-
-  \hlstd{master} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/MasterDatasheet_out.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),}
-    \hlkwc{row.names} \hlstd{=} \hlnum{1}\hlstd{)}
-
-  \hlcom{# Create Surv object}
-        \hlstd{germ} \hlkwb{<-} \hlkwd{with}\hlstd{(}\hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{MinGermDay)),} \hlkwd{Surv}\hlstd{(MinGermDay, MaxGermDay,}
-                \hlkwc{type} \hlstd{=} \hlstr{"interval2"}\hlstd{))}
-
-        \hlcom{# lognormal produced best fit to data}
-        \hlstd{fitGerm} \hlkwb{<-} \hlkwd{survreg}\hlstd{(germ} \hlopt{~} \hlstd{PopID} \hlopt{+} \hlkwd{frailty}\hlstd{(Family,} \hlkwc{sparse} \hlstd{= F),}
-                \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{MinGermDay)),} \hlkwc{dist} \hlstd{=} \hlstr{"lognormal"}\hlstd{)}
-        \hlkwd{suppressWarnings}\hlstd{(}\hlkwd{anova}\hlstd{(fitGerm))} \hlcom{# significant family and population effects}
-\end{alltt}
-\begin{verbatim}
-##                                   Df Deviance Resid. Df    -2*LL
-## NULL                              NA       NA  700.0000 2796.417
-## PopID                        15.0000 151.9495  685.0000 2644.468
-## frailty(Family, sparse = F) 106.9738 254.7233  578.0262 2389.745
-##                                 Pr(>Chi)
-## NULL                                  NA
-## PopID                       9.887489e-25
-## frailty(Family, sparse = F) 4.530519e-14
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
 \paragraph{Growth rate: leaf expansion and shoot elongation}
@@
 Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-  \hlkwd{library}\hlstd{(lme4)}
-\end{alltt}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: Matrix\\\#\# Loading required package: Rcpp}}\begin{alltt}
-  \hlkwd{library}\hlstd{(lmerTest)}
-\end{alltt}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'lmerTest'\\\#\# \\\#\# The following object is masked from 'package:lme4':\\\#\# \\\#\#\ \ \ \  lmer\\\#\# \\\#\# The following object is masked from 'package:stats':\\\#\# \\\#\#\ \ \ \  step}}\begin{alltt}
-  \hlcom{# New (blank) datasheet on which to add data}
-  \hlstd{newdata} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/MasterDatasheet_in.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-
-  \hlcom{# Read in LER data}
-  \hlstd{LERdata} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/LERdata.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-
-  \hlcom{#}
-        \hlcom{# Empirical Bayes Approach: get coefficients for every plant}
-        \hlcom{#}
-
-        \hlcom{# suppressWarnings(mmLER <- lmer(log(LLL) ~ poly(DayN, 2, raw = T) + }
-    \hlcom{# (poly(DayN, 2, raw = T)|indiv), data = LERdata))}
-  \hlcom{# save(mmLER, file = paste(pathDat, "/mmLER.RData", sep = ""))}
-  \hlkwd{load}\hlstd{(}\hlkwc{file} \hlstd{=} \hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/mmLER.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-
-  \hlcom{# Extract coefficients and add to datasheet}
-        \hlstd{Y} \hlkwb{<-} \hlkwd{t}\hlstd{(}\hlkwd{unlist}\hlstd{(}\hlkwd{fixef}\hlstd{(mmLER))} \hlopt{+} \hlkwd{t}\hlstd{(}\hlkwd{ranef}\hlstd{(mmLER)}\hlopt{$}\hlstd{indiv))[}\hlkwd{match}\hlstd{(newdata}\hlopt{$}\hlstd{indiv,}
-                \hlkwd{rownames}\hlstd{(}\hlkwd{ranef}\hlstd{(mmLER)}\hlopt{$}\hlstd{indiv)), ]}
-        \hlkwd{colnames}\hlstd{(Y)} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"b0_LER"}\hlstd{,} \hlstr{"b1_LER"}\hlstd{,} \hlstr{"b2_LER"}\hlstd{)}
-  \hlstd{newdata} \hlkwb{<-} \hlkwd{cbind}\hlstd{(newdata, Y)}
-
-  \hlcom{# Model-based LER}
-        \hlcom{# Timeframe for Cool treatment: Day 0 - 31}
-        \hlcom{# Timeframe for Hot treatment: Day 0 - 24}
-        \hlstd{newdata}\hlopt{$}\hlstd{LERstart} \hlkwb{<-} \hlkwd{exp}\hlstd{(newdata}\hlopt{$}\hlstd{b0_LER)}
-        \hlstd{newdata}\hlopt{$}\hlstd{LERend} \hlkwb{<-} \hlkwd{exp}\hlstd{(newdata}\hlopt{$}\hlstd{b0_LER} \hlopt{+}
-                                        \hlstd{newdata}\hlopt{$}\hlstd{b1_LER} \hlopt{*} \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31}\hlstd{,} \hlnum{24}\hlstd{)} \hlopt{+}
-                                        \hlstd{newdata}\hlopt{$}\hlstd{b2_LER} \hlopt{*} \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31} \hlopt{^} \hlnum{2}\hlstd{,} \hlnum{24} \hlopt{^} \hlnum{2}\hlstd{))}
-        \hlstd{newdata}\hlopt{$}\hlstd{LER_AbsGrowth} \hlkwb{<-} \hlstd{(newdata}\hlopt{$}\hlstd{LERend} \hlopt{-} \hlstd{newdata}\hlopt{$}\hlstd{LERstart)} \hlopt{/}
-                \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31}\hlstd{,} \hlnum{24}\hlstd{)}
-
-  \hlcom{# Mixed model ANOVA with family as random factor fit using step down procedure}
-        \hlstd{mm} \hlkwb{<-} \hlkwd{lmer}\hlstd{(LER_AbsGrowth} \hlopt{~} \hlstd{AvgGermDay} \hlopt{+} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt} \hlopt{+} \hlstd{(}\hlnum{1}\hlopt{|}\hlstd{Family),}
-                \hlkwc{data} \hlstd{= master)}
-        \hlstd{fitLER} \hlkwb{<-} \hlkwd{step}\hlstd{(mm,} \hlkwc{reduce.random} \hlstd{= F)} \hlcom{# includes pop, temp, h2o, population x water}
-        \hlcom{# fitLER <- step(mm, ddf = "Kenward-Roger", reduce.random = F) # same result using Satterthwaite ddf}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-  \hlcom{# Read in SER data}
-  \hlstd{SERdata} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/SERdata.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-
-  \hlcom{#}
-  \hlcom{# Empirical Bayes Approach: get coefficients for every plant}
-        \hlcom{#}
-
-        \hlcom{# suppressWarnings(mmSER <- lmer(log(height + 1) ~ poly(DayN, 2, raw = T) + }
-    \hlcom{# (poly(DayN, 2, raw = T)|indiv), data = SERdata))}
-  \hlcom{# save(mmSER, file = paste(pathDat, "/mmSER.RData", sep = ""))}
-  \hlkwd{load}\hlstd{(}\hlkwc{file} \hlstd{=} \hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/mmSER.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-
-  \hlcom{# Extract coefficients and add to datasheet}
-  \hlstd{Y} \hlkwb{<-} \hlkwd{t}\hlstd{(}\hlkwd{unlist}\hlstd{(}\hlkwd{fixef}\hlstd{(mmSER))} \hlopt{+} \hlkwd{t}\hlstd{(}\hlkwd{ranef}\hlstd{(mmSER)}\hlopt{$}\hlstd{indiv))[}\hlkwd{match}\hlstd{(newdata}\hlopt{$}\hlstd{indiv,}
-                \hlkwd{rownames}\hlstd{(}\hlkwd{ranef}\hlstd{(mmSER)}\hlopt{$}\hlstd{indiv)), ]}
-        \hlkwd{colnames}\hlstd{(Y)} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"b0_SER"}\hlstd{,} \hlstr{"b1_SER"}\hlstd{,} \hlstr{"b2_SER"}\hlstd{)}
-  \hlstd{newdata} \hlkwb{<-} \hlkwd{cbind}\hlstd{(newdata, Y)}
-
-  \hlcom{# Model-based SER}
-        \hlcom{# Timeframe for Cool treatment: Day 0 - 31}
-        \hlcom{# Timeframe for Hot treatment: Day 0 - 24}
-        \hlstd{newdata}\hlopt{$}\hlstd{SERstart} \hlkwb{<-} \hlkwd{exp}\hlstd{(newdata}\hlopt{$}\hlstd{b0_SER)}
-        \hlstd{newdata}\hlopt{$}\hlstd{SERend} \hlkwb{<-} \hlkwd{exp}\hlstd{(newdata}\hlopt{$}\hlstd{b0_SER} \hlopt{+}
-                                        \hlstd{newdata}\hlopt{$}\hlstd{b1_SER} \hlopt{*} \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31}\hlstd{,} \hlnum{24}\hlstd{)} \hlopt{+}
-                                        \hlstd{newdata}\hlopt{$}\hlstd{b2_SER} \hlopt{*} \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31} \hlopt{^} \hlnum{2}\hlstd{,} \hlnum{24} \hlopt{^} \hlnum{2}\hlstd{))}
-        \hlstd{newdata}\hlopt{$}\hlstd{SER_AbsGrowth} \hlkwb{<-} \hlstd{(newdata}\hlopt{$}\hlstd{SERend} \hlopt{-} \hlstd{newdata}\hlopt{$}\hlstd{SERstart)} \hlopt{/}
-                \hlkwd{ifelse}\hlstd{(newdata}\hlopt{$}\hlstd{TempTrt} \hlopt{==} \hlstr{"Cool"}\hlstd{,} \hlnum{31}\hlstd{,} \hlnum{24}\hlstd{)}
-
-  \hlstd{mm} \hlkwb{<-} \hlkwd{lmer}\hlstd{(SER_AbsGrowth} \hlopt{~} \hlstd{AvgGermDay} \hlopt{+} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt} \hlopt{+}
-                \hlstd{(}\hlnum{1}\hlopt{|}\hlstd{Family),} \hlkwc{data} \hlstd{= master)}
-        \hlstd{fitSER} \hlkwb{<-} \hlkwd{step}\hlstd{(mm,} \hlkwc{reduce.random} \hlstd{= F)} \hlcom{#includes pop, temp, h2o, temp x water}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-  \hlcom{# Photosynthetic rate}
-  \hlstd{mm} \hlkwb{<-} \hlkwd{lmer}\hlstd{(Photo} \hlopt{~} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt} \hlopt{+} \hlstd{(}\hlnum{1}\hlopt{|}\hlstd{Family),}
-    \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{Photo)))}
-  \hlstd{fitPhoto} \hlkwb{<-} \hlkwd{step}\hlstd{(mm,} \hlkwc{reduce.random} \hlstd{= F)}
-
-  \hlcom{# Intrinsic photosynthetic rate}
-  \hlcom{# = residuals of Conductance - Photosynthesis regression}
-  \hlcom{# Removed one outlier (may change if data are rearranged!)}
-        \hlstd{mm} \hlkwb{<-} \hlkwd{lmer}\hlstd{(resPhoto} \hlopt{~} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt} \hlopt{+} \hlstd{(}\hlnum{1}\hlopt{|}\hlstd{Family),}
-    \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{Photo))[}\hlopt{-}\hlnum{52}\hlstd{, ])}
-        \hlstd{fitIntrnPhoto} \hlkwb{<-} \hlkwd{step}\hlstd{(mm,} \hlkwc{reduce.random} \hlstd{= F)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
 \paragraph{Mortality}
@@
 We could not get GLMM with Family effects to converge, so we used GLM with a quasibinomial error structure and assessed signifiance using Type 2 Analysis of Deviance with the R package \pkg{car}.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-  \hlkwd{library}\hlstd{(car)}
-
-  \hlstd{fitMort} \hlkwb{<-} \hlkwd{glm}\hlstd{(DiedFromStress} \hlopt{~} \hlstd{PopID} \hlopt{*} \hlstd{TempTrt} \hlopt{*} \hlstd{WaterTrt,}
-    \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{DiedFromStress)),} \hlkwc{family} \hlstd{=} \hlstr{"quasibinomial"}\hlstd{)}
-
-  \hlkwd{Anova}\hlstd{(fitMort,} \hlkwc{type} \hlstd{=} \hlnum{2}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## Analysis of Deviance Table (Type II tests)
-## 
-## Response: DiedFromStress
-##                        LR Chisq Df Pr(>Chisq)    
-## PopID                    38.169 15  0.0008519 ***
-## TempTrt                 247.763  1  < 2.2e-16 ***
-## WaterTrt                 79.508  1  < 2.2e-16 ***
-## PopID:TempTrt            26.188 15  0.0360962 *  
-## PopID:WaterTrt           11.547 15  0.7129658    
-## TempTrt:WaterTrt         48.543  1  3.231e-12 ***
-## PopID:TempTrt:WaterTrt    8.685 15  0.8934160    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
 
@@
 \subsection*{Principal components of germination, growth, and photosynthesis}
-[maybe this goes in Results?]
 We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components.
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-  \hlstd{popMeans} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}
-    \hlkwc{germ} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{1}\hlstd{],} \hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{1}\hlstd{]} \hlopt{+} \hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{2}\hlopt{:}\hlnum{16}\hlstd{]),}
-    \hlkwc{LER} \hlstd{= fitLER}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{],}
-    \hlkwc{SER} \hlstd{= fitSER}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{],}
-    \hlkwc{photo} \hlstd{= fitIntrnPhoto}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{])}
-
-  \hlstd{traitPC} \hlkwb{<-} \hlkwd{prcomp}\hlstd{(popMeans,} \hlkwc{center} \hlstd{= T,} \hlkwc{scale.} \hlstd{= T)}
-  \hlstd{focClim}\hlopt{$}\hlstd{PC1} \hlkwb{<-} \hlstd{traitPC}\hlopt{$}\hlstd{x[,} \hlstr{"PC1"}\hlstd{]}
-  \hlstd{focClimSA}\hlopt{$}\hlstd{PC1} \hlkwb{<-} \hlstd{traitPC}\hlopt{$}\hlstd{x[,} \hlstr{"PC1"}\hlstd{]}
-
-  \hlcom{# Percent variance explained by first principal component}
-  \hlstd{pve} \hlkwb{<-} \hlkwd{paste}\hlstd{(}\hlkwd{round}\hlstd{(}\hlnum{100} \hlopt{*} \hlstd{traitPC}\hlopt{$}\hlstd{sdev[}\hlnum{1}\hlstd{]} \hlopt{^} \hlnum{2} \hlopt{/} \hlkwd{sum}\hlstd{(traitPC}\hlopt{$}\hlstd{sdev} \hlopt{^} \hlnum{2}\hlstd{),} \hlnum{1}\hlstd{),} \hlstr{"%"}\hlstd{)}
-
-  \hlcom{# Plot latitude versus PC1}
-  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_PC1vLat.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{5.5}\hlstd{,} \hlnum{5}\hlstd{)}
-  \hlkwd{par}\hlstd{(}\hlkwc{mai} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{1.5}\hlstd{,} \hlnum{0.25}\hlstd{,} \hlnum{0.25}\hlstd{))}
-  \hlkwd{plot}\hlstd{(focClim}\hlopt{$}\hlstd{Lat, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{32}\hlstd{,} \hlnum{44}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{axes} \hlstd{= F,}
-    \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
-  \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Latitude"}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
-  \hlkwd{title}\hlstd{(}\hlkwc{ylab} \hlstd{=} \hlkwd{substitute}\hlstd{(}\hlkwd{plain}\hlstd{(Principal}\hlopt{~}\hlstd{Component}\hlopt{~}\hlnum{1}\hlstd{)}\hlopt{~}\hlkwd{group}\hlstd{(}\hlstr{"("}\hlstd{, x,} \hlstr{")"}\hlstd{),}
-    \hlkwd{list}\hlstd{(}\hlkwc{x} \hlstd{= pve)))}
-  \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{32}\hlstd{,} \hlnum{44}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{labels} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{expression}\hlstd{(}\hlnum{32}\hlopt{*}\hlstd{degree),}
-                \hlkwd{expression}\hlstd{(}\hlnum{36}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{40}\hlopt{*}\hlstd{degree),} \hlkwd{expression}\hlstd{(}\hlnum{44}\hlopt{*}\hlstd{degree)),}
-                \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-  \hlkwd{axis}\hlstd{(}\hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{,} \hlnum{2}\hlstd{),} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-  \hlstd{fitLat} \hlkwb{<-} \hlkwd{lm}\hlstd{(PC1} \hlopt{~} \hlstd{Latitude,} \hlkwc{data} \hlstd{= focClim)}
-
-        \hlcom{# Label y-axis from slow to fast growth}
-        \hlkwd{mtext}\hlstd{(}\hlkwd{c}\hlstd{(}\hlstr{"Slow\textbackslash{}nGrowth"}\hlstd{,} \hlstr{"Fast\textbackslash{}nGrowth"}\hlstd{),} \hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{cex} \hlstd{=} \hlnum{1.25}\hlstd{,}
-                                \hlkwc{font} \hlstd{=} \hlnum{3}\hlstd{,} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{adj} \hlstd{=} \hlnum{0.5}\hlstd{)}
-
-  \hlcom{# polygon of confidence intervals}
-  \hlstd{x} \hlkwb{<-} \hlkwd{seq}\hlstd{(}\hlkwd{min}\hlstd{(focClim}\hlopt{$}\hlstd{Latitude),} \hlkwd{max}\hlstd{(focClim}\hlopt{$}\hlstd{Latitude),} \hlkwc{l} \hlstd{=} \hlnum{1e3}\hlstd{)}
-  \hlstd{u} \hlkwb{<-} \hlkwd{predict}\hlstd{(fitLat,} \hlkwc{new} \hlstd{=} \hlkwd{data.frame}\hlstd{(}\hlkwc{Latitude} \hlstd{= x))}
-  \hlstd{s} \hlkwb{<-} \hlkwd{sqrt}\hlstd{(}\hlkwd{sapply}\hlstd{(x,} \hlkwa{function}\hlstd{(}\hlkwc{X}\hlstd{)} \hlkwd{t}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X))} \hlopt{%*%} \hlkwd{vcov}\hlstd{(fitLat)} \hlopt{%*%} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X)))}
-  \hlstd{confInt} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwc{lowerCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.025}\hlstd{, u, s),} \hlkwc{upperCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.975}\hlstd{, u, s))}
-  \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(x,} \hlkwd{rev}\hlstd{(x)),} \hlkwd{c}\hlstd{(confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwd{rev}\hlstd{(confInt}\hlopt{$}\hlstd{upperCI)),}
-    \hlkwc{col} \hlstd{=} \hlstr{"grey80"}\hlstd{,} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{)}
-  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{upperCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{lines}\hlstd{(}\hlkwd{range}\hlstd{(focClim}\hlopt{$}\hlstd{Lat),} \hlkwd{rev}\hlstd{(}\hlkwd{range}\hlstd{(}\hlkwd{predict}\hlstd{(fitLat))),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{points}\hlstd{(focClim}\hlopt{$}\hlstd{Lat, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{= primShade0,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{,}
-    \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\begin{alltt}
-  \hlcom{# plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)}
-  \hlcom{#plot(focClim$Lat, traitPC$x[, "PC1"])}
-  \hlcom{#cor.test(focClim$Lat, traitPC$x[, "PC1"])}
-  \hlcom{#cor.test(popMeans[, 4], traitPC$x[, "PC1"])}
-  \hlcom{#plot(focClim$bio16_cva, traitPC$x[, "PC1"])}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# R data.frame of 178 M. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data}
-\hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/thinnedOccClimate.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast and slow growing populations.
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{library}\hlstd{(VSURF)}
-\end{alltt}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: randomForest\\\#\# randomForest 4.6-10\\\#\# Type rfNews() to see new features/changes/bug fixes.\\\#\# Loading required package: rpart\\\#\# Loading required package: doParallel\\\#\# Loading required package: foreach\\\#\# Loading required package: iterators\\\#\# Loading required package: parallel}}\begin{alltt}
-  \hlcom{# Load VSURF results of climatic variables associated with latitude at cardinalis occurence points}
-  \hlcom{# Note that occurences have been randomly thinned to account for variation in sampling effort}
-  \hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/clim2lat_fit_t1.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))} \hlcom{# fit_t1}
-  \hlstd{vsurfLat} \hlkwb{<-} \hlstd{fit_t1}
-
-        \hlcom{# Load climVars}
-        \hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/ExploratoryCimateVariables.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-
-  \hlcom{# Figure of variable importance for latitude}
-  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_clim2lat_imp.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{4}\hlstd{,} \hlnum{5}\hlstd{)}
-  \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{6}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
-  \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlkwd{ceiling}\hlstd{(vsurfLat}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x)[}\hlnum{1}\hlstd{]),}
-                \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)),} \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,}
-                \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
-        \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Climate-Latitude\textbackslash{}nImportance"}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
-  \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlnum{1}\hlopt{:}\hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred))}
-  \hlstd{\{}
-     \hlkwd{lines}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{1}\hlstd{,} \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(i, i),} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
-  \hlstd{\}}
-  \hlkwd{points}\hlstd{(vsurfLat}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)],}
-    \hlkwd{length}\hlstd{(vsurfLat}\hlopt{$}\hlstd{varselect.pred)}\hlopt{:}\hlnum{1}\hlstd{,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlcom{# mtext(climVars[vsurfLat$varselect.pred], side = 2, }
-    \hlcom{# at = length(vsurfLat$varselect.pred):1, las = 1, line = 1)}
-        \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-        \hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\begin{alltt}
-  \hlcom{# Variable selection using random forests (VSURF)}
-  \hlstd{linmod.PC1} \hlkwb{<-} \hlkwd{as.formula}\hlstd{(}\hlkwd{paste}\hlstd{(}\hlstr{"PC1 ~ "}\hlstd{,} \hlkwd{paste}\hlstd{(climVars,} \hlkwc{collapse} \hlstd{=} \hlstr{" + "}\hlstd{),} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
-  \hlkwd{set.seed}\hlstd{(}\hlnum{69647}\hlstd{)}
-  \hlstd{vsurfPC1} \hlkwb{<-} \hlkwd{VSURF.parallel}\hlstd{(linmod.PC1,} \hlkwc{data} \hlstd{= focClim[,} \hlkwd{c}\hlstd{(}\hlstr{"PC1"}\hlstd{, climVars)],}
-    \hlkwc{ncores} \hlstd{=} \hlnum{3}\hlstd{)}
-\end{alltt}
-
-{\ttfamily\noindent\color{warningcolor}{\#\# Warning in VSURF.parallel.formula(linmod.PC1, data = focClim[, c("{}PC1"{}, : VSURF with a formula-type call outputs selected variables\\\#\# which are indices of the input matrix based on the formula:\\\#\# you may reorder these to get indices of the original data}}\begin{alltt}
-  \hlstd{vsurfPC1_SA} \hlkwb{<-} \hlkwd{VSURF.parallel}\hlstd{(linmod.PC1,} \hlkwc{data} \hlstd{= focClimSA[,} \hlkwd{c}\hlstd{(}\hlstr{"PC1"}\hlstd{, climVars)],}
-    \hlkwc{ncores} \hlstd{=} \hlnum{3}\hlstd{)}
-\end{alltt}
-
-{\ttfamily\noindent\color{warningcolor}{\#\# Warning in VSURF.parallel.formula(linmod.PC1, data = focClimSA[, c("{}PC1"{}, : VSURF with a formula-type call outputs selected variables\\\#\# which are indices of the input matrix based on the formula:\\\#\# you may reorder these to get indices of the original data}}\begin{alltt}
-  \hlcom{# Plot most important climatic predictor versus PC1}
-  \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_bio16cva_v_Lat.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{5}\hlstd{,} \hlnum{5}\hlstd{)}
-  \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{))}
-  \hlkwd{plot}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0.25}\hlstd{,} \hlnum{0.55}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{),} \hlkwc{axes} \hlstd{= F,}
-    \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
-  \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Precipitation of Wettest Quarter"}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
-  \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"(Coefficient of variation)"}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{)}
-  \hlkwd{title}\hlstd{(}\hlkwc{ylab} \hlstd{=} \hlkwd{substitute}\hlstd{(}\hlkwd{plain}\hlstd{(Principal}\hlopt{~}\hlstd{Component}\hlopt{~}\hlnum{1}\hlstd{)}\hlopt{~}\hlkwd{group}\hlstd{(}\hlstr{"("}\hlstd{, x,} \hlstr{")"}\hlstd{),}
-    \hlkwd{list}\hlstd{(}\hlkwc{x} \hlstd{= pve)))}
-  \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-  \hlkwd{axis}\hlstd{(}\hlnum{2}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlopt{-}\hlnum{4}\hlstd{,} \hlnum{4}\hlstd{,} \hlnum{2}\hlstd{),} \hlkwc{las} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{)}
-  \hlstd{fitBio16cva} \hlkwb{<-} \hlkwd{lm}\hlstd{(PC1} \hlopt{~} \hlstd{bio16_cva,} \hlkwc{data} \hlstd{= focClim)}
-
-  \hlcom{# polygon of confidence intervals}
-  \hlstd{x} \hlkwb{<-} \hlkwd{seq}\hlstd{(}\hlkwd{min}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwd{max}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwc{l} \hlstd{=} \hlnum{1e3}\hlstd{)}
-  \hlstd{u} \hlkwb{<-} \hlkwd{predict}\hlstd{(fitBio16cva,} \hlkwc{new} \hlstd{=} \hlkwd{data.frame}\hlstd{(}\hlkwc{bio16_cva} \hlstd{= x))}
-  \hlstd{s} \hlkwb{<-} \hlkwd{sqrt}\hlstd{(}\hlkwd{sapply}\hlstd{(x,} \hlkwa{function}\hlstd{(}\hlkwc{X}\hlstd{)} \hlkwd{t}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X))} \hlopt{%*%} \hlkwd{vcov}\hlstd{(fitBio16cva)} \hlopt{%*%} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, X)))}
-  \hlstd{confInt} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwc{lowerCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.025}\hlstd{, u, s),} \hlkwc{upperCI} \hlstd{=} \hlkwd{qnorm}\hlstd{(}\hlnum{0.975}\hlstd{, u, s))}
-  \hlkwd{polygon}\hlstd{(}\hlkwd{c}\hlstd{(x,} \hlkwd{rev}\hlstd{(x)),} \hlkwd{c}\hlstd{(confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwd{rev}\hlstd{(confInt}\hlopt{$}\hlstd{upperCI)),}
-    \hlkwc{col} \hlstd{=} \hlstr{"grey80"}\hlstd{,} \hlkwc{border} \hlstd{=} \hlnum{NA}\hlstd{)}
-  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{lowerCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{lines}\hlstd{(x, confInt}\hlopt{$}\hlstd{upperCI,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{,} \hlkwc{lty} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{lines}\hlstd{(}\hlkwd{range}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva),} \hlkwd{range}\hlstd{(}\hlkwd{predict}\hlstd{(fitBio16cva)),} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{points}\hlstd{(focClim}\hlopt{$}\hlstd{bio16_cva, focClim}\hlopt{$}\hlstd{PC1,} \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{= primShade0,}
-    \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{2}\hlstd{)}
-  \hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\begin{alltt}
-  \hlcom{# Overlap between latitude predictors and phenotype predictors}
-\hlcom{#### NEED TO COMPARE RESULTS FROM POINT DATA AND SPATIALLY AVERAGED CLIMATE!}
-  \hlstd{imp2} \hlkwb{<-} \hlstd{climVars[vsurfPC1}\hlopt{$}\hlstd{varselect.pred]}
-  \hlcom{# imp2 <- climVars[vsurfPC1_SA$varselect.pred]}
-  \hlstd{nImpPC1} \hlkwb{<-} \hlkwd{length}\hlstd{(imp2)}
-
-  \hlstd{vsurfLat}\hlopt{$}\hlstd{varselect.pred}
-\end{alltt}
-\begin{verbatim}
-##  [1] 45 43 42 20 47 41 46 39 19 13
-\end{verbatim}
-\begin{alltt}
-  \hlstd{imp1} \hlkwb{<-} \hlstd{climVars[vsurfLat}\hlopt{$}\hlstd{varselect.pred]}
-  \hlstd{nImpLat} \hlkwb{<-} \hlkwd{length}\hlstd{(imp1)}
-
-  \hlcom{# Figure of variable importance for traits}
-        \hlkwd{pdf}\hlstd{(}\hlkwd{paste}\hlstd{(pathFig,} \hlstr{"/Figure_clim2pc1_imp.pdf"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),} \hlnum{4}\hlstd{,} \hlnum{5}\hlstd{)}
-        \hlkwd{par}\hlstd{(}\hlkwc{mar} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{5}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{6}\hlstd{))}
-
-        \hlkwd{plot}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0}\hlstd{,} \hlkwc{type} \hlstd{=} \hlstr{"n"}\hlstd{,} \hlkwc{xlim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.3}\hlstd{),} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{, nImpLat),}
-                \hlkwc{axes} \hlstd{= F,} \hlkwc{frame.plot} \hlstd{= T,} \hlkwc{xlab} \hlstd{=} \hlstr{""}\hlstd{,} \hlkwc{ylab} \hlstd{=} \hlstr{""}\hlstd{)}
-        \hlkwd{title}\hlstd{(}\hlkwc{xlab} \hlstd{=} \hlstr{"Climate-Trait PC1\textbackslash{}nImportance"}\hlstd{,} \hlkwc{line} \hlstd{=} \hlnum{4}\hlstd{,} \hlkwc{cex.lab} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlkwa{for} \hlstd{(i} \hlkwa{in} \hlstd{(nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{)))}
-        \hlstd{\{}
-                \hlkwd{lines}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{100}\hlstd{,} \hlnum{100}\hlstd{),} \hlkwc{y} \hlstd{=} \hlkwd{c}\hlstd{(i, i),} \hlkwc{col} \hlstd{=} \hlstr{"grey"}\hlstd{)}
-        \hlstd{\}}
-        \hlkwd{points}\hlstd{(vsurfPC1}\hlopt{$}\hlstd{ord.imp}\hlopt{$}\hlstd{x[}\hlnum{1}\hlopt{:}\hlstd{nImpPC1], nImpLat}\hlopt{:}\hlstd{(nImpLat} \hlopt{-} \hlstd{nImpPC1} \hlopt{+} \hlnum{1}\hlstd{),}
-                \hlkwc{pch} \hlstd{=} \hlnum{21}\hlstd{,} \hlkwc{col} \hlstd{=} \hlstr{"black"}\hlstd{,} \hlkwc{bg} \hlstd{=} \hlstr{"white"}\hlstd{,} \hlkwc{cex} \hlstd{=} \hlnum{1.5}\hlstd{)}
-        \hlcom{# mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, }
-  \hlcom{#  at = nImpLat:(nImpLat - nImpPC1 + 1), }
-        \hlcom{#	font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))}
-        \hlkwd{axis}\hlstd{(}\hlnum{1}\hlstd{,} \hlkwc{lwd} \hlstd{=} \hlnum{0}\hlstd{,} \hlkwc{lwd.ticks} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{at} \hlstd{=} \hlkwd{seq}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.3}\hlstd{,} \hlnum{0.1}\hlstd{))}
-  \hlkwd{dev.off}\hlstd{()}
-\end{alltt}
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
 For these analyses, we compiled a representative set of 178 recent (since 2000) known \textit{M.

### 2016-01-20 ms/ms.Rnw
Lots of changes.

@@
+lost on page 11 - extracted and used climate variables
+Maybe reordering stuff on paragraph
+cartoon of whole procedure for climate variables
+figure showing correlation b/t bio12 and 
+state that VSURF importance is different scales in different models
+moved up three hypotheses part to introduction
+play up lack of GxE and ???
+** correct # of Pop x Treatment
+
+* delve into interactions - do they change cline?
+
+* show figures of reaction norms
+* split plot, account for error variation in temp???
+* show PAR similarity
+* what CO2 concentration did I use?
+* some sites they are functionally annual
+
+* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
+	- maybe temporal source-sink makes more sense
+
+* check out matthew's blog on hydrology variables
+
+* explain rhizomes
+
+* put in biomass allocation cline
+% TO DO:
+% 1. Is water regime on figure 1 correct?
+% 2. Finalize Table 2
+% 3. Do anything with biomass data?
+% 4. Citations for latitudinal clines
+	% cases where clinal variation has 'obvious' selective source:
+	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
+	% Caicedo et al 2004?
+	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
+% 5. Citations for GxE as signature of local adaptation
+% 6. Fix figure 2
+% 7. Supplemental Figure and Table captions
+% 8. Supplemental figure showing agreement between assigned to climatic variables from point estimated and spatially averaged analyses
+
 \documentclass[11pt, oneside]{article}
@@
 \usepackage{lineno}
+\usepackage{textcomp}
 
 % Make helvetica the default sans-serif font
-\renewcommand\sfdefault{phv}
+% \renewcommand\sfdefault{phv}
 
@@
 
-<<Front Matter, echo=FALSE, results='hide'>>=
+<<Front Matter, echo=FALSE, results=hide>>=
 
@@
 \begin{document}
+\SweaveOpts{concordance=TRUE}
 \maketitle
 
-$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada
-
-$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca
-
-Original Article
-
-Key words: local adaptation, photosynthesis, growth rate, climatic variance, \textit{Mimulus}
-
-Word counts: TBA
-
-3 figures; 3 tables; ? supporting files
-
-Data will be archived on Dryad upon acceptance.
+$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada \\
+$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca \\
+
+Original Article \\
+
+Key words: local adaptation, cline, photosynthesis, growth rate, \textit{Mimulus} \\
+
+Word counts: TBA \\
+3 figures; 3 tables; ? supporting files \\
+Data will be archived on Dryad upon acceptance. \\
+
+\cdm{Chris' comments}\\
+\ala{Amy's comments}
 
@@
 \section*{Abstract}
+
+Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
+Correlation between traits and latitude, or latitudinal clines, are among the most common pieces of evidence for local adaptation, but identifying the traits under selection and the selective agents are challenging.
+Here, we investigated a latitudinal cline in growth and photosynthesis across 16 populations of the perennial herb \textit{Mimulus cardinalis} (Phrymaceae).
+Using machine learning methods, we identify interannual variation in precipitation as a likely selective agent: Southern populations from more variable environments had higher photosynthetic rates and grew faster.
+We hypothesize that selection may favor a more annualized life history -- grow now rather than save for next year -- in environments where severe droughts occur more often.
+Thus our study provides insight into how species may adapt if Meditteranean climates become more variable due to climate change.
 
@@
 
-Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
-Local adaptation within species most often involves genetic differences in ecologically relevant traits caused by selection to different environments, leading to predictable trait-environment correlations that persist in a common environment.
-Although local adaptation is well-supported by data from reciprocal transplant experiments, trait-environment associations, and ecological genetics/genomics, predicting which traits respond to selection is beyond our current understanding.
-In particular, the role of physiological traits is poorly understood, even though physiology explains adaptation to the abiotic environment, which we need to understand as a baseline for anticipating how organisms will respond to climate change.
+% Local adaptation and clines
+Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}.
+Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}.
+Phenotypic differences between populations along the cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
+Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
+In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment.
+We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change.
 A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures.
-Here, we  investigate which traits and causal selective agents explain range-wide patterns of local physiological adaptation to climate in \textit{Mimulus}, a model genus for local adaptation studies.
-
-% niche-tracking versus niche-shifting
-One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve.
-For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.
+Here, we examine physiological trait variation and possible selective agents in the perennail herb \textit{Mimulus cardinalis} Douglas ex Benth.
+(Phrymaceae), a model system for local adaptation studies.
+
+% Intrinsic versus plastic variation
+There are two basic approaches one can use to identify candidate traits underlying local adaptation in a common garden.
+First, if genetically-based trait differences between populations vary clinally with environmental differences, this may point to traits important for local adaptation.
+Second, genotype by environment interactions could indicate that variation in plasticity mediates local adaptation.
+We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively.
+There are classic cases adaptation involving both intrinsic and plastic trait variation.
+For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
+Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
 
 % Selective agents
-We also know little about the selective agents, such as climatic factors or biotic interactions, that are most important in driving adaptive divergence.
-In certain situations, such as pollinator shifts and edaphic adaptation, it is straightforward to identify putative selective agents and test predicted mechanisms.
-However, it is more challenging for cases in which traits vary with less specific factors like mean annual precipitation or latitude.
-For example, body size clines (larger size at higher latitude) have evolved repeatedly on three contients in \textit{Drosophila subobscura}, strongly indicating selection \citep{Huey_etal_2000}.
-But which one or more of the many climatic factors that covary with latitude actually controls selection on body size?
-Selection might be even more complex for perennial species that live over multiple seasons, meaning that environmental variation rather than the average environment may be a selective agent.
-
-% Motivate using Mimulus
-Linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions.
-Many classic and contemporary studies of local adaptation have been conducted using species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources [willis, lowry?].
-Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008}.
-
-Two common signatures of local adaptation are nonrandom genetic differences in ecologically important traits and variation in plasticity (genotype by environment interaction).
-In this study, we found strong genetically-based differences in photosynthesis and growth, but no evidence for variation in plasticity in response to temperature and drought treatments.
-Interannual variation in precipitaiton is associated intrinsic variation in these traits, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
+Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
+For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}.
+Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely related to climatic variation.
+Despite the fact that latitudinal clines in particular have been studied for a long time, the exact climatic factors (e.g.
+minimum or maximum temperature, growing season length, strength of biotic interactions) is rarely known.
+Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g.
+coastal vs.
+continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}.
+In plants especially, we know little about the prevelance and adaptive significance of clinal variation in fundamental physiological traits like photosynthesis and the impact on plant performance.
+
+% For the most part, we know much more about selective agents responsible for clines for conspicuous things like colouration, such as in mice \citep{Sumner_1932, Mullen_Hoekstra_2008}
+
+% Motivate primary questions using Mimulus
+% might add that Anacker and Strauss 2014 showed ecological shifts important in Mim speciation and we would like to know driving forces behind this???
+In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}.
+We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions.
+Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
+Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}.
+We measured genetic and genotype by environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
+We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity.
+Interannual variation in precipitaiton and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
 cardinalis}.
+We place these findings in the context of life history theory and consider future directions in the Discussion.
 
@@
 
-We used 16 populations from throughout the range of \textit{M. cardinalis} (Table~\ref{table:Table_FocalPops}). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
-
-<<focClim, echo = FALSE, results='hide', warning=FALSE>>=
+We used 16 populations from throughout the range of \textit{M.
+cardinalis} (Table~\ref{table:Table_FocalPops}).
+Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
+
+<<focClim, echo = FALSE, results=hide, warning=FALSE>>=
 # Read in climate data for 16 focal populations to use in making Table 1
-# [MAYBE SHOULD RENAMED "ExtendedTable1" or something?]
 load(paste(pathDat, "/focClim.RData", sep = ""))
@@
    \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
-   \begin{tabular}{@{} lllllll @{}}
+   \begin{tabular}{@{} lllll @{}}
       \toprule
-  Name& Region  & Demo?& Pop gen?& Latitude  & Longtiude  & Elevation (mas) \\
+  Name& Region  & Latitude  & Longtiude  & Elevation (mas) \\
       \midrule
-	HAU & South Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "HAU"], 3)}	& 
+	HAU & South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "HAU"], 3)}	& 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "HAU"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "HAU"]}   \\
-	CTC	& South Margin		& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "CTC"], 3)} & 
+	CTC	& South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "CTC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "CTC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "CTC"]}   \\
-	CUR	& South Margin		& yes	& yes & \Sexpr{round(focClim$Latitude[focClim$PopID == "CUR"], 3)} & 
+	CUR	& South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "CUR"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "CUR"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "CUR"]}   \\
-	GRP & South Margin		& no	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "GRP"], 3)} &
+	GRP & South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "GRP"], 3)} &
     \Sexpr{round(focClim$Longitude[focClim$PopID == "GRP"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "GRP"]}   \\
-	WWC &	Transverse		  & yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "WWC"], 3)} & 
+	WWC &	Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "WWC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WWC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WWC"]}   \\
-	MIL	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MIL"], 3)} & 
+	MIL	& Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "MIL"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "MIL"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "MIL"]}   \\
-	WFM	& Transverse	  	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "WFM"], 3)} & 
+	WFM	& Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "WFM"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WFM"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WFM"]}   \\
-	NMT	& South Sierras		& yes	& yes?& \Sexpr{round(focClim$Latitude[focClim$PopID == "NMT"], 3)} & 
+	NMT	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "NMT"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "NMT"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "NMT"]}   \\
-	PRD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "PRD"], 3)} & 
+	PRD	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "PRD"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "PRD"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "PRD"]}   \\
-	RWD	& South Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RWD"], 3)} & 
+	RWD	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "RWD"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RWD"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RWD"]}   \\
-	WNA	& Central Sierras	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "WNA"], 3)} & 
+	WNA	& Central Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "WNA"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WNA"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WNA"]}   \\
-	RBW	& Central Sierras	& yes	& no	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RBW"], 3)} & 
+	RBW	& Central Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RBW"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RBW"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RBW"]}   \\
-	MYU	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MYU"], 3)} & 
+	MYU	& North Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MYU"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "MYU"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "MYU"]}   \\
-	LIJ	& North Sierras		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "LIJ"], 3)} & 
+	LIJ	& North Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "LIJ"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "LIJ"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "LIJ"]}   \\
-	DPC	& North Coast	  	& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "DPC"], 3)} & 
+	DPC	& North Coast & \Sexpr{round(focClim$Latitude[focClim$PopID == "DPC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "DPC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "DPC"]}   \\
-	RCC	& North Margin		& yes	& yes	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RCC"], 3)} & 
+	RCC	& North Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "RCC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RCC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RCC"]}   \\
@@
 We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments.
-The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments (Table S\#).
+The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments Table~\ref{table:TableS_InitialSize}.
 
@@
 
-We imposed four treatments, a fully-factorial cross of two temperature treatments and two watering treatments.
-Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment and another to the Cool treatment.
+We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels.
+The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which designate as Hot and Cool treatments.
+Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments.
+A detailed description of treatments is described in the Supplemental Information and summarized in Fig~\ref{fig:Fig_ExptlDes}.
+Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level.
 Within each chamber, there were two Well-watered blocks and two Drought blocks.
-A detailed description of treatments is given below \cdm{should some be moved to supplement?} and summarized in Fig~\ref{fig:Fig_ExptlDes}.
 The irradiance in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$.
-The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature treatments (data not shown).
-
-\subsubsection*{Temperature treatments}
-
-We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool).
-We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
-Daily temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
-Angert, unpub.
-data).
-Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M.
-cardinalis} in nature.
-We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}.
-The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ).
-The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
-We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R.
-Simulated data closely resembled the autocorrelation and between-site correlation of the actual data.
-From simulated mean temperature, we next selected minimum and maximum daily temperatures.
-Mean, min, and max temperature were highly correlated at both sites.
-We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models.
-For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
-During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
-
-\subsubsection*{Watering treatments}
-
-For watering treatments, we simulated two extreme types of streams where \textit{M.
-cardinalis} grows.
-In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
-In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down as the winter snow pack melts.
-In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER.
-Plants in the well-watered treatment were fully saturated every two hours during the day.
-Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing).
-Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.
-
-\cdm{Are there any data from California about soil water content through the season?}
-
-<<ExperimentalDesign, echo=FALSE, results='hide', warning=FALSE>>=
+The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
+
+<<ExperimentalDesign, echo=FALSE, results=hide, warning=FALSE>>=
 library(scales) # for color transparency
@@
   Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
-  Mortality &  Probability?  \\
+  Mortality &  \cdm{Probability?}  \\
 	    \bottomrule
@@
 
-<<Germination rate, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
+<<Germination rate, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
 
-We censused longest leaf length $>1$ mm on 10 days (twice per week) between 12 May and 12 June (28 to 59 days after sowing).
+We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
 We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.
@@
 
-<<Leaf expansion rate, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
+<<Leaf expansion rate, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
   # Mixed model ANOVA with family as random factor fit using step down procedure
-	mm <- lmer(LER_AbsGrowth ~ AvgGermDay + PopID * TempTrt * WaterTrt + (1|Family),
-		data = master)
+	master$Chamber <- substr(master$Block, 1, 3)
+	mm <- lmer(LER_AbsGrowth ~ AvgGermDay + PopID * TempTrt * WaterTrt + 
+						 	(1|Family) + (1|Chamber/Block), data = master)
 	fitLER <- step(mm, reduce.random = F) # includes pop, temp, h2o, population x water
@@
 
-<<Shoot elongation rate, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
+<<Shoot elongation rate, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
   mm <- lmer(SER_AbsGrowth ~ AvgGermDay + PopID * TempTrt * WaterTrt + 
-		(1|Family), data = master)
-	fitSER <- step(mm, reduce.random = F) #includes pop, temp, h2o, temp x water
+		(1|Family) + (1|Block), data = master)
+	fitSER <- step(mm, reduce.random = F) #includes pop, temp
 @
@@
 During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations.
-The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
+The youngest, fully-expanded leaf \cdm{this is what I did...not sure exactly which node, as I didn't record that when I did measurements} acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
 All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity.
-During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ \cdm{get exact temps used}) so that all plants within a temperature treatment were measured under the same conditions.
-
-<<Photosynthesis, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
+During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.
+
+<<Photosynthesis, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
   # Photosynthetic rate
-  mm <- lmer(Photo ~ PopID * TempTrt * WaterTrt + (1|Family),
+  mm <- lmer(Photo ~ PopID * TempTrt * WaterTrt + (1|Family) + (1|Block),
     data = subset(master, !is.na(master$Photo)))
@@
   # Removed one outlier (may change if data are rearranged!)
-	mm <- lmer(resPhoto ~ PopID * TempTrt * WaterTrt + (1|Family),
+	mm <- lmer(resPhoto ~ PopID * TempTrt * WaterTrt + (1|Family) + (1|Block),
     data = subset(master, !is.na(master$Photo))[-52, ])
@@
 
-<<Mortality, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
+<<Mortality, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
   fitMort <- glm(DiedFromStress ~ PopID * TempTrt * WaterTrt,
-    data = subset(master, !is.na(master$DiedFromStress)), family = "quasibinomial")
-
+    data = subset(master, !is.na(master$DiedFromStress)), 
+    family = "quasibinomial")
+
+  fitMort <- glmer(DiedFromStress ~ PopID * TempTrt * WaterTrt + (1|Family) +
+  								 (1|Block), data = subset(master, !is.na(master$DiedFromStress)), family = "binomial")
   Anova(fitMort, type = 2)
@@
 
-\paragraph{Biomass (show correlation between growth rate and biomass?)}
+\paragraph{Biomass at harvest}
+
+\cdm{show correlation between growth rate and biomass?}
 
@@
 \subsection*{Principal components of germination, growth, and photosynthesis}
-We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components.
+For each single-trait model above, we extracted the Population coefficient (factoring out Treatment and other effects).
+The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA).
 The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast and slow growing populations.
 
-<<Principal Components, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
+<<Principal Components, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
   s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitLat) %*% c(1, X)))
-  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), upperCI = qnorm(0.975, u, s))
+  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
+  											upperCI = qnorm(0.975, u, s))
   polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
-    col = "grey80", border = NA)
+    col = "white", border = NA)
   lines(x, confInt$lowerCI, lwd = 2, lty = 2)
@@
   lines(range(focClim$Lat), rev(range(predict(fitLat))), lwd = 2)
-  points(focClim$Lat, focClim$PC1, pch = 21, col = "black", bg = primShade0, cex = 1.5,
-    lwd = 2)
+  points(focClim$Lat, focClim$PC1, pch = 21, col = "black", bg = "white", 
+  			 cex = 1.5, lwd = 2)
   dev.off()
@@
 
-<<thinnedOccClimate, echo=FALSE, results='hide', warning=FALSE>>=
+<<thinnedOccClimate, echo=FALSE, results=hide, warning=FALSE>>=
 # R data.frame of 178 M. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data
@@
 
-As with previous studies in \textit{Mimulus} and many other species [CITES, \citep{Huey_etal}, mosquitoes, what else?], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
-temperature or precipitation) that covary with latitude.
-We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results).
+Since latitude \textit{per se} cannot be a selective agent, we considered three possible explanations.
+First, latitude may be strongly correlated, in this species, with one or two climatic variables, such as temperature, precipitation, or growing degree-days.
+Second, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latiduinal cline.
+Third, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
+For example, average temperature of \textit{M.
+cardinalis} populations at a given latitude varies widely, but in aggregate, a Southern metapopulation experiences warmer climate than a Northern one.
+Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g.
+high elevation) site.
+
+To evaluate support for these hypotheses, we used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
 cardinalis}.
-We did this by looking for overlap between climatic variables that best predict latitude of \textit{M.
-cadinalis} occurence records and climatic variables that best predict trait variation across our 16 focal populations.
+We looked for overlap between climatic variables that best predict latitude of \textit{M.
+cadinalis} occurence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
 For brevity, we refer to these as Climate-Latitude and Climate-Trait variables.
 We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}).
-From VSURF models, we kept only variables selected for prediction, the most stringent criteria.
-
-<<VSURF, echo=FALSE, results='hide', warning=FALSE, message=FALSE>>=
+From VSURF models, we kept only variables selected for prediction, the most stringent criterion.
+
+The first hypothesis predicts that there should be one or two Climate-Latitude and Climate-Trait associations that are strongly correlated, whereas the second hypothesis predicts that several climatic variables should be weakly correlated.
+To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-Trait variables as above, except that we used spatially averaged climate variables.
+We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations.
+We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
+Since \textit{M.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
+For clarity, we distinguish between analyses where climate is inferred from a single point (`point estimated Climate-Trait') versus averaged across a 100-km buffer (`spatially averaged Climate-Trait').
+
+<<VSURF, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
   linmod.PC1 <- as.formula(paste("PC1 ~ ", paste(climVars, collapse = " + "), sep = ""))
-  set.seed(69647)
-  vsurfPC1 <- VSURF.parallel(linmod.PC1, data = focClim[, c("PC1", climVars)], 
-    ncores = 3)
-  vsurfPC1_SA <- VSURF.parallel(linmod.PC1, data = focClimSA[, c("PC1", climVars)], 
-    ncores = 3)
+  set.seed(69648)
+  # vsurfPC1 <- VSURF(linmod.PC1, data = focClim[, c("PC1", climVars)])
+  vsurfPC1 <- VSURF(focClim[, climVars], focClim[, "PC1"])
+  # vsurfPC1_SA <- VSURF(linmod.PC1, data = focClimSA[, c("PC1", climVars)])
+  vsurfPC1_SA <- VSURF(focClimSA[, climVars], focClimSA[, "PC1"])
+
+##### WORKING ON STUFF HERE #####
+vsurfPC1_SA$imp.mean.dec
+x <- match(vsurfPC1$imp.mean.dec.ind, vsurfPC1_SA$imp.mean.dec.ind)
+plot(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], type = "n")
+text(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], 
+		 labels = climVars[vsurfPC1$imp.mean.dec.ind])
+cor.test(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x])
 
@@
 
-  # Overlap between latitude predictors and phenotype predictors
-#### NEED TO COMPARE RESULTS FROM POINT DATA AND SPATIALLY AVERAGED CLIMATE!
+  # Overlap between latitude predictors and phenotype predictors (point measurements)
   imp2 <- climVars[vsurfPC1$varselect.pred]
-  # imp2 <- climVars[vsurfPC1_SA$varselect.pred]
   nImpPC1 <- length(imp2)
 
-  vsurfLat$varselect.pred
   imp1 <- climVars[vsurfLat$varselect.pred]
@@
 
-  # Figure of variable importance for traits
+  # Figure of variable importance for traits (point measurements)
 	pdf(paste(pathFig, "/Figure_clim2pc1_imp.pdf", sep = ""), 4, 5)
@@
 	}
-	points(vsurfPC1$ord.imp$x[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), 
+	points(vsurfPC1$imp.mean.dec[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), 
 		pch = 21, col = "black", bg = "white", cex = 1.5)
@@
   dev.off()
+
+  # Overlap between latitude predictors and phenotype predictors (spatial aveage)
+  imp3 <- climVars[vsurfPC1_SA$varselect.pred]
+  nImpPC1_SA <- length(imp3)
+
+  imp1 <- climVars[vsurfLat$varselect.pred]
+  nImpLat <- length(imp1)
+
+  # Figure of variable importance for traits (point measurements)
+	pdf(paste(pathFig, "/Figure_clim2pc1SA_imp.pdf", sep = ""), 4, 5)
+	par(mar = c(5, 1, 1, 6))
+
+	plot(0, 0, type = "n", xlim = c(0, 0.3), ylim = c(1, nImpLat), 
+		axes = F, frame.plot = T, xlab = "", ylab = "")
+	title(xlab = "Climate-Trait PC1\nImportance", line = 4, cex.lab = 1.5)
+	for (i in (nImpLat:(nImpLat - nImpPC1 + 1))) 
+	{
+		lines(x = c(-100, 100), y = c(i, i), col = "grey")
+	}
+	points(vsurfPC1$imp.mean.dec[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), 
+		pch = 21, col = "black", bg = "white", cex = 1.5)
+	# mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, 
+  #  at = nImpLat:(nImpLat - nImpPC1 + 1), 
+	#	font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
+	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.3, 0.1))
+  dev.off()
+
+# Table of important climatic variables
+
+clim2sym <- read.csv(paste(pathMS, "/climVar2plot.csv", sep = ""), 
+										 stringsAsFactors = F, row.names = 1)
+readLines(paste(pathMS, "/climVar2plot.csv", sep = ""))
+
+tableS <- file(paste(pathMS, "/Tables/TableS_ClimVarImp.txt", sep = ""), open = "w")
+for (i in 1:max(c(nImpLat, nImpPC1, nImpPC1_SA))) {
+clim2sym[imp1[i], "name"]
+	# Latitude variable
+	if (nImpLat >= i) {
+		cat(sprintf('%s ($\\%s$) & ', clim2sym[imp1[i], "name"], clim2sym[imp1[i], "what"]), file = tableS)
+	}else{
+		cat(' & ', file = tableS)
+	}
+
+	# Point estimated trait variable
+	if (nImpPC1 >= i) {
+		cat(sprintf('%s ($\\%s$) & ', clim2sym[imp2[i], "name"], clim2sym[imp2[i], "what"]), file = tableS)
+	}else{
+		cat(' & ', file = tableS)		
+	}
+
+	# Spatially averaged trait variable
+	if (nImpPC1_SA >= i) {
+		cat(sprintf('%s ($\\%s$) \\\\\n ', clim2sym[imp3[i], "name"], clim2sym[imp3[i], "what"]), file = tableS)
+	}else{
+		cat(' \\\\\n', file = tableS)		
+	}
+}
+close(tableS)
+
 @
@@
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included \Sexpr{length(climVars) / 2} cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+In total, we included \Sexpr{length(climVars) / 2} climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
 For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
@@
 
-Genetic differences between populations affecting originsmal function and performance point to traits that may be involved in local adaptation.
-Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis} (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured \Sexpr{pve}$\%$ of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
+Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
+cardinalis}, as evidenced by large and highly significant population effects (Table~\ref{table:Table_AnovaSummary}).
+A single principal component captured \Sexpr{gsub("%", "", pve)}\% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
 As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
@@
 Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods).
-The line and 95\% confidence intervals (grey polygon) were estimated using standard linear regression.}
+The line and 95\% confidence intervals were estimated using standard linear regression.}
 	\label{fig:Fig_PC1vLat}
@@
 
-Genotype $\times$ environment interactions are also a common signature of local adaptation [CITE].
+Genotype $\times$ environment interactions are also a common signature of local adaptation.
 For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
@@
 For \textit{M.
-cardinalis}, interannual variation in precipitation over the past 30 years is very closely related to the latitude of recently-recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
-The two most important Climate-Latitude variables were also strong predictors of position along the fast-slow growth axis (Fig.~\ref{fig:Fig_ClimVarImp}B).
-This overlap between Climate-Latitude and Climate-Trait variables suggests that interannual variation in precipitation is an important selective agent in \textit{M.
+cardinalis}, interannual variation in precipitation over the past 30 years is very closely related to the latitude of recently recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
+Between year variation in precipitation was also strongly correlated with position along the fast-slow growth axis, but climate based on point estimates gave somewhat different results than climate based on spatially averaged values.
+The two most important point estimated Climate-Latitude variables were also strong predictors of position along the fast-slow growth axis (Fig.~\ref{fig:Fig_ClimVarImp}B).
+For spatially averaged climate, interannual variation in precipitation seasonality was the only climatic variable selected, (Fig.~\ref{table:TableS_ClimVarImp}) \cdm{Do we need a figure for this?}.
+Overall, there was good but imperfect agreement between the weight assigned to climatic variables using point estimated and spatially averaged climate (Fig.~\ref{fig:Fig_TBD}) \cdm{This figure will show strong, positive correlation between weights given to climatic variables from point estimated and spatially averaged analyses}.
+
+Overlap between Climate-Latitude and point estimated Climate-Trait variables suggests that interannual variation in precipitation is an important selective agent in \textit{M.
 cardinalis}.
 Specifically, we hypothesize that more frequent droughts (greater precipitation cv) in Southern populations selects for an `annual-ized' life history, as we detail in the Discussion.
+Even though the spatially averaged Climate-Trait variable was not among the Climate-Latitude variables, it points to a similar conclusion.
+Specifically, this analysis showed that greater interannual variation in precipitation seasonality, which is also driven by precipitation variability (i.e.
+more frequent drought years) in Southern California, is associated with faster growth in southern populations.
 We must qualify these results because our analysis obviously cannot rule out that alternative variables not included in the analysis may be more important.
@@
 Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
-Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp) or Precipitation (Prec); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Dry Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
+Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp) or Precipitation (Prec); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
 	\label{fig:Fig_ClimVarImp}
@@
 
+% perhaps responses of Simiolus sp (annual, drought selects for escape rather than tolerance) differ from card (perennial, escape through growth rather than flowering time) because drying out of rivers is less severe than seeps and springs?
+
 \section*{Discussion}
 
-The observation that populations are locally adapted to their environment is old and well-documented in many species.
+% The observation that populations are locally adapted to their environment is old and well-documented in many species.
 New frontiers in local adaptation research seek a general explanation for why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
 Understanding why and how populations have adapted to different environments in the past is especially significant now as a scientific basis for predicting how species' ranges will shift under climate change [CITES, \citep{Catullo_etal_2015}].
-In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint.
-In particular, our data suggest that photosynthesis and intrinsic growth rate vary among populations, whereas the fundamental abiotic niche is more conserved, and that climatic variation may be a more important selective agent than the average.
-Next, we tie these results into the broader threads of evolutionary theory in order to figure out what factors explain the trait variation we see and that we do not.
-
-As long as environments vary and there exists genetic variation in fitness, why shouldn't local adapted genotypes evolve?
-Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001}.
-Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strong asymmetric, and/or maladaptive gene flow is too high.
+In this study, we found evidence for one of two common signatures of local adaptation.
+A latitudinal cline in traits such as germination rate, photosynthesis, and growth, suggests adaptive differentiation in fundamental physiological traits of the species.
+However, we found little evidence that populations respond differently to temperature or drought.
+As we discuss below, this may indicate that the fundamental abiotic niche is relatively conserved \ala{the results that underlie this statement need more emphasis in the methods}.
+Finally, we found that climatic variation between years may be a more important selective agent than the average climate.
+In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in photosynthesis and growth varies clinally, but plastic responses to temperature and drought are relatively conserved.
+
+Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001} and hence the type of variation maintained within species.
+Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strongly asymmetric, and/or maladaptive gene flow is too high.
 In \textit{M.
@@
 That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatial heterogenous selection.
-We also know from previous work that population size does not vary strongly with latitude \cdm{Amy - is this correct?
-JPs pop gen data seemed to suggest little difference in effective pop size across the range.
-Is this also true of census pop size?}.
-Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range.
+We also know from previous work that population size does not vary strongly with latitude.
+Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range \citep{Paul_etal_2016}.
 
 Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years (see next section), could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
-We found no evidence for this; all populations responded to drought and temperature treatments similarly (Table~\ref{table:Table_AnovaSummary}).
+We found no evidence for this; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
 We hypothesize that evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
@@
 cardinalis} live are highly heterogeneous at small spatial scales.
-Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Meditteranean climates of Western North America.
+Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Meditteranean climates of western North America.
 But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and vice versa, leading to strong fitness tradeoffs.
 Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
-Demography and gene flow may reinforce niche conservatism.
+Demography and gene flow may reinforce niche conservatism \ala{also bring in idea of temporal source sink: most individuals made in wet years, even if infrequent, so selection weighted towards wet environment even if dry states are frequent}.
 A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher resource environments.
 Finally, gene flow, which is generally high among \textit{M.
-cardinalis} populations within the same ecoregion \citep{Paul_etal_2015}, will thwart local adaptation and reinforce specialization.
+cardinalis} populations within the same ecoregion \citep{Paul_etal_2016}, will thwart local adaptation and reinforce specialization.
 Thus, the spatial grain of the environment, demographic asymmetry, and gene flow may conspire to constrain local adaptation via altered fundamental niche.
 
-Based on the available data, interannual variation in annual or winter precipitation (these are closely correlated in Mediterranean climates) may be the selective agent driving variation in growth and photosynthesis.
+Based on the available data, interannual variation in annual or winter precipitation or precipitaiton seasonality (these are closely correlated in Mediterranean climates) may be the selective agent driving variation in growth and photosynthesis.
 Variation in precipitation was best predicted latitude of recent \textit{M.
@@
 
-[Future work and broad conclusions].
+\cdm{I am still working a paragraph or two linking these results to empirical results in other systems and climatic adaptaiton in Mimulus sect Simiolus.}
+
+% [Future work and broad conclusions].
 Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
@@
 
+% niche-tracking versus niche-shifting
+% One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve.
+For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.
+
+% citations on life history allocation tradeoff between growth and reproduction:
+%% THEORY:
+% Bell 1980 - briefly mentionds on page 56 that increased environmental variation for adults will favor semelparity
+% Iwasa and Cohen 1989, pg.
+496 "A larger reliability of habitat, a higher storage efficiency, higher productivity of the environment, and a longer growing season, all tend to promote the evolution of perennial life, as suggested by previous models of life-historsy strategy (Stearns 1976).
+
+%% EMPIRICAL:
+% Sawada et al. 1994 - latitudinal cline dry matter accumulation and storage allocation
+% Stanton et al 2000 show selection in stressful environemts (artificual selection over 5 generations) favors stress avoidance rather than tolerance traits.
+i.e.
+might be easier to adjust life history rather than tolerance.
+% Hall and Willis 2006 find selection on flowering time in coastal-inland comparison of M guttatus
+% Ehleringer et al 1990 find lower WUE increased yield in ephemeral pops
+% Bazzaz 1979: higher photo in fast growing plants (early succession)
+
+% Reznick 1985; Lovett Doust 1989
+% Williams 1966; Charnov and Schaffer 1973; Bell 1980; Reznick 1985; Biere 1995; Johnson 2007).
+% for iteroparous sp: van Noordwijk and De Jong 1986.
+% from Remington et al.
+2013: "Genetic variation in traits subject to trade-offs results in what has been termed “structured pleiotropy,” in which genetic covariances between traits result from functional constraints imposed by the limiting resources (De Jong 1990; Stearns et al.
+1991; see Figure 1B).
+
 %--------------------------------------------------
@@
 
-\bibliography{refs}
+\bibliography{../../CommonResources/refs}
 \bibliographystyle{evolution}
@@
 	\end{center}
-	\label{tables:TableS_InitialSize}
+	\label{table:TableS_InitialSize}
 \end{table}
@@
 
+% Table of Climate-Latitude and Climate-Trait variables
+
+\begin{table}[ht]
+   \centering
+   \topcaption[Important climatic variables]{Important climatic variables predicting latitude of \textit{M.
+cardinalis} populations (`Climate-Latitude') and the first principal component of traits measured in a common garden (`Climate-Trait').
+Importance and significance were determined using the variable selection using random forests (VSURF) algorithm (see `Methods').
+Climatic variables are described in Table~\ref{table:TableS_ClimVars}.
+$\mu$ signifies the mean of the climate variables from 1981--2010; $\sigma$ indicates coeffiecient of variation among years}
+   \resizebox{6.5in}{!}{
+   \begin{tabular}{@{} lll @{}}
+      \toprule
+    	Climate-Latitude variables & Climate-Trait variables & \\
+                            	   & Point estimated         & Spatially averaged \\
+      \midrule
+			\input{Tables/TableS_ClimVarImp.txt}
+			\bottomrule
+   \end{tabular} }
+   \label{table:TableS_ClimVarImp}
+\end{table}
+
 % Figures showing latitudinal clines with specific traits
@@
 
-\subsection*{Spatially-averaged climate}
-
-% Methods for spatial-average selection
-Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
-We therefore calculated spatially averaged climate means and coefficients of variation.
-We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
+\section*{}
+
+\subsection*{Temperature treatments}
+
+We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool).
+We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
+This range was chosen because seeds used in the experiment were collected around 2012, thus their presence in that location at that time suggests that populations were able to persist there for at least some years before collection.
+Monthly temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
+Angert, unpub.
+data).
+Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M.
+cardinalis} in nature.
+We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}.
+The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ).
+The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
+We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R.
+Simulated data closely resembled the autocorrelation and between-site correlation of the actual data.
+From simulated mean temperature, we next selected minimum and maximum daily temperatures.
+Mean, min, and max temperature were highly correlated at both sites.
+We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models.
+For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
+During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
+
+\subsection*{Watering treatments}
+
+For watering treatments, we simulated two extreme types of streams where \textit{M.
+cardinalis} grows.
+In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
+In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down through the summer.
+In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER.
+Plants in the well-watered treatment were fully saturated every two hours during the day.
+Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing).
+Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.
 

### 2016-01-20 ms/ms.tex
Lots of changes.

@@
-\documentclass[11pt, oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
+% TO DO:
+% 1. Is water regime on figure 1 correct?
+% 2. Finalize Table 2
+% 3. Do anything with biomass data?
+% 4. Citations for latitudinal clines
+	% cases where clinal variation has 'obvious' selective source:
+	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
+	% Caicedo et al 2004?
+	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
+% 5. Citations for GxE as signature of local adaptation
+% 6. Fix figure 2
+% 7. Supplemental Figure and Table captions
+% 8. Supplemental figure showing agreement between assigned to climatic variables from point estimated and spatially averaged analyses
 
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
+\documentclass[11pt, oneside]{article}
 
@@
 %
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
+\usepackage{Sweave}
 \begin{document}
+\input{ms-concordance}
 \maketitle
 
-$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada
+$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada \\
+$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca \\
 
-$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca
+Original Article \\
 
-Original Article
+Key words: local adaptation, cline, photosynthesis, growth rate, \textit{Mimulus} \\
 
-Key words: local adaptation, photosynthesis, growth rate, climatic variance, \textit{Mimulus}
+Word counts: TBA \\
+3 figures; 3 tables; ? supporting files \\
+Data will be archived on Dryad upon acceptance. \\
 
-Word counts: TBA
-
-3 figures; 3 tables; ? supporting files
-
-Data will be archived on Dryad upon acceptance.
+\cdm{Chris' comments}\\
+\ala{Amy's comments}
 
@@
 
-Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
-Local adaptation within species most often involves genetic differences in ecologically relevant traits caused by selection to different environments, leading to predictable trait-environment correlations that persist in a common environment.
-Although local adaptation is well-supported by data from reciprocal transplant experiments, trait-environment associations, and ecological genetics/genomics, predicting which traits respond to selection is beyond our current understanding.
-In particular, the role of physiological traits is poorly understood, even though physiology explains adaptation to the abiotic environment, which we need to understand as a baseline for anticipating how organisms will respond to climate change.
+% Local adaptation and clines
+Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}.
+Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}.
+Phenotypic differences between populations along the cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
+Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
+In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment.
+We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change.
 A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures.
-Here, we  investigate which traits and causal selective agents explain range-wide patterns of local physiological adaptation to climate in \textit{Mimulus}, a model genus for local adaptation studies.
+Here, we examine physiological trait variation and possible selective agents in the perennail herb \textit{Mimulus cardinalis} Douglas ex Benth.
+(Phrymaceae), a model system for local adaptation studies.
 
-% niche-tracking versus niche-shifting
-One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve.
-For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.
+% Intrinsic versus plastic variation
+There are two basic approaches one can use to identify candidate traits underlying local adaptation in a common garden.
+First, if genetically-based trait differences between populations vary clinally with environmental differences, this may point to traits important for local adaptation.
+Second, genotype by environment interactions could indicate that variation in plasticity mediates local adaptation.
+We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively.
+There are classic cases adaptation involving both intrinsic and plastic trait variation.
+For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
+Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
 
 % Selective agents
-We also know little about the selective agents, such as climatic factors or biotic interactions, that are most important in driving adaptive divergence.
-In certain situations, such as pollinator shifts and edaphic adaptation, it is straightforward to identify putative selective agents and test predicted mechanisms.
-However, it is more challenging for cases in which traits vary with less specific factors like mean annual precipitation or latitude.
-For example, body size clines (larger size at higher latitude) have evolved repeatedly on three contients in \textit{Drosophila subobscura}, strongly indicating selection \citep{Huey_etal_2000}.
-But which one or more of the many climatic factors that covary with latitude actually controls selection on body size?
-Selection might be even more complex for perennial species that live over multiple seasons, meaning that environmental variation rather than the average environment may be a selective agent.
+Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
+Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
+For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}.
+Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely related to climatic variation.
+Despite the fact that latitudinal clines in particular have been studied for a long time, the exact climatic factors (e.g.
+minimum or maximum temperature, growing season length, strength of biotic interactions) is rarely known.
+Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g.
+coastal vs.
+continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}.
+In plants especially, we know little about the prevelance and adaptive significance of clinal variation in fundamental physiological traits like photosynthesis and the impact on plant performance.
 
-% Motivate using Mimulus
-Linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions.
-Many classic and contemporary studies of local adaptation have been conducted using species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources [willis, lowry?].
-Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008}.
+% For the most part, we know much more about selective agents responsible for clines for conspicuous things like colouration, such as in mice \citep{Sumner_1932, Mullen_Hoekstra_2008}
 
-Two common signatures of local adaptation are nonrandom genetic differences in ecologically important traits and variation in plasticity (genotype by environment interaction).
-In this study, we found strong genetically-based differences in photosynthesis and growth, but no evidence for variation in plasticity in response to temperature and drought treatments.
-Interannual variation in precipitaiton is associated intrinsic variation in these traits, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+% Motivate primary questions using Mimulus
+In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}.
+We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions.
+Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
+Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}.
+We measured genetic and genotype by environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
+We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity.
+Interannual variation in precipitaiton and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
 cardinalis}.
+We place these findings in the context of life history theory and consider future directions in the Discussion.
 
@@
 
-We used 16 populations from throughout the range of \textit{M. cardinalis} (Table~\ref{table:Table_FocalPops}). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.
+We used 16 populations from throughout the range of \textit{M.
+cardinalis} (Table~\ref{table:Table_FocalPops}).
+Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
 
@@
    \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
-   \begin{tabular}{@{} lllllll @{}}
+   \begin{tabular}{@{} lllll @{}}
       \toprule
-  Name& Region  & Demo?& Pop gen?& Latitude  & Longtiude  & Elevation (mas) \\
+  Name& Region  & Latitude  & Longtiude  & Elevation (mas) \\
       \midrule
-	HAU & South Margin		& yes	& yes	& 32.657	& 
-    \ensuremath{-116.532}	& 799   \\
-	CTC	& South Margin		& yes	& no	& 32.609 & 
-    \ensuremath{-116.7}	& 267   \\
-	CUR	& South Margin		& yes	& yes & 32.9 & 
-    \ensuremath{-116.585}	& 1180   \\
-	GRP & South Margin		& no	& no	& 33.314 &
-    \ensuremath{-116.871}	& 1577   \\
-	WWC &	Transverse		  & yes	& yes	& 33.994 & 
-    \ensuremath{-116.665}	& 705   \\
-	MIL	& Transverse	  	& yes	& no	& 34.077 & 
-    \ensuremath{-116.873}	& 2050   \\
-	WFM	& Transverse	  	& yes	& no	& 34.284 & 
-    \ensuremath{-117.378}	& 1120   \\
-	NMT	& South Sierras		& yes	& yes?& 36.201 & 
-    \ensuremath{-118.651}	& 1314   \\
-	PRD	& South Sierras		& yes	& yes	& 36.518 & 
-    \ensuremath{-118.759}	& 926   \\
-	RWD	& South Sierras		& yes	& yes	& 36.691 & 
-    \ensuremath{-118.91}	& 1727   \\
-	WNA	& Central Sierras	& yes	& yes	& 37.541 & 
-    \ensuremath{-119.649}	& 1224   \\
-	RBW	& Central Sierras	& yes	& no	& 37.819 & 
-    \ensuremath{-120.007}	& 876   \\
-	MYU	& North Sierras		& yes	& yes	& 39.397 & 
-    \ensuremath{-121.082}	& 455   \\
-	LIJ	& North Sierras		& yes	& yes	& 39.743 & 
-    \ensuremath{-120.704}	& 1603   \\
-	DPC	& North Coast	  	& yes	& yes	& 41.668 & 
-    \ensuremath{-123.11}	& 707   \\
-	RCC	& North Margin		& yes	& yes	& 43.374 & 
-    \ensuremath{-122.957}	& 326   \\
+	HAU & South Margin & 32.657	& 
+    -116.532	& 799   \\
+	CTC	& South Margin & 32.609 & 
+    -116.7	& 267   \\
+	CUR	& South Margin & 32.9 & 
+    -116.585	& 1180   \\
+	GRP & South Margin & 33.314 &
+    -116.871	& 1577   \\
+	WWC &	Transverse & 33.994 & 
+    -116.665	& 705   \\
+	MIL	& Transverse & 34.077 & 
+    -116.873	& 2050   \\
+	WFM	& Transverse & 34.284 & 
+    -117.378	& 1120   \\
+	NMT	& South Sierras & 36.201 & 
+    -118.651	& 1314   \\
+	PRD	& South Sierras & 36.518 & 
+    -118.759	& 926   \\
+	RWD	& South Sierras & 36.691 & 
+    -118.91	& 1727   \\
+	WNA	& Central Sierras & 37.541 & 
+    -119.649	& 1224   \\
+	RBW	& Central Sierras	& 37.819 & 
+    -120.007	& 876   \\
+	MYU	& North Sierras	& 39.397 & 
+    -121.082	& 455   \\
+	LIJ	& North Sierras	& 39.743 & 
+    -120.704	& 1603   \\
+	DPC	& North Coast & 41.668 & 
+    -123.11	& 707   \\
+	RCC	& North Margin & 43.374 & 
+    -122.957	& 326   \\
 	\bottomrule
@@
 
-We imposed four treatments, a fully-factorial cross of two temperature treatments and two watering treatments.
-Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment and another to the Cool treatment.
+We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels.
+The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which designate as Hot and Cool treatments.
+Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments.
+A detailed description of treatments is described in the Supplemental Information and summarized in Fig~\ref{fig:Fig_ExptlDes}.
+Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level.
 Within each chamber, there were two Well-watered blocks and two Drought blocks.
-A detailed description of treatments is given below \cdm{should some be moved to supplement?} and summarized in Fig~\ref{fig:Fig_ExptlDes}.
 The irradiance in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$.
-The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature treatments (data not shown).
-
-\subsubsection*{Temperature treatments}
-
-We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool).
-We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
-Daily temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
-Angert, unpub.
-data).
-Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M.
-cardinalis} in nature.
-We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}.
-The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ).
-The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
-We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R.
-Simulated data closely resembled the autocorrelation and between-site correlation of the actual data.
-From simulated mean temperature, we next selected minimum and maximum daily temperatures.
-Mean, min, and max temperature were highly correlated at both sites.
-We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models.
-For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
-During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
-
-\subsubsection*{Watering treatments}
-
-For watering treatments, we simulated two extreme types of streams where \textit{M.
-cardinalis} grows.
-In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
-In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down as the winter snow pack melts.
-In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER.
-Plants in the well-watered treatment were fully saturated every two hours during the day.
-Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing).
-Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.
-
-\cdm{Are there any data from California about soil water content through the season?}
+The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
 
@@
 During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of 329 plants evenly spread between treatments and families within populations.
-The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
+The youngest, fully-expanded leaf \cdm{this is what I did...not sure exactly which node, as I didn't record that when I did measurements} acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
 All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity.
-During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ \cdm{get exact temps used}) so that all plants within a temperature treatment were measured under the same conditions.
+During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.
 
@@
 \subsection*{Principal components of germination, growth, and photosynthesis}
-We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components.
+For each single-trait model above, we extracted the Population coefficient (factoring out Treatment and other effects).
+The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA).
 The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast and slow growing populations.
@@
 
-As with previous studies in \textit{Mimulus} and many other species [CITES, \citep{Huey_etal}, mosquitoes, what else?], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g.
-temperature or precipitation) that covary with latitude.
-We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results).
+Since latitude \textit{per se} cannot be a selective agent, we considered three possible explanations.
+First, latitude may be strongly correlated, in this species, with one or two climatic variables, such as temperature, precipitation, or growing degree-days.
+Second, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latiduinal cline.
+Third, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
+For example, average temperature of \textit{M.
+cardinalis} populations at a given latitude varies widely, but in aggregate, a Southern metapopulation experiences warmer climate than a Northern one.
+Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g.
+high elevation) site.
+
+To evaluate support for these hypotheses, we used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
 cardinalis}.
-We did this by looking for overlap between climatic variables that best predict latitude of \textit{M.
-cadinalis} occurence records and climatic variables that best predict trait variation across our 16 focal populations.
+We looked for overlap between climatic variables that best predict latitude of \textit{M.
+cadinalis} occurence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
 For brevity, we refer to these as Climate-Latitude and Climate-Trait variables.
 We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}).
-From VSURF models, we kept only variables selected for prediction, the most stringent criteria.
+From VSURF models, we kept only variables selected for prediction, the most stringent criterion.
+
+The first hypothesis predicts that there should be one or two Climate-Latitude and Climate-Trait associations that are strongly correlated, whereas the second hypothesis predicts that several climatic variables should be weakly correlated.
+To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-Trait variables as above, except that we used spatially averaged climate variables.
+We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations.
+We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
+Since \textit{M.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
+For clarity, we distinguish between analyses where climate is inferred from a single point (`point estimated Climate-Trait') versus averaged across a 100-km buffer (`spatially averaged Climate-Trait').
 
@@
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included 24 cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+In total, we included 24 climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
 For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
@@
 
-Genetic differences between populations affecting originsmal function and performance point to traits that may be involved in local adaptation.
-Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis} (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured 74.2 %$\%$ of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
+Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
+cardinalis}, as evidenced by large and highly significant population effects (Table~\ref{table:Table_AnovaSummary}).
+A single principal component captured 74.2 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
 As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
@@
 
+% perhaps responses of Simiolus sp (annual, drought selects for escape rather than tolerance) differ from card (perennial, escape through growth rather than flowering time) because drying out of rivers is less severe than seeps and springs?
+
 \section*{Discussion}
 
-The observation that populations are locally adapted to their environment is old and well-documented in many species.
+% The observation that populations are locally adapted to their environment is old and well-documented in many species.
 New frontiers in local adaptation research seek a general explanation for why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
 Understanding why and how populations have adapted to different environments in the past is especially significant now as a scientific basis for predicting how species' ranges will shift under climate change [CITES, \citep{Catullo_etal_2015}].
-In this study, we found evidence for one of two common signature of local adaptation, pointing to both phenotypic lability and constraint.
-In particular, our data suggest that photosynthesis and intrinsic growth rate vary among populations, whereas the fundamental abiotic niche is more conserved, and that climatic variation may be a more important selective agent than the average.
-Next, we tie these results into the broader threads of evolutionary theory in order to figure out what factors explain the trait variation we see and that we do not.
+In this study, we found evidence for one of two common signatures of local adaptation.
+A latitudinal cline in traits such as germination rate, photosynthesis, and growth, suggests adaptive differentiation in fundamental physiological traits of the species.
+However, we found little evidence that populations respond differently to temperature or drought.
+As we discuss below, this may indicate that the fundamental abiotic niche is relatively conserved \ala{the results that underlie this statement need more emphasis in the methods}.
+Finally, we found that climatic variation between years may be a more important selective agent than the average climate.
+In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in photosynthesis and growth varies clinally, but plastic responses to temperature and drought are relatively conserved.
 
-As long as environments vary and there exists genetic variation in fitness, why shouldn't local adapted genotypes evolve?
-Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001}.
-Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strong asymmetric, and/or maladaptive gene flow is too high.
+Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001} and hence the type of variation maintained within species.
+Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strongly asymmetric, and/or maladaptive gene flow is too high.
 In \textit{M.
@@
 \cdm{I will work more on this once I am more comfortable with the rest of the discussion}.
+
+% niche-tracking versus niche-shifting
+% One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve.
+For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.
+
+% citations on life history allocation tradeoff between growth and reproduction:
+%% THEORY:
+% Bell 1980 - briefly mentionds on page 56 that increased environmental variation for adults will favor semelparity
+% Iwasa and Cohen 1989, pg.
+496 "A larger reliability of habitat, a higher storage efficiency, higher productivity of the environment, and a longer growing season, all tend to promote the evolution of perennial life, as suggested by previous models of life-historsy strategy (Stearns 1976).
+
+%% EMPIRICAL:
+% Sawada et al. 1994 - latitudinal cline dry matter accumulation and storage allocation
+% Stanton et al 2000 show selection in stressful environemts (artificual selection over 5 generations) favors stress avoidance rather than tolerance traits.
+i.e.
+might be easier to adjust life history rather than tolerance.
+% Hall and Willis 2006 find selection on flowering time in coastal-inland comparison of M guttatus
+% Ehleringer et al 1990 find lower WUE increased yield in ephemeral pops
+% Bazzaz 1979: higher photo in fast growing plants (early succession)
+
+% Reznick 1985; Lovett Doust 1989
+% Williams 1966; Charnov and Schaffer 1973; Bell 1980; Reznick 1985; Biere 1995; Johnson 2007).
+% for iteroparous sp: van Noordwijk and De Jong 1986.
+% from Remington et al.
+2013: "Genetic variation in traits subject to trade-offs results in what has been termed “structured pleiotropy,” in which genetic covariances between traits result from functional constraints imposed by the limiting resources (De Jong 1990; Stearns et al.
+1991; see Figure 1B).
 
@@
 
-\subsection*{Spatially-averaged climate}
+\section*{}
 
-% Methods for spatial-average selection
-Selection at given site might not reflect simply the local environment, but rather the average environment in a surrounding region, since alleles will experience selection across populations linked by migration.
-We therefore calculated spatially averaged climate means and coefficients of variation.
-We sampled climate at 1000 random points (at 90-m resolition) within a 10 km buffer around focal populations.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
+\subsection*{Temperature treatments}
+
+We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool).
+We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
+This range was chosen because seeds used in the experiment were collected around 2012, thus their presence in that location at that time suggests that populations were able to persist there for at least some years before collection.
+Monthly temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
+Angert, unpub.
+data).
+Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M.
+cardinalis} in nature.
+We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}.
+The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ).
+The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
+We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R.
+Simulated data closely resembled the autocorrelation and between-site correlation of the actual data.
+From simulated mean temperature, we next selected minimum and maximum daily temperatures.
+Mean, min, and max temperature were highly correlated at both sites.
+We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models.
+For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
+During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
+
+\subsection*{Watering treatments}
+
+For watering treatments, we simulated two extreme types of streams where \textit{M.
+cardinalis} grows.
+In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
+In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down through the summer.
+In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER.
+Plants in the well-watered treatment were fully saturated every two hours during the day.
+Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing).
+Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.
 

### 2016-02-02 ms/ms.Rnw
Edited manuscript.

@@
 Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}.
-Phenotypic differences between populations along the cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
+Phenotypic differences between populations along a cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
 Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
@@
 A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures.
-Here, we examine physiological trait variation and possible selective agents in the perennail herb \textit{Mimulus cardinalis} Douglas ex Benth.
+Here, we examine physiological trait variation and possible selective agents in the perennial herb \textit{Mimulus cardinalis} Douglas ex Benth.
 (Phrymaceae), a model system for local adaptation studies.
@@
 We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively.
-There are classic cases adaptation involving both intrinsic and plastic trait variation.
+There are classic cases of adaptation involving both intrinsic and plastic trait variation.
 For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
 Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
+In plants especially, we know little about the prevalance and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
 
 % Selective agents
-Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
-Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
+Either instrinsic and/or plastic variation should vary clinally along environmental gradients.\ala{this doesn't make sense given definitions above} Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
 For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}.
 Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
-Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely related to climatic variation.
-Despite the fact that latitudinal clines in particular have been studied for a long time, the exact climatic factors (e.g.
-minimum or maximum temperature, growing season length, strength of biotic interactions) is rarely known.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely relate to climatic variation.
+
+Despite the fact that latitudinal clines in particular have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
+Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days, such that latitude is an effective proxy for the underlying climatic driver but nonetheless the key climatic variable(s) with stronger relationships to traits are identifiable.
+Alternatively, latitude may be more strongly related to traits than any single climatic variable for two possible reasons.
+First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline.
+Alternatively, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
+For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a Southern set of populations will experience warmer climate than a Northern one.
+Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g.
+high elevation) site.
 Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g.
@@
 continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}.
-In plants especially, we know little about the prevelance and adaptive significance of clinal variation in fundamental physiological traits like photosynthesis and the impact on plant performance.
+\ala{redo end sentence}
 
@@
 We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity.
-Interannual variation in precipitaiton and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+Interannual variation in precipitation and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
 cardinalis}.

### 2016-02-03 ms/ms.Rnw
Edited manuscript (intro, methods pertaining to climate-trait hypotheses.

@@
-lost on page 11 - extracted and used climate variables
-Maybe reordering stuff on paragraph
-cartoon of whole procedure for climate variables
-figure showing correlation b/t bio12 and 
-state that VSURF importance is different scales in different models
-moved up three hypotheses part to introduction
-play up lack of GxE and ???
-** correct # of Pop x Treatment
-
-* delve into interactions - do they change cline?
-
-* show figures of reaction norms
-* split plot, account for error variation in temp???
-* show PAR similarity
-* what CO2 concentration did I use?
-* some sites they are functionally annual
-
-* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
-	- maybe temporal source-sink makes more sense
-
-* check out matthew's blog on hydrology variables
-
-* explain rhizomes
-
-* put in biomass allocation cline
+%lost on page 11 - extracted and used climate variables
+%Maybe reordering stuff on paragraph
+%cartoon of whole procedure for climate variables
+%figure showing correlation b/t bio12 and 
+%state that VSURF importance is different scales in different models
+%moved up three hypotheses part to introduction
+%play up lack of GxE and ???
+%** correct # of Pop x Treatment
+
+%* delve into interactions - do they change cline?
+
+%* show figures of reaction norms
+%* split plot, account for error variation in temp???
+%* show PAR similarity
+%* what CO2 concentration did I use?
+%* some sites they are functionally annual
+
+%* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
+%	- maybe temporal source-sink makes more sense
+
+%* check out matthew's blog on hydrology variables
+
+%* explain rhizomes
+
+%* put in biomass allocation cline
 % TO DO:
@@
   # Manuscript files
-  pathMS <- "~/Google Drive/CardLocalAdaptation/ms" # Chris' MacBook
+  # pathMS <- "~/Google Drive/CardLocalAdaptation/ms" # Chris' MacBook
   # pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop
+  pathMS <- "/Users/amylauren/Desktop/CardAdapt/ms" # Amy's desktop
 
@@
 % Intrinsic versus plastic variation 
-There are two basic approaches one can use to identify candidate traits underlying local adaptation in a common garden.
-First, if genetically-based trait differences between populations vary clinally with environmental differences, this may point to traits important for local adaptation.
-Second, genotype by environment interactions could indicate that variation in plasticity mediates local adaptation.
-We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively.
+When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004} \ala{added citation - might give error}.
+The traits that underlie local adaptation, however, need not mirror this pattern.
+Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away.
+Alternatively, genotype-by-environment interactions could indicate that variation in plasticity mediates local adaptation.
+We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively.
 There are classic cases of adaptation involving both intrinsic and plastic trait variation.
 For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
-Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
+Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
 In plants especially, we know little about the prevalance and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
 
-% Selective agents
-Either instrinsic and/or plastic variation should vary clinally along environmental gradients.\ala{this doesn't make sense given definitions above} Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
+% Selective agents (trait-enviro assoc)
+A basic approach one can use to identify candidate traits underlying local adaptation is to find associations between traits and environments.
+Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
+Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
 For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}.
@@
 Despite the fact that latitudinal clines in particular have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
-Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days, such that latitude is an effective proxy for the underlying climatic driver but nonetheless the key climatic variable(s) with stronger relationships to traits are identifiable.
+Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days, such that latitude is an effective proxy for the underlying climatic driver but nonetheless  key climatic variable(s) with stronger relationships to traits exist.
 Alternatively, latitude may be more strongly related to traits than any single climatic variable for two possible reasons.
@@
 coastal vs.
-continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}.
-\ala{redo end sentence}
+continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local or regionally-averaged variables are more predictive of trait change.
+\ala{added this last clause to try to encompass gene flow idea, but now maybe the sentence is too clunky}
 
@@
 In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}.
-We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions.
+We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
 Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
 Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}.
-We measured genetic and genotype by environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
+We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
 We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity.
-Interannual variation in precipitation and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+Interannual variation in precipitation and/or precipitation seasonality is associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
 cardinalis}.
 We place these findings in the context of life history theory and consider future directions in the Discussion.
+\ala{wrap spatial buffer results into summary?}
 
@@
   library(survival)
-  pathMS <- "~/Google Drive/CardLocalAdaptation/ms"
-  pathDat <- paste(pathMS, "/Data", sep = "")
 
@@
 	mm <- lmer(LER_AbsGrowth ~ AvgGermDay + PopID * TempTrt * WaterTrt + 
-						 	(1|Family) + (1|Chamber/Block), data = master)
+						 	(1|Family) + (1|Block), data = master)
 	fitLER <- step(mm, reduce.random = F) # includes pop, temp, h2o, population x water
@@
 
-We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results).
-Since latitude \textit{per se} cannot be a selective agent, we considered three possible explanations.
-First, latitude may be strongly correlated, in this species, with one or two climatic variables, such as temperature, precipitation, or growing degree-days.
-Second, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latiduinal cline.
-Third, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
-For example, average temperature of \textit{M.
-cardinalis} populations at a given latitude varies widely, but in aggregate, a Southern metapopulation experiences warmer climate than a Northern one.
-Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g.
-high elevation) site.
-
-To evaluate support for these hypotheses, we used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
 cardinalis}.
 We looked for overlap between climatic variables that best predict latitude of \textit{M.
-cadinalis} occurence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
+cardinalis} occurrence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
 For brevity, we refer to these as Climate-Latitude and Climate-Trait variables.
@@
 
-The first hypothesis predicts that there should be one or two Climate-Latitude and Climate-Trait associations that are strongly correlated, whereas the second hypothesis predicts that several climatic variables should be weakly correlated.
+If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-Trait analyses \ala{and the correlation magnitudes should be high?
+maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}.
+If latitude is correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline, then several climatic variables should be weakly correlated \ala{fix this to mirror prior}.
+
 To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-Trait variables as above, except that we used spatially averaged climate variables.
-We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations.
+\ala{what's the actual prediction?
+that spatially averaged Climate-Trait will have higher correlation than point estimated Climate-Trait?} We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations.
 We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
@@
 ##### WORKING ON STUFF HERE #####
-vsurfPC1_SA$imp.mean.dec
-x <- match(vsurfPC1$imp.mean.dec.ind, vsurfPC1_SA$imp.mean.dec.ind)
-plot(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], type = "n")
-text(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], 
-		 labels = climVars[vsurfPC1$imp.mean.dec.ind])
-cor.test(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x])
+#vsurfPC1_SA$imp.mean.dec
+#x <- match(vsurfPC1$imp.mean.dec.ind, vsurfPC1_SA$imp.mean.dec.ind)
+#plot(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], type = "n")
+#text(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], 
+#		 labels = climVars[vsurfPC1$imp.mean.dec.ind])
+#cor.test(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x])
 

### 2016-02-03 ms/ms.tex
Edited manuscript (intro, methods pertaining to climate-trait hypotheses.

@@
+%lost on page 11 - extracted and used climate variables
+%Maybe reordering stuff on paragraph
+%cartoon of whole procedure for climate variables
+%figure showing correlation b/t bio12 and 
+%state that VSURF importance is different scales in different models
+%moved up three hypotheses part to introduction
+%play up lack of GxE and ???
+%** correct # of Pop x Treatment
+
+%* delve into interactions - do they change cline?
+
+%* show figures of reaction norms
+%* split plot, account for error variation in temp???
+%* show PAR similarity
+%* what CO2 concentration did I use?
+%* some sites they are functionally annual
+
+%* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
+%	- maybe temporal source-sink makes more sense
+
+%* check out matthew's blog on hydrology variables
+
+%* explain rhizomes
+
+%* put in biomass allocation cline
 % TO DO:
@@
 We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively.
-There are classic cases adaptation involving both intrinsic and plastic trait variation.
+There are classic cases of adaptation involving both intrinsic and plastic trait variation.
 For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
 Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
+In plants especially, we know little about the prevalance and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
 
 % Selective agents
-Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
-Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
+Either instrinsic and/or plastic variation should vary clinally along environmental gradients.\ala{this is confusing given definitions above} Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
 For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}.
 Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
-Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely related to climatic variation.
-Despite the fact that latitudinal clines in particular have been studied for a long time, the exact climatic factors (e.g.
-minimum or maximum temperature, growing season length, strength of biotic interactions) is rarely known.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely relate to climatic variation.
+
+Despite the fact that latitudinal clines in particular have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
+Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days, such that latitude is an effective proxy for the underlying climatic driver but nonetheless the key climatic variable(s) with stronger relationships to traits are identifiable.
+Alternatively, latitude may be more strongly related to traits than any single climatic variable for two possible reasons.
+First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline.
+Alternatively, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
+For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a Southern set of populations will experience warmer climate than a Northern one.
+Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g.
+high elevation) site.
 Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g.
@@
 % Motivate primary questions using Mimulus
+% might add that Anacker and Strauss 2014 showed ecological shifts important in Mim speciation and we would like to know driving forces behind this???
 In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}.
@@
 For clarity, we distinguish between analyses where climate is inferred from a single point (`point estimated Climate-Trait') versus averaged across a 100-km buffer (`spatially averaged Climate-Trait').
-
-For these analyses, we compiled a representative set of 178 recent (since 2000) known \textit{M.
-cardinalis} occurences.
-These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
-Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
-In total, we included 24 climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
-The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
-Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.
-
-%--------------------------------------------------
-% Results
-%--------------------------------------------------
-
-\section*{Results}
-
-\subsection*{A coordinated latitudinal cline in germination, growth, and photosynthesis}
-
-Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis}, as evidenced by large and highly significant population effects (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured 74.2 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
-As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Table of Population, Treatment, Population x Treatment interactions for varies traits
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}[ht]
-   \centering
-   \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects.
-We used different statistical modeling for the diverse traits assayed -- glm: generalized linear model using R \citep{R}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
-Note that temperature and water treatments were imposed after germination, hence are not application to this trait.
-Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}
-   \resizebox{6.5in}{!}{
-   \begin{tabular}{@{} llllllll @{}}
-      \toprule
-    {\raggedleft Trait}                          & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
-    Statistical model              & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
-      \midrule
-  Population                       &
-    ***  &
-    *** &
-    *** &
-    *** & 
-    *** & 
-    *** & \\
-  Temperature                      &
-    NA                                                          &
-    ***    &
-    *** &
-    *** & 
-     &
-    *** & \\
-  Water                            &
-    NA                                                          &
-    ***   &
-    ***   &
-    *** & 
-     &     
-    *** & \\
-  Pop $\times$ Temp                &
-    NA                                                          &
-     &
-     &
-    * & 
-     &     
-    * & \\
-  Pop $\times$ Water               &
-     NA                                                          &
-    * &
-     &
-     & 
-     &     
-     & \\
-  Temp $\times$ Water              &
-    NA                                                          &
-     &
-    *** &
-     & 
-     &     
-    *** & \\
-  Pop $\times$ Temp $\times$ Water &
-    NA                                                          &
-     &
-     &
-     & 
-     &     
-     & \\
-      \bottomrule
-   \end{tabular} }
-   \label{table:Table_AnovaSummary}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of latitude versus PC1 (growth axis)
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude.
-Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods).
-The line and 95\% confidence intervals were estimated using standard linear regression.}
-	\label{fig:Fig_PC1vLat}
-\end{figure}
-
-\subsection*{Remarkably little evidence for variation in plasticity}
-
-Genotype $\times$ environment interactions are also a common signature of local adaptation.
-For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
-We found remarkably little evidence for this pattern in \textit{M.
-cardinalis}.
-There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
-Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
-Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
-
-\subsection*{Climatic variability best explains phenotypic divergence}
-
-Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
-For \textit{M.
-cardinalis}, interannual variation in precipitation over the past 30 years is very closely related to the latitude of recently recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
-Between year variation in precipitation was also strongly correlated with position along the fast-slow growth axis, but climate based on point estimates gave somewhat different results than climate based on spatially averaged values.
-The two most important point estimated Climate-Latitude variables were also strong predictors of position along the fast-slow growth axis (Fig.~\ref{fig:Fig_ClimVarImp}B).
-For spatially averaged climate, interannual variation in precipitation seasonality was the only climatic variable selected, (Fig.~\ref{table:TableS_ClimVarImp}) \cdm{Do we need a figure for this?}.
-Overall, there was good but imperfect agreement between the weight assigned to climatic variables using point estimated and spatially averaged climate (Fig.~\ref{fig:Fig_TBD}) \cdm{This figure will show strong, positive correlation between weights given to climatic variables from point estimated and spatially averaged analyses}.
-
-Overlap between Climate-Latitude and point estimated Climate-Trait variables suggests that interannual variation in precipitation is an important selective agent in \textit{M.
-cardinalis}.
-Specifically, we hypothesize that more frequent droughts (greater precipitation cv) in Southern populations selects for an `annual-ized' life history, as we detail in the Discussion.
-Even though the spatially averaged Climate-Trait variable was not among the Climate-Latitude variables, it points to a similar conclusion.
-Specifically, this analysis showed that greater interannual variation in precipitation seasonality, which is also driven by precipitation variability (i.e.
-more frequent drought years) in Southern California, is associated with faster growth in southern populations.
-We must qualify these results because our analysis obviously cannot rule out that alternative variables not included in the analysis may be more important.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of climatic variable importance
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimVarImp.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{Interannual variation in precipitation is closely correlated with latitude and trait variation.
-A.
-Using Random Forest regression, we identified 10 climatic variables significantly (high importance) associated with latitude of \textit{M.
-cardinalis} occurences.
-B.
-The two most important Climate-Latitude variables were also the two most important Climate-Trait variables.
-Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
-Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp) or Precipitation (Prec); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
-	\label{fig:Fig_ClimVarImp}
-\end{figure}
-
-%--------------------------------------------------
-% Discussion
-%--------------------------------------------------
-
-% Maybe focus in on tradeoffs as sine qua non of specialization vs generalization, then mention that demography and/or gene flow could reinforce?
-
-% perhaps responses of Simiolus sp (annual, drought selects for escape rather than tolerance) differ from card (perennial, escape through growth rather than flowering time) because drying out of rivers is less severe than seeps and springs?
-
-\section*{Discussion}
-
-% The observation that populations are locally adapted to their environment is old and well-documented in many species.
-New frontiers in local adaptation research seek a general explanation for why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
-Understanding why and how populations have adapted to different environments in the past is especially significant now as a scientific basis for predicting how species' ranges will shift under climate change [CITES, \citep{Catullo_etal_2015}].
-In this study, we found evidence for one of two common signatures of local adaptation.
-A latitudinal cline in traits such as germination rate, photosynthesis, and growth, suggests adaptive differentiation in fundamental physiological traits of the species.
-However, we found little evidence that populations respond differently to temperature or drought.
-As we discuss below, this may indicate that the fundamental abiotic niche is relatively conserved \ala{the results that underlie this statement need more emphasis in the methods}.
-Finally, we found that climatic variation between years may be a more important selective agent than the average climate.
-In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in photosynthesis and growth varies clinally, but plastic responses to temperature and drought are relatively conserved.
-
-Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001} and hence the type of variation maintained within species.
-Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strongly asymmetric, and/or maladaptive gene flow is too high.
-In \textit{M.
-cardinalis} we found substantial genetic variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
-If this variation is adaptive, it suggests that the fitness tradeoff between doing well in low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
-That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatial heterogenous selection.
-We also know from previous work that population size does not vary strongly with latitude.
-Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range \citep{Paul_etal_2016}.
-
-Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years (see next section), could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
-We found no evidence for this; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
-We hypothesize that evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
-Riparian habitats where \textit{M.
-cardinalis} live are highly heterogeneous at small spatial scales.
-Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Meditteranean climates of western North America.
-But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and vice versa, leading to strong fitness tradeoffs.
-Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
-Demography and gene flow may reinforce niche conservatism \ala{also bring in idea of temporal source sink: most individuals made in wet years, even if infrequent, so selection weighted towards wet environment even if dry states are frequent}.
-A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher resource environments.
-Finally, gene flow, which is generally high among \textit{M.
-cardinalis} populations within the same ecoregion \citep{Paul_etal_2016}, will thwart local adaptation and reinforce specialization.
-Thus, the spatial grain of the environment, demographic asymmetry, and gene flow may conspire to constrain local adaptation via altered fundamental niche.
-
-Based on the available data, interannual variation in annual or winter precipitation or precipitaiton seasonality (these are closely correlated in Mediterranean climates) may be the selective agent driving variation in growth and photosynthesis.
-Variation in precipitation was best predicted latitude of recent \textit{M.
-cardinalis} occurences and trait variation along the fast-slow growth continuum (Fig.~\ref{fig:Fig_ClimVarImp}).
-A life history tradeoff between allocation to growth in the current year at the expense of future years could explain this pattern.
-In southern populations with more frequent droughts capable of killing rhizomes, a more annualized strategy could be favored.
-Conversely, in more predictable northern environments, lifetime fitness may be optimized when a significant fraction of assimilate is allocated below ground for future years.
-Although this hypothesis remains to be directly tested, a few indepdent lines of evidence are consistent with it.
-Preliminary surveys suggest that northern populations not only grow slower, but also produce greater numbers of rhizomes (C.D.
-Muir, unpub.
-data), suggesting an allocation tradeoff.
-Ecological niche models also show that occurence of southern populations is best predicted by recent climate (< 5 years), whereas northern occurences are best predicted by climate over the previous 30 years (M.
-Bayly \& A.
-Angert, unpub.
-data).
-Finally, demographic surveys of natural populations show greater variation in the size recruits in southern populations, suggesting higher maximum growth rates under natural conditions (M.
-Bayly \& A.
-Angert, unpub.
-data).
-There is a lot of interest in understanding how organisms will respond to changes in climatic variation, not just changes in the average climate.
-Our data indeed suggest that variation may be more important than the mean.
-
-\cdm{I am still working a paragraph or two linking these results to empirical results in other systems and climatic adaptaiton in Mimulus sect Simiolus.}
-
-% [Future work and broad conclusions].
-Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
-Next, we are testing whether phenotypic variation and constraint can be predicted by the shape of tradeoffs, as predicted by evolutionary theory.
-\cdm{I will work more on this once I am more comfortable with the rest of the discussion}.
-
-% niche-tracking versus niche-shifting
-% One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve.
-For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.
-
-% citations on life history allocation tradeoff between growth and reproduction:
-%% THEORY:
-% Bell 1980 - briefly mentionds on page 56 that increased environmental variation for adults will favor semelparity
-% Iwasa and Cohen 1989, pg.
-496 "A larger reliability of habitat, a higher storage efficiency, higher productivity of the environment, and a longer growing season, all tend to promote the evolution of perennial life, as suggested by previous models of life-historsy strategy (Stearns 1976).
-
-%% EMPIRICAL:
-% Sawada et al. 1994 - latitudinal cline dry matter accumulation and storage allocation
-% Stanton et al 2000 show selection in stressful environemts (artificual selection over 5 generations) favors stress avoidance rather than tolerance traits.
-i.e.
-might be easier to adjust life history rather than tolerance.
-% Hall and Willis 2006 find selection on flowering time in coastal-inland comparison of M guttatus
-% Ehleringer et al 1990 find lower WUE increased yield in ephemeral pops
-% Bazzaz 1979: higher photo in fast growing plants (early succession)
-
-% Reznick 1985; Lovett Doust 1989
-% Williams 1966; Charnov and Schaffer 1973; Bell 1980; Reznick 1985; Biere 1995; Johnson 2007).
-% for iteroparous sp: van Noordwijk and De Jong 1986.
-% from Remington et al.
-2013: "Genetic variation in traits subject to trade-offs results in what has been termed “structured pleiotropy,” in which genetic covariances between traits result from functional constraints imposed by the limiting resources (De Jong 1990; Stearns et al.
-1991; see Figure 1B).
-
-%--------------------------------------------------
-% Acknowledgements
-%--------------------------------------------------
-
-\section*{Acknowledgements}
-Erin Warkman and Lisa Lin helped collect data. CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program. \cdm{Please fill in other relevant funding info}
-
-%--------------------------------------------------
-% References
-%--------------------------------------------------
-
-\setlength{\bibsep}{6pt}
-\bigskip
-
-\bibliography{../../CommonResources/refs}
-\bibliographystyle{evolution}
-
-\clearpage
-
-%--------------------------------------------------
-% Supporting Information
-%--------------------------------------------------
-
-\section*{Supporting Information}
-
-% Modify and restart table/figure numbering for appendixes
-\renewcommand\thefigure{S\arabic{figure}}    
-\renewcommand\thetable{S\arabic{table}}    
-\renewcommand\theequation{S\arabic{equation}}    
-\setcounter{table}{0}    
-\setcounter{equation}{0}
-\setcounter{figure}{0}
-
-% DIC table for initial size (LLL on 5/12)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{Initial size of seedlings did not vary among Populations, Families, or Treatments.
-We used a censored Gaussian model of initial size at the outset of the experiment (longest leaf length of the first true leaves).
-The model was censored because we could not accurately measure leaves less than 0.25 mm with digital callipers (217 of 702, 30.9\%, were too small).
-We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package \pkg{MCMCglmm} version 2.17 \citep{Hadfield_2010}.
-We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in.
-We step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
-	\begin{center}
-	\begin{tabular}{>{\everypar{\hangindent1cm}{}\raggedright}p{6cm}lc}
-	\toprule
-
-	\input{Tables/Table_InitialSize.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-	\label{table:TableS_InitialSize}
-\end{table}
-
-\begin{table}[ht]
-   \centering
-   \topcaption{Climatic variables}
-   \begin{tabular}{@{} ll @{}}
-      \toprule
-  Climate variable & Abbreviation \\
-      \midrule
-  % climateWNA VARIABLES:
-	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
-	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
-	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
-	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
-	NFFD	& number of frost-free days                           \\
-	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
-	Eref	& Hargreaves reference evaporation (mm) \\
-	CMD	  & Hargreaves climativ moisture deficit (mm) \\
-	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
-	RH    &	mean annual relative humidity \\
-  % BIOCLIM VARIABLES:
-  bio1	&	Annual Mean Temperature	\\
-	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
-	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
-	bio4	&	Temperature Seasonality (standard deviation *100)	\\
-	bio5	&	Max Temperature of Warmest Month	\\
-	bio6	&	Min Temperature of Coldest Month	\\
-	bio7	&	Temperature Annual Range (bio5-bio6)	\\
-	bio8	&	Mean Temperature of Wettest Quarter	\\
-	bio9	&	Mean Temperature of Driest Quarter	\\
-	bio10	&	Mean Temperature of Warmest Quarter	\\
-	bio11	&	Mean Temperature of Coldest Quarter	\\
-	bio12	&	Annual Precipitation	\\
-	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
-	bio16	&	Precipitation of Wettest Quarter	\\
-	bio17	&	Precipitation of Driest Quarter	\\
-	bio18	&	Precipitation of Warmest Quarter	\\
-	bio19	&	Precipitation of Coldest Quarter	\\
-	    \bottomrule
-   \end{tabular}
-   \label{table:TableS_ClimVars}
-\end{table}
-
-% ANOVA table for LER (need to label tables)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{CAPTION}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-	\input{Tables/Table_lllGrowthAnova.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-% ANOVA table for SER (need to label tables)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, shoot elongation rate]{CAPTION}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-	\input{Tables/Table_heightGrowthAnova.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-\end{table}
-
-% Table of Climate-Latitude and Climate-Trait variables
-
-\begin{table}[ht]
-   \centering
-   \topcaption[Important climatic variables]{Important climatic variables predicting latitude of \textit{M.
-cardinalis} populations (`Climate-Latitude') and the first principal component of traits measured in a common garden (`Climate-Trait').
-Importance and significance were determined using the variable selection using random forests (VSURF) algorithm (see `Methods').
-Climatic variables are described in Table~\ref{table:TableS_ClimVars}.
-$\mu$ signifies the mean of the climate variables from 1981--2010; $\sigma$ indicates coeffiecient of variation among years}
-   \resizebox{6.5in}{!}{
-   \begin{tabular}{@{} lll @{}}
-      \toprule
-    	Climate-Latitude variables & Climate-Trait variables & \\
-                            	   & Point estimated         & Spatially averaged \\
-      \midrule
-			\input{Tables/TableS_ClimVarImp.txt}
-			\bottomrule
-   \end{tabular} }
-   \label{table:TableS_ClimVarImp}
-\end{table}
-
-% Figures showing latitudinal clines with specific traits
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations germinate sooner.]{CAPTION}
-	\label{fig:Fig_DoG}
-\end{figure}
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
-	\label{fig:Fig_LLL}
-\end{figure}
-
-\begin{figure}
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (shoot elongation rate).]{CAPTION}
-	\label{fig:Fig_height}
-\end{figure}
-
-\section*{}
-
-\subsection*{Temperature treatments}
-
-We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool).
-We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
-This range was chosen because seeds used in the experiment were collected around 2012, thus their presence in that location at that time suggests that populations were able to persist there for at least some years before collection.
-Monthly temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
-Angert, unpub.
-data).
-Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M.
-cardinalis} in nature.
-We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}.
-The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ).
-The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
-We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R.
-Simulated data closely resembled the autocorrelation and between-site correlation of the actual data.
-From simulated mean temperature, we next selected minimum and maximum daily temperatures.
-Mean, min, and max temperature were highly correlated at both sites.
-We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models.
-For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
-During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
-
-\subsection*{Watering treatments}
-
-For watering treatments, we simulated two extreme types of streams where \textit{M.
-cardinalis} grows.
-In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
-In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down through the summer.
-In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER.
-Plants in the well-watered treatment were fully saturated every two hours during the day.
-Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing).
-Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.
-
-\end{document}
-
-	\subsection{Structure}
-
-	Leaf traits method: \\
-	Day 1: \\
-	- select focal leaf (e.g. second or third leaf); \\
-	- place in H20, dark, cool overnight. \\
-	Day 2: \\
-	- wet weight; \\
-	- scan; \\
-	- place in drying oven. \\
-
-		\begin{enumerate}
-			\item{Stomatal density, guard cell length, ratio from leaf opposite focal leaf (above}
-			\item{Leaf size/shape (from scans)}
-			\item{Leaf thickness/SLA}
-			\item{Colorimetry (from scans, spec)}
-			\item{stem anthocyanin (qualitative)}
-		\end{enumerate}
-
-	OPEN QUESTIONS: \\
-	save wet samples in preservative (e.g. for leaf venation?) \\
-	chlorophyll content? \\
-	branching architecture (anything in plant protocols?)
-
-	\subsection{Function}
-
-		\subsubsection{during experiment}
-
-		\begin{enumerate}
-			\item{Instantaneous photosynthetic rate, $g_\text{s}$, water-use efficiency.}
-				\subitem{3 min per plant per LICOR}
-				\subitem{Do as many plants as possible in 1-2 weeks}
-				\subitem{Before and after drought?}
-			\item{Respiration}
-			\item{Whole plant water-use (gravimetric)}
-				\subitem{}
-			\item{Leaf temperature (subset)}
-			\item{Temperature response curve (subset)}
-				\subitem{see protocol here: http://prometheuswiki.publish.csiro.au/tiki-index.php?page=Temperature+response+of+photosynthesis+using+a+Li-Cor+6400}
-			\item{Light response curve}
-				\subitem{Sampling? Only in cool, wet?}
-				\subitem{At what temperature?}
-
-		\end{enumerate}
-
-		\subsubsection{after experiment}
-
-		\begin{enumerate}
-			\item{CN content}
-			\item{carbon isotope discrimination}
-		\end{enumerate}
-
-	\subsection{Development time and Life history}
-
-	\begin{enumerate}
-		\item{Flowering time: census daily}
-		\item{Starch accumulation: collect (subset) of plants}
-		\item{Rhizome formation: collect on subset of plants harvested for dry mass}
-	\end{enumerate}
-
-	\subsection{Floral (2nd/3rd/etc flower?)}
-
-	\begin{enumerate}
-		\item{Flower size}
-		\item{Herkogamy}
-	\end{enumerate}
-
-\section{Performance}
-
-	Possibilities: \\
-	1. (sub-)weekly census of plant height. $\approx$ 0.5 min per plant, so allow 1 day \\
-
-	$$RGR_\text{height} = \frac{\text{ln} H_2 - \text{ln} H_1}{t_2 - t_1}$$
-
-	2. early development (pre-treatment) leaf-area growth with camera \\
-	3. harvest dry biomass before and after treatments (get whole plant leaf area using scanner at same time to ground truth camera)
-
-	$$RGR_\text{mass} = \frac{\text{ln} M_2 - \text{ln} M_1}{t_2 - t_1}$$
-
-\section{Determining ecological gradients that matter}
-
-Performance tradeoffs
-Isolation by ecology
-
-\section{Tests of divergent selection}
-
-	\subsection{Trait-environment correlation}
-
-	We will account for relatedness by using genetic distance matrix in linear models.
-
-	\subsection{$P_\text{ST}$ -- $F_\text{ST}$}
-
-	Contrast trait classes with Fst distribution
-
-	\subsection{Isolation by ecology}
-
-	Is \textbf{G} better predicted by \textbf{E} than geographic distance? Use Gideon's method?
-
-	\subsection{Performance tradeoffs between treatments}
-
-\section{Ecological strategy, physiological mechanism, and tradeoff}
-
-	\subsection{Drought escape}
-		Pattern: \\
-			Flowering time shorter in drier/S populations \\
-			More between population variation in flowering time more variable than predicted by Fst \\
-			... \\
-
-	\subsection{Drought avoidance}
-		Pattern: \\
-			Maybe this should really be dehydration avoidance? \\
-			Populations from drier/S avoid dehydration by limiting water loss; more variable than predicted by Fst \\
-
-			Mechanism \\
-			closing stomata earlier during soil drying
-
-		Mechanisms: \\
-			lower water-use \\
-
-	\subsection{Drought tolerance}
-		Pattern: \\
-			drier/S populations have greater tolerance to low tissue water content
-
-		Mechanisms: \\
-			greater WUE \\
-			ability to photosynthesize during drought \\
-
-\section{Extension: incipient speciation as a byproduct or divergence?}
-
-Collect pollen to measure viability.
-
-(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
-Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
-Some hypotheses:
-\begin{itemize}
-  \item{H1: Latitude encompasses multivariate climatic dimension.}
-  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
-\end{itemize}
-
-Predictions:
-\begin{itemize}
-  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
-Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
-  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
-  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
-\end{itemize}
-
-\end{document}

### 2016-10-06 ms/ms.Rnw
Worked on analysis and ms

@@
   # Manuscript files
-  # pathMS <- "~/Google Drive/CardLocalAdaptation/ms" # Chris' MacBook
+  pathMS <- "~/Google Drive/CardAdapt/ms" # Chris' MacBook Air
   # pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop
-  pathMS <- "/Users/amylauren/Desktop/CardAdapt/ms" # Amy's desktop
+  # pathMS <- "/Users/amylauren/Desktop/CardAdapt/ms" # Amy's desktop
 
@@
 
-# Read in spatially-averaged (10 km buffer) for 16 focal populations
+# Read in spatially-averaged (100 km buffer) for 16 focal populations
 load(paste(pathDat, "/focClimSA.RData", sep = ""))
@@
   # Create Surv object
-	germ <- with(subset(master, !is.na(master$MinGermDay)), Surv(MinGermDay, MaxGermDay,	
-		type = "interval2"))
+	germ <- with(subset(master, !is.na(master$MinGermDay)), 
+	             Surv(MinGermDay, MaxGermDay,	type = "interval2"))
 
@@
     data = subset(master, !is.na(master$Photo))[-52, ])
-	fitIntrnPhoto <- step(mm, reduce.random = F)
+	fitIntrnPhoto <- step(mm, reduce.random = F) # only Pop
 
@@
 We assayed mortality during twice-weekly growth measurements.
-We could not get GLMM with Family effects to converge, so we used GLM with a quasibinomial error structure and assessed signifiance using Type 2 Analysis of Deviance with the R package \pkg{car}.
+We analyzed the probability of surviving until the end of the experiment as a function of population, treatment, and their interactions using a Generalized Linear Mixed Model (GLMM) assuming binomially distributed errors.
+As before, we included Family and Block as random factors.
+We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald \Chi$^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
 
@@
 
-  fitMort <- glm(DiedFromStress ~ PopID * TempTrt * WaterTrt,
+	# Code for GLM (remove if satisfied with GLMER)
+  # fitMort <- glm(DiedFromStress ~ PopID * TempTrt * WaterTrt,
+  #  data = subset(master, !is.na(master$DiedFromStress)), 
+  #  family = "quasibinomial")
+
+  fitMort <- glmer(DiedFromStress ~ PopID * TempTrt * WaterTrt + (1|Family) +
+  								 (1|Block), control = glmerControl(optimizer = "bobyqa"),
   								 data = subset(master, !is.na(master$DiedFromStress)), 
-    family = "quasibinomial")
-
-  fitMort <- glmer(DiedFromStress ~ PopID * TempTrt * WaterTrt + (1|Family) +
-  								 (1|Block), data = subset(master, !is.na(master$DiedFromStress)), family = "binomial")
-  Anova(fitMort, type = 2)
+  								 family = "binomial")
+  # Anova(fitMort, type = 2)
 @
@@
   par(mar = c(5, 6, 1, 1))
-  plot(0, 0, type = "n", xlim = c(0, ceiling(vsurfLat$ord.imp$x)[1]), 
+  plot(0, 0, type = "n", xlim = c(0, ceiling(vsurfLat$imp.mean.dec)[1]), 
 		ylim = c(1, length(vsurfLat$varselect.pred)), axes = F, frame.plot = T, 
@@
   }
-  points(vsurfLat$ord.imp$x[1:length(vsurfLat$varselect.pred)], 
+  points(vsurfLat$imp.mean.dec[1:length(vsurfLat$varselect.pred)], 
     length(vsurfLat$varselect.pred):1, pch = 21, col = "black", bg = "white", cex = 1.5)
@@
    \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects.
-We used different statistical modeling for the diverse traits assayed -- glm: generalized linear model using R \citep{R}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
+We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
 Note that temperature and water treatments were imposed after germination, hence are not application to this trait.
@@
     {\raggedleft Trait}                          & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
-    Statistical model              & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
+    Statistical model              & survreg     & lmer           & lmer            & lmer           & lmer          & glmer \\
       \midrule

### 2016-10-06 ms/Figures/CODE FOR GEO FIGURE EXPLANATION.R

@@ -1,0 +1,174 @@
+
+# Libraries
+
+	library(raster)
+	library(rgeos)
+
+# Color scheme
+	
+	colNorth <- "deepskyblue"
+	colSouth <- rgb(255, 131, 0, maxColorValue = 255)
+
+	# Get maps of California and Oregon
+usa <- getData('GADM', country = "USA", level = 1, path = pathDat)
+states <- gSimplify(usa[usa@data$NAME_1 %in% c("California", "Oregon", "Nevada", "Idaho"), ], 
+										tol = 0.01, topologyPreserve = TRUE)
+
+# Load thinned occurences used Climate-Lat analyses and convert to Spatial class
+occ <- read.csv(paste(pathDat, "/thinned1OccLatLon.csv", sep = ""), row.names = NULL)
+coordinates(occ) <- ~Longitude+Latitude
+projection(occ) <- CRS('+proj=longlat');
+
+# Load focal populations and convert to Spatial class
+load(paste(pathDat, "/focClim.RData", sep = ""))
+foc <- focClim[, c("Longitude", "Latitude")]
+coordinates(foc) <- ~Longitude+Latitude
+projection(foc) <- CRS('+proj=longlat');
+
+# Clip and map
+clipExtent <- as(extent(floor(min(occ$Longitude) * 10) / 10, 
+												ceiling(max(occ$Longitude) * 10) / 10, 
+												floor(min(occ$Latitude) * 10) / 10, 
+												ceiling(max(occ$Latitude) * 10) / 10), 
+								 "SpatialPolygons")
+proj4string(clipExtent) <- CRS(proj4string(states))
+states <- gIntersection(states, clipExtent, byid = TRUE)
+
+# Transform to Albert's Equal Area Projection
+projAEA <- sprintf("+proj=aea +lat_1=%s +lat_2=%s +lat_0=%s +lon_0=%s +x_0=0 +y_0=0",
+									 extent(clipExtent)[3], extent(clipExtent)[4], 
+									 extent(clipExtent)[1], extent(clipExtent)[2])
+states <- spTransform(states, CRS(projAEA))
+occAEA <- spTransform(occ, CRS(projAEA))
+focAEA <- spTransform(foc, CRS(projAEA))
+
+# PLOT
[…]

### 2016-10-07 ms/ms.Rnw
Worked on ms

@@
 % 8. Supplemental figure showing agreement between assigned to climatic variables from point estimated and spatially averaged analyses
+% 9. new spatially averaged analysis
+% 10. new analysis with updated ENM
+% 11. mention using LER_AbsGrowth as predictor in fitSER
+% 12. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
 
@@
 
-# Read in spatially-averaged (100 km buffer) for 16 focal populations
-load(paste(pathDat, "/focClimSA.RData", sep = ""))
+# Read in spatially-averaged (62 km buffer) for 16 focal populations
+# *** still need to make this filer ***
+load(paste(pathDat, "/focClimSA_62km.RData", sep = ""))
 
@@
 plot(0, 0, type = "n", xlim = c(0, max(exptlDes$DayN)), ylim = c(-1, 41), 
-		 axes = F, frame.plot = T, cex.lab = 1.5, xlab = "Key treatments and measurements", 
+		 axes = F, frame.plot = T, cex.lab = 1.5, 
+		 xlab = "Key treatments and measurements", 
      ylab = expression(paste("Temperature (", phantom()*degree, "C)")))
@@
 	border = "white", col = "grey"))
-#with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 3, 
-#	labels = "Photo", col = "white", font = 3, cex = 0.9))
 
@@
 
-# with(exptlDes, points(DayN, WaterFreq_dry, type = "l", lwd = 12, 
-	# col = alpha("white", 0.5)))
 with(exptlDes, points(DayN, WaterFreq_dry, type = "l", lwd = 3, col = "black"))
@@
 
+	# Note that growth data used in the mixed model are from the master datasheet.
+	# The code below shows how growth rates were calculated.
+
   # New (blank) datasheet on which to add data
@@
   # Mixed model ANOVA with family as random factor fit using step down procedure
-	master$Chamber <- substr(master$Block, 1, 3)
 	mm <- lmer(LER_AbsGrowth ~ AvgGermDay + PopID * TempTrt * WaterTrt + 
@@
 	fitLER <- step(mm, reduce.random = F) # includes pop, temp, h2o, population x water
-	# fitLER <- step(mm, ddf = "Kenward-Roger", reduce.random = F) # same result using Satterthwaite ddf
 
@@
 <<Shoot elongation rate, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+	# Note that growth data used in the mixed model are from the master datasheet.
+	# The code below shows how growth rates were calculated.
 
@@
 
-  mm <- lmer(SER_AbsGrowth ~ AvgGermDay + PopID * TempTrt * WaterTrt + 
-		(1|Family) + (1|Block), data = master)
-	fitSER <- step(mm, reduce.random = F) #includes pop, temp
+  mm <- lmer(SER_AbsGrowth ~ AvgGermDay + LER_AbsGrowth + 
+               PopID * TempTrt * WaterTrt + (1|Family) + (1|Block), 
+             data = master)
+	fitSER <- step(mm, reduce.random = F) #includes LER_AbsGrowth, pop, temp, h2o, and temp x h2o
+
 @
@@
 The youngest, fully-expanded leaf \cdm{this is what I did...not sure exactly which node, as I didn't record that when I did measurements} acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
-All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity.
+All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
 During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.
@@
   # Anova(fitMort, type = 2)
+
 @
@@
 
-  # plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
-  #plot(focClim$Lat, traitPC$x[, "PC1"])
-  #cor.test(focClim$Lat, traitPC$x[, "PC1"])
-  #cor.test(popMeans[, 4], traitPC$x[, "PC1"])
-  #plot(focClim$bio16_cva, traitPC$x[, "PC1"])
 @
@@
 	# Load climVars
-	load(paste(pathDat, "/ExploratoryCimateVariables.RData", sep = ""))
+	load(paste(pathDat, "/ExploratoryClimateVariables.RData", sep = ""))
 
@@
   {
-     lines(x = c(-1, length(vsurfLat$varselect.pred)), y = c(i, i), col = "grey")
+     lines(x = c(-1, length(vsurfLat$varselect.pred)), y = c(i, i), 
+           col = "grey")
   }
   points(vsurfLat$imp.mean.dec[1:length(vsurfLat$varselect.pred)], 
-    length(vsurfLat$varselect.pred):1, pch = 21, col = "black", bg = "white", cex = 1.5)
+    length(vsurfLat$varselect.pred):1, pch = 21, col = "black", bg = "white",
+    cex = 1.5)
 	# mtext(climVars[vsurfLat$varselect.pred], side = 2, 
@@
   # Variable selection using random forests (VSURF)
-  linmod.PC1 <- as.formula(paste("PC1 ~ ", paste(climVars, collapse = " + "), sep = ""))
+	# Climate variables that predict PC1 (point and spatially averaged)
   set.seed(69648)
-  # vsurfPC1 <- VSURF(linmod.PC1, data = focClim[, c("PC1", climVars)])
   vsurfPC1 <- VSURF(focClim[, climVars], focClim[, "PC1"])
-  # vsurfPC1_SA <- VSURF(linmod.PC1, data = focClimSA[, c("PC1", climVars)])
   vsurfPC1_SA <- VSURF(focClimSA[, climVars], focClimSA[, "PC1"])
 
-##### WORKING ON STUFF HERE #####
-#vsurfPC1_SA$imp.mean.dec
-#x <- match(vsurfPC1$imp.mean.dec.ind, vsurfPC1_SA$imp.mean.dec.ind)
-#plot(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], type = "n")
-#text(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], 
+##### WORKING ON STUFF HERE - can't finish until SA analysis with 62 km buffer finishes #####
+# x <- match(vsurfPC1$imp.mean.dec.ind, vsurfPC1_SA$imp.mean.dec.ind)
+# plot(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], type = "n")
+# text(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], 
 #		 labels = climVars[vsurfPC1$imp.mean.dec.ind])
-#cor.test(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x])
+# cor.test(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x])
 
@@
   par(mar = c(5, 5, 1, 1))
-  plot(focClim$bio16_cva, focClim$PC1, xlim = c(0.25, 0.55), ylim = c(-4, 4), axes = F,
-    frame.plot = T, type = "n", xlab = "", ylab = "")
+  plot(focClim$bio16_cva, focClim$PC1, xlim = c(0.25, 0.55), ylim = c(-4, 4),
+       axes = F, frame.plot = T, type = "n", xlab = "", ylab = "")
   title(xlab = "Precipitation of Wettest Quarter", cex.lab = 1.5)
@@
   s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitBio16cva) %*% c(1, X)))
-  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), upperCI = qnorm(0.975, u, s))
+  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
+                        upperCI = qnorm(0.975, u, s))
   polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
-    col = "grey80", border = NA)
+    col = "white", border = NA)
   lines(x, confInt$lowerCI, lwd = 2, lty = 2)
@@
   lines(range(focClim$bio16_cva), range(predict(fitBio16cva)), lwd = 2)
-  points(focClim$bio16_cva, focClim$PC1, pch = 21, col = "black", bg = primShade0, 
-    cex = 1.5, lwd = 2)
+  points(focClim$bio16_cva, focClim$PC1, pch = 21, col = "black", 
+         bg = "white", cex = 1.5, lwd = 2)
   dev.off()
@@
 
-	plot(0, 0, type = "n", xlim = c(0, 0.3), ylim = c(1, nImpLat), 
+	plot(0, 0, type = "n", xlim = c(0, 0.2), ylim = c(1, nImpLat), 
 		axes = F, frame.plot = T, xlab = "", ylab = "")
@@
 		pch = 21, col = "black", bg = "white", cex = 1.5)
-	# mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, 
-  #  at = nImpLat:(nImpLat - nImpPC1 + 1), 
-	#	font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
-	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.3, 0.1))
+	mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, 
+    at = nImpLat:(nImpLat - nImpPC1 + 1), 
+		font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
+	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.2, 0.05))
   dev.off()
@@
 \end{document}
-
-	\subsection{Structure}
-
-	Leaf traits method: \\
-	Day 1: \\
-	- select focal leaf (e.g. second or third leaf); \\
-	- place in H20, dark, cool overnight. \\
-	Day 2: \\
-	- wet weight; \\
-	- scan; \\
-	- place in drying oven. \\
-
-		\begin{enumerate}
-			\item{Stomatal density, guard cell length, ratio from leaf opposite focal leaf (above}
-			\item{Leaf size/shape (from scans)}
-			\item{Leaf thickness/SLA}
-			\item{Colorimetry (from scans, spec)}
-			\item{stem anthocyanin (qualitative)}
-		\end{enumerate}
-
-	OPEN QUESTIONS: \\
-	save wet samples in preservative (e.g. for leaf venation?) \\
-	chlorophyll content? \\
-	branching architecture (anything in plant protocols?)
-
-	\subsection{Function}
-
-		\subsubsection{during experiment}
-
-		\begin{enumerate}
-			\item{Instantaneous photosynthetic rate, $g_\text{s}$, water-use efficiency.}
-				\subitem{3 min per plant per LICOR}
-				\subitem{Do as many plants as possible in 1-2 weeks}
-				\subitem{Before and after drought?}
-			\item{Respiration}
-			\item{Whole plant water-use (gravimetric)}
-				\subitem{}
-			\item{Leaf temperature (subset)}
-			\item{Temperature response curve (subset)}
-				\subitem{see protocol here: http://prometheuswiki.publish.csiro.au/tiki-index.php?page=Temperature+response+of+photosynthesis+using+a+Li-Cor+6400}
-			\item{Light response curve}
-				\subitem{Sampling? Only in cool, wet?}
-				\subitem{At what temperature?}
-
-		\end{enumerate}
-
-		\subsubsection{after experiment}
-
-		\begin{enumerate}
-			\item{CN content}
-			\item{carbon isotope discrimination}
-		\end{enumerate}
-
-	\subsection{Development time and Life history}
-
-	\begin{enumerate}
-		\item{Flowering time: census daily}
-		\item{Starch accumulation: collect (subset) of plants}
-		\item{Rhizome formation: collect on subset of plants harvested for dry mass}
-	\end{enumerate}
-
-	\subsection{Floral (2nd/3rd/etc flower?)}
-
-	\begin{enumerate}
-		\item{Flower size}
-		\item{Herkogamy}
-	\end{enumerate}
-
-\section{Performance}
-
-	Possibilities: \\
-	1. (sub-)weekly census of plant height. $\approx$ 0.5 min per plant, so allow 1 day \\
-
-	$$RGR_\text{height} = \frac{\text{ln} H_2 - \text{ln} H_1}{t_2 - t_1}$$
-
-	2. early development (pre-treatment) leaf-area growth with camera \\
-	3. harvest dry biomass before and after treatments (get whole plant leaf area using scanner at same time to ground truth camera)
-
-	$$RGR_\text{mass} = \frac{\text{ln} M_2 - \text{ln} M_1}{t_2 - t_1}$$
-
-\section{Determining ecological gradients that matter}
-
-Performance tradeoffs
-Isolation by ecology
-
-\section{Tests of divergent selection}
-
-	\subsection{Trait-environment correlation}
-
-	We will account for relatedness by using genetic distance matrix in linear models.
-
-	\subsection{$P_\text{ST}$ -- $F_\text{ST}$}
-
-	Contrast trait classes with Fst distribution
-
-	\subsection{Isolation by ecology}
-
-	Is \textbf{G} better predicted by \textbf{E} than geographic distance? Use Gideon's method?
-
-	\subsection{Performance tradeoffs between treatments}
-
-\section{Ecological strategy, physiological mechanism, and tradeoff}
-
-	\subsection{Drought escape}
-		Pattern: \\
-			Flowering time shorter in drier/S populations \\
-			More between population variation in flowering time more variable than predicted by Fst \\
-			... \\
-
-	\subsection{Drought avoidance}
-		Pattern: \\
-			Maybe this should really be dehydration avoidance? \\
-			Populations from drier/S avoid dehydration by limiting water loss; more variable than predicted by Fst \\
-
-			Mechanism \\
-			closing stomata earlier during soil drying
-
-		Mechanisms: \\
-			lower water-use \\
-
-	\subsection{Drought tolerance}
-		Pattern: \\
-			drier/S populations have greater tolerance to low tissue water content
-
-		Mechanisms: \\
-			greater WUE \\
-			ability to photosynthesize during drought \\
-
-\section{Extension: incipient speciation as a byproduct or divergence?}
-
-Collect pollen to measure viability.
-
-(this also goes into the goal of trying to figure out ecological selective agent driving trait variation).
-Why does latitude predict trait variation, often better than any of the climatic variables that vary with latitude?
-Some hypotheses:
-\begin{itemize}
-  \item{H1: Latitude encompasses multivariate climatic dimension.}
-  \item{H2: Latitude is substitute for average selective environment. i.e. populations are connected by gene flow, such that alleles at a given site must do well there and in surrounding areas.}
-\end{itemize}
-
-Predictions:
-\begin{itemize}
-  \item{P1: Climatic axis in high dimensional climate space should correlate with latitude (across card sites especially, not just any site [using pseudoabsences]).
-Focal site position in high-dimensional climatic space should correlated with traits as well or better than latitude.}
-  \item{P2: Climate (averaged over nearby sites) varies with latitude and the nearby average climate predicts trait values as well or better than latitude.}
-  \item{P1+2: These are not mutually exclusive and both could be imortant. i.e. average selective environment in high-dimensional climate space predicts trait values as well or better than latitude.}
-\end{itemize}
-
-\end{document}

### 2016-10-12 ms/ms.Rnw
Worked on ms and figures

@@
 <<focClim, echo = FALSE, results=hide, warning=FALSE>>=
+
 # Read in climate data for 16 focal populations to use in making Table 1
@@
 # Read in spatially-averaged (62 km buffer) for 16 focal populations
-# *** still need to make this filer ***
 load(paste(pathDat, "/focClimSA_62km.RData", sep = ""))
@@
 
+  # Variable selection using random forests (VSURF)
+	# Climate variables that predict PC1 (point and spatially averaged)
+  set.seed(69648)
+  vsurfPC1 <- VSURF(focClim[, climVars], focClim[, "PC1"], parallel = TRUE,
+                    ncores = 3, ntree = 1e4)
+  climVars[vsurfPC1$varselect.pred]
+  vsurfPC1_SA <- VSURF(focClimSA[, climVars], focClimSA[, "PC1"], 
+                       parallel = TRUE, ncores = 3, ntree = 1e4)
+  climVars[vsurfPC1_SA$varselect.pred]
+
+  imp1 <- climVars[vsurfLat$varselect.pred]
+  nImpLat <- length(imp1)
+
+  imp2 <- climVars[vsurfPC1$varselect.pred]
+  nImpPC1 <- length(imp2)
+
+  imp3 <- climVars[vsurfPC1_SA$varselect.pred]
+  nImpPC1_SA <- length(imp3)
+
 	# Load climVars
@@
 
+	# Table of symbols for plotting climatic factor labels
+	clim2sym <- read.csv(paste(pathMS, "/climVar2plot.csv", sep = ""), 
+										   stringsAsFactors = F, row.names = 1)
+
   # Figure of variable importance for latitude
-  pdf(paste(pathFig, "/Figure_clim2lat_imp.pdf", sep = ""), 4, 5)
-  par(mar = c(5, 6, 1, 1))
+
+	# Calculatins for plot dimensions
+	sw1a <- strwidth(s = clim2sym[imp1, "climFactor"], units = "inches")
+	sw2a <- strwidth(s = clim2sym[imp1, "tempScale"], units = "inches")
+	sw3a <- strwidth(s = parse(text = clim2sym[imp1, "what"]), units = "inches")
+
+	sw1b <- strwidth(s = clim2sym[imp2, "climFactor"], units = "inches")
+	sw2b <- strwidth(s = clim2sym[imp2, "tempScale"], units = "inches")
+	sw3b <- strwidth(s = parse(text = clim2sym[imp2, "what"]), units = "inches")
+
+	panelWidth <- 3
+	panelHeight <- 4
+	panelGap <- 0.1
+	yLabWidth <- 4 * panelGap + sum(max(c(sw1a, sw1b)), max(c(sw2a, sw2b)),
+	                                max(c(sw3a, sw3b)))
+  xLabHeight <- 1
+
+  pdf(paste(pathFig, "/Figure_ClimVarImp.pdf", sep = ""), 
+      width = 2 * (yLabWidth + panelWidth) + panelGap, 
+      height = panelHeight + xLabHeight + panelGap)
+  par(mai = c(xLabHeight, yLabWidth, panelGap, panelGap / 2), mfrow = c(1, 2))
+
+  # Panel A: Climate-Latitude
   plot(0, 0, type = "n", xlim = c(0, ceiling(vsurfLat$imp.mean.dec)[1]), 
@@
 		xlab = "", ylab = "")
-	title(xlab = "Climate-Latitude\nImportance", line = 4, cex.lab = 1.5)
-  for (i in 1:length(vsurfLat$varselect.pred)) 
-  {
-     lines(x = c(-1, length(vsurfLat$varselect.pred)), y = c(i, i), 
-           col = "grey")
-  }
+  abline(h = 1:nImpLat, col = "grey")
   points(vsurfLat$imp.mean.dec[1:length(vsurfLat$varselect.pred)], 
@@
     cex = 1.5)
-	# mtext(climVars[vsurfLat$varselect.pred], side = 2, 
-  #   at = length(vsurfLat$varselect.pred):1, las = 1, line = 1)
+
+  # X axis & labels
 	axis(1, lwd = 0, lwd.ticks = 1)
+  title(xlab = "Climate-Latitude\nImportance", line = 4, cex.lab = 1.5)
+
+  # Y labels
+	clip(grconvertX(0, from = "nfc", to = "user"), 
+	     grconvertX(1, from = "nfc", to = "user"),
+	     grconvertY(0, from = "nfc", to = "user"),
+	     grconvertY(1, from = "nfc", to = "user"))
+
+	text(grconvertX(panelGap / 2, from = "in", to = "user"), 
+	     nImpLat:1, pos = 4, labels = clim2sym[imp1, "climFactor"])
+  text(grconvertX(panelGap + max(c(sw1a, sw1b)), from = "in", to = "user"), 
+       nImpLat:1, pos = 4, labels = clim2sym[imp1, "tempScale"])
+	text(grconvertX(3 / 2 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), 
+	                               from = "in", to = "user"), nImpLat:1, pos = 4,
+	                labels = parse(text = clim2sym[imp1, "what"]))
+
+	# Panel B: Climate-PC1
+	par(mai = c(xLabHeight, panelGap / 2, panelGap, yLabWidth))
+	plot(0, 0, type = "n", xlim = c(0, 0.2), ylim = c(1, nImpLat), 
+		axes = F, frame.plot = T, xlab = "", ylab = "")
+	abline(h = nImpLat:(nImpLat - nImpPC1 + 1), col = "grey") 
+	points(vsurfPC1$imp.mean.dec[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), 
+		pch = 21, col = "black", bg = "white", cex = 1.5)
+
+	# X axis & labels
+	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.2, 0.05))
+	title(xlab = "Climate-Trait PC1\nImportance", line = 4, cex.lab = 1.5)
+
+  # Y labels
+	clip(grconvertX(0, from = "nfc", to = "user"), 
+	     grconvertX(1, from = "nfc", to = "user"),
+	     grconvertY(0, from = "nfc", to = "user"),
+	     grconvertY(1, from = "nfc", to = "user"))
+
+	text(grconvertX(yLabWidth + 2 * panelWidth + 3 / 2 * panelGap, 
+	                from = "in", to = "user"), 
+	     nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
+	     labels = clim2sym[imp2, "climFactor"])
+  text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
+                    max(c(sw1a, sw1b)), from = "in", to = "user"), 
+       nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
+       labels = clim2sym[imp2, "tempScale"])
+	text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
+                    max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), from = "in", 
+	                to = "user"), nImpLat:(nImpLat - nImpPC1 + 1), pos = 4,
+	                labels = parse(text = clim2sym[imp2, "what"]))
+
 	dev.off()
 
-  # Variable selection using random forests (VSURF)
-	# Climate variables that predict PC1 (point and spatially averaged)
-  set.seed(69648)
-  vsurfPC1 <- VSURF(focClim[, climVars], focClim[, "PC1"])
-  vsurfPC1_SA <- VSURF(focClimSA[, climVars], focClimSA[, "PC1"])
-
-##### WORKING ON STUFF HERE - can't finish until SA analysis with 62 km buffer finishes #####
 # x <- match(vsurfPC1$imp.mean.dec.ind, vsurfPC1_SA$imp.mean.dec.ind)
@@
 
-  # Overlap between latitude predictors and phenotype predictors (point measurements)
-  imp2 <- climVars[vsurfPC1$varselect.pred]
-  nImpPC1 <- length(imp2)
-
-  imp1 <- climVars[vsurfLat$varselect.pred]
-  nImpLat <- length(imp1)
-
-  # Figure of variable importance for traits (point measurements)
-	pdf(paste(pathFig, "/Figure_clim2pc1_imp.pdf", sep = ""), 4, 5)
+  # Overlap between latitude predictors and phenotype predictors (spatial aveage)
+
+  # Figure of variable importance for traits (spatially averaged)
+	pdf(paste(pathFig, "/Figure_clim2pc1SA_imp.pdf", sep = ""), 4, 5)
 	par(mar = c(5, 1, 1, 6))
 
-	plot(0, 0, type = "n", xlim = c(0, 0.2), ylim = c(1, nImpLat), 
+	plot(0, 0, type = "n", xlim = c(0, 0.3), ylim = c(1, nImpLat), 
 		axes = F, frame.plot = T, xlab = "", ylab = "")
 	title(xlab = "Climate-Trait PC1\nImportance", line = 4, cex.lab = 1.5)
-	for (i in (nImpLat:(nImpLat - nImpPC1 + 1))) 
+	for (i in (nImpLat:(nImpLat - nImpPC1_SA + 1))) 
 	{
@@
 	}
-	points(vsurfPC1$imp.mean.dec[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), 
+	points(vsurfPC1_SA$imp.mean.dec[1:nImpPC1_SA], 
+	       nImpLat:(nImpLat - nImpPC1_SA + 1), 
 	       pch = 21, col = "black", bg = "white", cex = 1.5)
-	mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, 
-    at = nImpLat:(nImpLat - nImpPC1 + 1), 
-		font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
-	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.2, 0.05))
-  dev.off()
-
-  # Overlap between latitude predictors and phenotype predictors (spatial aveage)
-  imp3 <- climVars[vsurfPC1_SA$varselect.pred]
-  nImpPC1_SA <- length(imp3)
-
-  imp1 <- climVars[vsurfLat$varselect.pred]
-  nImpLat <- length(imp1)
-
-  # Figure of variable importance for traits (point measurements)
-	pdf(paste(pathFig, "/Figure_clim2pc1SA_imp.pdf", sep = ""), 4, 5)
-	par(mar = c(5, 1, 1, 6))
-
-	plot(0, 0, type = "n", xlim = c(0, 0.3), ylim = c(1, nImpLat), 
-		axes = F, frame.plot = T, xlab = "", ylab = "")
-	title(xlab = "Climate-Trait PC1\nImportance", line = 4, cex.lab = 1.5)
-	for (i in (nImpLat:(nImpLat - nImpPC1 + 1))) 
-	{
-		lines(x = c(-100, 100), y = c(i, i), col = "grey")
-	}
-	points(vsurfPC1$imp.mean.dec[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), 
-		pch = 21, col = "black", bg = "white", cex = 1.5)
-	# mtext(climVars[vsurfPC1$varselect.pred], side = 4, las = 1, line = 1, 
-  #  at = nImpLat:(nImpLat - nImpPC1 + 1), 
-	#	font = ifelse(vsurfPC1$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
+	mtext(climVars[vsurfPC1_SA$varselect.pred], side = 4, las = 1, line = 1, 
+    at = nImpLat:(nImpLat - nImpPC1_SA + 1), 
+		font = ifelse(vsurfPC1_SA$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
 	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.3, 0.1))
@@
 										 stringsAsFactors = F, row.names = 1)
-readLines(paste(pathMS, "/climVar2plot.csv", sep = ""))
 
@@
 
-\subsection*{Remarkably little evidence for variation in plasticity}
+\subsection*{Little evidence for variation in plasticity}
 

### 2016-10-12 ms/ms.Rnw
Worked on analyses, results, tables

@@
 %* show PAR similarity
-%* what CO2 concentration did I use?
 %* some sites they are functionally annual
@@
 
-%* check out matthew's blog on hydrology variables
-
 %* explain rhizomes
@@
 % TO DO:
-% 1. Is water regime on figure 1 correct?
-% 2. Finalize Table 2
+% 1. Is water regime on figure 1 correct? (data is in office)
+
 % 3. Do anything with biomass data?
@@
 % 5. Citations for GxE as signature of local adaptation
-% 6. Fix figure 2
 % 7. Supplemental Figure and Table captions
-% 8. Supplemental figure showing agreement between assigned to climatic variables from point estimated and spatially averaged analyses
+% 8. Supplemental figure showing comparison between assigned to climatic variables from point estimated and spatially averaged analyses
 % 9. new spatially averaged analysis
-% 10. new analysis with updated ENM
-% 11. mention using LER_AbsGrowth as predictor in fitSER
 % 12. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
@@
 
-  # Small function to make statistical significance asterisks
-  sigStar <- function(pvalue)
-  {
-    if (pvalue >= 0.05) return("")
-    if (pvalue < 0.05 & pvalue >= 0.01) return("*")
-    if (pvalue < 0.01 & pvalue >= 0.001) return("**")
-    if (pvalue < 0.001) return("***")
-  }
+  # Source custom functions
+  source(paste(pathMS, "/functions.R", sep = ""))
 
@@
 
-\title{Grow with the flow: faster growth is associated with more variable precipitation in a perennial herb}
+\title{Grow with the flow: a latitudinal cline in growth rate is associated with more variable precipitation in \textit{Mimulus cardinals}}
 \author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
@@
 
-Original Article \\
+Running Head: Latitudinal cline and climate in \textit{Mimulus} \\
 
@@
 
-Word counts: TBA \\
-3 figures; 3 tables; ? supporting files \\
-Data will be archived on Dryad upon acceptance. \\
-
-\cdm{Chris' comments}\\
-\ala{Amy's comments}
+Data will be archived on Dryad upon publication. \\
 
@@
 
-\subsection*{Treatments}
+\subsection*{Temperature and drought treatments}
 
@@
   Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
-  Mortality &  \cdm{Probability?}  \\
+  Mortality & probability of death  \\
 	    \bottomrule
@@
 The signifcance of the Population effect was determined using analysis of deviance.
+Note that, unlike other traits discussed below, we did not include Block or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not been applied.
 
@@
 
+We measured growth rate in two phases: leaf expansion as a rosette and shoot elongation after bolting.
 We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
@@
 We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing).
-Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.
+Both leaf expansion and shoot elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.
 
@@
 During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations.
-The youngest, fully-expanded leaf \cdm{this is what I did...not sure exactly which node, as I didn't record that when I did measurements} acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
+The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
 All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
 During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.
+As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out.
+Therefore, we calculated the `intrinsic' photosynthetic rate for each individual as its residual deviance from the regression between stomoatal conductance and photosynthetic rate across all plants.
+We also removed one outlier.
 
@@
 We analyzed the probability of surviving until the end of the experiment as a function of population, treatment, and their interactions using a Generalized Linear Mixed Model (GLMM) assuming binomially distributed errors.
-As before, we included Family and Block as random factors.
+We included Family and Block as random effects.
 We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald \Chi$^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
@@
   library(car)
-
-	# Code for GLM (remove if satisfied with GLMER)
-  # fitMort <- glm(DiedFromStress ~ PopID * TempTrt * WaterTrt,
-  #  data = subset(master, !is.na(master$DiedFromStress)), 
-  #  family = "quasibinomial")
 
@@
 
-\cdm{show correlation between growth rate and biomass?}
+We harvested plants that survived until the end of the experiment from 3 to 16 July (80 to 93 days after sowing).
+We separated individual plants into three biomass fractions: leaves, shoots, and belowground (roots plus rhizomes).
+Belowground fractions were rinsed thoroughly in water to remove sand.
+After completely drying in a forced air oven (Shel Lab, Sheldon Manufacturing Inc., Oregon, USA), we weighed the dry mass of each fraction using an analytical balance.
 
@@
 
-For all traits (Table~\ref{table:Table_traits}) we tested for Population, Treatment, and Population $\times$ Treatment interactions.
-We interpreted significant Population effects to indicate intrinsic variation and Population by Treatment effects to indicate variation in plasticity.
-As mentioned above, we survival and GLM models for germination rate and mortality, respectively.
-For all other traits, we used mixed model ANOVAs with Family included as a random factor.
+For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment, and Population $\times$ Treatment interactions.
+We interpreted significant Population effects to indicate intrinsic variation and Population $\times$ Treatment interactions to indicate variation in plasticity.
+As mentioned above, we used survival and GLMM models for germination rate and mortality, respectively.
+For all other traits, we used mixed model ANOVAs with Family and Block included as random factors.
 Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}.
@@
 Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation.
-Significant Population effect indicate intrinsic trait differences; significant Population $\times$ Treatment effects indicate population differences in plasticity.
-For growth rate, we also accounted for differences in germination rate by including day of germination as a factor.
+To ensure that Population and Treatment effects were specific to a particular growth phase, we included earlier growth phases as a covariate.
+Specifically, we included germination day as a covariate in leaf expansion and shoot elongation analyses; we included germination day and leaf expansion rate as a covariate in shoot elongation analyses.
 
@@
 The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA).
-The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast and slow growing populations.
+The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.
 
@@
 cardinalis} occurrence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
-For brevity, we refer to these as Climate-Latitude and Climate-Trait variables.
-We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}).
+For brevity, we refer to these as `Climate-Latitude' and `Climate-TraitPC1' variables.
+We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
 From VSURF models, we kept only variables selected for prediction, the most stringent criterion.
 
-If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-Trait analyses \ala{and the correlation magnitudes should be high?
+If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high?
 maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}.
@@
 
-To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-Trait variables as above, except that we used spatially averaged climate variables.
+To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-TraitPC1 variables as above, except that we used spatially averaged climate variables.
 \ala{what's the actual prediction?
-that spatially averaged Climate-Trait will have higher correlation than point estimated Climate-Trait?} We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations.
+that spatially averaged Climate-TraitPC1 will have higher correlation than point estimated Climate-Trait?} We sampled climate at 1000 random points (at 90-m resolition) within a 62-km buffer around focal populations.
 We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
+Significant spatial autocorrelation persisted for approximately 62 km.
 Since \textit{M.
@@
 Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
-For clarity, we distinguish between analyses where climate is inferred from a single point (`point estimated Climate-Trait') versus averaged across a 100-km buffer (`spatially averaged Climate-Trait').
+We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km buffer (`spatially averaged Climate-TraitPC1').
 
@@
 	# Calculatins for plot dimensions
-	sw1a <- strwidth(s = clim2sym[imp1, "climFactor"], units = "inches")
-	sw2a <- strwidth(s = clim2sym[imp1, "tempScale"], units = "inches")
-	sw3a <- strwidth(s = parse(text = clim2sym[imp1, "what"]), units = "inches")
-
-	sw1b <- strwidth(s = clim2sym[imp2, "climFactor"], units = "inches")
-	sw2b <- strwidth(s = clim2sym[imp2, "tempScale"], units = "inches")
-	sw3b <- strwidth(s = parse(text = clim2sym[imp2, "what"]), units = "inches")
+	sw1a <- strwidth(s = clim2sym[imp1, "climFactor"], units = "inches", font = 4)
+	sw2a <- strwidth(s = clim2sym[imp1, "tempScale"], units = "inches", font = 4)
+	sw3a <- strwidth(s = parse(text = clim2sym[imp1, "what"]), units = "inches",
+	                 font = 4)
+
+	sw1b <- strwidth(s = clim2sym[imp2, "climFactor"], units = "inches", font = 4)
+	sw2b <- strwidth(s = clim2sym[imp2, "tempScale"], units = "inches", font = 4)
+	sw3b <- strwidth(s = parse(text = clim2sym[imp2, "what"]), units = "inches",
+	                 font = 4)
 
@@
 	text(grconvertX(panelGap / 2, from = "in", to = "user"), 
-	     nImpLat:1, pos = 4, labels = clim2sym[imp1, "climFactor"])
+	     nImpLat:1, pos = 4, labels = clim2sym[imp1, "climFactor"],
+	     font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))
   text(grconvertX(panelGap + max(c(sw1a, sw1b)), from = "in", to = "user"), 
-       nImpLat:1, pos = 4, labels = clim2sym[imp1, "tempScale"])
+       nImpLat:1, pos = 4, labels = clim2sym[imp1, "tempScale"],
+       font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))
 	text(grconvertX(3 / 2 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), 
 	                               from = "in", to = "user"), nImpLat:1, pos = 4, 
-	                labels = parse(text = clim2sym[imp1, "what"]))
+	     labels = parse(text = clim2sym[imp1, "what"]),
+	     font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))
 
@@
 	     nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
-	     labels = clim2sym[imp2, "climFactor"])
+	     labels = clim2sym[imp2, "climFactor"],
+	     font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
   text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
@@
        nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
-       labels = clim2sym[imp2, "tempScale"])
+       labels = clim2sym[imp2, "tempScale"],
+       font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
 	text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
                     max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), from = "in", 
-	                to = "user"), nImpLat:(nImpLat - nImpPC1 + 1), pos = 4,
-	                labels = parse(text = clim2sym[imp2, "what"]))
+	                to = "user"), nImpLat:(nImpLat - nImpPC1 + 1), 
+	     pos = 4, labels = parse(text = clim2sym[imp2, "what"]),
+	     font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
 
@@
 These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
-Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
+For each occurence, we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
+Monthly interpolated climate layers were calculated using ClimateWNA version 5.30 \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
@@
 Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis}, as evidenced by large and highly significant population effects (Table~\ref{table:Table_AnovaSummary}).
+cardinalis}, as evidenced by large and significant population effects (Table~\ref{table:Table_AnovaSummary}).
 A single principal component captured \Sexpr{gsub("%", "", pve)}\% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
-As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
+As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Table of Population, Treatment, Population x Treatment interactions for varies traits
+% Table of Population, Treatment, Population x Treatment interactions
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
-Note that temperature and water treatments were imposed after germination, hence are not application to this trait.
+Note that temperature and water treatments were imposed after germination, hence are not applicable to this trait.
+Complete analysis of variance/deviance tables for each trait are available in the Supporting Information.
 Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}
@@
 	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude.
-Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods).
-The line and 95\% confidence intervals were estimated using standard linear regression.}
+Each point is a population's latitude of origin (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of five traits (see Methods).
+The line and 95\% confidence intervals were estimated using linear regression.}
 	\label{fig:Fig_PC1vLat}
@@
 
-Genotype $\times$ environment interactions are also a common signature of local adaptation.
+Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation.
 For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
-We found remarkably little evidence for this pattern in \textit{M.
+We found little evidence G$\times$E in \textit{M.
 cardinalis}.
-There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
-Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
+\cdm{Is this result still correct after revising analysis?} There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
+Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
 Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
@@
 % ANOVA table for LER (need to label tables)
+
+<<TableS_fitLER, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+	  source(paste(pathMS, "/functions.R", sep = ""))
+
+  ss <- sigDig(fitLER$anova.table$`Sum Sq`)
+  ms <- sigDig(fitLER$anova.table$`Mean Sq`)
+  df1 <- sigDig(fitLER$anova.table$NumDF)
+  df2 <- sigDig(fitLER$anova.table$DenDF)
+  fval <- sigDig(fitLER$anova.table$F.value)
+ pval <- pval2latex(fitLER$anova.table$`Pr(>F)`)
+@
+
 \begin{table}[htbp]
@@
 
-	\input{Tables/Table_lllGrowthAnova.txt}
+                     & SS & MS & df1 & df2 & \em{F} & $P$ \\ \midrule
+  Day of Germination & 18.2 & 18.2 & 1 & 633.1 & 35.2 & 4.78 $\times10^{-9}$ \\
+mmLER
+
+Population & 22.3 & 1.5 & 15 & 121.5 & 4.3 & 2.57 $\times10^{-6}$ \\
+Temperature & 114.4 & 114.4 & 1 & 567.7 & 352.3 & 1.64 $\times10^{-61}$ \\
+Water & 4.2 & 4.2 & 1 & 567.7 & 12.6 & 4.12 $\times10^{-4}$ \\
+Population $\times$ Temperature & 3.1 & 0.2 & 15 & 553.3 & 0.6 & 0.86 \\
+Population $\times$ Water & 8.9 & 0.6 & 15 & 567.8 & 1.8 & 0.03 \\
+Temperature $\times$ Water & 0 & 0 & 1 & 550.5 & 0 & 0.99 \\
+Population $\times$ Temperature $\times$ Water & 4 & 0.3 & 15 & 535.4 & 0.8 & 0.69 \\
 
@@
 	\end{tabular}
+	\label{table:TableS_fitLER}
 	\end{center}

### 2016-10-12 ms/functions.R

@@ -1,0 +1,34 @@
+# R functions associated with Muir and Angert 2016
+
+# Make statistical significance asterisks
+sigStar <- function(pvalue)
+{
+  if (pvalue >= 0.05) return("")
+  if (pvalue < 0.05 & pvalue >= 0.01) return("*")
+  if (pvalue < 0.01 & pvalue >= 0.001) return("**")
+  if (pvalue < 0.001) return("***")
+}
+
+# Determine a number's order of magnitude
+oom <- function(x)
+{
+  # x should be a vector of numbers
+  floor(log10(abs(x)))
+}
+
+# Determine number of significant digits to use for rounding in tables
+sigDig <- function(x)
+{
+  # x should be a vector of numeric elements to be rounded to same significant digit
+  ret <- round(x, -min(oom(x)))
+  return(ret)
+}
+
+# Text string for p-value in tables
+pval2latex <- function(x)
+{
+  ret <- ifelse(oom(x) > -3, 
+                sprintf("%.*f", 3, x), 
+                sprintf("%.*f $\times10^{%s}$", 1, stem, oom(x)))
+  return(ret)
+}

### 2016-10-13 ms/ms.Rnw
ANOVA tables and more

@@
 
-  mm <- lmer(SER_AbsGrowth ~ AvgGermDay + LER_AbsGrowth + 
+	# Mixed model including LER_AbsGrowth as covariate
+  # mm <- lmer(SER_AbsGrowth ~ AvgGermDay + LER_AbsGrowth + 
+  #             PopID * TempTrt * WaterTrt + (1|Family) + (1|Block), 
+  #           data = master)
+	# fitSER <- step(mm, reduce.random = F) #includes LER_AbsGrowth, pop, temp, h2o, and temp x h2o
+
+	# Mixed model without LER_AbsGrowth as covariate
+  mm <- lmer(SER_AbsGrowth ~ AvgGermDay + 
                PopID * TempTrt * WaterTrt + (1|Family) + (1|Block), 
              data = master)
-	fitSER <- step(mm, reduce.random = F) #includes LER_AbsGrowth, pop, temp, h2o, and temp x h2o
+	fitSER <- step(mm, reduce.random = F)
 
@@
 
-  # Photosynthetic rate
+  # Photosynthetic rate (remove is satisfied w/ resPhoto analysis)
   mm <- lmer(Photo ~ PopID * TempTrt * WaterTrt + (1|Family) + (1|Block),
@@
 We included Family and Block as random effects.
-We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald \Chi$^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
+We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
 
@@
   library(VSURF)
+
+	# Load climVars
+	load(paste(pathDat, "/ExploratoryClimateVariables.RData", sep = ""))
+
   # Load VSURF results of climatic variables associated with latitude at cardinalis occurence points
@@
   nImpPC1_SA <- length(imp3)
-
-	# Load climVars
-	load(paste(pathDat, "/ExploratoryClimateVariables.RData", sep = ""))
 
@@
       \toprule
-  Climate variable & Abbreviation \\
+   Abbreviation & Climate variable \\
       \midrule
@@
 	Eref	& Hargreaves reference evaporation (mm) \\
-	CMD	  & Hargreaves climativ moisture deficit (mm) \\
-	MAR   &	mean annual solar radiation (NOTE: removing because too many missing values) \\
+	CMD	  & Hargreaves climatic moisture deficit (mm) \\
 	RH    &	mean annual relative humidity \\
@@
 
-% ANOVA table for LER (need to label tables)
+% ANOVA table for LER
 
 <<TableS_fitLER, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  # this is sourced above. remove from here once testing is over  
   source(paste(pathMS, "/functions.R", sep = ""))
 
-  ss <- sigDig(fitLER$anova.table$`Sum Sq`)
-  ms <- sigDig(fitLER$anova.table$`Mean Sq`)
-  df1 <- sigDig(fitLER$anova.table$NumDF)
-  df2 <- sigDig(fitLER$anova.table$DenDF)
-  fval <- sigDig(fitLER$anova.table$F.value)
- pval <- pval2latex(fitLER$anova.table$`Pr(>F)`)
+  # Convert numbers from R to LaTeX characters
+  lerTab <- data.frame(ss = sigDig(fitLER$anova.table$`Sum Sq`),
+                       ms = sigDig(fitLER$anova.table$`Mean Sq`),
+                       df1 = sigDig(fitLER$anova.table$NumDF),
+                       df2 = sigDig(fitLER$anova.table$DenDF),
+                       fval = sigDig(fitLER$anova.table$F.value),
+                       pval = pval2latex(fitLER$anova.table$`Pr(>F)`),
+                       stringsAsFactors = F)
+  rownames(lerTab) <- rownames(fitLER$anova.table)
+
 @
@@
 	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{CAPTION}
+	\caption[ANOVA table, leaf expansion rate]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2); }
 	\begin{center}
@@
 
-                     & SS & MS & df1 & df2 & \em{F} & $P$ \\ \midrule
-  Day of Germination & 18.2 & 18.2 & 1 & 633.1 & 35.2 & 4.78 $\times10^{-9}$ \\
-mmLER
-
-Population & 22.3 & 1.5 & 15 & 121.5 & 4.3 & 2.57 $\times10^{-6}$ \\
-Temperature & 114.4 & 114.4 & 1 & 567.7 & 352.3 & 1.64 $\times10^{-61}$ \\
-Water & 4.2 & 4.2 & 1 & 567.7 & 12.6 & 4.12 $\times10^{-4}$ \\
-Population $\times$ Temperature & 3.1 & 0.2 & 15 & 553.3 & 0.6 & 0.86 \\
-Population $\times$ Water & 8.9 & 0.6 & 15 & 567.8 & 1.8 & 0.03 \\
-Temperature $\times$ Water & 0 & 0 & 1 & 550.5 & 0 & 0.99 \\
-Population $\times$ Temperature $\times$ Water & 4 & 0.3 & 15 & 535.4 & 0.8 & 0.69 \\
-
+  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
+  Day to Germination & \Sexpr{lerTab["AvgGermDay", "ss"]} & 
+    \Sexpr{lerTab["AvgGermDay", "ms"]} & \Sexpr{lerTab["AvgGermDay", "df1"]} & 
+    \Sexpr{lerTab["AvgGermDay", "df2"]} & \Sexpr{lerTab["AvgGermDay", "fval"]} &
+    \Sexpr{lerTab["AvgGermDay", "pval"]} \\
+  Population & \Sexpr{lerTab["PopID", "ss"]} & \Sexpr{lerTab["PopID", "ms"]} & 
+    \Sexpr{lerTab["PopID", "df1"]} & \Sexpr{lerTab["PopID", "df2"]} & 
+    \Sexpr{lerTab["PopID", "fval"]} & \Sexpr{lerTab["PopID", "pval"]} \\
+  Temperature & \Sexpr{lerTab["TempTrt", "ss"]} & 
+    \Sexpr{lerTab["TempTrt", "ms"]} & \Sexpr{lerTab["TempTrt", "df1"]} &
+    \Sexpr{lerTab["TempTrt", "df2"]} & \Sexpr{lerTab["TempTrt", "fval"]} &
+    \Sexpr{lerTab["TempTrt", "pval"]} \\
+  Water & \Sexpr{lerTab["WaterTrt", "ss"]} & \Sexpr{lerTab["WaterTrt", "ms"]} & 
+    \Sexpr{lerTab["WaterTrt", "df1"]} & \Sexpr{lerTab["WaterTrt", "df2"]} & 
+    \Sexpr{lerTab["WaterTrt", "fval"]} & \Sexpr{lerTab["WaterTrt", "pval"]} \\
+  Population $\times$ Temperature & \Sexpr{lerTab["PopID:TempTrt", "ss"]} & 
+    \Sexpr{lerTab["PopID:TempTrt", "ms"]} & 
+    \Sexpr{lerTab["PopID:TempTrt", "df1"]} & 
+    \Sexpr{lerTab["PopID:TempTrt", "df2"]} & 
+    \Sexpr{lerTab["PopID:TempTrt", "fval"]} &
+    \Sexpr{lerTab["PopID:TempTrt", "pval"]} \\
+  Population $\times$ Water & \Sexpr{lerTab["PopID:WaterTrt", "ss"]} & 
+    \Sexpr{lerTab["PopID:WaterTrt", "ms"]} & 
+    \Sexpr{lerTab["PopID:WaterTrt", "df1"]} &
+    \Sexpr{lerTab["PopID:WaterTrt", "df2"]} &
+    \Sexpr{lerTab["PopID:WaterTrt", "fval"]} & 
+    \Sexpr{lerTab["PopID:WaterTrt", "pval"]} \\
+  Population $\times$ Temperature $\times$ Water & 
+    \Sexpr{lerTab["PopID:TempTrt:WaterTrt", "ss"]} & 
+    \Sexpr{lerTab["PopID:TempTrt:WaterTrt", "ms"]} & 
+    \Sexpr{lerTab["PopID:TempTrt:WaterTrt", "df1"]} & 
+    \Sexpr{lerTab["PopID:TempTrt:WaterTrt", "df2"]} & 
+    \Sexpr{lerTab["PopID:TempTrt:WaterTrt", "fval"]} & 
+    \Sexpr{lerTab["PopID:TempTrt:WaterTrt", "pval"]} \\
 	\bottomrule
@@
 
-% ANOVA table for SER (need to label tables)
+% ANOVA table for SER
+
+<<TableS_fitSER, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  # Convert numbers from R to LaTeX characters
+  serTab <- data.frame(ss = sigDig(fitSER$anova.table$`Sum Sq`),
+                       ms = sigDig(fitSER$anova.table$`Mean Sq`),
+                       df1 = sigDig(fitSER$anova.table$NumDF),
+                       df2 = sigDig(fitSER$anova.table$DenDF),
+                       fval = sigDig(fitSER$anova.table$F.value),
+                       pval = pval2latex(fitSER$anova.table$`Pr(>F)`),
+                       stringsAsFactors = F)
+  rownames(serTab) <- rownames(fitSER$anova.table)
+
+@
+
 \begin{table}[htbp]
@@
 	\selectfont
-	\caption[ANOVA table, shoot elongation rate]{CAPTION}
+	\caption[ANOVA table, shoot elongation rate]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
 	\begin{center}
@@
 
-	\input{Tables/Table_heightGrowthAnova.txt}
-
+  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
+  Day to Germination & \Sexpr{serTab["AvgGermDay", "ss"]} & 
+    \Sexpr{serTab["AvgGermDay", "ms"]} & \Sexpr{serTab["AvgGermDay", "df1"]} & 
+    \Sexpr{serTab["AvgGermDay", "df2"]} & \Sexpr{serTab["AvgGermDay", "fval"]} &
+    \Sexpr{serTab["AvgGermDay", "pval"]} \\
+  Population & \Sexpr{serTab["PopID", "ss"]} & \Sexpr{serTab["PopID", "ms"]} & 
+    \Sexpr{serTab["PopID", "df1"]} & \Sexpr{serTab["PopID", "df2"]} & 
+    \Sexpr{serTab["PopID", "fval"]} & \Sexpr{serTab["PopID", "pval"]} \\
+  Temperature & \Sexpr{serTab["TempTrt", "ss"]} & 
+    \Sexpr{serTab["TempTrt", "ms"]} & \Sexpr{serTab["TempTrt", "df1"]} &
+    \Sexpr{serTab["TempTrt", "df2"]} & \Sexpr{serTab["TempTrt", "fval"]} &
+    \Sexpr{serTab["TempTrt", "pval"]} \\
+  Water & \Sexpr{serTab["WaterTrt", "ss"]} & \Sexpr{serTab["WaterTrt", "ms"]} & 
+    \Sexpr{serTab["WaterTrt", "df1"]} & \Sexpr{serTab["WaterTrt", "df2"]} & 
+    \Sexpr{serTab["WaterTrt", "fval"]} & \Sexpr{serTab["WaterTrt", "pval"]} \\
+  Population $\times$ Temperature & \Sexpr{serTab["PopID:TempTrt", "ss"]} & 
+    \Sexpr{serTab["PopID:TempTrt", "ms"]} & 
+    \Sexpr{serTab["PopID:TempTrt", "df1"]} & 
+    \Sexpr{serTab["PopID:TempTrt", "df2"]} & 
+    \Sexpr{serTab["PopID:TempTrt", "fval"]} &
+    \Sexpr{serTab["PopID:TempTrt", "pval"]} \\
+  Population $\times$ Water & \Sexpr{serTab["PopID:WaterTrt", "ss"]} & 
+    \Sexpr{serTab["PopID:WaterTrt", "ms"]} & 
+    \Sexpr{serTab["PopID:WaterTrt", "df1"]} &
+    \Sexpr{serTab["PopID:WaterTrt", "df2"]} &
+    \Sexpr{serTab["PopID:WaterTrt", "fval"]} & 
+    \Sexpr{serTab["PopID:WaterTrt", "pval"]} \\
+  Population $\times$ Temperature $\times$ Water & 
+    \Sexpr{serTab["PopID:TempTrt:WaterTrt", "ss"]} & 
+    \Sexpr{serTab["PopID:TempTrt:WaterTrt", "ms"]} & 
+    \Sexpr{serTab["PopID:TempTrt:WaterTrt", "df1"]} & 
+    \Sexpr{serTab["PopID:TempTrt:WaterTrt", "df2"]} & 
+    \Sexpr{serTab["PopID:TempTrt:WaterTrt", "fval"]} & 
+    \Sexpr{serTab["PopID:TempTrt:WaterTrt", "pval"]} \\
 	\bottomrule
 	\end{tabular}
+	\label{table:TableS_fitSER}
+	\end{center}
+\end{table}
+
+% ANOVA table for photosynthesis
+
+<<TableS_fitPhoto, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  # Convert numbers from R to LaTeX characters
+  photoTab <- data.frame(ss = sigDig(fitPhoto$anova.table$`Sum Sq`),
+                         ms = sigDig(fitPhoto$anova.table$`Mean Sq`),
+                         df1 = sigDig(fitPhoto$anova.table$NumDF),
+                         df2 = sigDig(fitPhoto$anova.table$DenDF),
+                         fval = sigDig(fitPhoto$anova.table$F.value),
+                         pval = pval2latex(fitPhoto$anova.table$`Pr(>F)`),
+                         stringsAsFactors = F)
+  rownames(photoTab) <- rownames(fitPhoto$anova.table)
+
+@
+
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, photosynthesis]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
+	\begin{center}
+	\begin{tabular}{lcccccc}
+	\toprule
+
+  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
+  Population & \Sexpr{photoTab["PopID", "ss"]} & 
+    \Sexpr{photoTab["PopID", "ms"]} & \Sexpr{photoTab["PopID", "df1"]} &
+    \Sexpr{photoTab["PopID", "df2"]} & \Sexpr{photoTab["PopID", "fval"]} &
+    \Sexpr{photoTab["PopID", "pval"]} \\
+  Temperature & \Sexpr{photoTab["TempTrt", "ss"]} & 
+    \Sexpr{photoTab["TempTrt", "ms"]} & \Sexpr{photoTab["TempTrt", "df1"]} &
+    \Sexpr{photoTab["TempTrt", "df2"]} & \Sexpr{photoTab["TempTrt", "fval"]} &
+    \Sexpr{photoTab["TempTrt", "pval"]} \\
+  Water & \Sexpr{photoTab["WaterTrt", "ss"]} & 
+    \Sexpr{photoTab["WaterTrt", "ms"]} & \Sexpr{photoTab["WaterTrt", "df1"]} &
+    \Sexpr{photoTab["WaterTrt", "df2"]} & \Sexpr{photoTab["WaterTrt", "fval"]} &
+    \Sexpr{photoTab["WaterTrt", "pval"]} \\
+  Population $\times$ Temperature & \Sexpr{photoTab["PopID:TempTrt", "ss"]} & 
+    \Sexpr{photoTab["PopID:TempTrt", "ms"]} & 
+    \Sexpr{photoTab["PopID:TempTrt", "df1"]} & 
+    \Sexpr{photoTab["PopID:TempTrt", "df2"]} & 
+    \Sexpr{photoTab["PopID:TempTrt", "fval"]} &
+    \Sexpr{photoTab["PopID:TempTrt", "pval"]} \\
+  Population $\times$ Water & \Sexpr{photoTab["PopID:WaterTrt", "ss"]} & 
+    \Sexpr{photoTab["PopID:WaterTrt", "ms"]} & 
+    \Sexpr{photoTab["PopID:WaterTrt", "df1"]} &
+    \Sexpr{photoTab["PopID:WaterTrt", "df2"]} &
+    \Sexpr{photoTab["PopID:WaterTrt", "fval"]} & 
+    \Sexpr{photoTab["PopID:WaterTrt", "pval"]} \\
+  Population $\times$ Temperature $\times$ Water & 
+    \Sexpr{photoTab["PopID:TempTrt:WaterTrt", "ss"]} & 
+    \Sexpr{photoTab["PopID:TempTrt:WaterTrt", "ms"]} & 
+    \Sexpr{photoTab["PopID:TempTrt:WaterTrt", "df1"]} & 
+    \Sexpr{photoTab["PopID:TempTrt:WaterTrt", "df2"]} & 
+    \Sexpr{photoTab["PopID:TempTrt:WaterTrt", "fval"]} & 
+    \Sexpr{photoTab["PopID:TempTrt:WaterTrt", "pval"]} \\
+	\bottomrule
+	\end{tabular}
+	\label{table:TableS_fitPhoto}
+	\end{center}
+\end{table}
+
+% ANOVA table for intrinsic photosynthetic rate
+
+<<TableS_fitIntrnPhoto, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  # Convert numbers from R to LaTeX characters
+  intrnPhotoTab <- data.frame(ss = sigDig(fitIntrnPhoto$anova.table$`Sum Sq`),
+                              ms = sigDig(fitIntrnPhoto$anova.table$`Mean Sq`),
+                              df1 = sigDig(fitIntrnPhoto$anova.table$NumDF),
+                              df2 = sigDig(fitIntrnPhoto$anova.table$DenDF),
+                              fval = sigDig(fitIntrnPhoto$anova.table$F.value),
+                              pval = 
+                                pval2latex(fitIntrnPhoto$anova.table$`Pr(>F)`),
+                              stringsAsFactors = F)
+  rownames(intrnPhotoTab) <- rownames(fitIntrnPhoto$anova.table)
+
+@
+
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, photosynthesis]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
+	\begin{center}
+	\begin{tabular}{lcccccc}
+	\toprule
+
+  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
+  Population & \Sexpr{photoTab["PopID", "ss"]} & 
+    \Sexpr{intrnPhotoTab["PopID", "ms"]} & 
+    \Sexpr{intrnPhotoTab["PopID", "df1"]} &
+    \Sexpr{intrnPhotoTab["PopID", "df2"]} & 
+    \Sexpr{intrnPhotoTab["PopID", "fval"]} &
+    \Sexpr{intrnPhotoTab["PopID", "pval"]} \\
+  Temperature & \Sexpr{intrnPhotoTab["TempTrt", "ss"]} & 
+    \Sexpr{intrnPhotoTab["TempTrt", "ms"]} & 
+    \Sexpr{intrnPhotoTab["TempTrt", "df1"]} &
+    \Sexpr{intrnPhotoTab["TempTrt", "df2"]} & 
+    \Sexpr{intrnPhotoTab["TempTrt", "fval"]} &
+    \Sexpr{intrnPhotoTab["TempTrt", "pval"]} \\
+  Water & \Sexpr{intrnPhotoTab["WaterTrt", "ss"]} & 
+    \Sexpr{intrnPhotoTab["WaterTrt", "ms"]} & 
+    \Sexpr{intrnPhotoTab["WaterTrt", "df1"]} &
+    \Sexpr{intrnPhotoTab["WaterTrt", "df2"]} & 
+    \Sexpr{intrnPhotoTab["WaterTrt", "fval"]} &
+    \Sexpr{intrnPhotoTab["WaterTrt", "pval"]} \\
+  Population $\times$ Temperature & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt", "ss"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt", "ms"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt", "df1"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt", "df2"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt", "fval"]} &
+    \Sexpr{intrnPhotoTab["PopID:TempTrt", "pval"]} \\
+  Population $\times$ Water & \Sexpr{intrnPhotoTab["PopID:WaterTrt", "ss"]} & 
+    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "ms"]} & 
+    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "df1"]} &
+    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "df2"]} &
+    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "fval"]} & 
+    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "pval"]} \\
+  Population $\times$ Temperature $\times$ Water & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "ss"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "ms"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "df1"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "df2"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "fval"]} & 
+    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "pval"]} \\
+	\bottomrule
+	\end{tabular}
+	\label{table:TableS_fitIntrnPhoto}
+	\end{center}
+\end{table}
+
+% Analysis of deviance table for mortality
+
+<<TableS_fitPhoto, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  # Convert numbers from R to LaTeX characters
+  tab <- Anova(fitMort, type = 2)
+  mortTab <- data.frame(chisq = sigDig(tab$Chisq), df = tab$Df,
+                         pval = pval2latex(tab$`Pr(>Chisq)`),
+                         stringsAsFactors = F)
+  rownames(mortTab) <- rownames(tab)
+
+@
+
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, mortality]{CAPTION}
+	\begin{center}
+	\begin{tabular}{lccc}
+	\toprule
+
+  & $\chi ^2$ & df & $P$-value \\ \midrule
+  Population & \Sexpr{mortTab["PopID", "chisq"]} & 
+    \Sexpr{mortTab["PopID", "df"]} & \Sexpr{mortTab["PopID", "pval"]} & \\
+  Temperature & \Sexpr{mortTab["TempTrt", "chisq"]} & 
+    \Sexpr{mortTab["TempTrt", "df"]} & \Sexpr{mortTab["TempTrt", "pval"]} & \\
+  Water & \Sexpr{mortTab["WaterTrt", "chisq"]} & 
+    \Sexpr{mortTab["WaterTrt", "df"]} & \Sexpr{mortTab["WaterTrt", "pval"]} & \\
+  Population $\times$ Temperature & \Sexpr{mortTab["PopID:TempTrt", "chisq"]} & 
+    \Sexpr{mortTab["PopID:TempTrt", "df"]} & 
+    \Sexpr{mortTab["PopID:TempTrt", "pval"]} & \\
+  Population $\times$ Water & \Sexpr{mortTab["PopID:WaterTrt", "chisq"]} & 
+    \Sexpr{mortTab["PopID:WaterTrt", "df"]} & 
+    \Sexpr{mortTab["PopID:WaterTrt", "pval"]} & \\
+  Population $\times$ Temperature $\times$ Water & 
+    \Sexpr{mortTab["PopID:TempTrt:WaterTrt", "chisq"]} & 
+    \Sexpr{mortTab["PopID:TempTrt:WaterTrt", "df"]} & 
+    \Sexpr{mortTab["PopID:TempTrt:WaterTrt", "pval"]} & \\	\bottomrule
+	\end{tabular}
+	\label{table:TableS_fitMort}
 	\end{center}

### 2016-10-13 ms/ms.tex
ANOVA tables and more

--- 2016-01-20 ms/ms.tex
@@
+%lost on page 11 - extracted and used climate variables
+%Maybe reordering stuff on paragraph
+%cartoon of whole procedure for climate variables
+%figure showing correlation b/t bio12 and 
+%state that VSURF importance is different scales in different models
+%moved up three hypotheses part to introduction
+%play up lack of GxE and ???
+%** correct # of Pop x Treatment
+
+%* delve into interactions - do they change cline?
+
+%* show figures of reaction norms
+%* split plot, account for error variation in temp???
+%* show PAR similarity
+%* some sites they are functionally annual
+
+%* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
+%	- maybe temporal source-sink makes more sense
+
+%* explain rhizomes
+
+%* put in biomass allocation cline
 % TO DO:
-% 1. Is water regime on figure 1 correct?
-% 2. Finalize Table 2
+% 1. Is water regime on figure 1 correct? (data is in office)
+
 % 3. Do anything with biomass data?
@@
 % 5. Citations for GxE as signature of local adaptation
-% 6. Fix figure 2
 % 7. Supplemental Figure and Table captions
-% 8. Supplemental figure showing agreement between assigned to climatic variables from point estimated and spatially averaged analyses
+% 8. Supplemental figure showing comparison between assigned to climatic variables from point estimated and spatially averaged analyses
+% 9. new spatially averaged analysis
+% 12. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
 
@@
 % Intrinsic versus plastic variation 
-There are two basic approaches one can use to identify candidate traits underlying local adaptation in a common garden.
-First, if genetically-based trait differences between populations vary clinally with environmental differences, this may point to traits important for local adaptation.
-Second, genotype by environment interactions could indicate that variation in plasticity mediates local adaptation.
-We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively.
-There are classic cases adaptation involving both intrinsic and plastic trait variation.
+When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004} \ala{added citation - might give error}.
+The traits that underlie local adaptation, however, need not mirror this pattern.
+Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away.
+Alternatively, genotype-by-environment interactions could indicate that variation in plasticity mediates local adaptation.
+We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively.
+There are classic cases of adaptation involving both intrinsic and plastic trait variation.
 For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
-Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
-
-% Selective agents
+Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
+In plants especially, we know little about the prevalance and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
+
+% Selective agents (trait-enviro assoc)
+A basic approach one can use to identify candidate traits underlying local adaptation is to find associations between traits and environments.
 Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
@@
 Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
-Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely related to climatic variation.
-Despite the fact that latitudinal clines in particular have been studied for a long time, the exact climatic factors (e.g.
-minimum or maximum temperature, growing season length, strength of biotic interactions) is rarely known.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely relate to climatic variation.
+
+Despite the fact that latitudinal clines in particular have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
+Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days, such that latitude is an effective proxy for the underlying climatic driver but nonetheless  key climatic variable(s) with stronger relationships to traits exist.
+Alternatively, latitude may be more strongly related to traits than any single climatic variable for two possible reasons.
+First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline.
+Alternatively, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
+For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a Southern set of populations will experience warmer climate than a Northern one.
+Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g.
+high elevation) site.
 Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g.
 coastal vs.
-continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}.
-In plants especially, we know little about the prevelance and adaptive significance of clinal variation in fundamental physiological traits like photosynthesis and the impact on plant performance.
+continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local or regionally-averaged variables are more predictive of trait change.
+\ala{added this last clause to try to encompass gene flow idea, but now maybe the sentence is too clunky}
 
@@
 % Motivate primary questions using Mimulus
+% might add that Anacker and Strauss 2014 showed ecological shifts important in Mim speciation and we would like to know driving forces behind this???
 In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}.
-We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions.
+We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
 Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
 Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}.
-We measured genetic and genotype by environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
+We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
 We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity.
-Interannual variation in precipitaiton and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+Interannual variation in precipitation and/or precipitation seasonality is associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
 cardinalis}.
 We place these findings in the context of life history theory and consider future directions in the Discussion.
+\ala{wrap spatial buffer results into summary?}
 
@@
 During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of 329 plants evenly spread between treatments and families within populations.
-The youngest, fully-expanded leaf \cdm{this is what I did...not sure exactly which node, as I didn't record that when I did measurements} acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
-All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity.
+The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
+All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
 During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.
+As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out.
+Therefore, we calculated the `intrinsic' photosynthetic rate for each individual as its residual deviance from the regression between stomoatal conductance and photosynthetic rate across all plants.
+We also removed one outlier.
 
@@
 We assayed mortality during twice-weekly growth measurements.
-We could not get GLMM with Family effects to converge, so we used GLM with a quasibinomial error structure and assessed signifiance using Type 2 Analysis of Deviance with the R package \pkg{car}.
+We analyzed the probability of surviving until the end of the experiment as a function of population, treatment, and their interactions using a Generalized Linear Mixed Model (GLMM) assuming binomially distributed errors.
+We included Family and Block as random effects.
+We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
 
@@
 
-We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results).
-Since latitude \textit{per se} cannot be a selective agent, we considered three possible explanations.
-First, latitude may be strongly correlated, in this species, with one or two climatic variables, such as temperature, precipitation, or growing degree-days.
-Second, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latiduinal cline.
-Third, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
-For example, average temperature of \textit{M.
-cardinalis} populations at a given latitude varies widely, but in aggregate, a Southern metapopulation experiences warmer climate than a Northern one.
-Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g.
-high elevation) site.
-
-To evaluate support for these hypotheses, we used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
 cardinalis}.
 We looked for overlap between climatic variables that best predict latitude of \textit{M.
-cadinalis} occurence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
-For brevity, we refer to these as Climate-Latitude and Climate-Trait variables.
-We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}).
+cardinalis} occurrence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
+For brevity, we refer to these as `Climate-Latitude' and `Climate-TraitPC1' variables.
+We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
 From VSURF models, we kept only variables selected for prediction, the most stringent criterion.
 
-The first hypothesis predicts that there should be one or two Climate-Latitude and Climate-Trait associations that are strongly correlated, whereas the second hypothesis predicts that several climatic variables should be weakly correlated.
-To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-Trait variables as above, except that we used spatially averaged climate variables.
-We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations.
+If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high?
+maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}.
+If latitude is correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline, then several climatic variables should be weakly correlated \ala{fix this to mirror prior}.
+
+To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-TraitPC1 variables as above, except that we used spatially averaged climate variables.
+\ala{what's the actual prediction?
+that spatially averaged Climate-TraitPC1 will have higher correlation than point estimated Climate-Trait?} We sampled climate at 1000 random points (at 90-m resolition) within a 62-km buffer around focal populations.
 We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
+Significant spatial autocorrelation persisted for approximately 62 km.
 Since \textit{M.
@@
 Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis}, as evidenced by large and highly significant population effects (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured 74.2 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
-As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Table of Population, Treatment, Population x Treatment interactions for varies traits
+cardinalis}, as evidenced by large and significant population effects (Table~\ref{table:Table_AnovaSummary}).
+A single principal component captured 74.4 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
+As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of Population, Treatment, Population x Treatment interactions
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
    \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects.
-We used different statistical modeling for the diverse traits assayed -- glm: generalized linear model using R \citep{R}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
-Note that temperature and water treatments were imposed after germination, hence are not application to this trait.
+We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
+Note that temperature and water treatments were imposed after germination, hence are not applicable to this trait.
+Complete analysis of variance/deviance tables for each trait are available in the Supporting Information.
 Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}
@@
     *** & 
-    *** & \\
+     & \\
   Temperature                      &
@@
     *** &
-    *** & 
+    ** & 
      &
@@
     NA                                                          &
-    ***   &
-    ***   &
-    *** & 
+    *   &
+       &
+     & 
      &     
@@
      &     
-    * & \\
+     & \\
   Pop $\times$ Water               &
@@
      &
-    *** &
+     &
      & 
@@
 
-\subsection*{Remarkably little evidence for variation in plasticity}
-
-Genotype $\times$ environment interactions are also a common signature of local adaptation.
+\subsection*{Little evidence for variation in plasticity}
+
+Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation.
 For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
-We found remarkably little evidence for this pattern in \textit{M.
+We found little evidence G$\times$E in \textit{M.
 cardinalis}.
-There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
-Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
+\cdm{Is this result still correct after revising analysis?} There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
+Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
 Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
@@
 
-% ANOVA table for LER (need to label tables)
+% ANOVA table for LER
+
 \begin{table}[htbp]
@@
 
-	\input{Tables/Table_lllGrowthAnova.txt}
-
+  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
+  Day to Germination & 12.12 & 
+    12.12 & 1 & 
+    637 & 35.21 &
+    4.9 $\times10^{-9}$ \\
+  Population & 22.22 & 1.48 & 
+    15 & 118 & 
+    4.3 & 2.5 $\times10^{-6}$ \\
+  Temperature & 80.42 & 
+    80.42 & 1 &
+    5 & 233.61 &
+    2.6 $\times10^{-5}$ \\
+  Water & 4.1 & 4.1 & 
+    1 & 5 & 
+    11.92 & 0.019 \\
+  Population $\times$ Temperature & 2.76 & 
+    0.18 & 
+    15 & 
+    547 & 
+    0.53 &
+    0.925 \\
+  Population $\times$ Water & 9.66 & 
+    0.64 & 
+    15 &
+    562 &
+    1.87 & 
+    0.024 \\
+  Population $\times$ Temperature $\times$ Water & 
+    4.11 & 
+    0.27 & 
+    15 & 
+    530 & 
+    0.78 & 
+    0.700 \\
 	\bottomrule
 	\end{tabular}
+	\label{table:TableS_fitLER}
 	\end{center}
@@
 
-% ANOVA table for SER (need to label tables)
+% ANOVA table for SER
+
 \begin{table}[htbp]
@@
 
-	\input{Tables/Table_heightGrowthAnova.txt}
-
+  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
+  Day to Germination & 3.6 & 
+    3.6 & 1 & 
+    662 & 21.1 &
+    5.1 $\times10^{-6}$ \\
+  Population & 12 & 0.8 & 
+    15 & 113 & 
+    4.7 & 5.8 $\times10^{-7}$ \\
+  Temperature & 12.4 & 
+    12.4 & 1 &
+    6 & 72.8 &
+    1.5 $\times10^{-4}$ \\
+  Water & 0.6 & 0.6 & 
+    1 & 5 & 
+    3.7 & 0.113 \\
+  Population $\times$ Temperature & 3.6 & 
+    0.2 & 
+    15 & 
+    549 & 
+    1.4 &
+    0.126 \\
+  Population $\times$ Water & 2.8 & 
+    0.2 & 
+    15 &
+    536 &
+    1.1 & 
+    0.330 \\
+  Population $\times$ Temperature $\times$ Water & 
+    1.5 & 
+    0.1 & 
+    15 & 
+    518 & 
+    0.6 & 
+    0.874 \\
 	\bottomrule
 	\end{tabular}
+	\label{table:TableS_fitLER}
+	\end{center}
+\end{table}
+
+% ANOVA table for photosynthesis
+
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, photosynthesis]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
+	\begin{center}
+	\begin{tabular}{lcccccc}
+	\toprule
+
+  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
+  Population & 347.7 & 
+    23.2 & 15 &
+    78 & 3.02 &
+    7.5 $\times10^{-4}$ \\
+  Temperature & 134.1 & 
+    134.1 & 1 &
+    6 & 17.46 &
+    6.4 $\times10^{-3}$ \\
+  Water & 51 & 
+    51 & 1 &
+    4 & 6.64 &
+    0.066 \\
+  Population $\times$ Temperature & 218.6 & 
+    14.6 & 
+    15 & 
+    263 & 
+    1.9 &
+    0.024 \\
+  Population $\times$ Water & 87.7 & 
+    5.8 & 
+    15 &
+    233 &
+    0.76 & 
+    0.724 \\
+  Population $\times$ Temperature $\times$ Water & 
+    91.4 & 
+    6.1 & 
+    15 & 
+    208 & 
+    0.79 & 
+    0.686 \\
+	\bottomrule
+	\end{tabular}
+	\label{table:TableS_fitPhoto}
+	\end{center}
+\end{table}
+
+% ANOVA table for intrinsic photosynthetic rate
+
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, photosynthesis]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
+	\begin{center}
+	\begin{tabular}{lcccccc}
+	\toprule
+
+  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
+  Population & 347.7 & 
+    0.1074 & 
+    15 &
+    305 & 
+    2.87 &
+    3.1 $\times10^{-4}$ \\
+  Temperature & 0.0105 & 
+    0.0105 & 
+    1 &
+    6 & 
+    0.28 &
+    0.616 \\
+  Water & 8e-04 & 
+    8e-04 & 
+    1 &
+    5 & 
+    0.02 &
+    0.888 \\
+  Population $\times$ Temperature & 
+    0.6463 & 
+    0.0431 & 
+    15 & 
+    289 & 
+    1.16 &
+    0.300 \\
+  Population $\times$ Water & 0.4592 & 
+    0.0306 & 
+    15 &
+    274 &
+    0.82 & 
+    0.656 \\
+  Population $\times$ Temperature $\times$ Water & 
+    0.4692 & 
+    0.0313 & 
+    15 & 
+    259 & 
+    0.83 & 
+    0.644 \\
+	\bottomrule
+	\end{tabular}
+	\label{table:TableS_fitIntrnPhoto}
+	\end{center}
+\end{table}
+
+% Analysis of deviance table for mortality
+
+\begin{table}[htbp]
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[ANOVA table, mortality]{CAPTION}
+	\begin{center}
+	\begin{tabular}{lccc}
+	\toprule
+
+  & $\chi ^2$ & df & $P$-value \\ \midrule
+  Population & 32 & 
+    31 & 0.419 & \\
+  Temperature & 31.8 & 
+    6 & 1.8 $\times10^{-5}$ & \\
+  Water & 69.2 & 
+    12 & 4.6 $\times10^{-10}$ & \\
+  Population $\times$ Temperature & 5.6 & 
+    15 & 
+    0.985 & \\
+  Population $\times$ Water & 8.6 & 
+    15 & 
+    0.897 & \\
+  Population $\times$ Temperature $\times$ Water & 
+    0.2 & 
+    15 & 
+    1.000 & \\	\bottomrule
+	\end{tabular}
+	\label{table:TableS_fitMort}
 	\end{center}

### 2016-10-13 ms/functions.R

@@ -27,8 +27,11 @@ sigDig <- function(x)
 # Text string for p-value in tables
 pval2latex <- function(x)
 {
+  stem <- x * 10 ^ -oom(x)
   ret <- ifelse(oom(x) > -3, 
                 sprintf("%.*f", 3, x), 
-                sprintf("%.*f $\times10^{%s}$", 1, stem, oom(x)))
+                sprintf("%.*f $\\\\times10^{%s}$", 1, stem, oom(x)))
   return(ret)
 }
+
+

### 2016-10-14 ms/ms.Rnw
Finished shitty version for bioarxiv

@@
 %lost on page 11 - extracted and used climate variables
-%Maybe reordering stuff on paragraph
 %cartoon of whole procedure for climate variables
@@
 %** correct # of Pop x Treatment
+% model of growth chamber
+% need to explain 'importance'
 
@@
 % 1. Is water regime on figure 1 correct? (data is in office)
-
-% 3. Do anything with biomass data?
-% 4. Citations for latitudinal clines
+% 2. Double check spatially averaged analysis
+% 3. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
+% 4. Do anything with biomass data? Currently, NO.
+
+% 5. Citations for latitudinal clines
 	% cases where clinal variation has 'obvious' selective source:
@@
 	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
-% 5. Citations for GxE as signature of local adaptation
-% 7. Supplemental Figure and Table captions
-% 8. Supplemental figure showing comparison between assigned to climatic variables from point estimated and spatially averaged analyses
-% 9. new spatially averaged analysis
-% 12. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
+% 6. Citations for GxE as signature of local adaptation
 
@@
 
-<<Front Matter, echo=FALSE, results=hide>>=
+<<Setup, echo=FALSE, results=hide>>=
 
@@
 
-\title{Grow with the flow: a latitudinal cline in growth rate is associated with more variable precipitation in \textit{Mimulus cardinals}}
+\title{Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Mimulus cardinalis}}
 \author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
@@
 
-\listoffigures
-\listoftables
+% \listoffigures
+% \listoftables
+
+\setstretch{\stretchy}
 
@@
 We hypothesize that selection may favor a more annualized life history -- grow now rather than save for next year -- in environments where severe droughts occur more often.
-Thus our study provides insight into how species may adapt if Meditteranean climates become more variable due to climate change.
-
-\setstretch{\stretchy}
+Thus our study provides insight into how species may adapt if Mediterranean climates become more variable due to climate change.
 
@@
 A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures.
-Here, we examine physiological trait variation and possible selective agents in the perennial herb \textit{Mimulus cardinalis} Douglas ex Benth.
+% Here, we examine physiological trait variation and possible selective agents in the perennial herb \textit{Mimulus cardinalis} Douglas ex Benth.
 (Phrymaceae), a model system for local adaptation studies.
@@
 % Intrinsic versus plastic variation 
-When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004} \ala{added citation - might give error}.
+When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}.
 The traits that underlie local adaptation, however, need not mirror this pattern.
@@
 There are classic cases of adaptation involving both intrinsic and plastic trait variation.
-For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
+For example, intrinsic differences in photoperiod \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
 Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
@@
 % Selective agents (trait-enviro assoc)
-A basic approach one can use to identify candidate traits underlying local adaptation is to find associations between traits and environments.
+A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments.
 Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
@@
 Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
-Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely relate to climatic variation.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008}, and drought response \citep{Kooyers_etal_2015} that likely relate to climatic variation.
 
 Despite the fact that latitudinal clines in particular have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
-Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days, such that latitude is an effective proxy for the underlying climatic driver but nonetheless  key climatic variable(s) with stronger relationships to traits exist.
-Alternatively, latitude may be more strongly related to traits than any single climatic variable for two possible reasons.
+Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days.
+Hence, latitude is an effective proxy for the underlying climatic driver, but we would expect a yet stronger relationship between traits and the key climatic variable(s) driving selection.
+Alternatively, latitude may be more strongly related to traits than any single climatic variable for at least two reasons.
 First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline.
 Alternatively, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
-For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a Southern set of populations will experience warmer climate than a Northern one.
-Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g.
+We refer to this as the `climatic neighborhood'.
+For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a lower latitude set of populations will experience warmer climate than a higher latitude one.
+Thus, any particular low latitude population would be warm-adapted, even if it was located in a cooler (e.g.
 high elevation) site.
@@
 coastal vs.
-continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local or regionally-averaged variables are more predictive of trait change.
-\ala{added this last clause to try to encompass gene flow idea, but now maybe the sentence is too clunky}
+continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local versus regional climate shape selective pressures.
 
@@
 % might add that Anacker and Strauss 2014 showed ecological shifts important in Mim speciation and we would like to know driving forces behind this???
-In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}.
-We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
+In this study, we intestigated two major questions: 1) whether intrinsic or plastic physiological trait variation corresponds with latitude; and 2) what climatic factor(s) could plausibly be repsonsible for latitudinal clines.
+Within question 2, we tested three hypotheses outlined in the previous paragraph: latitudinal clines are explained by a single dominant climatic factor, multiple climatic factors, or the climatic neighborhood experienced by nearby population connected through gene flow.
+These hypotheses are not mutually exclusive since, for example, single or multiple factors in a climatic neighborhood may lead to latitudinal clines.
+
+We examined these questions in \textit{Mimulus cardinalis} because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
 Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
@@
 We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
-We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity.
-Interannual variation in precipitation and/or precipitation seasonality is associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+We found a latitudinal cline of intrinsic variation in photosynthesis and growth, but little evidence for variation in plasticity.
+Interannual variation in precipitation is associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
 cardinalis}.
+However, no single climatic factor explained trait variation better than latitude alone, indicating that additional climatic factors are likely important, albeit less so.
+The climatic neighborhoods around populations likewise did not explain trait variation better than latitude alone.
 We place these findings in the context of life history theory and consider future directions in the Discussion.
-\ala{wrap spatial buffer results into summary?}
-
-\section*{Methods}
+
+\section*{Material and Methods}
 
@@
 Two jumbo-sized cotton balls at the bottom of cone-tainers prevented sand from washing out.
-Cone-tainers were continuously bottom-watered during germination by placing them in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) filled part way with water, placed on benches in greenhouses at the University British Columbia campus in Vancouver, Canada (49\degree 15' N, 123\degree 15' W).
+Cone-tainers sat in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) to continuously bottom-water plants during germination in greenhouses at the University British Columbia campus in Vancouver, Canada (49\degree 15' N, 123\degree 15' W).
 Misters thoroughly wetted the top of the sand every two hours during the day.
 Most seeds germinated between 1 and 2 weeks, but we allowed 3 weeks before transferring seedlings to growth chambers.
-Germination was recorded daily from one to two weeks after sowing, and every few days thereafter.
-On 5 May (21 days after sowing), seedlings were transferred to one of two MODEL Growth Chambers (Conviron, Manitoba, Canada).
+We recorded germination daily between one to two weeks after sowing, and every 2-3 days thereafter.
+On 5 May (21 days after sowing), we transferred seedlings to one of two growth chambers (Conviron, Manitoba, Canada).
 We thinned seedlings to one plant per cone-tainer, leaving the center-most plant.
@@
 We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments.
-The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments Table~\ref{table:TableS_InitialSize}.
+The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments (Table~\ref{table:TableS_InitialSize}).
 
@@
 We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels.
-The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which designate as Hot and Cool treatments.
+The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which we designate as Hot and Cool treatments.
 Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments.
-A detailed description of treatments is described in the Supplemental Information and summarized in Fig~\ref{fig:Fig_ExptlDes}.
+A detailed description of treatments is provided in the Supplemental Materials and Methods and summarized in Fig~\ref{fig:Fig_ExptlDes}.
 Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level.
 Within each chamber, there were two Well-watered blocks and two Drought blocks.
-The irradiance in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$.
+The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$.
 The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
@@
 All plants germinated within 21 days of sowing.
-At that time, we began temperature treatments (left axis), simulating a typical June-August weather pattern at Hot (red) and Cold (blue) sites.
+At that time, we began temperature treatments (left axis), simulating a typical June-August weather pattern at Hot (red) and Cool (blue) sites.
 The bold lines track the average daily temperatures.
@@
 The drought treatment commenced later by ramping down the frequency of bottom-watering episodes (black line; right axis).
-Grey boxes on the bottom of the plot outline the period of key measurements described in the Methods.}
+Grey boxes on the bottom of the plot outline the period of key measurements described in the Material and Methods.}
 	\label{fig:Fig_ExptlDes}
@@
       \midrule
-  Day of germination  & day \\
+  Days to germination  & day \\
   Leaf expansion rate  &  mm day$^{-1}$  \\
-  Shoot elongation rate  &  mm or cm day$^{-1}$  \\
-  Harvest dry mass &  g  \\
+  Stem elongation rate  &  mm or cm day$^{-1}$  \\
   Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
@@
 
-\paragraph{Day of germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}.
-The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
-The signifcance of the Population effect was determined using analysis of deviance.
-Note that, unlike other traits discussed below, we did not include Block or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not been applied.
+\paragraph{Days to germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}.
+We treated Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
+Statistical signifcance of the Population effect was determined using analysis of deviance.
+Note that, unlike other traits discussed below, we did not include Block, Treatment, or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not yet been applied.
 
@@
 
-\paragraph{Growth rate: leaf expansion and shoot elongation}
-
-We measured growth rate in two phases: leaf expansion as a rosette and shoot elongation after bolting.
+\paragraph{Growth rate: leaf expansion and stem elongation}
+
+We measured growth rate during two phases: leaf expansion as a rosette and stem elongation after bolting.
 We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
-We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.
-We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing).
-Both leaf expansion and shoot elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.
+We ceased measuring leaf length once it appeared to asymptote and growth shifted to stem elongation.
+We also censused plant height on 7 occasions (twice per week) between 29 May and 20 June (45 to 67 days after sowing).
+Both leaf expansion and stem elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and stem growth) using empirical Bayes' estimates from linear mixed-effects models fit with the R package \pkg{lme4} version 1.1-12 \citep{Bates_etal_2015}.
 
@@
 
-<<Shoot elongation rate, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+<<Stem elongation rate, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
 \paragraph{Photosynthesis}
-During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations.
-The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
-All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
-During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.
-As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out.
+During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations.
+The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6-cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
+We made all measurements at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$ of photosynthetically active radiation), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
+During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level could be measured under the same conditions.
+% As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out.
 Therefore, we calculated the `intrinsic' photosynthetic rate for each individual as its residual deviance from the regression between stomoatal conductance and photosynthetic rate across all plants.
@@
 
-  # Intrinsic photosynthetic rate
+  # Intrinsic photosynthetic rate (not using at the moment)
   # = residuals of Conductance - Photosynthesis regression
   # Removed one outlier (may change if data are rearranged!)
-	mm <- lmer(resPhoto ~ PopID * TempTrt * WaterTrt + (1|Family) + (1|Block),
-    data = subset(master, !is.na(master$Photo))[-52, ])
-	fitIntrnPhoto <- step(mm, reduce.random = F) # only Pop
+	# mm <- lmer(resPhoto ~ PopID * TempTrt * WaterTrt + (1|Family) + (1|Block),
+  #  data = subset(master, !is.na(master$Photo))[-52, ])
+	# fitIntrnPhoto <- step(mm, reduce.random = F) # only Pop
 
@@
 We included Family and Block as random effects.
-We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
+We assessed significance of fixed effects using Type-II Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
 
@@
 
-\paragraph{Biomass at harvest}
-
-We harvested plants that survived until the end of the experiment from 3 to 16 July (80 to 93 days after sowing).
-We separated individual plants into three biomass fractions: leaves, shoots, and belowground (roots plus rhizomes).
-Belowground fractions were rinsed thoroughly in water to remove sand.
-After completely drying in a forced air oven (Shel Lab, Sheldon Manufacturing Inc., Oregon, USA), we weighed the dry mass of each fraction using an analytical balance.
-
 \subsection*{Intrinsic variation and plasticity}
@@
 For all other traits, we used mixed model ANOVAs with Family and Block included as random factors.
-Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}.
-Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-11 \citep{Kuznetsova_etal_2014}.
-Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation.
-To ensure that Population and Treatment effects were specific to a particular growth phase, we included earlier growth phases as a covariate.
-Specifically, we included germination day as a covariate in leaf expansion and shoot elongation analyses; we included germination day and leaf expansion rate as a covariate in shoot elongation analyses.
+We fit models using restricted maximum likelihood in lmer, a function in the R package \pkg{lme4} \citep{Bates_etal_2015}.
+We determined significant fixed effect terms using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-32 \citep{Kuznetsova_etal_2016}.
+This package uses Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests.
+We also included days to germination as a covariate in growth analyses.
+% To ensure that Population and Treatment effects were specific to a particular growth phase, we included earlier growth phases as a covariate.
+Specifically, we included germination day as a covariate in leaf expansion and stem elongation analyses; we included germination day and leaf expansion rate as a covariate in stem elongation analyses.
 
@@
 The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA).
-The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.
+The first principal component of these traits (TraitPC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.
 
@@
     SER = fitSER$lsmeans.table[1:16, "Estimate"],
-    photo = fitIntrnPhoto$lsmeans.table[1:16, "Estimate"])
+    photo = fitPhoto$lsmeans.table[1:16, "Estimate"])
+  #  photo = fitIntrnPhoto$lsmeans.table[1:16, "Estimate"])
 
@@
   			 cex = 1.5, lwd = 2)
+
+  # P-value
+  text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1),
+	     labels = bquote(italic(P) == .(round(summary(fitLat)[[4]][2, 4], 3))))
+
   dev.off()
@@
 
-\subsection*{Selective agents and environmental correlates}
-
-We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+\subsection*{Identifying putative selective agents}
+
+We found that a population's position along TraitPC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
 cardinalis}.
-We looked for overlap between climatic variables that best predict latitude of \textit{M.
-cardinalis} occurrence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
-For brevity, we refer to these as `Climate-Latitude' and `Climate-TraitPC1' variables.
-We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
-From VSURF models, we kept only variables selected for prediction, the most stringent criterion.
-
-If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high?
+Hypothesis 1: if a single climatic factor explained more trait variation than latitude alone, this would suggest that that factor is a selective agent underlying the latitudinal cline in our 16 focal populations.
+Hypothesis 2: if multiple climatic factors together explained trait variation,  we interpreted this as evidence that multiple climatic factors together have generated the latitudinal cline.
+We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables.
+In addition, to help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{M.
+cardinalis} occurrence records, not just the 16 focial populations.
+We refer to these climatic factors as `Climate-Latitude' variables.
+The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{M.
+cardinalis} range.
+We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}.
+VSURF ranks variables by their importance over regression trees in the forest.
+We kept only variables selected for prediction, the most stringent criterion.
+
+% This is now stated above.
+%If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high?
 maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}.
@@
 
-To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-TraitPC1 variables as above, except that we used spatially averaged climate variables.
-\ala{what's the actual prediction?
-that spatially averaged Climate-TraitPC1 will have higher correlation than point estimated Climate-Trait?} We sampled climate at 1000 random points (at 90-m resolition) within a 62-km buffer around focal populations.
+To test the third hypothesis about climatic neighborhoods driving selection, we repeated the same procedure to identify Climate-TraitPC1 variables as above, except that we used spatially averaged climate variables.
+We sampled climate at 1000 random points (at 90-m resolution) within a 62-km buffer around the 16 focal populations.
 We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
@@
 Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
-We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km buffer (`spatially averaged Climate-TraitPC1').
+We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km climatic neighborhood (`spatially averaged Climate-TraitPC1').
 
@@
     cex = 1.5)
+  text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
+       labels = "A", font = 2, cex = 2)
 
@@
 		pch = 21, col = "black", bg = "white", cex = 1.5)
+  text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
+       labels = "B", font = 2, cex = 2)
 
@@
 
+  # *** SHOULD BE MOVED TO MORE LOGICAL POSITION *** #
+  fitLat <- lm(PC1 ~ Latitude, data = focClim)
+  fitBio16_cva <- lm(PC1 ~ bio16_cva, data = focClim)
+  fitBio12_cva <- lm(PC1 ~ bio12_cva, data = focClim)
+  fitDD_0_cva <- lm(PC1 ~ DD_0_cva, data = focClim)
+  fit2var <- lm(PC1 ~ bio16_cva + DD_0_cva, data = focClim)
+
   # Overlap between latitude predictors and phenotype predictors (spatial aveage)
@@
 
-For these analyses, we compiled a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
+For Climate-Latitude analyses, we compiled a representative set of \Sexpr{2 * nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
 cardinalis} occurences.
 These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
+For both Climate-TraitPC1 analyses (16 focal populations) and Climate-Latitude (many populations), we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
 Monthly interpolated climate layers were calculated using ClimateWNA version 5.30 \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
+For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} version 1.1-1 \citep{Hijmans_etal_2016}.
 In total, we included \Sexpr{length(climVars) / 2} climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
@@
 
-Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis}, as evidenced by large and significant population effects (Table~\ref{table:Table_AnovaSummary}).
+There are strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
+cardinalis}, as evidenced by large and significant population effects for these traits (Table~\ref{table:Table_AnovaSummary}).
 A single principal component captured \Sexpr{gsub("%", "", pve)}\% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
 As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.
+There were similar latitudinal clines for individual traits underlying PC1 (Fig S1-S4).
 
@@
 
+% *** Commented out results from Intrinsic Photosynthetic rate *** %
+
 \begin{table}[ht]
@@
    \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects.
-We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
+We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2015}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2015}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
 Note that temperature and water treatments were imposed after germination, hence are not applicable to this trait.
@@
    \resizebox{6.5in}{!}{
-   \begin{tabular}{@{} llllllll @{}}
+   \begin{tabular}{@{} lllllll @{}}
       \toprule
-    {\raggedleft Trait}                          & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
-    Statistical model              & survreg     & lmer           & lmer            & lmer           & lmer          & glmer \\
+    {\raggedleft Trait}                          & Germination & Leaf expansion & Stem elongation & Photosynthesis & Mortality \\
+    Statistical model              & survreg     & lmer           & lmer            & lmer           & glmer \\
       \midrule
@@
     \Sexpr{sigStar(fitPhoto$anova.table["PopID", "Pr(>F)"])} & 
-    \Sexpr{sigStar(fitIntrnPhoto$anova.table["PopID", "Pr(>F)"])} & 
     \Sexpr{sigStar(Anova(fitMort, type = 2)["PopID", "Pr(>Chisq)"])} & \\
@@
     \Sexpr{sigStar(fitPhoto$anova.table["TempTrt", "Pr(>F)"])} & 
-    \Sexpr{sigStar(fitIntrnPhoto$anova.table["TempTrt", "Pr(>F)"])} &
     \Sexpr{sigStar(Anova(fitMort, type = 2)["TempTrt", "Pr(>Chisq)"])} & \\
@@
     \Sexpr{sigStar(fitPhoto$anova.table["WaterTrt", "Pr(>F)"])} & 
-    \Sexpr{sigStar(fitIntrnPhoto$anova.table["WaterTrt", "Pr(>F)"])} &     
     \Sexpr{sigStar(Anova(fitMort, type = 2)["WaterTrt", "Pr(>Chisq)"])} & \\
@@
     \Sexpr{sigStar(fitPhoto$anova.table["PopID:TempTrt", "Pr(>F)"])} & 
-    \Sexpr{sigStar(fitIntrnPhoto$anova.table["PopID:TempTrt", "Pr(>F)"])} &     
     \Sexpr{sigStar(Anova(fitMort, type = 2)["PopID:TempTrt", "Pr(>Chisq)"])} & \\
@@
     \Sexpr{sigStar(fitPhoto$anova.table["PopID:WaterTrt", "Pr(>F)"])} & 
-    \Sexpr{sigStar(fitIntrnPhoto$anova.table["PopID:WaterTrt", "Pr(>F)"])} &     
     \Sexpr{sigStar(Anova(fitMort, type = 2)["PopID:WaterTrt", "Pr(>Chisq)"])} & \\
@@
     \Sexpr{sigStar(fitPhoto$anova.table["TempTrt:WaterTrt", "Pr(>F)"])} & 
-    \Sexpr{sigStar(fitIntrnPhoto$anova.table["TempTrt:WaterTrt", "Pr(>F)"])} &     
     \Sexpr{sigStar(Anova(fitMort, type = 2)["TempTrt:WaterTrt", "Pr(>Chisq)"])} & \\
@@
     \Sexpr{sigStar(fitPhoto$anova.table["PopID:TempTrt:WaterTrt", "Pr(>F)"])} & 
-    \Sexpr{sigStar(fitIntrnPhoto$anova.table["PopID:TempTrt:WaterTrt", "Pr(>F)"])} &     
     \Sexpr{sigStar(Anova(fitMort, type = 2)["PopID:TempTrt:WaterTrt", "Pr(>Chisq)"])} & \\
@@
 	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude.
-Each point is a population's latitude of origin (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of five traits (see Methods).
+Each point is a population's latitude of origin (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of four traits (see Material and Methods).
 The line and 95\% confidence intervals were estimated using linear regression.}
@@
 Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation.
-For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions.
 We found little evidence G$\times$E in \textit{M.
 cardinalis}.
-\cdm{Is this result still correct after revising analysis?} There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing.
+There were only two statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but these were not strong compared to Population and Temperature effects.
 Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
 Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
-
-\subsection*{Climatic variability best explains phenotypic divergence}
+However, the fact that the Population and Treatment effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
+Complete ANOVA tables are available in the Supporting Information (Table S3-S6).
+
+\subsection*{Climatic variability best explains latitudinal cline}
 
@@
 For \textit{M.
-cardinalis}, interannual variation in precipitation over the past 30 years is very closely related to the latitude of recently recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
-Between year variation in precipitation was also strongly correlated with position along the fast-slow growth axis, but climate based on point estimates gave somewhat different results than climate based on spatially averaged values.
-The two most important point estimated Climate-Latitude variables were also strong predictors of position along the fast-slow growth axis (Fig.~\ref{fig:Fig_ClimVarImp}B).
-For spatially averaged climate, interannual variation in precipitation seasonality was the only climatic variable selected, (Fig.~\ref{table:TableS_ClimVarImp}) \cdm{Do we need a figure for this?}.
-Overall, there was good but imperfect agreement between the weight assigned to climatic variables using point estimated and spatially averaged climate (Fig.~\ref{fig:Fig_TBD}) \cdm{This figure will show strong, positive correlation between weights given to climatic variables from point estimated and spatially averaged analyses}.
-
-Overlap between Climate-Latitude and point estimated Climate-Trait variables suggests that interannual variation in precipitation is an important selective agent in \textit{M.
-cardinalis}.
-Specifically, we hypothesize that more frequent droughts (greater precipitation cv) in Southern populations selects for an `annual-ized' life history, as we detail in the Discussion.
-Even though the spatially averaged Climate-Trait variable was not among the Climate-Latitude variables, it points to a similar conclusion.
-Specifically, this analysis showed that greater interannual variation in precipitation seasonality, which is also driven by precipitation variability (i.e.
-more frequent drought years) in Southern California, is associated with faster growth in southern populations.
-We must qualify these results because our analysis obviously cannot rule out that alternative variables not included in the analysis may be more important.
+cardinalis}, interannual variation in precipitation over 30 years (1981--2010) is very closely related to the latitude of recently recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
+Specifically, precipitation is more variable at lower latitude.
+The two most important Climate-Latitude variables were the interannual variation in total precipitation (bio12$_\sigma$; see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations) and precipitation in the wettest quarter (bio16$_\sigma$).
+Note that the coefficient of variation of a climatic factor is subscripted with $\sigma$ whereas the mean is subscripted with $\mu$.
+Bio12$_\sigma$ and bio16$_\sigma$ are very similar because in Mediterranean climates of California, most precipitation occurs in the winter quarter.
+These Climate-Latitude variables were also the two most important point estimated Climate-TraitPC1 variables (Fig.~\ref{fig:Fig_ClimVarImp}B).
+Populations from Southern, more variable environments grew faster.
+However, neither climatic variable alone explained more variation in TraitPC1 than latitude (latitude $r^2=\Sexpr{round(summary(fitLat)["r.squared"]$r.squared, 2)}$; bio16$_\sigma$ $r^2=\Sexpr{round(summary(fitBio16_cva)["r.squared"]$r.squared, 2)}$; bio12$_\sigma$ $r^2=\Sexpr{round(summary(fitBio12_cva)["r.squared"]$r.squared, 2)}$).
+Interannual variation in chilling degree-days (DD\_0$_\sigma$) was also common to both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
+Southern, faster growing populations are from climates with less variation in chilling degree-days.
+However, DD\_0$_\sigma$ and bio16$_\sigma$ together still only explained slightly more variation in TraitPC1 than latitude alone (latitude $r^2=\Sexpr{round(summary(fitLat)["r.squared"]$r.squared, 2)}$ versus DD\_0$_\sigma$+bio16$_\sigma$ $r^2=\Sexpr{round(summary(fit2var)["r.squared"]$r.squared, 2)}$).
+This suggests that interannual variation in precipitation explains most of the latitudinal cline, but other variables such as variation in chilling degree-days may contribute.
+
+There was no overlap between Climate-Latitude and spatially averaged Climate-TraitPC1 (Fig.~\ref{table:TableS_ClimVarImp}).
+This is not because climatic neighboorhood did not predict TraitPC1; there were significant correlations between important spatially averaged climatic factors and TraitPC1 that explained as much variation as latitude (data not shown).
+If these correlations are spurious, this indicates that climatic neighborhood is not as imporant as climate in the immediate vicinity of a population.
+Conversely, the climatic neighborhood may be important among our focal populations and depart from rangewide correlations between latitude and climate.
+Since these analyses are exploratory and prone to overinterpretation, we believe the closer overlap between Climate-Latitude and point estimated Climate-TraitPC1 variables provides more compelling evidence for putative selective agents.
 
@@
 A.
-Using Random Forest regression, we identified 10 climatic variables significantly (high importance) associated with latitude of \textit{M.
+Using Random Forest regression, we identified \Sexpr{nImpLat} climatic variables significantly (high importance) associated with latitude of \textit{M.
 cardinalis} occurences.
 B.
-The two most important Climate-Latitude variables were also the two most important Climate-Trait variables.
+The two most important Climate-Latitude variables were also the two most important Climate-TraitPC1 variables.
 Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
-Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp) or Precipitation (Prec); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
+Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp), Precipitation (Prec), Chilling (chilling degree-days), Snow (precipitation as snow); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
 	\label{fig:Fig_ClimVarImp}
@@
 
-% The observation that populations are locally adapted to their environment is old and well-documented in many species.
-New frontiers in local adaptation research seek a general explanation for why local adaptation often suceeds but sometimes fails, why certain traits respond to selection whereas others are constrained, and what environmental drivers are most important [CITES such as Kawecki].
-Understanding why and how populations have adapted to different environments in the past is especially significant now as a scientific basis for predicting how species' ranges will shift under climate change [CITES, \citep{Catullo_etal_2015}].
-In this study, we found evidence for one of two common signatures of local adaptation.
-A latitudinal cline in traits such as germination rate, photosynthesis, and growth, suggests adaptive differentiation in fundamental physiological traits of the species.
+% From Amy: \ala{the results that underlie this statement [niche conservation] need more emphasis in the methods}
+We found evidence for one of two common signatures of local adaptation in the perennial herb \textit{Mimulus cardinalis}.
+Latitudinal clines in germination rate, photosynthesis, and growth, suggests adaptive differentiation in fundamental physiological traits of the species.
 However, we found little evidence that populations respond differently to temperature or drought.
-As we discuss below, this may indicate that the fundamental abiotic niche is relatively conserved \ala{the results that underlie this statement need more emphasis in the methods}.
+As we discuss below, this may indicate that the fundamental abiotic niche is relatively conserved.
 Finally, we found that climatic variation between years may be a more important selective agent than the average climate.
@@
 In \textit{M.
-cardinalis} we found substantial genetic variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
+cardinalis} we found substantial genetically based variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
 If this variation is adaptive, it suggests that the fitness tradeoff between doing well in low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
-That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatial heterogenous selection.
+That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatially heterogenous selection.
 We also know from previous work that population size does not vary strongly with latitude.
@@
 We found no evidence for this; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
+Plants grew faster in the Hot treatment, but there was little effect of drought on growth.
+Rather, the effects of drought took longer to materialize but resulted in high mortality, especially in the Hot chamber.
+However, there was no differential mortality among populations in this treatment.
 We hypothesize that evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
@@
 cardinalis} live are highly heterogeneous at small spatial scales.
-Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Meditteranean climates of western North America.
+Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Mediterranean climates of western North America.
 But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and vice versa, leading to strong fitness tradeoffs.
 Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
-Demography and gene flow may reinforce niche conservatism \ala{also bring in idea of temporal source sink: most individuals made in wet years, even if infrequent, so selection weighted towards wet environment even if dry states are frequent}.
+Demography and gene flow may reinforce niche conservatism.
 A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher resource environments.
+Infrequent wet years may also produce most seeds, so selection weighted towards wet environment even if dry years are more frequent.
 Finally, gene flow, which is generally high among \textit{M.
@@
 
-Based on the available data, interannual variation in annual or winter precipitation or precipitaiton seasonality (these are closely correlated in Mediterranean climates) may be the selective agent driving variation in growth and photosynthesis.
+Based on the available data, interannual variation in annual or winter precipitation (these are closely correlated in Mediterranean climates) may be the selective agent driving variation in physiological traits.
 Variation in precipitation was best predicted latitude of recent \textit{M.
@@
 Conversely, in more predictable northern environments, lifetime fitness may be optimized when a significant fraction of assimilate is allocated below ground for future years.
-Although this hypothesis remains to be directly tested, a few indepdent lines of evidence are consistent with it.
+Although this hypothesis remains to be directly tested, a few independent lines of evidence are consistent with it.
 Preliminary surveys suggest that northern populations not only grow slower, but also produce greater numbers of rhizomes (C.D.
@@
 data).
-Finally, demographic surveys of natural populations show greater variation in the size recruits in southern populations, suggesting higher maximum growth rates under natural conditions (M.
+Finally, demographic surveys of natural populations show greater variation in the size of recruits in southern populations, suggesting higher maximum growth rates under natural conditions (M.
 Bayly \& A.
@@
 
-\cdm{I am still working a paragraph or two linking these results to empirical results in other systems and climatic adaptaiton in Mimulus sect Simiolus.}
+In summary, we found evidence for a coordinated latitudinal cline in germination rate, photosynthesis, and growth, suggesting local adaptation.
+These fixed differences between populations suggest different trait optima in different climates.
+We did not find evidence that the relative performance of populations shifts with temperature or watering regime, suggesting relatively little variation in plasticity.
+Exploratory analysis implicate that more variable precipitation regimes at lower latitude drive much of the latitudinal cline, though other climatic factors could also contribute.
+There was little compelling evidence that selection over a broad climatic neighborhood surrounding populations strongly explained the latitudinal cline.
+In the future, we will use field experiments to test whether greater variation in precipitation selects for faster growth and that selection on temperature/drought responses does not vary among populations.
+By doing so, we aim to understand why certain physiological and developmental mechanisms, but not others, contribute to local adaptation.
 
@@
 
+% *** Need to fill in exact grant numbers *** %
 \section*{Acknowledgements}
-Erin Warkman and Lisa Lin helped collect data. CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program. \cdm{Please fill in other relevant funding info}
+Erin Warkman and Lisa Lin helped collect data.
+CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program.
+ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation.
 
@@
 
+%--------------------------------------------------
+% Supporting Tables
+%--------------------------------------------------
+
+\subsection*{Supporting Tables}
+
 % DIC table for initial size (LLL on 5/12)
@@
 	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{Initial size of seedlings did not vary among Populations, Families, or Treatments.
+	\caption[DIC table for initial size of seedlings]{Initial size of seedlings did not vary among Populations, Families, or Treatments.
 We used a censored Gaussian model of initial size at the outset of the experiment (longest leaf length of the first true leaves).
@@
 We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in.
-We step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
+We used step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
 	\begin{center}
@@
 \begin{table}[ht]
+	\caption[Climate variables]{Climatic variables used }
    \centering
-   \topcaption{Climatic variables}
    \begin{tabular}{@{} ll @{}}
@@
 
+% Analysis of deviance table for germination
+
+<<TableS_fitGerm, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  # Convert numbers from R to LaTeX characters
+  tab <- suppressWarnings(anova(fitGerm))
+  germTab <- data.frame(pval = pval2latex(tab$`Pr(>Chi)`),
+                         stringsAsFactors = F)
+  rownames(germTab) <- rownames(tab)
+
+@
+
+% *** NEED TO DO TABLE IF I WANT IT ***
+
 % ANOVA table for LER
@@
 	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2); }
+	\caption[ANOVA table, leaf expansion rate]{Analysis of varianace (ANOVA) table on leaf expansion rate (LER) using \pkg{lmerTest} \citep{Kuznetsova_etal_2016}.
+Family and Block were included as random effects.
+Abbreviations: SS = sum of squares; MS = mean sum of squares (SS / df1); df1 = numerator degrees of freedom; df2 = denominator degrees of freedom.}
 	\begin{center}
@@
     \Sexpr{lerTab["WaterTrt", "fval"]} & \Sexpr{lerTab["WaterTrt", "pval"]} \\
+  Temperature $\times$ Water & \Sexpr{lerTab["TempTrt:WaterTrt", "ss"]} & 
+    \Sexpr{lerTab["TempTrt:WaterTrt", "ms"]} & 
+    \Sexpr{lerTab["TempTrt:WaterTrt", "df1"]} & 
+    \Sexpr{lerTab["TempTrt:WaterTrt", "df2"]} & 
+    \Sexpr{lerTab["TempTrt:WaterTrt", "fval"]} &
+    \Sexpr{lerTab["TempTrt:WaterTrt", "pval"]} \\
   Population $\times$ Temperature & \Sexpr{lerTab["PopID:TempTrt", "ss"]} & 
@@
 	\selectfont
-	\caption[ANOVA table, shoot elongation rate]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
+	\caption[ANOVA table, stem elongation rate]{Analysis of varianace (ANOVA) table on stem elongation rate (SER) using \pkg{lmerTest} \citep{Kuznetsova_etal_2016}.
+Family and Block were included as random effects.
+Abbreviations: SS = sum of squares; MS = mean sum of squares (SS / df1); df1 = numerator degrees of freedom; df2 = denominator degrees of freedom.}
 	\begin{center}
@@
     \Sexpr{serTab["WaterTrt", "fval"]} & \Sexpr{serTab["WaterTrt", "pval"]} \\
+  Temperature $\times$ Water & \Sexpr{serTab["TempTrt:WaterTrt", "ss"]} & 
+    \Sexpr{serTab["TempTrt:WaterTrt", "ms"]} & 
+    \Sexpr{serTab["TempTrt:WaterTrt", "df1"]} & 
+    \Sexpr{serTab["TempTrt:WaterTrt", "df2"]} & 
+    \Sexpr{serTab["TempTrt:WaterTrt", "fval"]} &
+    \Sexpr{serTab["TempTrt:WaterTrt", "pval"]} \\
   Population $\times$ Temperature & \Sexpr{serTab["PopID:TempTrt", "ss"]} & 
@@
 	\selectfont
-	\caption[ANOVA table, photosynthesis]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
+	\caption[ANOVA table, photosynthesis]{Analysis of varianace (ANOVA) table on photosynthetic rate using \pkg{lmerTest} \citep{Kuznetsova_etal_2016}.
+Family and Block were included as random effects.
+Abbreviations: SS = sum of squares; MS = mean sum of squares (SS / df1); df1 = numerator degrees of freedom; df2 = denominator degrees of freedom.}
 	\begin{center}
@@
     \Sexpr{photoTab["WaterTrt", "pval"]} \\
+  Temperature $\times$ Water & \Sexpr{photoTab["TempTrt:WaterTrt", "ss"]} & 
+    \Sexpr{photoTab["TempTrt:WaterTrt", "ms"]} & 
+    \Sexpr{photoTab["TempTrt:WaterTrt", "df1"]} & 
+    \Sexpr{photoTab["TempTrt:WaterTrt", "df2"]} & 
+    \Sexpr{photoTab["TempTrt:WaterTrt", "fval"]} &
+    \Sexpr{photoTab["TempTrt:WaterTrt", "pval"]} \\
   Population $\times$ Temperature & \Sexpr{photoTab["PopID:TempTrt", "ss"]} & 
@@
 
-% ANOVA table for intrinsic photosynthetic rate
-
-<<TableS_fitIntrnPhoto, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
-
-  # Convert numbers from R to LaTeX characters
-  intrnPhotoTab <- data.frame(ss = sigDig(fitIntrnPhoto$anova.table$`Sum Sq`),
-                              ms = sigDig(fitIntrnPhoto$anova.table$`Mean Sq`),
-                              df1 = sigDig(fitIntrnPhoto$anova.table$NumDF),
-                              df2 = sigDig(fitIntrnPhoto$anova.table$DenDF),
-                              fval = sigDig(fitIntrnPhoto$anova.table$F.value),
-                              pval = 
-                                pval2latex(fitIntrnPhoto$anova.table$`Pr(>F)`),
-                              stringsAsFactors = F)
-  rownames(intrnPhotoTab) <- rownames(fitIntrnPhoto$anova.table)
-
-@
-
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, photosynthesis]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
-  Population & \Sexpr{photoTab["PopID", "ss"]} & 
-    \Sexpr{intrnPhotoTab["PopID", "ms"]} & 
-    \Sexpr{intrnPhotoTab["PopID", "df1"]} &
-    \Sexpr{intrnPhotoTab["PopID", "df2"]} & 
-    \Sexpr{intrnPhotoTab["PopID", "fval"]} &
-    \Sexpr{intrnPhotoTab["PopID", "pval"]} \\
-  Temperature & \Sexpr{intrnPhotoTab["TempTrt", "ss"]} & 
-    \Sexpr{intrnPhotoTab["TempTrt", "ms"]} & 
-    \Sexpr{intrnPhotoTab["TempTrt", "df1"]} &
-    \Sexpr{intrnPhotoTab["TempTrt", "df2"]} & 
-    \Sexpr{intrnPhotoTab["TempTrt", "fval"]} &
-    \Sexpr{intrnPhotoTab["TempTrt", "pval"]} \\
-  Water & \Sexpr{intrnPhotoTab["WaterTrt", "ss"]} & 
-    \Sexpr{intrnPhotoTab["WaterTrt", "ms"]} & 
-    \Sexpr{intrnPhotoTab["WaterTrt", "df1"]} &
-    \Sexpr{intrnPhotoTab["WaterTrt", "df2"]} & 
-    \Sexpr{intrnPhotoTab["WaterTrt", "fval"]} &
-    \Sexpr{intrnPhotoTab["WaterTrt", "pval"]} \\
-  Population $\times$ Temperature & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt", "ss"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt", "ms"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt", "df1"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt", "df2"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt", "fval"]} &
-    \Sexpr{intrnPhotoTab["PopID:TempTrt", "pval"]} \\
-  Population $\times$ Water & \Sexpr{intrnPhotoTab["PopID:WaterTrt", "ss"]} & 
-    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "ms"]} & 
-    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "df1"]} &
-    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "df2"]} &
-    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "fval"]} & 
-    \Sexpr{intrnPhotoTab["PopID:WaterTrt", "pval"]} \\
-  Population $\times$ Temperature $\times$ Water & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "ss"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "ms"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "df1"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "df2"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "fval"]} & 
-    \Sexpr{intrnPhotoTab["PopID:TempTrt:WaterTrt", "pval"]} \\
-	\bottomrule
-	\end{tabular}
-	\label{table:TableS_fitIntrnPhoto}
-	\end{center}
-\end{table}
-
 % Analysis of deviance table for mortality
@@
 	\selectfont
-	\caption[ANOVA table, mortality]{CAPTION}
+	\caption[Analysis of deviance table, mortality]{Analysis of deviance table on the probability of mortality by the end of the experiment using Type-II Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
+Family and Block were included as random effects.
+Abbreviations: df = degrees of freedom}
 	\begin{center}
@@
   Population & \Sexpr{mortTab["PopID", "chisq"]} & 
-    \Sexpr{mortTab["PopID", "df"]} & \Sexpr{mortTab["PopID", "pval"]} & \\
+    \Sexpr{mortTab["PopID", "df"]} & \Sexpr{mortTab["PopID", "pval"]} \\
   Temperature & \Sexpr{mortTab["TempTrt", "chisq"]} & 
-    \Sexpr{mortTab["TempTrt", "df"]} & \Sexpr{mortTab["TempTrt", "pval"]} & \\
+    \Sexpr{mortTab["TempTrt", "df"]} & \Sexpr{mortTab["TempTrt", "pval"]} \\
   Water & \Sexpr{mortTab["WaterTrt", "chisq"]} & 
-    \Sexpr{mortTab["WaterTrt", "df"]} & \Sexpr{mortTab["WaterTrt", "pval"]} & \\
+    \Sexpr{mortTab["WaterTrt", "df"]} & \Sexpr{mortTab["WaterTrt", "pval"]} \\
+  Temperature $\times$ Water & \Sexpr{mortTab["TempTrt:WaterTrt", "chisq"]} & 
+    \Sexpr{mortTab["TempTrt:WaterTrt", "df"]} & 
+    \Sexpr{mortTab["TempTrt:WaterTrt", "pval"]} \\
   Population $\times$ Temperature & \Sexpr{mortTab["PopID:TempTrt", "chisq"]} & 
     \Sexpr{mortTab["PopID:TempTrt", "df"]} & 
-    \Sexpr{mortTab["PopID:TempTrt", "pval"]} & \\
+    \Sexpr{mortTab["PopID:TempTrt", "pval"]} \\
   Population $\times$ Water & \Sexpr{mortTab["PopID:WaterTrt", "chisq"]} & 
     \Sexpr{mortTab["PopID:WaterTrt", "df"]} & 
-    \Sexpr{mortTab["PopID:WaterTrt", "pval"]} & \\
+    \Sexpr{mortTab["PopID:WaterTrt", "pval"]} \\
   Population $\times$ Temperature $\times$ Water & 
@@
     \Sexpr{mortTab["PopID:TempTrt:WaterTrt", "df"]} & 
-    \Sexpr{mortTab["PopID:TempTrt:WaterTrt", "pval"]} & \\	\bottomrule
+    \Sexpr{mortTab["PopID:TempTrt:WaterTrt", "pval"]} \\	\bottomrule
 	\end{tabular}
@@
    \topcaption[Important climatic variables]{Important climatic variables predicting latitude of \textit{M.
-cardinalis} populations (`Climate-Latitude') and the first principal component of traits measured in a common garden (`Climate-Trait').
-Importance and significance were determined using the variable selection using random forests (VSURF) algorithm (see `Methods').
+cardinalis} populations (`Climate-Latitude') and the first principal component of traits measured in a common garden (`Climate-TraitPC1').
+Point estimated climatic variables were measured from the exact location of collection; spatially averaged climatic variables were averaged from a 62-km neighborhood around population (see Material and Methods).
+Importance and significance were determined using the variable selection using random forests (VSURF) algorithm (see Material and Methods).
 Climatic variables are described in Table~\ref{table:TableS_ClimVars}.
-$\mu$ signifies the mean of the climate variables from 1981--2010; $\sigma$ indicates coeffiecient of variation among years}
+$\mu$ signifies the mean of the climate variables from 1981--2010; $\sigma$ indicates coeffiecient of variation among years.
+}
    \resizebox{6.5in}{!}{
@@
       \toprule
-    	Climate-Latitude variables & Climate-Trait variables & \\
+    	Climate-Latitude variables & Climate-TraitPC1 variables & \\
                             	   & Point estimated         & Spatially averaged \\
@@
 
+\clearpage
+
+%--------------------------------------------------
+% Supporting Figures
+%--------------------------------------------------
+
+\subsection*{Supporting Figures}
+
 % Figures showing latitudinal clines with specific traits
 
+<<FigS_GermLat, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+	pdf(paste(pathFig, "/FigureS_GermLat.pdf", sep = ""), 5, 5)
+	par(mai = c(1, 1, 0.1, 0.1))
+	plot(focClim$Latitude, 
+	     c(coef(fitGerm)[1], coef(fitGerm)[1] + coef(fitGerm)[2:16]), 
+	     xlim = c(30, 45), ylim = c(1.65, 2.65), xlab = "Latitude of Origin", 
+	     ylab = "Days to Germination [log scale]", cex.lab = 1.5, type = "n", 
+	     axes = F, frame.plot = T)
+	lowCI <- c(confint(fitGerm)[1, 1], 
+	           coef(fitGerm)[1] + confint(fitGerm)[2:16, 1])
+	uppCI <- c(confint(fitGerm)[1, 2], 
+	           coef(fitGerm)[1] + confint(fitGerm)[2:16, 2])
+	arrows(focClim$Latitude, lowCI, focClim$Lat, uppCI, length = 0.05, angle = 90,
+	       code = 3)
+	points(focClim$Latitude, c(coef(fitGerm)[1], 
+	                           coef(fitGerm)[1] + coef(fitGerm)[2:16]), 
+	       col = "white", bg = "black", pch = 21, cex = 1.5)
+
+	axis(1, at = seq(30, 45, 5), lwd = 0, lwd.ticks = 1,
+	     labels = expression(30*degree, 35*degree, 40*degree,	45*degree))	
+	axis(2, lwd = 0, lwd.ticks = 1, las = 1, at = log(seq(6, 14, 2)),
+	     labels = seq(6, 14, 2))
+
+	fit <- lm(c(coef(fitGerm)[1], coef(fitGerm)[1] + coef(fitGerm)[2:16]) ~
+	            Latitude, data = focClim)
+	points(range(focClim$Latitude), 
+	       predict(fit, new = data.frame(Latitude = range(focClim$Latitude))),
+	       type = "l", lwd = 2)
+
+	text(grconvertX(0.05, "npc"), grconvertY(0.95, "npc"), adj = c(0, 1),
+	     labels = bquote(italic(P) == .(round(summary(fit)[[4]][2, 4], 3))),)
+
+	dev.off()
+
+@
+
 \begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_GermLat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-	\caption[Southern populations germinate sooner.]{CAPTION}
-	\label{fig:Fig_DoG}
+	\caption[Southern populations germinate faster]{Southern populations germinate faster.
+Each point is a population of \textit{M.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted days to germination in days under growth chamber conditions (see Material and Methods).
+Bars around each point are 95\% confidence intervals.
+Predicted time to germination and confidence intervals are based on survival regression (see Materials and Materials).
+The line is the linear regression of log(model-predicated days to germination) $\sim$ latitude.
+The $P$-value of the regression is given in the upper left corner.}
+	\label{fig:FigS_GermLat}
 \end{figure}
 
+<<FigS_LERlat, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+	pdf(paste(pathFig, "/FigureS_LERlat.pdf", sep = ""), 5, 5)
+	par(mai = c(1, 1, 0.1, 0.1))
+	plot(focClim$Latitude, fitLER$lsmeans.table[1:16, "Estimate"], 
+	     xlim = c(30, 45), ylim = c(1.5, 3.1), xlab = "Latitude of Origin", 
+	     ylab = expression(paste("Leaf expansion rate (", mm~day^{-1}, ")")), 
+	     las = 1, cex.lab = 1.5, type = "n", axes = F, frame.plot = T)
+	arrows(focClim$Lat, fitLER$lsmeans.table[1:16, "Lower CI"], 
+	       focClim$Lat, fitLER$lsmeans.table[1:16, "Upper CI"], 
+	       angle = 90, length = 0.05, code = 3)
+	points(focClim$Lat, fitLER$lsmeans.table[1:16, "Estimate"], 
+	       col = "white", bg = "black", pch = 21, cex = 1.5)
+
+	axis(1, at = seq(30, 45, 5), lwd = 0, lwd.ticks = 1,
+	     labels = expression(30*degree, 35*degree, 40*degree, 45*degree))	
+	axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+
+	fit <- lm(fitLER$lsmeans.table[1:16, "Estimate"] ~ Latitude, data = focClim)
+
+	points(range(focClim$Latitude), 
+	       predict(fit, new = data.frame(Latitude = range(focClim$Latitude))),
+	       type = "l", lwd = 2)
+
+	text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1),
+	     labels = bquote(italic(P) == .(round(summary(fit)[[4]][2, 4], 3))))
+
+	dev.off()
+
+@
+
 \begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_LERlat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
-	\label{fig:Fig_LLL}
+	\caption[Southern populations grow faster (leaf expansion rate).]{Southern populations grow faster.
+Each point is a population of \textit{M.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted leaf expansion rate during the rosette phase.
+Bars around each point are 95\% confidence intervals.
+Predicted leaf expansion rate based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
+The line is the linear regression of model-predicated leaf expansion rate $\sim$ latitude.
+The $P$-value of the regression is given in the upper right corner.}
+	\label{fig:FigS_LERlat}
 \end{figure}
 
-\begin{figure}
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
+<<FigS_SERlat, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+	pdf(paste(pathFig, "/FigureS_SERlat.pdf", sep = ""), 5, 5)
+	par(mai = c(1, 1, 0.1, 0.1))
+	plot(focClim$Latitude, fitSER$lsmeans.table[1:16, "Estimate"], 
+	     xlim = c(30, 45), ylim = c(0.75, 1.65), xlab = "Latitude of Origin", 
+	     ylab = expression(paste("Stem elongation rate (", mm~day^{-1}, ")")), 
+	     las = 1, cex.lab = 1.5, type = "n", axes = F, frame.plot = T)
+	arrows(focClim$Lat, fitSER$lsmeans.table[1:16, "Lower CI"], 
+	       focClim$Lat, fitSER$lsmeans.table[1:16, "Upper CI"], 
+	       angle = 90, length = 0.05, code = 3)
+	points(focClim$Lat, fitSER$lsmeans.table[1:16, "Estimate"], 
+	       col = "white", bg = "black", pch = 21, cex = 1.5)
+
+	axis(1, at = seq(30, 45, 5), lwd = 0, lwd.ticks = 1,
+	     labels = expression(30*degree, 35*degree, 40*degree, 45*degree))	
+	axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+
+	fit <- lm(fitSER$lsmeans.table[1:16, "Estimate"] ~ Latitude, data = focClim)
+
+	points(range(focClim$Latitude), 
+	       predict(fit, new = data.frame(Latitude = range(focClim$Latitude))),
+	       type = "l", lwd = 2)
+
+	text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1),
+	     labels = bquote(italic(P) == .(round(summary(fit)[[4]][2, 4], 3))))
+
+	dev.off()
+
+@
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_SERlat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-	\caption[Southern populations grow faster (shoot elongation rate).]{CAPTION}
-	\label{fig:Fig_height}
+	\caption[Southern populations grow faster (stem elongation rate).]{Southern populations grow faster.
+Each point is a population of \textit{M.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted stem elongation rate during the bolting phase.
+Bars around each point are 95\% confidence intervals.
+Predicted stem elongation rate based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
+The line is the linear regression of model-predicated stem elongation rate $\sim$ latitude.
+The $P$-value of the regression is given in the upper right corner.}
+	\label{fig:FigS_SERlat}
 \end{figure}
 
-\section*{}
-
-\subsection*{Temperature treatments}
-
-We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool).
+<<FigS_PhotoLat, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+	pdf(paste(pathFig, "/FigureS_PhotoLat.pdf", sep = ""), 5.5, 5)
+	par(mai = c(1, 1.5, 0.1, 0.1))
+	plot(focClim$Latitude, fitPhoto$lsmeans.table[1:16, "Estimate"], 
+	     xlim = c(30, 45), ylim = c(8, 16), xlab = "Latitude of Origin", 
+	     ylab = expression(paste(atop("Photosynthetic rate", paste("(", mu, 
+	                             mol~CO[2]~s^{-1}~m^{-2}, ")")))), 
+	     las = 1, cex.lab = 1.5, type = "n", axes = F, frame.plot = T)
+	arrows(focClim$Lat, fitPhoto$lsmeans.table[1:16, "Lower CI"], 
+	       focClim$Lat, fitPhoto$lsmeans.table[1:16, "Upper CI"], 
+	       angle = 90, length = 0.05, code = 3)
+	points(focClim$Lat, fitPhoto$lsmeans.table[1:16, "Estimate"], 
+	       col = "white", bg = "black", pch = 21, cex = 1.5)
+
+	axis(1, at = seq(30, 45, 5), lwd = 0, lwd.ticks = 1,
+	     labels = expression(30*degree, 35*degree, 40*degree, 45*degree))	
+	axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+
+	fit <- lm(fitPhoto$lsmeans.table[1:16, "Estimate"] ~ Latitude, data = focClim)
+
+	points(range(focClim$Latitude), 
+	       predict(fit, new = data.frame(Latitude = range(focClim$Latitude))),
+	       type = "l", lwd = 2)
+
+	text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1),
+	     labels = bquote(italic(P) == .(round(summary(fit)[[4]][2, 4], 3))))
+
+	dev.off()
+
+@
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_Photolat.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Southern populations photosynthesize faster.]{Southern populations photosynthesize faster.
+Each point is a population of \textit{M.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted instantaneous photosynthetic rate.
+Bars around each point are 95\% confidence intervals.
+Predicted photosynthetic rates based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
+The line is the linear regression of model-predicated photosynthetic rate $\sim$ latitude.
+The $P$-value of the regression is given in the upper right corner.}
+	\label{fig:FigS_Photolat}
+\end{figure}
+
+% Figure comparing importance of climatic variables from point estimated and spatially averaged analyses
+
+<<FigS_PEvsSA, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+	pdf(paste(pathFig, "/FigureS_PEvsSA.pdf", sep = ""), 5.5, 5.5)
+	par(mai = c(1.5, 1.5, 0.1, 0.1))
+	plot(vsurfPC1$imp.mean.dec[vsurfPC1$imp.mean.dec.ind], 
+	     vsurfPC1_SA$imp.mean.dec[vsurfPC1_SA$imp.mean.dec.ind], 
+	     xlab = "", ylab = "", type = "n", axes = F, frame.plot = T)
+	 title(xlab = "Climate-TraitPC1 Importance\nPoint estimated", cex.lab = 1.5,
+	       line = 5) 
+	 title(ylab = "Climate-TraitPC1 Importance\nSpatially averaged", 
+	       cex.lab = 1.5, line = 4)
+
+	 text(vsurfPC1$imp.mean.dec[vsurfPC1$imp.mean.dec.ind], 
+	      vsurfPC1_SA$imp.mean.dec[vsurfPC1_SA$imp.mean.dec.ind],
+	      labels = climVars)
+
+	axis(1, lwd = 0, lwd.ticks = 1)
+	axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+
+	dev.off()
+
+@
+
+% *** Need to double check 62-km buffer results and make figure prettier *** %
+% \begin{figure}[h!]
+%	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_PEvsSA.pdf}}
+%	\usefont{T1}{phv}{m}{n}
+%	\fontsize{10}{12}
+%	\selectfont
+%	\caption[Point-estimated versus spatially averaged importance]{CAPTION}
+%	\label{fig:FigS_PEvsSA}
+% \end{figure}
+
+\clearpage
+
+\subsection*{Supporting Material and Methods}
+
+\subsubsection*{Temperature treatments}
+
+We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (WWC, Hot) and Little Jameson (LIJ, Cool).
 We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
@@
 Mean, min, and max temperature were highly correlated at both sites.
-We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models.
+We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally distributed random deviates with variance equal to the residual variance of the linear models.
 For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
@@
 
-\subsection*{Watering treatments}
-
+\subsubsection*{Watering treatments}
+
+%*** need to check chiller temp and make of chiller
 For watering treatments, we simulated two extreme types of streams where \textit{M.
@@
 In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
-In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down through the summer.
-In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER.
+In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season due to rain and snow melt, but gradually dries down through the summer.
+In both treatments, plants were bottom-watered using  water chilled to 7.5\celsius.
 Plants in the well-watered treatment were fully saturated every two hours during the day.

### 2016-10-14 ms/ms.tex
Finished shitty version for bioarxiv

@@
 
-\paragraph{Day of germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}.
-The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
-The signifcance of the Population effect was determined using analysis of deviance.
-Note that, unlike other traits discussed below, we did not include Block or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not been applied.
-
-\paragraph{Growth rate: leaf expansion and shoot elongation}
-
-We measured growth rate in two phases: leaf expansion as a rosette and shoot elongation after bolting.
+\paragraph{Days to germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}.
+We treated Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
+Statistical signifcance of the Population effect was determined using analysis of deviance.
+Note that, unlike other traits discussed below, we did not include Block, Treatment, or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not yet been applied.
+
+\paragraph{Growth rate: leaf expansion and stem elongation}
+
+We measured growth rate during two phases: leaf expansion as a rosette and stem elongation after bolting.
 We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
-We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.
-We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing).
-Both leaf expansion and shoot elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.
+We ceased measuring leaf length once it appeared to asymptote and growth shifted to stem elongation.
+We also censused plant height on 7 occasions (twice per week) between 29 May and 20 June (45 to 67 days after sowing).
+Both leaf expansion and stem elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and stem growth) using empirical Bayes' estimates from linear mixed-effects models fit with the R package \pkg{lme4} version 1.1-12 \citep{Bates_etal_2015}.
 
 \paragraph{Photosynthesis}
-During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of 329 plants evenly spread between treatments and families within populations.
-The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
-All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
-During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.
-As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out.
+During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate on a subset of 329 plants evenly spread between treatments and families within populations.
+The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6-cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
+We made all measurements at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$ of photosynthetically active radiation), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
+During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level could be measured under the same conditions.
+% As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out.
 Therefore, we calculated the `intrinsic' photosynthetic rate for each individual as its residual deviance from the regression between stomoatal conductance and photosynthetic rate across all plants.
@@
 We included Family and Block as random effects.
-We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
-
-\paragraph{Biomass at harvest}
-
-We harvested plants that survived until the end of the experiment from 3 to 16 July (80 to 93 days after sowing).
-We separated individual plants into three biomass fractions: leaves, shoots, and belowground (roots plus rhizomes).
-Belowground fractions were rinsed thoroughly in water to remove sand.
-After completely drying in a forced air oven (Shel Lab, Sheldon Manufacturing Inc., Oregon, USA), we weighed the dry mass of each fraction using an analytical balance.
+We assessed significance of fixed effects using Type-II Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
 
@@
 The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA).
-The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.
-
-\subsection*{Selective agents and environmental correlates}
-
-We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+The first principal component of these traits (TraitPC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.
+
+\subsection*{Identifying putative selective agents}
+
+We found that a population's position along TraitPC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
 cardinalis}.
-We looked for overlap between climatic variables that best predict latitude of \textit{M.
-cardinalis} occurrence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations.
-For brevity, we refer to these as `Climate-Latitude' and `Climate-TraitPC1' variables.
-We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}.
-From VSURF models, we kept only variables selected for prediction, the most stringent criterion.
-
-If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high?
+Hypothesis 1: if a single climatic factor explained more trait variation than latitude alone, this would suggest that that factor is a selective agent underlying the latitudinal cline in our 16 focal populations.
+Hypothesis 2: if multiple climatic factors together explained trait variation,  we interpreted this as evidence that multiple climatic factors together have generated the latitudinal cline.
+We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables.
+In addition, to help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{M.
+cardinalis} occurrence records, not just the 16 focial populations.
+We refer to these climatic factors as `Climate-Latitude' variables.
+The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{M.
+cardinalis} range.
+We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}.
+VSURF ranks variables by their importance over regression trees in the forest.
+We kept only variables selected for prediction, the most stringent criterion.
+
+% This is now stated above.
+%If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high?
 maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}.
@@
 Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
-We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km buffer (`spatially averaged Climate-TraitPC1').
-
-For these analyses, we compiled a representative set of 178 recent (since 2000) known \textit{M.
+We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km climatic neighborhood (`spatially averaged Climate-TraitPC1').
+
+For Climate-Latitude analyses, we compiled a representative set of 356 recent (since 2000) known \textit{M.
 cardinalis} occurences.
 These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For each occurence, we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
+For both Climate-TraitPC1 analyses (16 focal populations) and Climate-Latitude (many populations), we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
 Monthly interpolated climate layers were calculated using ClimateWNA version 5.30 \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}.
+For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} version 1.1-1 \citep{Hijmans_etal_2016}.
 In total, we included 24 climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
@@
 
-Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis}, as evidenced by large and significant population effects (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured 74.4 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
+There are strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
+cardinalis}, as evidenced by large and significant population effects for these traits (Table~\ref{table:Table_AnovaSummary}).
+A single principal component captured 71.6 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
 As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.
+There were similar latitudinal clines for individual traits underlying PC1 (Fig S1-S4).
 
@@
    \resizebox{6.5in}{!}{
-   \begin{tabular}{@{} llllllll @{}}
+   \begin{tabular}{@{} lllllll @{}}
       \toprule
-    {\raggedleft Trait}                          & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
-    Statistical model              & survreg     & lmer           & lmer            & lmer           & lmer          & glmer \\
+    {\raggedleft Trait}                          & Germination & Leaf expansion & Stem elongation & Photosynthesis & Mortality \\
+    Statistical model              & survreg     & lmer           & lmer            & lmer           & glmer \\
       \midrule
   Population                       &
-    ***  &
     ***  &
@@
     ** & 
-     &
     *** & \\
@@
     *   &
-       &
        &
@@
     * & 
-     &     
      & \\
@@
      & 
-     &     
      & \\
@@
     NA                                                          &
-     &
      &
@@
 For \textit{M.
-cardinalis}, interannual variation in precipitation over the past 30 years is very closely related to the latitude of recently recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
-Between year variation in precipitation was also strongly correlated with position along the fast-slow growth axis, but climate based on point estimates gave somewhat different results than climate based on spatially averaged values.
-The two most important point estimated Climate-Latitude variables were also strong predictors of position along the fast-slow growth axis (Fig.~\ref{fig:Fig_ClimVarImp}B).
-For spatially averaged climate, interannual variation in precipitation seasonality was the only climatic variable selected, (Fig.~\ref{table:TableS_ClimVarImp}) \cdm{Do we need a figure for this?}.
-Overall, there was good but imperfect agreement between the weight assigned to climatic variables using point estimated and spatially averaged climate (Fig.~\ref{fig:Fig_TBD}) \cdm{This figure will show strong, positive correlation between weights given to climatic variables from point estimated and spatially averaged analyses}.
-
-Overlap between Climate-Latitude and point estimated Climate-Trait variables suggests that interannual variation in precipitation is an important selective agent in \textit{M.
-cardinalis}.
-Specifically, we hypothesize that more frequent droughts (greater precipitation cv) in Southern populations selects for an `annual-ized' life history, as we detail in the Discussion.
-Even though the spatially averaged Climate-Trait variable was not among the Climate-Latitude variables, it points to a similar conclusion.
-Specifically, this analysis showed that greater interannual variation in precipitation seasonality, which is also driven by precipitation variability (i.e.
-more frequent drought years) in Southern California, is associated with faster growth in southern populations.
-We must qualify these results because our analysis obviously cannot rule out that alternative variables not included in the analysis may be more important.
+cardinalis}, interannual variation in precipitation over 30 years (1981--2010) is very closely related to the latitude of recently recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
+Specifically, precipitation is more variable at lower latitude.
+The two most important Climate-Latitude variables were the interannual variation in total precipitation (bio12$_\sigma$; see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations) and precipitation in the wettest quarter (bio16$_\sigma$).
+Note that the coefficient of variation of a climatic factor is subscripted with $\sigma$ whereas the mean is subscripted with $\mu$.
+Bio12$_\sigma$ and bio16$_\sigma$ are very similar because in Mediterranean climates of California, most precipitation occurs in the winter quarter.
+These Climate-Latitude variables were also the two most important point estimated Climate-TraitPC1 variables (Fig.~\ref{fig:Fig_ClimVarImp}B).
+Populations from Southern, more variable environments grew faster.
+However, neither climatic variable alone explained more variation in TraitPC1 than latitude (latitude $r^2=0.55$; bio16$_\sigma$ $r^2=0.52$; bio12$_\sigma$ $r^2=0.47$).
+Interannual variation in chilling degree-days (DD\_0$_\sigma$) was also common to both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
+Southern, faster growing populations are from climates with less variation in chilling degree-days.
+However, DD\_0$_\sigma$ and bio16$_\sigma$ together still only explained slightly more variation in TraitPC1 than latitude alone (latitude $r^2=0.55$ versus DD\_0$_\sigma$+bio16$_\sigma$ $r^2=0.6$).
+This suggests that interannual variation in precipitation explains most of the latitudinal cline, but other variables such as variation in chilling degree-days may contribute.
+
+There was no overlap between Climate-Latitude and spatially averaged Climate-TraitPC1 (Fig.~\ref{table:TableS_ClimVarImp}).
+This is not because climatic neighboorhood did not predict TraitPC1; there were significant correlations between important spatially averaged climatic factors and TraitPC1 that explained as much variation as latitude (data not shown).
+If these correlations are spurious, this indicates that climatic neighborhood is not as imporant as climate in the immediate vicinity of a population.
+Conversely, the climatic neighborhood may be important among our focal populations and depart from rangewide correlations between latitude and climate.
+Since these analyses are exploratory and prone to overinterpretation, we believe the closer overlap between Climate-Latitude and point estimated Climate-TraitPC1 variables provides more compelling evidence for putative selective agents.
 
@@
 A.
-Using Random Forest regression, we identified 10 climatic variables significantly (high importance) associated with latitude of \textit{M.
+Using Random Forest regression, we identified 11 climatic variables significantly (high importance) associated with latitude of \textit{M.
 cardinalis} occurences.
 B.
-The two most important Climate-Latitude variables were also the two most important Climate-Trait variables.
+The two most important Climate-Latitude variables were also the two most important Climate-TraitPC1 variables.
 Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
-Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp) or Precipitation (Prec); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
+Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp), Precipitation (Prec), Chilling (chilling degree-days), Snow (precipitation as snow); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
 	\label{fig:Fig_ClimVarImp}
@@
 
+% Analysis of deviance table for germination
+
+% *** NEED TO DO TABLE IF I WANT IT ***
+
 % ANOVA table for LER
@@
     11.92 & 0.019 \\
+  Temperature $\times$ Water & 0.03 & 
+    0.03 & 
+    1 & 
+    4 & 
+    0.07 &
+    0.801 \\
   Population $\times$ Temperature & 2.76 & 
@@
     3.7 & 0.113 \\
+  Temperature $\times$ Water & 0.9 & 
+    0.9 & 
+    1 & 
+    4 & 
+    5.2 &
+    0.093 \\
   Population $\times$ Temperature & 3.6 & 
@@
 	\end{tabular}
-	\label{table:TableS_fitLER}
+	\label{table:TableS_fitSER}
 	\end{center}
@@
     0.066 \\
+  Temperature $\times$ Water & 0.7 & 
+    0.7 & 
+    1 & 
+    3 & 
+    0.09 &
+    0.781 \\
   Population $\times$ Temperature & 218.6 & 
@@
 
-% ANOVA table for intrinsic photosynthetic rate
+% Analysis of deviance table for mortality
 
@@
 	\selectfont
-	\caption[ANOVA table, photosynthesis]{Sum of squares (SS); Mean sum of squares (MS); Numerator degrees of freedom (df1); Denominator degrees of freedom (df2)}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
-  Population & 347.7 & 
-    0.1074 & 
-    15 &
-    305 & 
-    2.87 &
-    3.1 $\times10^{-4}$ \\
-  Temperature & 0.0105 & 
-    0.0105 & 
-    1 &
-    6 & 
-    0.28 &
-    0.616 \\
-  Water & 8e-04 & 
-    8e-04 & 
-    1 &
-    5 & 
-    0.02 &
-    0.888 \\
-  Population $\times$ Temperature & 
-    0.6463 & 
-    0.0431 & 
-    15 & 
-    289 & 
-    1.16 &
-    0.300 \\
-  Population $\times$ Water & 0.4592 & 
-    0.0306 & 
-    15 &
-    274 &
-    0.82 & 
-    0.656 \\
-  Population $\times$ Temperature $\times$ Water & 
-    0.4692 & 
-    0.0313 & 
-    15 & 
-    259 & 
-    0.83 & 
-    0.644 \\
-	\bottomrule
-	\end{tabular}
-	\label{table:TableS_fitIntrnPhoto}
-	\end{center}
-\end{table}
-
-% Analysis of deviance table for mortality
-
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, mortality]{CAPTION}
+	\caption[Analysis of deviance table, mortality]{Analysis of deviance table on the probability of mortality by the end of the experiment using Type-II Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
+Family and Block were included as random effects.
+Abbreviations: df = degrees of freedom}
 	\begin{center}
@@
   Population & 32 & 
-    31 & 0.419 & \\
+    31 & 0.419 \\
   Temperature & 31.8 & 
-    6 & 1.8 $\times10^{-5}$ & \\
+    6 & 1.8 $\times10^{-5}$ \\
   Water & 69.2 & 
-    12 & 4.6 $\times10^{-10}$ & \\
+    12 & 4.6 $\times10^{-10}$ \\
+  Temperature $\times$ Water & 20.7 & 
+    1 & 
+    5.3 $\times10^{-6}$ \\
   Population $\times$ Temperature & 5.6 & 
     15 & 
-    0.985 & \\
+    0.985 \\
   Population $\times$ Water & 8.6 & 
     15 & 
-    0.897 & \\
+    0.897 \\
   Population $\times$ Temperature $\times$ Water & 
@@
     15 & 
-    1.000 & \\	\bottomrule
+    1.000 \\	\bottomrule
 	\end{tabular}
@@
 
+\clearpage
+
+%--------------------------------------------------
+% Supporting Figures
+%--------------------------------------------------
+
+\subsection*{Supporting Figures}
+
 % Figures showing latitudinal clines with specific traits
@@
 \begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_DoG_Lat.pdf}}
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_GermLat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-	\caption[Southern populations germinate sooner.]{CAPTION}
-	\label{fig:Fig_DoG}
+	\caption[Southern populations germinate faster]{Southern populations germinate faster.
+Each point is a population of \textit{M.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted days to germination in days under growth chamber conditions (see Material and Methods).
+Bars around each point are 95\% confidence intervals.
+Predicted time to germination and confidence intervals are based on survival regression (see Materials and Materials).
+The line is the linear regression of log(model-predicated days to germination) $\sim$ latitude.
+The $P$-value of the regression is given in the upper left corner.}
+	\label{fig:FigS_GermLat}
 \end{figure}
@@
 \begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_LLL_Lat.pdf}}
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_LERlat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-	\caption[Southern populations grow faster (leaf expansion rate).]{CAPTION}
-	\label{fig:Fig_LLL}
+	\caption[Southern populations grow faster (leaf expansion rate).]{Southern populations grow faster.
+Each point is a population of \textit{M.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted leaf expansion rate during the rosette phase.
+Bars around each point are 95\% confidence intervals.
+Predicted leaf expansion rate based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
+The line is the linear regression of model-predicated leaf expansion rate $\sim$ latitude.
+The $P$-value of the regression is given in the upper right corner.}
+	\label{fig:FigS_LERlat}
 \end{figure}
 
-\begin{figure}
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_Height_Lat.pdf}}
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_SERlat.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-	\caption[Southern populations grow faster (shoot elongation rate).]{CAPTION}
-	\label{fig:Fig_height}
+	\caption[Southern populations grow faster (stem elongation rate).]{Southern populations grow faster.
+Each point is a population of \textit{M.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted stem elongation rate during the bolting phase.
+Bars around each point are 95\% confidence intervals.
+Predicted stem elongation rate based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
+The line is the linear regression of model-predicated stem elongation rate $\sim$ latitude.
+The $P$-value of the regression is given in the upper right corner.}
+	\label{fig:FigS_SERlat}
 \end{figure}
 
-\section*{}
-
-\subsection*{Temperature treatments}
-
-We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool).
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_Photolat.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Southern populations photosynthesize faster.]{Southern populations photosynthesize faster.
+Each point is a population of \textit{M.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted instantaneous photosynthetic rate.
+Bars around each point are 95\% confidence intervals.
+Predicted photosynthetic rates based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
+The line is the linear regression of model-predicated photosynthetic rate $\sim$ latitude.
+The $P$-value of the regression is given in the upper right corner.}
+	\label{fig:FigS_Photolat}
+\end{figure}
+
+% Figure comparing importance of climatic variables from point estimated and spatially averaged analyses
+
+% *** Need to double check 62-km buffer results and make figure prettier *** %
+% \begin{figure}[h!]
+%	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_PEvsSA.pdf}}
+%	\usefont{T1}{phv}{m}{n}
+%	\fontsize{10}{12}
+%	\selectfont
+%	\caption[Point-estimated versus spatially averaged importance]{CAPTION}
+%	\label{fig:FigS_PEvsSA}
+% \end{figure}
+
+\clearpage
+
+\subsection*{Supporting Material and Methods}
+
+\subsubsection*{Temperature treatments}
+
+We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (WWC, Hot) and Little Jameson (LIJ, Cool).
 We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.

### 2016-10-23 ms/ms.Rnw
Revised Discussion

@@
 %	- maybe temporal source-sink makes more sense
-
-%* explain rhizomes
 
@@
 cardinalis} we found substantial genetically based variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
-If this variation is adaptive, it suggests that the fitness tradeoff between doing well in low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
+If this variation is adaptive, it suggests that the fitness tradeoff between low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
 That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatially heterogenous selection.
@@
 
-Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years (see next section), could have evolved the ability to tolerate drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
+Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years (see below), could have adapted by tolerating drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
 We found no evidence for this; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
@@
 Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Mediterranean climates of western North America.
-But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and vice versa, leading to strong fitness tradeoffs.
+But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and \textit{vice versa}, leading to strong fitness tradeoffs.
 Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
 Demography and gene flow may reinforce niche conservatism.
-A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher resource environments.
-Infrequent wet years may also produce most seeds, so selection weighted towards wet environment even if dry years are more frequent.
+A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher-resource environments.
+Infrequent wet years may also produce most seeds, so selection is weighted towards alleles that have high fitness in the wet environment, even if dry years are more frequent.
 Finally, gene flow, which is generally high among \textit{M.
@@
 
-Based on the available data, interannual variation in annual or winter precipitation (these are closely correlated in Mediterranean climates) may be the selective agent driving variation in physiological traits.
-Variation in precipitation was best predicted latitude of recent \textit{M.
-cardinalis} occurences and trait variation along the fast-slow growth continuum (Fig.~\ref{fig:Fig_ClimVarImp}).
-A life history tradeoff between allocation to growth in the current year at the expense of future years could explain this pattern.
-In southern populations with more frequent droughts capable of killing rhizomes, a more annualized strategy could be favored.
-Conversely, in more predictable northern environments, lifetime fitness may be optimized when a significant fraction of assimilate is allocated below ground for future years.
-Although this hypothesis remains to be directly tested, a few independent lines of evidence are consistent with it.
+To reiterate, these results indicate that differences in physiology and growth, but not responses to temperature and drought, mediate local adaptation to climate in \textit{M.
+cardinalis}.
+Next, we would like to understand why variation in these particular traits may be adaptive.
+We argue that temporally more variable environments, as experienced by southern populations, select for a more ‘annualized’ life-history strategy.
+
+Greater investment in aboveground growth, as opposed to belowground storage for future seasons, may be favoured in climates with more frequent drought years, but maladaptive in climates with more consistent precipitation.
+Suppose plants that grow quickly and allocate new resources to continued growth rather than storage have higher fitness over a single growing season.
+However, by not allocating resources to storage, these fast-growing plants begin future seasons at a deficit.
+Therefore, in a stable environment where winter survivorship is assured in most years, failure to store resources may reduce lifetime fitness.
+But the environment is not constant.
+For perennial herbs in Mediterranean climates, a dry winter (rainy season) can kill the rhizomes (underground stems that store nutrients for future growth) before emergence or aboveground stems before flowering.
+If drought years occur frequently enough, selection may favour the fast-growing strategy because there is no advantage to storage if drought kills plants before flowering.
+Considering life-history strategy as a continuum from no storage (annual) to lots of storage (perennial), we hypothesize that the optimal allocation to aboveground growth is more ‘annualized’ in southern climates that have greater interannual variation in precipitation.
+
+The hypothesis that greater precipitation variability selects for an annualized life history is tentative, but consistent with theory, data from other species, and other observations in \textit{M.
+cardinalis}.
+Life history theory shows that less variable environments are one factor that favours the evolution of perenniality \citep{Stearns_1976, Iwasa_Cohen_1989, Friedman_Rubin_2015}.
+Populations of the perennial \textit{Plantago asiatica} show a similar latitudinal cline in growth and allocation to storage \citep{Sawada_etal_1994}, though these authors attribute the cline to variation in growing season length.
+There are also life history clines in the closely related species \textit{M.
+guttatus}, but the underlying traits and climatic drivers are quite different.
+Annual \textit{M.
+guttatus} flower sooner and produce fewer stolons in response to climates with shorter seasons and more intense summer drought \citep{Lowry_Willis_2010, Friedman_etal_2015, Kooyers_etal_2015}.
+In contrast, there are no truly annual (monocarpic and semelparous) populations of \textit{M.
+cardinalis}.
+Rather, our hypothesis states climatic variability selects on quantitative variation in allocation to growth versus storage.
+Several lines of evidence provide tentative support for this hypothesis.
 Preliminary surveys suggest that northern populations not only grow slower, but also produce greater numbers of rhizomes (C.D.
@@
 data), suggesting an allocation tradeoff.
-Ecological niche models also show that occurence of southern populations is best predicted by recent climate (< 5 years), whereas northern occurences are best predicted by climate over the previous 30 years (M.
+Ecological niche models also show that occurrence of southern populations is best predicted by recent climate (< 5 years), whereas northern occurrences are best predicted by climate over the previous 30 years (M.
 Bayly \& A.
@@
 data).
-There is a lot of interest in understanding how organisms will respond to changes in climatic variation, not just changes in the average climate.
-Our data indeed suggest that variation may be more important than the mean.
+Although much work remains, the present works indicates that future studies of \textit{M.
+cardinalis} will deliver new, complementary insight into the physiological traits and climatic drivers of local adaptation.
 

### 2016-10-26 ms/ms.Rnw
New figure of climate variable procedure

@@
 %cartoon of whole procedure for climate variables
-%figure showing correlation b/t bio12 and 
-%state that VSURF importance is different scales in different models
-%moved up three hypotheses part to introduction
-%play up lack of GxE and ???
-%** correct # of Pop x Treatment
 % model of growth chamber
@@
 % TO DO:
-% 1. Is water regime on figure 1 correct? (data is in office)
 % 2. Double check spatially averaged analysis
@@
 We kept only variables selected for prediction, the most stringent criterion.
-
-% This is now stated above.
-%If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high?
-maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}.
-If latitude is correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline, then several climatic variables should be weakly correlated \ala{fix this to mirror prior}.
+A visual overview of how we selected climatic variables is depicted in Fig~\ref{fig:Fig_VarSelect}.
 
@@
 We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km climatic neighborhood (`spatially averaged Climate-TraitPC1').
+
+<<Fig_VarSelect, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  library(raster)
+	library(rgeos)
+  library(extrafont)
+
+  # Get maps of California and Oregon
+  usa <- getData('GADM', country = "USA", level = 1, path = pathDat)
+  states <- gSimplify(usa[usa@data$NAME_1 %in% c("California", "Oregon",
+                                                 "Nevada", "Idaho"), ], 
+                      tol = 0.01, topologyPreserve = TRUE)
+
+  # Load thinned occurences used in Climate-Lat analyses and convert to Spatial class
+  occ <- read.csv(paste(pathDat, "/thinned1OccLatLon.csv", sep = ""), 
+                  row.names = NULL)
+  coordinates(occ) <- ~Longitude+Latitude
+  projection(occ) <- CRS('+proj=longlat');
+
+  # Convert focal population coordinates to Spatial class
+  foc <- focClim[, c("Longitude", "Latitude")]
+  coordinates(foc) <- ~Longitude+Latitude
+  projection(foc) <- CRS('+proj=longlat');
+
+  # Clip and map
+  clipExtent <- as(extent(floor(min(occ$Longitude) * 10) / 10, 
+	   											ceiling(max(occ$Longitude) * 10) / 10, 
+	  											floor(min(occ$Latitude) * 10) / 10, 
+		  										ceiling(max(occ$Latitude) * 10) / 10), 
+			  					 "SpatialPolygons")
+  proj4string(clipExtent) <- CRS(proj4string(states))
+  states <- gIntersection(states, clipExtent, byid = TRUE)
+
+  # Transform to Albert's Equal Area Projection
+  projAEA <- sprintf("+proj=aea +lat_1=%s +lat_2=%s +lat_0=%s +lon_0=%s +x_0=0 +y_0=0",
+                     extent(clipExtent)[3], extent(clipExtent)[4], 
+		  							 extent(clipExtent)[1], extent(clipExtent)[2])
+  states <- spTransform(states, CRS(projAEA))
+  occAEA <- spTransform(occ, CRS(projAEA))
+  focAEA <- spTransform(foc, CRS(projAEA))
+
+  # PLOT
+  pdf(paste(pathFig, "/Figure_VarSelectBase.pdf", sep = ""), w = 6, h = 7, 
+	  	useDingbat = F)
+  loadfonts()
+  mat <- matrix(c(1, 4, 7,
+	  							1, 4, 7,
+		  						1, 4, 7,
+			  					2, 5, 8,
+				  				2, 5, 8,
+					  			3, 6, 9,
+						  		3, 6, 9), ncol = 3, byrow = T)
+  layout(mat)
+
+  # A. Example Climate-Latitude variable
+
+	  ## Panel 1: Map
+	  par(mai = rep(0.5, 4), family = "sans")
+	  plot(states, main = "A. Climate-Latitude variables")
+	  points(occAEA, pch = 19, cex = 0.5)
+
+	  # Position of Cuyamaca Rancho population in device units
+	  xCUR <- grconvertX(focAEA$Longitude["CUR"], from = "user", to = "device")
+	  yCUR <- grconvertY(focAEA$Latitude["CUR"], from = "user", to = "device")
+
+	  ## Panel 2: Cartoon of point versus spatial average climate estimate
+	  par(mai = rep(0, 4), family = "LM Roman 10")
+
+	    ### River 1
+	    x1 <- seq(-1, 1, length.out = 1e2)
+	    r1 <- function(x) exp(x) - exp(0) + sin(x * 10) / 10
+	    y1 <- r1(x1)
+
+	    ### River 2
+	    x2 <- seq(0, 1, length.out = 1e2)
+	    r2 <- function(x) exp(-x) - exp(0) + sin(x * 10) / 10
+	    y2 <- r2(x2)
+
+	  plot(0, 0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1), axes = F)
+	  clip(-1, 1, -1, 1)
+	  points(x1, y1, type = "l", col = "blue", lwd = 5)
+	  points(x2, y2, type = "l", col = "blue", lwd = 5)
+
+	  # Draw buffer area
+	  x <- seq(-1, 0, length.out = 1e3)
+	  clip(-2, 2, -2, 2)
+	  polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
+	          col = "white", border = "white", lwd = 2)
+	  polygon(c(x + 1, rev(x + 1)), c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
+	          col = "white", border = "white", lwd = 2)
+	  polygon(c(x, rev(x)), -c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
+	          col = "white", border = "white", lwd = 2)
+	  polygon(c(x + 1, rev(x + 1)), -c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
+	          col = "white", border = "white", lwd = 2)
+
+	  # Draw point and label
+	  points(0, 0, pch = 19, cex = 3)
+	  text(0, 0, labels = expression(italic(x[paste(i, ",", 0)])), adj = 1.5, 
+	       cex = 1.5)
+	  # lines(c(grconvertX(xCUR, from = "device", to = "user"), 0),
+	  #     c(grconvertY(yCUR, from = "device", to = "user"), 0),
+	  #       lwd = 2)
+    text(0, 1, labels = expression(Climate[italic(i)] == 
+           italic(x[paste(i, ",", 0)])), cex = 1.5, pos = 3, offset = 1)
+
+	  ## Panel 3: Climate-Latitude relationship
+	  set.seed(01152016)
+  	y <- (occ$Latitude - mean(occ$Latitude)) / sd(occ$Latitude) + 
+		      rnorm(length(occ), 0, sqrt(0.1))
+	  par(mai = c(0.6, 0.6, 1/25, 1/25), family = "sans")
+	  plot(y, occ$Latitude, las = 1, axes = F, frame.plot = T, 
+		  	 xlim = c(-2.5, 2.5), ylim = c(32, 45), cex.lab = 1.5,
+			   pch = 19, col = "grey", xlab = "", ylab = "Latitude of origin")
+	  title(xlab = "Climate variable", cex.lab = 1.5, line = 1)
+
+  	for (i in seq(32, 44, 2)) {
+		  axis(2, at = i, labels = bquote(.(i)*degree), las = 1)
+	  	}
+	  fit <- lm(occ$Latitude ~ y)
+	  lines(range(y), range(predict(fit)), lwd = 2)
+
+   # B. Example Climate-Trait variable, point estimated
+
+  	## Panel 1: Map
+	  par(mai = rep(0.5, 4), family = "sans")
+	  plot(states, main = "B. Climate-Trait variables")
+	  mtext(text = "point estimated", side = 3)
+	  points(focAEA, pch = 19, cex = 0.5)
+
+	  # Position of Cuyamaca Rancho population in device units
+	  xCUR <- grconvertX(focAEA$Longitude["CUR"], from = "user", to = "device")
+	  yCUR <- grconvertY(focAEA$Latitude["CUR"], from = "user", to = "device")
+
+	  ## Panel 2: Cartoon of point versus spatial average climate estimate
+	  par(mai = rep(0, 4), family = "LM Roman 10")
+	  plot(0, 0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1), axes = F)
+	  clip(-1, 1, -1, 1)
+	  points(x1, y1, type = "l", col = "blue", lwd = 5)
+	  points(x2, y2, type = "l", col = "blue", lwd = 5)
+
+	  # Draw buffer area
+	  x <- seq(-1, 0, length.out = 1e3)
+	  clip(-2, 2, -2, 2)
+	  polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
+	          col = "white", border = "white", lwd = 2)
+	  polygon(c(x + 1, rev(x + 1)), c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
+	          col = "white", border = "white", lwd = 2)
+	  polygon(c(x, rev(x)), -c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
+	          col = "white", border = "white", lwd = 2)
+	  polygon(c(x + 1, rev(x + 1)), -c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
+	          col = "white", border = "white", lwd = 2)
+
+	  # Draw point and label
+	  points(0, 0, pch = 19, cex = 3)
+	  text(0, 0, labels = expression(italic(x[paste(i, ",", 0)])), adj = 1.5, 
+	       cex = 1.5)
+	  # lines(c(grconvertX(xCUR, from = "device", to = "user"), 0),
+	  #       c(grconvertY(yCUR, from = "device", to = "user"), 0),
+	  #       lwd = 2)
+    text(0, 1, labels = expression(Climate[italic(i)] == 
+           italic(x[paste(i, ",", 0)])), cex = 1.5, pos = 3, offset = 1)
+
+	  ## Panel 3: Climate-trait relationship
+	  par(mai = c(0.6, 0.6, 1/25, 1/25), family = "sans")
+	  y <- (foc$Latitude - mean(foc$Latitude)) / sd(foc$Latitude) + 
+		      rnorm(length(foc), 0, sqrt(0.1))
+	  plot(y, foc$Latitude, las = 1, axes = F, frame.plot = T, 
+		  	 xlim = c(-2.5, 2.5), ylim = c(32, 45),
+			   pch = 19, col = "grey", xlab = "", ylab = "")
+	  title(xlab = "Climate variable", ylab = "Trait", cex.lab = 1.5, line = 1)
+	  fit <- lm(foc$Latitude ~ y)
+	  lines(range(y), range(predict(fit)), lwd = 2)
+
+  # C. Example Climate-Trait variable, spatially averaged
+
+  	## Panel 1: Map
+  	par(mai = rep(0.5, 4), family = "sans")
+  	plot(states, main = "C. Climate-Trait variables")
+  	tmp <- raster(states)
+  	mtext(text = "spatially averaged", side = 3)
+  	points(focAEA, pch = 21, cex = 2, col = "black", bg = "white")
+	  points(focAEA, pch = 19, cex = 0.5)
+
+	  # Position of Cuyamaca Rancho population in device units
+	  xCUR <- grconvertX(focAEA$Longitude["CUR"], from = "user", to = "inches")
+	  yCUR <- grconvertY(focAEA$Latitude["CUR"], from = "user", to = "inches")
+
+  	## Panel 2: Cartoon of point versus spatial average climate estimate
+	  par(mai = rep(0, 4), family = "LM Roman 10")
+
+  		### Climate grid
+
+  			#### Functions to find closest point along "stream"
+				fn <- function(par, x, y) {
+					if (x < 0) {
+						d1 <- sqrt((par - x) ^ 2 + (r1(par) - y) ^ 2)
+						return(d1)
+					} else {
+						d1 <- sqrt((par - x) ^ 2 + (r1(par) - y) ^ 2)
+						d2 <- sqrt((par - x) ^ 2 + (r2(par) - y) ^ 2)
+						return(min(c(d1, d2)))
+					}
+				}
+
+				getStreamDist <- function(x, y) {
+					fit <- optim(x, fn, x = x, y = y, method = "Brent", lower = -2, 
+											 upper = 2)
+					retval <- setNames(c(fit$par, fit$value), c("s", "dist"))
+					return(retval)
+				}
+
+			  #### "Climate" is fxn of distance from stream and pos along x-axis
+			  xy <- expand.grid(x = seq(-0.95, 0.95, 0.1), 
+			                    y = seq(-0.95, 0.95, 0.1))
+			  xyz <- cbind(xy, 
+			               t(apply(xy, 1, function(X) getStreamDist(X[1], X[2]))))
+			  xyz$z <- xyz$dist + xyz$x * 0.3
+			  xyz$z <- xyz$z - min(xyz$z)
+			  xyz$z <- xyz$z / max(xyz$z)
+
+		  plot(0, 0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1), axes = F)
+		  clip(-1, 1, -1, 1)
+		  points(x1, y1, type = "l", col = "blue", lwd = 5)
+		  points(x2, y2, type = "l", col = "blue", lwd = 5)
+
+		  apply(xyz, 1, function(X) rect(X["x"] - 0.05, X["y"] - 0.05, 
+																	   X["x"] + 0.05, X["y"] + 0.05,
+														 			   col = rgb(0, 0, 1, alpha = X["z"]), 
+																	   border = NA))
+
+			# Draw buffer area and circle
+			x <- seq(-1, 0, length.out = 1e3)
+			clip(-2, 2, -2, 2)
+			polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
+			        col = "white", border = "white", lwd = 2)
+			polygon(c(x + 1, rev(x + 1)), c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
+			        col = "white", border = "white", lwd = 2)
+			polygon(c(x, rev(x)), -c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
+			        col = "white", border = "white", lwd = 2)
+			polygon(c(x + 1, rev(x + 1)), -c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
+			        col = "white", border = "white", lwd = 2)
+
+			x <- seq(-1, 1, length.out = 1e3)
+			polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), -sqrt(1 - x ^ 2)),
+			        border = "black", lwd = 2)
+
+			# Draw points and label
+	    points(0, 0, pch = 19, cex = 3)
+	    text(0, 0, labels = expression(italic(x[paste(i, ",", 0)])), adj = 1.5, 
+	         cex = 1.5)
+	    # Start line from outside of point
+	    # point radius = cex * size of point = cex * 0.75 * character height
+	    # ptRad <- 0.1 * 2 * 0.75 * par()$cin[2]
+	    # lines(c(grconvertX(xCUR - ptRad, from = "inches", to = "user"), -1),
+	    #     c(grconvertY(yCUR, from = "inches", to = "user"), 0),
+	    #       lwd = 2)
+      text(0, 1, labels = expression(Climate[italic(i)] == 
+           sum(over(x[paste(italic(i), ",", italic(j))], italic(n)), 
+               scriptstyle(italic(j) == 1), italic(n))), 
+           cex = 1.5, pos = 3, offset = 1)
+
+      points(c(-2/3, -1/3, 1/2, 1/3, 2/3),
+             c(r1(-2/3), r1(-1/3), r1(1/2), r2(1/3), r2(2/3)),
+             cex = 2, pch = 21, bg = "grey", col = "black")
+
+      text(-2/3 + 0.15, r1(-2/3) - 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 1)])))
+      text(-1/3, r1(-1/3) - 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 2)])))
+      text(1/3 - 0.15, r2(1/3) - 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 3)])))
+      text(1/2 - 0.15, r1(1/2) + 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 4)])))
+      text(2/3, r2(2/3) + 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 5)])))
+
+      par(family = "sans")
+      str <- "62-km\nbuffer"
+      rect(-0.4 - strwidth(str, cex = 1.75) / 2,
+           0.4 - strheight(str, cex = 1.75) / 2,
+           -0.4 + strwidth(str, cex = 1.75) / 2,
+           0.4 + strheight(str, cex = 1.75) / 2, 
+           col = rgb(1, 1, 1, alpha = 0.5), border = NA)
+      text(-0.4, 0.4, labels = str, cex = 1.5)
+
+      ## Panel 3: Climate-trait relationship
+
+		  par(mai = c(0.6, 0.6, 1/25, 1/25), family = "sans")
+		  plot(y, foc$Latitude, las = 1, axes = F, frame.plot = T, 
+				   xlim = c(-2.5, 2.5), ylim = c(32, 45),
+				   pch = 19, col = "grey", xlab = "", ylab = "")
+		  title(xlab = "Climate variable", ylab = "Trait", cex.lab = 1.5, 
+		        line = 1)
+		  lines(range(y), range(predict(fit)), lwd = 2)
+
+	dev.off()
+
+@
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of variable selection algorithm
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelectBase.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Climatic variable selection]{Overview of process for identifying putative climatic selective drivers underlying latitudinal cline.
+We looked for climate variables that explained both the latitude of \Sexpr{2 * nrow(thinnedOccClimate)} \textit{M.
+cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
+(A) For Climate-Latitude variables we extracted climate data from the specific location where a population was collected ($x_{i,0}$) throughout California and Oregon, USA (shown in map).
+For Climate-Trait variables, we extracted climatic data in two ways, but only for the 16 focal populations.
+(B) For point-estimated Climate-Trait variables, we extracted climate only from where a population was collected ($x_{i,0}$).
+(C) For spatially averaged Climate-Trait variables, we averaged over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M.
+cardinalis} is riparian.}
+	\label{fig:Fig_VarSelect}
+\end{figure}
 
@@
 	     font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
+
+  	# lines between panels
+  	# clip(grconvertX(0, from = "ndc", to = "user"),
+  	#     grconvertX(1, from = "ndc", to = "user"),
+  	#     grconvertY(0.05, from = "ndc", to = "user"),
+  	#     grconvertY(0.95, from = "ndc", to = "user"))
+	  # abline(v = grconvertX(1/3, from = "ndc", to = "user"))
 

### 2016-10-26 ms/ms.tex
New figure of climate variable procedure

@@
 We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km climatic neighborhood (`spatially averaged Climate-TraitPC1').
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of variable selection algorithm
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelectBase.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Climatic variable selection]{Overview of process for identifying putative climatic selective drivers underlying latitudinal cline.
+We looked for climate variables that explained both the latitude of 356 \textit{M.
+cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
+(A) For Climate-Latitude variables we extracted climate data from the specific location where a population was collected ($x_{i,0}$) throughout California and Oregon, USA (shown in map).
+For Climate-Trait variables, we extracted climatic data in two ways, but only for the 16 focal populations.
+(B) For point-estimated Climate-Trait variables, we extracted climate only from where a population was collected ($x_{i,0}$).
+(C) For spatially averaged Climate-Trait variables, we averaged over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M.
+cardinalis} is riparian.}
+	\label{fig:Fig_VarSelect}
+\end{figure}
 

### 2016-10-26 ms/Figures/CODE FOR GEO FIGURE EXPLANATION.R

@@ -5,129 +5,168 @@
 	library(rgeos)
 
 # Color scheme
-	
+
 	colNorth <- "deepskyblue"
 	colSouth <- rgb(255, 131, 0, maxColorValue = 255)
 
 	# Get maps of California and Oregon
-usa <- getData('GADM', country = "USA", level = 1, path = pathDat)
-states <- gSimplify(usa[usa@data$NAME_1 %in% c("California", "Oregon", "Nevada", "Idaho"), ], 
+  usa <- getData('GADM', country = "USA", level = 1, path = pathDat)
+  states <- gSimplify(usa[usa@data$NAME_1 %in% c("California", "Oregon", "Nevada", "Idaho"), ], 
 										tol = 0.01, topologyPreserve = TRUE)
 
-# Load thinned occurences used Climate-Lat analyses and convert to Spatial class
-occ <- read.csv(paste(pathDat, "/thinned1OccLatLon.csv", sep = ""), row.names = NULL)
-coordinates(occ) <- ~Longitude+Latitude
-projection(occ) <- CRS('+proj=longlat');
+  # Load thinned occurences used Climate-Lat analyses and convert to Spatial class
+  occ <- read.csv(paste(pathDat, "/thinned1OccLatLon.csv", sep = ""), row.names = NULL)
+  coordinates(occ) <- ~Longitude+Latitude
+  projection(occ) <- CRS('+proj=longlat');
 
-# Load focal populations and convert to Spatial class
-load(paste(pathDat, "/focClim.RData", sep = ""))
-foc <- focClim[, c("Longitude", "Latitude")]
-coordinates(foc) <- ~Longitude+Latitude
-projection(foc) <- CRS('+proj=longlat');
+  # Load focal populations and convert to Spatial class
+  load(paste(pathDat, "/focClim.RData", sep = ""))
+  foc <- focClim[, c("Longitude", "Latitude")]
+  coordinates(foc) <- ~Longitude+Latitude
+  projection(foc) <- CRS('+proj=longlat');
 
-# Clip and map
-clipExtent <- as(extent(floor(min(occ$Longitude) * 10) / 10, 
-												ceiling(max(occ$Longitude) * 10) / 10, 
[…]

### 2016-10-26 ms/Figures/CODE FOR GEO FIGURE EXPLANATION.R
Added funding info

@@ -1,233 +1,0 @@
-
-# Libraries
-
-	library(raster)
-	library(rgeos)
-
-# Color scheme
-
-	colNorth <- "deepskyblue"
-	colSouth <- rgb(255, 131, 0, maxColorValue = 255)
-
-	# Get maps of California and Oregon
-  usa <- getData('GADM', country = "USA", level = 1, path = pathDat)
-  states <- gSimplify(usa[usa@data$NAME_1 %in% c("California", "Oregon", "Nevada", "Idaho"), ], 
-										tol = 0.01, topologyPreserve = TRUE)
-
-  # Load thinned occurences used Climate-Lat analyses and convert to Spatial class
-  occ <- read.csv(paste(pathDat, "/thinned1OccLatLon.csv", sep = ""), row.names = NULL)
-  coordinates(occ) <- ~Longitude+Latitude
-  projection(occ) <- CRS('+proj=longlat');
-
-  # Load focal populations and convert to Spatial class
-  load(paste(pathDat, "/focClim.RData", sep = ""))
-  foc <- focClim[, c("Longitude", "Latitude")]
-  coordinates(foc) <- ~Longitude+Latitude
-  projection(foc) <- CRS('+proj=longlat');
-
-  # Clip and map
-  clipExtent <- as(extent(floor(min(occ$Longitude) * 10) / 10, 
-	   											ceiling(max(occ$Longitude) * 10) / 10, 
-	  											floor(min(occ$Latitude) * 10) / 10, 
-		  										ceiling(max(occ$Latitude) * 10) / 10), 
-			  					 "SpatialPolygons")
-  proj4string(clipExtent) <- CRS(proj4string(states))
-  states <- gIntersection(states, clipExtent, byid = TRUE)
-
-  # Transform to Albert's Equal Area Projection
-  projAEA <- sprintf("+proj=aea +lat_1=%s +lat_2=%s +lat_0=%s +lon_0=%s +x_0=0 +y_0=0",
-	  								 extent(clipExtent)[3], extent(clipExtent)[4], 
-		  							 extent(clipExtent)[1], extent(clipExtent)[2])
-  states <- spTransform(states, CRS(projAEA))
[…]

### 2016-11-17 ms/ms.Rnw
Added funding info
Deleted files and updated Readme file

@@
 %lost on page 11 - extracted and used climate variables
-%cartoon of whole procedure for climate variables
-% model of growth chamber
 % need to explain 'importance'
@@
 %* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
-%	- maybe temporal source-sink makes more sense
 
@@
 load(paste(pathDat, "/focClimSA_62km.RData", sep = ""))
+
+# How similar are focClim and focClimSA?
+# summary(mapply(cor, focClim[climVars], focClimSA[climVars]))
+# summary(mapply(cor, focClim[climVars], focClimSA[climVars]) ^ 2)
 
@@
 
-% *** Need to fill in exact grant numbers *** %
 \section*{Acknowledgements}
@@
 CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program.
-ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation.
+ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
 

### 2016-11-17 README.Rmd
Updated README

---
title: "Readme for Muir and Angert"
author: "Chris Muir and Amy Angert"
output: html_document
---

This repository contains the source for the paper Muir CD and Angert AL. Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in *Mimulus cardinalis*.

See the Sweave file 'ms/ms.Rnw' for code chunks. If you have R packages installed, you should be able to recreate the manuscript locally.

A preprint of this paper is available on biorxiv: http://dx.doi.org/10.1101/080952

### 2017-01-06 ms/ms.Rnw
Major revision

@@
-%lost on page 11 - extracted and used climate variables
+%%%%% ISSUES %%%%%%
 % need to explain 'importance'
 
-%* delve into interactions - do they change cline?
-
-%* show figures of reaction norms
-%* split plot, account for error variation in temp???
-%* show PAR similarity
-%* some sites they are functionally annual
-
-%* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
-
-%* put in biomass allocation cline
-% TO DO:
-% 2. Double check spatially averaged analysis
-% 3. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
-% 4. Do anything with biomass data? Currently, NO.
-
-% 5. Citations for latitudinal clines
-	% cases where clinal variation has 'obvious' selective source:
-	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
-	% Caicedo et al 2004?
-	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
-% 6. Citations for GxE as signature of local adaptation
+% delve into interactions - do they change cline?
+% show figures of reaction norms
+% show PAR similarity
+% hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
+% For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
 
@@
 %
-% Commenting
-%
-
-\newcommand{\cdm}[1]{{ \color{magenta} [{\bf{CDM:}} {\em#1}]}} % Chris' comments are in magenta.
-\newcommand{\ala}[1]{{ \color{blue} [{\bf{ALA:}} {\em#1}]}} % Amy's comments are in blue.
-
-%
 % Where to find files
@@
 \setstretch{\stretchy}
+\linenumbers
 
@@
 Phenotypic differences between populations along a cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
-Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
+Despite a long history of studying local adaptation and clines, it remains challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
 In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment.
@@
 We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively.
-There are classic cases of adaptation involving both intrinsic and plastic trait variation.
+Adaptation involves both intrinsic and plastic trait variation.
 For example, intrinsic differences in photoperiod \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
@@
 
-Despite the fact that latitudinal clines in particular have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
+Despite the fact that latitudinal clines have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
 Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days.
@@
 coastal vs.
-continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local versus regional climate shape selective pressures.
-
-% For the most part, we know much more about selective agents responsible for clines for conspicuous things like colouration, such as in mice \citep{Sumner_1932, Mullen_Hoekstra_2008}
-
-% Motivate primary questions using Mimulus
-% might add that Anacker and Strauss 2014 showed ecological shifts important in Mim speciation and we would like to know driving forces behind this???
+continental), dissecting the evolution of latitudinal clines across many species will help identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local versus regional climate shape selective pressures.
+
 In this study, we intestigated two major questions: 1) whether intrinsic or plastic physiological trait variation corresponds with latitude; and 2) what climatic factor(s) could plausibly be repsonsible for latitudinal clines.
@@
 We found a latitudinal cline of intrinsic variation in photosynthesis and growth, but little evidence for variation in plasticity.
-Interannual variation in precipitation is associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+Interannual variation in precipitation and temperature are associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
 cardinalis}.
-However, no single climatic factor explained trait variation better than latitude alone, indicating that additional climatic factors are likely important, albeit less so.
-The climatic neighborhoods around populations likewise did not explain trait variation better than latitude alone.
+The climatic neighborhoods around populations explained trait variation better than local climate, indicating that latitudinal clines may be common because latitude integrates effects of selection on populations connected through gene flow.
 We place these findings in the context of life history theory and consider future directions in the Discussion.
@@
 
-# Read in spatially-averaged (62 km buffer) for 16 focal populations
-load(paste(pathDat, "/focClimSA_62km.RData", sep = ""))
-
-# How similar are focClim and focClimSA?
+# How similar are focClim and focClimSA? ** NEEDS UPDATE WITH NEW DATASHEET **
 # summary(mapply(cor, focClim[climVars], focClimSA[climVars]))
@@
       \midrule
-	HAU & South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "HAU"], 3)}	& 
+	Hauser Creek & South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "HAU"], 3)}	& 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "HAU"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "HAU"]}   \\
-	CTC	& South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "CTC"], 3)} & 
+	Cottonwood Creek	& South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "CTC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "CTC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "CTC"]}   \\
-	CUR	& South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "CUR"], 3)} & 
+	Sweetwater River	& South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "CUR"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "CUR"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "CUR"]}   \\
-	GRP & South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "GRP"], 3)} &
+	Grade Road Palomar & South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "GRP"], 3)} &
     \Sexpr{round(focClim$Longitude[focClim$PopID == "GRP"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "GRP"]}   \\
-	WWC &	Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "WWC"], 3)} & 
+	Whitewater Canyon &	Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "WWC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WWC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WWC"]}   \\
-	MIL	& Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "MIL"], 3)} & 
+	Mill Creek	& Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "MIL"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "MIL"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "MIL"]}   \\
-	WFM	& Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "WFM"], 3)} & 
+	West Fork Mojave River	& Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "WFM"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WFM"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WFM"]}   \\
-	NMT	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "NMT"], 3)} & 
+	North Fork Middle Tule River	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "NMT"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "NMT"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "NMT"]}   \\
-	PRD	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "PRD"], 3)} & 
+	Paradise Creek	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "PRD"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "PRD"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "PRD"]}   \\
-	RWD	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "RWD"], 3)} & 
+	Redwood Creek	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "RWD"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RWD"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RWD"]}   \\
-	WNA	& Central Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "WNA"], 3)} & 
+	Wawona  & Central Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "WNA"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WNA"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WNA"]}   \\
-	RBW	& Central Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RBW"], 3)} & 
+	Rainbow Creek	& Central Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RBW"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RBW"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RBW"]}   \\
-	MYU	& North Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MYU"], 3)} & 
+	Middle Yuba River	& North Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MYU"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "MYU"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "MYU"]}   \\
-	LIJ	& North Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "LIJ"], 3)} & 
+	Little Jamison Creek	& North Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "LIJ"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "LIJ"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "LIJ"]}   \\
-	DPC	& North Coast & \Sexpr{round(focClim$Latitude[focClim$PopID == "DPC"], 3)} & 
+	Deep Creek	& North Coast & \Sexpr{round(focClim$Latitude[focClim$PopID == "DPC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "DPC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "DPC"]}   \\
-	RCC	& North Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "RCC"], 3)} & 
+	Rock Creek	& North Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "RCC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RCC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RCC"]}   \\
@@
 
-pdf(paste(pathFig, "/ExperimentalDesignBase.pdf", sep = ""), 7, 5)
+pdf(paste(pathFig, "/Figure_ExptlDes.pdf", sep = ""), 9, 5)
+layout(matrix(c(rep(1, 7), rep(2, 2)), nrow = 1))
 par(mai = rep(1, 4))
 plot(0, 0, type = "n", xlim = c(0, max(exptlDes$DayN)), ylim = c(-1, 41), 
-		 axes = F, frame.plot = T, cex.lab = 1.5, 
-		 xlab = "Key treatments and measurements", 
-     ylab = expression(paste("Temperature (", phantom()*degree, "C)")))
+		 axes = F, frame.plot = T, xlab = "", ylab = "")
+mtext("Key treatments and measurements", 1, cex = 1.5, line = 5, las = 1)
 
 # Temperature axis
-axis(2, at = seq(0, 40, 10), lwd = 0, lwd.ticks = 1, las = 1,
+axis(2, at = seq(0, 40, 10), lwd = 0, lwd.ticks = 1, las = 1, cex.axis = 1.5,
 	labels = c(expression(0*degree), expression(10*degree), expression(20*degree),
 	           expression(30*degree), expression(40*degree)))
+mtext(expression(paste("Temperature (", phantom()*degree, "C)")), 2, 
+	cex = 1.5, line = 4, las = 3)
 
 # Watering axis
-axis(4, at = c(0, 40), lwd = 0, lwd.ticks = 1, las = 1, labels = c(1, 12))
+axis(4, at = c(0, 40), lwd = 0, lwd.ticks = 1, las = 1, labels = c(1, 12), cex.axis = 1.5)
 mtext(expression(paste("Watering Freq (no. Day", phantom()^-1, ")")), 4, 
-	cex = 1.5, line = 3, las = 3)
+	cex = 1.5, line = 5, las = 3)
 
 # Axis of time on top
-mtext("Days after sowing", side = 3, line = 3.5, cex = 1.5)
-mtext("(Date for climate simulation)", side = 3, line = 2.25, cex = 1.5)
-axis(3, at = seq(0, 90, 30), lwd = 0, lwd.ticks = 1)
-axis(3, at = c(17, 47, 78), lwd = 0, lwd.ticks = 1, line = -0.5, 
+mtext("Days after sowing", side = 3, line = 5, cex = 1.5)
+mtext("(Date for climate simulation)", side = 3, line = 3, cex = 1.5)
+axis(3, at = seq(0, 90, 30), lwd = 0, lwd.ticks = 1, cex.axis = 1.5)
+axis(3, at = c(17, 47, 78), lwd = 0, lwd.ticks = 1, line = -0.5, cex.axis = 1.5, 
      labels = c("(June)", "(July)", "(August)"))
@@
 abline(v = 0, col = "grey", lwd = 2, lty = 2)
-mtext("Sow", side = 1, at = 0, line = 0)
+mtext("Sow", side = 1, at = 0, line = 1)
 abline(v = 17, col = "grey", lwd = 2, lty = 2)
-mtext("Temp\nTreatment", side = 1, at = 17, line = 1)
+mtext("Temp\nTreatment", side = 1, at = 17, line = 2)
 abline(v = 37, col = "grey", lwd = 2, lty = 2)
-mtext("Drought\nTreatment", side = 1, at = 37, line = 1)
+mtext("Drought\nTreatment", side = 1, at = 37, line = 2)
 
@@
 
-with(exptlDes, points(DayN, WaterFreq_dry, type = "l", lwd = 3, col = "black"))
-
-dev.off()
+with(exptlDes, points(DayN, WaterFreq_dry, type = "l", lwd = 3, col = "black",
+                      lty = 2))
+lines(range(exptlDes$DayN), rep(40, 2), lwd = 3)
 
 # Legend
-pdf(paste(pathFig, "/ExperimentalDesignLegend.pdf", sep = ""), 2, 2)
-par(mai = rep(0, 4))
-plot(0, 0, type = "n", xlim = c(-0.5, 1.5), ylim = c(0, 1), axes = F)
+# pdf(paste(pathFig, "/ExperimentalDesignLegend.pdf", sep = ""), 2, 2)
+par(mai = c(2, 0, 1, 0))
+plot(0, 0, type = "n", xlim = c(-0.5, 1.5), ylim = c(0, 1), axes = F, xlab = "")
+
+mtext("Legend for\ntreatments", 3, line = 1)
+axis(1, at = c(0, 0.5, 1), labels = c("n", "n + 1", "n + 2"),
+     font = 3)
+mtext("Key to\nabbreviations\n\nGerm = Germination\nLER = Leaf expansion rate\nSER = Stem elongation rate\nPhoto = Photosynthesis", 1, line = 12, adj = 0, at = -0.75)
 
@@
 # Drought
-lines(c(0, 1), c(1, 0), lwd = 2)
-text(-0.25, 0.95, labels = "Water\nFreq")
-text(0.5, 0.9, labels = "Hot", col = primShade3)
-text(0.5, 0.1, labels = "Cool", col = sec1shade3)
-text(1.25, 1, labels = "Max", col = alpha(primShade0, 0.25))
-text(1.25, 0.55, labels = "Min", col = alpha(primShade0, 0.25))
-text(1.25, 0.775, labels = "Mean", col = primShade3)
-
-text(1.25, 0, labels = "Min", col = alpha(sec1shade0, 0.25))
-text(1.25, 0.45, labels = "Max", col = alpha(sec1shade0, 0.25))
-text(1.25, 0.225, labels = "Mean", col = sec1shade3)
+lines(c(0, 1), c(1, 1), lwd = 2, lty = 1)
+lines(c(0, 1), c(1, 0), lwd = 2, lty = 2)
+text(-0.25, 0.95, labels = "Water\nFreq", cex = 1.5)
+text(0.5, 0.9, labels = "Hot", col = primShade3, cex = 1.5)
+text(0.5, 0.1, labels = "Cool", col = sec1shade3, cex = 1.5)
+text(1.25, 1, labels = "Max", col = alpha(primShade0, 0.25), cex = 1.5)
+text(1.25, 0.55, labels = "Min", col = alpha(primShade0, 0.25), cex = 1.5)
+text(1.25, 0.775, labels = "Mean", col = primShade3, cex = 1.5)
+
+text(1.25, 0, labels = "Min", col = alpha(sec1shade0, 0.25), cex = 1.5)
+text(1.25, 0.45, labels = "Max", col = alpha(sec1shade0, 0.25), cex = 1.5)
+text(1.25, 0.225, labels = "Mean", col = sec1shade3, cex = 1.5)
 
@@
 Within each day, there was a maximum daytime temperture (top of translucent polygons) and minimum nighttime temperature (bottom of translucent polygons).
-The drought treatment commenced later by ramping down the frequency of bottom-watering episodes (black line; right axis).
+The drought treatment commenced later by ramping down the frequency of bottom-watering episodes (dashed black line; right axis), while watering frequency was maintained in the control treatment (solid black line).
 Grey boxes on the bottom of the plot outline the period of key measurements described in the Material and Methods.}
@@
 
-\subsection*{Growth and photosynthesis}
+\subsection*{Trait measurements}
+
+We measured five traits in response to temperature and watering treatments (Table~\ref{table:Table_traits}).
 
@@
 
-We measured growth rate during two phases: leaf expansion as a rosette and stem elongation after bolting.
+We measured growth rate during two phases: leaf expansion and stem elongation.
 We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
@@
 For each single-trait model above, we extracted the Population coefficient (factoring out Treatment and other effects).
-The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA).
+The multivariate distribution of these coefficients was then summarized using principal components analysis.
 The first principal component of these traits (TraitPC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.
@@
   focClim$PC1 <- traitPC$x[, "PC1"]
-  focClimSA$PC1 <- traitPC$x[, "PC1"]
 
@@
   title(xlab = "Latitude", cex.lab = 1.5)
+  title(ylab = "TraitPC1", cex.lab = 2, line = 5)
   title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
@@
 In addition, to help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{M.
-cardinalis} occurrence records, not just the 16 focial populations.
+cardinalis} occurrence records (see detail below), not just the 16 focial populations.
 We refer to these climatic factors as `Climate-Latitude' variables.
@@
 cardinalis} range.
-We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}.
+Therefore, we did not consider Climate-TraitPC1 variables to be candidates selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
+We selected Climate-Latitude and Climate-TraitPC1 variables independently using Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}.
 VSURF ranks variables by their importance over regression trees in the forest.
@@
 
-To test the third hypothesis about climatic neighborhoods driving selection, we repeated the same procedure to identify Climate-TraitPC1 variables as above, except that we used spatially averaged climate variables.
-We sampled climate at 1000 random points (at 90-m resolution) within a 62-km buffer around the 16 focal populations.
+To test the third hypothesis about climatic neighborhoods driving selection, we directly competed local with neighborhood climate to see which correlated with traits (Climate-TraitPC1 variables) and latitude (Climate-Latitude variables) better.
+We used the immediate collection location for local climate.
+For climate neighborhoods, we sampled climate at 1000 random points (at 90-m resolution) within a 62-km radius buffer around the collection collection.
 We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
@@
 cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
-We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km climatic neighborhood (`spatially averaged Climate-TraitPC1').
+Climatic means and variances (see below) were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
 
@@
   # Load thinned occurences used in Climate-Lat analyses and convert to Spatial class
-  occ <- read.csv(paste(pathDat, "/thinned1OccLatLon.csv", sep = ""), 
+  occ <- read.csv(paste(pathDat, "/thinnedOccLatLon.csv", sep = ""), 
                   row.names = NULL)
@@
   # PLOT
-  pdf(paste(pathFig, "/Figure_VarSelectBase.pdf", sep = ""), w = 6, h = 7, 
-	  	useDingbat = F)
+  pdf(paste(pathFig, "/Figure_VarSelect.pdf", sep = ""), w = 8, h = 7)#, 
+	  	#useDingbat = F)
   loadfonts()
-  mat <- matrix(c(1, 4, 7,
-	  							1, 4, 7,
-		  						1, 4, 7,
-			  					2, 5, 8,
-				  				2, 5, 8,
-					  			3, 6, 9,
-						  		3, 6, 9), ncol = 3, byrow = T)
+  mat <- matrix(c(1, 4, 6, 9,
+	  							1, 4, 6, 9,
+		  						1, 4, 6, 9,
+			  					2, 5, 7, 10,
+				  				2, 5, 7, 10,
+					  			3, 3, 8, 8,
+						  		3, 3, 8, 8), ncol = 4, byrow = T)
   layout(mat)
 
-  # A. Example Climate-Latitude variable
+  # A. Climate-Latitude variables
 
@@
 	  par(mai = rep(0.5, 4), family = "sans")
-	  plot(states, main = "A. Climate-Latitude variables")
+	  plot(states)
+	  mtext(text = "Climate-Latitude variables", side = 3, at = grconvertX(2, "in"),
+	        font = 2, line = 2)
+	  mtext(text = "local", side = 3)
     points(occAEA, pch = 19, cex = 0.5)
@@
 	       cex = 1.5)
-	  # lines(c(grconvertX(xCUR, from = "device", to = "user"), 0),
-	  #     c(grconvertY(yCUR, from = "device", to = "user"), 0),
-	  #       lwd = 2)
     text(0, 1, labels = expression(Climate[italic(i)] == 
@@
 			   pch = 19, col = "grey", xlab = "", ylab = "Latitude of origin")
-	  title(xlab = "Climate variable", cex.lab = 1.5, line = 1)
+	  title(xlab = "Climate variable\n(local or neighborhood)", cex.lab = 1.5, line = 2)
 
@@
 
+	 	## Panel 4: Map
+	  par(mai = rep(0.5, 4), family = "sans")
+	  plot(states)
+  	mtext(text = "neighborhood", side = 3)
+    points(occAEA, pch = 21, cex = 2, col = "black", bg = "white")
+	  points(occAEA, pch = 19, cex = 0.5)
+
+	  ## Panel 5: Cartoon of neighborhood climate calculation
+	  par(mai = rep(0, 4), family = "LM Roman 10")
+
+  		### Climate grid
+
+  			#### Functions to find closest point along "stream"
+				fn <- function(par, x, y) {
+					if (x < 0) {
+						d1 <- sqrt((par - x) ^ 2 + (r1(par) - y) ^ 2)
+						return(d1)
+					} else {
+						d1 <- sqrt((par - x) ^ 2 + (r1(par) - y) ^ 2)
+						d2 <- sqrt((par - x) ^ 2 + (r2(par) - y) ^ 2)
+						return(min(c(d1, d2)))
+					}
+				}
+
+				getStreamDist <- function(x, y) {
+					fit <- optim(x, fn, x = x, y = y, method = "Brent", lower = -2, 
+											 upper = 2)
+					retval <- setNames(c(fit$par, fit$value), c("s", "dist"))
+					return(retval)
+				}
+
+			  #### "Climate" is fxn of distance from stream and pos along x-axis
+			  xy <- expand.grid(x = seq(-0.95, 0.95, 0.1), 
+			                    y = seq(-0.95, 0.95, 0.1))
+			  xyz <- cbind(xy, 
+			               t(apply(xy, 1, function(X) getStreamDist(X[1], X[2]))))
+			  xyz$z <- xyz$dist + xyz$x * 0.3
+			  xyz$z <- xyz$z - min(xyz$z)
+			  xyz$z <- xyz$z / max(xyz$z)
+
+		  plot(0, 0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1), axes = F)
+		  clip(-1, 1, -1, 1)
+		  points(x1, y1, type = "l", col = "blue", lwd = 5)
+		  points(x2, y2, type = "l", col = "blue", lwd = 5)
+
+		  apply(xyz, 1, function(X) rect(X["x"] - 0.05, X["y"] - 0.05, 
+																	   X["x"] + 0.05, X["y"] + 0.05,
+														 			   col = rgb(0, 0, 1, alpha = X["z"]), 
+																	   border = NA))
+
+			# Draw buffer area and circle
+			x <- seq(-1, 0, length.out = 1e3)
+			clip(-2, 2, -2, 2)
+			polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
+			        col = "white", border = "white", lwd = 2)
+			polygon(c(x + 1, rev(x + 1)), c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
+			        col = "white", border = "white", lwd = 2)
+			polygon(c(x, rev(x)), -c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
+			        col = "white", border = "white", lwd = 2)
+			polygon(c(x + 1, rev(x + 1)), -c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
+			        col = "white", border = "white", lwd = 2)
+
+			x <- seq(-1, 1, length.out = 1e3)
+			polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), -sqrt(1 - x ^ 2)),
+			        border = "black", lwd = 2)
+
+			# Draw points and label
+	    points(0, 0, pch = 19, cex = 3)
+	    text(0, 0, labels = expression(italic(x[paste(i, ",", 0)])), adj = 1.5, 
+	         cex = 1.5)
+      text(0, 1, labels = expression(Climate[italic(i)] == 
+           sum(over(italic(x)[paste(italic(i), ",", italic(j))], italic(n)), 
+               scriptstyle(italic(j) == 1), italic(n))), 
+           cex = 1.5, pos = 3, offset = 1)
+
+      points(c(-2/3, -1/3, 1/2, 1/3, 2/3),
+             c(r1(-2/3), r1(-1/3), r1(1/2), r2(1/3), r2(2/3)),
+             cex = 2, pch = 21, bg = "grey", col = "black")
+
+      text(-2/3 + 0.15, r1(-2/3) - 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 1)])))
+      text(-1/3, r1(-1/3) - 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 2)])))
+      text(1/3 - 0.15, r2(1/3) - 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 3)])))
+      text(1/2 - 0.15, r1(1/2) + 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 4)])))
+      text(2/3, r2(2/3) + 0.15, cex = 1.5,
+           labels = expression(italic(x[paste(i, ",", 5)])))
+
+      par(family = "sans")
+      str <- "62-km\nbuffer"
+      rect(-0.4 - strwidth(str, cex = 1.75) / 2,
+           0.4 - strheight(str, cex = 1.75) / 2,
+           -0.4 + strwidth(str, cex = 1.75) / 2,
+           0.4 + strheight(str, cex = 1.75) / 2, 
+           col = rgb(1, 1, 1, alpha = 0.5), border = NA)
+      text(-0.4, 0.4, labels = str, cex = 1.5)
+
    # B. Example Climate-Trait variable, point estimated
 
-  	## Panel 1: Map
+  	## Panel 6: Map
 	  par(mai = rep(0.5, 4), family = "sans")
-	  plot(states, main = "B. Climate-Trait variables")
-	  mtext(text = "point estimated", side = 3)
+	  plot(states)
+	  mtext(text = "Climate-Trait variables", side = 3, at = grconvertX(6, "in"),
+	        font = 2, line = 2)
+    mtext(text = "local", side = 3)
 	  points(focAEA, pch = 19, cex = 0.5)
 
-	  # Position of Cuyamaca Rancho population in device units
-	  xCUR <- grconvertX(focAEA$Longitude["CUR"], from = "user", to = "device")
-	  yCUR <- grconvertY(focAEA$Latitude["CUR"], from = "user", to = "device")
-
-	  ## Panel 2: Cartoon of point versus spatial average climate estimate
+	  ## Panel 7: Cartoon of local climate
 	  par(mai = rep(0, 4), family = "LM Roman 10")
@@
 	       cex = 1.5)
-	  # lines(c(grconvertX(xCUR, from = "device", to = "user"), 0),
-	  #       c(grconvertY(yCUR, from = "device", to = "user"), 0),
-	  #       lwd = 2)
     text(0, 1, labels = expression(Climate[italic(i)] == 
@@
 
-	  ## Panel 3: Climate-trait relationship
+	  ## Panel 8: Climate-trait relationship
 	  par(mai = c(0.6, 0.6, 1/25, 1/25), family = "sans")
@@
 			   pch = 19, col = "grey", xlab = "", ylab = "")
-	  title(xlab = "Climate variable", ylab = "Trait", cex.lab = 1.5, line = 1)
+	  title(ylab = "Trait", cex.lab = 1.5, line = 1)
+	  title(xlab = "Climate variable\n(local or neighborhood)", cex.lab = 1.5, line = 2)
     fit <- lm(foc$Latitude ~ y)
@@
 
-  # C. Example Climate-Trait variable, spatially averaged
-
-  	## Panel 1: Map
+  	## Panel 9: Map
   	par(mai = rep(0.5, 4), family = "sans")
-  	plot(states, main = "C. Climate-Trait variables")
-  	tmp <- raster(states)
-  	mtext(text = "spatially averaged", side = 3)
+  	plot(states)
+  	mtext(text = "neighborhood", side = 3)
   	points(focAEA, pch = 21, cex = 2, col = "black", bg = "white")
@@
 
-	  # Position of Cuyamaca Rancho population in device units
-	  xCUR <- grconvertX(focAEA$Longitude["CUR"], from = "user", to = "inches")
-	  yCUR <- grconvertY(focAEA$Latitude["CUR"], from = "user", to = "inches")
-
-  	## Panel 2: Cartoon of point versus spatial average climate estimate
+  	## Panel 10: Cartoon of point versus spatial average climate estimate
 	  par(mai = rep(0, 4), family = "LM Roman 10")
@@
   		### Climate grid
-
-  			#### Functions to find closest point along "stream"
-				fn <- function(par, x, y) {
-					if (x < 0) {
-						d1 <- sqrt((par - x) ^ 2 + (r1(par) - y) ^ 2)
-						return(d1)
-					} else {
-						d1 <- sqrt((par - x) ^ 2 + (r1(par) - y) ^ 2)
-						d2 <- sqrt((par - x) ^ 2 + (r2(par) - y) ^ 2)
-						return(min(c(d1, d2)))
-					}
-				}
-
-				getStreamDist <- function(x, y) {
-					fit <- optim(x, fn, x = x, y = y, method = "Brent", lower = -2, 
-											 upper = 2)
-					retval <- setNames(c(fit$par, fit$value), c("s", "dist"))
-					return(retval)
-				}
 
@@
 	         cex = 1.5)
-	    # Start line from outside of point
-	    # point radius = cex * size of point = cex * 0.75 * character height
-	    # ptRad <- 0.1 * 2 * 0.75 * par()$cin[2]
-	    # lines(c(grconvertX(xCUR - ptRad, from = "inches", to = "user"), -1),
-	    #     c(grconvertY(yCUR, from = "inches", to = "user"), 0),
-	    #       lwd = 2)
+
       text(0, 1, labels = expression(Climate[italic(i)] == 
-           sum(over(x[paste(italic(i), ",", italic(j))], italic(n)), 
+           sum(over(italic(x)[paste(italic(i), ",", italic(j))], italic(n)), 
                scriptstyle(italic(j) == 1), italic(n))), 
@@
 
-      ## Panel 3: Climate-trait relationship
-
-		  par(mai = c(0.6, 0.6, 1/25, 1/25), family = "sans")
-		  plot(y, foc$Latitude, las = 1, axes = F, frame.plot = T, 
-				   xlim = c(-2.5, 2.5), ylim = c(32, 45),
-				   pch = 19, col = "grey", xlab = "", ylab = "")
-		  title(xlab = "Climate variable", ylab = "Trait", cex.lab = 1.5, 
-		        line = 1)
-		  lines(range(y), range(predict(fit)), lwd = 2)
-
 	dev.off()
@@
 \begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelectBase.pdf}}
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelect.pdf}}
 	\usefont{T1}{phv}{m}{n}
@@
 	\selectfont
-	\caption[Climatic variable selection]{Overview of process for identifying putative climatic selective drivers underlying latitudinal cline.
+	\caption[Climatic variable selection]{Overview of method for identifying putative climatic selective agents underlying latitudinal cline.
 We looked for climate variables that explained both the latitude of \Sexpr{2 * nrow(thinnedOccClimate)} \textit{M.
 cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
-(A) For Climate-Latitude variables we extracted climate data from the specific location where a population was collected ($x_{i,0}$) throughout California and Oregon, USA (shown in map).
-For Climate-Trait variables, we extracted climatic data in two ways, but only for the 16 focal populations.
-(B) For point-estimated Climate-Trait variables, we extracted climate only from where a population was collected ($x_{i,0}$).
-(C) For spatially averaged Climate-Trait variables, we averaged over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M.
-cardinalis} is riparian.}
+For Climate-Latitude variables we extracted climate data from recent occurences located throughout California and Oregon, USA (shown in map).
+For Climate-Trait variables, we extracted climatic data for the 16 focal populations.
+For both analyses, we extracted local and neighborhood climate.
+Local climate refers to climate only from where a population was collected ($x_{i,0}$).
+Neighborhood climate was calculated as the average over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M.
+cardinalis} is riparian.
+We identified climatic factors that most strongly predicted latitude of occurences (Climate-Latitude variables) and traits (Climate-Trait variables), as shown for hypothetical data in plots at the bottom of the figure.}
 	\label{fig:Fig_VarSelect}
@@
   # Note that occurences have been randomly thinned to account for variation in sampling effort
-  load(paste(pathDat, "/clim2lat_fit_t1.RData", sep = "")) # fit_t1
-  vsurfLat <- fit_t1
+  vsurfLat <- get(load(paste(pathDat, "/clim2lat_fit_t1.RData", sep = "")))
 
@@
   set.seed(69648)
-  vsurfPC1 <- VSURF(focClim[, climVars], focClim[, "PC1"], parallel = TRUE,
-                    ncores = 3, ntree = 1e4)
-  climVars[vsurfPC1$varselect.pred]
-  vsurfPC1_SA <- VSURF(focClimSA[, climVars], focClimSA[, "PC1"], 
-                       parallel = TRUE, ncores = 3, ntree = 1e4)
-  climVars[vsurfPC1_SA$varselect.pred]
+  vsurfPC1 <- VSURF(focClim[, climVars], focClim[, "PC1"], parallel = TRUE, ncores = 3, 
+                    ntree = 1e4)
 
@@
   nImpPC1 <- length(imp2)
-
-  imp3 <- climVars[vsurfPC1_SA$varselect.pred]
-  nImpPC1_SA <- length(imp3)
 
@@
 
-  # Figure of variable importance for latitude
+#####	Figure_ClimVarImp.pdf ####
 
@@
 sw2a <- strwidth(s = clim2sym[imp1, "tempScale"], units = "inches", font = 4)
-	sw3a <- strwidth(s = parse(text = clim2sym[imp1, "what"]), units = "inches",
-	                 font = 4)
 
@@
 sw2b <- strwidth(s = clim2sym[imp2, "tempScale"], units = "inches", font = 4)
-	sw3b <- strwidth(s = parse(text = clim2sym[imp2, "what"]), units = "inches",
-	                 font = 4)
-
+
+# Figure size in inches
 panelWidth <- 3
@@
 panelGap <- 0.1
-	yLabWidth <- 4 * panelGap + sum(max(c(sw1a, sw1b)), max(c(sw2a, sw2b)),
-	                                max(c(sw3a, sw3b)))
+yLabWidth <- 4 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)) +
+  (panelHeight / max(nImpLat + 1, nImpPC1 + 1) * 5)
 xLabHeight <- 1
+heightA <- (panelHeight + 2 * xLabHeight)
+heightB <- (panelHeight * (nImpPC1) / (nImpLat) + xLabHeight)
 
 pdf(paste(pathFig, "/Figure_ClimVarImp.pdf", sep = ""), 
-      width = 2 * (yLabWidth + panelWidth) + panelGap, 
-      height = panelHeight + xLabHeight + panelGap)
-  par(mai = c(xLabHeight, yLabWidth, panelGap, panelGap / 2), mfrow = c(1, 2))
+       width = yLabWidth + panelWidth + 2 * panelGap, 
+       height = heightA + heightB)
+par(mai = c(xLabHeight, yLabWidth, xLabHeight, 2 * panelGap), ann = F, yaxs = "i")
+layout(matrix(1:2, ncol = 1), widths = rep(1, 2), heights = lcm(2.54 * c(heightA, heightB)))
 
@@
 plot(0, 0, type = "n", xlim = c(0, ceiling(vsurfLat$imp.mean.dec)[1]), 
-		ylim = c(1, length(vsurfLat$varselect.pred)), axes = F, frame.plot = T, 
-		xlab = "", ylab = "")
-  abline(h = 1:nImpLat, col = "grey")
-  points(vsurfLat$imp.mean.dec[1:length(vsurfLat$varselect.pred)], 
-    length(vsurfLat$varselect.pred):1, pch = 21, col = "black", bg = "white",
-    cex = 1.5)
-  text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
-       labels = "A", font = 2, cex = 2)
+     ylim = c(0.5, nImpLat + 0.5), axes = F, frame.plot = T)
+abline(h = nImpLat:1, col = "grey")
+points(vsurfLat$imp.mean.dec[1:length(vsurfLat$varselect.pred)], nImpLat:1, 
+       pch = 21, col = "black", bg = "white", cex = 1.5)
+text(grconvertX(0, "npc", "user") + strwidth("A", "user", cex = 2, font = 2), 
+     grconvertY(1, "npc", "user"), labels = "A", font = 2, cex = 2, pos = 1)
 
@@
 axis(1, lwd = 0, lwd.ticks = 1)
-  title(xlab = "Climate-Latitude\nImportance", line = 4, cex.lab = 1.5)
+title(xlab = "Climate-Latitude Importance", line = 3, cex.lab = 1.5)
 
@@
 
-	text(grconvertX(panelGap / 2, from = "in", to = "user"), 
-	     nImpLat:1, pos = 4, labels = clim2sym[imp1, "climFactor"],
+# Climatic factor
+text(grconvertX(panelGap, from = "in", to = "user"), nImpLat:1, 
+     pos = 4, labels = clim2sym[imp1, "climFactor"],
      font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))
-  text(grconvertX(panelGap + max(c(sw1a, sw1b)), from = "in", to = "user"), 
-       nImpLat:1, pos = 4, labels = clim2sym[imp1, "tempScale"],
+text(grconvertX(panelGap + max(c(sw1a, sw1b)) / 2, from = "in", to = "user"), 
+     grconvertY(1, "ndc", "user"), pos = 1, labels = "Climatic\nfactor")
+
+# Temporal scale
+text(grconvertX(2 * panelGap + max(c(sw1a, sw1b)), from = "in", to = "user"), nImpLat:1,
+     pos = 4, labels = clim2sym[imp1, "tempScale"],
      font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))
-	text(grconvertX(3 / 2 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), 
-	                               from = "in", to = "user"), nImpLat:1, pos = 4, 
-	     labels = parse(text = clim2sym[imp1, "what"]),
-	     font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))
-
-	# Panel B: Climate-PC1
-	par(mai = c(xLabHeight, panelGap / 2, panelGap, yLabWidth))
-	plot(0, 0, type = "n", xlim = c(0, 0.2), ylim = c(1, nImpLat), 
-		axes = F, frame.plot = T, xlab = "", ylab = "")
-	abline(h = nImpLat:(nImpLat - nImpPC1 + 1), col = "grey") 
-	points(vsurfPC1$imp.mean.dec[1:nImpPC1], nImpLat:(nImpLat - nImpPC1 + 1), 
+text(grconvertX(2 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)) / 2, from = "in", to = "user"), 
+     grconvertY(1, "ndc", "user"), pos = 1, labels = "Temporal\nScale")
+
+# Climatic average or variation?
+bx <- panelHeight / max(nImpLat + 1, nImpPC1 + 1)
+xs <- 3 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b))
+for (i in nImpLat:0) {
+  lines(x = c(grconvertX(xs, "in", "user"), grconvertX(xs + 2 * bx, "in", "user")), 
+        rep(i + 0.5, 2), col = "grey")
+}
+
+for (i in c(grconvertX(xs, "in", "user"), 
+            grconvertX(xs + bx, "in", "user"),
+            grconvertX(xs + 2 * bx, "in", "user"))) {
+  lines(rep(i, 2), c(max(nImpLat, nImpPC1) - nImpLat + 0.5, max(nImpLat, nImpPC1) + 0.5),
+        col = "grey")
+}
+
+x <- ifelse(grepl("_avg", imp1), xs + bx / 2, xs + 3 * bx / 2)
+points(grconvertX(x, "in", "user"), nImpLat:1, pch = 19)
+
+text(grconvertX(xs, from = "in", to = "user"), 
+     max(nImpLat, nImpPC1) + 1, pos = 4, labels = "Avg.", srt= 45)
+text(grconvertX(xs + bx, from = "in", to = "user"), 
+     max(nImpLat, nImpPC1) + 1, pos = 4, labels = "Var.", srt= 45)
+
+text(grconvertX(xs + bx, from = "in", to = "user"), 
+     grconvertY(1, "ndc", "user"), pos = 1, labels = "Type")
+
+# Spatial scale
+xs <- xs + 3 * bx
+for (i in nImpLat:0) {
+  lines(x = c(grconvertX(xs, "in", "user"), grconvertX(xs + 2 * bx, "in", "user")), 
+        rep(i + 0.5, 2), col = "grey")
+}
+
+x <- ifelse(!grepl("_SA", imp1), xs + bx / 2, xs + 3 * bx / 2)
+points(grconvertX(x, "in", "user"), nImpLat:1, pch = 19)
+
+for (i in c(grconvertX(xs, "in", "user"), 
+            grconvertX(xs + bx, "in", "user"),
+            grconvertX(xs + 2 * bx, "in", "user"))) {
+  lines(rep(i, 2), c(max(nImpLat, nImpPC1) - nImpLat + 0.5, max(nImpLat, nImpPC1) + 0.5),
+        col = "grey")
+}
+
+text(grconvertX(xs, from = "in", to = "user"), 
+     max(nImpLat, nImpPC1) + 1, pos = 4, labels = "Local", srt= 45)
+text(grconvertX(xs + bx, from = "in", to = "user"), 
+     max(nImpLat, nImpPC1) + 1, pos = 4, labels = "Neighborhood", srt= 45)
+
+text(grconvertX(xs + bx, from = "in", to = "user"), 
+     grconvertY(1, "ndc", "user"), pos = 1, labels = "Spatial\nscale")
+
+# Panel B: Climate-Trait
+par(mai = c(xLabHeight, yLabWidth, 0, 2 * panelGap))
+plot(0, 0, type = "n", xlim = c(0, 0.1 * ceiling(10 * vsurfPC1$imp.mean.dec)[1]), 
+     ylim = c(0.5, nImpPC1 + 0.5), axes = F, frame.plot = T)
+abline(h = nImpPC1:1, col = "grey")
+points(vsurfPC1$imp.mean.dec[1:length(vsurfPC1$varselect.pred)], nImpPC1:1, 
        pch = 21, col = "black", bg = "white", cex = 1.5)
-  text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
-       labels = "B", font = 2, cex = 2)
+text(grconvertX(0, "npc", "user") + strwidth("B", "user", cex = 2, font = 2), 
+     grconvertY(1, "npc", "user"), labels = "B", font = 2, cex = 2, pos = 1)
 
 # X axis & labels
-	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.2, 0.05))
-	title(xlab = "Climate-Trait PC1\nImportance", line = 4, cex.lab = 1.5)
+axis(1, lwd = 0, lwd.ticks = 1)
+title(xlab = "Climate-TraitPC1 Importance", line = 3, cex.lab = 1.5)
 
@@
 
-	text(grconvertX(yLabWidth + 2 * panelWidth + 3 / 2 * panelGap, 
-	                from = "in", to = "user"), 
-	     nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
-	     labels = clim2sym[imp2, "climFactor"],
+# Climatic factor
+text(grconvertX(panelGap, from = "in", to = "user"), nImpPC1:1, 
+     pos = 4, labels = clim2sym[imp2, "climFactor"],
      font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
-  text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
-                    max(c(sw1a, sw1b)), from = "in", to = "user"), 
-       nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
-       labels = clim2sym[imp2, "tempScale"],
+
+# Temporal scale
+text(grconvertX(2 * panelGap + max(c(sw1a, sw1b)), from = "in", to = "user"), nImpPC1:1,
+     pos = 4, labels = clim2sym[imp2, "tempScale"],
      font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
-	text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
-                    max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), from = "in", 
-	                to = "user"), nImpLat:(nImpLat - nImpPC1 + 1), 
-	     pos = 4, labels = parse(text = clim2sym[imp2, "what"]),
-	     font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
-
-  	# lines between panels
-  	# clip(grconvertX(0, from = "ndc", to = "user"),
-  	#     grconvertX(1, from = "ndc", to = "user"),
-  	#     grconvertY(0.05, from = "ndc", to = "user"),
-  	#     grconvertY(0.95, from = "ndc", to = "user"))
-	  # abline(v = grconvertX(1/3, from = "ndc", to = "user"))
+
+# Climatic average or variation?
+xs <- 3 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b))
+for (i in nImpPC1:0) {
+  lines(x = c(grconvertX(xs, "in", "user"), grconvertX(xs + 2 * bx, "in", "user")), 
+        rep(i + 0.5, 2), col = "grey")
+}
+
+for (i in c(grconvertX(xs, "in", "user"), 
+            grconvertX(xs + bx, "in", "user"),
+            grconvertX(xs + 2 * bx, "in", "user"))) {
+  lines(rep(i, 2), c(0.5, nImpPC1 + 0.5), col = "grey")
+}
+
+x <- ifelse(grepl("_avg", imp2), xs + bx / 2, xs + 3 * bx / 2)
+points(grconvertX(x, "in", "user"), nImpPC1:1, pch = 19)
+
+# Spatial scale
+xs <- xs + 3 * bx
+for (i in nImpPC1:0) {
+  lines(x = c(grconvertX(xs, "in", "user"), grconvertX(xs + 2 * bx, "in", "user")), 
+        rep(i + 0.5, 2), col = "grey")
+}
+
+x <- ifelse(!grepl("_SA", imp2), xs + bx / 2, xs + 3 * bx / 2)
+points(grconvertX(x, "in", "user"), nImpPC1:1, pch = 19)
+
+for (i in c(grconvertX(xs, "in", "user"), 
+            grconvertX(xs + bx, "in", "user"),
+            grconvertX(xs + 2 * bx, "in", "user"))) {
+  lines(rep(i, 2), c(0.5, nImpPC1 + 0.5), col = "grey")
+}
 
@@
 
-# x <- match(vsurfPC1$imp.mean.dec.ind, vsurfPC1_SA$imp.mean.dec.ind)
-# plot(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], type = "n")
-# text(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x], 
-#		 labels = climVars[vsurfPC1$imp.mean.dec.ind])
-# cor.test(vsurfPC1$imp.mean.dec, vsurfPC1_SA$imp.mean.dec[x])
-
-  # Plot most important climatic predictor versus PC1
-  pdf(paste(pathFig, "/Figure_bio16cva_v_Lat.pdf", sep = ""), 5, 5)
-  par(mar = c(5, 5, 1, 1))
-  plot(focClim$bio16_cva, focClim$PC1, xlim = c(0.25, 0.55), ylim = c(-4, 4),
+##### Plot most important climatic predictors versus PC1 #####
+
+  pdf(paste(pathFig, "/Figure_ClimPC1.pdf", sep = ""), 8, 8)
+
+	# Panel A. bio16_cva_SA
+  par(mai = c(1, 1, 0.25, 0.25), mfrow = c(2, 2))
+  plot(focClim$bio16_cva_SA, focClim$PC1, xlim = c(0.25, 0.55), ylim = c(-4, 4),
        axes = F, frame.plot = T, type = "n", xlab = "", ylab = "")
+  text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
+       labels = "A", font = 2, cex = 2)
   title(xlab = "Precipitation of Wettest Quarter", cex.lab = 1.5)
   title(xlab = "(Coefficient of variation)", cex.lab = 1, line = 4)
+  title(ylab = "TraitPC1", cex.lab = 1.5, line = 4)
   title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
@@
   axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
-  fitBio16cva <- lm(PC1 ~ bio16_cva, data = focClim)
+  fitBio16cva <- lm(PC1 ~ bio16_cva_SA, data = focClim)
 
   # polygon of confidence intervals
-  x <- seq(min(focClim$bio16_cva), max(focClim$bio16_cva), l = 1e3)
-  u <- predict(fitBio16cva, new = data.frame(bio16_cva = x))
+  x <- seq(min(focClim$bio16_cva_SA), max(focClim$bio16_cva_SA), l = 1e3)
+  u <- predict(fitBio16cva, new = data.frame(bio16_cva_SA = x))
   s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitBio16cva) %*% c(1, X)))
@@
   lines(x, confInt$upperCI, lwd = 2, lty = 2)
-  lines(range(focClim$bio16_cva), range(predict(fitBio16cva)), lwd = 2)
-  points(focClim$bio16_cva, focClim$PC1, pch = 21, col = "black", 
+  lines(range(focClim$bio16_cva_SA), range(predict(fitBio16cva)), lwd = 2)
+  points(focClim$bio16_cva_SA, focClim$PC1, pch = 21, col = "black", 
          bg = "white", cex = 1.5, lwd = 2)
+
+  # R2-value
+  text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1), 
+	     labels = bquote(italic(R)^2 == .(round(summary(fitBio16cva)$r.squared, 2))))
+
+  # Panel B. bio15_cva_SA
+  plot(focClim$bio15_cva_SA, focClim$PC1, xlim = c(0.14, 0.24), ylim = c(-4, 4),
+       axes = F, frame.plot = T, type = "n", xlab = "", ylab = "")
+  text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
+       labels = "B", font = 2, cex = 2)
+  title(xlab = "Precipitation Seasonality", cex.lab = 1.5)
+  title(xlab = "(Coefficient of variation)", cex.lab = 1, line = 4)
+  title(ylab = "TraitPC1", cex.lab = 1.5, line = 4)
+  title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
+    list(x = pve)))
+  axis(1, lwd = 0, lwd.ticks = 1) 
+  axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
+  fitBio15cva <- lm(PC1 ~ bio15_cva_SA, data = focClim)
+
+  # polygon of confidence intervals
+  x <- seq(min(focClim$bio15_cva_SA), max(focClim$bio15_cva_SA), l = 1e3)
+  u <- predict(fitBio15cva, new = data.frame(bio15_cva_SA = x))
+  s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitBio15cva) %*% c(1, X)))
+  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
+                        upperCI = qnorm(0.975, u, s))
+  polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
+    col = "white", border = NA)
+  lines(x, confInt$lowerCI, lwd = 2, lty = 2)
+  lines(x, confInt$upperCI, lwd = 2, lty = 2)
+  lines(x, u, lwd = 2)
+  points(focClim$bio15_cva_SA, focClim$PC1, pch = 21, col = "black", 
+         bg = "white", cex = 1.5, lwd = 2)
+
+  # R2-value
+  text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1), 
+	     labels = bquote(italic(R)^2 == .(round(summary(fitBio15cva)$r.squared, 2))))
+
+  # Panel C. bio15_cva_SA vs bio16_cva_SA
+  plot(focClim$bio16_cva_SA, focClim$bio15_cva_SA, xlim = c(0.25, 0.55), ylim = c(0.14, 0.24),
+       axes = F, frame.plot = T, type = "n", xlab = "", ylab = "")
+  text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
+       labels = "C", font = 2, cex = 2)
+  title(xlab = "Precipitation of Wettest Quarter", cex.lab = 1.5)
+  title(xlab = "(Coefficient of variation)", cex.lab = 1, line = 4)
+  title(ylab = "Precipitation Seasonality", cex.lab = 1.5, line = 4)
+  title(ylab = "(Coefficient of variation)", cex.lab = 1)
+  axis(1, lwd = 0, lwd.ticks = 1) 
+  axis(2, las = 1, lwd = 0, lwd.ticks = 1)    
+  fitBio16cva_Bio15cva <- lm(bio15_cva_SA ~ bio16_cva_SA, data = focClim)
+
+  # polygon of confidence intervals
+  x <- seq(min(focClim$bio16_cva_SA), max(focClim$bio16_cva_SA), l = 1e3)
+  u <- predict(fitBio16cva_Bio15cva, new = data.frame(bio16_cva_SA = x))
+  s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitBio16cva_Bio15cva) %*% c(1, X)))
+  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
+                        upperCI = qnorm(0.975, u, s))
+  polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
+    col = "white", border = NA)
+  lines(x, confInt$lowerCI, lwd = 2, lty = 2)
+  lines(x, confInt$upperCI, lwd = 2, lty = 2)
+  lines(x, u, lwd = 2)
+  points(focClim$bio16_cva_SA, focClim$bio15_cva_SA, pch = 21, col = "black", 
+         bg = "white", cex = 1.5, lwd = 2)
+
+  # R2-value
+  text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1), 
+	     labels = bquote(italic(R)^2 == .(round(summary(fitBio16cva_Bio15cva)$r.squared, 2))))
+
+  # Panel D. Latitude vs bio16_cva_SA
+  plot(thinnedOccClimate$Latitude, thinnedOccClimate$bio16_cva_SA, 
+       xlim = c(32, 44), ylim = c(0.25, 0.55), axes = F, frame.plot = T, type = "n", 
+       xlab = "", ylab = "")
+  text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
+       labels = "D", font = 2, cex = 2)
+  title(xlab = "Latitude", cex.lab = 1.5)
+  axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), 
+		expression(36*degree), expression(40*degree), expression(44*degree)), 
+		lwd = 0, lwd.ticks = 1) 
+  title(ylab = "Precipitation of Wettest Quarter", cex.lab = 1.5, line = 4)
+  title(ylab = "(Coefficient of variation)", cex.lab = 1)
+  axis(2, las = 1, lwd = 0, lwd.ticks = 1)    
+  points(thinnedOccClimate$Latitude, thinnedOccClimate$bio16_cva_SA, 
+         col = "grey", pch = 19)
+  points(focClim$Latitude, focClim$bio16_cva_SA, pch = 21, col = "black", 
+         bg = "white", cex = 1.5, lwd = 2)
+  legend(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), pch = c(19, 21),
+         col = c("grey", "black"), bg = "white", bty = "n", xjust = 1, yjust = 1,
+         legend = c("All occurences", "Focal populations"))
+
   dev.off()
 
-  # *** SHOULD BE MOVED TO MORE LOGICAL POSITION *** #
-  fitLat <- lm(PC1 ~ Latitude, data = focClim)
-  fitBio16_cva <- lm(PC1 ~ bio16_cva, data = focClim)
-  fitBio12_cva <- lm(PC1 ~ bio12_cva, data = focClim)
-  fitDD_0_cva <- lm(PC1 ~ DD_0_cva, data = focClim)
-  fit2var <- lm(PC1 ~ bio16_cva + DD_0_cva, data = focClim)
-
-  # Overlap between latitude predictors and phenotype predictors (spatial aveage)
-
-  # Figure of variable importance for traits (spatially averaged)
-	pdf(paste(pathFig, "/Figure_clim2pc1SA_imp.pdf", sep = ""), 4, 5)
-	par(mar = c(5, 1, 1, 6))
-
-	plot(0, 0, type = "n", xlim = c(0, 0.3), ylim = c(1, nImpLat), 
-		axes = F, frame.plot = T, xlab = "", ylab = "")
-	title(xlab = "Climate-Trait PC1\nImportance", line = 4, cex.lab = 1.5)
-	for (i in (nImpLat:(nImpLat - nImpPC1_SA + 1))) 
-	{
-		lines(x = c(-100, 100), y = c(i, i), col = "grey")
-	}
-	points(vsurfPC1_SA$imp.mean.dec[1:nImpPC1_SA], 
-	       nImpLat:(nImpLat - nImpPC1_SA + 1), 
-	       pch = 21, col = "black", bg = "white", cex = 1.5)
-	mtext(climVars[vsurfPC1_SA$varselect.pred], side = 4, las = 1, line = 1, 
-    at = nImpLat:(nImpLat - nImpPC1_SA + 1), 
-		font = ifelse(vsurfPC1_SA$varselect.pred %in% vsurfLat$varselect.pred,	2, 1))
-	axis(1, lwd = 0, lwd.ticks = 1, at = seq(0, 0.3, 0.1))
+##### Figure of bio11_cva_SA versus PC1 #####
+
+  pdf(paste(pathFig, "/FigureS_bio11PC1.pdf", sep = ""), 5, 5)
+
+  par(mar = c(5, 5, 1, 1))
+  plot(focClim$bio11_cva_SA, focClim$PC1, xlim = c(0.002, 0.005), ylim = c(-4, 4),
+       axes = F, frame.plot = T, type = "n", xlab = "", ylab = "")
+  title(xlab = "Temperature of Coldest Quarter", cex.lab = 1.5)
+  title(xlab = "(Coefficient of variation)", cex.lab = 1, line = 4)
+  title(ylab = "TraitPC1", cex.lab = 1.5, line = 3)
+  title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
+    list(x = pve)), line = 2)
+  axis(1, lwd = 0, lwd.ticks = 1) 
+  axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
+  fitBio11cva <- lm(PC1 ~ bio11_cva_SA, data = focClim)
+
+  # polygon of confidence intervals
+  x <- seq(min(focClim$bio11_cva_SA), max(focClim$bio11_cva_SA), l = 1e3)
+  u <- predict(fitBio11cva, new = data.frame(bio11_cva_SA = x))
+  s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitBio11cva) %*% c(1, X)))
+  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
+                        upperCI = qnorm(0.975, u, s))
+  polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
+    col = "white", border = NA)
+  lines(x, confInt$lowerCI, lwd = 2, lty = 2)
+  lines(x, confInt$upperCI, lwd = 2, lty = 2)
+  points(x, u, type = "l", lwd = 2)
+  points(focClim$bio11_cva_SA, focClim$PC1, pch = 21, col = "black", 
+         bg = "white", cex = 1.5, lwd = 2)
+
+  # R2-value
+  text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1), 
+	     labels = bquote(italic(R)^2 == .(round(summary(fitBio11cva)$r.squared, 2))))
+
   dev.off()
 
-# Table of important climatic variables
+##### Table of important climatic variables #####
 
@@
 tableS <- file(paste(pathMS, "/Tables/TableS_ClimVarImp.txt", sep = ""), open = "w")
-for (i in 1:max(c(nImpLat, nImpPC1, nImpPC1_SA))) {
+for (i in 1:max(c(nImpLat, nImpPC1))) {
 clim2sym[imp1[i], "name"]
-	# Latitude variable
+	# Climate-Latitude
 	if (nImpLat >= i) {
-		cat(sprintf('%s ($\\%s$) & ', clim2sym[imp1[i], "name"], clim2sym[imp1[i], "what"]), file = tableS)
+		cat(sprintf('%s ($\\%s$, %s) & ', clim2sym[imp1[i], "name"], clim2sym[imp1[i], "what"], clim2sym[imp1[i], "spatialScale"]), file = tableS)
 	}else{
@@
 
-	# Point estimated trait variable
+	# Climate-TraitPC1
 	if (nImpPC1 >= i) {
-		cat(sprintf('%s ($\\%s$) & ', clim2sym[imp2[i], "name"], clim2sym[imp2[i], "what"]), file = tableS)
-	}else{
-		cat(' & ', file = tableS)		
-	}
-
-	# Spatially averaged trait variable
-	if (nImpPC1_SA >= i) {
-		cat(sprintf('%s ($\\%s$) \\\\\n ', clim2sym[imp3[i], "name"], clim2sym[imp3[i], "what"]), file = tableS)
+		cat(sprintf('%s ($\\%s$, %s) \\\\\n', clim2sym[imp2[i], "name"], clim2sym[imp2[i], "what"], clim2sym[imp2[i], "spatialScale"]), file = tableS)
 	}else{
@@
 close(tableS)
+
+##### Two-factor model #####
+
+  fit2var1 <- lm(PC1 ~ bio15_cva_SA + bio11_cva_SA, data = focClim)
+  fit2var2 <- lm(PC1 ~ bio16_cva_SA + bio11_cva_SA, data = focClim)
 
@@
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} version 1.1-1 \citep{Hijmans_etal_2016}.
-In total, we included \Sexpr{length(climVars) / 2} climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+We included \Sexpr{length(climVars) / 4} climate factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
 The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
 For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
-Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.
+Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation (CV).
+In total, the VSURF algorithm selected among \Sexpr{length(climVars)} climate variables: \Sexpr{length(climVars) / 4} climatic factors $\times$ 2 types (30-year average and CV) $\times$ 2 spatial scales (local and neighborhood).
 
@@
 Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Treatment effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
+However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
 Complete ANOVA tables are available in the Supporting Information (Table S3-S6).
 
-\subsection*{Climatic variability best explains latitudinal cline}
+\subsection*{Neighborhood climatic variability best explains latitudinal cline}
 
 Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
-For \textit{M.
-cardinalis}, interannual variation in precipitation over 30 years (1981--2010) is very closely related to the latitude of recently recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
-Specifically, precipitation is more variable at lower latitude.
-The two most important Climate-Latitude variables were the interannual variation in total precipitation (bio12$_\sigma$; see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations) and precipitation in the wettest quarter (bio16$_\sigma$).
+To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variable explains latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
+Based on exploratory variable selection analyses, one or two aspects of interannual climatic variation (hypothesis 1) and climatic neighborhoods (hypothesis 3) are likely more important drivers of trait divergence than average climate immediately surrounding populations.
+These general conclusions are based on the fact that interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{M.
+cardinalis} occurences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
+All \Sexpr{nImpLat} Climate-Latitude and \Sexpr{nImpPC1} Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
+The most important climatic variables related to the interannual variation rather than average climate.
+
+Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{M.
+cardinalis} (see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
 Note that the coefficient of variation of a climatic factor is subscripted with $\sigma$ whereas the mean is subscripted with $\mu$.
-Bio12$_\sigma$ and bio16$_\sigma$ are very similar because in Mediterranean climates of California, most precipitation occurs in the winter quarter.
-These Climate-Latitude variables were also the two most important point estimated Climate-TraitPC1 variables (Fig.~\ref{fig:Fig_ClimVarImp}B).
-Populations from Southern, more variable environments grew faster.
-However, neither climatic variable alone explained more variation in TraitPC1 than latitude (latitude $r^2=\Sexpr{round(summary(fitLat)["r.squared"]$r.squared, 2)}$; bio16$_\sigma$ $r^2=\Sexpr{round(summary(fitBio16_cva)["r.squared"]$r.squared, 2)}$; bio12$_\sigma$ $r^2=\Sexpr{round(summary(fitBio12_cva)["r.squared"]$r.squared, 2)}$).
-Interannual variation in chilling degree-days (DD\_0$_\sigma$) was also common to both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
-Southern, faster growing populations are from climates with less variation in chilling degree-days.
-However, DD\_0$_\sigma$ and bio16$_\sigma$ together still only explained slightly more variation in TraitPC1 than latitude alone (latitude $r^2=\Sexpr{round(summary(fitLat)["r.squared"]$r.squared, 2)}$ versus DD\_0$_\sigma$+bio16$_\sigma$ $r^2=\Sexpr{round(summary(fit2var)["r.squared"]$r.squared, 2)}$).
-This suggests that interannual variation in precipitation explains most of the latitudinal cline, but other variables such as variation in chilling degree-days may contribute.
-
-There was no overlap between Climate-Latitude and spatially averaged Climate-TraitPC1 (Fig.~\ref{table:TableS_ClimVarImp}).
-This is not because climatic neighboorhood did not predict TraitPC1; there were significant correlations between important spatially averaged climatic factors and TraitPC1 that explained as much variation as latitude (data not shown).
-If these correlations are spurious, this indicates that climatic neighborhood is not as imporant as climate in the immediate vicinity of a population.
-Conversely, the climatic neighborhood may be important among our focal populations and depart from rangewide correlations between latitude and climate.
-Since these analyses are exploratory and prone to overinterpretation, we believe the closer overlap between Climate-Latitude and point estimated Climate-TraitPC1 variables provides more compelling evidence for putative selective agents.
+More specifically, greater winter precipitation variability and lower winter temperature variability are associted with Southern latitudes and higher TraitPC1 values (Fig.~\ref{fig:Fig_ClimPC1}A,B).
+Neighborhood interannual variation in winter precipitation (bio16$_\sigma$) was the most important Climate-Latitude variable (Fig.~\ref{fig:Fig_ClimVarImp}A).
+However, neighborhood bio16$_\sigma$ did not overlap with Climate-TraitPC1 variables (Fig.~\ref{fig:Fig_ClimVarImp}B).
+We nevertheless consider it a plausible candidate for two reasons.
+First, neighborhood bio16$_\sigma$ correlated strongly with TraitPC1 (Fig.~\ref{fig:Fig_ClimPC1}A).
+Second, one of the most important Climate-TraitPC1 variables (neighborhood bio15$_\sigma$; Fig.~\ref{fig:Fig_ClimPC1}B,C) is very similar to bio16$_\sigma$.
+In Mediterranean climates like California, most precipitation occurs in the wettest quarter (winter), so years with low winter precipitation also have low precipitation seasonality.
+Hence, highly variable year-to-year winter precipitation at lower latitude (Fig.~\ref{fig:Fig_ClimPC1}D) is closely associated with large swings in precipitation seasonality (Fig.~\ref{fig:Fig_ClimPC1}C).
+
+Interannual variation in temperature of the coldest quarter (neighborhood bio11$_\sigma$) is another plausible candidate because it was only variable in both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
+Neighborhood bio11$_\sigma$ explained more variation in TraitPC1 than latitude (latitude $r^2=\Sexpr{round(summary(fitLat)["r.squared"]$r.squared, 2)}$ vs.
+bio11$_\sigma$ $r^2=\Sexpr{round(summary(fitBio11cva)["r.squared"]$r.squared, 2)}$; Fig.~\ref{fig:FigS_bio11PC1}), whereas neighborhood bio16$_\sigma$ did slightly worse (bio16$_\sigma$ $r^2=\Sexpr{round(summary(fitBio16cva)["r.squared"]$r.squared, 2)}$).
+Models using bio15$_\sigma$ or bio11$_\sigma$ to predict TraitPC1 also had significantly lower Akaike Information Criteria (AIC) than the latitude model (AIC of different models -- bio15$_\sigma$: \Sexpr{round(AIC(fitBio15cva), 1)}; bio11$_\sigma$: \Sexpr{round(AIC(fitBio11cva), 1)}; latitude: \Sexpr{round(AIC(fitLat), 1)}).
+The best two-factor model including both neighborhood bio15$_\sigma$ and bio11$_\sigma$ did not significantly improve explanatory power ($r^2=\Sexpr{round(summary(fit2var1)["r.squared"]$r.squared, 2)}$, AIC$=\Sexpr{round(AIC(fit2var1), 1)}$).
+In summary, either variation in precipitation or temperature seasonality may be important selective agents, but there is no strong evidence that they are both important.
+The most important Climate-TraitPC1 variable, neighborhood variation in mean diurnal range (bio2$_\sigma$; Fig.~\ref{fig:Fig_ClimVarImp}B) did not have any obvious similarity to Climate-Latitude variables.
+Given the large number of potential associations, we therefore think this may be a spuriously strong relationship.
 
@@
 	\selectfont
-	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{Interannual variation in precipitation is closely correlated with latitude and trait variation.
+	\caption[Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
+cardinalis} and trait variation]{Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
+cardinalis} and trait variation.
 A.
@@
 B.
-The two most important Climate-Latitude variables were also the two most important Climate-TraitPC1 variables.
+Only one of of the most important Climate-Latitude variables (in bold) was among the most important Climate-TraitPC1 variables.
 Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
-Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp), Precipitation (Prec), Chilling (chilling degree-days), Snow (precipitation as snow); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
+Climatic variables (left of A; right of B) are defined by four qualities: Climatic factor -- Temperature (Temp), Precipitation (Prec), Heating degree-days (Heating), Snow (precipitation as snow); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Type -- 30-year average (Avg.) or coefficient of variation (Var.); Spatial scale -- local or 62-km radius climatic neighborhood.}
 	\label{fig:Fig_ClimVarImp}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of key Climate-TraitPC1 relationships
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimPC1.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Variation in precipitation is correlated with TraitPC1 and latitude]{Variation in precipitation is correlated with TraitPC1 and latitude.
+A.
+Greater values of TraitPC1 are associated with greater interannual variation in precipitation of the wettest quarter.
+This was the most important Climate-Latitude variable, but not among the most important Climate-TraitPC1 variables.
+B.
+However, a closely related parameter, interannaul variation in precipitation seasonality, was among the most important Climate-TraitPC1 variables.
+C.
+Across focal populations, variation in precipitation of the wettest quarter and seasonality are closely correlated.
+D.
+Southern populations of \textit{M.
+cardinalis} experience much greater interannual variationi in precipitation.
+In all panels, we report climatic neighborhood values (see Material and Methods).
+Regression lines, 95\% confidence intervals, and coefficients of determination ($R^2$) were calculated using linear regression.}
+	\label{fig:Fig_ClimPC1}
 \end{figure}
@@
 
-% Maybe focus in on tradeoffs as sine qua non of specialization vs generalization, then mention that demography and/or gene flow could reinforce?
-
-% perhaps responses of Simiolus sp (annual, drought selects for escape rather than tolerance) differ from card (perennial, escape through growth rather than flowering time) because drying out of rivers is less severe than seeps and springs?
-
 \section*{Discussion}
 
-% From Amy: \ala{the results that underlie this statement [niche conservation] need more emphasis in the methods}
 We found evidence for one of two common signatures of local adaptation in the perennial herb \textit{Mimulus cardinalis}.
-Latitudinal clines in germination rate, photosynthesis, and growth, suggests adaptive differentiation in fundamental physiological traits of the species.
+Latitudinal clines in germination rate, photosynthesis, and growth, suggest adaptive differentiation in fundamental physiological traits of the species.
 However, we found little evidence that populations respond differently to temperature or drought.
-As we discuss below, this may indicate that the fundamental abiotic niche is relatively conserved.
-Finally, we found that climatic variation between years may be a more important selective agent than the average climate.
-In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in photosynthesis and growth varies clinally, but plastic responses to temperature and drought are relatively conserved.
+As we discuss below, this latter result may indicate that the fundamental abiotic niche is relatively conserved.
+Finally, we found that neighborhood climatic variation between years may be a more important selective agent than the local average climate.
+In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in physiology changes clinally, whereas plastic responses to temperature and drought are relatively similar.
 
@@
 Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range \citep{Paul_etal_2016}.
+Nevertheless, local gene flow from similar environments may shape how selection varies with latitude.
+Climatic neighborhoods, defined here as the average climate in a 62-km radius around populations, correlated more strongly with trait and population latitude than local climate.
+This may not greatly constrain local adaptation because local and neighborhood climate values were generally similar in \textit{M.
+cardinalis} populations (data not shown).
 
@@
 Next, we would like to understand why variation in these particular traits may be adaptive.
-We argue that temporally more variable environments, as experienced by southern populations, select for a more ‘annualized’ life-history strategy.
+We argue that temporally more variable environments, as experienced by southern populations, select for a more `annualized` life-history strategy.
 
@@
 If drought years occur frequently enough, selection may favour the fast-growing strategy because there is no advantage to storage if drought kills plants before flowering.
-Considering life-history strategy as a continuum from no storage (annual) to lots of storage (perennial), we hypothesize that the optimal allocation to aboveground growth is more ‘annualized’ in southern climates that have greater interannual variation in precipitation.
+Considering life-history strategy as a continuum from no storage (annual) to lots of storage (perennial), we hypothesize that the optimal allocation to aboveground growth is more `annualized` in southern climates that have greater interannual variation in precipitation.
 
@@
 data).
-Although much work remains, the present works indicates that future studies of \textit{M.
-cardinalis} will deliver new, complementary insight into the physiological traits and climatic drivers of local adaptation.
+Exploratory analysis also pointed to variation in winter temperature as a potential selective agent.
+Specifically, winter temperature is more variable in the north, which may lead to less predictability in the start of the growing season or late frost.
+This could affect selection on phenology, which we did not study here, but it is not obviously connected to the latitudinal cline in physiological traits.
 
 In summary, we found evidence for a coordinated latitudinal cline in germination rate, photosynthesis, and growth, suggesting local adaptation.
-These fixed differences between populations suggest different trait optima in different climates.
+These differences between populations suggest different trait optima in different climates.
 We did not find evidence that the relative performance of populations shifts with temperature or watering regime, suggesting relatively little variation in plasticity.
 Exploratory analysis implicate that more variable precipitation regimes at lower latitude drive much of the latitudinal cline, though other climatic factors could also contribute.
-There was little compelling evidence that selection over a broad climatic neighborhood surrounding populations strongly explained the latitudinal cline.
+Interestingly, the climatic neighborhood may shape selective pressures than local climate.
 In the future, we will use field experiments to test whether greater variation in precipitation selects for faster growth and that selection on temperature/drought responses does not vary among populations.
@@
    \centering
-   \begin{tabular}{@{} ll @{}}
+   \begin{tabular}{@{} lp{12cm} @{}}
       \toprule
@@
   % BIOCLIM VARIABLES:
-  bio1	&	Annual Mean Temperature	\\
-	bio2	&	Mean Diurnal Range (Mean of monthly (max temp - min temp))	\\
-	bio3	&	Isothermality (bio2/bio7) (* 100)	\\
-	bio4	&	Temperature Seasonality (standard deviation *100)	\\
-	bio5	&	Max Temperature of Warmest Month	\\
-	bio6	&	Min Temperature of Coldest Month	\\
-	bio7	&	Temperature Annual Range (bio5-bio6)	\\
-	bio8	&	Mean Temperature of Wettest Quarter	\\
-	bio9	&	Mean Temperature of Driest Quarter	\\
-	bio10	&	Mean Temperature of Warmest Quarter	\\
-	bio11	&	Mean Temperature of Coldest Quarter	\\
-	bio12	&	Annual Precipitation	\\
-	bio15	&	Precipitation Seasonality (Coefficient of Variation)	\\
-	bio16	&	Precipitation of Wettest Quarter	\\
-	bio17	&	Precipitation of Driest Quarter	\\
-	bio18	&	Precipitation of Warmest Quarter	\\
-	bio19	&	Precipitation of Coldest Quarter	\\
+  bio1	&	annual mean temperature	\\
+	bio2	&	mean diurnal range (mean of monthly (max temp - min temp))	\\
+	bio3	&	isothermality (bio2/bio7) (* 100)	\\
+	bio4	&	temperature seasonality (standard deviation *100)	\\
+	bio5	&	max temperature of warmest month	\\
+	bio6	&	min temperature of coldest month	\\
+	bio7	&	temperature annual range (bio5-bio6)	\\
+	bio8	&	mean temperature of wettest quarter	\\
+	bio9	&	mean temperature of driest quarter	\\
+	bio10	&	mean temperature of warmest quarter	\\
+	bio11	&	mean temperature of coldest quarter	\\
+	bio12	&	annual precipitation	\\
+	bio15	&	precipitation seasonality (coefficient of variation)	\\
+	bio16	&	precipitation of wettest quarter	\\
+	bio17	&	precipitation of driest quarter	\\
+	bio18	&	precipitation of warmest quarter	\\
+	bio19	&	precipitation of coldest quarter	\\
 	    \bottomrule
@@
 \end{table}
-
-% Analysis of deviance table for germination
-
-<<TableS_fitGerm, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
-
-  # Convert numbers from R to LaTeX characters
-  tab <- suppressWarnings(anova(fitGerm))
-  germTab <- data.frame(pval = pval2latex(tab$`Pr(>Chi)`),
-                         stringsAsFactors = F)
-  rownames(germTab) <- rownames(tab)
-
-@
-
-% *** NEED TO DO TABLE IF I WANT IT ***
 
@@
 
-% Table of Climate-Latitude and Climate-Trait variables
+% Table of Climate-Latitude and Climate-TraitPC1 variables
 
 \begin{table}[ht]
-   \centering
-   \topcaption[Important climatic variables]{Important climatic variables predicting latitude of \textit{M.
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+  \caption[Important climatic variables]{Important climatic variables predicting latitude of \textit{M.
 cardinalis} populations (`Climate-Latitude') and the first principal component of traits measured in a common garden (`Climate-TraitPC1').
-Point estimated climatic variables were measured from the exact location of collection; spatially averaged climatic variables were averaged from a 62-km neighborhood around population (see Material and Methods).
+Local climatic variables were measured from the exact location of collection; neighborhood climatic variables were averaged from a 62-km neighborhood around population (see Material and Methods).
 Importance and significance were determined using the variable selection using random forests (VSURF) algorithm (see Material and Methods).
@@
 }
-   \resizebox{6.5in}{!}{
-   \begin{tabular}{@{} lll @{}}
+   \begin{center}
+   \begin{tabular}{@{} ll @{}}
       \toprule
-    	Climate-Latitude variables & Climate-TraitPC1 variables & \\
-                            	   & Point estimated         & Spatially averaged \\
+    	Climate-Latitude variables & Climate-TraitPC1 variables \\
       \midrule
@@
 			\bottomrule
-   \end{tabular} }
+   \end{tabular}
+   \end{center}
    \label{table:TableS_ClimVarImp}
@@
 Each point is a population of \textit{M.
-cardinalis} showing its latitude of origin (x-axis) and model-predicted stem elongation rate during the bolting phase.
+cardinalis} showing its latitude of origin (x-axis) and model-predicted stem elongation rate.
 Bars around each point are 95\% confidence intervals.
@@
 
-% Figure comparing importance of climatic variables from point estimated and spatially averaged analyses
-
-<<FigS_PEvsSA, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
-
-	pdf(paste(pathFig, "/FigureS_PEvsSA.pdf", sep = ""), 5.5, 5.5)
-	par(mai = c(1.5, 1.5, 0.1, 0.1))
-	plot(vsurfPC1$imp.mean.dec[vsurfPC1$imp.mean.dec.ind], 
-	     vsurfPC1_SA$imp.mean.dec[vsurfPC1_SA$imp.mean.dec.ind], 
-	     xlab = "", ylab = "", type = "n", axes = F, frame.plot = T)
-	 title(xlab = "Climate-TraitPC1 Importance\nPoint estimated", cex.lab = 1.5,
-	       line = 5) 
-	 title(ylab = "Climate-TraitPC1 Importance\nSpatially averaged", 
-	       cex.lab = 1.5, line = 4)
-
-	 text(vsurfPC1$imp.mean.dec[vsurfPC1$imp.mean.dec.ind], 
-	      vsurfPC1_SA$imp.mean.dec[vsurfPC1_SA$imp.mean.dec.ind],
-	      labels = climVars)
-
-	axis(1, lwd = 0, lwd.ticks = 1)
-	axis(2, lwd = 0, lwd.ticks = 1, las = 1)
-
-	dev.off()
-
-@
-
-% *** Need to double check 62-km buffer results and make figure prettier *** %
-% \begin{figure}[h!]
-%	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_PEvsSA.pdf}}
-%	\usefont{T1}{phv}{m}{n}
-%	\fontsize{10}{12}
-%	\selectfont
-%	\caption[Point-estimated versus spatially averaged importance]{CAPTION}
-%	\label{fig:FigS_PEvsSA}
-% \end{figure}
+% Figure of bio11_cva_SA versus TraitPC1
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_bio11PC1.pdf}}
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\caption[Trait PC1 versus bio11_cva_SA]{Trait variation, from fast to slow growth, is closely associated with neighborhood variation in temperature of the coldest quarter (bio11$_\sigma$) Each point is a population coefficient of variation in bio11 averaged over a 62-km climatic neighborhood  (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of four traits (see Material and Methods).
+The line and 95\% confidence intervals were estimated using linear regression.}
+\label{fig:FigS_bio11PC1}
+\end{figure}
 

### 2017-01-06 ms/ms.tex
Added funding info
Major revision

@@
-%lost on page 11 - extracted and used climate variables
-%cartoon of whole procedure for climate variables
-% model of growth chamber
+%%%%% ISSUES %%%%%%
 % need to explain 'importance'
 
-%* delve into interactions - do they change cline?
-
-%* show figures of reaction norms
-%* split plot, account for error variation in temp???
-%* show PAR similarity
-%* some sites they are functionally annual
-
-%* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
-%	- maybe temporal source-sink makes more sense
-
-%* put in biomass allocation cline
-% TO DO:
-% 2. Double check spatially averaged analysis
-% 3. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
-% 4. Do anything with biomass data? Currently, NO.
-
-% 5. Citations for latitudinal clines
-	% cases where clinal variation has 'obvious' selective source:
-	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
-	% Caicedo et al 2004?
-	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
-% 6. Citations for GxE as signature of local adaptation
+% delve into interactions - do they change cline?
+% show figures of reaction norms
+% show PAR similarity
+% hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
+% For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
 
@@
       \midrule
-	HAU & South Margin & 32.657	& 
+	Hauser Creek & South Margin & 32.657	& 
     -116.532	& 799   \\
-	CTC	& South Margin & 32.609 & 
+	Cottonwood Creek	& South Margin & 32.609 & 
     -116.7	& 267   \\
-	CUR	& South Margin & 32.9 & 
+	Sweetwater River	& South Margin & 32.9 & 
     -116.585	& 1180   \\
-	GRP & South Margin & 33.314 &
+	Grade Road Palomar & South Margin & 33.314 &
     -116.871	& 1577   \\
-	WWC &	Transverse & 33.994 & 
+	Whitewater Canyon &	Transverse & 33.994 & 
     -116.665	& 705   \\
-	MIL	& Transverse & 34.077 & 
+	Mill Creek	& Transverse & 34.077 & 
     -116.873	& 2050   \\
-	WFM	& Transverse & 34.284 & 
+	West Fork Mojave River	& Transverse & 34.284 & 
     -117.378	& 1120   \\
-	NMT	& South Sierras & 36.201 & 
+	North Fork Middle Tule River	& South Sierras & 36.201 & 
     -118.651	& 1314   \\
-	PRD	& South Sierras & 36.518 & 
+	Paradise Creek	& South Sierras & 36.518 & 
     -118.759	& 926   \\
-	RWD	& South Sierras & 36.691 & 
+	Redwood Creek	& South Sierras & 36.691 & 
     -118.91	& 1727   \\
-	WNA	& Central Sierras & 37.541 & 
+	Wawona  & Central Sierras & 37.541 & 
     -119.649	& 1224   \\
-	RBW	& Central Sierras	& 37.819 & 
+	Rainbow Creek	& Central Sierras	& 37.819 & 
     -120.007	& 876   \\
-	MYU	& North Sierras	& 39.397 & 
+	Middle Yuba River	& North Sierras	& 39.397 & 
     -121.082	& 455   \\
-	LIJ	& North Sierras	& 39.743 & 
+	Little Jamison Creek	& North Sierras	& 39.743 & 
     -120.704	& 1603   \\
-	DPC	& North Coast & 41.668 & 
+	Deep Creek	& North Coast & 41.668 & 
     -123.11	& 707   \\
-	RCC	& North Margin & 43.374 & 
+	Rock Creek	& North Margin & 43.374 & 
     -122.957	& 326   \\
@@
 \begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelectBase.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Climatic variable selection]{Overview of process for identifying putative climatic selective drivers underlying latitudinal cline.
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelect.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Climatic variable selection]{Overview of method for identifying putative climatic selective agents underlying latitudinal cline.
 We looked for climate variables that explained both the latitude of 356 \textit{M.
 cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
-(A) For Climate-Latitude variables we extracted climate data from the specific location where a population was collected ($x_{i,0}$) throughout California and Oregon, USA (shown in map).
-For Climate-Trait variables, we extracted climatic data in two ways, but only for the 16 focal populations.
-(B) For point-estimated Climate-Trait variables, we extracted climate only from where a population was collected ($x_{i,0}$).
-(C) For spatially averaged Climate-Trait variables, we averaged over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M.
-cardinalis} is riparian.}
+For Climate-Latitude variables we extracted climate data from recent occurences located throughout California and Oregon, USA (shown in map).
+For Climate-Trait variables, we extracted climatic data for the 16 focal populations.
+For both analyses, we extracted local and neighborhood climate.
+Local climate refers to climate only from where a population was collected ($x_{i,0}$).
+Neighborhood climate was calculated as the average over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M.
+cardinalis} is riparian.
+We identified climatic factors that most strongly predicted latitude of occurences (Climate-Latitude variables) and traits (Climate-Trait variables), as shown for hypothetical data in plots at the bottom of the figure.}
 	\label{fig:Fig_VarSelect}
@@
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} version 1.1-1 \citep{Hijmans_etal_2016}.
-In total, we included 24 climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+We included 24 climate factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
 The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
 For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
-Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.
+Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation (CV).
+In total, the VSURF algorithm selected among 96 climate variables: 24 climatic factors $\times$ 2 types (30-year average and CV) $\times$ 2 spatial scales (local and neighborhood).
 
@@
 Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Treatment effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
+However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
 Complete ANOVA tables are available in the Supporting Information (Table S3-S6).
 
-\subsection*{Climatic variability best explains latitudinal cline}
+\subsection*{Neighborhood climatic variability best explains latitudinal cline}
 
 Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
-For \textit{M.
-cardinalis}, interannual variation in precipitation over 30 years (1981--2010) is very closely related to the latitude of recently recorded occurences of this species (Fig.~\ref{fig:Fig_ClimVarImp}A).
-Specifically, precipitation is more variable at lower latitude.
-The two most important Climate-Latitude variables were the interannual variation in total precipitation (bio12$_\sigma$; see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations) and precipitation in the wettest quarter (bio16$_\sigma$).
+To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variable explains latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
+Based on exploratory variable selection analyses, one or two aspects of interannual climatic variation (hypothesis 1) and climatic neighborhoods (hypothesis 3) are likely more important drivers of trait divergence than average climate immediately surrounding populations.
+These general conclusions are based on the fact that interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{M.
+cardinalis} occurences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
+All 16 Climate-Latitude and 3 Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
+The most important climatic variables related to the interannual variation rather than average climate.
+
+Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{M.
+cardinalis} (see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
 Note that the coefficient of variation of a climatic factor is subscripted with $\sigma$ whereas the mean is subscripted with $\mu$.
-Bio12$_\sigma$ and bio16$_\sigma$ are very similar because in Mediterranean climates of California, most precipitation occurs in the winter quarter.
-These Climate-Latitude variables were also the two most important point estimated Climate-TraitPC1 variables (Fig.~\ref{fig:Fig_ClimVarImp}B).
-Populations from Southern, more variable environments grew faster.
-However, neither climatic variable alone explained more variation in TraitPC1 than latitude (latitude $r^2=0.55$; bio16$_\sigma$ $r^2=0.52$; bio12$_\sigma$ $r^2=0.47$).
-Interannual variation in chilling degree-days (DD\_0$_\sigma$) was also common to both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
-Southern, faster growing populations are from climates with less variation in chilling degree-days.
-However, DD\_0$_\sigma$ and bio16$_\sigma$ together still only explained slightly more variation in TraitPC1 than latitude alone (latitude $r^2=0.55$ versus DD\_0$_\sigma$+bio16$_\sigma$ $r^2=0.6$).
-This suggests that interannual variation in precipitation explains most of the latitudinal cline, but other variables such as variation in chilling degree-days may contribute.
-
-There was no overlap between Climate-Latitude and spatially averaged Climate-TraitPC1 (Fig.~\ref{table:TableS_ClimVarImp}).
-This is not because climatic neighboorhood did not predict TraitPC1; there were significant correlations between important spatially averaged climatic factors and TraitPC1 that explained as much variation as latitude (data not shown).
-If these correlations are spurious, this indicates that climatic neighborhood is not as imporant as climate in the immediate vicinity of a population.
-Conversely, the climatic neighborhood may be important among our focal populations and depart from rangewide correlations between latitude and climate.
-Since these analyses are exploratory and prone to overinterpretation, we believe the closer overlap between Climate-Latitude and point estimated Climate-TraitPC1 variables provides more compelling evidence for putative selective agents.
+More specifically, greater winter precipitation variability and lower winter temperature variability are associted with Southern latitudes and higher TraitPC1 values (Fig.~\ref{fig:Fig_ClimPC1}A,B).
+Neighborhood interannual variation in winter precipitation (bio16$_\sigma$) was the most important Climate-Latitude variable (Fig.~\ref{fig:Fig_ClimVarImp}A).
+However, neighborhood bio16$_\sigma$ did not overlap with Climate-TraitPC1 variables (Fig.~\ref{fig:Fig_ClimVarImp}B).
+We nevertheless consider it a plausible candidate for two reasons.
+First, neighborhood bio16$_\sigma$ correlated strongly with TraitPC1 (Fig.~\ref{fig:Fig_ClimPC1}A).
+Second, one of the most important Climate-TraitPC1 variables (neighborhood bio15$_\sigma$; Fig.~\ref{fig:Fig_ClimPC1}B,C) is very similar to bio16$_\sigma$.
+In Mediterranean climates like California, most precipitation occurs in the wettest quarter (winter), so years with low winter precipitation also have low precipitation seasonality.
+Hence, highly variable year-to-year winter precipitation at lower latitude (Fig.~\ref{fig:Fig_ClimPC1}D) is closely associated with large swings in precipitation seasonality (Fig.~\ref{fig:Fig_ClimPC1}C).
+
+Interannual variation in temperature of the coldest quarter (neighborhood bio11$_\sigma$) is another plausible candidate because it was only variable in both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
+Neighborhood bio11$_\sigma$ explained more variation in TraitPC1 than latitude (latitude $r^2=0.55$ vs.
+bio11$_\sigma$ $r^2=0.6$; Fig.~\ref{fig:FigS_bio11PC1}), whereas neighborhood bio16$_\sigma$ did slightly worse (bio16$_\sigma$ $r^2=0.49$).
+Models using bio15$_\sigma$ or bio11$_\sigma$ to predict TraitPC1 also had significantly lower Akaike Information Criteria (AIC) than the latitude model (AIC of different models -- bio15$_\sigma$: 48.5; bio11$_\sigma$: 52.4; latitude: 54.5).
+The best two-factor model including both neighborhood bio15$_\sigma$ and bio11$_\sigma$ did not significantly improve explanatory power ($r^2=0.71$, AIC$=49.2$).
+In summary, either variation in precipitation or temperature seasonality may be important selective agents, but there is no strong evidence that they are both important.
+The most important Climate-TraitPC1 variable, neighborhood variation in mean diurnal range (bio2$_\sigma$; Fig.~\ref{fig:Fig_ClimVarImp}B) did not have any obvious similarity to Climate-Latitude variables.
+Given the large number of potential associations, we therefore think this may be a spuriously strong relationship.
 
@@
 	\selectfont
-	\caption[Interannual variation in precipitation is closely correlated with latitude and trait variation]{Interannual variation in precipitation is closely correlated with latitude and trait variation.
+	\caption[Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
+cardinalis} and trait variation]{Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
+cardinalis} and trait variation.
 A.
-Using Random Forest regression, we identified 11 climatic variables significantly (high importance) associated with latitude of \textit{M.
+Using Random Forest regression, we identified 16 climatic variables significantly (high importance) associated with latitude of \textit{M.
 cardinalis} occurences.
 B.
-The two most important Climate-Latitude variables were also the two most important Climate-TraitPC1 variables.
+Only one of of the most important Climate-Latitude variables (in bold) was among the most important Climate-TraitPC1 variables.
 Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
-Climatic variables (left of A; right of B) are defined by three qualities: Climatic factor -- Temperature (Temp), Precipitation (Prec), Chilling (chilling degree-days), Snow (precipitation as snow); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Summary statistic -- average ($\mu$) or coefficient of variation ($\sigma$)}
+Climatic variables (left of A; right of B) are defined by four qualities: Climatic factor -- Temperature (Temp), Precipitation (Prec), Heating degree-days (Heating), Snow (precipitation as snow); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Type -- 30-year average (Avg.) or coefficient of variation (Var.); Spatial scale -- local or 62-km radius climatic neighborhood.}
 	\label{fig:Fig_ClimVarImp}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of key Climate-TraitPC1 relationships
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimPC1.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Variation in precipitation is correlated with TraitPC1 and latitude]{Variation in precipitation is correlated with TraitPC1 and latitude.
+A.
+Greater values of TraitPC1 are associated with greater interannual variation in precipitation of the wettest quarter.
+This was the most important Climate-Latitude variable, but not among the most important Climate-TraitPC1 variables.
+B.
+However, a closely related parameter, interannaul variation in precipitation seasonality, was among the most important Climate-TraitPC1 variables.
+C.
+Across focal populations, variation in precipitation of the wettest quarter and seasonality are closely correlated.
+D.
+Southern populations of \textit{M.
+cardinalis} experience much greater interannual variationi in precipitation.
+In all panels, we report climatic neighborhood values (see Material and Methods).
+Regression lines, 95\% confidence intervals, and coefficients of determination ($R^2$) were calculated using linear regression.}
+	\label{fig:Fig_ClimPC1}
+\end{figure}
+
 %--------------------------------------------------
@@
 \end{table}
-
-% Analysis of deviance table for germination
-
-% *** NEED TO DO TABLE IF I WANT IT ***
 
@@
 
-% Figure comparing importance of climatic variables from point estimated and spatially averaged analyses
-
-% *** Need to double check 62-km buffer results and make figure prettier *** %
-% \begin{figure}[h!]
-%	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_PEvsSA.pdf}}
-%	\usefont{T1}{phv}{m}{n}
-%	\fontsize{10}{12}
-%	\selectfont
-%	\caption[Point-estimated versus spatially averaged importance]{CAPTION}
-%	\label{fig:FigS_PEvsSA}
-% \end{figure}
+% Figure of bio11_cva_SA versus TraitPC1
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_bio11PC1.pdf}}
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\caption[Trait PC1 versus bio11_cva_SA]{Trait variation, from fast to slow growth, is closely associated with neighborhood variation in temperature of the coldest quarter (bio11$_\sigma$) Each point is a population coefficient of variation in bio11 averaged over a 62-km climatic neighborhood  (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of four traits (see Material and Methods).
+The line and 95\% confidence intervals were estimated using linear regression.}
+\label{fig:FigS_bio11PC1}
+\end{figure}
 

### 2017-03-08 ms/ms.Rnw
Revised ms

@@
-%%%%% ISSUES %%%%%%
-% need to explain 'importance'
-
-% delve into interactions - do they change cline?
-% show figures of reaction norms
-% show PAR similarity
-% hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
-% For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
-
 \documentclass[11pt, oneside]{article}
@@
 \usepackage{textcomp}
+\usepackage{cleveref}
+\usepackage{subcaption}
 
@@
 
-\title{Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Mimulus cardinalis}}
+\title{Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}}
 \author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
@@
 
-Running Head: Latitudinal cline and climate in \textit{Mimulus} \\
-
-Key words: local adaptation, cline, photosynthesis, growth rate, \textit{Mimulus} \\
+Running Head: Latitudinal cline and climate in \textit{Erythranthe} \\
+
+Key words: local adaptation, cline, photosynthesis, growth rate, \textit{Erythranthe} \\
 
@@
 
+%--------------------------------------------------
+% Acknowledgements
+%--------------------------------------------------
+
+\section*{Acknowledgements}
+Erin Warkman and Lisa Lin helped collect data.
+CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program.
+ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
+
 \clearpage
 
-% \listoffigures
-% \listoftables
-
 \setstretch{\stretchy}
-\linenumbers
+
+\begin{center}
+{\LARGE  \centering Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}}
+\end{center}
 
@@
 
-Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
+Local adaptation is commonly observed in nature: organisms perform well in their natal environment, but poorly outside it.
 Correlation between traits and latitude, or latitudinal clines, are among the most common pieces of evidence for local adaptation, but identifying the traits under selection and the selective agents are challenging.
-Here, we investigated a latitudinal cline in growth and photosynthesis across 16 populations of the perennial herb \textit{Mimulus cardinalis} (Phrymaceae).
+Here, we investigated a latitudinal cline in growth and photosynthesis across 16 populations of the perennial herb \textit{Erythranthe cardinalis} (Phrymaceae).
 Using machine learning methods, we identify interannual variation in precipitation as a likely selective agent: Southern populations from more variable environments had higher photosynthetic rates and grew faster.
@@
 % Toggle these commands to get version for doing word count
-% \linenumbers
+\linenumbers
 % \pagenumbering{gobble}
@@
 % Local adaptation and clines
-Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}.
-Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}.
+Local adaptation within species has been documented in numerous species; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Leimu_Fischer_2008, Hereford_2009}.
+However, the prevalance of local adaptation remains difficult to assess because researchers rarely test for local adaptation unless there are obvious phenotypic or environmental differences (but see \citeauthor{Hereford_Winn_2008} \citeyear{Hereford_Winn_2008}).
+When local adaptation occurs, it frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977, Barton_1999}.
 Phenotypic differences between populations along a cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
-Despite a long history of studying local adaptation and clines, it remains challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
+Despite a long history of studying local adaptation and clines, it remains challenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
 In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment.
-We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change.
-A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures.
-% Here, we examine physiological trait variation and possible selective agents in the perennial herb \textit{Mimulus cardinalis} Douglas ex Benth.
-(Phrymaceae), a model system for local adaptation studies.
+A related problem is identifying which of the myriad and often covarying aspects of the environment causes spatially varying selective pressures.
 
@@
 Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away.
-Alternatively, genotype-by-environment interactions could indicate that variation in plasticity mediates local adaptation.
+Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive.
 We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively.
-Adaptation involves both intrinsic and plastic trait variation.
-For example, intrinsic differences in photoperiod \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
+Both intrinsic and plastic trait variation can be explained by genetic differences and both are involved in adaptation.
+For example, intrinsic differences in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
 Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
-In plants especially, we know little about the prevalance and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
+In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
 
@@
 A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments.
-Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
+Either intrinsic and/or plastic variation should vary clinally along environmental gradients.
 Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
@@
 Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
-Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008}, and drought response \citep{Kooyers_etal_2015} that likely relate to climatic variation.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008, Stock_etal_2014}, and drought response \citep{Kooyers_etal_2015} that likely relate to climatic variation.
 
@@
 Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days.
-Hence, latitude is an effective proxy for the underlying climatic driver, but we would expect a yet stronger relationship between traits and the key climatic variable(s) driving selection.
+Latitude may also correlate with the strength of biotic interactions \citep{Schemske_etal_2009} or other nonclimatic aspects of the environment, though as we explain below, we do not yet have compelling data that these are important in our study system.
+Hence, we focus on whether latitude could be an effective proxy for an underlying climatic driver, in which case we would expect a yet stronger relationship between traits and the key climatic variable(s) driving selection.
 Alternatively, latitude may be more strongly related to traits than any single climatic variable for at least two reasons.
 First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline.
-Alternatively, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
+Alternatively, gene flow among neighbouring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration \citep{Slatkin_1978, Paul_etal_2011, Hadfield_2016}.
 We refer to this as the `climatic neighborhood'.
@@
 
-In this study, we intestigated two major questions: 1) whether intrinsic or plastic physiological trait variation corresponds with latitude; and 2) what climatic factor(s) could plausibly be repsonsible for latitudinal clines.
+In this study, we investigated two major questions: 1) whether intrinsic or plastic physiological trait variation corresponds with latitude; and 2) what climatic factor(s) could plausibly be responsible for latitudinal clines.
 Within question 2, we tested three hypotheses outlined in the previous paragraph: latitudinal clines are explained by a single dominant climatic factor, multiple climatic factors, or the climatic neighborhood experienced by nearby population connected through gene flow.
 These hypotheses are not mutually exclusive since, for example, single or multiple factors in a climatic neighborhood may lead to latitudinal clines.
-
-We examined these questions in \textit{Mimulus cardinalis} because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
-Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
-Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}.
+We focused on climate because climate often determines and where species are found and also can exert strong selection on populations within species [CITATION?].
+There is also a compelling need to know how populations are (or are not) locally adapted to climate so as to predict how they’ll respond to climate change [CITATION?].
+We did not test for local adaptation to other environmental characteristicss, such as edaphic or biotic (e.g.
+competitors, pollinators) factors, because we did not have compelling observations from which to make testable hypotheses about the latter in our focal species (described below).
+
+We examined these questions in \textit{Erythranthe cardinalis} (formerly \textit{Mimulus cardinalis} [\citeauthor{Nesom_2014} \citeyear{Nesom_2014}]) because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
+Many classic and contemporary studies of local adaptation use \textit{Mimulus sensu lato} species because of its natural history, easy propagation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
+Yet, there is a conspicuous deficiency of links between local adaptation and physiological mechanisms (\cite{Angert_2006, Angert_etal_2008, Wu_etal_2010}, but see \cite{Wright_etal_2013} on mechanisms of copper tolerance).
 We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
 We found a latitudinal cline of intrinsic variation in photosynthesis and growth, but little evidence for variation in plasticity.
-Interannual variation in precipitation and temperature are associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
+Interannual variation in precipitation and temperature are associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{E.
 cardinalis}.
@@
 
+<<Summary Statistics, echo = FALSE, results=hide, warning=FALSE>>=
+
+  master <- read.csv(paste(pathDat, "/MasterDatasheet_out.csv", sep = ""), 
+    row.names = 1)
+
+  # Number of families per population
+  famTab <- tapply(master$Family, list(master$PopID), function(X) length(unique(X)))
+
+  # Individuals per population per treatment
+  indPerPop <- tapply(master$indiv, list(master$TempTrt, master$WaterTrt, master$PopID), length)
+
+@
+
 \subsection*{Population Selection}
 
-We used 16 populations from throughout the range of \textit{M.
+We used 16 populations from throughout the range of \textit{E.
 cardinalis} (Table~\ref{table:Table_FocalPops}).
+These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below).
 Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
+We used seeds from \Sexpr{nlevels(master$Family)} families, \Sexpr{min(famTab)}--\Sexpr{max(famTab)} (mean = \Sexpr{round(mean(famTab), 1)}, median = \Sexpr{median(famTab)}) families per population.
 
@@
 load(paste(pathDat, "/focClim.RData", sep = ""))
-
-# How similar are focClim and focClimSA? ** NEEDS UPDATE WITH NEW DATASHEET **
-# summary(mapply(cor, focClim[climVars], focClimSA[climVars]))
-# summary(mapply(cor, focClim[climVars], focClimSA[climVars]) ^ 2)
 
@@
 On 14 April, 2014, 3-5 seeds per family were sown directly on sand (Quikrete Play Sand, Georgia, USA) watered to field capacity in RLC4 Ray Leach cone-tainers placed in RL98 98-well trays (Stuewe \& Sons, Inc., Oregon, USA).
-We used pure sand both to facilitate root-washing and because \textit{M.
+We used pure sand because \textit{E.
 cardinalis} typically grows in sandy, riparian soils (A.
@@
 We thinned seedlings to one plant per cone-tainer, leaving the center-most plant.
-702 of 768 (91.4\%) had plants that could be used in the experiment.
+\Sexpr{nrow(master)} of 768 (\Sexpr{round(100 * nrow(master) / 768, 1)}\%) had plants that could be used in the experiment.
 We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments.
@@
 The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
+Lower humidity would have made the drought more severe,but low soil moisture is stressful in and of itself.
+The total number of plants in each treatment was: $n_{\text{cool,dry}} = \Sexpr{sum(indPerPop["Cool", "Dry", ])}$; $n_{\text{cool,ww}} = \Sexpr{sum(indPerPop["Cool", "Wet", ])}$; $n_{\text{hot,dry}} = \Sexpr{sum(indPerPop["Hot", "Dry", ])}$; $n_{\text{hot,ww}} = \Sexpr{sum(indPerPop["Hot", "Wet", ])}$.
+Each population had \Sexpr{min(indPerPop)}--\Sexpr{max(indPerPop)} individuals per treatment level (mean = \Sexpr{round(mean(indPerPop), 1)}, median = \Sexpr{median(indPerPop)}).
 
@@
 The bold lines track the average daily temperatures.
-Within each day, there was a maximum daytime temperture (top of translucent polygons) and minimum nighttime temperature (bottom of translucent polygons).
+Within each day, there was a maximum daytime temperature (top of translucent polygons) and minimum nighttime temperature (bottom of translucent polygons).
 The drought treatment commenced later by ramping down the frequency of bottom-watering episodes (dashed black line; right axis), while watering frequency was maintained in the control treatment (solid black line).
@@
   Leaf expansion rate  &  mm day$^{-1}$  \\
-  Stem elongation rate  &  mm or cm day$^{-1}$  \\
+  Stem elongation rate  &  cm day$^{-1}$  \\
   Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
@@
 We treated Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
-Statistical signifcance of the Population effect was determined using analysis of deviance.
+Statistical significance of the Population effect was determined using analysis of deviance.
 Note that, unlike other traits discussed below, we did not include Block, Treatment, or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not yet been applied.
@@
 We measured growth rate during two phases: leaf expansion and stem elongation.
-We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
+Growth measurements were taken during the early vegetative stage.
+We censused leaf length twice per week shortly after the emergence of true leaves from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
 We ceased measuring leaf length once it appeared to asymptote and growth shifted to stem elongation.
-We also censused plant height on 7 occasions (twice per week) between 29 May and 20 June (45 to 67 days after sowing).
-Both leaf expansion and stem elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and stem growth) using empirical Bayes' estimates from linear mixed-effects models fit with the R package \pkg{lme4} version 1.1-12 \citep{Bates_etal_2015}.
+We also censused plant height on 7 occasions (twice per week) between 29 May and 20 June (45 to 67 days after sowing) until plants began to initiate floral buds.
+Thus all growth measurements occured during the vegetative, prereproductive phase.
+Both leaf expansion and stem elongation were modelled separately as second-order polynomials.
+We used empirical Bayes' estimates of growth for each individual plant from linear mixed-effects models fit with the R package \pkg{lme4} version 1.1-12 \citep{Bates_etal_2015}.
 
@@
 
+	# Figure of reaction norms
+  tab <- tapply(newdata$LER_AbsGrowth, 
+                list(newdata$Population, newdata$WaterTrt, newdata$TempTrt), mean)
+  pdf(paste(pathFig, "/FigureS_LERrxnNorm.pdf", sep = ""), 4, 4, useDingbats = F)
+  par(mai = c(0.5, 1, 0.5, 0.25))
+  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, xlim = c(1, 4), 
+       ylim = c(1, 3.5))
+  abline(v = 2.5)
+  mtext(side = 3, c("Cool", "Hot"), at = c(1.625, 3.375))
+  mtext(side = 1, c("Wet", "Dry", "Wet", "Dry"), at = c(1, 2.25, 2.75, 4))
+  mtext(side = 1, "Water Treatment", font = 2, cex = 1.25, line = 1.5)
+  mtext(side = 3, "Temperature Treatment", font = 2, cex = 1.25, line = 1.5)
+  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+  title(ylab = expression(paste("Leaf expansion rate (mm ", day ^-1, ")")), cex.lab = 1.5)
+
+  for(i in 1:dim(tab)[1]) {
+    lines(c(1, 2.75), c(tab[i, "Wet", "Cool"], tab[i, "Wet", "Hot"]), col = "grey")
+    lines(c(2.25, 4), c(tab[i, "Dry", "Cool"], tab[i, "Dry", "Hot"]), col = "grey", lty = 2)
+  }
+
+  for(i in 1:dim(tab)[1]) {
+    lines(c(1, 2.25), c(tab[i, "Wet", "Cool"], tab[i, "Dry", "Cool"]), lty = 2)
+    lines(c(2.75, 4), c(tab[i, "Wet", "Hot"], tab[i, "Dry", "Hot"]))
+  }
+
+  mtext(side = 3, "A", font = 2, cex = 2, at = grconvertX(-0.1, "npc", "user"))
+  dev.off()
+
 @
@@
 
+	# Figure of reaction norms
+  tab <- tapply(newdata$SER_AbsGrowth, 
+                list(newdata$Population, newdata$WaterTrt, newdata$TempTrt), mean, na.rm = T)
+  pdf(paste(pathFig, "/FigureS_SERrxnNorm.pdf", sep = ""), 4, 4, useDingbats = F)
+  par(mai = c(0.5, 1, 0.5, 0.25))
+  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, xlim = c(1, 4), 
+       ylim = c(0.15, 0.50))
+  abline(v = 2.5)
+  mtext(side = 3, c("Cool", "Hot"), at = c(1.625, 3.375))
+  mtext(side = 1, c("Wet", "Dry", "Wet", "Dry"), at = c(1, 2.25, 2.75, 4))
+  mtext(side = 1, "Water Treatment", font = 2, cex = 1.25, line = 1.5)
+  mtext(side = 3, "Temperature Treatment", font = 2, cex = 1.25, line = 1.5)
+  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+  title(ylab = expression(paste("Stem elongation rate (cm ", day ^-1, ")")), cex.lab = 1.5)
+
+  for(i in 1:dim(tab)[1]) {
+    lines(c(1, 2.75), c(tab[i, "Wet", "Cool"], tab[i, "Wet", "Hot"]), col = "grey")
+    lines(c(2.25, 4), c(tab[i, "Dry", "Cool"], tab[i, "Dry", "Hot"]), col = "grey", lty = 2)
+  }
+
+  for(i in 1:dim(tab)[1]) {
+    lines(c(1, 2.25), c(tab[i, "Wet", "Cool"], tab[i, "Dry", "Cool"]), lty = 2)
+    lines(c(2.75, 4), c(tab[i, "Wet", "Hot"], tab[i, "Dry", "Hot"]))
+  }
+
+  mtext(side = 3, "B", font = 2, cex = 2, at = grconvertX(-0.1, "npc", "user"))
+
+  dev.off()
+
 @
@@
 We made all measurements at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$ of photosynthetically active radiation), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
-During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level could be measured under the same conditions.
-% As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out.
-Therefore, we calculated the `intrinsic' photosynthetic rate for each individual as its residual deviance from the regression between stomoatal conductance and photosynthetic rate across all plants.
-We also removed one outlier.
+During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to the average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$) so that all plants within a temperature level could be measured under the same conditions.
 
@@
 
-  # Intrinsic photosynthetic rate (not using at the moment)
-  # = residuals of Conductance - Photosynthesis regression
-  # Removed one outlier (may change if data are rearranged!)
-	# mm <- lmer(resPhoto ~ PopID * TempTrt * WaterTrt + (1|Family) + (1|Block),
-  #  data = subset(master, !is.na(master$Photo))[-52, ])
-	# fitIntrnPhoto <- step(mm, reduce.random = F) # only Pop
+  # Figure of reaction norms
+  tab <- tapply(master$Photo, 
+                list(master$Population, master$WaterTrt, master$TempTrt), mean, na.rm = T)
+  pdf(paste(pathFig, "/FigureS_PhotoRxnNorm.pdf", sep = ""), 4, 4, useDingbats = F)
+  par(mai = c(0.5, 1, 0.5, 0.25))
+  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, xlim = c(1, 4), 
+       ylim = c(5, 16))
+  abline(v = 2.5)
+  mtext(side = 3, c("Cool", "Hot"), at = c(1.625, 3.375))
+  mtext(side = 1, c("Wet", "Dry", "Wet", "Dry"), at = c(1, 2.25, 2.75, 4))
+  mtext(side = 1, "Water Treatment", font = 2, cex = 1.25, line = 1.5)
+  mtext(side = 3, "Temperature Treatment", font = 2, cex = 1.25, line = 1.5)
+  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+  title(ylab = expression(paste("Photosynthesis (", mu, "mol ", CO[2]~m^-2~s^-1, ")")), 
+        cex.lab = 1.5)
+
+  for(i in 1:dim(tab)[1]) {
+    lines(c(1, 2.75), c(tab[i, "Wet", "Cool"], tab[i, "Wet", "Hot"]), col = "grey")
+    lines(c(2.25, 4), c(tab[i, "Dry", "Cool"], tab[i, "Dry", "Hot"]), col = "grey", lty = 2)
+  }
+
+  for(i in 1:dim(tab)[1]) {
+    lines(c(1, 2.25), c(tab[i, "Wet", "Cool"], tab[i, "Dry", "Cool"]), lty = 2)
+    lines(c(2.75, 4), c(tab[i, "Wet", "Hot"], tab[i, "Dry", "Hot"]))
+  }
+
+  mtext(side = 3, "C", font = 2, cex = 2, at = grconvertX(-0.1, "npc", "user"))
+
+  dev.off()
 
@@
 
+  # Figure of reaction norms
+  d <- subset(master, !is.na(master$DiedFromStress))
+  tab <- tapply(d$DiedFromStress, list(d$PopID, d$WaterTrt, d$TempTrt), 
+                function(X) 1 - length(which(X)) / length(X))
+  pdf(paste(pathFig, "/FigureS_MortRxnNorm.pdf", sep = ""), 4, 4, useDingbats = F)
+  par(mai = c(0.5, 1, 0.5, 0.25))
+  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, xlim = c(1, 4), 
+       ylim = c(0, 1))
+  abline(v = 2.5)
+  mtext(side = 3, c("Cool", "Hot"), at = c(1.625, 3.375))
+  mtext(side = 1, c("Wet", "Dry", "Wet", "Dry"), at = c(1, 2.25, 2.75, 4))
+  mtext(side = 1, "Water Treatment", font = 2, cex = 1.25, line = 1.5)
+  mtext(side = 3, "Temperature Treatment", font = 2, cex = 1.25, line = 1.5)
+  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+  title(ylab = "Survivorship", cex.lab = 1.5)
+
+  for(i in 1:dim(tab)[1]) {
+    lines(c(1, 2.75), c(tab[i, "Wet", "Cool"], tab[i, "Wet", "Hot"]), col = "grey")
+    lines(c(2.25, 4), c(tab[i, "Dry", "Cool"], tab[i, "Dry", "Hot"]), col = "grey", lty = 2)
+  }
+
+  for(i in 1:dim(tab)[1]) {
+    lines(c(1, 2.25), c(tab[i, "Wet", "Cool"], tab[i, "Dry", "Cool"]), lty = 2)
+    lines(c(2.75, 4), c(tab[i, "Wet", "Hot"], tab[i, "Dry", "Hot"]))
+  }
+
+  mtext(side = 3, "D", font = 2, cex = 2, at = grconvertX(-0.1, "npc", "user"))
+
+  dev.off()
+
+  ##### TEMPORARY: CODE TO EXPORT FIGURE FOR TALK #####
+  d <- subset(master, !is.na(master$DiedFromStress))
+  x <- tapply(d$DiedFromStress, list(d$PopID, d$WaterTrt, d$TempTrt), function(X) 1 - length(which(X)) / length(X))
+
+  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/DroughtResponse.pdf", w = 5.5, h = 5, useDingbats = F)
+  par(mai = c(1, 1.5, 0.25, 0.25))
+
+  plot(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], axes = F, frame.plot = T,
+       pch = 21, col = "black", bg = primShade0, cex = 2, type = "n",
+       xlab = "Precipitation of Wettest Quarter (mm)", 
+       ylab = "Relative Survivorship\n(Hot/Dry : Hot/Wet)", cex.lab = 1.5)
+  points(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], pch = 21, col = "black", 
+         bg = primShade0, cex = 1.5, lwd = 2)
+  axis(1, lwd = 0, lwd.ticks = 1)
+  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+  dev.off()
+
 @
@@
 
-For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment, and Population $\times$ Treatment interactions.
+For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment (Temperature, Water, and Temperature $\times$ Water), and Population $\times$ Treatment interactions (Population $\times$ Temperature, Population $\times$ Water, and Population $\times$ Temperature $\times$ Water).
 We interpreted significant Population effects to indicate intrinsic variation and Population $\times$ Treatment interactions to indicate variation in plasticity.
@@
 We also included days to germination as a covariate in growth analyses.
-% To ensure that Population and Treatment effects were specific to a particular growth phase, we included earlier growth phases as a covariate.
-Specifically, we included germination day as a covariate in leaf expansion and stem elongation analyses; we included germination day and leaf expansion rate as a covariate in stem elongation analyses.
+To ensure that Population and Treatment effects were specific to a particular growth phase, we included germination day as a covariate in leaf expansion and stem elongation analyses.
 
@@
 The multivariate distribution of these coefficients was then summarized using principal components analysis.
-The first principal component of these traits (TraitPC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.
+The first principal component of these traits (TraitPC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis delineating fast to slow growth.
 
@@
     photo = fitPhoto$lsmeans.table[1:16, "Estimate"])
-  #  photo = fitIntrnPhoto$lsmeans.table[1:16, "Estimate"])
 
@@
 
+  ##### TEMPORARY VERSION FOR TALK #####
+  # Plot latitude versus PC1
+  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/Figure_PC1vLat.pdf", 5.5, 5, useDingbats = F)
+  par(mai = c(1, 1.5, 0.25, 0.25))
+  plot(focClim$Lat, focClim$PC1, xlim = c(32, 44), ylim = c(-4, 4), axes = F,
+    frame.plot = T, type = "n", xlab = "", ylab = "")
+  title(xlab = "Latitude", cex.lab = 1.5)
+  title(ylab = "TraitPC1", cex.lab = 2, line = 5)
+  title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
+    list(x = pve)))
+  axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), 
+		expression(36*degree), expression(40*degree), expression(44*degree)), 
+		lwd = 0, lwd.ticks = 1) 
+  axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
+  fitLat <- lm(PC1 ~ Latitude, data = focClim)
+
+	# Label y-axis from slow to fast growth
+	mtext(c("Slow\nGrowth", "Fast\nGrowth"), 2, at = c(-4, 4), cex = 1.25, 
+				font = 3, las = 1, line = 4, adj = 0.5)
+
+  # polygon of confidence intervals
+  x <- seq(min(focClim$Latitude), max(focClim$Latitude), l = 1e3)
+  u <- predict(fitLat, new = data.frame(Latitude = x))
+  s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitLat) %*% c(1, X)))
+  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
+  											upperCI = qnorm(0.975, u, s))
+  polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
+    col = "grey", border = NA)
+  lines(x, confInt$lowerCI, lwd = 2, lty = 2)
+  lines(x, confInt$upperCI, lwd = 2, lty = 2)
+  lines(range(focClim$Lat), rev(range(predict(fitLat))), lwd = 2)
+  points(focClim$Lat, focClim$PC1, pch = 21, col = "black", bg = primShade0, 
+  			 cex = 1.5, lwd = 2)
+
+  # P-value
+  text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1),
+	     labels = bquote(italic(P) == .(round(summary(fitLat)[[4]][2, 4], 3))))
+
+  dev.off()
+
 @
@@
 <<thinnedOccClimate, echo=FALSE, results=hide, warning=FALSE>>=
-# R data.frame of 178 M. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data
+# R data.frame of 178 E. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data
 load(paste(pathDat, "/thinnedOccClimate.RData", sep = ""))
+
 @
@@
 
-We found that a population's position along TraitPC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
+To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variable explains latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
+We found that a population's position along TraitPC1 correlated strongly with the latitude of origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{E.
 cardinalis}.
-Hypothesis 1: if a single climatic factor explained more trait variation than latitude alone, this would suggest that that factor is a selective agent underlying the latitudinal cline in our 16 focal populations.
-Hypothesis 2: if multiple climatic factors together explained trait variation,  we interpreted this as evidence that multiple climatic factors together have generated the latitudinal cline.
+We reasoned that if we identified a single climatic factor that explained more trait variation than latitude, then this would suggest that factor is a key selective agent underlying the latitudinal cline (Hypothesis 1).
+On the other hand, if multiple climatic factors together are necessary to explain trait variation, then this would suggest that many climatic factors together have imposed selection for the latitudinal cline (Hypothesis 2).
 We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables.
-In addition, to help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{M.
-cardinalis} occurrence records (see detail below), not just the 16 focial populations.
+To test Hypothesis 3 about climatic neighborhoods driving selection, we directly competed local with neighborhood climate.
+We used the immediate collection location for local climate.
+For climate neighborhoods, we sampled climate at 1000 random points (at 90-m resolution) within a 62-km radius buffer around the collection and took the average.
+We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
+Significant spatial autocorrelation persisted for approximately 62 km.
+Since \textit{E.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and variances (see below) were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
+In addition to competing local and neighborhood climate, we compared the univariate correlation between local and neighborhood climate with TraitPC1 and Latitude using paired $t$-tests.
+We adjusted degrees of freedom to account for the fact that many climatic factors are highly correlated and not independent.
+Specifically, we calculated the effective number of independent climatic factors ($M_\text{eff}$) using the formula $M_\text{eff} = 1 + (M - 1) (1 - \text{Var}(\lambda) / M)$ \citep{Chevrud_2001}, where $M$ is the original number of climatic factors and $\lambda$ are the eigenvalues of the correlation matrix of all climatic factors.
+
+To help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{E.
+cardinalis} occurrence records (see detail below), not just the 16 focal populations.
 We refer to these climatic factors as `Climate-Latitude' variables.
-The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{M.
+The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{E.
 cardinalis} range.
-Therefore, we did not consider Climate-TraitPC1 variables to be candidates selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
+Therefore, we did not consider Climate-TraitPC1 variables to be candidate selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
+
 We selected Climate-Latitude and Climate-TraitPC1 variables independently using Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}.
-VSURF ranks variables by their importance over regression trees in the forest.
+Random Forest regression is useful for cases like ours when the number of potential predictors is similar to or greater than the number of observations (`high $p$, low $n$' problem).
+VSURF is a multistip algorithm that progressively retains or eliminates variables based on their importance over regression trees in the forest.
+Variable importance is defined as the average amount a climate variable reduces mean-squared error in the predicted response (TraitPC1 or Latitude), compared to a randomly permuted dataset, across all trees in the random forest (see \citeauthor{Genuer_etal_2015} [\citeyear{Genuer_etal_2015}] for further detail).
+Hence, VSURF automatically eliminates unimportant and redundant variables based on the data without having to arbitrarily choose among colinear climate variables before the analysis.
 We kept only variables selected for prediction, the most stringent criterion.
 A visual overview of how we selected climatic variables is depicted in Fig~\ref{fig:Fig_VarSelect}.
-
-To test the third hypothesis about climatic neighborhoods driving selection, we directly competed local with neighborhood climate to see which correlated with traits (Climate-TraitPC1 variables) and latitude (Climate-Latitude variables) better.
-We used the immediate collection location for local climate.
-For climate neighborhoods, we sampled climate at 1000 random points (at 90-m resolution) within a 62-km radius buffer around the collection collection.
-We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
-Significant spatial autocorrelation persisted for approximately 62 km.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and variances (see below) were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
 
@@
 
-  # Load thinned occurences used in Climate-Lat analyses and convert to Spatial class
+  # Load thinned occurrences used in Climate-Lat analyses and convert to Spatial class
   occ <- read.csv(paste(pathDat, "/thinnedOccLatLon.csv", sep = ""), 
@@
 	\caption[Climatic variable selection]{Overview of method for identifying putative climatic selective agents underlying latitudinal cline.
-We looked for climate variables that explained both the latitude of \Sexpr{2 * nrow(thinnedOccClimate)} \textit{M.
-cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
-For Climate-Latitude variables we extracted climate data from recent occurences located throughout California and Oregon, USA (shown in map).
+We looked for climate variables that explained both the latitude of \Sexpr{2 * nrow(thinnedOccClimate)} \textit{E.
+cardinalis} occurrences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
+For Climate-Latitude variables we extracted climate data from recent occurrences located throughout California and Oregon, USA (shown in map).
 For Climate-Trait variables, we extracted climatic data for the 16 focal populations.
@@
 Local climate refers to climate only from where a population was collected ($x_{i,0}$).
-Neighborhood climate was calculated as the average over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M.
+Neighborhood climate was calculated as the average over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{E.
 cardinalis} is riparian.
-We identified climatic factors that most strongly predicted latitude of occurences (Climate-Latitude variables) and traits (Climate-Trait variables), as shown for hypothetical data in plots at the bottom of the figure.}
+We identified climatic factors that most strongly predicted latitude of occurrences (Climate-Latitude variables) and traits (Climate-Trait variables), as shown for hypothetical data in plots at the bottom of the figure.}
 	\label{fig:Fig_VarSelect}
@@
   # Load VSURF results of climatic variables associated with latitude at cardinalis occurence points
-  # Note that occurences have been randomly thinned to account for variation in sampling effort
+  # Note that occurrences have been randomly thinned to account for variation in sampling effort
   vsurfLat <- get(load(paste(pathDat, "/clim2lat_fit_t1.RData", sep = "")))
@@
          col = c("grey", "black"), bg = "white", bty = "n", xjust = 1, yjust = 1,
-         legend = c("All occurences", "Focal populations"))
+         legend = c("All occurrences", "Focal populations"))
 
@@
 
+##### Figure of local versus neighborhood Climate-TraitPC1 and Climate-Latitude correlations #####
+
+  pdf(paste(pathFig, "/FigureS_LocalvNeighCorr.pdf", sep = ""), w = 8, h = 4, useDingbats = F)
+  par(mai = c(1, 1, 0.2, 0.2), mfrow = c(1, 2))
+
+  # Panel A. Local versus neighborhood Climate-TraitPC1 correlations
+
+    # Vector of correlations between climVars and TraitPC1
+    r <- apply(focClim[, climVars], 2, cor, y = focClim$PC1)
+
+    # Matrix comparing correlation coefficients of local and neighborhood climate
+    corTab <- as.data.frame(cbind(r[!grepl("_SA", climVars)], r[grepl("_SA", climVars)]))
+    corTab <- abs(corTab)
+    colnames(corTab) <- c("local", "neighborhood")
+    corTab$type <- ifelse(grepl("_avg", rownames(corTab)), "avg", "cva")
+
+    plot(corTab[, "local"], corTab[, "neighborhood"], xlim = c(0, 1), ylim = c(0, 1),
+         xlab = "", ylab = "", type = "n", axes = F, frame.plot = T)
+    abline(0, 1, lty = 2)
+    axis(1, lwd = 0, lwd.ticks = 1)
+    axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+    title(xlab = expression(paste("local |", italic(r)[paste(climate-TraitPC1)], "|")),
+          ylab = expression(paste("neighborhood |", italic(r)[paste(climate-TraitPC1)], "|")),
+          cex.lab = 1.5)
+    points(corTab[, "local"], corTab[, "neighborhood"], pch = 21, col = "black",
+           bg = ifelse(corTab$type == "avg", "black", "white"))
+
+    legend(1, 0, xjust = 1, yjust = 0, legend = c("Avg.", "Var."), pch = 21, col = "black", 
+           pt.bg = c("black", "white"))
+    text(0.975, 0.95, labels = "1:1", pos = 1)
+    text(0 + strwidth("A", cex = 2, font = 2) / 2, 1 - strheight("A", cex = 2, font = 2) / 2, 
+         labels = "A", font = 2, cex = 2)
+
+    # Is neighborhood greater than local |r|?  Paired t-test
+    # Effective number of variables (Meff) from Chevrud et al 1983, 2001:
+    # Meff = 1 + (M - 1) * ((1 - Var(lambda)) / M)
+
+    mat <- cor(focClim[, climVars]) # Correlation matrix
+    M <- nrow(mat) # Correlation matrix rank
+    lambda <- eigen(mat)$values # Eigenvalues
+    Meff1 <- 1 + (M - 1) * (1 - var(lambda) / M) # Effective number of variables
+    tt1 <- t.test(corTab$local, corTab$neighborhood, paired = T) # paired t-test
+    pt1 <- pt(tt1$statistic, Meff1 / 2 - 1) * 2 # Corrected p-value
+
+  # Panel B. Local versus neighborhood Climate-Latitude correlations
+
+    # Vector of correlations between climVars and TraitPC1
+    r <- apply(thinnedOccClimate[, climVars], 2, cor, y = thinnedOccClimate$Latitude)
+
+    # Matrix comparing correlation coefficients of local and neighborhood climate
+    corTab <- as.data.frame(cbind(r[!grepl("_SA", climVars)], r[grepl("_SA", climVars)]))
+    corTab <- abs(corTab)
+    colnames(corTab) <- c("local", "neighborhood")
+    corTab$type <- ifelse(grepl("_avg", rownames(corTab)), "avg", "cva")
+
+    plot(corTab[, "local"], corTab[, "neighborhood"], xlim = c(0, 1), ylim = c(0, 1),
+         xlab = "", ylab = "", type = "n", axes = F, frame.plot = T)
+    abline(0, 1, lty = 2)
+    axis(1, lwd = 0, lwd.ticks = 1)
+    axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+    title(xlab = expression(paste("local |", italic(r)[paste(climate-latitude)], "|")),
+          ylab = expression(paste("neighborhood |", italic(r)[paste(climate-latitude)], "|")),
+          cex.lab = 1.5)
+    points(corTab[, "local"], corTab[, "neighborhood"], pch = 21, col = "black",
+           bg = ifelse(corTab$type == "avg", "black", "white"))
+
+    legend(1, 0, xjust = 1, yjust = 0, legend = c("Avg.", "Var."), pch = 21, col = "black", 
+           pt.bg = c("black", "white"))
+    text(0.975, 0.95, labels = "1:1", pos = 1)
+    text(0 + strwidth("B", cex = 2, font = 2) / 2, 1 - strheight("B", cex = 2, font = 2) / 2, 
+         labels = "B", font = 2, cex = 2)
+
+    # Is neighborhood greater than local |r|?  Paired t-test
+    # Effective number of variables (Meff) from Chevrud et al 1983, 2001:
+    # Meff = 1 + (M - 1) * ((1 - Var(lambda)) / M)
+
+    mat <- cor(thinnedOccClimate[, climVars]) # Correlation matrix
+    M <- nrow(mat) # Correlation matrix rank
+    lambda <- eigen(mat)$values # Eigenvalues
+    Meff2 <- 1 + (M - 1) * (1 - var(lambda) / M) # Effective number of variables
+    tt2 <- t.test(corTab$local, corTab$neighborhood, paired = T) # paired t-test
+    pt2 <- pt(tt2$statistic, Meff2 / 2 - 1) * 2 # Corrected p-value
+
+    dev.off()
+
+##### TEMPORARRY FIGURE FOR TALK #####
+
+path.root <- "~/Google Drive/CardAdapt" 
+setwd(path.root)
+source("Data/Climate/Rcode/1_Setup_directories.R")
+
+load(paste(path.obj, "/AllOcc_avg1981-2010.RData", sep = "")) # all.avg30
+load(paste(path.obj, "/AllOcc_cva1981-2010.RData", sep = "")) # all.cva30
+load(paste(path.obj, "/AllOccSA_avg1981-2010_62km.RData", sep = "")) # allSA.avg
+load(paste(path.obj, "/AllOccSA_cva1981-2010_62km.RData", sep = "")) # allSA.cva
+
+# Change column names for SA variables
+sharedCols <- c("Latitude", "Longitude", "Elevation", "PopID", "ID", "PopID")
+colnames(allSA.avg)[!colnames(allSA.avg) %in% sharedCols] <-
+  paste0(colnames(allSA.avg)[!colnames(allSA.avg) %in% sharedCols], "_SA")
+
+colnames(allSA.cva)[!colnames(allSA.cva) %in% sharedCols] <-
+  paste0(colnames(allSA.cva)[!colnames(allSA.cva) %in% sharedCols], "_SA")
+
+# Combine into single data.frame
+allClim <- cbind(all.avg30, all.cva30[rownames(all.avg30), !(colnames(all.cva30) %in% sharedCols)],
+                 allSA.avg[rownames(all.avg30), !(colnames(allSA.avg) %in% sharedCols)],
+                 allSA.cva[rownames(all.avg30), !(colnames(allSA.cva) %in% sharedCols)])
+
+  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/ClimateSpace.pdf", w = 5.5, h = 5, useDingbats = F)
+  par(mai = c(1, 1.5, 0.25, 0.25))
+
+  with(allClim, plot(bio10_avg - 273.16, bio16_avg, xlim = c(15, 27), ylim = c(100, 1500),
+                     xlab = "", ylab = "", axes = F, frame.plot = T, col = primShade1, pch = 19))
+  with(focClim, points(bio10_avg - 273.16, bio16_avg, cex = 2, col = "black", bg = primShade0, 
+                       lwd = 2, pch = 21))
+  axis(1, at = seq(16, 26, 2), labels = c(expression(16*degree), expression(18*degree), 
+                                          expression(20*degree), expression(22*degree), 
+                                          expression(24*degree), expression(26*degree)),
+		lwd = 0, lwd.ticks = 1) 
+  axis(2, las = 1, lwd = 0, lwd.ticks = 1)
+  title(xlab = "Temperature of Warmest Quarter", cex.lab = 1.5, line = 2.5)
+  title(xlab = expression(paste("(", degree, "C, 1981-2010 average)")), cex.lab = 1, line = 4)
+  title(ylab = "Precipitation of Wettest Quarter", line = 5, cex.lab = 1.5)
+  title(ylab = "(mm, 1981-2010 average)", cex.lab = 1, line = 4)
+
+  legend(grconvertX(0.96, "nfc", "user"), grconvertY(0.95, "nfc", "user"), xjust = 1, 
+         legend = c("all recorded populations", "focal populations"),
+         pch = c(19, 21), col = c(primShade1, "black"), pt.bg = primShade0, pt.cex = c(1, 2))
+dev.off()
+
+##### TEMPORARY FOR REVIEW RESPONSE #####
+  # This predicts that the difference between local and neighbourhood climate estimates should be lower 
+# for northern populations than all others
+
+# Separate into different datasets
+c1 <- !grepl("_SA", climVars)
+c2 <- grepl("_SA", climVars)
+
+dat1 <- thinnedOccClimate[, climVars[c1]]
+dat2 <- thinnedOccClimate[, climVars[c2]]
+
+# Standardize
+for (i in ncol(dat1)) {
+  x <- c(dat1[, i], dat2[, i])
+  x <- (x - mean(x)) / sd(x)
+  dat1[, i] <- x[1:nrow(dat1)]
+  dat2[, i] <- x[(nrow(dat1) + 1):(2 * nrow(dat1))]
+}
+
+# Calculate difference
+d12 <- dat1
+r <- numeric(ncol(d12))
+for (i in 1:ncol(d12)) {
+  d12[, i] <- abs(dat2[, i] - dat1[, i])
+  r[i] <- cor(thinnedOccClimate$Latitude, d12[, i])
+}
+hist(r[1:24])
+hist(r[25:48])
+summary(r)
+
+plot(thinnedOccClimate$Latitude, apply(d12, 1, function(X) sum(abs(X))))
+plot(thinnedOccClimate$Latitude, d12$bio12_avg)
+plot(thinnedOccClimate$Latitude, d12$bio12_cva)
+plot(thinnedOccClimate$Latitude, d12$bio1_avg)
+plot(thinnedOccClimate$Latitude, d12$bio1_cva)
+
 @
 
-For Climate-Latitude analyses, we compiled a representative set of \Sexpr{2 * nrow(thinnedOccClimate)} recent (since 2000) known \textit{M.
-cardinalis} occurences.
-These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+For Climate-Latitude analyses, we compiled a representative set of \Sexpr{2 * nrow(thinnedOccClimate)} recent (since 2000) known \textit{E.
+cardinalis} occurrences.
+These occurrences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
 For both Climate-TraitPC1 analyses (16 focal populations) and Climate-Latitude (many populations), we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
@@
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} version 1.1-1 \citep{Hijmans_etal_2016}.
-We included \Sexpr{length(climVars) / 4} climate factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+We included \Sexpr{length(climVars) / 4} climatic factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
 The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
+For each variable, we calculated both a 30-year normal by averaging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
 Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation (CV).
@@
 
-There are strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
+There are strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{E.
 cardinalis}, as evidenced by large and significant population effects for these traits (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured \Sexpr{gsub("%", "", pve)}\% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
+A single principal component captured \Sexpr{gsub("%", "", pve)}\% of the trait variation among populations, defining an axis of variation from fast to slow growth.
+A population's position along this axis strongly covaried with its latitude of origin; southern populations grew faster than northern populations (Fig~\ref{fig:Fig_PC1vLat}).
 As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.
-There were similar latitudinal clines for individual traits underlying PC1 (Fig S1-S4).
+There were similar latitudinal clines for individual traits underlying PC1 (\Cref{fig:FigS_GermLat,fig:FigS_LERlat,fig:FigS_SERlat,fig:FigS_Photolat}).
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% *** Commented out results from Intrinsic Photosynthetic rate *** %
 
@@
 Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation.
-We found little evidence G$\times$E in \textit{M.
+We found little evidence of G$\times$E in \textit{E.
 cardinalis}.
-There were only two statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but these were not strong compared to Population and Temperature effects.
+There were only two statistically significant Population $\times$ Treatment interactions (Table~\ref{table:Table_AnovaSummary}, Fig.~\ref{fig:FigS_RxnNorms}), but these were not strong compared to Population and Temperature effects.
 Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
@@
 However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
-Complete ANOVA tables are available in the Supporting Information (Table S3-S6).
+Complete ANOVA tables are available in the Supporting Information (\Cref{table:TableS_fitLER,table:TableS_fitSER,table:TableS_fitPhoto,table:TableS_fitMort})
 
@@
 
-Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
-To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variable explains latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
-Based on exploratory variable selection analyses, one or two aspects of interannual climatic variation (hypothesis 1) and climatic neighborhoods (hypothesis 3) are likely more important drivers of trait divergence than average climate immediately surrounding populations.
-These general conclusions are based on the fact that interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{M.
-cardinalis} occurences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
+Interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{E.
+cardinalis} occurrences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
 All \Sexpr{nImpLat} Climate-Latitude and \Sexpr{nImpPC1} Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
+In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:FigS_LocalvNeighCorr}).
+On average, neighborhood Climate-TraitPC1 correlation coefficients were \Sexpr{-round(tt1$estimate, 2)} higher than correlations with local-scale climate variables (paired \textit{t}-test, $t = \Sexpr{-round(tt1$statistic, 2)}$, d.f.
+= \Sexpr{round(Meff1 / 2 - 1, 1)}, $P = \Sexpr{round(pt1 * 10 ^ -floor(log10(pt1)), 2)} \times 10 ^ \Sexpr{floor(log10(pt1))}$).
+Likewise, neighborhood Climate-Latitude correlation coefficients were \Sexpr{-round(tt2$estimate, 2)} higher than those for local-scale climate (paired \textit{t}-test, $t = \Sexpr{-round(tt2$statistic, 2)}$, d.f.
+= \Sexpr{round(Meff2 / 2 - 1, 1)}, $P = \Sexpr{round(pt2 * 10 ^ -floor(log10(pt2)), 2)} \times 10 ^ \Sexpr{floor(log10(pt2))}$).
+
 The most important climatic variables related to the interannual variation rather than average climate.
-
-Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{M.
-cardinalis} (see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
+Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{E.
+cardinalis} (see Table~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
 Note that the coefficient of variation of a climatic factor is subscripted with $\sigma$ whereas the mean is subscripted with $\mu$.
-More specifically, greater winter precipitation variability and lower winter temperature variability are associted with Southern latitudes and higher TraitPC1 values (Fig.~\ref{fig:Fig_ClimPC1}A,B).
+More specifically, greater winter precipitation variability and lower winter temperature variability are associated with Southern latitudes and higher TraitPC1 values (Fig.~\ref{fig:Fig_ClimPC1}A,B).
 Neighborhood interannual variation in winter precipitation (bio16$_\sigma$) was the most important Climate-Latitude variable (Fig.~\ref{fig:Fig_ClimVarImp}A).
@@
 The most important Climate-TraitPC1 variable, neighborhood variation in mean diurnal range (bio2$_\sigma$; Fig.~\ref{fig:Fig_ClimVarImp}B) did not have any obvious similarity to Climate-Latitude variables.
-Given the large number of potential associations, we therefore think this may be a spuriously strong relationship.
+Given the large number of potential associations, we therefore think this may be a spuriously strong relationship and do not consider it further.
 
@@
 	\selectfont
-	\caption[Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
-cardinalis} and trait variation]{Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
+	\caption[Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{E.
+cardinalis} and trait variation]{Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{E.
 cardinalis} and trait variation.
 A.
-Using Random Forest regression, we identified \Sexpr{nImpLat} climatic variables significantly (high importance) associated with latitude of \textit{M.
-cardinalis} occurences.
+Using Random Forest regression, we identified \Sexpr{nImpLat} climatic variables significantly (high importance) associated with latitude of \textit{E.
+cardinalis} occurrences.
 B.
 Only one of of the most important Climate-Latitude variables (in bold) was among the most important Climate-TraitPC1 variables.
+Variable importance is defined as the average amount a climate variable reduces mean-squared error in the predicted response (TraitPC1 or Latitude), compared to a randomly permuted dataset, across all trees in the random forest (see \citeauthor{Genuer_etal_2015} [\citeyear{Genuer_etal_2015}] for further detail).
 Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
@@
 D.
-Southern populations of \textit{M.
+Southern populations of \textit{E.
 cardinalis} experience much greater interannual variationi in precipitation.
@@
 
-We found evidence for one of two common signatures of local adaptation in the perennial herb \textit{Mimulus cardinalis}.
-Latitudinal clines in germination rate, photosynthesis, and growth, suggest adaptive differentiation in fundamental physiological traits of the species.
+We found evidence for one of two common signatures of local adaptation in the perennial herb \textit{Erythranthe cardinalis}.
+Latitudinal clines in germination rate, photosynthesis, and growth suggest adaptive differentiation in important physiological traits of the species.
 However, we found little evidence that populations respond differently to temperature or drought.
-As we discuss below, this latter result may indicate that the fundamental abiotic niche is relatively conserved.
-Finally, we found that neighborhood climatic variation between years may be a more important selective agent than the local average climate.
+Due to low replication within families, we did not have power to assess within population genotype-by-environment interactions, which may be present.
+As we discuss below, low variation in plasticity among populations result may indicate that some dimensions of the fundamental abiotic niche are relatively conserved.
+Finally, our results suggest that neighborhood-scale climate and interannual variation are more important selective agents than local averages.
 In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in physiology changes clinally, whereas plastic responses to temperature and drought are relatively similar.
-
-Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001} and hence the type of variation maintained within species.
+One caveat to bear in mind is that we are limited by the size of the climate grid ($\approx 90~\text{m}^2$) and therefore unable to detect very fine-scale local adaptation.
+
+Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001, Lenormand_2002} and hence the type of variation maintained within species.
 Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strongly asymmetric, and/or maladaptive gene flow is too high.
-In \textit{M.
+To clarify, strong tradeoffs can prevent local adaptation in variable environments because selection favors habitat specialists that track a specific habitat regardless of its frequency in the environment \citep{Levins_1968}.
+For example, a riparian specialist may experience similar selection in rivers of highfall regions and deserts, even though the habitat is much rarer in the latter.
+In \textit{E.
 cardinalis} we found substantial genetically based variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
-If this variation is adaptive, it suggests that the fitness tradeoff between low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
+If this variation is adaptive, it suggests one of several possibilities to investigate in the future: fitness tradeoff between low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
 That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatially heterogenous selection.
@@
 Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range \citep{Paul_etal_2016}.
+
 Nevertheless, local gene flow from similar environments may shape how selection varies with latitude.
-Climatic neighborhoods, defined here as the average climate in a 62-km radius around populations, correlated more strongly with trait and population latitude than local climate.
-This may not greatly constrain local adaptation because local and neighborhood climate values were generally similar in \textit{M.
-cardinalis} populations (data not shown).
-
-Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years (see below), could have adapted by tolerating drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
-We found no evidence for this; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
+Theory predicts that populations will not be perfectly adapted to their immediate habitat when there is gene flow from surrounding populations with different optima \citep{Lenormand_2002}.
+With spatial heterogeneity and gene flow, traits will not covary perfectly with the local optimum \citep{Slatkin_1978, Paul_etal_2011, Hadfield_2016}, but should instead better match the average environment experienced by nearby populations connected through gene flow, which we refer to as the climatic neighborhood.
+Gene flow and spatial heterogeneity may therefore be important in maintaining genetic variation \citep{Yeaman_Jarvis_2006}.
+As this hypothesis predicts, climatic neighborhoods (62-km buffer around populations) correlated with traits and latitude of occurrences better than local climate (Fig.~\ref{fig:Fig_ClimVarImp}).
+We interpret this as suggestive evidence that gene flow between neighboring \textit{E.
+cardinalis} populations shapes selection -- populations are locally adapted to prevailing climate in their neighborhood, but perhaps not perfectly adapted to their local climate.
+This may not greatly constrain local adaptation because local and neighborhood climate values at the resolution of ClimateWNA (90 m$^2$) were generally similar in \textit{E.
+cardinalis} populations (Fig.~\ref{fig:FigS_LocalvNeighCorr}).
+Therefore, we would predict in reciprocal transplants that populations whose local climate is farther from their neighborhood average would be less well adapted than those close to their neighborhood average.
+
+It is reasonable to predict that southern populations, which appear to experience more frequent drought years (see below), might have physiological adaptation to survive and grow in drier soil.
+We found no evidence for this type of drought tolerance; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
 Plants grew faster in the Hot treatment, but there was little effect of drought on growth.
-Rather, the effects of drought took longer to materialize but resulted in high mortality, especially in the Hot chamber.
+Rather, the effects of drought took longer to materialize but resulted in high mortality, especially in the Hot treatment.
 However, there was no differential mortality among populations in this treatment.
-We hypothesize that evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
-Riparian habitats where \textit{M.
-cardinalis} live are highly heterogeneous at small spatial scales.
+Although our results indicate that this axis of the species niche may be constrained, plants have multiple ways to resist drought through both tolerance and escape \citep{Ludlow_1989, Kooyers_2015}.
+Next, we consider why drought tolerance may less important in local adaptation than a form of escape for this species.
+
+We hypothesize that tolerance to dry soil may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
+Soil moisture in riparian habitats where \textit{E.
+cardinalis} live are highly heterogeneous at very small spatial scales (several meters).
 Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Mediterranean climates of western North America.
-But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and \textit{vice versa}, leading to strong fitness tradeoffs.
-Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
+We hypothesize alleles that confer greater drought tolerance may be quite costly in well-watered soils, and \textit{vice versa}, leading to strong fitness tradeoffs.
+Such tradeoffs would promote specialization to one soil moisture or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
 Demography and gene flow may reinforce niche conservatism.
-A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher-resource environments.
-Infrequent wet years may also produce most seeds, so selection is weighted towards alleles that have high fitness in the wet environment, even if dry years are more frequent.
-Finally, gene flow, which is generally high among \textit{M.
+A new mutant with increased drought tolerance that could survive at the resource-poor margin of a population would likely be demographically overwhelmed by the larger census populations that can be maintained in higher-resource environments.
+Infrequent wet years may also produce most seeds, so selection is weighted towards alleles that have high fitness in the wet environment, even if dry years are more frequent \citep{Templeton_Levin_1979, Brown_Venable_1986}.
+Finally, gene flow, which is generally high among \textit{E.
 cardinalis} populations within the same ecoregion \citep{Paul_etal_2016}, will thwart local adaptation and reinforce specialization.
-Thus, the spatial grain of the environment, demographic asymmetry, and gene flow may conspire to constrain local adaptation via altered fundamental niche.
-
-To reiterate, these results indicate that differences in physiology and growth, but not responses to temperature and drought, mediate local adaptation to climate in \textit{M.
+Thus, the spatial grain of the environment, demographic asymmetry, and gene flow may conspire to constrain local adaptation along this environmental axis.
+Consistent with this hypothesis, recent record-setting droughts have caused the decline or even local extinction of some natural populations of \textit{E.
+cardinalis} \citep{Sheth_Angert_2017}.
+
+In sum, these results indicate that differences in physiology and growth, but not responses to temperature and drought, mediate local adaptation to climate in \textit{E.
 cardinalis}.
 Next, we would like to understand why variation in these particular traits may be adaptive.
-We argue that temporally more variable environments, as experienced by southern populations, select for a more `annualized` life-history strategy.
+We argue that temporally more variable environments, as experienced by southern populations, select for a more `annualized' life-history strategy, a form of drought escape.
+Demographic observations in natural populations of \textit{E.
+cardinalis} reveal that southern populations tend to flower earlier at a smaller size, while northern populations invest more in vegetative growth \citep{Sheth_Angert_2017}.
+The association between position along the `fast-slow' continuum and associated traits in \textit{E.
+cardinalis} is similar to interspecific relationships between growth, functional traits, and life history \citep{Adler_etal_2014, Salguero-Gomez_etal_2016}.
+However, we cannot exclude unexplored factors (e.g.
+edaphic conditions, competitors, pollinators, etc.) which may drive the latitudinal cline.
 
 Greater investment in aboveground growth, as opposed to belowground storage for future seasons, may be favoured in climates with more frequent drought years, but maladaptive in climates with more consistent precipitation.
+This is a form of drought escape in that plants are investing more reproduction in the present to avoid possible drought in subsequent years.
 Suppose plants that grow quickly and allocate new resources to continued growth rather than storage have higher fitness over a single growing season.
@@
 If drought years occur frequently enough, selection may favour the fast-growing strategy because there is no advantage to storage if drought kills plants before flowering.
-Considering life-history strategy as a continuum from no storage (annual) to lots of storage (perennial), we hypothesize that the optimal allocation to aboveground growth is more `annualized` in southern climates that have greater interannual variation in precipitation.
-
-The hypothesis that greater precipitation variability selects for an annualized life history is tentative, but consistent with theory, data from other species, and other observations in \textit{M.
-cardinalis}.
+Considering life-history strategy as a continuum from no storage (annual) to lots of storage (perennial), we hypothesize that the optimal allocation to aboveground growth is more `annualized' in southern climates that have greater interannual variation in precipitation.
+This scenario differs from classic drought escape syndromes in which plants speed up development early in the season before the onset of drought.
+
+The hypothesis that greater precipitation variability selects for an annualized life history is tentative, but consistent with theory and data from other species.
 Life history theory shows that less variable environments are one factor that favours the evolution of perenniality \citep{Stearns_1976, Iwasa_Cohen_1989, Friedman_Rubin_2015}.
 Populations of the perennial \textit{Plantago asiatica} show a similar latitudinal cline in growth and allocation to storage \citep{Sawada_etal_1994}, though these authors attribute the cline to variation in growing season length.
-There are also life history clines in the closely related species \textit{M.
-guttatus}, but the underlying traits and climatic drivers are quite different.
-Annual \textit{M.
-guttatus} flower sooner and produce fewer stolons in response to climates with shorter seasons and more intense summer drought \citep{Lowry_Willis_2010, Friedman_etal_2015, Kooyers_etal_2015}.
-In contrast, there are no truly annual (monocarpic and semelparous) populations of \textit{M.
+There are also life history clines in the closely related species \textit{E.
+guttata}, but the underlying traits and climatic drivers are quite different.
+Annual \textit{E.
+guttata} flower sooner and produce fewer stolons in response to climates with shorter seasons and more intense summer drought \citep{Lowry_Willis_2010, Friedman_etal_2015, Kooyers_etal_2015}.
+In contrast, there are no truly annual (monocarpic and semelparous) populations of \textit{E.
 cardinalis}.
-Rather, our hypothesis states climatic variability selects on quantitative variation in allocation to growth versus storage.
-Several lines of evidence provide tentative support for this hypothesis.
-Preliminary surveys suggest that northern populations not only grow slower, but also produce greater numbers of rhizomes (C.D.
-Muir, unpub.
-data), suggesting an allocation tradeoff.
-Ecological niche models also show that occurrence of southern populations is best predicted by recent climate (< 5 years), whereas northern occurrences are best predicted by climate over the previous 30 years (M.
-Bayly \& A.
-Angert, unpub.
-data).
-Finally, demographic surveys of natural populations show greater variation in the size of recruits in southern populations, suggesting higher maximum growth rates under natural conditions (M.
-Bayly \& A.
-Angert, unpub.
-data).
+Rather, our hypothesis states that climatic variability selects on quantitative variation in allocation to growth versus storage.
+This hypothesis makes several independent, testable predictions.
+The allocation tradeoff predicts that northern populations will provision more assimilate to rhizomes compared with southern populations.
+If southern populations are indeed more `annualized' because more frequent droughts cause mortality, then we predict that species distribution models using recent climate would best predict occurrences in the south, whereas longer term climate would be a better predictor in the north.
+Finally, we predict that southern populations would show greater variation in the size of recruits and higher maximum population growth rates.
 Exploratory analysis also pointed to variation in winter temperature as a potential selective agent.
@@
 In summary, we found evidence for a coordinated latitudinal cline in germination rate, photosynthesis, and growth, suggesting local adaptation.
-These differences between populations suggest different trait optima in different climates.
+We therefore predict to find different optima for  these traits in different climates.
 We did not find evidence that the relative performance of populations shifts with temperature or watering regime, suggesting relatively little variation in plasticity.
 Exploratory analysis implicate that more variable precipitation regimes at lower latitude drive much of the latitudinal cline, though other climatic factors could also contribute.
-Interestingly, the climatic neighborhood may shape selective pressures than local climate.
+Interestingly, the climatic neighborhood may shape selective pressures more than local climate.
 In the future, we will use field experiments to test whether greater variation in precipitation selects for faster growth and that selection on temperature/drought responses does not vary among populations.
 By doing so, we aim to understand why certain physiological and developmental mechanisms, but not others, contribute to local adaptation.
-
-% [Future work and broad conclusions].
-Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
-Next, we are testing whether phenotypic variation and constraint can be predicted by the shape of tradeoffs, as predicted by evolutionary theory.
-\cdm{I will work more on this once I am more comfortable with the rest of the discussion}.
-
-% niche-tracking versus niche-shifting
-% One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve.
-For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.
-
-% citations on life history allocation tradeoff between growth and reproduction:
-%% THEORY:
-% Bell 1980 - briefly mentionds on page 56 that increased environmental variation for adults will favor semelparity
-% Iwasa and Cohen 1989, pg.
-496 "A larger reliability of habitat, a higher storage efficiency, higher productivity of the environment, and a longer growing season, all tend to promote the evolution of perennial life, as suggested by previous models of life-historsy strategy (Stearns 1976).
-
-%% EMPIRICAL:
-% Sawada et al. 1994 - latitudinal cline dry matter accumulation and storage allocation
-% Stanton et al 2000 show selection in stressful environemts (artificual selection over 5 generations) favors stress avoidance rather than tolerance traits.
-i.e.
-might be easier to adjust life history rather than tolerance.
-% Hall and Willis 2006 find selection on flowering time in coastal-inland comparison of M guttatus
-% Ehleringer et al 1990 find lower WUE increased yield in ephemeral pops
-% Bazzaz 1979: higher photo in fast growing plants (early succession)
-
-% Reznick 1985; Lovett Doust 1989
-% Williams 1966; Charnov and Schaffer 1973; Bell 1980; Reznick 1985; Biere 1995; Johnson 2007).
-% for iteroparous sp: van Noordwijk and De Jong 1986.
-% from Remington et al.
-2013: "Genetic variation in traits subject to trade-offs results in what has been termed “structured pleiotropy,” in which genetic covariances between traits result from functional constraints imposed by the limiting resources (De Jong 1990; Stearns et al.
-1991; see Figure 1B).
-
-%--------------------------------------------------
-% Acknowledgements
-%--------------------------------------------------
-
-\section*{Acknowledgements}
-Erin Warkman and Lisa Lin helped collect data.
-CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program.
-ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
 
@@
 	\selectfont
-  \caption[Important climatic variables]{Important climatic variables predicting latitude of \textit{M.
+  \caption[Important climatic variables]{Important climatic variables predicting latitude of \textit{E.
 cardinalis} populations (`Climate-Latitude') and the first principal component of traits measured in a common garden (`Climate-TraitPC1').
@@
 	text(grconvertX(0.05, "npc"), grconvertY(0.95, "npc"), adj = c(0, 1),
-	     labels = bquote(italic(P) == .(round(summary(fit)[[4]][2, 4], 3))),)
+	     labels = bquote(italic(P) == .(round(summary(fit)[[4]][2, 4], 3))))
 
@@
 	\caption[Southern populations germinate faster]{Southern populations germinate faster.
-Each point is a population of \textit{M.
+Each point is a population of \textit{E.
 cardinalis} showing its latitude of origin (x-axis) and model-predicted days to germination in days under growth chamber conditions (see Material and Methods).
@@
 	\caption[Southern populations grow faster (leaf expansion rate).]{Southern populations grow faster.
-Each point is a population of \textit{M.
+Each point is a population of \textit{E.
 cardinalis} showing its latitude of origin (x-axis) and model-predicted leaf expansion rate during the rosette phase.
@@
 	\caption[Southern populations grow faster (stem elongation rate).]{Southern populations grow faster.
-Each point is a population of \textit{M.
+Each point is a population of \textit{E.
 cardinalis} showing its latitude of origin (x-axis) and model-predicted stem elongation rate.
@@
 	\selectfont
-	\caption[Southern populations photosynthesize faster.]{Southern populations photosynthesize faster.
-Each point is a population of \textit{M.
+	\caption[Southern populations photosynthesize faster]{Southern populations photosynthesize faster.
+Each point is a population of \textit{E.
 cardinalis} showing its latitude of origin (x-axis) and model-predicted instantaneous photosynthetic rate.
@@
 	\label{fig:FigS_Photolat}
+\end{figure}
+
+% Reaction norms of LER, SER, Photo, and Mort
+
+\begin{figure}
+\centering
+    \begin{subfigure}{.4\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{Figures/FigureS_LERrxnNorm.pdf}
+        \phantomcaption{}\label{fig:FigS_LERrxnNorm}
+    \end{subfigure} %
+    \begin{subfigure}{.4\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{Figures/FigureS_SERrxnNorm.pdf}
+        \phantomcaption{}\label{fig:FigS_SERrxnNorm}
+    \end{subfigure} %
+    \begin{subfigure}{0.4\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{Figures/FigureS_PhotoRxnNorm.pdf}
+        \phantomcaption{}\label{fig:FigS_PhotoRxnNorm}
+    \end{subfigure}
+    \begin{subfigure}{0.4\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{Figures/FigureS_MortRxnNorm.pdf}
+        \phantomcaption{}\label{fig:FigS_MortRxnNorm}
+    \end{subfigure}
+\caption[Reaction norms]{Reaction norms signify little Population $\times$ Treatment interactions.
+For all panels, black lines represent population-level reaction norms from Wet to Dry in the Cool temperature treatment (dashed black lines) and Hot temperature treatment (solid black lines); gray lines represent reaction norms from Cool to Hot in the Wet treatment (solid gray lines) and Dry treatment (dashed gray lines).
+The responses shown are (A) leaf expansion rate, (B) stem elongation rate, (C) photosynthesis, and (D) survivorship (= 1 - mortality).}
+	\label{fig:FigS_RxnNorms}
+\end{figure}
+
+% Figure of local versus neighborhood Climate-TraitPC1 and Climate-Latitude correlations
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=1\textwidth]{Figures/FigureS_LocalvNeighCorr.pdf}}
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\caption[Local versus neighborhood correlations]{Neighborhood climate predicts TraitPC1 (`Climate-trait', panel A) and Latitude of occurences (`Climate-latitude', panel B) better than local climate.
+Each point is the absolute value of the Pearson correlation coefficient (|$r$|) between TraitPC1 (A) or latitude (B) for \Sexpr{length(climVars)/4} climatic factors, for which we used both the 30-year mean (closed circles) and coefficient of variation (open circles).
+Most points lie above the 1:1 line, indicating stronger correlations with neighborhood compared to local climate.
+Neighborhood climate was integrated over a 62-km radius around focal populations (see text for further detail).}
+\label{fig:FigS_LocalvNeighCorr}
 \end{figure}
@@
 data).
-Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M.
+Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{E.
 cardinalis} in nature.
@@
 
-%*** need to check chiller temp and make of chiller
-For watering treatments, we simulated two extreme types of streams where \textit{M.
+For watering treatments, we simulated two extreme types of streams where \textit{E.
 cardinalis} grows.

### 2017-03-08 ms/ms.tex
Revised ms

@@
 
-We used 16 populations from throughout the range of \textit{M.
+We used 16 populations from throughout the range of \textit{E.
 cardinalis} (Table~\ref{table:Table_FocalPops}).
+These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below).
 Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
+We used seeds from 154 families, 4--12 (mean = 9.6, median = 12) families per population.
 
@@
 The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
+Lower humidity would have made the drought more severe,but low soil moisture is stressful in and of itself.
+The total number of plants in each treatment was: $n_{\text{cool,dry}} = 169$; $n_{\text{cool,ww}} = 174$; $n_{\text{hot,dry}} = 176$; $n_{\text{hot,ww}} = 183$.
+Each population had 8--12 individuals per treatment level (mean = 11, median = 11).
 
@@
 
-We found that a population's position along TraitPC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
+Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
+To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variable explains latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
+We found that a population's position along TraitPC1 correlated strongly with the latitude of origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{E.
 cardinalis}.
-Hypothesis 1: if a single climatic factor explained more trait variation than latitude alone, this would suggest that that factor is a selective agent underlying the latitudinal cline in our 16 focal populations.
-Hypothesis 2: if multiple climatic factors together explained trait variation,  we interpreted this as evidence that multiple climatic factors together have generated the latitudinal cline.
+We reasoned that if we identified a single climatic factor that explained more trait variation than latitude, then this would suggest that factor is a key selective agent underlying the latitudinal cline (Hypothesis 1).
+On the other hand, if multiple climatic factors together are necessary to explain trait variation, then this would suggest that many climatic factors together have imposed selection for the latitudinal cline (Hypothesis 2).
 We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables.
-In addition, to help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{M.
-cardinalis} occurrence records (see detail below), not just the 16 focial populations.
+To test Hypothesis 3 about climatic neighborhoods driving selection, we directly competed local with neighborhood climate.
+We used the immediate collection location for local climate.
+For climate neighborhoods, we sampled climate at 1000 random points (at 90-m resolution) within a 62-km radius buffer around the collection and took the average.
+We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
+Significant spatial autocorrelation persisted for approximately 62 km.
+Since \textit{E.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and variances (see below) were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
+In addition to competing local and neighborhood climate, we compared the univariate correlation between local and neighborhood climate with TraitPC1 and Latitude using paired $t$-tests.
+We adjusted degrees of freedom to account for the fact that many climatic factors are highly correlated and not independent.
+Specifically, we calculated the effective number of independent climatic factors ($M_\text{eff}$) using the formula $M_\text{eff} = 1 + (M - 1) (1 - \text{Var}(\lambda) / M)$ \citep{Chevrud_2001}, where $M$ is the original number of climatic factors and $\lambda$ are the eigenvalues of the correlation matrix of all climatic factors.
+
+To help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{E.
+cardinalis} occurrence records (see detail below), not just the 16 focal populations.
 We refer to these climatic factors as `Climate-Latitude' variables.
-The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{M.
+The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{E.
 cardinalis} range.
-Therefore, we did not consider Climate-TraitPC1 variables to be candidates selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
+Therefore, we did not consider Climate-TraitPC1 variables to be candidate selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
+
 We selected Climate-Latitude and Climate-TraitPC1 variables independently using Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}.
-VSURF ranks variables by their importance over regression trees in the forest.
+Random Forest regression is useful for cases like ours when the number of potential predictors is similar to or greater than the number of observations (`high $p$, low $n$' problem).
+VSURF is a multistip algorithm that progressively retains or eliminates variables based on their importance over regression trees in the forest.
+Variable importance is defined as the average amount a climate variable reduces mean-squared error in the predicted response (TraitPC1 or Latitude), compared to a randomly permuted dataset, across all trees in the random forest (see \citeauthor{Genuer_etal_2015} [\citeyear{Genuer_etal_2015}] for further detail).
+Hence, VSURF automatically eliminates unimportant and redundant variables based on the data without having to arbitrarily choose among colinear climate variables before the analysis.
 We kept only variables selected for prediction, the most stringent criterion.
@@
 
-To test the third hypothesis about climatic neighborhoods driving selection, we directly competed local with neighborhood climate to see which correlated with traits (Climate-TraitPC1 variables) and latitude (Climate-Latitude variables) better.
-We used the immediate collection location for local climate.
-For climate neighborhoods, we sampled climate at 1000 random points (at 90-m resolution) within a 62-km radius buffer around the collection collection.
-We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
-Significant spatial autocorrelation persisted for approximately 62 km.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and variances (see below) were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 	\caption[Climatic variable selection]{Overview of method for identifying putative climatic selective agents underlying latitudinal cline.
-We looked for climate variables that explained both the latitude of 356 \textit{M.
-cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
-For Climate-Latitude variables we extracted climate data from recent occurences located throughout California and Oregon, USA (shown in map).
+We looked for climate variables that explained both the latitude of 356 \textit{E.
+cardinalis} occurrences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
+For Climate-Latitude variables we extracted climate data from recent occurrences located throughout California and Oregon, USA (shown in map).
 For Climate-Trait variables, we extracted climatic data for the 16 focal populations.
@@
 
-For Climate-Latitude analyses, we compiled a representative set of 356 recent (since 2000) known \textit{M.
-cardinalis} occurences.
-These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+For Climate-Latitude analyses, we compiled a representative set of 356 recent (since 2000) known \textit{E.
+cardinalis} occurrences.
+These occurrences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
 For both Climate-TraitPC1 analyses (16 focal populations) and Climate-Latitude (many populations), we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
@@
 For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} version 1.1-1 \citep{Hijmans_etal_2016}.
-We included 24 climate factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
+We included 24 climatic factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
 The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
+For each variable, we calculated both a 30-year normal by averaging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
 Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation (CV).
@@
 
-There are strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
+There are strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{E.
 cardinalis}, as evidenced by large and significant population effects for these traits (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured 71.6 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
+A single principal component captured 71.6 \% of the trait variation among populations, defining an axis of variation from fast to slow growth.
+A population's position along this axis strongly covaried with its latitude of origin; southern populations grew faster than northern populations (Fig~\ref{fig:Fig_PC1vLat}).
 As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.
-There were similar latitudinal clines for individual traits underlying PC1 (Fig S1-S4).
+There were similar latitudinal clines for individual traits underlying PC1 (\Cref{fig:FigS_GermLat,fig:FigS_LERlat,fig:FigS_SERlat,fig:FigS_Photolat}).
 
@@
 
-Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
-To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variable explains latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
-Based on exploratory variable selection analyses, one or two aspects of interannual climatic variation (hypothesis 1) and climatic neighborhoods (hypothesis 3) are likely more important drivers of trait divergence than average climate immediately surrounding populations.
-These general conclusions are based on the fact that interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{M.
-cardinalis} occurences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
+Interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{E.
+cardinalis} occurrences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
 All 16 Climate-Latitude and 3 Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
+In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:FigS_LocalvNeighCorr}).
+On average, neighborhood Climate-TraitPC1 correlation coefficients were 0.16 higher than correlations with local-scale climate variables (paired \textit{t}-test, $t = 7.87$, d.f.
+= 33.6, $P = 3.94 \times 10 ^ -9$).
+Likewise, neighborhood Climate-Latitude correlation coefficients were 0.13 higher than those for local-scale climate (paired \textit{t}-test, $t = 6.71$, d.f.
+= 36.8, $P = 7.22 \times 10 ^ -8$).
+
 The most important climatic variables related to the interannual variation rather than average climate.
-
-Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{M.
-cardinalis} (see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
+Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{E.
+cardinalis} (see Table~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
 Note that the coefficient of variation of a climatic factor is subscripted with $\sigma$ whereas the mean is subscripted with $\mu$.
-More specifically, greater winter precipitation variability and lower winter temperature variability are associted with Southern latitudes and higher TraitPC1 values (Fig.~\ref{fig:Fig_ClimPC1}A,B).
+More specifically, greater winter precipitation variability and lower winter temperature variability are associated with Southern latitudes and higher TraitPC1 values (Fig.~\ref{fig:Fig_ClimPC1}A,B).
 Neighborhood interannual variation in winter precipitation (bio16$_\sigma$) was the most important Climate-Latitude variable (Fig.~\ref{fig:Fig_ClimVarImp}A).
@@
 	\selectfont
-	\caption[Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
-cardinalis} and trait variation]{Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
+	\caption[Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{E.
+cardinalis} and trait variation]{Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{E.
 cardinalis} and trait variation.
 A.
-Using Random Forest regression, we identified 16 climatic variables significantly (high importance) associated with latitude of \textit{M.
-cardinalis} occurences.
+Using Random Forest regression, we identified 16 climatic variables significantly (high importance) associated with latitude of \textit{E.
+cardinalis} occurrences.
 B.
 Only one of of the most important Climate-Latitude variables (in bold) was among the most important Climate-TraitPC1 variables.
+Variable importance is defined as the average amount a climate variable reduces mean-squared error in the predicted response (TraitPC1 or Latitude), compared to a randomly permuted dataset, across all trees in the random forest (see \citeauthor{Genuer_etal_2015} [\citeyear{Genuer_etal_2015}] for further detail).
 Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
@@
 
+% Reaction norms of LER, SER, Photo, and Mort
+
+\begin{figure}
+\centering
+    \begin{subfigure}{.4\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{Figures/FigureS_LERrxnNorm.pdf}
+        \phantomcaption{}\label{fig:FigS_LERrxnNorm}
+    \end{subfigure} %
+    \begin{subfigure}{.4\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{Figures/FigureS_SERrxnNorm.pdf}
+        \phantomcaption{}\label{fig:FigS_SERrxnNorm}
+    \end{subfigure} %
+    \begin{subfigure}{0.4\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{Figures/FigureS_PhotoRxnNorm.pdf}
+        \phantomcaption{}\label{fig:FigS_PhotoRxnNorm}
+    \end{subfigure}
+    \begin{subfigure}{0.4\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{Figures/FigureS_MortRxnNorm.pdf}
+        \phantomcaption{}\label{fig:FigS_MortRxnNorm}
+    \end{subfigure}
+\caption[Reaction norms]{Reaction norms signify little Population $\times$ Treatment interactions.
+For all panels, black lines represent population-level reaction norms from Wet to Dry in the Cool temperature treatment (dashed black lines) and Hot temperature treatment (solid black lines); gray lines represent reaction norms from Cool to Hot in the Wet treatment (solid gray lines) and Dry treatment (dashed gray lines).
+The responses shown are (A) leaf expansion rate, (B) stem elongation rate, (C) photosynthesis, and (D) survivorship (= 1 - mortality).}
+	\label{fig:FigS_RxnNorms}
+\end{figure}
+
+% Figure of local versus neighborhood Climate-TraitPC1 and Climate-Latitude correlations
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=1\textwidth]{Figures/FigureS_LocalvNeighCorr.pdf}}
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\caption[Local versus neighborhood correlations]{Neighborhood climate predicts TraitPC1 (`Climate-trait', panel A) and Latitude of occurences (`Climate-latitude', panel B) better than local climate.
+Each point is the absolute value of the Pearson correlation coefficient (|$r$|) between TraitPC1 (A) or latitude (B) for 24 climatic factors, for which we used both the 30-year mean (closed circles) and coefficient of variation (open circles).
+Most points lie above the 1:1 line, indicating stronger correlations with neighborhood compared to local climate.
+Neighborhood climate was integrated over a 62-km radius around focal populations (see text for further detail).}
+\label{fig:FigS_LocalvNeighCorr}
+\end{figure}
+
 % Figure of bio11_cva_SA versus TraitPC1

### 2017-05-30 ms/ms.Rnw
revised ms

@@
 
-Key words: local adaptation, cline, photosynthesis, growth rate, \textit{Erythranthe} \\
+Key words: local adaptation, cline, photosynthesis, growth rate, monkeyflower, \textit{Mimulus} \\
 
@@
 ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
+Two anonymous referees provided constructive comments on earlier versions of this manuscript.
 
@@
 Local adaptation is commonly observed in nature: organisms perform well in their natal environment, but poorly outside it.
-Correlation between traits and latitude, or latitudinal clines, are among the most common pieces of evidence for local adaptation, but identifying the traits under selection and the selective agents are challenging.
+Correlations between traits and latitude, or latitudinal clines, are among the most common pieces of evidence for local adaptation, but identifying the traits under selection and the selective agents is challenging.
 Here, we investigated a latitudinal cline in growth and photosynthesis across 16 populations of the perennial herb \textit{Erythranthe cardinalis} (Phrymaceae).
@@
 % Local adaptation and clines
-Local adaptation within species has been documented in numerous species; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Leimu_Fischer_2008, Hereford_2009}.
+Local adaptation has been documented within numerous species; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Leimu_Fischer_2008, Hereford_2009}.
 However, the prevalance of local adaptation remains difficult to assess because researchers rarely test for local adaptation unless there are obvious phenotypic or environmental differences (but see \citeauthor{Hereford_Winn_2008} \citeyear{Hereford_Winn_2008}).
 When local adaptation occurs, it frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977, Barton_1999}.
-Phenotypic differences between populations along a cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
+Phenotypic differences between populations along a cline often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
 Despite a long history of studying local adaptation and clines, it remains challenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
@@
 coastal vs.
-continental), dissecting the evolution of latitudinal clines across many species will help identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local versus regional climate shape selective pressures.
+continental), dissecting the evolution of latitudinal clines across many species will help identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, and whether local or regional climate shapes selective pressures.
 
@@
 These hypotheses are not mutually exclusive since, for example, single or multiple factors in a climatic neighborhood may lead to latitudinal clines.
-We focused on climate because climate often determines and where species are found and also can exert strong selection on populations within species [CITATION?].
-There is also a compelling need to know how populations are (or are not) locally adapted to climate so as to predict how they’ll respond to climate change [CITATION?].
-We did not test for local adaptation to other environmental characteristicss, such as edaphic or biotic (e.g.
-competitors, pollinators) factors, because we did not have compelling observations from which to make testable hypotheses about the latter in our focal species (described below).
+We focused on climate because climate often determines and where species are found and also can exert strong selection on populations within species, though we acknowledge that other abiotic and biotic factors could also contribute to selection and the overall pattern of local adaptation.
+There is also a compelling need to know how populations are (or are not) locally adapted to climate so as to predict how they will respond to climate change \citep{Aitken_Whitlock_2013}.
 
 We examined these questions in \textit{Erythranthe cardinalis} (formerly \textit{Mimulus cardinalis} [\citeauthor{Nesom_2014} \citeyear{Nesom_2014}]) because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
-Many classic and contemporary studies of local adaptation use \textit{Mimulus sensu lato} species because of its natural history, easy propagation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
-Yet, there is a conspicuous deficiency of links between local adaptation and physiological mechanisms (\cite{Angert_2006, Angert_etal_2008, Wu_etal_2010}, but see \cite{Wright_etal_2013} on mechanisms of copper tolerance).
+Many classic and contemporary studies of local adaptation use \textit{Mimulus sensu lato} species because of their natural history, easy propagation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
+Yet, there is a deficiency of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010, Wright_etal_2013}.
 We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
@@
 
+\textit{E.
+cardinalis} is a perennial forb native to the Western US (California and Oregon).
+It is predomintantly outcrossing, self-compatible, and pollinated primarily by hummingbirds.
 We used 16 populations from throughout the range of \textit{E.
@@
    \centering
-   \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
-   \begin{tabular}{@{} lllll @{}}
+   \topcaption[Focal Populations]{Latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
+   \begin{tabular}{@{} llll @{}}
       \toprule
-  Name& Region  & Latitude  & Longtiude  & Elevation (mas) \\
+  Name  & Latitude  & Longtiude  & Elevation (mas) \\
       \midrule
-	Hauser Creek & South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "HAU"], 3)}	& 
+	Hauser Creek & \Sexpr{round(focClim$Latitude[focClim$PopID == "HAU"], 3)}	& 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "HAU"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "HAU"]}   \\
-	Cottonwood Creek	& South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "CTC"], 3)} & 
+	Cottonwood Creek	& \Sexpr{round(focClim$Latitude[focClim$PopID == "CTC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "CTC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "CTC"]}   \\
-	Sweetwater River	& South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "CUR"], 3)} & 
+	Sweetwater River	& \Sexpr{round(focClim$Latitude[focClim$PopID == "CUR"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "CUR"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "CUR"]}   \\
-	Grade Road Palomar & South Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "GRP"], 3)} &
+	Grade Road Palomar & \Sexpr{round(focClim$Latitude[focClim$PopID == "GRP"], 3)} &
     \Sexpr{round(focClim$Longitude[focClim$PopID == "GRP"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "GRP"]}   \\
-	Whitewater Canyon &	Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "WWC"], 3)} & 
+	Whitewater Canyon & \Sexpr{round(focClim$Latitude[focClim$PopID == "WWC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WWC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WWC"]}   \\
-	Mill Creek	& Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "MIL"], 3)} & 
+	Mill Creek	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MIL"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "MIL"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "MIL"]}   \\
-	West Fork Mojave River	& Transverse & \Sexpr{round(focClim$Latitude[focClim$PopID == "WFM"], 3)} & 
+	West Fork Mojave River	& \Sexpr{round(focClim$Latitude[focClim$PopID == "WFM"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WFM"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WFM"]}   \\
-	North Fork Middle Tule River	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "NMT"], 3)} & 
+	North Fork Middle Tule River	& \Sexpr{round(focClim$Latitude[focClim$PopID == "NMT"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "NMT"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "NMT"]}   \\
-	Paradise Creek	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "PRD"], 3)} & 
+	Paradise Creek	& \Sexpr{round(focClim$Latitude[focClim$PopID == "PRD"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "PRD"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "PRD"]}   \\
-	Redwood Creek	& South Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "RWD"], 3)} & 
+	Redwood Creek	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RWD"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RWD"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RWD"]}   \\
-	Wawona  & Central Sierras & \Sexpr{round(focClim$Latitude[focClim$PopID == "WNA"], 3)} & 
+	Wawona  & \Sexpr{round(focClim$Latitude[focClim$PopID == "WNA"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "WNA"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "WNA"]}   \\
-	Rainbow Creek	& Central Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RBW"], 3)} & 
+	Rainbow Creek	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RBW"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RBW"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RBW"]}   \\
-	Middle Yuba River	& North Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MYU"], 3)} & 
+	Middle Yuba River	& \Sexpr{round(focClim$Latitude[focClim$PopID == "MYU"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "MYU"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "MYU"]}   \\
-	Little Jamison Creek	& North Sierras	& \Sexpr{round(focClim$Latitude[focClim$PopID == "LIJ"], 3)} & 
+	Little Jamison Creek	& \Sexpr{round(focClim$Latitude[focClim$PopID == "LIJ"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "LIJ"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "LIJ"]}   \\
-	Deep Creek	& North Coast & \Sexpr{round(focClim$Latitude[focClim$PopID == "DPC"], 3)} & 
+	Deep Creek	& \Sexpr{round(focClim$Latitude[focClim$PopID == "DPC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "DPC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "DPC"]}   \\
-	Rock Creek	& North Margin & \Sexpr{round(focClim$Latitude[focClim$PopID == "RCC"], 3)} & 
+	Rock Creek	& \Sexpr{round(focClim$Latitude[focClim$PopID == "RCC"], 3)} & 
     \Sexpr{round(focClim$Longitude[focClim$PopID == "RCC"], 3)}	& \Sexpr{focClim$Elevation[focClim$PopID == "RCC"]}   \\
@@
 We recorded germination daily between one to two weeks after sowing, and every 2-3 days thereafter.
-On 5 May (21 days after sowing), we transferred seedlings to one of two growth chambers (Conviron, Manitoba, Canada).
+On 5 May (21 days after sowing), we transferred seedlings to one of two growth chambers (Model E-15 Conviron, Manitoba, Canada).
 We thinned seedlings to one plant per cone-tainer, leaving the center-most plant.
@@
 The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
-Lower humidity would have made the drought more severe,but low soil moisture is stressful in and of itself.
+Lower humidity would have made the drought more severe, but low soil moisture is stressful in and of itself.
 The total number of plants in each treatment was: $n_{\text{cool,dry}} = \Sexpr{sum(indPerPop["Cool", "Dry", ])}$; $n_{\text{cool,ww}} = \Sexpr{sum(indPerPop["Cool", "Wet", ])}$; $n_{\text{hot,dry}} = \Sexpr{sum(indPerPop["Hot", "Dry", ])}$; $n_{\text{hot,ww}} = \Sexpr{sum(indPerPop["Hot", "Wet", ])}$.
@@
 Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
-To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variable explains latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
+To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variables explain latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
 We found that a population's position along TraitPC1 correlated strongly with the latitude of origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{E.
@@
 For climate neighborhoods, we sampled climate at 1000 random points (at 90-m resolution) within a 62-km radius buffer around the collection and took the average.
-We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
-Significant spatial autocorrelation persisted for approximately 62 km.
+We chose this buffer radius based on population genetic structure, as inferred from $\approx$25,000 restriction-site associated SNPs among 49 populations from across the range \citep{Paul_etal_2016}.
+Spatial autocorrelation in allele frequencies persists for 62 km.
+However radii of 10 km$^2$ and 100 km$^2$ resulted in similar outcomes (data not shown).
 Since \textit{E.
@@
 Therefore, we did not consider Climate-TraitPC1 variables to be candidate selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
+However, we do interpret potential selective agents identified in Climate-Latiude analyses alone, because the goal was to explain the latitudinal clines in traits, not all aspects of climate that vary with latitude.
 
@@
 
-  pdf(paste(pathFig, "/FigureS_LocalvNeighCorr.pdf", sep = ""), w = 8, h = 4, useDingbats = F)
+  pdf(paste(pathFig, "/Figure_LocalvNeighCorr.pdf", sep = ""), w = 8, h = 4, useDingbats = F)
   par(mai = c(1, 1, 0.2, 0.2), mfrow = c(1, 2))
@@
 For Climate-Latitude analyses, we compiled a representative set of \Sexpr{2 * nrow(thinnedOccClimate)} recent (since 2000) known \textit{E.
-cardinalis} occurrences.
-These occurrences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+cardinalis} occurrences from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+These occurrences were thinned by 50\% to correct for uneven sampling.
 For both Climate-TraitPC1 analyses (16 focal populations) and Climate-Latitude (many populations), we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
@@
 We included \Sexpr{length(climVars) / 4} climatic factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
-The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
+The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, and annual temperature range.
 For each variable, we calculated both a 30-year normal by averaging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
@@
 A population's position along this axis strongly covaried with its latitude of origin; southern populations grew faster than northern populations (Fig~\ref{fig:Fig_PC1vLat}).
-As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.
 There were similar latitudinal clines for individual traits underlying PC1 (\Cref{fig:FigS_GermLat,fig:FigS_LERlat,fig:FigS_SERlat,fig:FigS_Photolat}).
@@
 Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation.
-We found little evidence of G$\times$E in \textit{E.
+In contrast to the intrinsic differences described above, we found little evidence of G$\times$E in \textit{E.
 cardinalis}.
@@
 Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
-Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
 Complete ANOVA tables are available in the Supporting Information (\Cref{table:TableS_fitLER,table:TableS_fitSER,table:TableS_fitPhoto,table:TableS_fitMort})
@@
 
-Interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{E.
+Interannual variation in climate averaged over each populations's climatic neighborhood correlated most strongly with trait variation and latitude of \textit{E.
 cardinalis} occurrences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
 All \Sexpr{nImpLat} Climate-Latitude and \Sexpr{nImpPC1} Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
-In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:FigS_LocalvNeighCorr}).
+In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:Fig_LocalvNeighCorr}).
 On average, neighborhood Climate-TraitPC1 correlation coefficients were \Sexpr{-round(tt1$estimate, 2)} higher than correlations with local-scale climate variables (paired \textit{t}-test, $t = \Sexpr{-round(tt1$statistic, 2)}$, d.f.
-= \Sexpr{round(Meff1 / 2 - 1, 1)}, $P = \Sexpr{round(pt1 * 10 ^ -floor(log10(pt1)), 2)} \times 10 ^ \Sexpr{floor(log10(pt1))}$).
+= \Sexpr{round(Meff1 / 2 - 1, 1)}, $P = \Sexpr{round(pt1 * 10 ^ -floor(log10(pt1)), 2)} \times 10 ^ {\Sexpr{floor(log10(pt1))}}$).
 Likewise, neighborhood Climate-Latitude correlation coefficients were \Sexpr{-round(tt2$estimate, 2)} higher than those for local-scale climate (paired \textit{t}-test, $t = \Sexpr{-round(tt2$statistic, 2)}$, d.f.
-= \Sexpr{round(Meff2 / 2 - 1, 1)}, $P = \Sexpr{round(pt2 * 10 ^ -floor(log10(pt2)), 2)} \times 10 ^ \Sexpr{floor(log10(pt2))}$).
-
-The most important climatic variables related to the interannual variation rather than average climate.
-Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{E.
+= \Sexpr{round(Meff2 / 2 - 1, 1)}, $P = \Sexpr{round(pt2 * 10 ^ -floor(log10(pt2)), 2)} \times 10 ^ {\Sexpr{floor(log10(pt2))}}$).
+
+Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates to explain the latitudinal cline in \textit{E.
 cardinalis} (see Table~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
@@
 
-Interannual variation in temperature of the coldest quarter (neighborhood bio11$_\sigma$) is another plausible candidate because it was only variable in both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
+Interannual variation in temperature of the coldest quarter (neighborhood bio11$_\sigma$) is another plausible candidate because it was the only variable in both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
 Neighborhood bio11$_\sigma$ explained more variation in TraitPC1 than latitude (latitude $r^2=\Sexpr{round(summary(fitLat)["r.squared"]$r.squared, 2)}$ vs.
@@
 The most important Climate-TraitPC1 variable, neighborhood variation in mean diurnal range (bio2$_\sigma$; Fig.~\ref{fig:Fig_ClimVarImp}B) did not have any obvious similarity to Climate-Latitude variables.
-Given the large number of potential associations, we therefore think this may be a spuriously strong relationship and do not consider it further.
+Given the large number of potential associations, we therefore think this may be a spuriously strong relationship.
 
@@
 	\label{fig:Fig_ClimVarImp}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of local versus neighborhood Climate-TraitPC1 and Climate-Latitude correlations
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_LocalvNeighCorr.pdf}}
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\caption[Local versus neighborhood correlations]{Neighborhood climate predicts TraitPC1 (`Climate-trait', panel A) and Latitude of occurences (`Climate-latitude', panel B) better than local climate.
+Each point is the absolute value of the Pearson correlation coefficient (|$r$|) between TraitPC1 (A) or latitude (B) for \Sexpr{length(climVars)/4} climatic factors, for which we used both the 30-year mean (closed circles) and coefficient of variation (open circles).
+Most points lie above the 1:1 line, indicating stronger correlations with neighborhood compared to local climate.
+Neighborhood climate was integrated over a 62-km radius around focal populations (see text for further detail).}
+\label{fig:Fig_LocalvNeighCorr}
 \end{figure}
@@
 However, we found little evidence that populations respond differently to temperature or drought.
-Due to low replication within families, we did not have power to assess within population genotype-by-environment interactions, which may be present.
-As we discuss below, low variation in plasticity among populations result may indicate that some dimensions of the fundamental abiotic niche are relatively conserved.
+Due to low replication within families, we did not have power to assess within-population genotype-by-environment interactions, which may be present.
+As we discuss below, low variation in plasticity among populations may indicate that some dimensions of the fundamental abiotic niche are relatively conserved.
+Note that statistical power to detect significant plasticity is lower than that for intrinsic differences.
+However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain low variation in plasticity.
 Finally, our results suggest that neighborhood-scale climate and interannual variation are more important selective agents than local averages.
-In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in physiology changes clinally, whereas plastic responses to temperature and drought are relatively similar.
+In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in physiology changes clinally, whereas plastic responses to temperature and drought are relatively static.
 One caveat to bear in mind is that we are limited by the size of the climate grid ($\approx 90~\text{m}^2$) and therefore unable to detect very fine-scale local adaptation.
@@
 Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001, Lenormand_2002} and hence the type of variation maintained within species.
-Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strongly asymmetric, and/or maladaptive gene flow is too high.
-To clarify, strong tradeoffs can prevent local adaptation in variable environments because selection favors habitat specialists that track a specific habitat regardless of its frequency in the environment \citep{Levins_1968}.
-For example, a riparian specialist may experience similar selection in rivers of highfall regions and deserts, even though the habitat is much rarer in the latter.
+Specifically, adaptive variation can be maintained by spatially varying selection if tradeoffs are not too strong, demography is symmetric, and/or maladaptive gene flow is low.
+Strong tradeoffs can prevent local adaptation in spatially variable environments because selection favors habitat specialists that track a specific habitat regardless of its frequency in the environment \citep{Levins_1968}.
+For example, a riparian specialist may experience similar selection in rivers of high rainfall regions and deserts, even though the habitat is much rarer in the latter.
 In \textit{E.
 cardinalis} we found substantial genetically based variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
-If this variation is adaptive, it suggests one of several possibilities to investigate in the future: fitness tradeoff between low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
+If this variation is adaptive, it suggests one of several possibilities to investigate in the future: the fitness tradeoff between low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
 That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatially heterogenous selection.
-We also know from previous work that population size does not vary strongly with latitude.
+We also know from previous work that population size does not vary strongly with latitude (Angert, unpub.
+data).
 Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range \citep{Paul_etal_2016}.
@@
 cardinalis} populations shapes selection -- populations are locally adapted to prevailing climate in their neighborhood, but perhaps not perfectly adapted to their local climate.
-This may not greatly constrain local adaptation because local and neighborhood climate values at the resolution of ClimateWNA (90 m$^2$) were generally similar in \textit{E.
-cardinalis} populations (Fig.~\ref{fig:FigS_LocalvNeighCorr}).
+This may not greatly constrain local adaptation because local and neighborhood climate values were generally similar in \textit{E.
+cardinalis} populations (Fig.~\ref{fig:Fig_LocalvNeighCorr}), at least at the resolution of ClimateWNA (90 m$^2$).
 Therefore, we would predict in reciprocal transplants that populations whose local climate is farther from their neighborhood average would be less well adapted than those close to their neighborhood average.
@@
 Soil moisture in riparian habitats where \textit{E.
-cardinalis} live are highly heterogeneous at very small spatial scales (several meters).
+cardinalis} lives is highly heterogeneous at very small spatial scales (several meters).
 Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Mediterranean climates of western North America.
@@
 
-In sum, these results indicate that differences in physiology and growth, but not responses to temperature and drought, mediate local adaptation to climate in \textit{E.
+In sum, these results indicate that intrinsic differences in physiology and growth, but not plastic responses to temperature and drought, mediate local adaptation to climate in \textit{E.
 cardinalis}.
@@
 However, we cannot exclude unexplored factors (e.g.
-edaphic conditions, competitors, pollinators, etc.) which may drive the latitudinal cline.
+edaphic conditions, competitors, pollinators, etc.) which may also contribute to the latitudinal cline.
 
@@
 Therefore, in a stable environment where winter survivorship is assured in most years, failure to store resources may reduce lifetime fitness.
-But the environment is not constant.
-For perennial herbs in Mediterranean climates, a dry winter (rainy season) can kill the rhizomes (underground stems that store nutrients for future growth) before emergence or aboveground stems before flowering.
+But for perennial herbs in Mediterranean climates, a dry winter (rainy season) can kill the rhizomes (underground stems that store nutrients for future growth) before emergence or aboveground stems before flowering.
 If drought years occur frequently enough, selection may favour the fast-growing strategy because there is no advantage to storage if drought kills plants before flowering.
@@
 This hypothesis makes several independent, testable predictions.
-The allocation tradeoff predicts that northern populations will provision more assimilate to rhizomes compared with southern populations.
+The allocation tradeoff predicts that northern populations will provision more photosynthetic assimilate to rhizomes compared with southern populations.
 If southern populations are indeed more `annualized' because more frequent droughts cause mortality, then we predict that species distribution models using recent climate would best predict occurrences in the south, whereas longer term climate would be a better predictor in the north.
 Finally, we predict that southern populations would show greater variation in the size of recruits and higher maximum population growth rates.
-Exploratory analysis also pointed to variation in winter temperature as a potential selective agent.
-Specifically, winter temperature is more variable in the north, which may lead to less predictability in the start of the growing season or late frost.
-This could affect selection on phenology, which we did not study here, but it is not obviously connected to the latitudinal cline in physiological traits.
 
@@
 Interestingly, the climatic neighborhood may shape selective pressures more than local climate.
-In the future, we will use field experiments to test whether greater variation in precipitation selects for faster growth and that selection on temperature/drought responses does not vary among populations.
+In the future, we will use field experiments to test whether greater variation in precipitation selects for faster growth and if selection on temperature/drought responses does not vary among populations.
 By doing so, we aim to understand why certain physiological and developmental mechanisms, but not others, contribute to local adaptation.
@@
 
-% Figure of local versus neighborhood Climate-TraitPC1 and Climate-Latitude correlations
-
-\begin{figure}[h!]
-\centerline{\includegraphics[width=1\textwidth]{Figures/FigureS_LocalvNeighCorr.pdf}}
-\usefont{T1}{phv}{m}{n}
-\fontsize{10}{12}
-\selectfont
-\caption[Local versus neighborhood correlations]{Neighborhood climate predicts TraitPC1 (`Climate-trait', panel A) and Latitude of occurences (`Climate-latitude', panel B) better than local climate.
-Each point is the absolute value of the Pearson correlation coefficient (|$r$|) between TraitPC1 (A) or latitude (B) for \Sexpr{length(climVars)/4} climatic factors, for which we used both the 30-year mean (closed circles) and coefficient of variation (open circles).
-Most points lie above the 1:1 line, indicating stronger correlations with neighborhood compared to local climate.
-Neighborhood climate was integrated over a 62-km radius around focal populations (see text for further detail).}
-\label{fig:FigS_LocalvNeighCorr}
-\end{figure}
-
 % Figure of bio11_cva_SA versus TraitPC1

### 2017-05-30 ms/ms.tex
revised ms

@@
    \centering
-   \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
-   \begin{tabular}{@{} lllll @{}}
+   \topcaption[Focal Populations]{Latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
+   \begin{tabular}{@{} llll @{}}
       \toprule
-  Name& Region  & Latitude  & Longtiude  & Elevation (mas) \\
+  Name  & Latitude  & Longtiude  & Elevation (mas) \\
       \midrule
-	Hauser Creek & South Margin & 32.657	& 
+	Hauser Creek & 32.657	& 
     -116.532	& 799   \\
-	Cottonwood Creek	& South Margin & 32.609 & 
+	Cottonwood Creek	& 32.609 & 
     -116.7	& 267   \\
-	Sweetwater River	& South Margin & 32.9 & 
+	Sweetwater River	& 32.9 & 
     -116.585	& 1180   \\
-	Grade Road Palomar & South Margin & 33.314 &
+	Grade Road Palomar & 33.314 &
     -116.871	& 1577   \\
-	Whitewater Canyon &	Transverse & 33.994 & 
+	Whitewater Canyon & 33.994 & 
     -116.665	& 705   \\
-	Mill Creek	& Transverse & 34.077 & 
+	Mill Creek	& 34.077 & 
     -116.873	& 2050   \\
-	West Fork Mojave River	& Transverse & 34.284 & 
+	West Fork Mojave River	& 34.284 & 
     -117.378	& 1120   \\
-	North Fork Middle Tule River	& South Sierras & 36.201 & 
+	North Fork Middle Tule River	& 36.201 & 
     -118.651	& 1314   \\
-	Paradise Creek	& South Sierras & 36.518 & 
+	Paradise Creek	& 36.518 & 
     -118.759	& 926   \\
-	Redwood Creek	& South Sierras & 36.691 & 
+	Redwood Creek	& 36.691 & 
     -118.91	& 1727   \\
-	Wawona  & Central Sierras & 37.541 & 
+	Wawona  & 37.541 & 
     -119.649	& 1224   \\
-	Rainbow Creek	& Central Sierras	& 37.819 & 
+	Rainbow Creek	& 37.819 & 
     -120.007	& 876   \\
-	Middle Yuba River	& North Sierras	& 39.397 & 
+	Middle Yuba River	& 39.397 & 
     -121.082	& 455   \\
-	Little Jamison Creek	& North Sierras	& 39.743 & 
+	Little Jamison Creek	& 39.743 & 
     -120.704	& 1603   \\
-	Deep Creek	& North Coast & 41.668 & 
+	Deep Creek	& 41.668 & 
     -123.11	& 707   \\
-	Rock Creek	& North Margin & 43.374 & 
+	Rock Creek	& 43.374 & 
     -122.957	& 326   \\
@@
 The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
-Lower humidity would have made the drought more severe,but low soil moisture is stressful in and of itself.
+Lower humidity would have made the drought more severe, but low soil moisture is stressful in and of itself.
 The total number of plants in each treatment was: $n_{\text{cool,dry}} = 169$; $n_{\text{cool,ww}} = 174$; $n_{\text{hot,dry}} = 176$; $n_{\text{hot,ww}} = 183$.
@@
 For Climate-Latitude analyses, we compiled a representative set of 356 recent (since 2000) known \textit{E.
-cardinalis} occurrences.
-These occurrences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+cardinalis} occurrences from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
+These occurrences were thinned by 50\% to correct for uneven sampling.
 For both Climate-TraitPC1 analyses (16 focal populations) and Climate-Latitude (many populations), we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
@@
 We included 24 climatic factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
-The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
+The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, and annual temperature range.
 For each variable, we calculated both a 30-year normal by averaging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
@@
 
-Interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{E.
+Interannual variation in climate averaged over each populations's climatic neighborhood correlated most strongly with trait variation and latitude of \textit{E.
 cardinalis} occurrences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
 All 16 Climate-Latitude and 3 Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
-In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:FigS_LocalvNeighCorr}).
+In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:Fig_LocalvNeighCorr}).
 On average, neighborhood Climate-TraitPC1 correlation coefficients were 0.16 higher than correlations with local-scale climate variables (paired \textit{t}-test, $t = 7.87$, d.f.
-= 33.6, $P = 3.94 \times 10 ^ -9$).
+= 33.6, $P = 3.94 \times 10 ^ {-9}$).
 Likewise, neighborhood Climate-Latitude correlation coefficients were 0.13 higher than those for local-scale climate (paired \textit{t}-test, $t = 6.71$, d.f.
-= 36.8, $P = 7.22 \times 10 ^ -8$).
-
-The most important climatic variables related to the interannual variation rather than average climate.
-Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{E.
+= 36.8, $P = 7.22 \times 10 ^ {-8}$).
+
+Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates to explain the latitudinal cline in \textit{E.
 cardinalis} (see Table~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
@@
 
-Interannual variation in temperature of the coldest quarter (neighborhood bio11$_\sigma$) is another plausible candidate because it was only variable in both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
+Interannual variation in temperature of the coldest quarter (neighborhood bio11$_\sigma$) is another plausible candidate because it was the only variable in both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
 Neighborhood bio11$_\sigma$ explained more variation in TraitPC1 than latitude (latitude $r^2=0.55$ vs.
@@
 	\label{fig:Fig_ClimVarImp}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure of local versus neighborhood Climate-TraitPC1 and Climate-Latitude correlations
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_LocalvNeighCorr.pdf}}
+\usefont{T1}{phv}{m}{n}
+\fontsize{10}{12}
+\selectfont
+\caption[Local versus neighborhood correlations]{Neighborhood climate predicts TraitPC1 (`Climate-trait', panel A) and Latitude of occurences (`Climate-latitude', panel B) better than local climate.
+Each point is the absolute value of the Pearson correlation coefficient (|$r$|) between TraitPC1 (A) or latitude (B) for 24 climatic factors, for which we used both the 30-year mean (closed circles) and coefficient of variation (open circles).
+Most points lie above the 1:1 line, indicating stronger correlations with neighborhood compared to local climate.
+Neighborhood climate was integrated over a 62-km radius around focal populations (see text for further detail).}
+\label{fig:Fig_LocalvNeighCorr}
 \end{figure}
@@
 
-% Figure of local versus neighborhood Climate-TraitPC1 and Climate-Latitude correlations
-
-\begin{figure}[h!]
-\centerline{\includegraphics[width=1\textwidth]{Figures/FigureS_LocalvNeighCorr.pdf}}
-\usefont{T1}{phv}{m}{n}
-\fontsize{10}{12}
-\selectfont
-\caption[Local versus neighborhood correlations]{Neighborhood climate predicts TraitPC1 (`Climate-trait', panel A) and Latitude of occurences (`Climate-latitude', panel B) better than local climate.
-Each point is the absolute value of the Pearson correlation coefficient (|$r$|) between TraitPC1 (A) or latitude (B) for 24 climatic factors, for which we used both the 30-year mean (closed circles) and coefficient of variation (open circles).
-Most points lie above the 1:1 line, indicating stronger correlations with neighborhood compared to local climate.
-Neighborhood climate was integrated over a 62-km radius around focal populations (see text for further detail).}
-\label{fig:FigS_LocalvNeighCorr}
-\end{figure}
-
 % Figure of bio11_cva_SA versus TraitPC1

### 2017-08-28 ms/ms.tex
test

--- 2017-01-06 ms/ms.tex
@@
 %
-
-%
-% Title and authors
-%
-
-\title{Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Mimulus cardinalis}}
-\author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
-\date{}
-
-%
-% Start document
-%
-
-\usepackage{Sweave}
-\begin{document}
-\input{ms-concordance}
-\maketitle
-
-$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada \\
-$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca \\
-
-Running Head: Latitudinal cline and climate in \textit{Mimulus} \\
-
-Key words: local adaptation, cline, photosynthesis, growth rate, \textit{Mimulus} \\
-
-Data will be archived on Dryad upon publication. \\
-
-\clearpage
-
-% \listoffigures
-% \listoftables
-
-\setstretch{\stretchy}
-\linenumbers
-
-\section*{Abstract}
-
-Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it.
-Correlation between traits and latitude, or latitudinal clines, are among the most common pieces of evidence for local adaptation, but identifying the traits under selection and the selective agents are challenging.
-Here, we investigated a latitudinal cline in growth and photosynthesis across 16 populations of the perennial herb \textit{Mimulus cardinalis} (Phrymaceae).
-Using machine learning methods, we identify interannual variation in precipitation as a likely selective agent: Southern populations from more variable environments had higher photosynthetic rates and grew faster.
-We hypothesize that selection may favor a more annualized life history -- grow now rather than save for next year -- in environments where severe droughts occur more often.
-Thus our study provides insight into how species may adapt if Mediterranean climates become more variable due to climate change.
-
-\section*{Introduction}
-
-% Toggle these commands to get version for doing word count
-% \linenumbers
-% \pagenumbering{gobble}
-% \raggedright
-
-% Local adaptation and clines
-Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}.
-Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}.
-Phenotypic differences between populations along a cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
-Despite a long history of studying local adaptation and clines, it remains challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
-In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment.
-We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change.
-A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures.
-% Here, we examine physiological trait variation and possible selective agents in the perennial herb \textit{Mimulus cardinalis} Douglas ex Benth.
-(Phrymaceae), a model system for local adaptation studies.
-
-% Intrinsic versus plastic variation 
-When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}.
-The traits that underlie local adaptation, however, need not mirror this pattern.
-Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away.
-Alternatively, genotype-by-environment interactions could indicate that variation in plasticity mediates local adaptation.
-We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively.
-Adaptation involves both intrinsic and plastic trait variation.
-For example, intrinsic differences in photoperiod \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
-Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
-In plants especially, we know little about the prevalance and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
-
-% Selective agents (trait-enviro assoc)
-A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments.
-Either instrinsic and/or plastic variation should vary clinally along environmental gradients.
-Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
-For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}.
-Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
-Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008}, and drought response \citep{Kooyers_etal_2015} that likely relate to climatic variation.
-
-Despite the fact that latitudinal clines have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
-Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days.
-Hence, latitude is an effective proxy for the underlying climatic driver, but we would expect a yet stronger relationship between traits and the key climatic variable(s) driving selection.
-Alternatively, latitude may be more strongly related to traits than any single climatic variable for at least two reasons.
-First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline.
-Alternatively, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration.
-We refer to this as the `climatic neighborhood'.
-For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a lower latitude set of populations will experience warmer climate than a higher latitude one.
-Thus, any particular low latitude population would be warm-adapted, even if it was located in a cooler (e.g.
-high elevation) site.
-Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g.
-coastal vs.
-continental), dissecting the evolution of latitudinal clines across many species will help identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local versus regional climate shape selective pressures.
-
-In this study, we intestigated two major questions: 1) whether intrinsic or plastic physiological trait variation corresponds with latitude; and 2) what climatic factor(s) could plausibly be repsonsible for latitudinal clines.
-Within question 2, we tested three hypotheses outlined in the previous paragraph: latitudinal clines are explained by a single dominant climatic factor, multiple climatic factors, or the climatic neighborhood experienced by nearby population connected through gene flow.
-These hypotheses are not mutually exclusive since, for example, single or multiple factors in a climatic neighborhood may lead to latitudinal clines.
-
-We examined these questions in \textit{Mimulus cardinalis} because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
-Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
-Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}.
-We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
-We found a latitudinal cline of intrinsic variation in photosynthesis and growth, but little evidence for variation in plasticity.
-Interannual variation in precipitation and temperature are associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M.
-cardinalis}.
-The climatic neighborhoods around populations explained trait variation better than local climate, indicating that latitudinal clines may be common because latitude integrates effects of selection on populations connected through gene flow.
-We place these findings in the context of life history theory and consider future directions in the Discussion.
-
-\section*{Material and Methods}
-
-\subsection*{Population Selection}
-
-We used 16 populations from throughout the range of \textit{M.
-cardinalis} (Table~\ref{table:Table_FocalPops}).
-Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Table of focal population
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}[ht]
-   \centering
-   \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
-   \begin{tabular}{@{} lllll @{}}
-      \toprule
-  Name& Region  & Latitude  & Longtiude  & Elevation (mas) \\
-      \midrule
-	Hauser Creek & South Margin & 32.657	& 
-    -116.532	& 799   \\
-	Cottonwood Creek	& South Margin & 32.609 & 
-    -116.7	& 267   \\
-	Sweetwater River	& South Margin & 32.9 & 
-    -116.585	& 1180   \\
-	Grade Road Palomar & South Margin & 33.314 &
-    -116.871	& 1577   \\
-	Whitewater Canyon &	Transverse & 33.994 & 
-    -116.665	& 705   \\
-	Mill Creek	& Transverse & 34.077 & 
-    -116.873	& 2050   \\
-	West Fork Mojave River	& Transverse & 34.284 & 
-    -117.378	& 1120   \\
-	North Fork Middle Tule River	& South Sierras & 36.201 & 
-    -118.651	& 1314   \\
-	Paradise Creek	& South Sierras & 36.518 & 
-    -118.759	& 926   \\
-	Redwood Creek	& South Sierras & 36.691 & 
-    -118.91	& 1727   \\
-	Wawona  & Central Sierras & 37.541 & 
-    -119.649	& 1224   \\
-	Rainbow Creek	& Central Sierras	& 37.819 & 
-    -120.007	& 876   \\
-	Middle Yuba River	& North Sierras	& 39.397 & 
-    -121.082	& 455   \\
-	Little Jamison Creek	& North Sierras	& 39.743 & 
-    -120.704	& 1603   \\
-	Deep Creek	& North Coast & 41.668 & 
-    -123.11	& 707   \\
-	Rock Creek	& North Margin & 43.374 & 
-    -122.957	& 326   \\
-	\bottomrule
-	\end{tabular}
-	\label{table:Table_FocalPops}
-\end{table}
-
-\subsection*{Plant propagation}
-
-On 14 April, 2014, 3-5 seeds per family were sown directly on sand (Quikrete Play Sand, Georgia, USA) watered to field capacity in RLC4 Ray Leach cone-tainers placed in RL98 98-well trays (Stuewe \& Sons, Inc., Oregon, USA).
-We used pure sand both to facilitate root-washing and because \textit{M.
-cardinalis} typically grows in sandy, riparian soils (A.
-Angert, pers.
-obs.).
-Two jumbo-sized cotton balls at the bottom of cone-tainers prevented sand from washing out.
-Cone-tainers sat in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) to continuously bottom-water plants during germination in greenhouses at the University British Columbia campus in Vancouver, Canada (49\degree 15' N, 123\degree 15' W).
-Misters thoroughly wetted the top of the sand every two hours during the day.
-Most seeds germinated between 1 and 2 weeks, but we allowed 3 weeks before transferring seedlings to growth chambers.
-We recorded germination daily between one to two weeks after sowing, and every 2-3 days thereafter.
-On 5 May (21 days after sowing), we transferred seedlings to one of two growth chambers (Conviron, Manitoba, Canada).
-We thinned seedlings to one plant per cone-tainer, leaving the center-most plant.
-702 of 768 (91.4\%) had plants that could be used in the experiment.
-We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments.
-The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments (Table~\ref{table:TableS_InitialSize}).
-
-\subsection*{Temperature and drought treatments}
-
-We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels.
-The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which we designate as Hot and Cool treatments.
-Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments.
-A detailed description of treatments is provided in the Supplemental Materials and Methods and summarized in Fig~\ref{fig:Fig_ExptlDes}.
-Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level.
-Within each chamber, there were two Well-watered blocks and two Drought blocks.
-The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$.
-The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure summarizing experimental design
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ExptlDes.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Experimental Design]{Overview of experimental treatments and timing of key trait measurements.
-All plants germinated within 21 days of sowing.
-At that time, we began temperature treatments (left axis), simulating a typical June-August weather pattern at Hot (red) and Cool (blue) sites.
-The bold lines track the average daily temperatures.
-Within each day, there was a maximum daytime temperture (top of translucent polygons) and minimum nighttime temperature (bottom of translucent polygons).
-The drought treatment commenced later by ramping down the frequency of bottom-watering episodes (dashed black line; right axis), while watering frequency was maintained in the control treatment (solid black line).
-Grey boxes on the bottom of the plot outline the period of key measurements described in the Material and Methods.}
-	\label{fig:Fig_ExptlDes}
-\end{figure}
-
-\subsection*{Trait measurements}
-
-We measured five traits in response to temperature and watering treatments (Table~\ref{table:Table_traits}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Table of key traits
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}[ht]
-   \centering
-   \topcaption{Key traits measured in this study.}
-   \begin{tabular}{@{} ll @{}}
-      \toprule
-  Trait & Units \\
-      \midrule
-  Days to germination  & day \\
-  Leaf expansion rate  &  mm day$^{-1}$  \\
-  Stem elongation rate  &  mm or cm day$^{-1}$  \\
-  Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
-  Mortality & probability of death  \\
-	    \bottomrule
-   \end{tabular}
-   \label{table:Table_traits}
-\end{table}
-
-\paragraph{Days to germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}.
-We treated Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
-Statistical signifcance of the Population effect was determined using analysis of deviance.
-Note that, unlike other traits discussed below, we did not include Block, Treatment, or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not yet been applied.
-
-\paragraph{Growth rate: leaf expansion and stem elongation}
-
-We measured growth rate during two phases: leaf expansion and stem elongation.
-We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
-We ceased measuring leaf length once it appeared to asymptote and growth shifted to stem elongation.
-We also censused plant height on 7 occasions (twice per week) between 29 May and 20 June (45 to 67 days after sowing).
-Both leaf expansion and stem elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and stem growth) using empirical Bayes' estimates from linear mixed-effects models fit with the R package \pkg{lme4} version 1.1-12 \citep{Bates_etal_2015}.
-
-\paragraph{Photosynthesis}
-During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate on a subset of 329 plants evenly spread between treatments and families within populations.
-The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6-cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
-We made all measurements at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$ of photosynthetically active radiation), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
-During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level could be measured under the same conditions.
-% As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out.
-Therefore, we calculated the `intrinsic' photosynthetic rate for each individual as its residual deviance from the regression between stomoatal conductance and photosynthetic rate across all plants.
-We also removed one outlier.
-
-\paragraph{Mortality}
-We assayed mortality during twice-weekly growth measurements.
-We analyzed the probability of surviving until the end of the experiment as a function of population, treatment, and their interactions using a Generalized Linear Mixed Model (GLMM) assuming binomially distributed errors.
-We included Family and Block as random effects.
-We assessed significance of fixed effects using Type-II Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
-
-\subsection*{Intrinsic variation and plasticity}
-
-For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment, and Population $\times$ Treatment interactions.
-We interpreted significant Population effects to indicate intrinsic variation and Population $\times$ Treatment interactions to indicate variation in plasticity.
-As mentioned above, we used survival and GLMM models for germination rate and mortality, respectively.
-For all other traits, we used mixed model ANOVAs with Family and Block included as random factors.
-We fit models using restricted maximum likelihood in lmer, a function in the R package \pkg{lme4} \citep{Bates_etal_2015}.
-We determined significant fixed effect terms using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-32 \citep{Kuznetsova_etal_2016}.
-This package uses Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests.
-We also included days to germination as a covariate in growth analyses.
-% To ensure that Population and Treatment effects were specific to a particular growth phase, we included earlier growth phases as a covariate.
-Specifically, we included germination day as a covariate in leaf expansion and stem elongation analyses; we included germination day and leaf expansion rate as a covariate in stem elongation analyses.
-
-\subsection*{Principal components of germination, growth, and photosynthesis}
-For each single-trait model above, we extracted the Population coefficient (factoring out Treatment and other effects).
-The multivariate distribution of these coefficients was then summarized using principal components analysis.
-The first principal component of these traits (TraitPC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.
-
-\subsection*{Identifying putative selective agents}
-
-We found that a population's position along TraitPC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M.
-cardinalis}.
-Hypothesis 1: if a single climatic factor explained more trait variation than latitude alone, this would suggest that that factor is a selective agent underlying the latitudinal cline in our 16 focal populations.
-Hypothesis 2: if multiple climatic factors together explained trait variation,  we interpreted this as evidence that multiple climatic factors together have generated the latitudinal cline.
-We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables.
-In addition, to help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{M.
-cardinalis} occurrence records (see detail below), not just the 16 focial populations.
-We refer to these climatic factors as `Climate-Latitude' variables.
-The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{M.
-cardinalis} range.
-Therefore, we did not consider Climate-TraitPC1 variables to be candidates selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
-We selected Climate-Latitude and Climate-TraitPC1 variables independently using Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}.
-VSURF ranks variables by their importance over regression trees in the forest.
-We kept only variables selected for prediction, the most stringent criterion.
-A visual overview of how we selected climatic variables is depicted in Fig~\ref{fig:Fig_VarSelect}.
-
-To test the third hypothesis about climatic neighborhoods driving selection, we directly competed local with neighborhood climate to see which correlated with traits (Climate-TraitPC1 variables) and latitude (Climate-Latitude variables) better.
-We used the immediate collection location for local climate.
-For climate neighborhoods, we sampled climate at 1000 random points (at 90-m resolution) within a 62-km radius buffer around the collection collection.
-We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}.
-Significant spatial autocorrelation persisted for approximately 62 km.
-Since \textit{M.
-cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
-Climatic means and variances (see below) were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of variable selection algorithm
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelect.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Climatic variable selection]{Overview of method for identifying putative climatic selective agents underlying latitudinal cline.
-We looked for climate variables that explained both the latitude of 356 \textit{M.
-cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
-For Climate-Latitude variables we extracted climate data from recent occurences located throughout California and Oregon, USA (shown in map).
-For Climate-Trait variables, we extracted climatic data for the 16 focal populations.
-For both analyses, we extracted local and neighborhood climate.
-Local climate refers to climate only from where a population was collected ($x_{i,0}$).
-Neighborhood climate was calculated as the average over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M.
-cardinalis} is riparian.
-We identified climatic factors that most strongly predicted latitude of occurences (Climate-Latitude variables) and traits (Climate-Trait variables), as shown for hypothetical data in plots at the bottom of the figure.}
-	\label{fig:Fig_VarSelect}
-\end{figure}
-
-For Climate-Latitude analyses, we compiled a representative set of 356 recent (since 2000) known \textit{M.
-cardinalis} occurences.
-These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}.
-For both Climate-TraitPC1 analyses (16 focal populations) and Climate-Latitude (many populations), we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation.
-Monthly interpolated climate layers were calculated using ClimateWNA version 5.30 \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America.
-For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} version 1.1-1 \citep{Hijmans_etal_2016}.
-We included 24 climate factors, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}).
-The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range.
-For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation.
-Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation (CV).
-In total, the VSURF algorithm selected among 96 climate variables: 24 climatic factors $\times$ 2 types (30-year average and CV) $\times$ 2 spatial scales (local and neighborhood).
-
-%--------------------------------------------------
-% Results
-%--------------------------------------------------
-
-\section*{Results}
-
-\subsection*{A coordinated latitudinal cline in germination, growth, and photosynthesis}
-
-There are strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M.
-cardinalis}, as evidenced by large and significant population effects for these traits (Table~\ref{table:Table_AnovaSummary}).
-A single principal component captured 71.6 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}).
-As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.
-There were similar latitudinal clines for individual traits underlying PC1 (Fig S1-S4).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Table of Population, Treatment, Population x Treatment interactions
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% *** Commented out results from Intrinsic Photosynthetic rate *** %
-
-\begin{table}[ht]
-   \centering
-   \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects.
-We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2015}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2015}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}.
-Note that temperature and water treatments were imposed after germination, hence are not applicable to this trait.
-Complete analysis of variance/deviance tables for each trait are available in the Supporting Information.
-Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}
-   \resizebox{6.5in}{!}{
-   \begin{tabular}{@{} lllllll @{}}
-      \toprule
-    {\raggedleft Trait}                          & Germination & Leaf expansion & Stem elongation & Photosynthesis & Mortality \\
-    Statistical model              & survreg     & lmer           & lmer            & lmer           & glmer \\
-      \midrule
-  Population                       &
-    ***  &
-    *** &
-    *** &
-    *** & 
-     & \\
-  Temperature                      &
-    NA                                                          &
-    ***    &
-    *** &
-    ** & 
-    *** & \\
-  Water                            &
-    NA                                                          &
-    *   &
-       &
-     & 
-    *** & \\
-  Pop $\times$ Temp                &
-    NA                                                          &
-     &
-     &
-    * & 
-     & \\
-  Pop $\times$ Water               &
-     NA                                                          &
-    * &
-     &
-     & 
-     & \\
-  Temp $\times$ Water              &
-    NA                                                          &
-     &
-     &
-     & 
-    *** & \\
-  Pop $\times$ Temp $\times$ Water &
-    NA                                                          &
-     &
-     &
-     & 
-     & \\
-      \bottomrule
-   \end{tabular} }
-   \label{table:Table_AnovaSummary}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of latitude versus PC1 (growth axis)
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude.
-Each point is a population's latitude of origin (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of four traits (see Material and Methods).
-The line and 95\% confidence intervals were estimated using linear regression.}
-	\label{fig:Fig_PC1vLat}
-\end{figure}
-
-\subsection*{Little evidence for variation in plasticity}
-
-Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation.
-We found little evidence G$\times$E in \textit{M.
-cardinalis}.
-There were only two statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but these were not strong compared to Population and Temperature effects.
-Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
-Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects.
-However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
-Complete ANOVA tables are available in the Supporting Information (Table S3-S6).
-
-\subsection*{Neighborhood climatic variability best explains latitudinal cline}
-
-Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
-To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variable explains latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
-Based on exploratory variable selection analyses, one or two aspects of interannual climatic variation (hypothesis 1) and climatic neighborhoods (hypothesis 3) are likely more important drivers of trait divergence than average climate immediately surrounding populations.
-These general conclusions are based on the fact that interannual variation in climate averaged over 62-km radius climatic neighborhoods around populations correlated most strongly with trait variation and latitude of \textit{M.
-cardinalis} occurences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
-All 16 Climate-Latitude and 3 Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
-The most important climatic variables related to the interannual variation rather than average climate.
-
-Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates driving the latitudinal cline in \textit{M.
-cardinalis} (see Table.~\ref{table:TableS_ClimVars} for a key to climate variable abbreviations).
-Note that the coefficient of variation of a climatic factor is subscripted with $\sigma$ whereas the mean is subscripted with $\mu$.
-More specifically, greater winter precipitation variability and lower winter temperature variability are associted with Southern latitudes and higher TraitPC1 values (Fig.~\ref{fig:Fig_ClimPC1}A,B).
-Neighborhood interannual variation in winter precipitation (bio16$_\sigma$) was the most important Climate-Latitude variable (Fig.~\ref{fig:Fig_ClimVarImp}A).
-However, neighborhood bio16$_\sigma$ did not overlap with Climate-TraitPC1 variables (Fig.~\ref{fig:Fig_ClimVarImp}B).
-We nevertheless consider it a plausible candidate for two reasons.
-First, neighborhood bio16$_\sigma$ correlated strongly with TraitPC1 (Fig.~\ref{fig:Fig_ClimPC1}A).
-Second, one of the most important Climate-TraitPC1 variables (neighborhood bio15$_\sigma$; Fig.~\ref{fig:Fig_ClimPC1}B,C) is very similar to bio16$_\sigma$.
-In Mediterranean climates like California, most precipitation occurs in the wettest quarter (winter), so years with low winter precipitation also have low precipitation seasonality.
-Hence, highly variable year-to-year winter precipitation at lower latitude (Fig.~\ref{fig:Fig_ClimPC1}D) is closely associated with large swings in precipitation seasonality (Fig.~\ref{fig:Fig_ClimPC1}C).
-
-Interannual variation in temperature of the coldest quarter (neighborhood bio11$_\sigma$) is another plausible candidate because it was only variable in both Climate-Latitude and Climate-TraitPC1 analyses (Fig.~\ref{fig:Fig_ClimVarImp}).
-Neighborhood bio11$_\sigma$ explained more variation in TraitPC1 than latitude (latitude $r^2=0.55$ vs.
-bio11$_\sigma$ $r^2=0.6$; Fig.~\ref{fig:FigS_bio11PC1}), whereas neighborhood bio16$_\sigma$ did slightly worse (bio16$_\sigma$ $r^2=0.49$).
-Models using bio15$_\sigma$ or bio11$_\sigma$ to predict TraitPC1 also had significantly lower Akaike Information Criteria (AIC) than the latitude model (AIC of different models -- bio15$_\sigma$: 48.5; bio11$_\sigma$: 52.4; latitude: 54.5).
-The best two-factor model including both neighborhood bio15$_\sigma$ and bio11$_\sigma$ did not significantly improve explanatory power ($r^2=0.71$, AIC$=49.2$).
-In summary, either variation in precipitation or temperature seasonality may be important selective agents, but there is no strong evidence that they are both important.
-The most important Climate-TraitPC1 variable, neighborhood variation in mean diurnal range (bio2$_\sigma$; Fig.~\ref{fig:Fig_ClimVarImp}B) did not have any obvious similarity to Climate-Latitude variables.
-Given the large number of potential associations, we therefore think this may be a spuriously strong relationship.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of climatic variable importance
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimVarImp.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
-cardinalis} and trait variation]{Climatic variation integrated over climatic neighborhood is closely correlated with latitude of \textit{M.
-cardinalis} and trait variation.
-A.
-Using Random Forest regression, we identified 16 climatic variables significantly (high importance) associated with latitude of \textit{M.
-cardinalis} occurences.
-B.
-Only one of of the most important Climate-Latitude variables (in bold) was among the most important Climate-TraitPC1 variables.
-Note that the Importance values in A and B are not comparable because the dependent variables (Latitude and Trait PC1, respectively) are on different scales.
-Climatic variables (left of A; right of B) are defined by four qualities: Climatic factor -- Temperature (Temp), Precipitation (Prec), Heating degree-days (Heating), Snow (precipitation as snow); Temporal scale -- Annual, Coldest quarter (Cold Quar), Warmest Quarter (Warm Quar), Wettest quarter (Wet Quar), Driest Quarter (Dry Quar), or Seasonality; Type -- 30-year average (Avg.) or coefficient of variation (Var.); Spatial scale -- local or 62-km radius climatic neighborhood.}
-	\label{fig:Fig_ClimVarImp}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Figure of key Climate-TraitPC1 relationships
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ClimPC1.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Variation in precipitation is correlated with TraitPC1 and latitude]{Variation in precipitation is correlated with TraitPC1 and latitude.
-A.
-Greater values of TraitPC1 are associated with greater interannual variation in precipitation of the wettest quarter.
-This was the most important Climate-Latitude variable, but not among the most important Climate-TraitPC1 variables.
-B.
-However, a closely related parameter, interannaul variation in precipitation seasonality, was among the most important Climate-TraitPC1 variables.
-C.
-Across focal populations, variation in precipitation of the wettest quarter and seasonality are closely correlated.
-D.
-Southern populations of \textit{M.
-cardinalis} experience much greater interannual variationi in precipitation.
-In all panels, we report climatic neighborhood values (see Material and Methods).
-Regression lines, 95\% confidence intervals, and coefficients of determination ($R^2$) were calculated using linear regression.}
-	\label{fig:Fig_ClimPC1}
-\end{figure}
-
-%--------------------------------------------------
-% Discussion
-%--------------------------------------------------
-
-\section*{Discussion}
-
-We found evidence for one of two common signatures of local adaptation in the perennial herb \textit{Mimulus cardinalis}.
-Latitudinal clines in germination rate, photosynthesis, and growth, suggest adaptive differentiation in fundamental physiological traits of the species.
-However, we found little evidence that populations respond differently to temperature or drought.
-As we discuss below, this latter result may indicate that the fundamental abiotic niche is relatively conserved.
-Finally, we found that neighborhood climatic variation between years may be a more important selective agent than the local average climate.
-In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in physiology changes clinally, whereas plastic responses to temperature and drought are relatively similar.
-
-Evolutionary theory indicates that the shape of fitness tradeoffs, demography, and gene flow can constrain adaptation \citep{Levins_1968, Ronce_Kirkpatrick_2001} and hence the type of variation maintained within species.
-Specifically, adaptive variation cannot be maintained by spatially varying selection if tradeoffs are too strong, demography is strongly asymmetric, and/or maladaptive gene flow is too high.
-In \textit{M.
-cardinalis} we found substantial genetically based variation among populations along a phenotypic axis from fast to slow growth that varied over a large spatial scale (Fig.~\ref{fig:Fig_PC1vLat}).
-If this variation is adaptive, it suggests that the fitness tradeoff between low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
-That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatially heterogenous selection.
-We also know from previous work that population size does not vary strongly with latitude.
-Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range \citep{Paul_etal_2016}.
-Nevertheless, local gene flow from similar environments may shape how selection varies with latitude.
-Climatic neighborhoods, defined here as the average climate in a 62-km radius around populations, correlated more strongly with trait and population latitude than local climate.
-This may not greatly constrain local adaptation because local and neighborhood climate values were generally similar in \textit{M.
-cardinalis} populations (data not shown).
-
-Another possibility we could have seen is that southern populations, which appear to experience more frequent drought years (see below), could have adapted by tolerating drought better than northern populations, thereby expanding the fundamental niche of the species as a whole.
-We found no evidence for this; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
-Plants grew faster in the Hot treatment, but there was little effect of drought on growth.
-Rather, the effects of drought took longer to materialize but resulted in high mortality, especially in the Hot chamber.
-However, there was no differential mortality among populations in this treatment.
-We hypothesize that evolution of the fundamental niche may be constrained by a combination of strong fitness tradeoffs, demographic asymmetry, and gene flow.
-Riparian habitats where \textit{M.
-cardinalis} live are highly heterogeneous at small spatial scales.
-Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Mediterranean climates of western North America.
-But alleles that confer greater drought tolerance may be quite costly in well-watered soils, and \textit{vice versa}, leading to strong fitness tradeoffs.
-Such tradeoffs promote specialization to one soil type or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
-Demography and gene flow may reinforce niche conservatism.
-A new mutant with increased drought tolerance that can survive at the resource-poor margin of a population will be demographically overwhelmed by the larger census populations that can be maintained in higher-resource environments.
-Infrequent wet years may also produce most seeds, so selection is weighted towards alleles that have high fitness in the wet environment, even if dry years are more frequent.
-Finally, gene flow, which is generally high among \textit{M.
-cardinalis} populations within the same ecoregion \citep{Paul_etal_2016}, will thwart local adaptation and reinforce specialization.
-Thus, the spatial grain of the environment, demographic asymmetry, and gene flow may conspire to constrain local adaptation via altered fundamental niche.
-
-To reiterate, these results indicate that differences in physiology and growth, but not responses to temperature and drought, mediate local adaptation to climate in \textit{M.
-cardinalis}.
-Next, we would like to understand why variation in these particular traits may be adaptive.
-We argue that temporally more variable environments, as experienced by southern populations, select for a more `annualized` life-history strategy.
-
-Greater investment in aboveground growth, as opposed to belowground storage for future seasons, may be favoured in climates with more frequent drought years, but maladaptive in climates with more consistent precipitation.
-Suppose plants that grow quickly and allocate new resources to continued growth rather than storage have higher fitness over a single growing season.
-However, by not allocating resources to storage, these fast-growing plants begin future seasons at a deficit.
-Therefore, in a stable environment where winter survivorship is assured in most years, failure to store resources may reduce lifetime fitness.
-But the environment is not constant.
-For perennial herbs in Mediterranean climates, a dry winter (rainy season) can kill the rhizomes (underground stems that store nutrients for future growth) before emergence or aboveground stems before flowering.
-If drought years occur frequently enough, selection may favour the fast-growing strategy because there is no advantage to storage if drought kills plants before flowering.
-Considering life-history strategy as a continuum from no storage (annual) to lots of storage (perennial), we hypothesize that the optimal allocation to aboveground growth is more `annualized` in southern climates that have greater interannual variation in precipitation.
-
-The hypothesis that greater precipitation variability selects for an annualized life history is tentative, but consistent with theory, data from other species, and other observations in \textit{M.
-cardinalis}.
-Life history theory shows that less variable environments are one factor that favours the evolution of perenniality \citep{Stearns_1976, Iwasa_Cohen_1989, Friedman_Rubin_2015}.
-Populations of the perennial \textit{Plantago asiatica} show a similar latitudinal cline in growth and allocation to storage \citep{Sawada_etal_1994}, though these authors attribute the cline to variation in growing season length.
-There are also life history clines in the closely related species \textit{M.
-guttatus}, but the underlying traits and climatic drivers are quite different.
-Annual \textit{M.
-guttatus} flower sooner and produce fewer stolons in response to climates with shorter seasons and more intense summer drought \citep{Lowry_Willis_2010, Friedman_etal_2015, Kooyers_etal_2015}.
-In contrast, there are no truly annual (monocarpic and semelparous) populations of \textit{M.
-cardinalis}.
-Rather, our hypothesis states climatic variability selects on quantitative variation in allocation to growth versus storage.
-Several lines of evidence provide tentative support for this hypothesis.
-Preliminary surveys suggest that northern populations not only grow slower, but also produce greater numbers of rhizomes (C.D.
-Muir, unpub.
-data), suggesting an allocation tradeoff.
-Ecological niche models also show that occurrence of southern populations is best predicted by recent climate (< 5 years), whereas northern occurrences are best predicted by climate over the previous 30 years (M.
-Bayly \& A.
-Angert, unpub.
-data).
-Finally, demographic surveys of natural populations show greater variation in the size of recruits in southern populations, suggesting higher maximum growth rates under natural conditions (M.
-Bayly \& A.
-Angert, unpub.
-data).
-Exploratory analysis also pointed to variation in winter temperature as a potential selective agent.
-Specifically, winter temperature is more variable in the north, which may lead to less predictability in the start of the growing season or late frost.
-This could affect selection on phenology, which we did not study here, but it is not obviously connected to the latitudinal cline in physiological traits.
-
-In summary, we found evidence for a coordinated latitudinal cline in germination rate, photosynthesis, and growth, suggesting local adaptation.
-These differences between populations suggest different trait optima in different climates.
-We did not find evidence that the relative performance of populations shifts with temperature or watering regime, suggesting relatively little variation in plasticity.
-Exploratory analysis implicate that more variable precipitation regimes at lower latitude drive much of the latitudinal cline, though other climatic factors could also contribute.
-Interestingly, the climatic neighborhood may shape selective pressures than local climate.
-In the future, we will use field experiments to test whether greater variation in precipitation selects for faster growth and that selection on temperature/drought responses does not vary among populations.
-By doing so, we aim to understand why certain physiological and developmental mechanisms, but not others, contribute to local adaptation.
-
-% [Future work and broad conclusions].
-Supports general conclusions that traits related to timing and growth are evolutionarily labile whereas those related to the fundamental niche are constrained (cites like \cite{Emery_etal_2012, Emery_Ackerly_2014}).
-Next, we are testing whether phenotypic variation and constraint can be predicted by the shape of tradeoffs, as predicted by evolutionary theory.
-\cdm{I will work more on this once I am more comfortable with the rest of the discussion}.
-
-% niche-tracking versus niche-shifting
-% One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve.
-For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.
-
-% citations on life history allocation tradeoff between growth and reproduction:
-%% THEORY:
-% Bell 1980 - briefly mentionds on page 56 that increased environmental variation for adults will favor semelparity
-% Iwasa and Cohen 1989, pg.
-496 "A larger reliability of habitat, a higher storage efficiency, higher productivity of the environment, and a longer growing season, all tend to promote the evolution of perennial life, as suggested by previous models of life-historsy strategy (Stearns 1976).
-
-%% EMPIRICAL:
-% Sawada et al. 1994 - latitudinal cline dry matter accumulation and storage allocation
-% Stanton et al 2000 show selection in stressful environemts (artificual selection over 5 generations) favors stress avoidance rather than tolerance traits.
-i.e.
-might be easier to adjust life history rather than tolerance.
-% Hall and Willis 2006 find selection on flowering time in coastal-inland comparison of M guttatus
-% Ehleringer et al 1990 find lower WUE increased yield in ephemeral pops
-% Bazzaz 1979: higher photo in fast growing plants (early succession)
-
-% Reznick 1985; Lovett Doust 1989
-% Williams 1966; Charnov and Schaffer 1973; Bell 1980; Reznick 1985; Biere 1995; Johnson 2007).
-% for iteroparous sp: van Noordwijk and De Jong 1986.
-% from Remington et al.
-2013: "Genetic variation in traits subject to trade-offs results in what has been termed “structured pleiotropy,” in which genetic covariances between traits result from functional constraints imposed by the limiting resources (De Jong 1990; Stearns et al.
-1991; see Figure 1B).
-
-%--------------------------------------------------
-% Acknowledgements
-%--------------------------------------------------
-
-\section*{Acknowledgements}
-Erin Warkman and Lisa Lin helped collect data.
-CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program.
-ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
-
-%--------------------------------------------------
-% References
-%--------------------------------------------------
-
-\setlength{\bibsep}{6pt}
-\bigskip
-
-\bibliography{../../CommonResources/refs}
-\bibliographystyle{evolution}
-
-\clearpage
-
-%--------------------------------------------------
-% Supporting Information
-%--------------------------------------------------
-
-\section*{Supporting Information}
-
-% Modify and restart table/figure numbering for appendixes
-\renewcommand\thefigure{S\arabic{figure}}    
-\renewcommand\thetable{S\arabic{table}}    
-\renewcommand\theequation{S\arabic{equation}}    
-\setcounter{table}{0}    
-\setcounter{equation}{0}
-\setcounter{figure}{0}
-
-%--------------------------------------------------
-% Supporting Tables
-%--------------------------------------------------
-
-\subsection*{Supporting Tables}
-
-% DIC table for initial size (LLL on 5/12)
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[DIC table for initial size of seedlings]{Initial size of seedlings did not vary among Populations, Families, or Treatments.
-We used a censored Gaussian model of initial size at the outset of the experiment (longest leaf length of the first true leaves).
-The model was censored because we could not accurately measure leaves less than 0.25 mm with digital callipers (217 of 702, 30.9\%, were too small).
-We fit models using a Bayesian MCMC method implemented using the MCMCglmm function with default priors in the R package \pkg{MCMCglmm} version 2.17 \citep{Hadfield_2010}.
-We estimated the posterior distribution from 1000 samples of an MCMC chain run for $10 ^ 5$ steps after a $10^4$ step burn-in.
-We used step-wise backward elimination procedure to find the best-supported model according to Deviance Information Criterion (DIC).}
-	\begin{center}
-	\begin{tabular}{>{\everypar{\hangindent1cm}{}\raggedright}p{6cm}lc}
-	\toprule
-
-	\input{Tables/Table_InitialSize.txt}
-
-	\bottomrule
-	\end{tabular}
-	\end{center}
-	\label{table:TableS_InitialSize}
-\end{table}
-
-\begin{table}[ht]
-	\caption[Climate variables]{Climatic variables used }
-   \centering
-   \begin{tabular}{@{} lp{12cm} @{}}
-      \toprule
-   Abbreviation & Climate variable \\
-      \midrule
-  % climateWNA VARIABLES:
-	DD\_0	& degree-days below 0\celsius (chilling degree-days)  \\
-	DD5	  & degree-days above 5\celsius (growing degree-days)   \\
-	DD\_18	& degree-days below 18\celsius (heating degree-days)  \\
-	DD18	& degree-days above 18\celsius (cooling degree-days)  \\
-	NFFD	& number of frost-free days                           \\
-	PAS	  & precipitation as snow (mm) between August in previous year and July in current  \\
-	Eref	& Hargreaves reference evaporation (mm) \\
-	CMD	  & Hargreaves climatic moisture deficit (mm) \\
-	RH    &	mean annual relative humidity \\
-  % BIOCLIM VARIABLES:
-  bio1	&	annual mean temperature	\\
-	bio2	&	mean diurnal range (mean of monthly (max temp - min temp))	\\
-	bio3	&	isothermality (bio2/bio7) (* 100)	\\
-	bio4	&	temperature seasonality (standard deviation *100)	\\
-	bio5	&	max temperature of warmest month	\\
-	bio6	&	min temperature of coldest month	\\
-	bio7	&	temperature annual range (bio5-bio6)	\\
-	bio8	&	mean temperature of wettest quarter	\\
-	bio9	&	mean temperature of driest quarter	\\
-	bio10	&	mean temperature of warmest quarter	\\
-	bio11	&	mean temperature of coldest quarter	\\
-	bio12	&	annual precipitation	\\
-	bio15	&	precipitation seasonality (coefficient of variation)	\\
-	bio16	&	precipitation of wettest quarter	\\
-	bio17	&	precipitation of driest quarter	\\
-	bio18	&	precipitation of warmest quarter	\\
-	bio19	&	precipitation of coldest quarter	\\
-	    \bottomrule
-   \end{tabular}
-   \label{table:TableS_ClimVars}
-\end{table}
-
-% ANOVA table for LER
-
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, leaf expansion rate]{Analysis of varianace (ANOVA) table on leaf expansion rate (LER) using \pkg{lmerTest} \citep{Kuznetsova_etal_2016}.
-Family and Block were included as random effects.
-Abbreviations: SS = sum of squares; MS = mean sum of squares (SS / df1); df1 = numerator degrees of freedom; df2 = denominator degrees of freedom.}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
-  Day to Germination & 12.12 & 
-    12.12 & 1 & 
-    637 & 35.21 &
-    4.9 $\times10^{-9}$ \\
-  Population & 22.22 & 1.48 & 
-    15 & 118 & 
-    4.3 & 2.5 $\times10^{-6}$ \\
-  Temperature & 80.42 & 
-    80.42 & 1 &
-    5 & 233.61 &
-    2.6 $\times10^{-5}$ \\
-  Water & 4.1 & 4.1 & 
-    1 & 5 & 
-    11.92 & 0.019 \\
-  Temperature $\times$ Water & 0.03 & 
-    0.03 & 
-    1 & 
-    4 & 
-    0.07 &
-    0.801 \\
-  Population $\times$ Temperature & 2.76 & 
-    0.18 & 
-    15 & 
-    547 & 
-    0.53 &
-    0.925 \\
-  Population $\times$ Water & 9.66 & 
-    0.64 & 
-    15 &
-    562 &
-    1.87 & 
-    0.024 \\
-  Population $\times$ Temperature $\times$ Water & 
-    4.11 & 
-    0.27 & 
-    15 & 
-    530 & 
-    0.78 & 
-    0.700 \\
-	\bottomrule
-	\end{tabular}
-	\label{table:TableS_fitLER}
-	\end{center}
-\end{table}
-
-% ANOVA table for SER
-
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, stem elongation rate]{Analysis of varianace (ANOVA) table on stem elongation rate (SER) using \pkg{lmerTest} \citep{Kuznetsova_etal_2016}.
-Family and Block were included as random effects.
-Abbreviations: SS = sum of squares; MS = mean sum of squares (SS / df1); df1 = numerator degrees of freedom; df2 = denominator degrees of freedom.}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
-  Day to Germination & 3.6 & 
-    3.6 & 1 & 
-    662 & 21.1 &
-    5.1 $\times10^{-6}$ \\
-  Population & 12 & 0.8 & 
-    15 & 113 & 
-    4.7 & 5.8 $\times10^{-7}$ \\
-  Temperature & 12.4 & 
-    12.4 & 1 &
-    6 & 72.8 &
-    1.5 $\times10^{-4}$ \\
-  Water & 0.6 & 0.6 & 
-    1 & 5 & 
-    3.7 & 0.113 \\
-  Temperature $\times$ Water & 0.9 & 
-    0.9 & 
-    1 & 
-    4 & 
-    5.2 &
-    0.093 \\
-  Population $\times$ Temperature & 3.6 & 
-    0.2 & 
-    15 & 
-    549 & 
-    1.4 &
-    0.126 \\
-  Population $\times$ Water & 2.8 & 
-    0.2 & 
-    15 &
-    536 &
-    1.1 & 
-    0.330 \\
-  Population $\times$ Temperature $\times$ Water & 
-    1.5 & 
-    0.1 & 
-    15 & 
-    518 & 
-    0.6 & 
-    0.874 \\
-	\bottomrule
-	\end{tabular}
-	\label{table:TableS_fitSER}
-	\end{center}
-\end{table}
-
-% ANOVA table for photosynthesis
-
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[ANOVA table, photosynthesis]{Analysis of varianace (ANOVA) table on photosynthetic rate using \pkg{lmerTest} \citep{Kuznetsova_etal_2016}.
-Family and Block were included as random effects.
-Abbreviations: SS = sum of squares; MS = mean sum of squares (SS / df1); df1 = numerator degrees of freedom; df2 = denominator degrees of freedom.}
-	\begin{center}
-	\begin{tabular}{lcccccc}
-	\toprule
-
-  & SS & MS & df1 & df2 & \em{F}-value & $P$-value \\ \midrule
-  Population & 347.7 & 
-    23.2 & 15 &
-    78 & 3.02 &
-    7.5 $\times10^{-4}$ \\
-  Temperature & 134.1 & 
-    134.1 & 1 &
-    6 & 17.46 &
-    6.4 $\times10^{-3}$ \\
-  Water & 51 & 
-    51 & 1 &
-    4 & 6.64 &
-    0.066 \\
-  Temperature $\times$ Water & 0.7 & 
-    0.7 & 
-    1 & 
-    3 & 
-    0.09 &
-    0.781 \\
-  Population $\times$ Temperature & 218.6 & 
-    14.6 & 
-    15 & 
-    263 & 
-    1.9 &
-    0.024 \\
-  Population $\times$ Water & 87.7 & 
-    5.8 & 
-    15 &
-    233 &
-    0.76 & 
-    0.724 \\
-  Population $\times$ Temperature $\times$ Water & 
-    91.4 & 
-    6.1 & 
-    15 & 
-    208 & 
-    0.79 & 
-    0.686 \\
-	\bottomrule
-	\end{tabular}
-	\label{table:TableS_fitPhoto}
-	\end{center}
-\end{table}
-
-% Analysis of deviance table for mortality
-
-\begin{table}[htbp]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Analysis of deviance table, mortality]{Analysis of deviance table on the probability of mortality by the end of the experiment using Type-II Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
-Family and Block were included as random effects.
-Abbreviations: df = degrees of freedom}
-	\begin{center}
-	\begin{tabular}{lccc}
-	\toprule
-
-  & $\chi ^2$ & df & $P$-value \\ \midrule
-  Population & 32 & 
-    31 & 0.419 \\
-  Temperature & 31.8 & 
-    6 & 1.8 $\times10^{-5}$ \\
-  Water & 69.2 & 
-    12 & 4.6 $\times10^{-10}$ \\
-  Temperature $\times$ Water & 20.7 & 
-    1 & 
-    5.3 $\times10^{-6}$ \\
-  Population $\times$ Temperature & 5.6 & 
-    15 & 
-    0.985 \\
-  Population $\times$ Water & 8.6 & 
-    15 & 
-    0.897 \\
-  Population $\times$ Temperature $\times$ Water & 
-    0.2 & 
-    15 & 
-    1.000 \\	\bottomrule
-	\end{tabular}
-	\label{table:TableS_fitMort}
-	\end{center}
-\end{table}
-
-% Table of Climate-Latitude and Climate-TraitPC1 variables
-
-\begin{table}[ht]
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-  \caption[Important climatic variables]{Important climatic variables predicting latitude of \textit{M.
-cardinalis} populations (`Climate-Latitude') and the first principal component of traits measured in a common garden (`Climate-TraitPC1').
-Local climatic variables were measured from the exact location of collection; neighborhood climatic variables were averaged from a 62-km neighborhood around population (see Material and Methods).
-Importance and significance were determined using the variable selection using random forests (VSURF) algorithm (see Material and Methods).
-Climatic variables are described in Table~\ref{table:TableS_ClimVars}.
-$\mu$ signifies the mean of the climate variables from 1981--2010; $\sigma$ indicates coeffiecient of variation among years.
-}
-   \begin{center}
-   \begin{tabular}{@{} ll @{}}
-      \toprule
-    	Climate-Latitude variables & Climate-TraitPC1 variables \\
-      \midrule
-			\input{Tables/TableS_ClimVarImp.txt}
-			\bottomrule
-   \end{tabular}
-   \end{center}
-   \label{table:TableS_ClimVarImp}
-\end{table}
-
-\clearpage
-
-%--------------------------------------------------
-% Supporting Figures
-%--------------------------------------------------
-
-\subsection*{Supporting Figures}
-
-% Figures showing latitudinal clines with specific traits
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_GermLat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations germinate faster]{Southern populations germinate faster.
-Each point is a population of \textit{M.
-cardinalis} showing its latitude of origin (x-axis) and model-predicted days to germination in days under growth chamber conditions (see Material and Methods).
-Bars around each point are 95\% confidence intervals.
-Predicted time to germination and confidence intervals are based on survival regression (see Materials and Materials).
-The line is the linear regression of log(model-predicated days to germination) $\sim$ latitude.
-The $P$-value of the regression is given in the upper left corner.}
-	\label{fig:FigS_GermLat}
-\end{figure}
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_LERlat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (leaf expansion rate).]{Southern populations grow faster.
-Each point is a population of \textit{M.
-cardinalis} showing its latitude of origin (x-axis) and model-predicted leaf expansion rate during the rosette phase.
-Bars around each point are 95\% confidence intervals.
-Predicted leaf expansion rate based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
-The line is the linear regression of model-predicated leaf expansion rate $\sim$ latitude.
-The $P$-value of the regression is given in the upper right corner.}
-	\label{fig:FigS_LERlat}
-\end{figure}
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_SERlat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations grow faster (stem elongation rate).]{Southern populations grow faster.
-Each point is a population of \textit{M.
-cardinalis} showing its latitude of origin (x-axis) and model-predicted stem elongation rate.
-Bars around each point are 95\% confidence intervals.
-Predicted stem elongation rate based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
-The line is the linear regression of model-predicated stem elongation rate $\sim$ latitude.
-The $P$-value of the regression is given in the upper right corner.}
-	\label{fig:FigS_SERlat}
-\end{figure}
-
-\begin{figure}[h!]
-	\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_Photolat.pdf}}
-	\usefont{T1}{phv}{m}{n}
-	\fontsize{10}{12}
-	\selectfont
-	\caption[Southern populations photosynthesize faster.]{Southern populations photosynthesize faster.
-Each point is a population of \textit{M.
-cardinalis} showing its latitude of origin (x-axis) and model-predicted instantaneous photosynthetic rate.
-Bars around each point are 95\% confidence intervals.
-Predicted photosynthetic rates based least-square mean estimates and confidence intervals were calculated from linear mixed-effects models (see Materials and Materials).
-The line is the linear regression of model-predicated photosynthetic rate $\sim$ latitude.
-The $P$-value of the regression is given in the upper right corner.}
-	\label{fig:FigS_Photolat}
-\end{figure}
-
-% Figure of bio11_cva_SA versus TraitPC1
-
-\begin{figure}[h!]
-\centerline{\includegraphics[width=0.5\textwidth]{Figures/FigureS_bio11PC1.pdf}}
-\usefont{T1}{phv}{m}{n}
-\fontsize{10}{12}
-\selectfont
-\caption[Trait PC1 versus bio11_cva_SA]{Trait variation, from fast to slow growth, is closely associated with neighborhood variation in temperature of the coldest quarter (bio11$_\sigma$) Each point is a population coefficient of variation in bio11 averaged over a 62-km climatic neighborhood  (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of four traits (see Material and Methods).
-The line and 95\% confidence intervals were estimated using linear regression.}
-\label{fig:FigS_bio11PC1}
-\end{figure}
-
-\clearpage
-
-\subsection*{Supporting Material and Methods}
-
-\subsubsection*{Temperature treatments}
-
-We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (WWC, Hot) and Little Jameson (LIJ, Cool).
-We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}.
-This range was chosen because seeds used in the experiment were collected around 2012, thus their presence in that location at that time suggests that populations were able to persist there for at least some years before collection.
-Monthly temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A.
-Angert, unpub.
-data).
-Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M.
-cardinalis} in nature.
-We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}.
-The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ).
-The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals.
-We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R.
-Simulated data closely resembled the autocorrelation and between-site correlation of the actual data.
-From simulated mean temperature, we next selected minimum and maximum daily temperatures.
-Mean, min, and max temperature were highly correlated at both sites.
-We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally distributed random deviates with variance equal to the residual variance of the linear models.
-For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature.
-During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.
-
-\subsubsection*{Watering treatments}
-
-%*** need to check chiller temp and make of chiller
-For watering treatments, we simulated two extreme types of streams where \textit{M.
-cardinalis} grows.
-In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season.
-In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season due to rain and snow melt, but gradually dries down through the summer.
-In both treatments, plants were bottom-watered using  water chilled to 7.5\celsius.
-Plants in the well-watered treatment were fully saturated every two hours during the day.
-Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing).
-Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.
-
-\end{document}

### 2017-08-29 ms/ms.Rnw
starting revision

@@
 
-%
-% Where to find files
-%
+% Setup
 
@@
 
-  # The path to the ms directory may need to be changed for different users
   # Manuscript files
-  pathMS <- "~/Google Drive/CardAdapt/ms" # Chris' MacBook Air
-  # pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop
-  # pathMS <- "/Users/amylauren/Desktop/CardAdapt/ms" # Amy's desktop
+  pathMS <- getwd() # for sweave
+  # pathMS <- "ms" # for debugging
 
@@
   # Source custom functions
-  source(paste(pathMS, "/functions.R", sep = ""))
+  #source(paste(pathMS, "/functions.R", sep = ""))
+  source("functions.R")
 
@@
   # Individuals per population per treatment
-  indPerPop <- tapply(master$indiv, list(master$TempTrt, master$WaterTrt, master$PopID), length)
+  indPerPop <- tapply(master$indiv, list(master$TempTrt, master$WaterTrt, 
+                                         master$PopID), length)
 
@@
 # Watering axis
-axis(4, at = c(0, 40), lwd = 0, lwd.ticks = 1, las = 1, labels = c(1, 12), cex.axis = 1.5)
+axis(4, at = c(0, 40), lwd = 0, lwd.ticks = 1, las = 1, labels = c(1, 12), 
+     cex.axis = 1.5)
 mtext(expression(paste("Watering Freq (no. Day", phantom()^-1, ")")), 4,
@@
   par(mai = c(0.5, 1, 0.5, 0.25))
-  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, xlim = c(1, 4), 
-       ylim = c(1, 3.5))
+  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, 
+       xlim = c(1, 4), ylim = c(1, 3.5))
   abline(v = 2.5)
@@
   axis(2, lwd = 0, lwd.ticks = 1, las = 1)
-  title(ylab = expression(paste("Leaf expansion rate (mm ", day ^-1, ")")), cex.lab = 1.5)
+  title(ylab = expression(paste("Leaf expansion rate (mm ", day ^-1, ")")), 
+        cex.lab = 1.5)
 
@@
     lines(c(1, 2.75), c(tab[i, "Wet", "Cool"], tab[i, "Wet", "Hot"]), col = "grey")
-    lines(c(2.25, 4), c(tab[i, "Dry", "Cool"], tab[i, "Dry", "Hot"]), col = "grey", lty = 2)
+    lines(c(2.25, 4), c(tab[i, "Dry", "Cool"], tab[i, "Dry", "Hot"]), col = "grey",
+          lty = 2)
   }
@@
   tab <- tapply(newdata$SER_AbsGrowth, 
-                list(newdata$Population, newdata$WaterTrt, newdata$TempTrt), mean, na.rm = T)
+                list(newdata$Population, newdata$WaterTrt, newdata$TempTrt), mean, 
+                na.rm = T)
   pdf(paste(pathFig, "/FigureS_SERrxnNorm.pdf", sep = ""), 4, 4, useDingbats = F)
   par(mai = c(0.5, 1, 0.5, 0.25))
-  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, xlim = c(1, 4), 
-       ylim = c(0.15, 0.50))
+  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, 
+       xlim = c(1, 4), ylim = c(0.15, 0.50))
   abline(v = 2.5)
@@
   axis(2, lwd = 0, lwd.ticks = 1, las = 1)
-  title(ylab = expression(paste("Stem elongation rate (cm ", day ^-1, ")")), cex.lab = 1.5)
+  title(ylab = expression(paste("Stem elongation rate (cm ", day ^-1, ")")), 
+        cex.lab = 1.5)
 
@@
   tab <- tapply(master$Photo, 
-                list(master$Population, master$WaterTrt, master$TempTrt), mean, na.rm = T)
+                list(master$Population, master$WaterTrt, master$TempTrt), mean, 
+                na.rm = T)
   pdf(paste(pathFig, "/FigureS_PhotoRxnNorm.pdf", sep = ""), 4, 4, useDingbats = F)
@@
   par(mai = c(0.5, 1, 0.5, 0.25))
-  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, xlim = c(1, 4), 
-       ylim = c(0, 1))
+  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, 
+       xlim = c(1, 4), ylim = c(0, 1))
   abline(v = 2.5)
@@
   ##### TEMPORARY: CODE TO EXPORT FIGURE FOR TALK #####
-  d <- subset(master, !is.na(master$DiedFromStress))
-  x <- tapply(d$DiedFromStress, list(d$PopID, d$WaterTrt, d$TempTrt), function(X) 1 - length(which(X)) / length(X))
-
-  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/DroughtResponse.pdf", w = 5.5, h = 5, useDingbats = F)
-  par(mai = c(1, 1.5, 0.25, 0.25))
-
-  plot(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], axes = F, frame.plot = T,
-       pch = 21, col = "black", bg = primShade0, cex = 2, type = "n",
-       xlab = "Precipitation of Wettest Quarter (mm)", 
-       ylab = "Relative Survivorship\n(Hot/Dry : Hot/Wet)", cex.lab = 1.5)
-  points(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], pch = 21, col = "black", 
-         bg = primShade0, cex = 1.5, lwd = 2)
-  axis(1, lwd = 0, lwd.ticks = 1)
-  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
-  dev.off()
+  # d <- subset(master, !is.na(master$DiedFromStress))
+  # x <- tapply(d$DiedFromStress, list(d$PopID, d$WaterTrt, d$TempTrt), function(X) 1 - length(which(X)) / length(X))
+  # 
+  # pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/DroughtResponse.pdf", w = 5.5, h = 5, useDingbats = F)
+  # par(mai = c(1, 1.5, 0.25, 0.25))
+  # 
+  # plot(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], axes = F, frame.plot = T,
+  #      pch = 21, col = "black", bg = primShade0, cex = 2, type = "n",
+  #      xlab = "Precipitation of Wettest Quarter (mm)", 
+  #      ylab = "Relative Survivorship\n(Hot/Dry : Hot/Wet)", cex.lab = 1.5)
+  # points(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], pch = 21, col = "black", 
+  #        bg = primShade0, cex = 1.5, lwd = 2)
+  # axis(1, lwd = 0, lwd.ticks = 1)
+  # axis(2, lwd = 0, lwd.ticks = 1, las = 1)
+  # dev.off()
+  #
 
@@
   # Plot latitude versus PC1
-  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/Figure_PC1vLat.pdf", 5.5, 5, useDingbats = F)
-  par(mai = c(1, 1.5, 0.25, 0.25))
-  plot(focClim$Lat, focClim$PC1, xlim = c(32, 44), ylim = c(-4, 4), axes = F,
-    frame.plot = T, type = "n", xlab = "", ylab = "")
-  title(xlab = "Latitude", cex.lab = 1.5)
-  title(ylab = "TraitPC1", cex.lab = 2, line = 5)
-  title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
-    list(x = pve)))
-  axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), 
-		expression(36*degree), expression(40*degree), expression(44*degree)), 
-		lwd = 0, lwd.ticks = 1) 
-  axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
-  fitLat <- lm(PC1 ~ Latitude, data = focClim)
-
-	# Label y-axis from slow to fast growth
-	mtext(c("Slow\nGrowth", "Fast\nGrowth"), 2, at = c(-4, 4), cex = 1.25, 
-				font = 3, las = 1, line = 4, adj = 0.5)
-
-  # polygon of confidence intervals
-  x <- seq(min(focClim$Latitude), max(focClim$Latitude), l = 1e3)
-  u <- predict(fitLat, new = data.frame(Latitude = x))
-  s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitLat) %*% c(1, X)))
-  confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
-  											upperCI = qnorm(0.975, u, s))
-  polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
-    col = "grey", border = NA)
-  lines(x, confInt$lowerCI, lwd = 2, lty = 2)
-  lines(x, confInt$upperCI, lwd = 2, lty = 2)
-  lines(range(focClim$Lat), rev(range(predict(fitLat))), lwd = 2)
-  points(focClim$Lat, focClim$PC1, pch = 21, col = "black", bg = primShade0, 
-  			 cex = 1.5, lwd = 2)
-
-  # P-value
-  text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1),
-	     labels = bquote(italic(P) == .(round(summary(fitLat)[[4]][2, 4], 3))))
-
-  dev.off()
+#   pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/Figure_PC1vLat.pdf", 5.5, 5, useDingbats = F)
+#   par(mai = c(1, 1.5, 0.25, 0.25))
+#   plot(focClim$Lat, focClim$PC1, xlim = c(32, 44), ylim = c(-4, 4), axes = F,
+#     frame.plot = T, type = "n", xlab = "", ylab = "")
+#   title(xlab = "Latitude", cex.lab = 1.5)
+#   title(ylab = "TraitPC1", cex.lab = 2, line = 5)
+#   title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
+#     list(x = pve)))
+#   axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), 
+# 		expression(36*degree), expression(40*degree), expression(44*degree)), 
+# 		lwd = 0, lwd.ticks = 1) 
+#   axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
+#   fitLat <- lm(PC1 ~ Latitude, data = focClim)
+#   
+# 	# Label y-axis from slow to fast growth
+# 	mtext(c("Slow\nGrowth", "Fast\nGrowth"), 2, at = c(-4, 4), cex = 1.25, 
+# 				font = 3, las = 1, line = 4, adj = 0.5)
+# 
+#   # polygon of confidence intervals
+#   x <- seq(min(focClim$Latitude), max(focClim$Latitude), l = 1e3)
+#   u <- predict(fitLat, new = data.frame(Latitude = x))
+#   s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitLat) %*% c(1, X)))
+#   confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
+#   											upperCI = qnorm(0.975, u, s))
+#   polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
+#     col = "grey", border = NA)
+#   lines(x, confInt$lowerCI, lwd = 2, lty = 2)
+#   lines(x, confInt$upperCI, lwd = 2, lty = 2)
+#   lines(range(focClim$Lat), rev(range(predict(fitLat))), lwd = 2)
+#   points(focClim$Lat, focClim$PC1, pch = 21, col = "black", bg = primShade0, 
+#   			 cex = 1.5, lwd = 2)
+#   
+#   # P-value
+#   text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1),
+# 	     labels = bquote(italic(P) == .(round(summary(fitLat)[[4]][2, 4], 3))))
+# 
+#   dev.off()
 

### 2017-08-29 ms/ms.tex
starting revision

@@
 
+% Setup
+
 %
-% Where to find files
+% Title and authors
 %
+
+\title{Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}}
+\author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
+\date{}
+
+%
+% Start document
+%
+
+\usepackage{Sweave}
+\begin{document}
+\input{ms-concordance}
+\maketitle
+
+$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada \\
+$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca \\
+
+Running Head: Latitudinal cline and climate in \textit{Erythranthe} \\
+
+Key words: local adaptation, cline, photosynthesis, growth rate, monkeyflower, \textit{Mimulus} \\
+
+Data will be archived on Dryad upon publication. \\
+
+%--------------------------------------------------
+% Acknowledgements
+%--------------------------------------------------
+
+\section*{Acknowledgements}
+Erin Warkman and Lisa Lin helped collect data.
+CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program.
+ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
+Two anonymous referees provided constructive comments on earlier versions of this manuscript.
+
+\clearpage
+
+\setstretch{\stretchy}
+
+\begin{center}
+{\LARGE  \centering Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}}
+\end{center}
+
+\section*{Abstract}
+
+Local adaptation is commonly observed in nature: organisms perform well in their natal environment, but poorly outside it.
+Correlations between traits and latitude, or latitudinal clines, are among the most common pieces of evidence for local adaptation, but identifying the traits under selection and the selective agents is challenging.
+Here, we investigated a latitudinal cline in growth and photosynthesis across 16 populations of the perennial herb \textit{Erythranthe cardinalis} (Phrymaceae).
+Using machine learning methods, we identify interannual variation in precipitation as a likely selective agent: Southern populations from more variable environments had higher photosynthetic rates and grew faster.
+We hypothesize that selection may favor a more annualized life history -- grow now rather than save for next year -- in environments where severe droughts occur more often.
+Thus our study provides insight into how species may adapt if Mediterranean climates become more variable due to climate change.
+
+\section*{Introduction}
+
+% Toggle these commands to get version for doing word count
+\linenumbers
+% \pagenumbering{gobble}
+% \raggedright
+
+% Local adaptation and clines
+Local adaptation has been documented within numerous species; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Leimu_Fischer_2008, Hereford_2009}.
+However, the prevalance of local adaptation remains difficult to assess because researchers rarely test for local adaptation unless there are obvious phenotypic or environmental differences (but see \citeauthor{Hereford_Winn_2008} \citeyear{Hereford_Winn_2008}).
+When local adaptation occurs, it frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977, Barton_1999}.
+Phenotypic differences between populations along a cline often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}.
+Despite a long history of studying local adaptation and clines, it remains challenging to identify exactly which traits are under selection and which differ for nonadaptive reasons.
+In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment.
+A related problem is identifying which of the myriad and often covarying aspects of the environment causes spatially varying selective pressures.
+
+% Intrinsic versus plastic variation 
+When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}.
+The traits that underlie local adaptation, however, need not mirror this pattern.
+Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away.
+Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive.
+We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively.
+Both intrinsic and plastic trait variation can be explained by genetic differences and both are involved in adaptation.
+For example, intrinsic differences in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
+Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
+In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
+
+% Selective agents (trait-enviro assoc)
+A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments.
+Either intrinsic and/or plastic variation should vary clinally along environmental gradients.
+Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
+For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}.
+Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008, Stock_etal_2014}, and drought response \citep{Kooyers_etal_2015} that likely relate to climatic variation.
+
+Despite the fact that latitudinal clines have been studied for a long time, latitude \textit{per se} cannot be a selective agent.
+Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days.
+Latitude may also correlate with the strength of biotic interactions \citep{Schemske_etal_2009} or other nonclimatic aspects of the environment, though as we explain below, we do not yet have compelling data that these are important in our study system.
+Hence, we focus on whether latitude could be an effective proxy for an underlying climatic driver, in which case we would expect a yet stronger relationship between traits and the key climatic variable(s) driving selection.
+Alternatively, latitude may be more strongly related to traits than any single climatic variable for at least two reasons.
+First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline.
+Alternatively, gene flow among neighbouring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration \citep{Slatkin_1978, Paul_etal_2011, Hadfield_2016}.
+We refer to this as the `climatic neighborhood'.
+For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a lower latitude set of populations will experience warmer climate than a higher latitude one.
+Thus, any particular low latitude population would be warm-adapted, even if it was located in a cooler (e.g.
+high elevation) site.
+Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g.
+coastal vs.
+continental), dissecting the evolution of latitudinal clines across many species will help identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, and whether local or regional climate shapes selective pressures.
+
+In this study, we investigated two major questions: 1) whether intrinsic or plastic physiological trait variation corresponds with latitude; and 2) what climatic factor(s) could plausibly be responsible for latitudinal clines.
+Within question 2, we tested three hypotheses outlined in the previous paragraph: latitudinal clines are explained by a single dominant climatic factor, multiple climatic factors, or the climatic neighborhood experienced by nearby population connected through gene flow.
+These hypotheses are not mutually exclusive since, for example, single or multiple factors in a climatic neighborhood may lead to latitudinal clines.
+We focused on climate because climate often determines and where species are found and also can exert strong selection on populations within species, though we acknowledge that other abiotic and biotic factors could also contribute to selection and the overall pattern of local adaptation.
+There is also a compelling need to know how populations are (or are not) locally adapted to climate so as to predict how they will respond to climate change \citep{Aitken_Whitlock_2013}.
+
+We examined these questions in \textit{Erythranthe cardinalis} (formerly \textit{Mimulus cardinalis} [\citeauthor{Nesom_2014} \citeyear{Nesom_2014}]) because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions.
+Many classic and contemporary studies of local adaptation use \textit{Mimulus sensu lato} species because of their natural history, easy propagation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}.
+Yet, there is a deficiency of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010, Wright_etal_2013}.
+We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
+We found a latitudinal cline of intrinsic variation in photosynthesis and growth, but little evidence for variation in plasticity.
+Interannual variation in precipitation and temperature are associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{E.
+cardinalis}.
+The climatic neighborhoods around populations explained trait variation better than local climate, indicating that latitudinal clines may be common because latitude integrates effects of selection on populations connected through gene flow.
+We place these findings in the context of life history theory and consider future directions in the Discussion.
+
+\section*{Material and Methods}
+
+\subsection*{Population Selection}
+
+\textit{E.
+cardinalis} is a perennial forb native to the Western US (California and Oregon).
+It is predomintantly outcrossing, self-compatible, and pollinated primarily by hummingbirds.
+We used 16 populations from throughout the range of \textit{E.
+cardinalis} (Table~\ref{table:Table_FocalPops}).
+These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below).
+Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
+We used seeds from 154 families, 4--12 (mean = 9.6, median = 12) families per population.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of focal population
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}[ht]
+   \centering
+   \topcaption[Focal Populations]{Latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
+   \begin{tabular}{@{} llll @{}}
+      \toprule
+  Name  & Latitude  & Longtiude  & Elevation (mas) \\
+      \midrule
+	Hauser Creek & 32.657	& 
+    -116.532	& 799   \\
+	Cottonwood Creek	& 32.609 & 
+    -116.7	& 267   \\
+	Sweetwater River	& 32.9 & 
+    -116.585	& 1180   \\
+	Grade Road Palomar & 33.314 &
+    -116.871	& 1577   \\
+	Whitewater Canyon & 33.994 & 
+    -116.665	& 705   \\
+	Mill Creek	& 34.077 & 
+    -116.873	& 2050   \\
+	West Fork Mojave River	& 34.284 & 
+    -117.378	& 1120   \\
+	North Fork Middle Tule River	& 36.201 & 
+    -118.651	& 1314   \\
+	Paradise Creek	& 36.518 & 
+    -118.759	& 926   \\
+	Redwood Creek	& 36.691 & 
+    -118.91	& 1727   \\
+	Wawona  & 37.541 & 
+    -119.649	& 1224   \\
+	Rainbow Creek	& 37.819 & 
+    -120.007	& 876   \\
+	Middle Yuba River	& 39.397 & 
+    -121.082	& 455   \\
+	Little Jamison Creek	& 39.743 & 
+    -120.704	& 1603   \\
+	Deep Creek	& 41.668 & 
+    -123.11	& 707   \\
+	Rock Creek	& 43.374 & 
+    -122.957	& 326   \\
+	\bottomrule
+	\end{tabular}
+	\label{table:Table_FocalPops}
+\end{table}
+
+\subsection*{Plant propagation}
+
+On 14 April, 2014, 3-5 seeds per family were sown directly on sand (Quikrete Play Sand, Georgia, USA) watered to field capacity in RLC4 Ray Leach cone-tainers placed in RL98 98-well trays (Stuewe \& Sons, Inc., Oregon, USA).
+We used pure sand because \textit{E.
+cardinalis} typically grows in sandy, riparian soils (A.
+Angert, pers.
+obs.).
+Two jumbo-sized cotton balls at the bottom of cone-tainers prevented sand from washing out.
+Cone-tainers sat in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) to continuously bottom-water plants during germination in greenhouses at the University British Columbia campus in Vancouver, Canada (49\degree 15' N, 123\degree 15' W).
+Misters thoroughly wetted the top of the sand every two hours during the day.
+Most seeds germinated between 1 and 2 weeks, but we allowed 3 weeks before transferring seedlings to growth chambers.
+We recorded germination daily between one to two weeks after sowing, and every 2-3 days thereafter.
+On 5 May (21 days after sowing), we transferred seedlings to one of two growth chambers (Model E-15 Conviron, Manitoba, Canada).
+We thinned seedlings to one plant per cone-tainer, leaving the center-most plant.
+702 of 768 (91.4\%) had plants that could be used in the experiment.
+We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments.
+The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments (Table~\ref{table:TableS_InitialSize}).
+
+\subsection*{Temperature and drought treatments}
+
+We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels.
+The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which we designate as Hot and Cool treatments.
+Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments.
+A detailed description of treatments is provided in the Supplemental Materials and Methods and summarized in Fig~\ref{fig:Fig_ExptlDes}.
+Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level.
+Within each chamber, there were two Well-watered blocks and two Drought blocks.
+The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$.
+The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
+Lower humidity would have made the drought more severe, but low soil moisture is stressful in and of itself.
+The total number of plants in each treatment was: $n_{\text{cool,dry}} = 169$; $n_{\text{cool,ww}} = 174$; $n_{\text{hot,dry}} = 176$; $n_{\text{hot,ww}} = 183$.
+Each population had 8--12 individuals per treatment level (mean = 11, median = 11).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Figure summarizing experimental design
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_ExptlDes.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Experimental Design]{Overview of experimental treatments and timing of key trait measurements.
+All plants germinated within 21 days of sowing.
+At that time, we began temperature treatments (left axis), simulating a typical June-August weather pattern at Hot (red) and Cool (blue) sites.
+The bold lines track the average daily temperatures.
+Within each day, there was a maximum daytime temperature (top of translucent polygons) and minimum nighttime temperature (bottom of translucent polygons).
+The drought treatment commenced later by ramping down the frequency of bottom-watering episodes (dashed black line; right axis), while watering frequency was maintained in the control treatment (solid black line).
+Grey boxes on the bottom of the plot outline the period of key measurements described in the Material and Methods.}
+	\label{fig:Fig_ExptlDes}
+\end{figure}
+
+\subsection*{Trait measurements}
+
+We measured five traits in response to temperature and watering treatments (Table~\ref{table:Table_traits}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Table of key traits
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}[ht]
+   \centering
+   \topcaption{Key traits measured in this study.}
+   \begin{tabular}{@{} ll @{}}
+      \toprule
+  Trait & Units \\
+      \midrule
+  Days to germination  & day \\
+  Leaf expansion rate  &  mm day$^{-1}$  \\
+  Stem elongation rate  &  cm day$^{-1}$  \\
+  Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
+  Mortality & probability of death  \\
+	    \bottomrule
+   \end{tabular}
+   \label{table:Table_traits}
+\end{table}
+
+\paragraph{Days to germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}.
+We treated Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function.
+Statistical significance of the Population effect was determined using analysis of deviance.
+Note that, unlike other traits discussed below, we did not include Block, Treatment, or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not yet been applied.
+
+\paragraph{Growth rate: leaf expansion and stem elongation}
+
+We measured growth rate during two phases: leaf expansion and stem elongation.
+Growth measurements were taken during the early vegetative stage.
+We censused leaf length twice per week shortly after the emergence of true leaves from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements.
+We ceased measuring leaf length once it appeared to asymptote and growth shifted to stem elongation.
+We also censused plant height on 7 occasions (twice per week) between 29 May and 20 June (45 to 67 days after sowing) until plants began to initiate floral buds.
+Thus all growth measurements occured during the vegetative, prereproductive phase.
+Both leaf expansion and stem elongation were modelled separately as second-order polynomials.
+We used empirical Bayes' estimates of growth for each individual plant from linear mixed-effects models fit with the R package \pkg{lme4} version 1.1-12 \citep{Bates_etal_2015}.
+
+\paragraph{Photosynthesis}
+During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate on a subset of 329 plants evenly spread between treatments and families within populations.
+The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6-cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska).
+We made all measurements at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$ of photosynthetically active radiation), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
+During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to the average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$) so that all plants within a temperature level could be measured under the same conditions.
+
+\paragraph{Mortality}
+We assayed mortality during twice-weekly growth measurements.
+We analyzed the probability of surviving until the end of the experiment as a function of population, treatment, and their interactions using a Generalized Linear Mixed Model (GLMM) assuming binomially distributed errors.
+We included Family and Block as random effects.
+We assessed significance of fixed effects using Type-II Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.
+
+\subsection*{Intrinsic variation and plasticity}
+
+For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment (Temperature, Water, and Temperature $\times$ Water), and Population $\times$ Treatment interactions (Population $\times$ Temperature, Population $\times$ Water, and Population $\times$ Temperature $\times$ Water).
+We interpreted significant Population effects to indicate intrinsic variation and Population $\times$ Treatment interactions to indicate variation in plasticity.
+As mentioned above, we used survival and GLMM models for germination rate and mortality, respectively.
+For all other traits, we used mixed model ANOVAs with Family and Block included as random factors.
+We fit models using restricted maximum likelihood in lmer, a function in the R package \pkg{lme4} \citep{Bates_etal_2015}.
+We determined significant fixed effect terms using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-32 \citep{Kuznetsova_etal_2016}.
+This package uses Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests.
+We also included days to germination as a covariate in growth analyses.
+To ensure that Population and Treatment effects were specific to a particular growth phase, we included germination day as a covariate in leaf expansion and stem elongation analyses.
+
+\subsection*{Principal components of germination, growth, and photosynthesis}
+For each single-trait model above, we extracted the Population coefficient (factoring out Treatment and other effects).
+The multivariate distribution of these coefficients was then summarized using principal components analysis.
+The first principal component of these traits (TraitPC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis delineating fast to slow growth.
+
+\subsection*{Identifying putative selective agents}
+
+Latitudinal clines are common, but it is often difficult to ascribe this variation to a particular selective agent.
+To reiterate, we tested three non-mutually exclusive hypotheses about how such latitudinal clines emerge: 1) one or two climatic variables explain latitudinal trait variation; 2) latitude is a proxy for multiple climatic factors that together shape trait variation; and 3) latitude integrates selection in a broader climatic neighborhood.
+We found that a population's position along TraitPC1 correlated strongly with the latitude of origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{E.
+cardinalis}.
+We reasoned that if we identified a single climatic factor that explained more trait variation than latitude, then this would suggest that factor is a key selective agent underlying the latitudinal cline (Hypothesis 1).
+On the other hand, if multiple climatic factors together are necessary to explain trait variation, then this would suggest that many climatic factors together have imposed selection for the latitudinal cline (Hypothesis 2).
+We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables.
+To test Hypothesis 3 about climatic neighborhoods driving selection, we directly competed local with neighborhood climate.
+We used the immediate collection location for local climate.
+For climate neighborhoods, we sampled climate at 1000 random points (at 90-m resolution) within a 62-km radius buffer around the collection and took the average.
+We chose this buffer radius based on population genetic structure, as inferred from $\approx$25,000 restriction-site associated SNPs among 49 populations from across the range \citep{Paul_etal_2016}.
+Spatial autocorrelation in allele frequencies persists for 62 km.
+However radii of 10 km$^2$ and 100 km$^2$ resulted in similar outcomes (data not shown).
+Since \textit{E.
+cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}.
+Climatic means and variances (see below) were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}.
+In addition to competing local and neighborhood climate, we compared the univariate correlation between local and neighborhood climate with TraitPC1 and Latitude using paired $t$-tests.
+We adjusted degrees of freedom to account for the fact that many climatic factors are highly correlated and not independent.
+Specifically, we calculated the effective number of independent climatic factors ($M_\text{eff}$) using the formula $M_\text{eff} = 1 + (M - 1) (1 - \text{Var}(\lambda) / M)$ \citep{Chevrud_2001}, where $M$ is the original number of climatic factors and $\lambda$ are the eigenvalues of the correlation matrix of all climatic factors.
+
+To help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{E.
+cardinalis} occurrence records (see detail below), not just the 16 focal populations.
+We refer to these climatic factors as `Climate-Latitude' variables.
+The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{E.
+cardinalis} range.
+Therefore, we did not consider Climate-TraitPC1 variables to be candidate selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
+However, we do interpret potential selective agents identified in Climate-Latiude analyses alone, because the goal was to explain the latitudinal clines in traits, not all aspects of climate that vary with latitude.
+
+We selected Climate-Latitude and Climate-TraitPC1 variables independently using Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}.
+Random Forest regression is useful for cases like ours when the number of potential predictors is similar to or greater than the number of observations (`high $p$, low $n$' problem).
+VSURF is a multistip algorithm that progressively retains or eliminates variables based on their importance over regression trees in the forest.
+Variable importance is defined as the average amount a climate variable reduces mean-squared error in the predicted response (TraitPC1 or Latitude), compared to a randomly permuted dataset, across all trees in the random forest (see \citeauthor{Genuer_etal_2015} [\citeyear{Genuer_etal_2015}] for further detail).
+Hence, VSURF automatically eliminates unimportant and redundant variables based on the data without having to arbitrarily choose among colinear climate variables before the analysis.
+We kept only variables selected for prediction, the most stringent criterion.
+A visual overview of how we selected climatic variables is depicted in Fig~\ref{fig:Fig_VarSelect}.

### 2017-08-30 ms/ms.Rnw
minor change
restructuring

--- 2017-05-30 ms/ms.Rnw
@@
   pathMS <- "~/Google Drive/CardAdapt/ms" # Chris' MacBook Air
-  # pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop
+  # pathMS <- "~/Documents/Temporary/ms" # Chris' Poisson laptop
   # pathMS <- "/Users/amylauren/Desktop/CardAdapt/ms" # Amy's desktop
@@
 Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
-In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
+In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance \citep{Flood_etal_2011}.
 
@@
 Within each chamber, there were two Well-watered blocks and two Drought blocks.
-The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$.
+The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$ and set to a 16:8 light:dark cycle to simulate summer growing conditions.
 The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).
@@
 
-  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/DroughtResponse.pdf", w = 5.5, h = 5, useDingbats = F)
+  #pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/DroughtResponse.pdf", w = 5.5, h = 5, useDingbats = F)
   par(mai = c(1, 1.5, 0.25, 0.25))
@@
   axis(2, lwd = 0, lwd.ticks = 1, las = 1)
-  dev.off()
+  #dev.off()
 
@@
   # Plot latitude versus PC1
-  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/Figure_PC1vLat.pdf", 5.5, 5, useDingbats = F)
+  #pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/Figure_PC1vLat.pdf", 5.5, 5, useDingbats = F)
   par(mai = c(1, 1.5, 0.25, 0.25))
@@
 
-  dev.off()
+ # dev.off()
 
@@
 
-<<Fig_VarSelect, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+<<Fig_VarSelect, eval=FALSE, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
     dev.off()
-
-##### TEMPORARRY FIGURE FOR TALK #####
-
-path.root <- "~/Google Drive/CardAdapt" 
-setwd(path.root)
-source("Data/Climate/Rcode/1_Setup_directories.R")
-
-load(paste(path.obj, "/AllOcc_avg1981-2010.RData", sep = "")) # all.avg30
-load(paste(path.obj, "/AllOcc_cva1981-2010.RData", sep = "")) # all.cva30
-load(paste(path.obj, "/AllOccSA_avg1981-2010_62km.RData", sep = "")) # allSA.avg
-load(paste(path.obj, "/AllOccSA_cva1981-2010_62km.RData", sep = "")) # allSA.cva
-
-# Change column names for SA variables
-sharedCols <- c("Latitude", "Longitude", "Elevation", "PopID", "ID", "PopID")
-colnames(allSA.avg)[!colnames(allSA.avg) %in% sharedCols] <-
-  paste0(colnames(allSA.avg)[!colnames(allSA.avg) %in% sharedCols], "_SA")
-
-colnames(allSA.cva)[!colnames(allSA.cva) %in% sharedCols] <-
-  paste0(colnames(allSA.cva)[!colnames(allSA.cva) %in% sharedCols], "_SA")
-
-# Combine into single data.frame
-allClim <- cbind(all.avg30, all.cva30[rownames(all.avg30), !(colnames(all.cva30) %in% sharedCols)],
-                 allSA.avg[rownames(all.avg30), !(colnames(allSA.avg) %in% sharedCols)],
-                 allSA.cva[rownames(all.avg30), !(colnames(allSA.cva) %in% sharedCols)])
-
-  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/ClimateSpace.pdf", w = 5.5, h = 5, useDingbats = F)
-  par(mai = c(1, 1.5, 0.25, 0.25))
-
-  with(allClim, plot(bio10_avg - 273.16, bio16_avg, xlim = c(15, 27), ylim = c(100, 1500),
-                     xlab = "", ylab = "", axes = F, frame.plot = T, col = primShade1, pch = 19))
-  with(focClim, points(bio10_avg - 273.16, bio16_avg, cex = 2, col = "black", bg = primShade0, 
-                       lwd = 2, pch = 21))
-  axis(1, at = seq(16, 26, 2), labels = c(expression(16*degree), expression(18*degree), 
-                                          expression(20*degree), expression(22*degree), 
-                                          expression(24*degree), expression(26*degree)),
-		lwd = 0, lwd.ticks = 1) 
-  axis(2, las = 1, lwd = 0, lwd.ticks = 1)
-  title(xlab = "Temperature of Warmest Quarter", cex.lab = 1.5, line = 2.5)
-  title(xlab = expression(paste("(", degree, "C, 1981-2010 average)")), cex.lab = 1, line = 4)
-  title(ylab = "Precipitation of Wettest Quarter", line = 5, cex.lab = 1.5)
-  title(ylab = "(mm, 1981-2010 average)", cex.lab = 1, line = 4)
-
-  legend(grconvertX(0.96, "nfc", "user"), grconvertY(0.95, "nfc", "user"), xjust = 1, 
-         legend = c("all recorded populations", "focal populations"),
-         pch = c(19, 21), col = c(primShade1, "black"), pt.bg = primShade0, pt.cex = c(1, 2))
-dev.off()
-
-##### TEMPORARY FOR REVIEW RESPONSE #####
-  # This predicts that the difference between local and neighbourhood climate estimates should be lower 
-# for northern populations than all others
-
-# Separate into different datasets
-c1 <- !grepl("_SA", climVars)
-c2 <- grepl("_SA", climVars)
-
-dat1 <- thinnedOccClimate[, climVars[c1]]
-dat2 <- thinnedOccClimate[, climVars[c2]]
-
-# Standardize
-for (i in ncol(dat1)) {
-  x <- c(dat1[, i], dat2[, i])
-  x <- (x - mean(x)) / sd(x)
-  dat1[, i] <- x[1:nrow(dat1)]
-  dat2[, i] <- x[(nrow(dat1) + 1):(2 * nrow(dat1))]
-}
-
-# Calculate difference
-d12 <- dat1
-r <- numeric(ncol(d12))
-for (i in 1:ncol(d12)) {
-  d12[, i] <- abs(dat2[, i] - dat1[, i])
-  r[i] <- cor(thinnedOccClimate$Latitude, d12[, i])
-}
-hist(r[1:24])
-hist(r[25:48])
-summary(r)
-
-plot(thinnedOccClimate$Latitude, apply(d12, 1, function(X) sum(abs(X))))
-plot(thinnedOccClimate$Latitude, d12$bio12_avg)
-plot(thinnedOccClimate$Latitude, d12$bio12_cva)
-plot(thinnedOccClimate$Latitude, d12$bio1_avg)
-plot(thinnedOccClimate$Latitude, d12$bio1_cva)
 
@@
 
-\bibliography{../../CommonResources/refs}
+\bibliography{refs}
 \bibliographystyle{evolution}

### 2017-09-02 ms/ms.Rnw
effect sizes calculations

--- 2017-08-29 ms/ms.Rnw
@@
   # Source custom functions
-  #source(paste(pathMS, "/functions.R", sep = ""))
-  source("functions.R")
+  source("functions.R") # for sweave
+  # source(paste(pathMS, "/functions.R", sep = "")) # for debugging
 
@@
 Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive.
-We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively.
-Both intrinsic and plastic trait variation can be explained by genetic differences and both are involved in adaptation.
-For example, intrinsic differences in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
-Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
-In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.
+We distinguish between these patterns of adaptive trait differences by referring to `genetic variation' in trait means and `genetic variation in plasticity', respectively.
+Genetic variation in plasticity is synonymous with genotype-by-environment interactions, or simply (G$\times$E).
+Genetic variation in trait means and plasticity are both involved in adaptation.
+For example, genetic variation in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
+Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genetic variation in plasticity cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
+In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance \citep{Flood_etal_2011}.
 
@@
 A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments.
-Either intrinsic and/or plastic variation should vary clinally along environmental gradients.
+Either genetic variation in trait means and/or plasticity should vary clinally along environmental gradients.
 Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown.
@@
 Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation.
-Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008, Stock_etal_2014}, and drought response \citep{Kooyers_etal_2015} that likely relate to climatic variation.
+Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008, Stock_etal_2014}, and drought resistance \citep{Kooyers_etal_2015} that likely relate to climatic variation.
 
@@
 
-In this study, we investigated two major questions: 1) whether intrinsic or plastic physiological trait variation corresponds with latitude; and 2) what climatic factor(s) could plausibly be responsible for latitudinal clines.
+In this study, we investigated two major questions: 1) whether genetic variation in physiological trait means or plasticity corresponds with latitude; and 2) what climatic factor(s) could plausibly be responsible for latitudinal clines.
 Within question 2, we tested three hypotheses outlined in the previous paragraph: latitudinal clines are explained by a single dominant climatic factor, multiple climatic factors, or the climatic neighborhood experienced by nearby population connected through gene flow.
@@
 Yet, there is a deficiency of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010, Wright_etal_2013}.
-We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
-We found a latitudinal cline of intrinsic variation in photosynthesis and growth, but little evidence for variation in plasticity.
+We measured genetic variation in trait means and plasticity in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude.
+We found a latitudinal cline of trait means in photosynthesis and growth, but little evidence for variation in plasticity.
 Interannual variation in precipitation and temperature are associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{E.
@@
 Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
-We used seeds from \Sexpr{nlevels(master$Family)} families, \Sexpr{min(famTab)}--\Sexpr{max(famTab)} (mean = \Sexpr{round(mean(famTab), 1)}, median = \Sexpr{median(famTab)}) families per population.
+To control for maternal effects, we grew a large number of field-derived seeds in the greenhouse and generated seed families for this experiment by haphazardly crossing individuals from the same population.
+We selected seed families to maximize the number of field-derived individuals represented.
+Thus, we used seeds from \Sexpr{nlevels(master$Family)} greenhouse-derived seed families, \Sexpr{min(famTab)}--\Sexpr{max(famTab)} (mean = \Sexpr{round(mean(famTab), 1)}, median = \Sexpr{median(famTab)}) families per population.
 
@@
 rect(6, -2, 17, 0, border = "white", col = "grey")
-with(exptlDes, text(11.5, -1, labels = "Germ", col = "white", font = 3, 
+with(exptlDes, text(11.5, -1, labels = "Germ", col = "black", font = 3, 
   cex = 0.9))
@@
 with(exptlDes, text(min((which(LERmeas)) + max(which(LERmeas))) / 2, -1, 
-	labels = "LER", col = "white", font = 3, cex = 0.9))
+	labels = "LER", col = "black", font = 3, cex = 0.9))
 
@@
 with(exptlDes, text(min((which(SERmeas)) + max(which(SERmeas))) / 2 - 5, 1, 
-	labels = "SER", col = "white", font = 3, cex = 0.9))
+	labels = "SER", col = "black", font = 3, cex = 0.9))
 
@@
 with(exptlDes, text(min((which(Harvest)) + max(which(Harvest))) / 2, -1, 
-	labels = "Harvest", col = "white", font = 3, cex = 0.9))
-
-with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 5, 
-	labels = "Photo", col = "grey", font = 3, cex = 0.9))
+	labels = "Harvest", col = "black", font = 3, cex = 0.9))
+
 with(exptlDes, rect(min(which(PhotoMeas)), 2, max(which(PhotoMeas)), 4, 
 	border = "white", col = "grey"))
+with(exptlDes, text(min((which(PhotoMeas)) + max(which(PhotoMeas))) / 2, 5, 
+	labels = "Photo", col = "black", font = 3, cex = 0.9))
 
@@
 
+  # Confidence intervals on effect size (mean squared error)
+  fn <- function(merMod) {
+    x <- anova(merMod)
+    ret <- x[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"]
+    ret <- c(ret, x["PopID", "Mean Sq"] - x["PopID:TempTrt", "Mean Sq"],
+             x["PopID", "Mean Sq"] - x["PopID:WaterTrt", "Mean Sq"])
+    ret
+  }
+
+  ciMerModLER <- confint(mm, method = "boot", FUN = fn, nsim = 1e3, seed = 50717035,
+                         parallel = "multicore", ncpus = 2L, level = 0.95)
+  ciMerModLER <- as.data.frame(ciMerModLER)
+  fit <- anova(mm)
+  ciMerModLER$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
+                            fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
+                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
+  rownames(ciMerModLER) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+                             "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
+  rm(fit)
+
 @
@@
 
+  # Confidence intervals on effect size (mean squared error)
+  ciMerModSER <- confint(mm, method = "boot", FUN = fn, nsim = 1e3, seed = 46887099,
+                         parallel = "multicore", ncpus = 2L)
+  ciMerModSER <- as.data.frame(ciMerModSER)
+  fit <- anova(mm)
+  ciMerModSER$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
+                            fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
+                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
+  rownames(ciMerModSER) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+                             "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
+  rm(fit)
+
 @
@@
 We made all measurements at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$ of photosynthetically active radiation), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity.
+All measurements were taken between 9:00 AM and 5:00 PM (3 hours after lights turned on and 5 hours before lights turned off).
 During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to the average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$) so that all plants within a temperature level could be measured under the same conditions.
+We were photosynthesis after dry down had progressed to assess differences in photosynthetic responses to drought.
 
@@
 
+  # Confidence intervals on effect size (mean squared error)
+  ciMerModPhoto <- confint(mm, method = "boot", FUN = fn, nsim = 1e3, seed = 12657909,
+                           parallel = "multicore", ncpus = 2L)
+  ciMerModPhoto <- as.data.frame(ciMerModPhoto)
+  fit <- anova(mm)
+  ciMerModPhoto$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
+                              fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
+                              fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
+  rownames(ciMerModPhoto) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+                               "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
+  rm(fit)
+
 @
@@
 
+  fn <- function(merMod) {
+    x <- Anova(merMod, type = 2)
+    ret <- x[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Chisq"]
+    ret <- c(ret, x["PopID", "Chisq"] - x["PopID:TempTrt", "Chisq"],
+             x["PopID", "Chisq"] - x["PopID:WaterTrt", "Chisq"])
+    ret
+  }
+
+  system.time(
+  ciMerModMort <- confint(fitMort, method = "boot", FUN = fn, nsim = 1e1, seed = 925200709,
+                          parallel = "multicore", ncpus = 2L)
+)
   # Figure of reaction norms
@@
 
-\subsection*{Intrinsic variation and plasticity}
+<<Effect size confidence intervals, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  pathFig <- "/Users/cdmuir/Google Drive/CardAdapt/ms/Figures"
+
+pdf(paste0(pathFig, "/FigureS_EffectSizes.pdf"), 12, 4, useDingbats = F)
+par(mai = c(1, 1, 0.5, 0.1), mfrow = c(1, 3))
+
+for (mod in list(ciMerModLER, ciMerModSER, ciMerModPhoto)) {
+
+#  assign("mod", mod)
+
+  plot(0, 0, type = "n", axes = FALSE, frame.plot = TRUE,
+       ylim = range(mod), xlim = c(1, 5),
+       xlab = "", ylab = "Effect size (Mean squared error)", cex.lab = 1.5,
+       main = )
+
+  abline(h = 0, lty = 2, col = "grey50")
+  for (i in 1:5) {
+    lines(rep(i, 2), c(mod[i, "2.5 %"], mod[i, "97.5 %"]))
+    points(i, mod[i, "estimate"], pch = 19)
+  }
+
+  axis(2, las = 1, lwd = 0, lwd.ticks = 1)
+  clip(-100, 100, -100, 100)
+  text(x = 1:5, y = grconvertY(-0.025, "npc", "user"), srt = 45, pos = 2, 
+       labels = c("Pop", "Pop x Temp", "Pop x Water", "Pop - Pop x Temp",
+                  "Pop - Pop x Water"))
+
+}
+
+dev.off()
+
+@
+
+\subsection*{Genetic variation in trait means and plasticity}
 
 For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment (Temperature, Water, and Temperature $\times$ Water), and Population $\times$ Treatment interactions (Population $\times$ Temperature, Population $\times$ Water, and Population $\times$ Temperature $\times$ Water).
-We interpreted significant Population effects to indicate intrinsic variation and Population $\times$ Treatment interactions to indicate variation in plasticity.
+We interpreted significant Population effects to indicate genetic variation in trait means and Population $\times$ Treatment interactions to indicate genetic variation in plasticity.
 As mentioned above, we used survival and GLMM models for germination rate and mortality, respectively.
@@
 We determined significant fixed effect terms using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-32 \citep{Kuznetsova_etal_2016}.
-This package uses Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests.
+We used Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests.
 We also included days to germination as a covariate in growth analyses.
 To ensure that Population and Treatment effects were specific to a particular growth phase, we included germination day as a covariate in leaf expansion and stem elongation analyses.
+
+Failure to detect a significant effect could be the result of Type-2 error, so we complemented the above analysis by comparing effect sizes in the full model.
+The full model contains all main effects, two-way interactions (Population $\times$ Temperature, Population $\times$ Water), a three-way interaction (Population $\times$ Temperature $\times$ Water), and random effects.
+We compared the effect sizes (mean squared error for LMM, ?
+for GLMM) of Population to Population $\times$ Treatment in 1000 parametric bootstrap samples simulated from fitted models.
+MARKER
 
@@
 Random Forest regression is useful for cases like ours when the number of potential predictors is similar to or greater than the number of observations (`high $p$, low $n$' problem).
-VSURF is a multistip algorithm that progressively retains or eliminates variables based on their importance over regression trees in the forest.
+VSURF is a multistep algorithm that progressively retains or eliminates variables based on their importance over regression trees in the forest.
 Variable importance is defined as the average amount a climate variable reduces mean-squared error in the predicted response (TraitPC1 or Latitude), compared to a randomly permuted dataset, across all trees in the random forest (see \citeauthor{Genuer_etal_2015} [\citeyear{Genuer_etal_2015}] for further detail).
@@
 We kept only variables selected for prediction, the most stringent criterion.
-A visual overview of how we selected climatic variables is depicted in Fig~\ref{fig:Fig_VarSelect}.
-
-<<Fig_VarSelect, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+We visually depict how we selected climatic variables in Fig~\ref{fig:Fig_VarSelect}.
+
+<<Fig_VarSelect, eval=FALSE, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
 
@@
 
-Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation.
-In contrast to the intrinsic differences described above, we found little evidence of G$\times$E in \textit{E.
+In contrast to the genetic variation in trait means described above, we found little evidence of G$\times$E in \textit{E.
 cardinalis}.
@@
 = \Sexpr{round(Meff2 / 2 - 1, 1)}, $P = \Sexpr{round(pt2 * 10 ^ -floor(log10(pt2)), 2)} \times 10 ^ {\Sexpr{floor(log10(pt2))}}$).
-
 Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates to explain the latitudinal cline in \textit{E.
@@
 Latitudinal clines in germination rate, photosynthesis, and growth suggest adaptive differentiation in important physiological traits of the species.
-However, we found little evidence that populations respond differently to temperature or drought.
+However, we found little evidence for variation in plasticity to temperature or drought.
 Due to low replication within families, we did not have power to assess within-population genotype-by-environment interactions, which may be present.
 As we discuss below, low variation in plasticity among populations may indicate that some dimensions of the fundamental abiotic niche are relatively conserved.
-Note that statistical power to detect significant plasticity is lower than that for intrinsic differences.
+Note that statistical power to detect significant plasticity is lower than that for differences in trait means.
 However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain low variation in plasticity.
 Finally, our results suggest that neighborhood-scale climate and interannual variation are more important selective agents than local averages.
-In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in physiology changes clinally, whereas plastic responses to temperature and drought are relatively static.
+In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why variation in physiological trait means changes clinally, whereas plastic responses to temperature and drought are relatively static.
 One caveat to bear in mind is that we are limited by the size of the climate grid ($\approx 90~\text{m}^2$) and therefore unable to detect very fine-scale local adaptation.
@@
 
-It is reasonable to predict that southern populations, which appear to experience more frequent drought years (see below), might have physiological adaptation to survive and grow in drier soil.
+It is reasonable to predict that southern populations, which appear to experience more frequent drought years (see below), might have physiological adaptations to respond to drought stress to survive and grow in drier soil.
 We found no evidence for this type of drought tolerance; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
@@
 Plants in the stream never have to tolerate drought whereas plants only a few meters away may experience extreme drought since there is little direct precipitation during the growing season in Mediterranean climates of western North America.
-We hypothesize alleles that confer greater drought tolerance may be quite costly in well-watered soils, and \textit{vice versa}, leading to strong fitness tradeoffs.
+We hypothesize that alleles confering greater drought tolerance may be quite costly in well-watered soils, and \textit{vice versa}, leading to strong fitness tradeoffs.
 Such tradeoffs would promote specialization to one soil moisture or another, thereby inhibiting the evolution of broad environmental tolerance within a population.
@@
 Infrequent wet years may also produce most seeds, so selection is weighted towards alleles that have high fitness in the wet environment, even if dry years are more frequent \citep{Templeton_Levin_1979, Brown_Venable_1986}.
+However, demographic asymmetry should equally hinder the evolution of both drought tolerance and escape, so it should not explain why one mechanism evolves but not the other.
 Finally, gene flow, which is generally high among \textit{E.
@@
 
-In sum, these results indicate that intrinsic differences in physiology and growth, but not plastic responses to temperature and drought, mediate local adaptation to climate in \textit{E.
+In sum, these results indicate that genetic differences in physiology and growth are better candidates than plastic responses to temperature and drought as mediators of local adaptation to climate in \textit{E.
 cardinalis}.

### 2017-09-02 README.Rmd
updated README

@@
----
-title: "Readme for Muir and Angert"
-author: "Chris Muir and Amy Angert"
-output: html_document
----
+# card-cline
+Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in *Erythranthe cardinalis*
 
-This repository contains the source for the paper Muir CD and Angert AL. Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in *Mimulus cardinalis*.
+This project was developed by [Chris Muir](www.chrisdmuir.com) and [Amy Angert](http://angert.botany.ubc.ca/) .
 
-See the Sweave file 'ms/ms.Rnw' for code chunks. If you have R packages installed, you should be able to recreate the manuscript locally.
+More information about the study is available in a preprint which you can find on [biorxiv](https://doi.org/10.1101/080952) or on [github](https://github.com/cdmuir/card-cline/blob/master/ms/ms.pdf).
 
-A preprint of this paper is available on biorxiv: http://dx.doi.org/10.1101/080952
+## Downloading data and code
+
+1. Download or clone this repository to your machine.
+2. Open `card-cline.Rproj` in [RStudio](https://www.rstudio.com/)
+
+## Generating manuscript
+
+Simply open `ms/ms.Rnw` and compile using RStudio.
+
+[![CC BY](http://i.creativecommons.org/l/by/3.0/88x31.png)](http://creativecommons.org/licenses/by/3.0/)

### 2017-09-02 ms/ms.Rnw
added link to github repo
adding mortality effect size analysis

@@
 \section*{Material and Methods}
+
+Data and annotated source code to reproduce these analyses and manuscript are available on GitHub (https://github.com/cdmuir/card-cline).
 
@@
 
-  # Confidence intervals on effect size (mean squared error)
+  # Function for bootstrapping effect sizes
   fn <- function(merMod) {
@@
   par(mai = c(0.5, 1, 0.5, 0.25))
-  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, xlim = c(1, 4), 
-       ylim = c(5, 16))
+  plot(0, 0, type = "n", xlab = "", ylab = "", axes = F, frame.plot = T, 
+       xlim = c(1, 4), ylim = c(5, 16))
   abline(v = 2.5)
@@
   # Confidence intervals on effect size (mean squared error)
-  ciMerModPhoto <- confint(mm, method = "boot", FUN = fn, nsim = 1e3, seed = 12657909,
-                           parallel = "multicore", ncpus = 2L)
+  ciMerModPhoto <- confint(mm, method = "boot", FUN = fn, nsim = 1e3, 
+                           seed = 12657909, parallel = "multicore", ncpus = 2L)
   ciMerModPhoto <- as.data.frame(ciMerModPhoto)
@@
   								 family = "binomial")
-  # Anova(fitMort, type = 2)
-
+  fit <- Anova(fitMort, type = 2)
+
+  # Function for bootstrapping effect sizes
   fn <- function(merMod) {
     x <- Anova(merMod, type = 2)
-    ret <- x[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Chisq"]
+    ret <- x[c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+               "PopID:TempTrt:WaterTrt"), "Chisq"]
     ret <- c(ret, x["PopID", "Chisq"] - x["PopID:TempTrt", "Chisq"],
-             x["PopID", "Chisq"] - x["PopID:WaterTrt", "Chisq"])
+             x["PopID", "Chisq"] - x["PopID:WaterTrt", "Chisq"],
+             x["PopID", "Chisq"] - x["PopID:TempTrt:WaterTrt", "Chisq"])
     ret
@@
   system.time(
-  ciMerModMort <- confint(fitMort, method = "boot", FUN = fn, nsim = 1e1, seed = 925200709,
-                          parallel = "multicore", ncpus = 2L)
+  ciMerModMort <- confint(fitMort, method = "boot", FUN = fn, nsim = 1e3, seed = 925200709,
+                          parallel = "multicore", ncpus = 4L)
 )
+
+  # tmp  <- ciMerModMort
+  # ciMerModMort <- tmp
+  ciMerModMort <- as.data.frame(ciMerModMort)
+  ciMerModMort$estimate <- c(
+    fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt", "PopID:TempTrt:WaterTrt"),
+        "Chisq"],
+    fit["PopID", "Chisq"] - fit["PopID:TempTrt", "Chisq"],
+    fit["PopID", "Chisq"] - fit["PopID:WaterTrt", "Chisq"],
+    fit["PopID", "Chisq"] - fit["PopID:TempTrt:WaterTrt", "Chisq"])
+
+      rownames(ciMerModPhoto) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+                               "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
+  rm(fit)
+
+  saveRDS(ciMerModMort, paste0(pathDat, "/ciMerModMort.rds"))
+  ciMerModMort <- readRDS(paste0(pathDat, "/ciMerModMort.rds"))
   # Figure of reaction norms

### 2017-09-02 ms/ms.Rnw
revising ms

@@
 We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables.
+
 To test Hypothesis 3 about climatic neighborhoods driving selection, we directly competed local with neighborhood climate.
+The logic is that if the climatic analysis can identify candidate climatic factors important for local adaptation, then stronger correlations with neighbourhood climate would suggest a role for gene flow.
 We used the immediate collection location for local climate.
@@
 cardinalis} range.
+If a climatic factor is driving the latitudinal cline in TraitPC1, then we expect that climatic factor will correlate strongly with latitude of occurrence localities.
 Therefore, we did not consider Climate-TraitPC1 variables to be candidate selective agents unless the same or very similar variable was found in the Climate-Latitude analysis.
-However, we do interpret potential selective agents identified in Climate-Latiude analyses alone, because the goal was to explain the latitudinal clines in traits, not all aspects of climate that vary with latitude.
+However, we do not interpret potential selective agents identified in Climate-Latitude analyses alone, because the goal was to explain the latitudinal clines in traits, not all aspects of climate that vary with latitude.
 
@@
 Latitudinal clines in germination rate, photosynthesis, and growth suggest adaptive differentiation in important physiological traits of the species.
-However, we found little evidence for variation in plasticity to temperature or drought.
+However, we caution that we cannot yet rule nonadaptive demographic processes such as a recent range expansion toward higher latitude \citep{Paul_etal_2016, Sheth_Angert_2017}.
+In contrast, we found little evidence for variation in plasticity to temperature or drought.
 Due to low replication within families, we did not have power to assess within-population genotype-by-environment interactions, which may be present.
@@
 Greater investment in aboveground growth, as opposed to belowground storage for future seasons, may be favoured in climates with more frequent drought years, but maladaptive in climates with more consistent precipitation.
-This is a form of drought escape in that plants are investing more reproduction in the present to avoid possible drought in subsequent years.
-Suppose plants that grow quickly and allocate new resources to continued growth rather than storage have higher fitness over a single growing season.
-However, by not allocating resources to storage, these fast-growing plants begin future seasons at a deficit.
-Therefore, in a stable environment where winter survivorship is assured in most years, failure to store resources may reduce lifetime fitness.
+In a stable environment where winter survivorship is assured in most years, failure to store resources may reduce lifetime fitness.
 But for perennial herbs in Mediterranean climates, a dry winter (rainy season) can kill the rhizomes (underground stems that store nutrients for future growth) before emergence or aboveground stems before flowering.
 If drought years occur frequently enough, selection may favour the fast-growing strategy because there is no advantage to storage if drought kills plants before flowering.
-Considering life-history strategy as a continuum from no storage (annual) to lots of storage (perennial), we hypothesize that the optimal allocation to aboveground growth is more `annualized' in southern climates that have greater interannual variation in precipitation.
-This scenario differs from classic drought escape syndromes in which plants speed up development early in the season before the onset of drought.
+Considering life-history strategy as a continuum from no storage (annual) to lots of storage (perennial), we hypothesize that the optimal allocation to aboveground growth is more `annualized' in southern climates that have greater interannual variation in precipitation.This is a form of drought escape in that plants are investing more reproduction in the present to avoid possible drought in subsequent years, but is distinct from classic drought escape syndromes in which plants speed up development early in the season before the onset of drought.
 
@@
 Life history theory shows that less variable environments are one factor that favours the evolution of perenniality \citep{Stearns_1976, Iwasa_Cohen_1989, Friedman_Rubin_2015}.
-Populations of the perennial \textit{Plantago asiatica} show a similar latitudinal cline in growth and allocation to storage \citep{Sawada_etal_1994}, though these authors attribute the cline to variation in growing season length.
+Populations of the perennial \textit{Plantago asiatica} show a similar latitudinal cline in growth and allocation to storage \citep{Sawada_etal_1994} but attribute the cline to variation in growing season length.
 There are also life history clines in the closely related species \textit{E.
@@
 Rather, our hypothesis states that climatic variability selects on quantitative variation in allocation to growth versus storage.
-This hypothesis makes several independent, testable predictions.
+
+% This hypothesis makes several independent, testable predictions.
 The allocation tradeoff predicts that northern populations will provision more photosynthetic assimilate to rhizomes compared with southern populations.

### 2017-09-02 ms/ms.Rnw
mortality effect size analysis

@@
 
-  system.time(
-  ciMerModMort <- confint(fitMort, method = "boot", FUN = fn, nsim = 1e3, seed = 925200709,
-                          parallel = "multicore", ncpus = 4L)
-)
-
-  # tmp  <- ciMerModMort
-  # ciMerModMort <- tmp
-  ciMerModMort <- as.data.frame(ciMerModMort)
-  ciMerModMort$estimate <- c(
-    fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt", "PopID:TempTrt:WaterTrt"),
-        "Chisq"],
-    fit["PopID", "Chisq"] - fit["PopID:TempTrt", "Chisq"],
-    fit["PopID", "Chisq"] - fit["PopID:WaterTrt", "Chisq"],
-    fit["PopID", "Chisq"] - fit["PopID:TempTrt:WaterTrt", "Chisq"])
-
-      rownames(ciMerModPhoto) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
-                               "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
-  rm(fit)
-
-  saveRDS(ciMerModMort, paste0(pathDat, "/ciMerModMort.rds"))
+  # Takes ~3 hours to run; use saved output
+  # ciMerModMort <- confint(fitMort, method = "boot", FUN = fn, nsim = 1e3, 
+  #                         seed = 925200709, parallel = "multicore", ncpus = 4L)
+  # 
+  # ciMerModMort <- as.data.frame(ciMerModMort)
+  # ciMerModMort$estimate <- c(
+  #   fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt", "PopID:TempTrt:WaterTrt"),
+  #       "Chisq"],
+  #   fit["PopID", "Chisq"] - fit["PopID:TempTrt", "Chisq"],
+  #   fit["PopID", "Chisq"] - fit["PopID:WaterTrt", "Chisq"],
+  #   fit["PopID", "Chisq"] - fit["PopID:TempTrt:WaterTrt", "Chisq"])
+  # 
+  # rownames(ciMerModMort) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+  #                             "PopID:TempTrt:WaterTrt", "PopID - PopID:TempTrt",
+  #                             "PopID - PopID:WaterTrt", 
+  #                             "PopID - PopID:TempTrt:WaterTrt")
+  # rm(fit)
+  # 
+  # saveRDS(ciMerModMort, paste0(pathDat, "/ciMerModMort.rds"))
   ciMerModMort <- readRDS(paste0(pathDat, "/ciMerModMort.rds"))
+
   # Figure of reaction norms

### 2017-09-02 ms/ms.Rnw
revising ms

@@
     x <- anova(merMod)
-    ret <- x[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"]
+    ret <- x[c("PopID", "PopID:TempTrt", "PopID:WaterTrt", "PopID:TempTrt:WaterTrt"), "Mean Sq"]
     ret <- c(ret, x["PopID", "Mean Sq"] - x["PopID:TempTrt", "Mean Sq"],
-             x["PopID", "Mean Sq"] - x["PopID:WaterTrt", "Mean Sq"])
+             x["PopID", "Mean Sq"] - x["PopID:WaterTrt", "Mean Sq"],
+             x["PopID", "Mean Sq"] - x["PopID:TempTrt:WaterTrt", "Mean Sq"])
     ret
@@
   fit <- anova(mm)
-  ciMerModLER$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
+  ciMerModLER$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+                                  "PopID:TempTrt:WaterTrt"), "Mean Sq"],
                             fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
-                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
-  rownames(ciMerModLER) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
-                             "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
+                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"],
+                            fit["PopID", "Mean Sq"] - fit["PopID:TempTrt:WaterTrt", "Mean Sq"])
+  rownames(ciMerModLER) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", "PopID:TempTrt:WaterTrt", 
+                             "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt",
+                             "PopID - PopID:TempTrt:WaterTrt")
   rm(fit)
@@
   fit <- anova(mm)
-  ciMerModSER$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
+  ciMerModSER$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+                                  "PopID:TempTrt:WaterTrt"), "Mean Sq"],
                             fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
-                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
-  rownames(ciMerModSER) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
-                             "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
+                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"],
+                            fit["PopID", "Mean Sq"] - fit["PopID:TempTrt:WaterTrt", "Mean Sq"])
+  rownames(ciMerModSER) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", "PopID:TempTrt:WaterTrt", 
+                             "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt",
+                             "PopID - PopID:TempTrt:WaterTrt")
   rm(fit)
@@
   fit <- anova(mm)
-  ciMerModPhoto$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt"), "Mean Sq"],
+  ciMerModPhoto$estimate <- c(fit[c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
+                                  "PopID:TempTrt:WaterTrt"), "Mean Sq"],
                             fit["PopID", "Mean Sq"] - fit["PopID:TempTrt", "Mean Sq"],
-                              fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"])
+                            fit["PopID", "Mean Sq"] - fit["PopID:WaterTrt", "Mean Sq"],
+                            fit["PopID", "Mean Sq"] - fit["PopID:TempTrt:WaterTrt", "Mean Sq"])
   rownames(ciMerModPhoto) <- c("PopID", "PopID:TempTrt", "PopID:WaterTrt", 
-                               "PopID - PopID:TempTrt", "PopID - PopID:WaterTrt")
+                               "PopID:TempTrt:WaterTrt", "PopID - PopID:TempTrt", 
+                               "PopID - PopID:WaterTrt",
+                               "PopID - PopID:TempTrt:WaterTrt")
   rm(fit)
@@
 
-  ##### TEMPORARY: CODE TO EXPORT FIGURE FOR TALK #####
-  # d <- subset(master, !is.na(master$DiedFromStress))
-  # x <- tapply(d$DiedFromStress, list(d$PopID, d$WaterTrt, d$TempTrt), function(X) 1 - length(which(X)) / length(X))
-  # 
-  # pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/DroughtResponse.pdf", w = 5.5, h = 5, useDingbats = F)
-  # par(mai = c(1, 1.5, 0.25, 0.25))
-  # 
-  # plot(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], axes = F, frame.plot = T,
-  #      pch = 21, col = "black", bg = primShade0, cex = 2, type = "n",
-  #      xlab = "Precipitation of Wettest Quarter (mm)", 
-  #      ylab = "Relative Survivorship\n(Hot/Dry : Hot/Wet)", cex.lab = 1.5)
-  # points(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], pch = 21, col = "black", 
-  #        bg = primShade0, cex = 1.5, lwd = 2)
-  # axis(1, lwd = 0, lwd.ticks = 1)
-  # axis(2, lwd = 0, lwd.ticks = 1, las = 1)
-  # dev.off()
-  #
-
 @
@@
 
-  pathFig <- "/Users/cdmuir/Google Drive/CardAdapt/ms/Figures"
-
-pdf(paste0(pathFig, "/FigureS_EffectSizes.pdf"), 12, 4, useDingbats = F)
-par(mai = c(1, 1, 0.5, 0.1), mfrow = c(1, 3))
-
-for (mod in list(ciMerModLER, ciMerModSER, ciMerModPhoto)) {
-
-#  assign("mod", mod)
-
+  # pathFig <- "/Users/cdmuir/Google Drive/CardAdapt/ms/Figures" # for debug
+
+  pdf(paste0(pathFig, "/FigureS_EffectSizes.pdf"), 8, 8, useDingbats = F)
+  par(mai = c(1, 1, 0.5, 0.1), mfrow = c(2, 2))
+
+  mods <- list(ciMerModLER, ciMerModSER, ciMerModPhoto, ciMerModMort)
+  for (i in 1:length(mods)) {
+
+    mod <- mods[[i]]
+    main <- c("Leaf expansion", "Stem elongation", "Photosynthesis", "Mortality")[i]
     plot(0, 0, type = "n", axes = FALSE, frame.plot = TRUE,
-       ylim = range(mod), xlim = c(1, 5),
-       xlab = "", ylab = "Effect size (Mean squared error)", cex.lab = 1.5,
-       main = )
+         ylim = range(mod), xlim = c(1, 7),
+         xlab = "", ylab = "Effect size", cex.lab = 1.5,
+         main = main)
 
     abline(h = 0, lty = 2, col = "grey50")
-  for (i in 1:5) {
+    for (i in 1:7) {
     lines(rep(i, 2), c(mod[i, "2.5 %"], mod[i, "97.5 %"]))
@@
     clip(-100, 100, -100, 100)
-  text(x = 1:5, y = grconvertY(-0.025, "npc", "user"), srt = 45, pos = 2, 
-       labels = c("Pop", "Pop x Temp", "Pop x Water", "Pop - Pop x Temp",
-                  "Pop - Pop x Water"))
+    text(x = 1:7, y = grconvertY(-0.025, "npc", "user"), srt = 45, pos = 2, 
+         labels = c("Pop", "Pop x Temp", "Pop x Water", "Pop x Temp x Water", 
+                    "Pop - Pop x Temp", "Pop - Pop x Water", "Pop - Pop x Temp x Water"))
 
@@
   dev.off()
-
-  ##### TEMPORARY VERSION FOR TALK #####
-  # Plot latitude versus PC1
-#   pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/Figure_PC1vLat.pdf", 5.5, 5, useDingbats = F)
-#   par(mai = c(1, 1.5, 0.25, 0.25))
-#   plot(focClim$Lat, focClim$PC1, xlim = c(32, 44), ylim = c(-4, 4), axes = F,
-#     frame.plot = T, type = "n", xlab = "", ylab = "")
-#   title(xlab = "Latitude", cex.lab = 1.5)
-#   title(ylab = "TraitPC1", cex.lab = 2, line = 5)
-#   title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
-#     list(x = pve)))
-#   axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), 
-# 		expression(36*degree), expression(40*degree), expression(44*degree)), 
-# 		lwd = 0, lwd.ticks = 1) 
-#   axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
-#   fitLat <- lm(PC1 ~ Latitude, data = focClim)
-#   
-# 	# Label y-axis from slow to fast growth
-# 	mtext(c("Slow\nGrowth", "Fast\nGrowth"), 2, at = c(-4, 4), cex = 1.25, 
-# 				font = 3, las = 1, line = 4, adj = 0.5)
-# 
-#   # polygon of confidence intervals
-#   x <- seq(min(focClim$Latitude), max(focClim$Latitude), l = 1e3)
-#   u <- predict(fitLat, new = data.frame(Latitude = x))
-#   s <- sqrt(sapply(x, function(X) t(c(1, X)) %*% vcov(fitLat) %*% c(1, X)))
-#   confInt <- data.frame(lowerCI = qnorm(0.025, u, s), 
-#   											upperCI = qnorm(0.975, u, s))
-#   polygon(c(x, rev(x)), c(confInt$lowerCI, rev(confInt$upperCI)), 
-#     col = "grey", border = NA)
-#   lines(x, confInt$lowerCI, lwd = 2, lty = 2)
-#   lines(x, confInt$upperCI, lwd = 2, lty = 2)
-#   lines(range(focClim$Lat), rev(range(predict(fitLat))), lwd = 2)
-#   points(focClim$Lat, focClim$PC1, pch = 21, col = "black", bg = primShade0, 
-#   			 cex = 1.5, lwd = 2)
-#   
-#   # P-value
-#   text(grconvertX(0.95, "npc"), grconvertY(0.95, "npc"), adj = c(1, 1),
-# 	     labels = bquote(italic(P) == .(round(summary(fitLat)[[4]][2, 4], 3))))
-# 
-#   dev.off()
 

### 2017-09-03 ms/ms.Rnw
added effect size figure to ms

@@
 
-<<Effect size confidence intervals, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
-
-  # pathFig <- "/Users/cdmuir/Google Drive/CardAdapt/ms/Figures" # for debug
-
-  pdf(paste0(pathFig, "/FigureS_EffectSizes.pdf"), 8, 8, useDingbats = F)
-  par(mai = c(1, 1, 0.5, 0.1), mfrow = c(2, 2))
-
-  mods <- list(ciMerModLER, ciMerModSER, ciMerModPhoto, ciMerModMort)
-  for (i in 1:length(mods)) {
-
-    mod <- mods[[i]]
-    main <- c("Leaf expansion", "Stem elongation", "Photosynthesis", "Mortality")[i]
-    plot(0, 0, type = "n", axes = FALSE, frame.plot = TRUE,
-         ylim = range(mod), xlim = c(1, 7),
-         xlab = "", ylab = "Effect size", cex.lab = 1.5,
-         main = main)
-
-    abline(h = 0, lty = 2, col = "grey50")
-    for (i in 1:7) {
-    lines(rep(i, 2), c(mod[i, "2.5 %"], mod[i, "97.5 %"]))
-      points(i, mod[i, "estimate"], pch = 19)
-    }
-
-    axis(2, las = 1, lwd = 0, lwd.ticks = 1)
-    clip(-100, 100, -100, 100)
-    text(x = 1:7, y = grconvertY(-0.025, "npc", "user"), srt = 45, pos = 2, 
-         labels = c("Pop", "Pop x Temp", "Pop x Water", "Pop x Temp x Water", 
-                    "Pop - Pop x Temp", "Pop - Pop x Water", "Pop - Pop x Temp x Water"))
-
-}
-
-dev.off()
-
-@
-
 \subsection*{Genetic variation in trait means and plasticity}
@@
 
-Failure to detect a significant effect could be the result of Type-2 error, so we complemented the above analysis by comparing effect sizes in the full model.
+Failure to detect a significant effect could be the result of Type-2 error, so we complemented step-wise ANOVA (see above) by comparing effect sizes calculated in the full model.
 The full model contains all main effects, two-way interactions (Population $\times$ Temperature, Population $\times$ Water), a three-way interaction (Population $\times$ Temperature $\times$ Water), and random effects.
-We compared the effect sizes (mean squared error for LMM, ?
-for GLMM) of Population to Population $\times$ Treatment in 1000 parametric bootstrap samples simulated from fitted models.
-MARKER
+For linear mixed-effects models (leaf expansion, stem elongation, and photosynthesis) we used mean-squared error as a measure of effect size; for GLMM (mortality) we used $\chi^2$ as a measure of effect size.
+We did not include germination rate because no Population $\times$ Treatment effects were estimated.
+The difference in effect size of Population versus Population $\times$ Treatment is:
+
+$$ \Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})} = \textrm{Effect Size}_\textrm{Pop} - \textrm{Effect Size}_{\textrm{Pop} \times \textrm{Trt}} $$
+
+We calculated $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ for all two- and three-way Population $\times$ Treatment interactions for each trait.
+To determine whether $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ was significantly different than 0, we calculated 95\% confidence intervals using 1000 parametric bootstrap samples simulated from fitted models.
+If the 95\% confidence interval for a given $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$, was greater than zero then we concluded that the Population effect size was significantly larger than the Population $\times$ Treatment effect size, and vice versa if the confidence internval was less than zero.
+If the confidence spanned zero, then the effect sizes are not significantly different.
+
+<<Effect sizes, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
+
+  pdf(paste0(pathFig, "/FigureS_EffectSizes.pdf"), 8, 8, useDingbats = F)
+  par(mfrow = c(2, 2), oma = c(0, 2, 0, 0))
+
+  mods <- list(ciMerModLER, ciMerModSER, ciMerModPhoto, ciMerModMort)
+  for (i in 1:length(mods)) {
+
+    if (i %in% c(1, 3)) {
+      par(mar = c(5, 10, 5, 0))
+    } else {
+      par(mar = c(5, 0, 5, 10))
+    }
+
+    mod <- mods[[i]]
+    main <- paste0(LETTERS[i], ". ", 
+                   c("Leaf expansion", "Stem elongation", "Photosynthesis", "Mortality")[i])
+    plot(0, 0, type = "n", axes = FALSE, frame.plot = TRUE,
+         xlim = range(mod), ylim = c(7.5, 0.5),
+         ylab = "", xlab = "Effect size", cex.lab = 1.5, main = main)
+
+    abline(v = 0, lty = 2, col = "grey50")
+    for (j in 1:7) {
+      lines(y = rep(j, 2), c(mod[j, "2.5 %"], mod[j, "97.5 %"]))
+      x = points(mod[j, "estimate"], j, pch = 19)
+    }
+
+    axis(1, las = 1, lwd = 0, lwd.ticks = 1)
+    clip(-100, 100, -100, 100)
+   if (i %in% c(1, 3)) {
+      text(y = 1:7, x = grconvertX(-0.025, "npc", "user"), srt = 0, pos = 2, 
+           labels = c("Pop", "Pop x Temp", "Pop x Water", "Pop x Temp x Water", 
+                      "Pop - (Pop x Temp)", "Pop - (Pop x Water)", 
+                      "Pop - (Pop x Temp x Water)"))
+    }
+  }
+
+  dev.off()
+
+  # Numbers for Results
+  comparisons <- c(
+    "PopID - PopID:TempTrt",
+    "PopID - PopID:WaterTrt",
+    "PopID - PopID:TempTrt:WaterTrt"
+  )
+
+  fn <- function(ciMerMod, comparisons) {
+    ret <- numeric(3)
+
+    # Pop has larger effect than Pop x Trt
+    ret[1] <- length(which(ciMerMod[comparisons, "2.5 %"] > 0))
+
+    # Pop has smaller effect than Pop x Trt
+    ret[2] <- length(which(ciMerMod[comparisons, "97.5 %"] < 0))
+
+    # Pop and Pop x Trt have similar effect
+    ret[3] <- length(which(ciMerMod[comparisons, "2.5 %"] < 0 & 
+                             ciMerMod[comparisons, "97.5 %"] > 0))
+
+    stopifnot(sum(ret) == length(comparisons))
+
+    ret
+
+  }
+
+  effSize <- sapply(list(ciMerModLER, ciMerModSER, ciMerModPhoto, ciMerModMort),
+                    fn, comparisons = comparisons)
+  colnames(effSize) <- c("LER", "SER", "Photo", "Mort")
+  rownames(effSize) <- c("larger", "smaller", "same")
+
+@
 
@@
 Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
-Complete ANOVA tables are available in the Supporting Information (\Cref{table:TableS_fitLER,table:TableS_fitSER,table:TableS_fitPhoto,table:TableS_fitMort})
+Complete ANOVA tables are available in the Supporting Information (\Cref{table:TableS_fitLER,table:TableS_fitSER,table:TableS_fitPhoto,table:TableS_fitMort}).
+
+The effect size of Population was significantly larger than that for Population $\times$ Treatment interactions (Fig.~\ref{fig:FigS_EffectSizes}) in most cases.
+For leaf expansion, Population had a signicantly larger effect size than Population $\times$ Treatment interactions in \Sexpr{effSize["larger", "LER"]} of \Sexpr{sum(effSize[, "LER"])} comparisons (Fig.~\ref{fig:FigS_EffectSizes}A).
+For stem elongation (Fig.~\ref{fig:FigS_EffectSizes}B) and mortality (Fig.~\ref{fig:FigS_EffectSizes}D), Population effect sizes were significantly larger than all Population $\times$ Treatment interactions.
+For Photosynthesis, Population and Population $\times$ Treatment effect sizes were not significantly different (Fig.~\ref{fig:FigS_EffectSizes}C), presumably because we had a smaller sample size.
 
@@
 
+% Effect sizes of LER, SER, Photo, and Mort
+
+\begin{figure}
+	\centerline{\includegraphics{Figures/FigureS_EffectSizes.pdf}}
+	\usefont{T1}{phv}{m}{n}
+	\fontsize{10}{12}
+	\selectfont
+	\caption[Effect sizes]{Population effect sizes are usually larger than Population $\times$ Treatment effect sizes.
+In each panel, we plot estimated effect size (points) and 95\% confidence intervals (lines) inferred using parametric bootstrap (see Materials and Methods).
+At the top, we plot the effect sizes of Population (`Pop'), two-way interactions between Population and Temperature (`Pop $\times$ Temp') or Water (`Pop $\times$ Water'), and the three-way interaction between Population, Temperature, and Water (`Pop $\times$ Temp $\times$ Water').
+Below that we plot the difference in the Population minus the Population $\times$ Treatment effect size (e.g.
+`Pop - (Pop $\times$ Temp)').
+When confidence intervals do no overlap zero (dashed line), this means that Population has a significantly greater effect size than the interaction.
+Effect sizes were measured using unstandarized mean square error for linear mixed-effects models (leaf expansion, stem elongation, and photosynthesis) and $\chi^2$ for GLMM (mortality).
+Hence, the effect size values are not comparable between different traits.}
+	\label{fig:FigS_EffectSizes}
+\end{figure}
+
 % Figure of bio11_cva_SA versus TraitPC1

### 2017-09-03 ms/ms.tex
effect sizes calculations
added effect size figure to ms

@@
 Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive.
-We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively.
-Both intrinsic and plastic trait variation can be explained by genetic differences and both are involved in adaptation.
-For example, intrinsic differences in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
-Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
+We distinguish between these patterns of adaptive trait differences by referring to `genetic variation' in trait means and `genetic variation in plasticity', respectively.
+Genetic variation in plasticity is synonymous with genotype-by-environment interactions, or simply (G$\times$E).
+Genetic variation in trait means and plasticity are both involved in adaptation.
+For example, genetic variation in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment.
+Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genetic variation in plasticity cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.
 In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance \citep{Flood_etal_2011}.
@@
 Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
-We used seeds from 154 families, 4--12 (mean = 9.6, median = 12) families per population.
+To control for maternal effects, we grew a large number of field-derived seeds in the greenhouse and generated seed families for this experiment by haphazardly crossing individuals from the same population.
+We selected seed families to maximize the number of field-derived individuals represented.
+Thus, we used seeds from 154 greenhouse-derived seed families, 4--12 (mean = 9.6, median = 12) families per population.
 
@@
 
-\subsection*{Intrinsic variation and plasticity}
+\subsection*{Genetic variation in trait means and plasticity}
 
 For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment (Temperature, Water, and Temperature $\times$ Water), and Population $\times$ Treatment interactions (Population $\times$ Temperature, Population $\times$ Water, and Population $\times$ Temperature $\times$ Water).
-We interpreted significant Population effects to indicate intrinsic variation and Population $\times$ Treatment interactions to indicate variation in plasticity.
+We interpreted significant Population effects to indicate genetic variation in trait means and Population $\times$ Treatment interactions to indicate genetic variation in plasticity.
 As mentioned above, we used survival and GLMM models for germination rate and mortality, respectively.
@@
 We determined significant fixed effect terms using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-32 \citep{Kuznetsova_etal_2016}.
-This package uses Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests.
+We used Satterthwaite's approximation to calculate denominator degrees of freedom for $F$-tests.
 We also included days to germination as a covariate in growth analyses.
 To ensure that Population and Treatment effects were specific to a particular growth phase, we included germination day as a covariate in leaf expansion and stem elongation analyses.
+
+Failure to detect a significant effect could be the result of Type-2 error, so we complemented step-wise ANOVA (see above) by comparing effect sizes calculated in the full model.
+The full model contains all main effects, two-way interactions (Population $\times$ Temperature, Population $\times$ Water), a three-way interaction (Population $\times$ Temperature $\times$ Water), and random effects.
+For linear mixed-effects models (leaf expansion, stem elongation, and photosynthesis) we used mean-squared error as a measure of effect size; for GLMM (mortality) we used $\chi^2$ as a measure of effect size.
+We did not include germination rate because no Population $\times$ Treatment effects were estimated.
+The difference in effect size of Population versus Population $\times$ Treatment is:
+
+$$ \Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})} = \textrm{Effect Size}_\textrm{Pop} - \textrm{Effect Size}_{\textrm{Pop} \times \textrm{Trt}} $$
+
+We calculated $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ for all two- and three-way Population $\times$ Treatment interactions for each trait.
+To determine whether $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ was significantly different than 0, we calculated 95\% confidence intervals using 1000 parametric bootstrap samples simulated from fitted models.
+If the 95\% confidence interval for a given $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$, was greater than zero then we concluded that the Population effect size was significantly larger than the Population $\times$ Treatment effect size, and vice versa if the confidence internval was less than zero.
+If the confidence spanned zero, then the effect sizes are not significantly different.
 
@@
 We kept only variables selected for prediction, the most stringent criterion.
-A visual overview of how we selected climatic variables is depicted in Fig~\ref{fig:Fig_VarSelect}.
+We visually depict how we selected climatic variables in Fig~\ref{fig:Fig_VarSelect}.
 
@@
 Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}).
-Complete ANOVA tables are available in the Supporting Information (\Cref{table:TableS_fitLER,table:TableS_fitSER,table:TableS_fitPhoto,table:TableS_fitMort})
+Complete ANOVA tables are available in the Supporting Information (\Cref{table:TableS_fitLER,table:TableS_fitSER,table:TableS_fitPhoto,table:TableS_fitMort}).
+
+The effect size of Population was significantly larger than that for Population $\times$ Treatment interactions (Fig.~\ref{fig:FigS_EffectSizes}) in most cases.
+For leaf expansion, Population had a signicantly larger effect size than Population $\times$ Treatment interactions in 2 of 3 comparisons (Fig.~\ref{fig:FigS_EffectSizes}A).
+For stem elongation (Fig.~\ref{fig:FigS_EffectSizes}B) and mortality (Fig.~\ref{fig:FigS_EffectSizes}D), Population effect sizes were significantly larger than all Population $\times$ Treatment interactions.
+For Photosynthesis, Population and Population $\times$ Treatment effect sizes were not significantly different (Fig.~\ref{fig:FigS_EffectSizes}C), presumably because we had a smaller sample size.
 
@@
 cardinalis} occurrences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
-All 16 Climate-Latitude and 3 Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
+All 16 Climate-Latitude and 4 Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
 In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:Fig_LocalvNeighCorr}).
@@
 = 36.8, $P = 7.22 \times 10 ^ {-8}$).
-
 Among Climate-Latitude and Climate-TraitPC1 variables, neighborhood climatic variability over 30 years (1981--2010) in either winter precipitation (bio16$_\sigma$) and/or temperature (bio11$_\sigma$) are the strongest candidates to explain the latitudinal cline in \textit{E.
@@
 Latitudinal clines in germination rate, photosynthesis, and growth suggest adaptive differentiation in important physiological traits of the species.
-However, we found little evidence that populations respond differently to temperature or drought.
+However, we caution that we cannot yet rule nonadaptive demographic processes such as a recent range expansion toward higher latitude \citep{Paul_etal_2016, Sheth_Angert_2017}.
+In contrast, we found little evidence for variation in plasticity to temperature or drought.
 Due to low replication within families, we did not have power to assess within-population genotype-by-environment interactions, which may be present.
 As we discuss below, low variation in plasticity among populations may indicate that some dimensions of the fundamental abiotic niche are relatively conserved.
-Note that statistical power to detect significant plasticity is lower than that for intrinsic differences.
+Note that statistical power to detect significant plasticity is lower than that for differences in trait means.
 However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain low variation in plasticity.
 Finally, our results suggest that neighborhood-scale climate and interannual variation are more important selective agents than local averages.
-In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why intrinsic variation in physiology changes clinally, whereas plastic responses to temperature and drought are relatively static.
+In the paragraphs that follow, we tie these results into the broader threads of evolutionary theory that might help explain why variation in physiological trait means changes clinally, whereas plastic responses to temperature and drought are relatively static.
 One caveat to bear in mind is that we are limited by the size of the climate grid ($\approx 90~\text{m}^2$) and therefore unable to detect very fine-scale local adaptation.

### 2017-09-03 ms/ms.Rnw
version for biorxiv

@@
 \geometry{letterpaper}
-\usepackage[parfill]{parskip}      		% Activate to begin paragraphs with an empty line rather than an indent
+\usepackage[parfill]{parskip}
 \usepackage{graphicx}
@@
   pathMS <- getwd() # for sweave
-  # pathMS <- "ms" # for debugging
 
@@
   source("functions.R") # for sweave
-  # source(paste(pathMS, "/functions.R", sep = "")) # for debugging
 
@@
 \title{Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}}
-\author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
+\author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^{1,2}$}
 \date{}
@@
 
-$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada \\
-$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca \\
+$^1$ Biodiversity Research Centre and Department of Botany, University of British Columbia, Vancouver, BC, Canada \\
+$^2$ Department of Zoology, University of British Columbia, Vancouver, BC, Canada \\
+$^*$corresponding author: Chris Muir, chrisdmuir@gmail.com \\
 
@@
 
-Data will be archived on Dryad upon publication. \\
+Data are currently available on GitHub and will be archived on Dryad upon publication. \\
 
@@
 ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
-Two anonymous referees provided constructive comments on earlier versions of this manuscript.
+Four anonymous referees provided constructive comments on earlier versions of this manuscript.
 
@@
 In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment.
-A related problem is identifying which of the myriad and often covarying aspects of the environment causes spatially varying selective pressures.
+A related problem is identifying which of the myriad and often covarying aspects of the environment cause spatially varying selective pressures.
 
@@
 These hypotheses are not mutually exclusive since, for example, single or multiple factors in a climatic neighborhood may lead to latitudinal clines.
-We focused on climate because climate often determines and where species are found and also can exert strong selection on populations within species, though we acknowledge that other abiotic and biotic factors could also contribute to selection and the overall pattern of local adaptation.
-There is also a compelling need to know how populations are (or are not) locally adapted to climate so as to predict how they will respond to climate change \citep{Aitken_Whitlock_2013}.
+We focused on climate because climate often determines where species are found and also can exert strong selection on populations within species.
+We acknowledge that other abiotic and biotic factors could contribute to selection and the overall pattern of local adaptation.
+Furthermore, there is a compelling need to know how populations are (or are not) locally adapted to climate so as to predict how they will respond to climate change \citep{Aitken_Whitlock_2013}.
 
@@
 During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to the average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$) so that all plants within a temperature level could be measured under the same conditions.
-We were photosynthesis after dry down had progressed to assess differences in photosynthetic responses to drought.
+We measured photosynthesis after dry down had progressed to assess differences in photosynthetic responses to drought.
 
@@
 We looked for climate variables that explained both the latitude of \Sexpr{2 * nrow(thinnedOccClimate)} \textit{E.
-cardinalis} occurrences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
+cardinalis} occurrences (`Climate-Latitude variables') and traits (`Climate-Trait variables').
 For Climate-Latitude variables we extracted climate data from recent occurrences located throughout California and Oregon, USA (shown in map).
@@
 The effect size of Population was significantly larger than that for Population $\times$ Treatment interactions (Fig.~\ref{fig:FigS_EffectSizes}) in most cases.
-For leaf expansion, Population had a signicantly larger effect size than Population $\times$ Treatment interactions in \Sexpr{effSize["larger", "LER"]} of \Sexpr{sum(effSize[, "LER"])} comparisons (Fig.~\ref{fig:FigS_EffectSizes}A).
+For leaf expansion, Population had a significantly larger effect size than Population $\times$ Treatment interactions in \Sexpr{effSize["larger", "LER"]} of \Sexpr{sum(effSize[, "LER"])} comparisons (Fig.~\ref{fig:FigS_EffectSizes}A).
 For stem elongation (Fig.~\ref{fig:FigS_EffectSizes}B) and mortality (Fig.~\ref{fig:FigS_EffectSizes}D), Population effect sizes were significantly larger than all Population $\times$ Treatment interactions.
@@
 Latitudinal clines in germination rate, photosynthesis, and growth suggest adaptive differentiation in important physiological traits of the species.
-However, we caution that we cannot yet rule nonadaptive demographic processes such as a recent range expansion toward higher latitude \citep{Paul_etal_2016, Sheth_Angert_2017}.
+However, we caution that these are candidate adaptive traits and that we cannot yet rule out nonadaptive demographic processes such as a recent range expansion toward higher latitude \citep{Paul_etal_2016, Sheth_Angert_2017}.
 In contrast, we found little evidence for variation in plasticity to temperature or drought.
@@
 As we discuss below, low variation in plasticity among populations may indicate that some dimensions of the fundamental abiotic niche are relatively conserved.
-Note that statistical power to detect significant plasticity is lower than that for differences in trait means.
-However, the fact that the Population and Temperature effects were often highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain low variation in plasticity.
+Note that statistical power to detect significant plasticity is lower than that for differences in trait means, but the effect size of variation in plasticity was significantly less than that for trait means in most cases (Fig.~\ref{fig:FigS_EffectSizes}).
 Finally, our results suggest that neighborhood-scale climate and interannual variation are more important selective agents than local averages.
@@
 If this variation is adaptive, it suggests one of several possibilities to investigate in the future: the fitness tradeoff between low versus high latitude environments is not too strong nor swamped by demographic asymmetry or maladaptive gene flow.
-That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatially heterogenous selection.
+That is, alleles favoured at one latitude are not strongly selected against when they flow to another population, allowing locally adaptive genetic variation to be maintained by spatially heterogeneous selection.
 We also know from previous work that population size does not vary strongly with latitude (Angert, unpub.
 data).
-Gene flow appears to be high, but attenuates at broad spatial scales, especially between Southern ($<35\degree$N) and Northern portions of the range \citep{Paul_etal_2016}.
+Gene flow appears to be high, but attenuates at broad spatial scales, especially between southern ($<35\degree$N) and northern portions of the range \citep{Paul_etal_2016}.
 
@@
 It is reasonable to predict that southern populations, which appear to experience more frequent drought years (see below), might have physiological adaptations to respond to drought stress to survive and grow in drier soil.
-We found no evidence for this type of drought tolerance; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
+We found little evidence for this type of drought tolerance; all populations responded to drought and temperature similarly (Table~\ref{table:Table_AnovaSummary}).
 Plants grew faster in the Hot treatment, but there was little effect of drought on growth.
@@
 cardinalis} reveal that southern populations tend to flower earlier at a smaller size, while northern populations invest more in vegetative growth \citep{Sheth_Angert_2017}.
+In this experiment, the fastest growing plants began producing flowers in $\sim$ 60 days (data not shown), suggesting that rapid vegetative develop may likewise affect flowering time.
 The association between position along the `fast-slow' continuum and associated traits in \textit{E.
@@
 We did not find evidence that the relative performance of populations shifts with temperature or watering regime, suggesting relatively little variation in plasticity.
-Exploratory analysis implicate that more variable precipitation regimes at lower latitude drive much of the latitudinal cline, though other climatic factors could also contribute.
+Exploratory analysis implicate that more variable precipitation regimes at lower latitude could drive much of the latitudinal cline, though other climatic factors could also contribute.
 Interestingly, the climatic neighborhood may shape selective pressures more than local climate.

### 2017-09-03 ms/ms.tex
version for biorxiv

@@
 We looked for climate variables that explained both the latitude of 356 \textit{E.
-cardinalis} occurrences (`Climate-Latitude variables') and with traits (`Climate-Trait variables').
+cardinalis} occurrences (`Climate-Latitude variables') and traits (`Climate-Trait variables').
 For Climate-Latitude variables we extracted climate data from recent occurrences located throughout California and Oregon, USA (shown in map).
@@
 The effect size of Population was significantly larger than that for Population $\times$ Treatment interactions (Fig.~\ref{fig:FigS_EffectSizes}) in most cases.
-For leaf expansion, Population had a signicantly larger effect size than Population $\times$ Treatment interactions in 2 of 3 comparisons (Fig.~\ref{fig:FigS_EffectSizes}A).
+For leaf expansion, Population had a significantly larger effect size than Population $\times$ Treatment interactions in 2 of 3 comparisons (Fig.~\ref{fig:FigS_EffectSizes}A).
 For stem elongation (Fig.~\ref{fig:FigS_EffectSizes}B) and mortality (Fig.~\ref{fig:FigS_EffectSizes}D), Population effect sizes were significantly larger than all Population $\times$ Treatment interactions.
@@
 cardinalis} occurrences (Fig.~\ref{fig:Fig_ClimVarImp}, Table~\ref{table:TableS_ClimVarImp}).
-All 16 Climate-Latitude and 4 Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
+All 16 Climate-Latitude and 3 Climate-TraitPC1 variables were neighborhood rather than local variables (Fig.~\ref{fig:Fig_ClimVarImp}).
 In fact, neighborhood climate almost always correlated better with TraitPC1 and Latitude than local climate (Fig.~\ref{fig:Fig_LocalvNeighCorr}).

### 2017-09-24 ms/ms.Rnw
made minor revisions for JEB

@@
 
+<<focClim, echo = FALSE, results=hide, warning=FALSE>>=
+
+# Read in climate data for 16 focal populations to use in making Table 1
+load(paste(pathDat, "/focClim.RData", sep = ""))
+
+@
+
 \textit{E.
 cardinalis} is a perennial forb native to the Western US (California and Oregon).
-It is predomintantly outcrossing, self-compatible, and pollinated primarily by hummingbirds.
+It is predominantantly outcrossing, self-compatible, and pollinated primarily by hummingbirds.
 We used 16 populations from throughout the range of \textit{E.
@@
 These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below).
-Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
+Although the elevation range of \textit{E.
+cardinalis} compresses slightly from south to north, latitude is not strongly confounded by elevation among our focal populations.
+The correlation between Latitude and Elevation among the 16 focal populations was weak ($r = \Sexpr{round(cor(focClim$Latitude, focClim$Elevation), 2)}$, $P = \Sexpr{round(cor.test(focClim$Latitude, focClim$Elevation)$p.value, 2)}$).
+Seeds were collected in the field from mature, undehisced fruits left open for 2-4 weeks to dry, then stored at room temperature.
 To control for maternal effects, we grew a large number of field-derived seeds in the greenhouse and generated seed families for this experiment by haphazardly crossing individuals from the same population.
@@
 Thus, we used seeds from \Sexpr{nlevels(master$Family)} greenhouse-derived seed families, \Sexpr{min(famTab)}--\Sexpr{max(famTab)} (mean = \Sexpr{round(mean(famTab), 1)}, median = \Sexpr{median(famTab)}) families per population.
-
-<<focClim, echo = FALSE, results=hide, warning=FALSE>>=
-
-# Read in climate data for 16 focal populations to use in making Table 1
-load(paste(pathDat, "/focClim.RData", sep = ""))
-
-@
 
@@
 We thinned seedlings to one plant per cone-tainer, leaving the center-most plant.
-\Sexpr{nrow(master)} of 768 (\Sexpr{round(100 * nrow(master) / 768, 1)}\%) had plants that could be used in the experiment.
+\Sexpr{nrow(master)} of 768 (\Sexpr{round(100 * nrow(master) / 768, 1)}\%) cone-tainers had plants that could be used in the experiment.
 We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments.
@@
   set.seed(69648)
-  vsurfPC1 <- VSURF(focClim[, climVars], focClim[, "PC1"], parallel = TRUE, ncores = 3, 
-                    ntree = 1e4)
+  vsurfPC1 <- VSURF(focClim[, climVars], focClim[, "PC1"], parallel = TRUE, 
+                    ncores = 3, ntree = 1e4)
 
@@
 par(mai = c(xLabHeight, yLabWidth, xLabHeight, 2 * panelGap), ann = F, yaxs = "i")
-layout(matrix(1:2, ncol = 1), widths = rep(1, 2), heights = lcm(2.54 * c(heightA, heightB)))
+layout(matrix(1:2, ncol = 1), widths = rep(1, 2), 
+       heights = lcm(2.54 * c(heightA, heightB)))
 
@@
   # Panel C. bio15_cva_SA vs bio16_cva_SA
-  plot(focClim$bio16_cva_SA, focClim$bio15_cva_SA, xlim = c(0.25, 0.55), ylim = c(0.14, 0.24),
-       axes = F, frame.plot = T, type = "n", xlab = "", ylab = "")
+  plot(focClim$bio16_cva_SA, focClim$bio15_cva_SA, xlim = c(0.25, 0.55), 
+       ylim = c(0.14, 0.24), axes = F, frame.plot = T, type = "n", 
+       xlab = "", ylab = "")
   text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
@@
   plot(thinnedOccClimate$Latitude, thinnedOccClimate$bio16_cva_SA, 
-       xlim = c(32, 44), ylim = c(0.25, 0.55), axes = F, frame.plot = T, type = "n", 
-       xlab = "", ylab = "")
+       xlim = c(32, 44), ylim = c(0.25, 0.55), axes = F, frame.plot = T, 
+       type = "n", xlab = "", ylab = "")
   text(grconvertX(0.075, "npc", "user"), grconvertY(0.925, "npc", "user"),
@@
 
-  pdf(paste(pathFig, "/Figure_LocalvNeighCorr.pdf", sep = ""), w = 8, h = 4, useDingbats = F)
+  pdf(paste(pathFig, "/Figure_LocalvNeighCorr.pdf", sep = ""), w = 8, h = 4,
+      useDingbats = F)
   par(mai = c(1, 1, 0.2, 0.2), mfrow = c(1, 2))
@@
 
-    plot(corTab[, "local"], corTab[, "neighborhood"], xlim = c(0, 1), ylim = c(0, 1),
-         xlab = "", ylab = "", type = "n", axes = F, frame.plot = T)
+    plot(corTab[, "local"], corTab[, "neighborhood"], xlim = c(0, 1), 
+         ylim = c(0, 1), xlab = "", ylab = "", type = "n", axes = F, frame.plot = T)
     abline(0, 1, lty = 2)
@@
     axis(2, lwd = 0, lwd.ticks = 1, las = 1)
-    title(xlab = expression(paste("local |", italic(r)[paste(climate-TraitPC1)], "|")),
-          ylab = expression(paste("neighborhood |", italic(r)[paste(climate-TraitPC1)], "|")),
+    title(xlab = expression(paste("local |", italic(r)[paste(climate-TraitPC1)], 
+                                  "|")),
+          ylab = expression(paste("neighborhood |", 
+                                  italic(r)[paste(climate-TraitPC1)], "|")),
           cex.lab = 1.5)
@@
 
-    legend(1, 0, xjust = 1, yjust = 0, legend = c("Avg.", "Var."), pch = 21, col = "black", 
-           pt.bg = c("black", "white"))
+    legend(1, 0, xjust = 1, yjust = 0, legend = c("Avg.", "Var."), pch = 21, 
+           col = "black", pt.bg = c("black", "white"))
     text(0.975, 0.95, labels = "1:1", pos = 1)
-    text(0 + strwidth("A", cex = 2, font = 2) / 2, 1 - strheight("A", cex = 2, font = 2) / 2, 
+    text(0 + strwidth("A", cex = 2, font = 2) / 2, 1 - 
+           strheight("A", cex = 2, font = 2) / 2, 
          labels = "A", font = 2, cex = 2)
@@
     # Matrix comparing correlation coefficients of local and neighborhood climate
-    corTab <- as.data.frame(cbind(r[!grepl("_SA", climVars)], r[grepl("_SA", climVars)]))
+    corTab <- as.data.frame(cbind(r[!grepl("_SA", climVars)], 
+                                  r[grepl("_SA", climVars)]))
     corTab <- abs(corTab)
@@
     axis(2, lwd = 0, lwd.ticks = 1, las = 1)
-    title(xlab = expression(paste("local |", italic(r)[paste(climate-latitude)], "|")),
-          ylab = expression(paste("neighborhood |", italic(r)[paste(climate-latitude)], "|")),
+    title(xlab = expression(paste("local |", 
+                                  italic(r)[paste(climate-latitude)], "|")),
+          ylab = expression(paste("neighborhood |", 
+                                  italic(r)[paste(climate-latitude)], "|")),
           cex.lab = 1.5)
@@
 
-    legend(1, 0, xjust = 1, yjust = 0, legend = c("Avg.", "Var."), pch = 21, col = "black", 
-           pt.bg = c("black", "white"))
+    legend(1, 0, xjust = 1, yjust = 0, legend = c("Avg.", "Var."), pch = 21, 
+           col = "black", pt.bg = c("black", "white"))
     text(0.975, 0.95, labels = "1:1", pos = 1)
-    text(0 + strwidth("B", cex = 2, font = 2) / 2, 1 - strheight("B", cex = 2, font = 2) / 2, 
+    text(0 + strwidth("B", cex = 2, font = 2) / 2, 1 - 
+           strheight("B", cex = 2, font = 2) / 2, 
          labels = "B", font = 2, cex = 2)
@@
 
+An alternative, nonbiological, explanation for why neighborhood climate correlates better than local climate is that averaging climate over a wider area integrates out error in ClimateWNA values.
+ClimateWNA interpolates between weather stations using a fine-scale digital elevation model to accurately downscale climate data \citep{Wang_etal_2012}, but there is obviously some error.
+If the error were randomly distributed about the true climatic values, then averaging climate over a larger sample of points could give a more precise estimate, which might explain why traits and latitude correlate better with climatic neighbourhood.
+For this technical artefact to be important, the measurement error in local climate would have to be large relative to true variation in climate among locations within the climatic neighbourhoods.
+We do not know the relative magnitude of these factors, but future work should compare ClimateWNA-predicted local climate with actual plant-level measurements to estimate the magnitude of this error.
+
 It is reasonable to predict that southern populations, which appear to experience more frequent drought years (see below), might have physiological adaptations to respond to drought stress to survive and grow in drier soil.

### 2017-09-24 ms/ms.tex
made minor revisions for JEB

@@
 cardinalis} is a perennial forb native to the Western US (California and Oregon).
-It is predomintantly outcrossing, self-compatible, and pollinated primarily by hummingbirds.
+It is predominantantly outcrossing, self-compatible, and pollinated primarily by hummingbirds.
 We used 16 populations from throughout the range of \textit{E.
@@
 These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below).
-Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.
+Although the elevation range of \textit{E.
+cardinalis} compresses slightly from south to north, latitude is not strongly confounded by elevation among our focal populations.
+The correlation between Latitude and Elevation among the 16 focal populations was weak ($r = -0.25$, $P = 0.34$).
+Seeds were collected in the field from mature, undehisced fruits left open for 2-4 weeks to dry, then stored at room temperature.
 To control for maternal effects, we grew a large number of field-derived seeds in the greenhouse and generated seed families for this experiment by haphazardly crossing individuals from the same population.
@@
 We thinned seedlings to one plant per cone-tainer, leaving the center-most plant.
-702 of 768 (91.4\%) had plants that could be used in the experiment.
+702 of 768 (91.4\%) cone-tainers had plants that could be used in the experiment.
 We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments.

### 2017-10-13 ms/ms.Rnw
release 1

@@
 
-% Make helvetica the default sans-serif font
-% \renewcommand\sfdefault{phv}
-
 % Package command for citing R packages
@@
 
-Data are currently available on GitHub and will be archived on Dryad upon publication. \\
-
-%--------------------------------------------------
-% Acknowledgements
-%--------------------------------------------------
-
-\section*{Acknowledgements}
-Erin Warkman and Lisa Lin helped collect data.
-CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program.
-ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
-Four anonymous referees provided constructive comments on earlier versions of this manuscript.
+Data are currently available on GitHub and Dryad \citep{Muir_Angert_2017} \\
 
 \clearpage
-
-\setstretch{\stretchy}
-
-\begin{center}
-{\LARGE  \centering Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}}
-\end{center}
 
@@
 % Toggle these commands to get version for doing word count
-\linenumbers
+% \linenumbers
 % \pagenumbering{gobble}
@@
 
-% Local adaptation and clines
 Local adaptation has been documented within numerous species; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Leimu_Fischer_2008, Hereford_2009}.
@@
 
-% Intrinsic versus plastic variation 
 When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}.
@@
 Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive.
-We distinguish between these patterns of adaptive trait differences by referring to `genetic variation' in trait means and `genetic variation in plasticity', respectively.
+We distinguish between these patterns of adaptive trait differences by referring to `genetic variation in trait means' and `genetic variation in plasticity', respectively.
 Genetic variation in plasticity is synonymous with genotype-by-environment interactions, or simply (G$\times$E).
@@
 
-% Selective agents (trait-enviro assoc)
 A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments.
@@
 # Legend
-# pdf(paste(pathFig, "/ExperimentalDesignLegend.pdf", sep = ""), 2, 2)
 par(mai = c(2, 0, 1, 0))
@@
 To determine whether $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ was significantly different than 0, we calculated 95\% confidence intervals using 1000 parametric bootstrap samples simulated from fitted models.
-If the 95\% confidence interval for a given $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$, was greater than zero then we concluded that the Population effect size was significantly larger than the Population $\times$ Treatment effect size, and vice versa if the confidence internval was less than zero.
-If the confidence spanned zero, then the effect sizes are not significantly different.
+If the 95\% confidence interval for a given $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$, was greater than zero then we concluded that the Population effect size was significantly larger than the Population $\times$ Treatment effect size, and vice versa if the confidence interval was less than zero.
+If the confidence interval spanned zero, then the effect sizes are not significantly different.
 
@@
 
-% This hypothesis makes several independent, testable predictions.
-The allocation tradeoff predicts that northern populations will provision more photosynthetic assimilate to rhizomes compared with southern populations.
-If southern populations are indeed more `annualized' because more frequent droughts cause mortality, then we predict that species distribution models using recent climate would best predict occurrences in the south, whereas longer term climate would be a better predictor in the north.
-Finally, we predict that southern populations would show greater variation in the size of recruits and higher maximum population growth rates.
-
 In summary, we found evidence for a coordinated latitudinal cline in germination rate, photosynthesis, and growth, suggesting local adaptation.
@@
 By doing so, we aim to understand why certain physiological and developmental mechanisms, but not others, contribute to local adaptation.
+
+%--------------------------------------------------
+% Acknowledgements
+%--------------------------------------------------
+
+\section*{Acknowledgements}
+Erin Warkman and Lisa Lin helped collect data.
+CDM was supported by a Biodiversity Postdoctoral Fellowship funded by the NSERC CREATE program.
+ALA was supported by an NSERC Discovery Grant and a grant from the National Science Foundation (DEB 0950171).
+Four anonymous referees provided constructive comments on earlier versions of this manuscript.
 

### 2017-10-13 README.Rmd
test

@@
 # card-cline
-Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in *Erythranthe cardinalis*
+Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in *Erythranthe cardinalis*. In Press at *Journal of Evolutionary Biology*.
 
@@
 
-## Downloading data and code
+## Downloading repository
 
@@
 2. Open `card-cline.Rproj` in [RStudio](https://www.rstudio.com/)
+3. Install R packages if necessary.
 

### 2017-10-13 README.Rmd
Update readme

@@
 # card-cline
-Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in *Erythranthe cardinalis*. In Press at *Journal of Evolutionary Biology*.
+[Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in *Erythranthe cardinalis*](https://dx.doi.org/10.1111/jeb.13184).
+*Journal of Evolutionary Biology*.
 
